### 2019-01-08 doc/manuscript/paper.tex
Starting the paper

\documentclass[11 pts]{article}

\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{commath}

\usepackage[font={small}]{caption}
\usepackage{longtable}

\newcommand{\todo}[1]{\textbf{[TODO\@: #1]}}
\newcommand{\note}[1]{\textbf{[NOTE\@: #1]}}
\newcommand{\fixme}[1]{\textbf{[FIXME\@: #1]}}

\begin{document}
\title{A pipeline to analyse transcriptomic time-course data}
\maketitle

\begin{abstract}

The phenotypic diversity of cells is governed by a complex equilibrium between
their genetic identity and their environmental interactions, and understanding
the dynamics of gene expression is a fundamental question of biology. However,
analysing time-course transcriptomic data raises challenging statistical and
computational questions, requiring the development of novel methods and
software. We present a case study of the lifetime of S. bicolor under
droughted and watered condition using an integrated workflow providing a
step-by-step tutorial to the methodology and associated software for the
following four main tasks.

Novel single-cell transcriptome sequencing assays allow researchers to measure
gene expression levels at the resolution of single cells and offer the
unprecendented opportunity to investigate at the molecular level fundamental
biological questions, such as stem cell differentiation or the discovery and
characterization of rare cell types. However, such assays raise challenging
statistical and computational questions and require the development of novel
methodology and software. Using stem cell differentiation in the mouse
olfactory epithelium as a case study, this integrated workflow provides a
step-by-step tutorial to the methodology and associated software for the
following four main tasks: (1) dimensionality reduction accounting for zero
inflation and over dispersion and adjusting for gene and cell-level
covariates; (2) cell clustering using resampling-based sequential ensemble
clustering; (3) inference of cell lineages and pseudotimes; and (4)
differential expression analysis along lineages
\end{abstract}

\end{document}

### 2019-02-07 scripts/shoemaker2015.Rmd
Adding Rmd file containing shoemaker analysis

---
title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
    %\VignetteIndexEntry{Title of your vignette}
      %\usepackage[UTF-8]{inputenc}
---

```{r}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)

# library(moanin)
load_all("../bin/moanin")
```

```{r}
# Set options
df = 6

# Preprocessing & filtering
take_log = FALSE    
filter_expression = FALSE

# Differential expression analysis
timecourse_contrasts = c("C-K", "C-M")

# Clustering
filter_genes = FALSE
n_clusters = 20
percentage_genes_to_label = 0.5
```

```{r}
# Set data files and options
data("shoemaker2015")
data = shoemaker2015$data
meta = shoemaker2015$meta
```

# Normalization

```{r}
# Normalize
data_set = EDASeq::newSeqExpressionSet(as.matrix(data), phenoData=meta)
data_normalized = EDASeq::betweenLaneNormalization(
    data_set, which="upper", offset=TRUE, round=FALSE)
data_normalized = EDASeq::normCounts(data_normalized)
```

```{r}

```

# Differential expression analysis

## Time-course differential expression analysis between two groups

```{r}
# Define contrast
de_analysis = data.frame(row.names=row.names(data))
for(contrast_formula in timecourse_contrasts){
    contrast = limma::makeContrasts(
	contrast_formula,
	levels=levels(meta$Group))
    model = edgeWithContrasts(as.matrix(data), meta, contrast)
    contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
    contrast_name_pval = paste(contrast_name, "-pval", sep="")
    contrast_name_qval = paste(contrast_name, "-qval", sep="")
    de_analysis[contrast_name_pval] = model$pval
    de_analysis[contrast_name_qval] = model$qval
}
```

## Developmental differential expression analysis

```{r}
# Define contrast
contrast_formula = "C"
contrast = limma::makeContrasts(
    contrast_formula,
    levels=levels(meta$Group))
# Why do I have to do this again?
mask = with(meta, Group == contrast_formula)
model = edgeWithContrasts(
    as.matrix(data), meta, contrast, developmental=TRUE,
    mask=mask)
```

## Weekly differential expression analysis

```{r}
# Define contrast
contrasts = c("C.0-K.0",
	      "C.3-K.3",
	      "C.6-K.6",
	      "C.9-K.9",
	      "C.12-K.12",
	      "C.18-K.18",
	      "C.24-K.24",
	      "C.30-K.30",
	      "C.48-K.48",
	      "C.60-K.60",
	      "C.120-K.120",
	      "C.168-K.168")

meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
# de_analysis =  fit_weekly_analysis(
#     data, meta, contrasts,
#     use_voom_weights=FALSE)
```

# Clustering

```{r}
if(filter_genes){
    # Filter out q-values for the pvalues table
    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
    pvalues = pvalues[, col_to_keep]
    fishers_pval = fisher_method(pvalues)
    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
    y = as.matrix(data[genes_to_keep, ])
}else{
    y = as.matrix(data)
}

###############################################################################
# Fit the splines

full_model = ~Group:ns(Time, df=df) + Group + 0
X = model.matrix(full_model, data=meta)

fitted_data = fit_predict_splines(y, X)
rescaled_fitted_data = rescale_values(fitted_data, meta)

# FIXME should use a package with a smarter initialization
kmeans_clusters = kmeans(rescaled_fitted_data, n_clusters, nstart=10)
all_scores = data.frame(row.names=row.names(data))
for(k in 1:n_clusters){
    scores = score_genes_centroid(data, kmeans_clusters$centers[k,])
    all_scores[paste0("cluster_", k)] = scores
}

# Only assign labels to X% of the genes
max_score = quantile(row_min(all_scores), c(percentage_genes_to_label))
genes_to_not_consider = row_min(all_scores) >= max_score
labels = row_argmin(all_scores)
labels[genes_to_not_consider] = NA
```

# Downstream analysis

```{r}

```

### 2019-02-11 scripts/shoemaker2015.Rmd
MAINT Adding TODO list

@@
 # Clustering
-filter_genes = FALSE
+filter_genes = TRUE
+n_genes_to_keep = 5000
 n_clusters = 20
@@
 
+FIXME this is not the normalization to use (this is the one for the EPICON
+data)
 ```{r}
@@
 
-```{r}
-
-```
-
 # Differential expression analysis
@@
 ## Time-course differential expression analysis between two groups
+
+FIXME add estimation of log fold change.
 
@@
 }
+
+pvalues = de_analysis
 ```
@@
 
+FIXME should use a package with a smarter initialization (kmeans++)
+
 ```{r}
@@
 
-# FIXME should use a package with a smarter initialization
 kmeans_clusters = kmeans(rescaled_fitted_data, n_clusters, nstart=10)

### 2019-02-11 bin/moanin/R/de_analysis.R

@@ -4,6 +4,7 @@ library("edge")
 library("splines")
 library("MASS")
 
+
 center_data = function(y, ng_labels){
   for(g in levels(ng_labels)){
     whKeep = which(ng_labels == g)
@@ -88,7 +89,11 @@ lrtStat = function(resNull, resFull, ng_labels=NULL) {
   return(stat)
 }
 
-compute_pvalue = function(X, y, beta, beta_null, ng_labels, statistics="lrt",
+compute_pvalue = function(X, y, beta, beta_null, ng_labels,
+			  n_groups=NULL,
+			  n_samples=NULL,
+			  degrees_of_freedom=NULL,
+			  statistics="lrt",
 			  df2=NULL, weights=NULL, mask=NULL, developmental=FALSE){
 
     fitFull = beta %*% t(X)
@@ -115,19 +120,30 @@ compute_pvalue = function(X, y, beta, beta_null, ng_labels, statistics="lrt",
       resFull = y - fitFull
     }
 
-    ng = nlevels(ng_labels)
-    df = ncol(X) / ng
+    # estimate degrees of freedom.
+    if(is.null(n_groups)){
+        n_groups = nlevels(ng_labels)
+	# FIXME Raise warning
+    }
+
+    if(is.null(n_samples)){
+	n_samples = ncol(X)
+	# FIXME raise warning
+    }
+    
+    df = degrees_of_freedom
 
     if(statistics == "ftest"){
-      stat = lrtStat(resNull, resFull)
-      if(is.null(df2)){
-        df2 = nrow(X) - ncol(X)
-      }
-      df1 = df
-      pval = pf(stat * df2 / df1, df1=df1, df2=df2, lower.tail=FALSE)
+	stat = lrtStat(resNull, resFull)
+	if(is.null(df2)){
+	    # FIXME Check this.
+	    df2 = n_samples - degrees_of_freedom * n_groups
+	}
+	df1 = df
+	pval = pf(stat * df2 / df1, df1=df1, df2=df2, lower.tail=FALSE)
     }else{
-      lstat = lrtStat(resNull, resFull, ng_labels=ng_labels)
-      pval = pchisq(lstat, df=df, lower.tail=FALSE)
+	lstat = lrtStat(resNull, resFull, ng_labels=ng_labels)
+	pval = pchisq(lstat, df=degrees_of_freedom, lower.tail=FALSE)
     }
     return(pval)
 }
[…]

### 2019-02-14 scripts/shoemaker2015.Rmd
MAINT using better clustering algorithm

@@
----
 title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
 output: rmarkdown::html_vignette
-vignette: >
-  %\VignetteEngine{knitr::knitr}
-    %\VignetteIndexEntry{Title of your vignette}
-      %\usepackage[UTF-8]{inputenc}
+
 ---
 
-```{r}
+```{r echo=TRUE, results="hide"}
 # Loading dependencies
@@
 library(stats)
+library(moanin)
 
 # library(moanin)
-load_all("../bin/moanin")
 ```
 
-```{r}
+```{r echo=TRUE}
 # Set options
@@
 # Preprocessing & filtering
+is_count = FALSE
 take_log = FALSE    
@@
 
-```{r}
+```{r echo=FALSE}
 # Set data files and options
@@
 meta = shoemaker2015$meta
+
+# Right now, the package doesn't contain the normalized data
 ```
@@
 data)
-```{r}
+
+```{r normalization}
+if(is_count){
     # Normalize
@@
     data_normalized = EDASeq::normCounts(data_normalized)
+}
 ```
@@
 
-```{r}
+```{r de_analysis}
 # Define contrast
@@
 	levels=levels(meta$Group))
-    model = edgeWithContrasts(as.matrix(data), meta, contrast)
+    model = moanin:::edgeWithContrasts(as.matrix(data), meta, contrast)
     contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
@@
     contrast_name_qval = paste(contrast_name, "-qval", sep="")
-    de_analysis[contrast_name_pval] = model$pval
-    de_analysis[contrast_name_qval] = model$qval
+    de_analysis[contrast_name_pval] = model$pval_lrt
+    de_analysis[contrast_name_qval] = model$qval_lrt
 }
@@
 mask = with(meta, Group == contrast_formula)
-model = edgeWithContrasts(
+model = moanin:::edgeWithContrasts(
     as.matrix(data), meta, contrast, developmental=TRUE,
@@
 
-FIXME should use a package with a smarter initialization (kmeans++)
-
 ```{r}
@@
     pvalues = pvalues[, col_to_keep]
-    fishers_pval = fisher_method(pvalues)
+    fishers_pval = moanin:::fisher_method(pvalues)
     genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
@@
 
-full_model = ~Group:ns(Time, df=df) + Group + 0
-X = model.matrix(full_model, data=meta)
+kmeans_clusters = moanin::splines_kmeans(
+    y, meta, n_clusters=20,
+    random_seed=42,
+    degrees_of_freedom=df)
 
-fitted_data = fit_predict_splines(y, X)
-rescaled_fitted_data = rescale_values(fitted_data, meta)
-
-kmeans_clusters = kmeans(rescaled_fitted_data, n_clusters, nstart=10)
+# Rescale the centroids
+centroids = moanin::rescale_values(kmeans_clusters$centroids, meta)
 all_scores = data.frame(row.names=row.names(data))
 for(k in 1:n_clusters){
-    scores = score_genes_centroid(data, kmeans_clusters$centers[k,])
+    scores = moanin:::score_genes_centroid(data, centroids[k,])
     all_scores[paste0("cluster_", k)] = scores
@@
 # Only assign labels to X% of the genes
-max_score = quantile(row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = row_min(all_scores) >= max_score
-labels = row_argmin(all_scores)
+max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
+genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
+labels = moanin:::row_argmin(all_scores)
 labels[genes_to_not_consider] = NA
+```
+
+```{r}
+print(unique(labels))
+```
+
+```{r}
+moanin::plot_centroids(centroids, meta)
 ```

### 2019-02-16 scripts/00_introduction.Rmd
Splitting the paper into several Rmd files

---
title: "XXX: a pipeline to analyse time-course gene expression data"
output: html_document
---

# Abstract

The phenotypic diversity of cells is governed by a complex equilibrium between
their genetic identity and their environmental interactions, and understanding
the dynamics of gene expression is a fundamental question of biology. However,
analysing time-course transcriptomic data raises challenging statistical and
computational questions, requiring the development of novel methods and
software.

# Introduction

- TODO why is time-course gene expression analysis important for biology. Why
  are novel workflows necessary.

Time-course gene expression analysis workflows include steps used more broadly
for gene expression data analasys such as quality control (QC), exploratory
data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
typical gene expression data analysis workflow is not entirely adequate for
analysing time-course data. In particular, differential expression analysis
needs to be adapted to time-course data, and often needs to be followed by a
clustering step to group genes sharing similar patterns of expression.

Here, we propose an integrated workflow for downstream analysis of time-course
data, with the following main steps: (1) quality control and normalization;
(2) differential expression analysis of time-course data; (3) clustering; (4)
downstream analysis of clusters with GO term and pathway enrichment.

XXX Add the packages used in the workflow

# Analysis of the dynamical response of mouse lung tissue to influenza

## Overview

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to different strains of influenza, and collecting
lung tissue during 12 time-points (..).

### 2019-02-16 scripts/01_quality_control_normalization.Rmd
Splitting the paper into several Rmd files

--- 2019-02-14 scripts/shoemaker2015.Rmd
@@
-title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
+---
+title: "Normalization: A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
-
 ---
 
-```{r echo=TRUE, results="hide"}
+```{r echo=FALSE, results="hide"}
 # Loading dependencies
@@
 
-```{r echo=TRUE}
+```{r echo=FALSE}
 # Set options
@@
 ```
-
-```{r echo=FALSE}
-# Set data files and options
-data("shoemaker2015")
-data = shoemaker2015$data
-meta = shoemaker2015$meta
-
-# Right now, the package doesn't contain the normalized data
-```
-
-# Normalization
-
-FIXME this is not the normalization to use (this is the one for the EPICON
-data)
-
-```{r normalization}
-if(is_count){
-    # Normalize
-    data_set = EDASeq::newSeqExpressionSet(as.matrix(data), phenoData=meta)
-    data_normalized = EDASeq::betweenLaneNormalization(
-	data_set, which="upper", offset=TRUE, round=FALSE)
-    data_normalized = EDASeq::normCounts(data_normalized)
-}
-```
-
-# Differential expression analysis
-
-## Time-course differential expression analysis between two groups
-
-FIXME add estimation of log fold change.
-
-```{r de_analysis}
-# Define contrast
-de_analysis = data.frame(row.names=row.names(data))
-for(contrast_formula in timecourse_contrasts){
-    contrast = limma::makeContrasts(
-	contrast_formula,
-	levels=levels(meta$Group))
-    model = moanin:::edgeWithContrasts(as.matrix(data), meta, contrast)
-    contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
-    contrast_name_pval = paste(contrast_name, "-pval", sep="")
-    contrast_name_qval = paste(contrast_name, "-qval", sep="")
-    de_analysis[contrast_name_pval] = model$pval_lrt
-    de_analysis[contrast_name_qval] = model$qval_lrt
-}
-
-pvalues = de_analysis
-```
-
-## Developmental differential expression analysis
-
-```{r}
-# Define contrast
-contrast_formula = "C"
-contrast = limma::makeContrasts(
-    contrast_formula,
-    levels=levels(meta$Group))
-# Why do I have to do this again?
-mask = with(meta, Group == contrast_formula)
-model = moanin:::edgeWithContrasts(
-    as.matrix(data), meta, contrast, developmental=TRUE,
-    mask=mask)
-```
-
-## Weekly differential expression analysis
-
-```{r}
-# Define contrast
-contrasts = c("C.0-K.0",
-	      "C.3-K.3",
-	      "C.6-K.6",
-	      "C.9-K.9",
-	      "C.12-K.12",
-	      "C.18-K.18",
-	      "C.24-K.24",
-	      "C.30-K.30",
-	      "C.48-K.48",
-	      "C.60-K.60",
-	      "C.120-K.120",
-	      "C.168-K.168")
-
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
-# de_analysis =  fit_weekly_analysis(
-#     data, meta, contrasts,
-#     use_voom_weights=FALSE)
-```
-
-# Clustering
-
-```{r}
-if(filter_genes){
-    # Filter out q-values for the pvalues table
-    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
-    pvalues = pvalues[, col_to_keep]
-    fishers_pval = moanin:::fisher_method(pvalues)
-    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
-    y = as.matrix(data[genes_to_keep, ])
-}else{
-    y = as.matrix(data)
-}
-
-###############################################################################
-# Fit the splines
-
-kmeans_clusters = moanin::splines_kmeans(
-    y, meta, n_clusters=20,
-    random_seed=42,
-    degrees_of_freedom=df)
-
-# Rescale the centroids
-centroids = moanin::rescale_values(kmeans_clusters$centroids, meta)
-all_scores = data.frame(row.names=row.names(data))
-for(k in 1:n_clusters){
-    scores = moanin:::score_genes_centroid(data, centroids[k,])
-    all_scores[paste0("cluster_", k)] = scores
-}
-
-# Only assign labels to X% of the genes
-max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-labels = moanin:::row_argmin(all_scores)
-labels[genes_to_not_consider] = NA
-```
-
-```{r}
-print(unique(labels))
-```
-
-```{r}
-moanin::plot_centroids(centroids, meta)
-```
-
-# Downstream analysis
-
-```{r}
-
-```

### 2019-02-16 scripts/02_differential_expression_analysis.Rmd
Splitting the paper into several Rmd files

--- 2019-02-14 scripts/shoemaker2015.Rmd
@@
-title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
+---
+title: "Normalization: A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
-
 ---
 
-```{r echo=TRUE, results="hide"}
-# Loading dependencies
+## Differential expression analysis of time-course data
+
+```{r echo=FALSE, results="hide"}
+# Reloading necessary data and preliminary results
 library(devtools)
@@
 library(moanin)
-
-# library(moanin)
 ```
 
-```{r echo=TRUE}
-# Set options
-df = 6
-
-# Preprocessing & filtering
-is_count = FALSE
-take_log = FALSE    
-filter_expression = FALSE
-
-# Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M")
-
-# Clustering
-filter_genes = TRUE
-n_genes_to_keep = 5000
-n_clusters = 20
-percentage_genes_to_label = 0.5
-```
-
-```{r echo=FALSE}
+```{r echo=FALSE, results="hide"}
 # Set data files and options
@@
 
-# Normalization
-
-FIXME this is not the normalization to use (this is the one for the EPICON
-data)
-
-```{r normalization}
-if(is_count){
-    # Normalize
-    data_set = EDASeq::newSeqExpressionSet(as.matrix(data), phenoData=meta)
-    data_normalized = EDASeq::betweenLaneNormalization(
-	data_set, which="upper", offset=TRUE, round=FALSE)
-    data_normalized = EDASeq::normCounts(data_normalized)
-}
-```
-
-# Differential expression analysis
-
-## Time-course differential expression analysis between two groups
+### Time-course differential expression analysis between two groups
 
@@
 pvalues = de_analysis
+outname = ".results/pvalues.txt"
+write.table(pvalues, outname)
 ```
 
-## Developmental differential expression analysis
+### Developmental differential expression analysis
 
@@
 
-## Weekly differential expression analysis
+### Weekly differential expression analysis
 
@@
 ```
-
-# Clustering
-
-```{r}
-if(filter_genes){
-    # Filter out q-values for the pvalues table
-    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
-    pvalues = pvalues[, col_to_keep]
-    fishers_pval = moanin:::fisher_method(pvalues)
-    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
-    y = as.matrix(data[genes_to_keep, ])
-}else{
-    y = as.matrix(data)
-}
-
-###############################################################################
-# Fit the splines
-
-kmeans_clusters = moanin::splines_kmeans(
-    y, meta, n_clusters=20,
-    random_seed=42,
-    degrees_of_freedom=df)
-
-# Rescale the centroids
-centroids = moanin::rescale_values(kmeans_clusters$centroids, meta)
-all_scores = data.frame(row.names=row.names(data))
-for(k in 1:n_clusters){
-    scores = moanin:::score_genes_centroid(data, centroids[k,])
-    all_scores[paste0("cluster_", k)] = scores
-}
-
-# Only assign labels to X% of the genes
-max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-labels = moanin:::row_argmin(all_scores)
-labels[genes_to_not_consider] = NA
-```
-
-```{r}
-print(unique(labels))
-```
-
-```{r}
-moanin::plot_centroids(centroids, meta)
-```
-
-# Downstream analysis
-
-```{r}
-
-```

### 2019-02-16 scripts/03_clustering.Rmd
Splitting the paper into several Rmd files

--- 2019-02-14 scripts/shoemaker2015.Rmd
@@
-title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
+---
+title: "Clustering: A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
-
 ---
 
-```{r echo=TRUE, results="hide"}
+```{r echo=FALSE, results="hide"}
 # Loading dependencies
@@
 
-```{r echo=TRUE}
+```{r echo=FALSE, results="hide"}
 # Set options
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, results="hide"}
 # Set data files and options
@@
 # Right now, the package doesn't contain the normalized data
+pvalues = read.table(".results/pvalues.txt", sep="\t")
 ```
 
-# Normalization
-
-FIXME this is not the normalization to use (this is the one for the EPICON
-data)
-
-```{r normalization}
-if(is_count){
-    # Normalize
-    data_set = EDASeq::newSeqExpressionSet(as.matrix(data), phenoData=meta)
-    data_normalized = EDASeq::betweenLaneNormalization(
-	data_set, which="upper", offset=TRUE, round=FALSE)
-    data_normalized = EDASeq::normCounts(data_normalized)
-}
-```
-
-# Differential expression analysis
-
-## Time-course differential expression analysis between two groups
-
-FIXME add estimation of log fold change.
-
-```{r de_analysis}
-# Define contrast
-de_analysis = data.frame(row.names=row.names(data))
-for(contrast_formula in timecourse_contrasts){
-    contrast = limma::makeContrasts(
-	contrast_formula,
-	levels=levels(meta$Group))
-    model = moanin:::edgeWithContrasts(as.matrix(data), meta, contrast)
-    contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
-    contrast_name_pval = paste(contrast_name, "-pval", sep="")
-    contrast_name_qval = paste(contrast_name, "-qval", sep="")
-    de_analysis[contrast_name_pval] = model$pval_lrt
-    de_analysis[contrast_name_qval] = model$qval_lrt
-}
-
-pvalues = de_analysis
-```
-
-## Developmental differential expression analysis
-
-```{r}
-# Define contrast
-contrast_formula = "C"
-contrast = limma::makeContrasts(
-    contrast_formula,
-    levels=levels(meta$Group))
-# Why do I have to do this again?
-mask = with(meta, Group == contrast_formula)
-model = moanin:::edgeWithContrasts(
-    as.matrix(data), meta, contrast, developmental=TRUE,
-    mask=mask)
-```
-
-## Weekly differential expression analysis
-
-```{r}
-# Define contrast
-contrasts = c("C.0-K.0",
-	      "C.3-K.3",
-	      "C.6-K.6",
-	      "C.9-K.9",
-	      "C.12-K.12",
-	      "C.18-K.18",
-	      "C.24-K.24",
-	      "C.30-K.30",
-	      "C.48-K.48",
-	      "C.60-K.60",
-	      "C.120-K.120",
-	      "C.168-K.168")
-
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
-# de_analysis =  fit_weekly_analysis(
-#     data, meta, contrasts,
-#     use_voom_weights=FALSE)
-```
-
-# Clustering
+# Clustering of time-course data
 
@@
 
-###############################################################################
+#############################################################################
 # Fit the splines
@@
 ```{r}
-print(unique(labels))
+table(labels)
 ```
@@
 ```
-
-# Downstream analysis
-
-```{r}
-
-```

### 2019-02-16 scripts/shoemaker2015.Rmd
Splitting the paper into several Rmd files

@@
-title: "Shoemaker et al, 2015: An Ultrasensitive Mechanism Regulates Influenza Virus-Induced Inflammation"
+---
+title: "A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
-
 ---
 
-```{r echo=TRUE, results="hide"}
+```{r echo=FALSE, results="hide"}
 # Loading dependencies

### 2019-02-16 scripts/02_differential_expression_analysis.Rmd
FIX small stuff due to R changing names

@@
 library(moanin)
+```
+
+```{r echo=FALSE}
+# Set options
+df = 6
+
+# Preprocessing & filtering
+is_count = FALSE
+take_log = FALSE    
+filter_expression = FALSE
+
+# Differential expression analysis
+timecourse_contrasts = c("C-K", "C-M")
+
+# Clustering
+filter_genes = TRUE
+n_genes_to_keep = 5000
+n_clusters = 20
+percentage_genes_to_label = 0.5
 ```
@@
 outname = ".results/pvalues.txt"
-write.table(pvalues, outname)
+write.table(pvalues, outname, sep="\t")
 ```

### 2019-02-16 scripts/03_clustering.Rmd
FIX small stuff due to R changing names

@@
 # Right now, the package doesn't contain the normalized data
-pvalues = read.table(".results/pvalues.txt", sep="\t")
+pvalues = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)
 ```
@@
 # Rescale the centroids
-centroids = moanin::rescale_values(kmeans_clusters$centroids, meta)
+centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)
 all_scores = data.frame(row.names=row.names(data))

### 2019-02-16 bin/moanin/.scripts/build_documentation.R

@@ -1,3 +1,6 @@
 library("devtools")
-document()
+library("pkgdown")
+
+devtools::document()
+pkgdown::build_site()
 

### 2019-02-16 bin/moanin/tests/testthat/test-de_analysis_utils.R

@@ -14,8 +14,9 @@ test_that("Estimating log fold change smoke tests", {
     contrasts = c("C-K")
     methods = moanin:::ALL_LFC_METHODS
     for(method in methods){
-        log_fold_change_sum = estimate_log_fold_change(
-	    data, meta, contrasts, method=method)
+        expect_silent(
+	    estimate_log_fold_change(
+		data, meta, contrasts, method=method))
     }
 })
 

### 2019-02-16 scripts/step3_DE_analysis/run_timecourse_differential_expression_analysis.R

@@ -54,8 +54,8 @@ for(contrast_formula in contrasts){
     contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
     contrast_name_pval = paste(contrast_name, "-pval", sep="")
     contrast_name_qval = paste(contrast_name, "-qval", sep="")
-    de_analysis[contrast_name_pval] = model$pval
-    de_analysis[contrast_name_qval] = model$qval
+    de_analysis[contrast_name_pval] = model$pval_lrt
+    de_analysis[contrast_name_qval] = model$qval_lrt
 }
 
 outname = file.path(data_dir, "timecourse_de_analysis.tsv")

### 2019-02-16 scripts/01_quality_control_normalization.Rmd
Now normalizing data as part of the paper

@@
 ```
+
+```{r echo=FALSE, include=FALSE}
+library(knitr)
+knitr::opts_chunk$set(
+  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
+  echo=FALSE,
+  duplicate.label="allow"
+)
+
+library(genefilter)
+library(dplyr)
+library(ggplot2)
+library(NMF)
+library(reshape2)
+library(limma)
+```
+
+The paper describes a normalization strategy that I am going to use here:
+
+  - First, normalize probe intensity using background information using the
+    "norm-exponential" strategy.
+  - Then normalize across micro-arrrays using quantily normalization.
+
+Unfortunately, that means going back to the raw data, which can be downloaded
+in the data folder, by clicking on the link specified in the README (it might
+take a while…). Untar the main file into a `rawdata` folder.
+
+```{r reading_targets, echo=FALSE, include=FALSE}
+
+targets = readTargets()
+RG = read.maimages(
+  targets, source="agilent",
+  path="data/shoemaker2015/rawdata",
+  other.columns="gIsWellAboveBG",
+  green.only=TRUE)
+
+```
+
+Before quantile normalization:
+
+```{r echo=FALSE, include=FALSE}
+RG = backgroundCorrect(RG, method="normexp")
+```
+
+```{r plot_densities, echo=FALSE}
+xlim = c(0, 14)
+plotDensities(RG, col="black", legend=FALSE)
+```
+
+After quantile normalization:
+
+```{r normalize_between_arrays, echo=FALSE}
+RG.q = normalizeBetweenArrays(RG, method="quantile")
+plotDensities(RG.q, col="black", legend=FALSE)
+
+sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
+colnames(RG.q) = sample_names
+row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
+idx_to_keep = which(
+  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
+			       "NegativeControl")))
+
+data = RG.q[idx_to_keep,]
+
+# Filter out control probes.
+control_probes = data$genes[, "ControlType"] == 1
+is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4
+
+data = data[!control_probes & is_expressed, ]$E
+
+write.table(data, "results/quantile_normalized.txt")
+
+meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
+
+```
+
+```{r order_genes, echo=FALSE}
+# Reorder genes on condition, time, and replicate
+ord = order(
+  meta$Group,
+  meta$Time,
+  meta$Replicate)
+data = data[, ord]
+meta = meta[ord, ]
+meta["Time"] = as.factor(unlist(meta["Time"]))
+meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
+```
+
+## Plot the RLE and PCA
+
+```{r leafRleRaw, fig.width=6}
+library(EDASeq)
+ylim = c(-2.5,2.5)
+plotRLE(exp(data), outline=FALSE,
+        whisklty=1, whisklwd=2, col=meta$Replicate, main="RLE, data")
+plotRLE(exp(data), outline=FALSE,
+        whisklty=1, whisklwd=2, col=meta$Group, main="RLE, data")
+plotRLE(exp(data), outline=FALSE,
+        whisklty=1, whisklwd=2, col=meta$Time, main="RLE, data")
+
+```
+```{r leafPCARaw}
+library(ggfortify)
+pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
+percent_var = round(100 * attr(pca_data, "percentVar"))
+
+autoplot(pca_data, data=meta, colour="Group")
+autoplot(pca_data, data=meta, colour="Time")
+autoplot(pca_data, data=meta, colour="Replicate")
+
+```
+
+```{r fig.width=6, fig.height=6}
+library(NMF)
+
+data_corr = cor(data, method="pearson")
+aheatmap(data_corr,
+         Colv=NA, Rowv=NA,
+         annCol=meta[,
+		     c("Group", "Time",
+			"Replicate")],
+         annLegend=FALSE, 
+         main="Replicates\nAfter Normalization")
+```
+
+```{r fig.width=6, fig.height=6}
+library(NMF)
+
+data_corr = cor(data, method="spearman")
+aheatmap(data_corr,
+         Colv=NA, Rowv=NA,
+         annCol=meta[,
+		     c("Group", "Time",
+			"Replicate")],
+         annLegend=FALSE, 
+         main="Replicates\nAfter Normalization")
+```

### 2019-02-16 scripts/02_differential_expression_analysis.Rmd
FIX taking the means of lfc for epicon-style computation

@@
     de_analysis[contrast_name_qval] = model$qval_lrt
+
+    # Now compute log fold change using different methods
+    # XXX This fails very badly if the contrasts provided is wrong.
+    log_fold_change_epicon = moanin::estimate_log_fold_change(
+	data, meta, c(contrast_formula), method="epicon")
+    contrast_name_lfc_epicon = paste(contrast_name, "-lfc_epicon", sep="")
+    de_analysis[contrast_name_lfc_epicon] = log_fold_change_epicon
+
+    # XXX Something is wrong with method max
+    log_fold_change_max = moanin::estimate_log_fold_change(
+	data, meta, c(contrast_formula), method="max")
+    contrast_name_lfc_max = paste(contrast_name, "-lfc_max", sep="")
+    de_analysis[contrast_name_lfc_max] = log_fold_change_max
+
 }

### 2019-02-16 bin/moanin/R/de_analysis_utils.R

@@ -46,7 +46,7 @@ estimate_log_fold_change = function(data, meta, contrasts, method="epicon"){
 		log_fold_changes[, contrast] = rowSums(timely_lfc[, mask]**2)
 	    }else if(method == "epicon"){
 		log_fold_changes[, contrast] = (
-		    rowSums(abs(timely_lfc[, mask])) * sign(rowSums(timely_lfc[, mask])))
+		    rowMeans(abs(timely_lfc[, mask])) * sign(rowSums(timely_lfc[, mask])))
 	    }
 	}
 

### 2019-02-16 scripts/03_clustering.Rmd
FIX rowMin and rowMax were wrong

@@
 if(filter_genes){
+    # First, filter out any genes that doesn't have a log fold change of at
+    # least 2
+    # XXX I should actually filter on the max, but it is broken right now.
+    lfc_col_to_keep = colnames(pvalues)[
+	grepl("-lfc_epicon", colnames(pvalues))]
+    pvalues = pvalues[rowSums(abs(pvalues[, lfc_col_to_keep]) > 2) > 0, ]
+
+    if(dim(pvalues)[1] > n_genes_to_keep){
+
+	# Then rank by fisher's p-value and take max the number of genes of
+	# interest
 	# Filter out q-values for the pvalues table
-    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
-    pvalues = pvalues[, col_to_keep]
+	pval_col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
+	pvalues = pvalues[, pval_col_to_keep]
 	fishers_pval = moanin:::fisher_method(pvalues)
+    } else {
+	genes_to_keep = names(pvalues)
+    }
     genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
@@
 ```
+
+```{r results="hide", echo=FALSE}
+# Save the clustering for the next step
+write.table(labels, ".results/clustering_labels.txt", sep="\t")
+```

### 2019-02-16 bin/moanin/R/utils.R

@@ -1,10 +1,12 @@
 
 # This should override the Biobase function
 rowMax = function(data){
-    return(as.vector(lapply(data, function(x) max(x))))
+    check_is_2d(data)
+    return(as.vector(apply(data, 1, max)))
 }
 
 rowMin = function(data){
-    return(as.vector(lapply(data, function(x) min(x))))
+    check_is_2d(data)
+    return(as.vector(apply(data, 1, min)))
 }
 

### 2019-02-16 bin/moanin/R/validation.R

@@ -17,7 +17,7 @@ check_meta = function(meta){
     if(!("Time" %in% metadata_column_names)){
 	stop(
 	    "Metadata doesn't contain expected information." +
-	    " Group column is missing.")
+	    " Time column is missing.")
     }
 
     # Just create this one.
@@ -43,4 +43,13 @@ check_data_meta = function(data, meta){
 	stop("Data should be of type numeric")
     }
     meta = check_meta(meta)
+}
+
+
+#' Check is 2D
+check_is_2d = function(X){
+    dim_data = dim(X)
+    if(is.null(dim_data)){
+	stop("Data is expected to be 2D. No dimension found.")
+    }
 }

### 2019-02-16 scripts/02_differential_expression_analysis.Rmd
adding a volcano plot

@@
 # Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M")
+timecourse_contrasts = c("C-K", "C-M", "C-VL", "C-VH")
 
@@
 pvalues = de_analysis
+```
+
+Let's explore a bit the number of genes found differentially expressed.
+
+```{r}
+print(colSums(pvalues[, grepl("-qval", colnames(pvalues))] < 0.05))
+```
+
+And now a volcano plot for quick visualization
+
+```{r fig.height=8, fig.width=8}
+
+pvalue = de_analysis[, "C-K-qval"]
+lfc_epicon = de_analysis[, "C-K-lfc_epicon"]
+
+plot(lfc_epicon, -log10(pvalue), pch=20, main="Volcano plot",
+     xlim=c(-2.5, 2))
+
+# Add colored points: red if padj<0.05, orange of log2FC>1, green if both)
+points(lfc_epicon[pvalue < 0.01], -log10(pvalue[pvalue < 0.01]), pch=20,
+       col="red")
+points(lfc_epicon[abs(lfc_epicon) > 1],
+       -log10(pvalue[abs(lfc_epicon) > 1]), pch=20, col="orange")
+points(lfc_epicon[(pvalue < 0.01) & (abs(lfc_epicon) > 1)],
+       -log10(pvalue[(pvalue < 0.01) & (abs(lfc_epicon) > 1)]), pch=20, col="green")
+
+```
+
+```{r echo=FALSE, results="hide"}
 outname = ".results/pvalues.txt"

### 2019-02-17 scripts/04_downstream_analysis.Rmd
ENH Starting downstream analysis

--- 2019-02-16 scripts/03_clustering.Rmd
@@
 ---
-title: "Clustering: A pipeline to analyse time-course gene expression data"
+title: "Downstream analysis: A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
@@
 library(moanin)
+library(KEGGprofile)
 
@@
 
-# Right now, the package doesn't contain the normalized data
-pvalues = read.table(".results/pvalues.txt", sep="\t")
+labels = read.table(".results/clustering_labels.txt", sep="\t",
+		    check.names=FALSE)
 ```
 
-# Clustering of time-course data
+### Finding enriched pathways
 
 ```{r}
-if(filter_genes){
-    # Filter out q-values for the pvalues table
-    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
-    pvalues = pvalues[, col_to_keep]
-    fishers_pval = moanin:::fisher_method(pvalues)
-    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
-    y = as.matrix(data[genes_to_keep, ])
-}else{
-    y = as.matrix(data)
+clusters = as.vector(unlist(unique(labels)))
+for(cluster in clusters){
+    genes = row.names(labels[labels == cluster])
+    pathways = KEGGprofile::find_enriched_pathway(
+	genes, species="mmu", download_latest=TRUE)
 }
-
-#############################################################################
-# Fit the splines
-
-kmeans_clusters = moanin::splines_kmeans(
-    y, meta, n_clusters=20,
-    random_seed=42,
-    degrees_of_freedom=df)
-
-# Rescale the centroids
-centroids = moanin::rescale_values(kmeans_clusters$centroids, meta)
-all_scores = data.frame(row.names=row.names(data))
-for(k in 1:n_clusters){
-    scores = moanin:::score_genes_centroid(data, centroids[k,])
-    all_scores[paste0("cluster_", k)] = scores
-}
-
-# Only assign labels to X% of the genes
-max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-labels = moanin:::row_argmin(all_scores)
-labels[genes_to_not_consider] = NA
 ```
-
-```{r}
-table(labels)
-```
-
-```{r}
-moanin::plot_centroids(centroids, meta)
-```

### 2019-02-17 scripts/03_clustering.Rmd
ENH filtering on lfc
ENH switching to viridis for better colors

@@
     lfc_col_to_keep = colnames(pvalues)[
-	grepl("-lfc_epicon", colnames(pvalues))]
+	grepl("-lfc_max", colnames(pvalues))]
     pvalues = pvalues[rowSums(abs(pvalues[, lfc_col_to_keep]) > 2) > 0, ]
@@
 ```{r}
+table(kmeans_clusters$clusters)
+```
+
+```{r}
 table(labels)
@@
 
-```{r}
+```{r fig.height=8, fig.width=8}
 moanin::plot_centroids(centroids, meta)
@@
 # Save the clustering for the next step
-write.table(labels, ".results/clustering_labels.txt", sep="\t")
+write.table(kmeans_clusters$clusters,
+	    ".results/clustering_labels.txt",
+	    sep="\t")
 ```

### 2019-02-17 bin/moanin/R/visualization.R

@@ -39,7 +39,7 @@ plot_centroid_individual = function(centroid, meta){
     yrange = range(centroid)
     
     plot(xrange, yrange, type="n")
-    colors = grDevices::terrain.colors(length(groups))
+    colors = viridis(length(groups))
     
     for(i in 1:length(groups)){
         group = groups[i]

### 2019-02-19 scripts/00_introduction.Rmd
DOC More details on the data set used

@@
 ---
-title: "XXX: a pipeline to analyse time-course gene expression data"
+title: "a pipeline to analyse time-course gene expression data"
 output: html_document
@@
 This workflow is illustrated using data from a micro-array time-course
-experiment, exposing mice to different strains of influenza, and collecting
-lung tissue during 12 time-points (..).
+experiment, exposing mice to three different strains of influenza, and
+collecting lung tissue during 14 time-points.  The three strains of influenza
+used in the study are (1) a low pathogenicity seasonal H1N1 influenza virus
+(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
+pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
+avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
+10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
+of the Vietnam avian influenza virus (10^3 PFU).
+
+XXX Add summary of the findings of the paper.

### 2019-02-19 scripts/04_downstream_analysis.Rmd
ENH Pathway enrichment analysis finally works

@@
 
+```{r}
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = labels[!is.na(labels)]
+```
+
 ### Finding enriched pathways
+
+```{r}
+library(biomaRt)
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+
+```
 
@@
 for(cluster in clusters){
-    genes = row.names(labels[labels == cluster])
+    print(cluster)
+    genes = gene_names[labels == cluster]
+
+    # convert gene names
+    genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		       filters="refseq_mrna", values=genes,
+		       mart=ensembl)["entrezgene"]
+    genes = as.vector(unlist(genes))
     pathways = KEGGprofile::find_enriched_pathway(
-	genes, species="mmu", download_latest=TRUE)
+	genes, species="mmu",
+	download_latest=TRUE)
+    print(pathways$stastic)
 }
 ```
+
+### Finding enriched GO terms

### 2019-02-19 scripts/03_clustering.Rmd
ENH Adding nicer colors to centroids

@@
 library(moanin)
-
+source("utils.R")
 # library(moanin)
@@
 
-kmeans_clusters = moanin::splines_kmeans(
+kmeans_clusters = moanin:::splines_kmeans(
     y, meta, n_clusters=20,
@@
 ```{r fig.height=8, fig.width=8}
-moanin::plot_centroids(centroids, meta)
+moanin:::plot_centroids(centroids, meta,
+			colors=as.vector(unlist(ann_colors["Group"])))
 ```
@@
 # Save the clustering for the next step
-write.table(kmeans_clusters$clusters,
+kmeans_labels = as.data.frame(kmeans_clusters$clusters)
+row.names(kmeans_labels) = row.names(y)
+write.table(kmeans_labels,
 	    ".results/clustering_labels.txt",

### 2019-02-19 scripts/02_differential_expression_analysis.Rmd
ENH Weekly DE analysis finally works

@@
 ---
-title: "Normalization: A pipeline to analyse time-course gene expression data"
+title: "Differential expression analysis: A pipeline to analyse time-course gene expression data"
 output: rmarkdown::html_vignette
 ---
-
-## Differential expression analysis of time-course data
 
@@
 # Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M", "C-VL", "C-VH")
+timecourse_contrasts = c("M-K", "M-C", "M-VL", "M-VH")
 
@@
 # Right now, the package doesn't contain the normalized data
+```
+
+```{r, echo=FALSE, results="hide"}
+source("utils.R")
+```
+
+## Differential expression analysis of time-course data
+
+The next step is typically to run a differential expression analysis.
+
+### Weekly differential expression analysis
+
+`moanin` provides a simple interface to perform a time by time differential
+expression analysis. Under the hood, it simply calls `limma` on the set of
+contrasts provided. Here, we show an example between the control mouse and the
+mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
+(H5N1), but the function works with any form contrasts.
+
+```{r weekly_differential_analysis}
+meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
+
+# Define contrast	
+contrasts = c("M.0-VH.0",
+	      "M.3-VH.3",
+	      "M.6-VH.6",
+	      "M.9-VH.9",
+	      "M.12-VH.12",
+	      "M.18-VH.18",
+	      "M.24-VH.24",
+	      "M.30-VH.30",
+	      "M.48-VH.48",
+	      "M.60-VH.60",
+	      "M.72-VH.72",
+	      "M.120-VH.120",
+	      "M.168-VH.168")
+weekly_de_analysis =  moanin::fit_weekly_analysis(
+     data, meta, contrasts,
+     use_voom_weights=FALSE)
+```
+
+```{r echo=FALSE, results="hide"}
+# Run the DE analysis on the rest of the influenza strain.
+contrasts = c("M.0-K.0",
+	      "M.3-K.3",
+	      "M.6-K.6",
+	      "M.9-K.9",
+	      "M.12-K.12",
+	      "M.18-K.18",
+	      "M.24-K.24",
+	      "M.30-K.30",
+	      "M.48-K.48",
+	      "M.60-K.60",
+	      "M.72-K.72",
+	      "M.120-K.120",
+	      "M.168-K.168")
+K_weekly_de_analysis =  moanin::fit_weekly_analysis(
+     data, meta, contrasts,
+     use_voom_weights=FALSE)
+contrasts = c("M.0-C.0",
+	      "M.3-C.3",
+	      "M.6-C.6",
+	      "M.9-C.9",
+	      "M.12-C.12",
+	      "M.18-C.18",
+	      "M.24-C.24",
+	      "M.30-C.30",
+	      "M.48-C.48",
+	      "M.60-C.60",
+	      "M.72-C.72",
+	      "M.120-C.120",
+	      "M.168-C.168")
+C_weekly_de_analysis =  moanin::fit_weekly_analysis(
+     data, meta, contrasts,
+     use_voom_weights=FALSE)
+contrasts = c("M.0-VH.0",
+	      "M.3-VH.3",
+	      "M.6-VH.6",
+	      "M.9-VH.9",
+	      "M.12-VH.12",
+	      "M.18-VH.18",
+	      "M.24-VH.24",
+	      "M.30-VH.30",
+	      "M.48-VH.48",
+	      "M.60-VH.60",
+	      "M.72-VH.72",
+	      "M.120-VH.120",
+	      "M.168-VH.168")
+VH_weekly_de_analysis =  moanin::fit_weekly_analysis(
+     data, meta, contrasts,
+     use_voom_weights=FALSE)
+
+```
+
+Let's look at the distribution of genes found differentially expressed per
+week between control and each of the influenza strains.
+
+```{r fig.width=8, fig.height=8, echo=FALSE}
+par(mfrow=c(2, 2))
+
+time = c(0, 3, 6, 9, 12, 18, 24, 30, 48, 60, 72, 120, 168)
+
+timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
+    qval_colnames = colnames(de_analysis)[
+	grepl("qval", colnames(de_analysis))]
+    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
+    names(number_de_genes_per_time) = labels
+    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
+	    xlab="Time", ylab="Number of DE genes", main=title)
+
+}
+
+timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
+timely_barplot(M_weekly_de_analysis, time, title="California")
+timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high")
+timely_barplot(weekly_de_analysis, time, title="Vietnam low")
 ```
@@
 
-pvalue = de_analysis[, "C-K-qval"]
-lfc_epicon = de_analysis[, "C-K-lfc_epicon"]
+pvalue = de_analysis[, "M-K-qval"]
+lfc_epicon = de_analysis[, "M-K-lfc_epicon"]
 
@@
 ```
-
-### Developmental differential expression analysis
-
-```{r}
-# Define contrast
-contrast_formula = "C"
-contrast = limma::makeContrasts(
-    contrast_formula,
-    levels=levels(meta$Group))
-# Why do I have to do this again?
-mask = with(meta, Group == contrast_formula)
-model = moanin:::edgeWithContrasts(
-    as.matrix(data), meta, contrast, developmental=TRUE,
-    mask=mask)
-```
-
-### Weekly differential expression analysis
-
-```{r}
-# Define contrast
-contrasts = c("C.0-K.0",
-	      "C.3-K.3",
-	      "C.6-K.6",
-	      "C.9-K.9",
-	      "C.12-K.12",
-	      "C.18-K.18",
-	      "C.24-K.24",
-	      "C.30-K.30",
-	      "C.48-K.48",
-	      "C.60-K.60",
-	      "C.120-K.120",
-	      "C.168-K.168")
-
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
-# de_analysis =  fit_weekly_analysis(
-#     data, meta, contrasts,
-#     use_voom_weights=FALSE)
-```

### 2019-02-19 scripts/01_quality_control_normalization.Rmd
FIX saving temporary results
Attempting to add colors

@@
 library(moanin)
+source("utils.R")
 
@@
   green.only=TRUE)
-
 ```
@@
 
-write.table(data, "results/quantile_normalized.txt")
+write.table(data, ".results/quantile_normalized.txt")
 
@@
          annLegend=FALSE, 
+	 annColors=ann_colors,
          main="Replicates\nAfter Normalization")

### 2019-02-20 scripts/00_introduction.Rmd
DOC chapter 0 & 1 of the paper

@@
 experiment, exposing mice to three different strains of influenza, and
-collecting lung tissue during 14 time-points.  The three strains of influenza
-used in the study are (1) a low pathogenicity seasonal H1N1 influenza virus
-(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
-pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
-avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
-10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
-of the Vietnam avian influenza virus (10^3 PFU).
+collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
+18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later). The three strains
+of influenza used in the study are (1) a low pathogenicity seasonal H1N1
+influenza virus (A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from
+the 2009 pandemic season (A/California/04/2009 [H1N1]), and a highly
+pathogenic H5N1 avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were
+injected with 10^5 PFU of each virus. An adidtional 42 mice were injected with
+a lower dose of the Vietnam avian influenza virus (10^3 PFU).
 
-XXX Add summary of the findings of the paper.
+**XXX Add summary of the key findings of the the paper**

### 2019-02-20 scripts/01_quality_control_normalization.Rmd
DOC chapter 0 & 1 of the paper

--- 2019-02-16 scripts/01_quality_control_normalization.Rmd
@@
 ```
+
+```{r echo=FALSE, include=FALSE}
+library(knitr)
+knitr::opts_chunk$set(
+  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
+  echo=TRUE,
+  duplicate.label="allow"
+)
+
+library(genefilter)
+library(dplyr)
+library(ggplot2)
+library(NMF)
+library(reshape2)
+library(limma)
+```
+
+## Pre-processing
+
+The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
+with accession number GSE63786. The dataset has a total of 209 samples, with
+45018 probes.
+
+Note that in the following, we assume that the user has access to a data
+folder `data/shoemaker2015/rawdata`, containing the microarray data.
+
+```{r}
+data_dir = "data/shoemaker2015/rawdata"
+```
+
+The normalization strategy we apply here relies on `limma` and is as follows:
+
+  - First, normalize probe intensity using background information using the
+    "norm-exponential" strategy.
+  - Then normalize across micro-arrrays using quantile normalization.
+
+```{r reading_targets, results="hide", fig.width=8, fig.height=8}
+# Read in the data using limma.
+targets = limma::readTargets()
+RG = limma::read.maimages(
+  targets, source="agilent",
+  path=data_dir,
+  other.columns="gIsWellAboveBG",
+  green.only=TRUE)
+
+# Preform background correction using norm-exponential strategy and normalize
+# across samples using quantile normalization.
+RG = backgroundCorrect(RG, method="normexp")
+RG.q = normalizeBetweenArrays(RG, method="quantile")
+
+# Plot the densities of the probes before and after normalization.
+par(mfrow=c(1, 2))
+plotDensities(RG, col="black", legend=FALSE)
+plotDensities(RG.q, col="black", legend=FALSE)
+```
+
+```{r echo=FALSE, results="hide"}
+# Get the number for the text. Do not display results or computation
+control_probes = RG.q$genes[, "ControlType"] == 1
+is_expressed = rowSums(RG.q$other$"gIsWellAboveBG" > 0) >= 4
+n_control_probes = sum(control_probes)
+n_lowly_expressed = sum(!is_expressed)
+```
+
+We then clean up the names of each gene and each sample to match the original
+data. Finally, we further filter out probes based on the following criteria:
+(1) removing the `r n_control_probes` control probes; (2) removing `r
+n_lowly_expressed` low-expressed genes.
+
+```{r clean_up_names}
+
+sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
+colnames(RG.q) = sample_names
+row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
+idx_to_keep = which(
+  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
+			       "NegativeControl")))
+
+data = RG.q[idx_to_keep,]
+
+# Filter out control probes.
+control_probes = data$genes[, "ControlType"] == 1
+is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4
+
+data = data[!control_probes & is_expressed, ]$E
+```
+
+ XXX The original paper says the data is "averaged over unique probes IDs."
+
+```{r echo=FALSE, results="hide"}
+# Do not display. setting variables for text
+n_samples = dim(data)[2]
+n_probes = dim(data)[1]
+```
+
+After filtering, we are left with `r n_probes` probes of interest in `r
+n_samples`.
+
+```{r}
+
+# Now load in the metadata
+meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
+write.table(data, ".results/quantile_normalized.txt")
+```
+
+```{r order_genes}
+# Reorder genes on condition, time, and replicate
+
+library(ggfortify)
+pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
+percent_var = round(100 * attr(pca_data, "percentVar"))
+
+meta["Time"] = as.factor(unlist(meta["Time"]))
+meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
+par(mfrow=c(1, 2))
+autoplot(pca_data, data=meta, colour="Group", shape="Replicate")
+autoplot(pca_data, data=meta, colour="Time", shape="Replicate")
+```
+
+We then plot the correlation matrix
+
+```{r fig.width=6, fig.height=6}
+library(NMF)
+
+# Reorder genes on condition, time, and replicate
+ord = order(
+  meta$Group,
+  meta$Time,
+  meta$Replicate)
+data = data[, ord]
+meta = meta[ord, ]
+
+data_corr = cor(data, method="pearson")
+aheatmap(data_corr,
+         Colv=NA, Rowv=NA,
+         annCol=meta[,
+		     c("Group", "Time",
+			"Replicate")],
+         annLegend=FALSE, 
+	 annColors=ann_colors,
+         main="Correlation plot after Normalization")
+```
+
+## Structure of the data set
+
+```{r}
+```

### 2019-02-20 scripts/01_quality_control_normalization.Rmd
FIX colors in PCA Plot

@@
 par(mfrow=c(1, 2))
-autoplot(pca_data, data=meta, colour="Group", shape="Replicate")
-autoplot(pca_data, data=meta, colour="Time", shape="Replicate")
+autoplot(pca_data, data=meta, colour="Group", shape="Replicate") +
+    scale_color_manual(values=as.vector(unlist(ann_colors["Group"])))
+
+autoplot(pca_data, data=meta, colour="Time", shape="Replicate") +
+    scale_color_manual(values=as.vector(unlist(ann_colors["Group"])))
 ```

### 2019-02-20 scripts/02_differential_expression_analysis.Rmd
FIX barplot of California strain was wrong

@@
 timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
-timely_barplot(M_weekly_de_analysis, time, title="California")
+timely_barplot(C_weekly_de_analysis, time, title="California")
 timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high")

### 2019-02-22 scripts/01_quality_control_normalization.Rmd
FIX 01_quality_control_normalization.Rmd
FIX plot in quality_control_normalization

--- 2019-02-20 scripts/01_quality_control_normalization.Rmd
@@
 
-```{r order_genes}
+```{r order_genes, fig.width=8, fig.height=4}
 # Reorder genes on condition, time, and replicate
@@
 
-meta["Time"] = as.factor(unlist(meta["Time"]))
-meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
+# FIXME Add legend
 par(mfrow=c(1, 2))
-autoplot(pca_data, data=meta, colour="Group", shape="Replicate")
-autoplot(pca_data, data=meta, colour="Time", shape="Replicate")
+plot(pca_data$x[, "PC1"], pca_data$x[, "PC2"],
+     col=ann_colors$Group[meta$Group], pch=16, xlab="PC1", ylab="PC2")
+plot(pca_data$x[, "PC1"], pca_data$x[, "PC2"],
+     col=ann_colors$Time[meta$Time], pch=16, xlab="PC1", ylab="PC2")
+
 ```

### 2019-02-22 scripts/03_clustering.Rmd
ENH Can now keep all the DE genes above the lfc threshold

@@
 
+###############################################################################
 # Clustering
+
+# Whether to filter genes
 filter_genes = TRUE
+keep_all_significant = TRUE
 n_genes_to_keep = 5000
@@
     }
+
+    if(keep_all_significant){
+	genes_to_keep = names(fishers_pval[fishers_pval < 0.05])
+    }else{
         genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
+    }
     y = as.matrix(data[genes_to_keep, ])
@@
     random_seed=42,
+    n_init=20,
     degrees_of_freedom=df)

### 2019-02-22 scripts/04_downstream_analysis.Rmd
ENH Better table for pathway + text

--- 2019-02-17 scripts/04_downstream_analysis.Rmd
@@
 library(KEGGprofile)
-
-# library(moanin)
 ```
@@
 
+## Downstream analysis of clusters.
+
+Once good clusters are obtained, the next step is to leverage the clustering
+to ease interpretation. Classic enrichment analysis step can then be performed
+in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
+term enrichment analysis, motif enrichment analysis, etc.
+
 ```{r echo=FALSE, results="hide"}
@@
 
-### Finding enriched pathways
+First, let us clean up the gene obtained and only select the genes we are
+going to use in the enrichment analysis. One can either use the whole set of
+genes, only the set of differentially expressed genes in each cluster, or a
+subset of genes that fit well to a cluster (based on some criterion).
 
 ```{r}
-clusters = as.vector(unlist(unique(labels)))
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = labels[!is.na(labels)]
+```
+
+### Finding enriched pathways using `biomaRt` and `KEGGprofile`
+
+Let us first tackle the case of pathway enrichment analysis. We will leverage
+the packages `biomaRt` and `KEGGprofile` for this step. `KEGGprofile` is a
+package that easily allows to perform pathway enrichment analysis on a set of
+genes labeled with the ensembl annotation. We thus need to convert the gene
+names into the appropriate format. This is where `biomaRt` comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use
+it to convert the gene names from the Refseq annotation to the ensembl one.
+
+```{r}
+library(biomaRt)
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+```
+
+```{r enrichment_analysis, results="hide"}
+clusters = sort(as.vector(unlist(unique(labels))))
+all_pathways = data.frame()
 for(cluster in clusters){
-    genes = row.names(labels[labels == cluster])
+    genes = gene_names[labels == cluster]
+
+    # convert gene names
+    genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		  filters="refseq_mrna", values=genes,
+		  mart=ensembl)["entrezgene"]
+    genes = as.vector(unlist(genes))
     pathways = KEGGprofile::find_enriched_pathway(
-	genes, species="mmu", download_latest=TRUE)
+	genes, species="mmu",
+	download_latest=TRUE)
+
+    if((dim(all_pathways)[1] == 0) & (dim(pathways$stastic)[1] != 0)){
+	print("Initialize")
+	pathways$stastic["Cluster"] = cluster
+	all_pathways = pathways$stastic
+    }else if(dim(pathways$stastic)[1] != 0) {
+	print("Merge")
+	pathways$stastic["Cluster"] = cluster
+	all_pathways = merge(all_pathways, pathways$stastic, all=TRUE)
+    }
 }
 ```
+
+```{r echo=FALSE}
+library(kableExtra)
+library(viridis)
+
+all_pathways = all_pathways[
+    order(all_pathways["Cluster"],
+	  all_pathways["pvalueAdj"]),]
+
+# Select the columns to print
+all_pathways = subset(
+    all_pathways,
+    select=c(
+	"Pathway_Name", "Percentage",
+	"pvalueAdj", "Cluster"))
+
+colnames(all_pathways) = c("Pathway", "Percentage", "Adj. p-value", "Cluster")
+all_pathways["Percentage"] = all_pathways["Percentage"] * 100
+kable(all_pathways, "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+### Finding enriched GO terms
+
+```{r go_term_enrichment}
+
+```

### 2019-02-23 scripts/04_downstream_analysis.Rmd
ENH GO term enrichment analysis now works

@@
 
-```{r go_term_enrichment}
+```{r go_to_gene_mapping}
 
+genes = getBM(attributes=c("go_id", "refseq_mrna"), values=gene_names, filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = NULL
+i = 1
+for(gene in gene_names){
+    go_terms = genes[genes[, "refseq_mrna"] == gene, "go_id"]
+    if(length(go_terms) != 0){
+        gene_id_go_mapping$gene = go_terms
+	names(gene_id_go_mapping)[i] = gene
+	i = i + 1
+    }
+}
 ```
+
+```{r go_term_enrichment, results="hide"}
+source("GO_terms_analysis.R")
+
+all_go_terms = data.frame()
+for(cluster in clusters){
+
+    scores = as.numeric(labels == cluster)
+    names(scores) = gene_names
+    go_terms_enriched = find_enriched_go_terms(
+	scores,
+    	gene_id_go_mapping)
+
+    if((dim(all_go_terms)[1] == 0) & (dim(go_terms_enriched)[1] != 0)){
+	go_terms_enriched["Cluster"] = cluster
+	all_go_terms = go_terms_enriched
+    }else if(dim(go_terms_enriched)[1] != 0){
+	go_terms_enriched["Cluster"] = cluster
+	all_go_terms = merge(all_go_terms, go_terms_enriched, all=TRUE)
+    }
+
+}
+```
+
+```{r echo=FALSE}
+library(kableExtra)
+library(viridis)
+
+all_go_terms = all_go_terms[
+    order(all_go_terms["Cluster",],
+	  all_go_terms["resultFisher.p.adj",])]
+
+kable(all_go_terms, "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```

### 2019-02-23 scripts/01_quality_control_normalization.Rmd
FIX More text + small fixes in QC + normalization

--- 2019-02-20 scripts/01_quality_control_normalization.Rmd
@@
 
-## Pre-processing
+## Pre-processing and normalization
 
@@
 with accession number GSE63786. The dataset has a total of 209 samples, with
-45018 probes.
+45018 probes (1 sample was removed by the authors prior to uploading the
+dataset on GEo).
 
@@
 
-```{r reading_targets, results="hide", fig.width=8, fig.height=8}
+```{r reading_targets, results="hide", fig.width=8, fig.height=4}
 # Read in the data using limma.
@@
 RG.q = normalizeBetweenArrays(RG, method="quantile")
+```
 
+Plotting the smoothed empirical densities of the probe values before
+normalization shows considerable variations betweens arrays. Once the data has
+been normalized across array using quantile normalization, the smoothed
+empirical densities of each sample perfectly overlap.
+
+```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
 # Plot the densities of the probes before and after normalization.
@@
 
-```{r order_genes}
+## Exploratory analysis and quality control
+
+Plotting the smoothed empirical densities of probe values for each sample and
+estimating the variations across samples before and after normalization is a
+first quality control. Typically, two other quality control and exploratory
+analysis steps are also performed before and after normalization: (1) low
+dimensionality embedding of the samples; (2) correlation plots between each
+samples. In both cases, we expect a strong biological signal, while replicate
+samples should be strongly clustered or correlated with one another.
+
+Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
+centered and scaled gene expression data.
+
+```{r pca, fig.width=8, fig.height=8}
 # Reorder genes on condition, time, and replicate
@@
 library(ggfortify)
-pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
+pca_data = prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
 percent_var = round(100 * attr(pca_data, "percentVar"))
-
-meta["Time"] = as.factor(unlist(meta["Time"]))
-meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
-par(mfrow=c(1, 2))
-autoplot(pca_data, data=meta, colour="Group", shape="Replicate")
-autoplot(pca_data, data=meta, colour="Time", shape="Replicate")
 ```
 
-We then plot the correlation matrix
+We then plot the two first components, and color each sample by (1) its
+condition; (2) its sampling time. We use different markers for each replicate.
+We also plot the second and third components in the second row.
 
-```{r fig.width=6, fig.height=6}
+```{r pca_plots, fig.width=8, fig.height=8}
+# FIXME Add legend
+par(mfrow=c(2, 2))
+plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+     col=ann_colors$Group[meta$Group],
+     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
+     xlab="PC2", ylab="PC1")
+plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+     col=ann_colors$Time[as.factor(meta$Time)],
+     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
+     xlab="PC2", ylab="PC1")
+
+plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
+     col=ann_colors$Group[meta$Group],
+     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
+     xlab="PC2", ylab="PC3")
+plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
+     col=ann_colors$Time[as.factor(meta$Time)],
+     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
+     xlab="PC2", ylab="PC3")
+```
+
+Then, we plot the pearson correlation across each samples. We order the
+samples by their Group (treatment) and Time (the time of sampling).
+Additionally, in this example, we order each treatment by strength of the
+pathogenicity of the treatment: Control, Kawasaki, California, low dose of
+Vietnam, then high dose of Vietnam.
+
+```{r  correlation_plot, fig.width=6, fig.height=6}
 library(NMF)
@@
 # Reorder genes on condition, time, and replicate
+meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 1, 2, 5, 4)])
 ord = order(
@@
 data_corr = cor(data, method="pearson")
+
 aheatmap(data_corr,
@@
 	 annColors=ann_colors,
-         main="Correlation plot after Normalization")
+         main="Correlation plot")
 ```
+
+We can already see interesting patterns: the cross-correlation amongst samples
+taking from the untreated mice is higher than the cross correlation amongst
+the rest of the treatments.
 

### 2019-02-23 scripts/utils.R

@@ -5,10 +5,14 @@ group_colors = c(
     "VH"="red4",
     "VL"="red2")
 time_colors = grDevices::rainbow(14)
-names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 48, 60, 72, 120, 168)
+names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
 
 ann_colors = list(
     Time=time_colors,
     Group=group_colors
     )
 
+replicate_markers = c(15, 17, 19)
+names(replicate_markers) = c(1, 2, 3)
+ann_markers = list(
+    Replicate=replicate_markers)

### 2019-02-23 scripts/01_quality_control_normalization.Rmd
More text on the QC normalization section

@@
 
+Before performing any additional exploratory analysis, let us only keep highly
+variable genes.
+
+```{r}
+variance_cutoff = 0.5
+variance_per_genes = apply(data, 1, mad)
+min_variance = quantile(variance_per_genes, c(variance_cutoff))
+data = data[variance_per_genes > min_variance,]
+```
+
+```{r echo=FALSE, fig.width=8, fig.height=4}
+hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
+     ylab="# probes")
+abline(v=min_variance, col="#AB0000", lw=2)
+```
+
 Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
@@
 
+meta$Time = as.factor(meta$Time)
 aheatmap(data_corr,
@@
          annCol=meta[,
-		     c("Group", "Time",
-		       "Replicate")],
+		     c("Group", "Time")],
          annLegend=FALSE, 
@@
 
-We can already see interesting patterns: the cross-correlation amongst samples
-taking from the untreated mice is higher than the cross correlation amongst
-the rest of the treatments.
-
-## Structure of the data set
-
-```{r}
-```
+We can already see interesting patterns emerging from the correlation plot.
+First, the cross-correlation amongst samples taking from the control mice is
+higher than the cross correlation amongst the rest of the treatments. Second,
+the influenza-infected mice midly react until time point 36. Third, the
+California strain samples seem closer to control samples than the
+lowly-pathogenic Kawasaki samples. Fourth, the Vietnam samples at time point
+120 and 168 are the one that are the most different from control samples.

### 2019-02-26 scripts/02_differential_expression_analysis.Rmd
ENH small fixes and enhancements

@@
 ```{r, echo=FALSE, results="hide"}
+# XXX FIXME
+data = read.table(".results/quantile_normalized.txt") 
 source("utils.R")
@@
 # Define contrast	
-contrasts = c("M.0-VH.0",
-	      "M.3-VH.3",
-	      "M.6-VH.6",
-	      "M.9-VH.9",
-	      "M.12-VH.12",
-	      "M.18-VH.18",
-	      "M.24-VH.24",
-	      "M.30-VH.30",
-	      "M.48-VH.48",
-	      "M.60-VH.60",
-	      "M.72-VH.72",
-	      "M.120-VH.120",
-	      "M.168-VH.168")
+contrasts = c("M.0-VL.0",
+	      "M.3-VL.3",
+	      "M.6-VL.6",
+	      "M.9-VL.9",
+	      "M.12-VL.12",
+	      "M.18-VL.18",
+	      "M.24-VL.24",
+	      "M.30-VL.30",
+	      "M.36-VL.36",
+	      "M.48-VL.48",
+	      "M.60-VL.60",
+	      "M.72-VL.72",
+	      "M.120-VL.120",
+	      "M.168-VL.168")
 weekly_de_analysis =  moanin::fit_weekly_analysis(
@@
 	      "M.30-K.30",
+	      "M.36-K.36",
 	      "M.48-K.48",
@@
 	      "M.30-C.30",
+	      "M.36-K.36",
 	      "M.48-C.48",
@@
 	      "M.30-VH.30",
+	      "M.36-VH.36",
 	      "M.48-VH.48",
@@
      use_voom_weights=FALSE)
-
 ```
@@
 
-time = c(0, 3, 6, 9, 12, 18, 24, 30, 48, 60, 72, 120, 168)
+time = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
 
@@
 timely_barplot(C_weekly_de_analysis, time, title="California")
-timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high")
-timely_barplot(weekly_de_analysis, time, title="Vietnam low")
+timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose")
+timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose")
 ```

### 2019-02-28 scripts/02_differential_expression_analysis.Rmd
FIX renamed some functions

@@
 	      "M.168-VL.168")
-weekly_de_analysis =  moanin::fit_weekly_analysis(
+weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
      data, meta, contrasts,
@@
 	      "M.168-K.168")
-K_weekly_de_analysis =  moanin::fit_weekly_analysis(
+K_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
      data, meta, contrasts,
@@
 	      "M.168-C.168")
-C_weekly_de_analysis =  moanin::fit_weekly_analysis(
+C_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
      data, meta, contrasts,
@@
 	      "M.168-VH.168")
-VH_weekly_de_analysis =  moanin::fit_weekly_analysis(
+VH_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
      data, meta, contrasts,
@@
     contrast = limma::makeContrasts(
-	contrast_formula,
+	contrasts=contrast_formula,
 	levels=levels(meta$Group))
-    model = moanin:::edgeWithContrasts(as.matrix(data), meta, contrast)
+    model = moanin::timecourse_differential_expression_analysis(
+	as.matrix(data), meta, contrast)
     contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)

### 2019-02-28 bin/moanin/R/timecourse_de_analysis.R

@@ -173,9 +173,15 @@ summarise = function(X, ng_levels) {
 #'	  useful when you want to use all data points for fitting the splines,
 #'	  but not all the data points for the tests.
 #' @export
-edgeWithContrasts = function(data, meta, contrasts=NULL, center=FALSE,
-			     weights=NULL, df=4,
-			     basis=NULL, mask=NULL, developmental=FALSE){
+timecourse_differential_expression_analysis = function(data,
+						       meta,
+						       contrasts=NULL,
+						       center=FALSE,
+						       weights=NULL,
+						       df=4,
+						       basis=NULL,
+						       mask=NULL,
+						       developmental=FALSE){
     ng = nlevels(meta$Group)
     ng_labels = meta$Group
 

### 2019-02-28 bin/moanin/R/weekly_de_analysis.R

@@ -11,7 +11,9 @@ library("edgeR")
 #' @param use_voom_weights boolean: whether to use voom weights or not
 #'
 #' @export
-fit_weekly_analysis = function(counts, meta, contrasts, use_voom_weights=TRUE){
+weekly_differential_expression_analysis = function(counts, meta,
+						   contrasts,
+						   use_voom_weights=TRUE){
     meta = check_meta(meta)
 
     design = stats::model.matrix(~WeeklyGroup + 0, data=meta)

### 2019-03-01 scripts/02_differential_expression_analysis.Rmd
ENH Every DE related method now take limma contrasts

@@
 
-FIXME add estimation of log fold change.
-
 ```{r de_analysis}
 # Define contrast
-de_analysis = data.frame(row.names=row.names(data))
-for(contrast_formula in timecourse_contrasts){
-    contrast = limma::makeContrasts(
-	contrasts=contrast_formula,
+contrasts = limma::makeContrasts(
+	contrasts=timecourse_contrasts,
 	levels=levels(meta$Group))
-    model = moanin::timecourse_differential_expression_analysis(
-	as.matrix(data), meta, contrast)
-    contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
-    contrast_name_pval = paste(contrast_name, "-pval", sep="")
-    contrast_name_qval = paste(contrast_name, "-qval", sep="")
-    de_analysis[contrast_name_pval] = model$pval_lrt
-    de_analysis[contrast_name_qval] = model$qval_lrt
 
-    # Now compute log fold change using different methods
-    # XXX This fails very badly if the contrasts provided is wrong.
+pvalues = moanin::timecourse_differential_expression_analysis(
+    data, meta, contrasts)
+qvalues = apply(pvalues, 2, stats::p.adjust)
+
 log_fold_change_epicon = moanin::estimate_log_fold_change(
-	data, meta, c(contrast_formula), method="epicon")
-    contrast_name_lfc_epicon = paste(contrast_name, "-lfc_epicon", sep="")
-    de_analysis[contrast_name_lfc_epicon] = log_fold_change_epicon
+    data, meta, contrasts,  method="epicon")
 
-    # XXX Something is wrong with method max
 log_fold_change_max = moanin::estimate_log_fold_change(
-	data, meta, c(contrast_formula), method="max")
-    contrast_name_lfc_max = paste(contrast_name, "-lfc_max", sep="")
-    de_analysis[contrast_name_lfc_max] = log_fold_change_max
-
-}
-
-pvalues = de_analysis
+    data, meta, contrasts, method="max")
 ```
@@
 
-pvalue = de_analysis[, "M-K-qval"]
-lfc_epicon = de_analysis[, "M-K-lfc_epicon"]
+pvalue = qvalues[, "M-K"]
+lfc_epicon = log_fold_change_epicon[, "M-K"]
 

### 2019-03-01 bin/moanin/R/de_analysis_utils.R

@@ -23,20 +23,13 @@ estimate_log_fold_change = function(data, meta, contrasts, method="epicon"){
 		  all_methods, sep=""))
     }
 
-    if(method == "sum"){
-
-	log_fold_changes = estimate_log_fold_change_sum(data, meta, contrasts)
-
-    }else if(method == "timely"){
-
+    if(method == "timely"){
 	log_fold_changes = lfc_per_time(data, meta, contrasts)
-
-    }else if(method %in% c("max", "min", "abs_sum", "abs_squared_sum", "epicon")){
-
+    }else if(method %in% c("sum", "max", "min", "abs_sum", "abs_squared_sum", "epicon")){
 	timely_lfc = lfc_per_time(data, meta, contrasts)
 	timely_lfc_meta = reconstruct_meta_from_lfc(timely_lfc)
     	log_fold_changes = data.frame(row.names=row.names(data))
-	for(contrast in contrasts){
+	for(contrast in colnames(contrasts)){
 	    mask = timely_lfc_meta$Group == contrast
 	    if(method == "max"){
 		log_fold_changes[, contrast] = rowMax(abs(timely_lfc[, mask]))
@@ -49,6 +42,8 @@ estimate_log_fold_change = function(data, meta, contrasts, method="epicon"){
 	    }else if(method == "epicon"){
 		log_fold_changes[, contrast] = (
 		    rowMeans(abs(timely_lfc[, mask])) * sign(rowSums(timely_lfc[, mask])))
+	    }else if(method == "sum"){
+		log_fold_changes[, contrast] = rowSums(timely_lfc[, mask])
 	    }
 	}
 
@@ -60,9 +55,7 @@ estimate_log_fold_change = function(data, meta, contrasts, method="epicon"){
 
 
 estimate_log_fold_change_sum = function(data, meta, contrasts){
-    contrasts_coef = limma::makeContrasts(contrasts=contrasts,
-					  levels=levels(meta$Group))
-    sample_coefficients = lapply(meta$Group, function(x) return(contrasts_coef[x, ]))
+    sample_coefficients = lapply(meta$Group, function(x) return(contrasts[x, ]))
     sample_coefficients = as.matrix(sample_coefficients)
 
     # First, do weekly contrasts
@@ -103,13 +96,10 @@ reconstruct_meta_from_lfc = function(data_per_time){
 lfc_per_time = function(data, meta, contrasts){
     meta$Time = as.factor(meta$Time)
 
[…]

### 2019-03-01 bin/moanin/R/timecourse_de_analysis.R

@@ -89,23 +89,11 @@ compute_pvalue = function(X, y, beta, beta_null, ng_labels,
 			  n_samples=NULL,
 			  degrees_of_freedom=NULL,
 			  statistics="lrt",
-			  df2=NULL, weights=NULL, mask=NULL, developmental=FALSE){
+			  df2=NULL, weights=NULL){
 
     fitFull = beta %*% t(X)
-    if(developmental){
-      fitNull = row_mean(y[, mask])
-    }else{
-      fitNull = beta_null %*% t(X)
-    }
-  
-    if(!is.null(mask)){
-      fitFull = fitFull[, mask]
-      if(!developmental){
-        fitNull = fitNull[, mask]
-      }
-      y = y[, mask]
-      ng_labels = ng_labels[mask]
-    }
+
+    fitNull = beta_null %*% t(X)
 
     if(!is.null(weights)){
       resNull = weights^(1/2) * (y - fitNull)
@@ -165,46 +153,24 @@ summarise = function(X, ng_levels) {
 #' @param meta Meta data which should be consistent with the column names \code{data}.
 #' @param contrasts Contrast using \code{makeContrasts} from \code{limma}.
 #' @param center boolean, whether to center the data matrix
-#' @param developmental boolean, is that a developmental test?
 #' @param weights Weights matrix for the fitting.
 #' @param df Degrees of freedom to for the splines.
 #' @param basis the basis matrix. If provided, ignore some of the parameters
-#' @param mask boolean mask of which points to use for the tests. This is
-#'	  useful when you want to use all data points for fitting the splines,
-#'	  but not all the data points for the tests.
 #' @export
 timecourse_differential_expression_analysis = function(data,
 						       meta,
-						       contrasts=NULL,
+						       contrasts,
 						       center=FALSE,
 						       weights=NULL,
 						       df=4,
-						       basis=NULL,
-						       mask=NULL,
-						       developmental=FALSE){
+						       basis=NULL){
     ng = nlevels(meta$Group)
     ng_labels = meta$Group
 
     meta = check_meta(meta)
-
-    if(is.null(contrasts) & !developmental){
-	# FIXME need better error message
-	stop("Needs either contrasts or developmental")
-    }
-
-    contrasts_coef = c(contrasts)
     meta = droplevels(meta)
[…]

### 2019-03-01 scripts/02_differential_expression_analysis.Rmd
FIX saving the pvalues in the right format

@@
 ```{r echo=FALSE, results="hide"}
+# Rename stuff for saving.
 outname = ".results/pvalues.txt"
-write.table(pvalues, outname, sep="\t")
+colnames(pvalues) = unlist(
+    lapply(colnames(pvalues), function(x){paste0(x, "-pval")}))
+
+colnames(qvalues) = unlist(
+    lapply(colnames(qvalues), function(x){paste0(x, "-qval")}))
+
+colnames(log_fold_change_epicon) = unlist(
+    lapply(colnames(log_fold_change_epicon),
+    function(x){paste0(x, "-lfc_epicon")}))
+
+colnames(log_fold_change_max) = unlist(
+    lapply(colnames(log_fold_change_max),
+    function(x){paste0(x, "-lfc_max")}))
+
+de_analysis = merge(pvalues, qvalues, all=TRUE, by="row.names")
+row.names(de_analysis) = row.names(pvalues)
+de_analysis = subset(de_analysis, select=-c(Row.names))
+
+de_analysis = merge(de_analysis, log_fold_change_max, all=TRUE, by="row.names")
+row.names(de_analysis) = row.names(pvalues)
+de_analysis = subset(de_analysis, select=-c(Row.names))
+
+de_analysis = merge(de_analysis, log_fold_change_epicon, all=TRUE,
+		    by="row.names")
+row.names(de_analysis) = row.names(pvalues)
+de_analysis = subset(de_analysis, select=-c(Row.names))
+
+write.table(de_analysis, outname, sep="\t")
 ```

### 2019-03-01 scripts/03_clustering.Rmd
ENH small fixes and enhancements
FIX saving the pvalues in the right format

@@
 data("shoemaker2015")
+
 data = shoemaker2015$data
 meta = shoemaker2015$meta
+
+data = read.table(".results/quantile_normalized.txt")
 
@@
 	fishers_pval = moanin:::fisher_method(pvalues)
-    } else {
-	genes_to_keep = names(pvalues)
-    }
-
         if(keep_all_significant){
@@
 	}
+    }else{
+	genes_to_keep = names(pvalues)
+    }
+
     y = as.matrix(data[genes_to_keep, ])

### 2019-03-01 scripts/03_clustering.Rmd
FIX reactivating the max number of genes used for the clustering

@@
 
-# Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M")
-
 ###############################################################################
@@
 filter_genes = TRUE
-keep_all_significant = TRUE
-n_genes_to_keep = 5000
+keep_all_significant = FALSE
+n_genes_to_keep = 10000
 n_clusters = 20
@@
 # Right now, the package doesn't contain the normalized data
-pvalues = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)
+de_analysis = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)
 ```
@@
 
-```{r}
+```{r filter_genes}
 if(filter_genes){
@@
     # XXX I should actually filter on the max, but it is broken right now.
-    lfc_col_to_keep = colnames(pvalues)[
-	grepl("-lfc_max", colnames(pvalues))]
-    pvalues = pvalues[rowSums(abs(pvalues[, lfc_col_to_keep]) > 2) > 0, ]
+    lfc_col_to_keep = colnames(de_analysis)[
+	grepl("-lfc_max", colnames(de_analysis))]
+    de_analysis = de_analysis[rowSums(abs(de_analysis[, lfc_col_to_keep]) > 2) > 0, ]
 
-    if(dim(pvalues)[1] > n_genes_to_keep){
+    if(dim(de_analysis)[1] > n_genes_to_keep){
 
@@
 	# Filter out q-values for the pvalues table
-	pval_col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
-	pvalues = pvalues[, pval_col_to_keep]
+	pval_col_to_keep = colnames(de_analysis)[grepl("-pval", colnames(de_analysis))]
+	pvalues = de_analysis[, pval_col_to_keep]
 	fishers_pval = moanin:::fisher_method(pvalues)
+	fishers_qval = stats::p.adjust(fishers_pval)
         if(keep_all_significant){
-	   genes_to_keep = names(fishers_pval[fishers_pval < 0.05])
+	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
 	}else{
@@
 }
+```
+
+After filtering, we consider `r dim(y)[1]` genes for the clustering.
+
+```{r clustering}
 

### 2019-03-01 bin/moanin/R/de_analysis_utils.R

@@ -108,7 +108,7 @@ lfc_per_time = function(data, meta, contrasts){
 	for(time in meta$Time){
 	    mask = meta$Time == time
 	    colname = paste0(column, ".", as.character(time))
-	    log_fold_changes[, colname] = rowSums(coef_data[, mask])
+	    log_fold_changes[, colname] = rowMeans(coef_data[, mask])
 	}
     }
     return(log_fold_changes)

### 2019-03-01 scripts/04_downstream_analysis.Rmd
ENH small fixes and enhancements
ENH running GO term enrichment only on three clusters, but on the three ontologies

--- 2019-02-22 scripts/04_downstream_analysis.Rmd
@@
 
-```{r}
+```{r results="hide"}
 library(biomaRt)
@@
 
-```{r enrichment_analysis, results="hide"}
+```{r enrichment_analysis, results="hide", message=FALSE}
 clusters = sort(as.vector(unlist(unique(labels))))
+clusters = c(1, 2, 3)
 all_pathways = data.frame()
@@
 
-```{r go_term_enrichment}
+To find GO terms, we use `biomaRt` to find the mapping between GO terms and
+gene mapping.
 
+```{r go_to_gene_mapping, results="hide", message=FALSE}
+
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = NULL
+i = 1
+for(gene in gene_names){
+    go_terms = genes[genes[, "refseq_mrna"] == gene, "go_id"]
+    if(length(go_terms) != 0){
+        gene_id_go_mapping$gene = go_terms
+	names(gene_id_go_mapping)[i] = gene
+	i = i + 1
+    }
+}
 ```
+
+```{r go_term_enrichment, results="hide", message=FALSE}
+source("GO_terms_analysis.R")
+
+all_go_terms = data.frame()
+for(cluster in clusters){
+    for(ontogoly in c("BP", "NF", "CC")){ 
+	scores = as.numeric(labels == cluster)
+	names(scores) = gene_names
+	go_terms_enriched = find_enriched_go_terms(
+	    scores,
+	    gene_id_go_mapping)
+
+	if((dim(all_go_terms)[1] == 0) & (dim(go_terms_enriched)[1] != 0)){
+	    go_terms_enriched["Cluster"] = cluster
+	    go_terms_enriched["Ontology"] = ontology
+	    all_go_terms = go_terms_enriched
+	}else if(dim(go_terms_enriched)[1] != 0){
+	    go_terms_enriched["Cluster"] = cluster
+	    go_terms_enriched["Ontogoly"] = ontology
+	    all_go_terms = merge(all_go_terms, go_terms_enriched, all=TRUE)
+	}
+    }
+}
+```
+
+```{r echo=FALSE}
+library(kableExtra)
+
+all_go_terms = all_go_terms[
+    order(all_go_terms[, "Cluster"],
+	  all_go_terms[, "resultFisher_padj"]), ]
+
+kable(all_go_terms, "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```

### 2019-03-01 scripts/GO_terms_analysis.R

@@ -1,0 +1,49 @@
+library(topGO)
+
+
+# ontology: BP, CC, NF
+find_enriched_go_terms = function(scores, geneID2GO,
+				  ontology="BP", 
+				  weighted=FALSE,
+				  node_size=10){
+    if(!(ontology %in% c("BP", "CC", "NF"))){
+	error_message = paste(
+	    "moanin::find_enriched_go_terms: Ontology should be 'BP', 'CC',",
+	    "or 'NF'. Ontology provided is",
+	    ontology)
+	stop(error_message)
+    }
+
+
+    getTopDiffGenes = function(data, cutOff=NULL){
+      return(data < 0.5)
+    }
+
+    GOdata = new("topGOdata", ontology=ontology, allGenes=scores, nodeSize=node_size,
+		 geneSel=getTopDiffGenes,
+		 annot=annFUN.gene2GO, gene2GO=geneID2GO)
+
+    if(weighted){
+        resultFisher = runTest(GOdata, algorithm="classic", statistic="fisher")
+    }else{
+        resultFisher = runTest(GOdata, algorithm="weight", statistic="fisher")
+    }
+    n_nodes = length(resultFisher@score)
+
+    allRes = GenTable(
+	GOdata, 
+	resultFisher=resultFisher,
+	orderBy="resultFisher", ranksOf="resultFisher",
+	topNodes=n_nodes)
+        
+    # P-value correct
+    allRes[, "resultFisher_padj"] = stats::p.adjust(allRes$resultFisher, method="BH")
+    wh = which(allRes[, "resultFisher_padj"] <= 0.05)
+
+    allRes = allRes[wh,]
+    wh = which(apply(allRes[, c("Significant", "Expected")],
+    	              1, function(x){x["Significant"] > x["Expected"]}))
+    allRes = allRes[wh,]
+
+    return(allRes)
+}

### 2019-03-02 scripts/04_downstream_analysis.Rmd
ENH more text on the downstream analysis part of the manuscript

--- 2019-02-22 scripts/04_downstream_analysis.Rmd
@@
 
-```{r go_term_enrichment}
+To find GO terms, we use `biomaRt` to find the mapping between GO terms and
+gene mapping. The GO enrichment library `topGO` expects the GO term to gene
+mapping to be a list where each item is a mapping between a gene name and a GO
+term ID vector.
 
+```text
+$NM_199153
+[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
+[6] "GO:0050896" "GO:0050909"
+
+$NM_201361
+ [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
+ [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
+[11] "GO:0005819" "GO:0000922" "GO:0072686"
 ```
+
+`biomaRt` queries results a matrix with two named columns of gene names and GO
+term ID. **FIXME** matrix or data.frame??
+
+```{r go_to_gene_mapping, results="hide", message=FALSE}
+# FIXME move this to the package
+source("GO_terms_analysis.R")
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = create_go_term_mapping(genes)
+```
+
+Once the gene ID to GO mapping list is created, `moanin` provides a simple
+interface to `topGO` to fetch enriched GO terms.
+
+```{r go_term_enrichment, results="hide", message=FALSE}
+
+cluster = c(3)
+ontologies = c("BP", "NF", "CC")
+# It's taking too long to compile with everything. Just limit ourselves for
+# now to BP
+ontologies = c("BP")
+all_go_terms = data.frame()
+for(cluster in clusters){
+    for(ontology in c("BP", "NF", "CC")){ 
+	scores = as.numeric(labels == cluster)
+	names(scores) = gene_names
+	go_terms_enriched = find_enriched_go_terms(
+	    scores,
+	    gene_id_go_mapping)
+
+	if((dim(all_go_terms)[1] == 0) & (dim(go_terms_enriched)[1] != 0)){
+	    go_terms_enriched["Cluster"] = cluster
+	    go_terms_enriched["Ontology"] = ontology
+	    all_go_terms = go_terms_enriched
+	}else if(dim(go_terms_enriched)[1] != 0){
+	    go_terms_enriched["Cluster"] = cluster
+	    go_terms_enriched["Ontology"] = ontology
+	    all_go_terms = merge(all_go_terms, go_terms_enriched, all=TRUE)
+	}
+    }
+}
+```
+
+```{r echo=FALSE}
+library(kableExtra)
+
+all_go_terms = all_go_terms[
+    order(all_go_terms[, "Cluster"],
+	  all_go_terms[, "resultFisher_padj"]), ]
+
+kable(all_go_terms, "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```

### 2019-03-02 scripts/GO_terms_analysis.R

@@ -1,8 +1,18 @@
 library(topGO)
 
 
-# ontology: BP, CC, NF
-find_enriched_go_terms = function(scores, geneID2GO,
+#' Find enriched GO terms
+#'
+#' @param labels
+#' @param gene_id_to_go
+#' @param ontology string, optional, default: BP
+#'	specficies which ontology to use. Can be 'BP', 'CC', or 'NF'
+#' @param weighted, boolean, optional, default: FALSE
+#'	Whether to use the weighted algorithm or not.
+#' @param node_size integer, optional, default: 10
+#'	Consider only GO terms with node_size number of genes.
+#' @param ontology: BP, CC, NF
+find_enriched_go_terms = function(labels, gene_id_to_go,
 				  ontology="BP", 
 				  weighted=FALSE,
 				  node_size=10){
@@ -19,9 +29,9 @@ find_enriched_go_terms = function(scores, geneID2GO,
       return(data < 0.5)
     }
 
-    GOdata = new("topGOdata", ontology=ontology, allGenes=scores, nodeSize=node_size,
+    GOdata = new("topGOdata", ontology=ontology, allGenes=labels, nodeSize=node_size,
 		 geneSel=getTopDiffGenes,
-		 annot=annFUN.gene2GO, gene2GO=geneID2GO)
+		 annot=annFUN.gene2GO, gene2GO=gene_id_to_go)
 
     if(weighted){
         resultFisher = runTest(GOdata, algorithm="classic", statistic="fisher")
@@ -46,4 +56,23 @@ find_enriched_go_terms = function(scores, geneID2GO,
     allRes = allRes[wh,]
 
     return(allRes)
+}
+
+
+#' Create the Gene to GO Term mapping
+#'
+create_go_term_mapping = function(genes, gene_col="refseq_mrna"){
+    gene_id_go_mapping = NULL
+    gene_names = unique(genes[, "refseq_mrna"])
+
+    i = 1
+    for(gene in gene_names){
+	go_terms = genes[genes[, gene_col] == gene, "go_id"]
+	if(length(go_terms) != 0){
+	    gene_id_go_mapping$gene = go_terms
+	    names(gene_id_go_mapping)[i] = gene
+	    i = i + 1
+	}
+    }
+    return(gene_id_go_mapping)
 }

### 2019-03-05 scripts/02_differential_expression_analysis.Rmd
DOC more on the paper

@@
 ```{r echo=FALSE, results="hide"}
+# Hide those results
 # Run the DE analysis on the rest of the influenza strain.
@@
 
+The distribution of number of genes found differentially expressed by
+considering each time-point independantly highlights the challenges of such
+approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
+genes found differentially expressed than timepoints 3H and 9H: this is likely
+due to biological or technical variances for specific genes at specific
+timepoints.
+
+As a summary, classic differential expression methods are appropriate for
+unordered treatments, but fail to use the temporal structure of the data.
+
 ### Time-course differential expression analysis between two groups
+
+To leverage this temporal structure, Storey et al [CITE FIXME] proposes to
+model time-course micro-array with splines functions, and to use a log-ratio
+likelihood test to detect differentially expressed genes. `moanin` extends
+this idea by providing functions to perform time-course differential
+expression analysis using `limma` contrasts.
 
@@
 
+# The function takes the data (data.frame or named matrix), the meta data
+# (data.frame containing a timepoint and group column, the first corresponding
+# to the time-course information, the latter corresponding to the
+# treatment).
 pvalues = moanin::timecourse_differential_expression_analysis(
@@
 ```{r echo=FALSE, results="hide"}
-# Rename stuff for saving.
+# Rename stuff for saving
 outname = ".results/pvalues.txt"

### 2019-03-05 scripts/03_clustering.Rmd
FIX pushd cannot be used in makefiles

@@
 
+The next step is to perform clustering.
+
 ```{r filter_genes}
@@
     # First, filter out any genes that doesn't have a log fold change of at
-    # least 2
-    # XXX I should actually filter on the max, but it is broken right now.
+    # least 2 at at least one of the time points.
     lfc_col_to_keep = colnames(de_analysis)[

### 2019-03-05 scripts/01_quality_control_normalization.Rmd
MAINT removing a lot of unused packages

@@
 library(moanin)
+library(ggplot2)
+library(stats)
+
 source("utils.R")
@@
 
-library(genefilter)
-library(dplyr)
-library(ggplot2)
-library(NMF)
-library(reshape2)
-library(limma)
 ```
@@
 # across samples using quantile normalization.
-RG = backgroundCorrect(RG, method="normexp")
-RG.q = normalizeBetweenArrays(RG, method="quantile")
+RG = limma::backgroundCorrect(RG, method="normexp")
+RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
 ```
@@
 par(mfrow=c(1, 2))
-plotDensities(RG, col="black", legend=FALSE)
-plotDensities(RG.q, col="black", legend=FALSE)
+limma::plotDensities(RG, col="black", legend=FALSE)
+limma::plotDensities(RG.q, col="black", legend=FALSE)
 ```
@@
 variance_per_genes = apply(data, 1, mad)
-min_variance = quantile(variance_per_genes, c(variance_cutoff))
+min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
 data = data[variance_per_genes > min_variance,]
@@
 ```{r echo=FALSE, fig.width=8, fig.height=4}
-hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
+graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
 	       ylab="# probes")
-abline(v=min_variance, col="#AB0000", lw=2)
+graphics::abline(v=min_variance, col="#AB0000", lw=2)
 ```
@@
 
-library(ggfortify)
-pca_data = prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
+pca_data = stats::prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
 percent_var = round(100 * attr(pca_data, "percentVar"))
@@
 # FIXME Add legend
-par(mfrow=c(2, 2))
-plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+graphics::par(mfrow=c(2, 2))
+graphics::plot(
+    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Group[meta$Group],
@@
     xlab="PC2", ylab="PC1")
-plot(pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+graphics::plot(
+    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
     col=ann_colors$Time[as.factor(meta$Time)],
@@
 
-plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
+graphics::plot(
+    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Group[meta$Group],
@@
     xlab="PC2", ylab="PC3")
-plot(pca_data$x[, "PC2"], pca_data$x[, "PC3"],
+graphics::plot(
+    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
     col=ann_colors$Time[as.factor(meta$Time)],
@@
 
-data_corr = cor(data, method="pearson")
+data_corr = stats::cor(data, method="pearson")
 
 meta$Time = as.factor(meta$Time)
-aheatmap(data_corr,
+NMF::aheatmap(
+    data_corr,
     Colv=NA, Rowv=NA,

### 2019-03-05 scripts/03_clustering.Rmd
MAINT removing a lot of unused packages

@@
 de_analysis = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)
+
+# Extract pvalues and log_fold change
+lfc_col_to_keep = colnames(de_analysis)[
+    grepl("-lfc_max", colnames(de_analysis))]
+lfc_max = de_analysis[, lfc_col_to_keep]
+
+pval_col_to_keep = colnames(de_analysis)[
+    grepl("-pval", colnames(de_analysis))]
+pvalues = de_analysis[, pval_col_to_keep]
 ```
@@
     # least 2 at at least one of the time points.
-    lfc_col_to_keep = colnames(de_analysis)[
-	grepl("-lfc_max", colnames(de_analysis))]
-    de_analysis = de_analysis[rowSums(abs(de_analysis[, lfc_col_to_keep]) > 2) > 0, ]
 
-    if(dim(de_analysis)[1] > n_genes_to_keep){
+    genes_to_keep = row.names(lfc_max[rowSums(lfc_max > 2) > 0, ])
+    pvalues = pvalues[genes_to_keep, ]
+
+    if(length(genes_to_keep) > n_genes_to_keep){
 
@@
 	# Filter out q-values for the pvalues table
-	pval_col_to_keep = colnames(de_analysis)[grepl("-pval", colnames(de_analysis))]
-	pvalues = de_analysis[, pval_col_to_keep]
 	fishers_pval = moanin:::fisher_method(pvalues)
@@
 	}
-    }else{
-	genes_to_keep = names(pvalues)
     }
@@
 centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)
-all_scores = data.frame(row.names=row.names(data))
-for(k in 1:n_clusters){
-    scores = moanin:::score_genes_centroid(data, centroids[k,])
-    all_scores[paste0("cluster_", k)] = scores
-}
+
+# XXX is there a bug here?
+# all_scores = data.frame(row.names=row.names(data))
+# for(k in 1:n_clusters){
+#     scores = moanin:::score_genes_centroid(data, centroids[k,])
+#     all_scores[paste0("cluster_", k)] = scores
+# }
 
 # Only assign labels to X% of the genes
-max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-labels = moanin:::row_argmin(all_scores)
-labels[genes_to_not_consider] = NA
+# max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
+# genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
+# labels = moanin:::row_argmin(all_scores)
+# labels[genes_to_not_consider] = NA
 ```
@@
 
-```{r}
-table(labels)
-```
-
 ```{r fig.height=8, fig.width=8}
@@
 ```
+
+## How to choose the number of clusters.
+
+A common question that arises when performing clustering is how to choose the
+number of clusters.
 

### 2019-03-05 scripts/install.R

@@ -45,18 +45,7 @@ if(!require(NMF)){
     install.packages("NMF")
 }
 
-if(!require(ggfortify)){
-    install.packages("ggfortify")
-}
-
-if(!require(reshape2)){
-    install.packages("reshape2")
-}
-
 source("https://bioconductor.org/biocLite.R")
-if(!require(genefilter)){
-    biocLite("genefilter")
-}
 
 if(!require(limma)){
     biocLite("limma")

### 2019-03-05 scripts/03_clustering.Rmd
DOC 03_clustering.Rmd

@@
 A common question that arises when performing clustering is how to choose the
-number of clusters.
+number of clusters. A choice for the number of clusters K depends on the goal. 
+In this particular case, the end goal is not the clustering, but to facilitate
+interpretation of the differential expression analysis step. As a result, the
+number of clusters should not exceed the number of gene sets the user wants to
+interpret. This allows to set a maximum number of clusters. Let us assume here
+that this number is 20 clusters.
+
+Once the maximum number of clusters is set, several strategies allow to
+identify the number of clusters:
+
+    - *Within cluster similarity as a function of the number of cluster* XXX
+      GAP statistics
+    - *Stability analysis*: this method assumes that the results of the
+      clustering will be more stable for the "true" number of clusters than
+      for other number of clusters.
 

### 2019-03-05 scripts/01_quality_control_normalization.Rmd
FIX 01_quality_control_normalization.Rmd doesn't depend on written data

@@
 
-```{r echo=FALSE, results="hide"}
+```{r echo=FALSE, results="hide", message=FALSE}
 # Loading dependencies
@@
 ```{r}
-
 # Now load in the metadata
-meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
+# meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
+data(shoemaker2015)
+meta = shoemaker2015$meta
 write.table(data, ".results/quantile_normalized.txt")

### 2019-03-06 scripts/02_differential_expression_analysis.Rmd
MAINT reactivating travis tests

@@
      use_voom_weights=FALSE)
+
 ```
@@
 
-```{r}
-print(colSums(pvalues[, grepl("-qval", colnames(pvalues))] < 0.05))
+```{r echo=FALSE, fig.width=6}
+num_de = colSums(qvalues < 0.05)
+names(num_de) = c(
+    "Kawasaki", "California", "Vietnam low-dose", "Vietname high-dose")
+barplot(num_de, col="black", ylab="Number of DE genes")
+
 ```

### 2019-03-06 scripts/04_downstream_analysis.Rmd
MAINT reactivating travis tests

@@
 library(stats)
+library(KEGGprofile)
+library(kableExtra)
+library(viridis)
+
 library(moanin)
-library(KEGGprofile)
 ```
@@
 ```{r echo=FALSE}
-library(kableExtra)
-library(viridis)
 
@@
 
-cluster = c(3)
+cluster = c(1, 2, 3)
 ontologies = c("BP", "NF", "CC")
@@
 ```{r echo=FALSE}
-library(kableExtra)
 
@@
     order(all_go_terms[, "Cluster"],
+	  all_go_terms[, "Ontology"],
 	  all_go_terms[, "resultFisher_padj"]), ]

### 2019-03-06 scripts/04_downstream_analysis.Rmd
ENH Let's attempt to deploy automatically the html to another private repository

@@
 	    scores,
-	    gene_id_go_mapping)
+	    gene_id_go_mapping, ontology=ontology)
 

### 2019-03-06 .build_tools/travis/after_success.sh

@@ -1,0 +1,36 @@
+#!/bin/bash
+# This script is meant to be called by the "after_success" step defined in
+# .travis.yml. See https://docs.travis-ci.com/ for more details.
+
+# License: 3-clause BSD
+
+# Ok, now that we succeeded, I want this to be published somewhere not too
+# public. How can I do that?
+
+set -e
+
+
+# commands taken shamelessly from https://gist.github.com/willprice/e07efd73fb7f13f917ea
+setup_git() {
+  git config --global user.email "travis@travis-ci.org"
+  git config --global user.name "Travis CI"
+}
+
+commit_website_files() {
+  git add . *.html
+  git commit --message "Travis build: $TRAVIS_BUILD_NUMBER"
+}
+
+upload_files() {
+  git remote add origin-pages https://${GH_TOKEN}@github.com:NelleV/2019timecourse-rnaseq-pipeline_html_outputs.git > /dev/null 2>&1
+  git push --quiet --set-upstream origin-pages master 
+}
+
+if [ "$TRAVIS_PULL_REQUEST" = "false" ]; then
+    git clone git@github.com:NelleV/2019timecourse-rnaseq-pipeline_html_outputs.git /tmp/html_outputs
+    cp -r scripts/reports/*.html /tmp/html_outputs
+    cd /tmp/html_outputs
+    setup_git
+    commit_website_files
+    upload_files
+fi;

### 2019-03-06 scripts/04_downstream_analysis.Rmd
FIX looping over pre-defined ontologies

--- 2019-03-06 scripts/04_downstream_analysis.Rmd
@@
 for(cluster in clusters){
-    for(ontology in c("BP", "NF", "CC")){ 
+    for(ontology in ontologies){ 
 	scores = as.numeric(labels == cluster)

### 2019-03-06 scripts/00_introduction.Rmd
FIX title on 00_introduction.Rmd
MAINT reactivating travis tests
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ---
-title: "a pipeline to analyse time-course gene expression data"
-output: html_document
+title: "Introduction: a pipeline to analyse time-course gene expression data"
+output: rmarkdown::html_vignette
 ---
@@
 The phenotypic diversity of cells is governed by a complex equilibrium between
-their genetic identity and their environmental interactions, and understanding
-the dynamics of gene expression is a fundamental question of biology. However,
-analysing time-course transcriptomic data raises challenging statistical and
-computational questions, requiring the development of novel methods and
-software.
+their genetic identity and their environmental interactions: Understanding the
+dynamics of gene expression is a fundamental question of biology. However,
+analysing time-course transcriptomic data raises unique challenging
+statistical and computational questions, requiring the development of novel
+methods and software.  Using as case study time-course transcriptomics data
+from mice exposed to different strains of influenza, ihis workflows provides a
+step-by-step tutorial of the methodology used to analyse time-course data:
+(1) normalization of the micro-array dataset; (2) differential expression
+analysis using functional data analysis; (3) clustering fo time-course data;
+(4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
 

### 2019-03-06 scripts/05_conclusion.Rmd
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

---
title: "Conclusion: A pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
---

# Author contributions

NV and EP wrote the workflow.

# Competing interests
The authors declare that they have no competing interests.

# Grant information

This research was funded in part by a Department of Energy (DOE) grant
(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
California, Berkeley [N.V.]; by a ENS-CFM Data Science Chair [E.P.].

*I confirm that the funders had no role in study design, data collection and
analysis, decision to publish, or preparation of the manuscript.*

# Acknowledgments

The authors thank Karthik Ram and the Ropensci community for valuable
feedback.

### 2019-03-06 .build_tools/travis/install.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+# This script is meant to be called by the "install" step defined in
+# .travis.yml. See http://docs.travis-ci.com/ for more details.
+# The behavior of the script is controlled by environment variabled defined
+# in the .travis.yml in the top level folder of the project.
+
+# License: 3-clause BSD
+
+set -e
+
+Rscript scripts/install.R
+
+pushd bin/moanin
+make install
+popd

### 2019-03-06 bin/moanin/R/de_analysis_utils.R

@@ -1,0 +1,115 @@
+library("limma")
+
+ALL_LFC_METHODS = c("sum", "max", "timely", "epicon", "abs_sum", "abs_squared_sum", "min")
+
+#' Estimates log fold change
+#'
+#' @param data The data in a matrix
+#' @param meta The metadata containing Groups and Time
+#' @param contrasts The contrasts to consider
+#' @param method ["sum", "max", "min", "timely", "epicon"]
+#'
+#' @export
+estimate_log_fold_change = function(data, meta, contrasts, method="epicon"){
+    # Should check that data and meta is sorted identically
+    meta = check_meta(meta)
+    # Should check that the method is a known method
+
+    if(!(method %in% ALL_LFC_METHODS)){
+	all_methods = paste(ALL_LFC_METHODS, sep=", ")
+	stop(
+	    paste("moanin::estimate_log_fold_change: '", method, "' is an unknown",
+		  " method to compute log fold change. Known methods are",
+		  all_methods, sep=""))
+    }
+
+    if(method == "timely"){
+	log_fold_changes = lfc_per_time(data, meta, contrasts)
+    }else if(method %in% c("sum", "max", "min", "abs_sum", "abs_squared_sum", "epicon")){
+	timely_lfc = lfc_per_time(data, meta, contrasts)
+	timely_lfc_meta = reconstruct_meta_from_lfc(timely_lfc)
+    	log_fold_changes = data.frame(row.names=row.names(data))
+	for(contrast in colnames(contrasts)){
+	    mask = timely_lfc_meta$Group == contrast
+	    if(method == "max"){
+		log_fold_changes[, contrast] = rowMax(abs(timely_lfc[, mask]))
+	    } else if(method == "min") {
+		log_fold_changes[, contrast] = rowMin(abs(timely_lfc[, mask])) 
+	    }else if(method == "abs_sum"){
+		log_fold_changes[, contrast] = rowSums(abs(timely_lfc[, mask]))
+	    }else if(method == "abs_squared_sum"){
+		log_fold_changes[, contrast] = rowSums(timely_lfc[, mask]**2)
+	    }else if(method == "epicon"){
+		log_fold_changes[, contrast] = (
+		    rowMeans(abs(timely_lfc[, mask])) * sign(rowSums(timely_lfc[, mask])))
+	    }else if(method == "sum"){
+		log_fold_changes[, contrast] = rowSums(timely_lfc[, mask])
+	    }
+	}
+
+    }
+
+    return(log_fold_changes)
[…]

### 2019-03-06 bin/moanin/R/splines.R

@@ -1,0 +1,142 @@
+# Splines utility
+
+library(splines)
+library(MASS)
+library(stats)
+
+
+#' Fit splines
+#'
+#' @param data the data
+#' @param basis the basis
+#' @param weights weigts
+#'
+#' @return beta coefficients
+#'
+#' @export
+fit_splines = function(data, basis, weights=NULL){
+    n = ncol(basis)
+    nr = nrow(data)
+    
+    if(!is.null(weights)){
+	beta = matrix(nrow=nr, ncol=n)
+	for(i in 1:nr){
+	    beta[i,] = stats::lm.wfit(basis, data[i,], weights[i,])$coefficients
+        }
+	row.names(beta) = row.names(data)
+    }else{
+	# Don't inverse directly the matrix
+	beta = t(stats::lm.fit(basis, t(data))$coefficients)
+    }
+    return(beta)
+}
+
+#' Fit and predict splines
+#'
+#' @param data the data
+#' @param basis the basis
+#' @param weights weigts
+#'
+#' @return y_fitted the fitted y values
+#'
+#' @export
+fit_predict_splines = function(data, basis, weights=NULL){
+    y_fitted = t(stats::lm.fit(basis, t(data))$fitted.values)
+    return(y_fitted)
+}
+
+
+rescale_values = function(y, meta, group=NULL){
+    if(is.null(group)){
+	ymin = row_min(y) 
+	y = y - ymin
+	ymax = row_max(y)
+	# We may have a division by 0 here
+	y = y / ymax
+    }else{
+	factors_to_consider = levels(unlist(meta[group]))
+	for(factor in factors_to_consider){
+	    mask = meta["Group"] == factor
+	    ymin = row_min(y[, mask]) 
+	    y[, mask] = y[, mask] - ymin
+	    ymax = row_max(y[, mask])
+	    # We may have a division by 0 here
+	    y[, mask] = y[, mask] / ymax
+	}
+    }
+    return(y)
+}
+
+
+# XXX It's wierd that this does not exists in R…
+# it probably exists but under another name?
+row_max = function(X){
+  return(apply(X, 1, max))
+}
+
+row_min = function(X){
+  return(apply(X, 1, min))
+}
+
+row_mean = function(X){
+    return(apply(X, 1, mean))
+}
+
+row_sum = function(X){
+    return(apply(X, 1, sum))
+}
+
+row_argmin = function(X){
+    return(apply(X, 1, which.min))
+}
+
+
[…]

### 2019-03-06 bin/moanin/R/timecourse_de_analysis.R

@@ -1,0 +1,220 @@
+# This file contains EDGE code adapted to work with limma contrasts
+
+center_data = function(y, ng_labels){
+  for(g in levels(ng_labels)){
+    whKeep = which(ng_labels == g)
+    sub_mean = rowMeans(y[, whKeep])
+    y[, whKeep] = y[, whKeep] - sub_mean
+  }
+  return(y)
+}
+
+compute_beta_null = function(X, beta, contrasts_coef){
+  ng = length(contrasts_coef)
+  df = ncol(X) / ng
+  contrasts_coef_ = rep(contrasts_coef, times=df)
+
+  # Reshape b so that each row corresponds to a group and drop the intercept
+  b_ = array(beta, dim=c(dim(beta)[1], ng, df))
+  # First start by constructing the matrix T
+  t_ = apply(b_, 1, function(x){colSums(x*contrasts_coef)})
+
+  # The observations can not be assumed to be balanced...
+  # We need to get rid of the intercept for this part
+  # FIXME don't invert this matrix…
+  part_K = MASS::ginv(t(X) %*% X)
+  K = part_K * contrasts_coef_**2
+
+  # We now need to sum all elements associated to the same pairs of splines.
+  # Which is, in a particular case every four elements in both directions.
+
+  K = sapply(1:df, function(jg) rowSums(K[, (jg-1)*ng + 1:ng]))
+  K = sapply(1:df, function(jg) colSums(K[(jg-1)*ng + 1:ng,]))
+
+  T_ = MASS::ginv(K) %*% t_
+
+  # We got T. Now, let's move on to the rest
+  tmp = as.array(rep(as.vector(T_), each=ng), dim=c(1, 1, 1))
+  dim(tmp) = c(ng * df, dim(beta)[1])
+  C_ = contrasts_coef * part_K %*% tmp
+
+  C_ = t(C_)
+  dim(C_) = c(dim(beta)[1], ng, df)
+  beta_null = b_ - C_
+
+  # Last step: reshape beta null so that it is of the same shape as beta
+  dim(beta_null) = dim(beta)
+  return(beta_null)
+}
+
+
+lrtStat = function(resNull, resFull, ng_labels=NULL) {
+  # FIXME I'm pretty sure that in the case of contrasts, the degrees of
+  # freedom computed here are wrong as they include part of the data that is
+  # not used for the test. This needs to be fixed
+  stat = 0
[…]

### 2019-03-06 bin/moanin/R/validation.R

@@ -1,0 +1,68 @@
+
+
+#' Check that the metadata provided is what we expect
+#'
+#' This method will raise errors if the metadata provided is not as expected.
+#'
+#' @param meta metadata
+#' @param check_replicates boolean, optional, default: FALSE
+#'	  If TRUE, checks whether metadata contains a column Replicate
+#' @return meta returns the metadata with additional columns if necessary.
+#'
+#' @keywords internal
+check_meta = function(meta, check_replicates=FALSE){
+    metadata_column_names = colnames(meta)
+    if(!("Group" %in% metadata_column_names)){
+	stop(
+	    "Metadata doesn't contain expected information." +
+	    " Group column is missing.")
+    }
+
+    if(!("Time" %in% metadata_column_names)){
+	stop(
+	    "Metadata doesn't contain expected information." +
+	    " Time column is missing.")
+    }
+
+    if(check_replicates & !("Replicate" %in% metadata_column_names)){
+	stop(
+	    "Metadata doesn't contain expected information." +
+	    "Replicate column is missing")
+    }
+
+    # Just create this one.
+    if(!("WeeklyGroup" %in% metadata_column_names)){
+	meta["WeeklyGroup"] = as.factor(make.names(meta$Group:as.factor(meta$Time)))
+    }
+    return(meta) 
+}
+
+
+#' Check data and meta
+#' @keywords internal
+check_data_meta = function(data, meta){
+    dim_data = dim(data)
+    dim_meta = dim(meta)
+    data = as.matrix(data)
+    if(dim_meta[1] != dim_data[2]){
+	stop(
+	    "Data and metadata are inconsistent. Data is of shape (Xx"+
+	    "Metadata is of shape XX")
+    }
+
+    if(!is.numeric(data)){
+	stop("Data should be of type numeric")
+    }
+    meta = check_meta(meta)
+}
+
+
+#' Check is 2D
+#'
+#' @keywords internal
+check_is_2d = function(X){
+    dim_data = dim(X)
+    if(is.null(dim_data)){
+	stop("Data is expected to be 2D. No dimension found.")
+    }
+}

### 2019-03-06 bin/moanin/R/visualization.R

@@ -1,0 +1,64 @@
+library(viridis)
+library(graphics)
+
+#' Plotting centroids
+#'
+#' @param centroids matrix (k, t) containing the centroids
+#' @param meta	data.frame (t, n) containing the metadata.
+#' @param colors vector, optional, default NULL
+#'		vector of colors
+#'
+#' @export
+plot_centroids = function(centroids, meta, colors=NULL){
+    n_centroids = dim(centroids)[1]
+    if(n_centroids <= 3){
+        graphics::par(
+	    mfrow=c(n_centroids, 1),
+            mar=c(0.1, 0.1, 0.1, 0.1))
+    }else if(n_centroids <= 6){
+        n_col = ceiling(n_centroids / 2)
+        graphics::par(
+	    mfrow=c(n_col, 2),
+            mar=c(0.1, 0.1, 0.1, 0.1))
+    }else if(n_centroids <= 12){
+        ncol = ceiling(n_centroids / 3)
+        graphics::par(
+	    mfrow=c(ncol, 3),
+            mar=c(0.1, 0.1, 0.1, 0.1))
+    }else{
+        nrow = round(n_centroids ** 0.5)
+        ncol = ceiling(n_centroids / nrow)
+        graphics::par(
+	    mfrow=c(ncol, nrow),
+            mar=c(0.1, 0.1, 0.1, 0.1))
+    }
+
+    for(i in 1:n_centroids){
+        plot_centroid_individual(centroids[i, ], meta, colors=colors)
+    }
+}
+
+
+plot_centroid_individual = function(centroid, meta, colors=NULL){
+    groups = levels(meta$Group)
+
+    xrange = range(meta$Time)
+    yrange = range(centroid)
+    
+    graphics::plot(xrange, yrange, type="n")
+    if(is.null(colors)){
+        colors = viridis::viridis(length(groups))
+    }
+    
+    for(i in 1:length(groups)){
+        group = groups[i]
+        color = colors[i]
+        
+        mask = meta$Group == group
+        time = meta$Time[mask]
+        indx = order(time)
+        graphics::lines(time[indx], centroid[mask][indx], type="b",
+		        col=color, pch=16,
+			lwd=2)
+    }         
+} 

### 2019-03-06 bin/moanin/R/weekly_de_analysis.R

@@ -1,0 +1,67 @@
+library("limma")
+library("stats")
+library("edgeR")
+
+
+#' Fit weekly differential expression analysis
+#'
+#' @param counts Gene expression data
+#' @param meta Metadata data frame
+#' @param contrasts Contrast to use.
+#' @param use_voom_weights boolean: whether to use voom weights or not
+#'
+#' @export
+weekly_differential_expression_analysis = function(counts, meta,
+						   contrasts,
+						   use_voom_weights=TRUE){
+    meta = check_meta(meta)
+
+    design = stats::model.matrix(~WeeklyGroup + 0, data=meta)
+
+    cleaned_colnames = gsub("WeeklyGroup", "", colnames(design))
+    colnames(design) = cleaned_colnames
+
+    allcontrasts = limma::makeContrasts(
+        contrasts=contrasts,
+        levels=design)
+
+    if(use_voom_weights){
+        y = edgeR::DGEList(counts=counts)
+	y = edgeR::calcNormFactors(y, method="upperquartile")
+        v = limma::voom(y, design, plot=FALSE)
+	v = limma::lmFit(v)
+    }else{
+	v = limma::lmFit(counts, design)	
+    }
+
+    fit = limma::contrasts.fit(v, allcontrasts)
+    fit = limma::eBayes(fit)
+    contrast_names = colnames(fit$p.value)
+    fit$adj.p.value = stats::p.adjust(fit$p.value, method="BH")
+    dim(fit$adj.p.value) = dim(fit$p.value)
+    colnames(fit$adj.p.value) = contrast_names
+
+    combine_results = function(ii, fit2){
+	contrast_formula = contrasts[ii]
+	de_analysis = data.frame(row.names=row.names(counts))
+
+        base_colname = gsub(" ", "", contrast_formula, fixed=TRUE)
+	colname_pval = paste(base_colname, "-pval", sep="")
+	colname_qval = paste(base_colname, "-qval", sep="")
+	colname_lfc = paste(base_colname, "-lfc", sep="")
+
+        tt = limma::topTable(
+            fit2, coef=ii, number=length(rownames(fit2$coef)),
+            p.value=1, adjust.method="none",
+            genelist=rownames(fit2$coef))
+	de_analysis[colname_pval] = fit2$p.value[, contrast_formula]
+	de_analysis[colname_qval] = fit2$adj.p.value[,  contrast_formula]
+	de_analysis[colname_lfc] = tt$logFC
[…]

### 2019-03-06 bin/moanin/tests/testthat/test-de_analysis_utils.R

@@ -1,0 +1,45 @@
+library("moanin")
+
+context("moanin::de_analysis_utils.R")
+
+
+test_that("Estimating log fold change smoke tests", {
+    # This is just a smoke test.
+    data(shoemaker2015)
+    data = shoemaker2015$data
+    meta = shoemaker2015$meta
+
+    # Reduce the data set
+    data = data[1:10, ]
+    methods = moanin:::ALL_LFC_METHODS
+
+    contrasts = limma::makeContrasts(contrasts="C-K",
+				     levels=meta$Group)
+    for(method in methods){
+        expect_silent(
+	    estimate_log_fold_change(
+		data, meta, contrasts, method=method))
+    }
+
+    # Now same test, but several contrasts
+    contrasts = limma::makeContrasts(contrasts=c("C-K", "C-M"),
+				     levels=meta$Group)
+    for(method in methods){
+        expect_silent(
+	    estimate_log_fold_change(
+		data, meta, contrasts, method=method))
+    }
+
+})
+
+test_that("Estimating log fold change with unknown error", {
+    data(shoemaker2015)
+    data = shoemaker2015$data
+    meta = shoemaker2015$meta
+
+    # Reduce the data set
+    data = data[1:10, ]
+    contrasts = c("C-K")
+    expect_error(estimate_log_fold_change(data, meta, contrast, method="hahaha"))
+
+})

### 2019-03-06 bin/moanin/tests/testthat/test-splines.R

@@ -1,0 +1,43 @@
+library("moanin")
+
+context("moanin::splines.R")
+
+test_that("splines:align_data_onto_centroid", {
+    set.seed(42)
+    n_samples = 20
+    n_genes = 5
+    centroid = runif(n_samples)
+
+    shift = runif(n_genes)
+    scale = 1 + runif(n_genes)
+
+    data = rep(centroid, each=n_genes)
+    dim(data) = c(n_genes, n_samples)
+    expect_equal(data, align_data_onto_centroid(data, centroid))
+
+    # Ok, now let's make this a bit more complicated.
+    scale = rep(scale, times=n_samples)
+    dim(scale) = dim(data)
+    shift = rep(shift, times=n_samples)
+    dim(shift) = dim(data)
+
+    shifted_scaled_data = scale * data + shift
+    shifted_scaled_centroid = scale * centroid + shift
+    expect_equal(data,
+		 align_data_onto_centroid(shifted_scaled_data, centroid))
+    expect_error(align_data_onto_centroid(data[, 1:10], centroid))
+})
+
+
+test_that("splines:rescale_values", {
+
+    n_genes = 5
+    data(shoemaker2015)
+    data = shoemaker2015$data
+    meta = shoemaker2015$meta
+    data = data[1:n_genes,]
+
+    rescaled_data = rescale_values(data, meta)
+    expect_equal(rep(0, n_genes), as.vector(row_min(rescaled_data)))
+    expect_equal(rep(1, n_genes), as.vector(row_max(rescaled_data)))
+})

### 2019-03-06 bin/moanin/tests/testthat/test-validation.R

@@ -1,0 +1,40 @@
+library(moanin)
+
+context("Testing validation functions")
+
+test_that("validation:check_meta", {
+    data(shoemaker2015)
+    meta = shoemaker2015$meta
+    expect_silent(check_meta(meta))
+
+    meta_without_group = subset(meta, select=c("Replicate", "Time"))
+    meta_without_time = subset(meta, select=c("Group", "Replicate"))
+    expect_error(check_meta(meta_without_group))
+    expect_error(check_meta(meta_without_time))
+
+    meta_without_replicate = subset(meta, select=c("Time", "Group"))
+    expect_silent(check_meta(meta_without_replicate))
+    expect_error(check_meta(meta_without_replicate, check_replicates=TRUE))
+})
+
+test_that("validation:check_data_meta", {
+    data(shoemaker2015)
+    data = shoemaker2015$data
+    meta = shoemaker2015$meta
+
+    data = data[1:10, ]
+
+    # Running check_data_meta should work fine
+    expect_silent(check_data_meta(data, meta))
+
+    expect_error(
+	check_data_meta(data[, 2:10], meta))
+})
+
+
+test_that("validation:check_is_2d", {
+    data_2d = matrix(1:9, nrow=3)
+    expect_silent(check_is_2d(data_2d))
+    expect_silent(check_is_2d(as.data.frame(data_2d)))
+    expect_error(check_is_2d(1:9))
+})

### 2019-03-06 scripts/.scripts/compile_rmarkdown.R

@@ -1,0 +1,7 @@
+library(rmarkdown)
+
+args = commandArgs(trailingOnly=TRUE)
+filename = args[1]
+outname = args[2]
+
+render(filename, output_file=outname)

### 2019-03-06 scripts/GO_terms_analysis.R

@@ -1,0 +1,78 @@
+library(topGO)
+
+
+#' Find enriched GO terms
+#'
+#' @param labels
+#' @param gene_id_to_go
+#' @param ontology string, optional, default: BP
+#'	specficies which ontology to use. Can be 'BP', 'CC', or 'NF'
+#' @param weighted, boolean, optional, default: FALSE
+#'	Whether to use the weighted algorithm or not.
+#' @param node_size integer, optional, default: 10
+#'	Consider only GO terms with node_size number of genes.
+#' @param ontology: BP, CC, NF
+find_enriched_go_terms = function(labels, gene_id_to_go,
+				  ontology="BP", 
+				  weighted=FALSE,
+				  node_size=10){
+    if(!(ontology %in% c("BP", "CC", "NF"))){
+	error_message = paste(
+	    "moanin::find_enriched_go_terms: Ontology should be 'BP', 'CC',",
+	    "or 'NF'. Ontology provided is",
+	    ontology)
+	stop(error_message)
+    }
+
+
+    getTopDiffGenes = function(data, cutOff=NULL){
+      return(data < 0.5)
+    }
+
+    GOdata = new("topGOdata", ontology=ontology, allGenes=labels, nodeSize=node_size,
+		 geneSel=getTopDiffGenes,
+		 annot=annFUN.gene2GO, gene2GO=gene_id_to_go)
+
+    if(weighted){
+        resultFisher = runTest(GOdata, algorithm="classic", statistic="fisher")
+    }else{
+        resultFisher = runTest(GOdata, algorithm="weight", statistic="fisher")
+    }
+    n_nodes = length(resultFisher@score)
+
+    allRes = GenTable(
+	GOdata, 
+	resultFisher=resultFisher,
+	orderBy="resultFisher", ranksOf="resultFisher",
+	topNodes=n_nodes)
+        
+    # P-value correct
+    allRes[, "resultFisher_padj"] = stats::p.adjust(allRes$resultFisher, method="BH")
+    wh = which(allRes[, "resultFisher_padj"] <= 0.05)
+
+    allRes = allRes[wh,]
+    wh = which(apply(allRes[, c("Significant", "Expected")],
+    	              1, function(x){x["Significant"] > x["Expected"]}))
+    allRes = allRes[wh,]
+
+    return(allRes)
+}
+
+
+#' Create the Gene to GO Term mapping
+#'
[…]

### 2019-03-06 scripts/install.R

@@ -1,0 +1,86 @@
+
+
+# This is a bit annoying, but I'm hoping to save time on travis-ci
+if(!require("devtools")){
+    install.packages("devtools")
+}
+
+if(!require(ggfortify)){
+    install.packages("ggfortify")
+}
+
+if(!require(kableExtra)){
+    install.packages("kableExtra")
+}
+
+if(!require(testthat)){
+    install.packages("testthat")
+}
+
+if(!require(viridis)){
+    install.packages("viridis")
+}
+
+if(!require(knitr)){
+    install.packages("knitr")
+}
+
+if(!require(rmarkdown)){
+    install.packages("rmarkdown")
+}
+
+if(!require(roxygen2)){
+    install.packages("roxygen2")
+}
+
+if(!require(ClusteR)){
+    install.packages("ClusteR")
+}
+
+if(!require(splines)){
+    install.packages("splines")
+}
+
+if(!require(NMF)){
+    install.packages("NMF")
+}
+
+source("https://bioconductor.org/biocLite.R")
+
+if(!require(limma)){
+    biocLite("limma")
+}
+
+if(!require(edgeR)){
+    biocLite("edgeR")
+}
+
+if(!require(edge)){
+    biocLite("edge")
+}
+
+if(!require(EDASeq)){
+    biocLite("EDASeq")
+}
+
+if(!require(topGO)){
+    biocLite("topGO")
+}
+
+if(!require(Biobase)){
+    biocLite("Biobase")
+}
+
+if(!require(MASS)){
+    biocLite("MASS")
+}
+
+if(!require("biomaRt")){
+    biocLite("biomaRt")
+}
+
+if(!require(KEGGprofile)){
+    biocLite("KEGGprofile")
+}
+
+

### 2019-03-06 scripts/step0_preprocessing/process_epicon_leaf.R

@@ -1,0 +1,45 @@
+
+###############################################################################
+# Options
+#
+
+data_dir = "data/varoquaux2019/leaf/"
+results_dir = "results/varoquaux2019/leaf/"
+
+###############################################################################
+# Load the data
+counts = read.delim(file.path(data_dir, "data.tsv"),
+    		    sep="\t",
+		    stringsAsFactors=FALSE,
+		    check.names=FALSE)
+meta = read.table(file.path(data_dir, "meta.tsv"),
+		  check.names=FALSE)
+
+###############################################################################
+# The pipeline requires:
+#   - a Group column in the metadata file, containing the "group" used to do
+#     the time-course differential expression analysis
+#   - a Time colmun, containing the time variable (in the case of EPICON, the
+#     week
+meta$Time = as.numeric(as.character(meta$Week))
+ng_labels = as.factor(
+  make.names(meta$Condition:meta$Genotype))
+weekly_ng_labels = as.factor(
+    make.names(meta$Condition:meta$Genotype:as.factor(meta$Time)))
+meta$Group = ng_labels
+meta$WeeklyGroup = weekly_ng_labels
+
+
+###############################################################################
+# Now, just save in the folder
+outname = file.path(results_dir, "data.tsv")
+write.table(counts, file=outname, sep="\t", row.names=TRUE, col.names=TRUE)
+
+outname = file.path(results_dir, "meta.tsv")
+write.table(meta, file=outname, sep="\t", row.names=TRUE, col.names=TRUE)
+
+file.copy(file.path(data_dir, "contrasts"),
+	  file.path(results_dir, "contrasts"))
+
+file.copy(file.path(data_dir, "config.yml"),
+	  file.path(results_dir, "config.yml"))

### 2019-03-06 scripts/step0_preprocessing/process_epicon_root.R

@@ -1,0 +1,46 @@
+
+###############################################################################
+# Options
+#
+
+data_dir = "data/varoquaux2019/root/"
+results_dir = "results/varoquaux2019/root/"
+
+###############################################################################
+# Load the data
+counts = read.delim(file.path(data_dir, "data.tsv"),
+    		    sep="\t",
+		    stringsAsFactors=FALSE,
+		    check.names=FALSE)
+meta = read.table(file.path(data_dir, "meta.tsv"),
+		  check.names=FALSE)
+
+###############################################################################
+# The pipeline requires:
+#   - a Group column in the metadata file, containing the "group" used to do
+#     the time-course differential expression analysis
+#   - a Time colmun, containing the time variable (in the case of EPICON, the
+#     week
+meta$Time = as.numeric(as.character(meta$Week))
+ng_labels = as.factor(
+  make.names(meta$Condition:meta$Genotype))
+meta$Group = ng_labels
+weekly_ng_labels = as.factor(
+    make.names(meta$Condition:meta$Genotype:as.factor(meta$Time)))
+meta$Group = ng_labels
+meta$WeeklyGroup = weekly_ng_labels
+
+
+###############################################################################
+# Now, just save in the folder
+outname = file.path(results_dir, "data.tsv")
+write.table(counts, file=outname, sep="\t", row.names=TRUE, col.names=TRUE)
+
+outname = file.path(results_dir, "meta.tsv")
+write.table(meta, file=outname, sep="\t", row.names=TRUE, col.names=TRUE)
+
+file.copy(file.path(data_dir, "contrasts"),
+	  file.path(results_dir, "contrasts"))
+
+file.copy(file.path(data_dir, "config.yml"),
+	  file.path(results_dir, "config.yml"))

### 2019-03-06 scripts/step3_DE_analysis/run_development_differential_expression_analysis.R

@@ -1,0 +1,73 @@
+library("limma")
+library("edge")
+library("devtools")
+load_all("../../bin/moanin")
+
+# FIXME XXX it works, but there's a lot of hacks going on…
+
+
+###############################################################################
+# Options
+#
+
+args = commandArgs(trailingOnly=TRUE)
+data_dir = args[1]
+# For debugging purposes
+# data_dir = "results/varoquaux2019/leaf/"
+
+config_file = file.path(data_dir, "config.yml")
+config = config::get(file=config_file)
+
+var_filtering = unlist(config["var_filtering"] )
+var_cutoff = unlist(config["var_cutoff"])
+take_log = unlist(config["take_log"])
+contrasts = as.vector(unlist(strsplit(unlist(config["development_contrasts"]), ",")))
+
+###############################################################################
+# Load the data
+counts = read.delim(file.path(data_dir, "counts_normed.tsv"),
+    		    sep="\t",
+		    stringsAsFactors=FALSE,
+		    check.names=FALSE)
+meta = read.table(file.path(data_dir, "meta.tsv"),
+		  check.names=FALSE)
+
+if(take_log){
+    counts = log2(counts + 1)
+}
+
+###############################################################################
+# Var filtering
+#
+# FIXME
+
+
+###############################################################################
+# DE analysis
+
+meta$Time = as.numeric(as.character(meta$Week))
+ng_labels = as.factor(
+  make.names(meta$Condition:meta$Genotype))
+meta$Group = ng_labels
+
+
+de_analysis = data.frame(row.names=row.names(counts))
+for(contrast_formula in contrasts){
+    contrast = limma::makeContrasts(
+	contrast_formula,
+	levels=levels(meta$Group))
+    mask = with(meta, Group == contrast_formula)
+    model = edgeWithContrasts(as.matrix(counts), meta, contrast, developmental=TRUE,
+			      mask=mask)
+    contrast_name = gsub(" ", "", contrast_formula, fixed=TRUE)
+    contrast_name_pval = paste(contrast_name, "-pval", sep="")
+    contrast_name_qval = paste(contrast_name, "-qval", sep="")
+    de_analysis[contrast_name_pval] = model$pval
+    de_analysis[contrast_name_qval] = model$qval
+}
+
+outname = file.path(data_dir, "development_de_analysis.tsv")
+write.table(
+    de_analysis,
+    file=outname,
+    sep="\t", row.names=TRUE, col.names=TRUE)

### 2019-03-06 scripts/step3_DE_analysis/run_weekly_differential_expression_analysis.R

@@ -1,0 +1,47 @@
+library("limma")
+source("utils_limma_voom.R")
+
+
+###############################################################################
+# Options
+#
+#args = commandArgs(trailingOnly=TRUE)
+#data_dir = args[1]
+# For debugging purpose
+data_dir = "results/varoquaux2019/leaf"
+
+config_file = file.path(data_dir, "config.yml")
+config = config::get(file=config_file)
+
+var_filtering = unlist(config["var_filtering"] )
+var_cutoff = unlist(config["var_cutoff"])
+take_log = unlist(config["take_log"])
+contrasts = as.vector(unlist(strsplit(unlist(config["weekly_contrasts"]), ",")))
+
+
+###############################################################################
+# Load the data
+counts = read.delim(file.path(data_dir, "counts_normed.tsv"),
+    		    sep="\t",
+		    stringsAsFactors=FALSE,
+		    check.names=FALSE)
+meta = read.table(file.path(data_dir, "meta.tsv"),
+		  check.names=FALSE)
+
+
+###############################################################################
+# Var filtering
+#
+# FIXME
+
+
+###############################################################################
+# DE analysis
+
+de_analysis = fit_limma_voom(counts, meta, contrasts)
+
+outname = file.path(data_dir, "weekly_de_analysis.tsv")
+write.table(
+    de_analysis,
+    file=outname,
+    sep="\t", row.names=TRUE, col.names=TRUE)

### 2019-03-06 scripts/step4_clustering/clustering_kmeans.R

@@ -1,0 +1,66 @@
+library(splines)
+library(stats)
+source("utils_splines.R")
+library(devtools)
+load_all("../../bin/moanin")
+
+###############################################################################
+# Options
+
+n_clusters = 20
+df = 6
+data_dir = "results/varoquaux2019/leaf/"
+filter_genes = TRUE
+n_genes_to_keep = 5000
+percentage_genes_to_label = 0.5
+
+###############################################################################
+# Load the data
+counts = read.delim(file.path(data_dir, "counts_normed.tsv"),
+    		    sep="\t",
+		    stringsAsFactors=FALSE,
+		    check.names=FALSE)
+meta = read.table(file.path(data_dir, "meta.tsv"),
+		  check.names=FALSE)
+pvalues = read.table(file.path(data_dir, "timecourse_de_analysis.tsv"),
+		     sep="\t",
+		     check.names=FALSE)
+
+###############################################################################
+# Select "best" genes
+
+if(filter_genes){
+    # Filter out q-values for the pvalues table
+    col_to_keep = colnames(pvalues)[grepl("-pval", colnames(pvalues))]
+    pvalues = pvalues[, col_to_keep]
+    fishers_pval = fisher_method(pvalues)
+    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
+    counts = counts[genes_to_keep, ]
+}
+
+meta$Time = as.numeric(as.character(meta$Week))
+ng_labels = as.factor(
+  make.names(meta$Condition:meta$Genotype))
+meta$Group = ng_labels
+
+###############################################################################
+# Fit the splines
+
+full_model = ~Group:ns(Time, df=df) + Group + 0
+X = model.matrix(full_model, data=meta)
+y = as.matrix(log2(counts + 1))
+
+fitted_counts = fit_predict_splines(y, X)
+rescaled_fitted_counts = rescale_values(fitted_counts, meta)
+kmeans_clusters = kmeans(rescaled_fitted_counts, n_clusters, nstart=10)
+all_scores = data.frame(row.names=row.names(counts))
+for(k in 1:n_clusters){
+    scores = score_genes_centroid(y, kmeans_clusters$centers[k,])
+    all_scores[paste0("cluster_", k)] = scores
+}
+
+# Only assign labels to X% of the genes
+max_score = quantile(row_min(all_scores), c(percentage_genes_to_label))
[…]

### 2019-03-17 scripts/02_differential_expression_analysis.Rmd
ENH Updating the rest of the functions to use the object

@@
 ```
+```{r}
+splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
+```
 
@@
 weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, meta, contrasts,
+     data, splines_model, contrasts,
      use_voom_weights=FALSE)
@@
 K_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, meta, contrasts,
+     data, splines_model, contrasts,
      use_voom_weights=FALSE)
@@
 C_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, meta, contrasts,
+     data, splines_model, contrasts,
      use_voom_weights=FALSE)
@@
 VH_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, meta, contrasts,
+     data, splines_model, contrasts,
      use_voom_weights=FALSE)
@@
 pvalues = moanin::timecourse_differential_expression_analysis(
-    data, meta, contrasts)
+    data, splines_model, contrasts)
 qvalues = apply(pvalues, 2, stats::p.adjust)
@@
 log_fold_change_epicon = moanin::estimate_log_fold_change(
-    data, meta, contrasts,  method="epicon")
+    data, splines_model, contrasts,  method="epicon")
 
 log_fold_change_max = moanin::estimate_log_fold_change(
-    data, meta, contrasts, method="max")
+    data, splines_model, contrasts, method="max")
 ```

### 2019-03-17 bin/moanin/R/splines_model.R

@@ -37,4 +37,5 @@ create_splines_model = function(meta, formula=NULL, basis=NULL,
     splines_model$"basis" = basis
     splines_model$"meta" = meta
     splines_model$"formula" = formula
+    return(splines_model)
 }

### 2019-03-17 bin/moanin/R/timecourse_de_analysis.R

@@ -9,9 +9,9 @@ center_data = function(y, ng_labels){
   return(y)
 }
 
-compute_beta_null = function(X, beta, contrasts_coef){
+compute_beta_null = function(basis, beta, contrasts_coef){
   ng = length(contrasts_coef)
-  df = ncol(X) / ng
+  df = ncol(basis) / ng
   contrasts_coef_ = rep(contrasts_coef, times=df)
 
   # Reshape b so that each row corresponds to a group and drop the intercept
@@ -21,8 +21,8 @@ compute_beta_null = function(X, beta, contrasts_coef){
 
   # The observations can not be assumed to be balanced...
   # We need to get rid of the intercept for this part
-  # FIXME don't invert this matrix…
-  part_K = MASS::ginv(t(X) %*% X)
+  # FIbasisME don't invert this matrix…
+  part_K = MASS::ginv(t(basis) %*% basis)
   K = part_K * contrasts_coef_**2
 
   # We now need to sum all elements associated to the same pairs of splines.
@@ -49,7 +49,7 @@ compute_beta_null = function(X, beta, contrasts_coef){
 
 
 lrtStat = function(resNull, resFull, ng_labels=NULL) {
-  # FIXME I'm pretty sure that in the case of contrasts, the degrees of
+  # FIbasisME I'm pretty sure that in the case of contrasts, the degrees of
   # freedom computed here are wrong as they include part of the data that is
   # not used for the test. This needs to be fixed
   stat = 0
@@ -84,16 +84,16 @@ lrtStat = function(resNull, resFull, ng_labels=NULL) {
   return(stat)
 }
 
-compute_pvalue = function(X, y, beta, beta_null, ng_labels,
+compute_pvalue = function(basis, y, beta, beta_null, ng_labels,
 			  n_groups=NULL,
 			  n_samples=NULL,
 			  degrees_of_freedom=NULL,
 			  statistics="lrt",
 			  df2=NULL, weights=NULL){
 
-    fitFull = beta %*% t(X)
+    fitFull = beta %*% t(basis)
 
-    fitNull = beta_null %*% t(X)
+    fitNull = beta_null %*% t(basis)
 
     if(!is.null(weights)){
       resNull = weights^(1/2) * (y - fitNull)
@@ -106,12 +106,12 @@ compute_pvalue = function(X, y, beta, beta_null, ng_labels,
     # estimate degrees of freedom.
     if(is.null(n_groups)){
[…]

### 2019-03-17 bin/moanin/tests/testthat/test-timecourse_de_analysis.R

@@ -1,0 +1,35 @@
+library(moanin)
+library(limma)
+
+context("moanin::de_analysis.R")
+
+test_that("time-course DE analysis", {
+    data(shoemaker2015)
+    data = shoemaker2015$data
+    meta = shoemaker2015$meta
+
+    splines_model = create_splines_model(meta)
+
+    # Limit ourselves to the top 5000 genes
+    data = data[1:5000, ]
+
+    # Do only one contrast
+    contrast_formulas = c("K-M")
+    contrast = limma::makeContrasts(
+	contrasts=contrast_formulas,
+	levels=levels(meta$Group))
+
+    expect_silent(moanin::timecourse_differential_expression_analysis(
+	as.matrix(data), splines_model, contrasts=contrast))
+
+    # Now test with several contrasts
+    contrast_formulas = c("K-M", "C-M")
+    contrasts = limma::makeContrasts(
+	contrasts=contrast_formulas,
+	levels=levels(meta$Group))
+
+    expect_silent(moanin::timecourse_differential_expression_analysis(
+    	as.matrix(data), splines_model, contrasts=contrast))
+
+
+})

### 2019-03-17 scripts/00_introduction.Rmd

@@
 
-- TODO why is time-course gene expression analysis important for biology. Why
-  are novel workflows necessary.
-
 Time-course gene expression analysis workflows include steps used more broadly
-for gene expression data analasys such as quality control (QC), exploratory
+for gene expression data analysis such as quality control (QC), exploratory
 data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
@@
 (2) differential expression analysis of time-course data; (3) clustering; (4)
-downstream analysis of clusters with GO term and pathway enrichment.
+downstream analysis of clusters with GO term, pathway, and motif enrichment.
 
-XXX Add the packages used in the workflow
+This workflow relies on several standard packages for analysing gene
+expression data, some specific for time-course data, others broadly used by
+the community. We also provide alongside this workflow a package that nicely
+wraps all calls to dependencies with a common API by defining a model object
+holding all necessary metadata to perform all steps of the pipeline.
 
@@
 
+The authors find that
+
 **XXX Add summary of the key findings of the the paper**

### 2019-03-17 scripts/03_clustering.Rmd
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 pvalues = de_analysis[, pval_col_to_keep]
+
+splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
 ```
@@
 
-The next step is to perform clustering.
+The next step is to perform clustering. First, we reduce the set of genes of
+interest to genes that are (1) significant; (2) "highly" differentially
+expressed. First, we select the genes that have a log-fold change of at least
+2 across at least one experiment for one time point. We thus use the object
+`lfc_max`.
+
+XXX
 
@@
 kmeans_clusters = moanin:::splines_kmeans(
-    y, meta, n_clusters=20,
+    y, splines_model, n_clusters=20,
     random_seed=42,
-    n_init=20,
-    degrees_of_freedom=df)
+    n_init=20)
 
 # Rescale the centroids
-centroids = moanin:::rescale_values(kmeans_clusters$centroids, meta)
+centroids = moanin:::rescale_values(kmeans_clusters$centroids, splines_model)
 
@@
 ```{r fig.height=8, fig.width=8}
-moanin:::plot_centroids(centroids, meta,
+moanin:::plot_centroids(centroids, splines_model,
 			colors=as.vector(unlist(ann_colors["Group"])))

### 2019-03-17 scripts/01_quality_control_normalization.Rmd
MAINT rename Time -> Timepoint

@@
     pca_data$x[, "PC2"], pca_data$x[, "PC1"],
-    col=ann_colors$Time[as.factor(meta$Time)],
+    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
@@
     pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Time[as.factor(meta$Time)],
+    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
@@
 Then, we plot the pearson correlation across each samples. We order the
-samples by their Group (treatment) and Time (the time of sampling).
+samples by their Group (treatment) and Timepoint (the time of sampling).
 Additionally, in this example, we order each treatment by strength of the
@@
   meta$Group,
-  meta$Time,
+  meta$Timepoint,
   meta$Replicate)
@@
 
-meta$Time = as.factor(meta$Time)
+meta$Timepoint = as.factor(meta$Timepoint)
 NMF::aheatmap(
@@
     annCol=meta[,
-		c("Group", "Time")],
+		c("Group", "Timepoint")],
     annLegend=FALSE, 

### 2019-03-17 scripts/02_differential_expression_analysis.Rmd
MAINT rename Time -> Timepoint

@@
 ```{r weekly_differential_analysis}
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Time))
+meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Timepoint))
 
@@
     barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
-	    xlab="Time", ylab="Number of DE genes", main=title)
+	    xlab="Timepoint", ylab="Number of DE genes", main=title)
 

### 2019-03-17 bin/moanin/R/de_analysis_utils.R

@@ -85,17 +85,18 @@ data_summarize_per_time = function(data, meta){
 
 # XXX helper function to reconstruct metadat from adat
 reconstruct_meta_from_lfc = function(data_per_time){
-    meta_per_time = t(as.data.frame(strsplit(colnames(data_per_time), ".", fixed=TRUE)))
+    meta_per_time = t(
+	as.data.frame(strsplit(colnames(data_per_time), ".", fixed=TRUE)))
     row.names(meta_per_time) = colnames(data_per_time)
-    colnames(meta_per_time) = c("Group", "Time")
+    colnames(meta_per_time) = c("Group", "Timepoint")
     meta_per_time = as.data.frame(meta_per_time)
-    meta_per_time[, "Time"] = as.numeric(meta_per_time[, "Time"])
+    meta_per_time[, "Timepoint"] = as.numeric(meta_per_time[, "Timepoint"])
     return(meta_per_time) 
 }
 
 
 lfc_per_time = function(data, meta, contrasts){
-    meta$Time = as.factor(meta$Time)
+    meta$Timepoint = as.factor(meta$Timepoint)
 
     sample_coefficients = lapply(meta$Group, function(x) return(contrasts[x, ]))
     sample_coefficients = as.matrix(unlist(sample_coefficients))
@@ -106,9 +107,9 @@ lfc_per_time = function(data, meta, contrasts){
     for(column in colnames(sample_coefficients)){
 	sample_coefficient = as.vector(unlist(sample_coefficients[, column]))
 	coef_data = t(t(data) * sample_coefficient)
-	for(time in meta$Time){
-	    mask = meta$Time == time
-	    colname = paste0(column, ".", as.character(time))
+	for(timepoint in meta$Timepoint){
+	    mask = meta$Timepoint == timepoint
+	    colname = paste0(column, ".", as.character(timepoint))
 	    log_fold_changes[, colname] = rowMeans(coef_data[, mask])
 	}
     }

### 2019-03-17 bin/moanin/R/visualization.R

@@ -44,7 +44,7 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL){
     meta = splines_model$meta
     groups = levels(meta$Group)
 
-    xrange = range(meta$Time)
+    xrange = range(meta$Timepoint)
     yrange = range(centroid)
     
     graphics::plot(xrange, yrange, type="n")
@@ -58,7 +58,7 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL){
         color = colors[i]
         
         mask = meta$Group == group
-        time = meta$Time[mask]
+        time = meta$Timepoint[mask]
         indx = order(time)
         graphics::lines(time[indx], centroid[mask][indx], type="b",
 		        col=color, pch=16,
@@ -78,6 +78,6 @@ plot_gene_splines = function(data, meta, gene_name, colors=NULL){
     # All of this should be extracted from the amazing model object that we
     # don't have implemented yet.
     degrees_of_freedom = 6
-    model = ~Group:splines::ns(Time, degrees_of_freedom) + Group + 0
+    model = ~Group:splines::ns(Timepoint, degrees_of_freedom) + Group + 0
     
 }

### 2019-03-17 bin/moanin/tests/testthat/test-validation.R

@@ -7,12 +7,12 @@ test_that("validation:check_meta", {
     meta = shoemaker2015$meta
     expect_silent(check_meta(meta))
 
-    meta_without_group = subset(meta, select=c("Replicate", "Time"))
+    meta_without_group = subset(meta, select=c("Replicate", "Timepoint"))
     meta_without_time = subset(meta, select=c("Group", "Replicate"))
     expect_error(check_meta(meta_without_group))
     expect_error(check_meta(meta_without_time))
 
-    meta_without_replicate = subset(meta, select=c("Time", "Group"))
+    meta_without_replicate = subset(meta, select=c("Timepoint", "Group"))
     expect_silent(check_meta(meta_without_replicate))
     expect_error(check_meta(meta_without_replicate, check_replicates=TRUE))
 })

### 2019-03-22 scripts/03_clustering.Rmd
ENH Adding method to select number of clusters for the clustering

@@
     - *Within cluster similarity as a function of the number of cluster* XXX
-      GAP statistics
+      Elbow method
+    - *Silhouette method* 
     - *Stability analysis*: this method assumes that the results of the
@@
       for other number of clusters.
+
+```{r}
+possible_n_clusters = 5:20
+
+```
+
+### Elbow method
+
+```{r elbow_method}
+
+# function to compute total within-cluster sum of square 
+within_cluster_sum_of_squares = function(n_cluster){
+    return(moanin:::splines_kmeans(y, splines_model,
+				   n_clusters=n_cluster, random_seed=42,
+				   n_init=10)$tot.withinss)
+}
+
+i = 0
+wss_values = list()
+for(n_cluster in possible_n_clusters){
+    wss_values[i] = within_cluster_sum_of_squares(n_cluster)
+    i = i + 1
+}
+
+plot(possible_n_clusters, wss_values,
+     type="b", pch=19, frame=FALSE, 
+     xlab="Number of clusters K",
+     ylab="Total within-clusters sum of squares")
+
+```
+
+### Average silhouette method
+
+```{r average_silhouette_method}
+# function to compute average silhouette for k clusters
+average_silhoutte = function(n_cluster) {
+    skmeans_results = moanin:::splines_kmeans(
+	y, splines_model, n_clusters=n_cluster,
+	random_seed=42, n_init=10)
+    silhouette_results = cluster::silhouette(skmeans_results$cluster, dist(y))
+    return(mean(silhouette_results[, 3]))
+}
+
+# extract the average silhoutte
+average_silhoutte_values = list()
+i = 1
+for(n_cluster in possible_n_clusters){
+    average_silhoutte_values[i] = average_silhoutte(n_cluster)
+    i = i + 1
+}
+
+plot(possible_n_clusters, average_silhoutte_values,
+     type="b", pch=19, frame=FALSE,
+     xlab="Number of clusters K",
+     ylab="Average Silhouettes")
+```
 

### 2019-03-22 bin/moanin/R/splines.R

@@ -43,6 +43,9 @@ fit_splines = function(data, splines_model, weights=NULL){
 #' @export
 fit_predict_splines = function(data, splines_model, weights=NULL){
     basis = splines_model$basis
+    if(!is.null(weights)){
+	stop("moanin::fit_predict_splines: not implemented")
+    }
     y_fitted = t(stats::lm.fit(basis, t(data))$fitted.values)
     return(y_fitted)
 }

### 2019-03-22 scripts/03_clustering.Rmd
FIX not the same API as normal Kmeans

@@
 possible_n_clusters = 5:20
-
 ```
@@
 				   n_clusters=n_cluster, random_seed=42,
-				   n_init=10)$tot.withinss)
+				   n_init=10)$WCSS_per_cluster)
 }

### 2019-03-22 scripts/03_clustering.Rmd
More fixes to model selection

--- 2019-03-17 scripts/03_clustering.Rmd
@@
     - *Within cluster similarity as a function of the number of cluster* XXX
-      GAP statistics
+      Elbow method
+    - *Silhouette method*
+
     - *Stability analysis*: this method assumes that the results of the
@@
       for other number of clusters.
+
+```{r running_clustering_on_all_k}
+all_possible_n_clusters = 5:20
+all_clustering = list()
+wss_values = list()
+
+i = 1
+for(n_cluster in all_possible_n_clusters){
+    clustering_results = moanin:::splines_kmeans(
+	y, splines_model,
+	n_clusters=n_cluster, random_seed=42,
+	n_init=10)
+    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
+    all_clustering[i] = clustering_results
+    i = i + 1
+}
+```
+
+### Elbow method
+
+```{r elbow_method}
+
+# function to compute total within-cluster sum of square
+
+plot(all_possible_n_clusters, wss_values,
+     type="b", pch=19, frame=FALSE, 
+     xlab="Number of clusters K",
+     ylab="Total within-clusters sum of squares")
+
+```
+
+### Average silhouette method
+
+```{r average_silhouette_method}
+# function to compute average silhouette for k clusters
+average_silhoutte = function(labels, y) {
+    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
+    return(mean(silhouette_results[, 3]))
+}
+
+# extract the average silhoutte
+average_silhoutte_values = list()
+i = 1
+for(i in 1:length(all_clustering)){
+    labels = all_clustering[i]
+    average_silhoutte_values[i] = average_silhoutte(labels, y)
+    i = i + 1
+}
+
+plot(all_possible_n_clusters, average_silhoutte_values,
+     type="b", pch=19, frame=FALSE,
+     xlab="Number of clusters K",
+     ylab="Average Silhouettes")
+```
 

### 2019-04-15 doc/manuscript/paper.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages
add some to intro

@@
 \usepackage{longtable}
-
+\usepackage{color}
 \newcommand{\todo}[1]{\textbf{[TODO\@: #1]}}
@@
 \newcommand{\fixme}[1]{\textbf{[FIXME\@: #1]}}
+\newcommand{\ep}[1]{\textcolor{cyan}{[EP\@: #1]}}
+\newcommand{\xxx}[1]{\textcolor{red}{XXX}}
 
@@
 differential expression analysis along lineages
+
+\ep{I assume this is, at this point, two different possible abstracts?}
 \end{abstract}
 
+\section{Introduction}
+
+Gene expression studies provide simultaneous quantification of the level of mRNA from all genes in a sample.
+High-throughput studies of gene expression have a long history, starting with microarray technologies in the 1990s through to single-cell technologies.
+While many expression studies are designed to compare the gene expression in distinct groups, there is also a long history of time-course expression studies, where the the gene expression is compared across time by measuring mRNA levels from different samples across time.\footnote{Because the collection of the mRNA is often destructive, samples at different time points are generally from different biological samples; longitudinal studies, for example tracking the same subject over time, are certainly possible in certain settings, but not directly considered here.} Such time course studies can vary from measuring a few distinct time points, to sampling ten to twenty time points.
+Many longer time series are particularly interested in investigating development over time.
+More recently, single-cell studies track single cells through their development, and a single cell is measured at a particular moment in its developmental progression -- a value that is not know but estimated from the data as its ``pseudo-time''.
+
+While there are many methods that have been proposed for discrete aspects of time course data, the entire workflow for analysis of such data remains difficult, particularly for long, developmental time series.
+Most methods proposed for time course data are concerned with detecting genes that are changing over time (differential expression analysis), examples being \texttt{edge}\todo[Other methods to mention: impulse models, what else?
+Citations!].
+However, with long time course datasets, particularly in developmental systems, a massive number of genes will show \emph{some} change.
+For example, in the \xxx system, over \xxx\% of genes are shown to be changing over time.
+The task in these settings is often not to detect changes in genes, but to categorize into biologically interpretable patterns the vast number of changes discovered.
+
+We present here a workflow for such an analysis that consists of \xxx main parts (Figure \ref{fig:workflow}):
+\begin{enumerate}
+	\item Identification of genes that are differentially expressed
+	\item Clustering of genes into distinct temporal patterns
+	\item Biological interpretation of the clusters
+\end{enumerate}
+\ep{This is probably not best enumeration of steps, but need something...}
+
+\begin{figure}
+\caption{Workflow for analysing time-course RNASeq experiments}
+\label{fig:workflow}
+\end{figure}
+
+This workflow represents an integration of both novel implementations of previously established methods and new methodologies for the settings of developmental time series.
+We provide the various steps of the workflow as functions in a R package, \citettt{\xxx}, along with command-line tools for running the entire workflow in a single command.
+
+\section{Results}
+\subsection{Quality control analysis}
+
+\begin{table}
+\caption{Summary of time-course RNASeq experiments and results of DE Analysis}
+\end{table}
+
+\subsection{Normalization}
+
+\subsection{Differential expression analysis}
+
+Several techniques for mRNA-expression analysis were
+
+\subsection{Clustering}
+
+\subsection{Downstream analysis}
+
+\section{Conclusion}
+
 \end{document}

### 2019-04-23 scripts/03_clustering.Rmd
Fixing silhouette scores + speed ups

--- 2019-03-22 scripts/03_clustering.Rmd
@@
 
-```{r}
-possible_n_clusters = 5:20
+```{r running_clustering_on_all_k}
+all_possible_n_clusters = c(5, 7, 9, 11)
+all_clustering = list()
+wss_values = list()
 
+i = 1
+for(n_cluster in all_possible_n_clusters){
+    clustering_results = moanin:::splines_kmeans(
+	y, splines_model,
+	n_clusters=n_cluster, random_seed=42,
+	n_init=10)
+    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
+    all_clustering[[i]] = clustering_results$clusters
+    i = i + 1
+}
 ```
@@
 
-```{r elbow_method}
+```{r elbow_method, fig.width=6, fig.height=4}
 
 # function to compute total within-cluster sum of square
-within_cluster_sum_of_squares = function(n_cluster){
-    return(moanin:::splines_kmeans(y, splines_model,
-				   n_clusters=n_cluster, random_seed=42,
-				   n_init=10)$tot.withinss)
-}
 
-i = 0
-wss_values = list()
-for(n_cluster in possible_n_clusters){
-    wss_values[i] = within_cluster_sum_of_squares(n_cluster)
-    i = i + 1
-}
-
-plot(possible_n_clusters, wss_values,
+plot(all_possible_n_clusters, wss_values,
      type="b", pch=19, frame=FALSE, 
@@
 
-```{r average_silhouette_method}
+```{r average_silhouette_method, fig.width=6, fig.height=4}
 # function to compute average silhouette for k clusters
-average_silhoutte = function(n_cluster) {
-    skmeans_results = moanin:::splines_kmeans(
-	y, splines_model, n_clusters=n_cluster,
-	random_seed=42, n_init=10)
-    silhouette_results = cluster::silhouette(skmeans_results$cluster, dist(y))
+average_silhouette = function(labels, y) {
+    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
     return(mean(silhouette_results[, 3]))
@@
 
-# extract the average silhoutte
-average_silhoutte_values = list()
+# extract the average silhouette
+average_silhouette_values = list()
 i = 1
-for(n_cluster in possible_n_clusters){
-    average_silhoutte_values[i] = average_silhoutte(n_cluster)
+for(i in 1:length(all_clustering)){
+    labels = all_clustering[i]
+    average_silhouette_values[i] = average_silhouette(labels, y)
     i = i + 1
@@
 
-plot(possible_n_clusters, average_silhoutte_values,
+plot(all_possible_n_clusters, average_silhouette_values,
      type="b", pch=19, frame=FALSE,

### 2019-04-23 scripts/04_downstream_analysis.Rmd
Fixing silhouette scores + speed ups

--- 2019-03-06 scripts/04_downstream_analysis.Rmd
@@
 
-```{r enrichment_analysis, results="hide", message=FALSE}
+```{r enrichment_analysis, results="hide", message=FALSE, eval=FALSE}
 clusters = sort(as.vector(unlist(unique(labels))))
-clusters = c(1, 2, 3)
 all_pathways = data.frame()
@@
 }
+```
+
+```{r echo=FALSE}
+all_pathways = read.table("results/all_pathways.tsv",
+			  sep="\t")
 ```
@@
 
-```{r go_term_enrichment, results="hide", message=FALSE}
+```{r go_term_enrichment, results="hide", message=FALSE, eval=FALSE}
 
-cluster = c(1, 2, 3)
-ontologies = c("BP", "NF", "CC")
-# It's taking too long to compile with everything. Just limit ourselves for
-# now to BP
 ontologies = c("BP")
@@
 ```{r echo=FALSE}
+all_go_terms = read.table(
+    "results/all_go_terms.tsv",
+    sep="\t")
+```
+
+```{r echo=FALSE}
 

### 2019-04-23 scripts/run_go_terms.R

@@ -1,0 +1,76 @@
+library(devtools)
+library(limma)
+library(splines)
+library(stats)
+library(KEGGprofile)
+library(kableExtra)
+library(viridis)
+
+library(moanin)
+library(biomaRt)
+df = 6
+
+# Preprocessing & filtering
+is_count = FALSE
+take_log = FALSE
+filter_expression = FALSE
+
+# Differential expression analysis
+timecourse_contrasts = c("C-K", "C-M")
+
+# Clustering
+filter_genes = TRUE
+n_genes_to_keep = 5000
+n_clusters = 20
+percentage_genes_to_label = 0.5
+# Set data files and options
+data("shoemaker2015")
+data = shoemaker2015$data
+meta = shoemaker2015$meta
+
+labels = read.table(".results/clustering_labels.txt", sep="\t",
+                    check.names=FALSE)
+
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = labels[!is.na(labels)]
+
+
+# FIXME move this to the package
+source("GO_terms_analysis.R")
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
+              values=gene_names,
+              filters="refseq_mrna",
+              mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = create_go_term_mapping(genes)
+
+clusters = sort(as.vector(unlist(unique(labels))))
+ontologies = c("BP")
+all_go_terms = data.frame()
+for(cluster in clusters){
+    for(ontology in ontologies){
+        scores = as.numeric(labels == cluster)
+        names(scores) = gene_names
+        go_terms_enriched = find_enriched_go_terms(
+            scores,
+            gene_id_go_mapping, ontology=ontology)
+
+        if((dim(all_go_terms)[1] == 0) & (dim(go_terms_enriched)[1] != 0)){
+            go_terms_enriched["Cluster"] = cluster
+            go_terms_enriched["Ontology"] = ontology
+            all_go_terms = go_terms_enriched
+        }else if(dim(go_terms_enriched)[1] != 0){
+            go_terms_enriched["Cluster"] = cluster
+            go_terms_enriched["Ontology"] = ontology
+            all_go_terms = merge(all_go_terms, go_terms_enriched, all=TRUE)
+        }
[…]

### 2019-04-23 scripts/run_pathways.R

@@ -1,0 +1,79 @@
+library(devtools)
+library(limma)
+library(splines)
+library(stats)
+library(KEGGprofile)
+library(kableExtra)
+library(viridis)
+
+library(moanin)
+library(biomaRt)
+df = 6
+
+# Preprocessing & filtering
+is_count = FALSE
+take_log = FALSE
+filter_expression = FALSE
+
+# Differential expression analysis
+timecourse_contrasts = c("C-K", "C-M")
+
+# Clustering
+filter_genes = TRUE
+n_genes_to_keep = 5000
+n_clusters = 20
+percentage_genes_to_label = 0.5
+# Set data files and options
+data("shoemaker2015")
+data = shoemaker2015$data
+meta = shoemaker2015$meta
+
+labels = read.table(".results/clustering_labels.txt", sep="\t",
+                    check.names=FALSE)
+
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = labels[!is.na(labels)]
+
+
+# FIXME move this to the package
+source("GO_terms_analysis.R")
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
+              values=gene_names,
+              filters="refseq_mrna",
+              mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = create_go_term_mapping(genes)
+
+clusters = sort(as.vector(unlist(unique(labels))))
+all_pathways = data.frame()
+for(cluster in clusters){
+    genes = gene_names[labels == cluster]
+
+    # convert gene names
+    genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		  filters="refseq_mrna", values=genes,
+		  mart=ensembl)["entrezgene"]
+    genes = as.vector(unlist(genes))
+    pathways = KEGGprofile::find_enriched_pathway(
+	genes, species="mmu",
+	download_latest=TRUE)
+
+    if((dim(all_pathways)[1] == 0) & (dim(pathways$stastic)[1] != 0)){
+	print("Initialize")
+	pathways$stastic["Cluster"] = cluster
+	all_pathways = pathways$stastic
+    }else if(dim(pathways$stastic)[1] != 0) {
+	print("Merge")
+	pathways$stastic["Cluster"] = cluster
[…]

### 2019-04-23 scripts/03_clustering.Rmd
Reducing the number of genes for faster clustering

@@
 keep_all_significant = FALSE
-n_genes_to_keep = 10000
+n_genes_to_keep = 5000
 n_clusters = 20
@@
 moanin:::plot_centroids(centroids, splines_model,
-			colors=as.vector(unlist(ann_colors["Group"])))
+			colors=as.vector(unlist(ann_colors["Group"])),
+			smooth=FALSE)
 ```

### 2019-04-23 bin/moanin/R/splines.R

@@ -36,20 +36,66 @@ fit_splines = function(data, splines_model, weights=NULL){
 #'
 #' @param data the data
 #' @param splines_model splines_model
-#' @param weights weigts
+#' @param weights weights
 #'
 #' @return y_fitted the fitted y values
 #'
 #' @export
-fit_predict_splines = function(data, splines_model, weights=NULL){
+fit_predict_splines = function(data, splines_model, weights=NULL, meta_prediction=NULL){
     basis = splines_model$basis
+    meta = splines_model$meta
     if(!is.null(weights)){
 	stop("moanin::fit_predict_splines: not implemented")
     }
-    y_fitted = t(stats::lm.fit(basis, t(data))$fitted.values)
+    if(is.null(meta_prediction)){
+        y_fitted = t(stats::lm.fit(basis, t(data))$fitted.values)
+    }else{
+	degrees_of_freedom = splines_model$degrees_of_freedom
+	fitting_data = t(as.matrix(data))
+	formula_data = list(
+	    "Group"=meta$Group,
+	    "Timepoint"=meta$Timepoint,
+	    "fitting_data"=fitting_data,
+	    "degrees_of_freedom"=splines_model$degrees_of_freedom)
+
+	updated_formula = update(splines_model$formula, fitting_data ~ .)
+	model = lm(updated_formula, formula_data)	
+	y_fitted = predict(model, meta_prediction)
+    }
     return(y_fitted)
 }
 
+
+#' Create prediction meta data from splines model
+#'
+#' @param splines_model a Splines model object
+#' @param num_timepoints integer, optional, default: 100
+#'	Number of timepoints to use for the prediction metadata
+#'
+#' @export
+create_meta_prediction = function(splines_model, num_timepoints=100){
+    # Create splines_model for prediction
+    timepoints_pred = NULL
+    groups_pred = NULL
+    meta = splines_model$meta
+    groups = levels(meta$Group) 
+    for(group in groups){
+	mask = meta$Group == group
+	time = meta$Timepoint[mask]
+
+	timepoints_pred = c(
+	    timepoints_pred,
+	    seq(min(time), max(time), length=100))
+	groups_pred = c(
+	    groups_pred,
+	    rep(group, 100))
+    }
+    meta_prediction = data.frame(
+	"Timepoint"=timepoints_pred,
+	"Replicates"=rep(1, length(timepoints_pred)),
+	"Group"=groups_pred)
+    return(meta_prediction)
+}
[…]

### 2019-04-23 bin/moanin/R/visualization.R

@@ -8,9 +8,10 @@ library(graphics)
 #' @param meta	data.frame (t, n) containing the metadata.
 #' @param colors vector, optional, default NULL
 #'		vector of colors
-#'
+#' @param smooth boolean, optional, default: FALSE
+#'  Whether to smooth the centroids or not.
 #' @export
-plot_centroids = function(centroids, splines_model, colors=NULL){
+plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
     n_centroids = dim(centroids)[1]
     if(n_centroids <= 3){
         graphics::par(
@@ -35,37 +36,60 @@ plot_centroids = function(centroids, splines_model, colors=NULL){
     }
 
     for(i in 1:n_centroids){
-        plot_centroid_individual(centroids[i, ], splines_model, colors=colors)
+        plot_centroid_individual(centroids[i, ], splines_model, colors=colors,
+				 smooth=smooth)
     }
 }
 
 
-plot_centroid_individual = function(centroid, splines_model, colors=NULL){
+plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth=FALSE){
     meta = splines_model$meta
     groups = levels(meta$Group)
 
     xrange = range(meta$Timepoint)
     yrange = range(centroid)
+
+    centroid = t(as.matrix(centroid))
     
     graphics::plot(xrange, yrange, type="n")
     if(is.null(colors)){
         colors = viridis::viridis(length(groups))
     }
-    
+    if(smooth){
+	# FIXME this is supposed to be on the fitted lines, but I'm not able
+	# to get this to work fine in R.
+	meta_prediction = create_meta_prediction(splines_model)
+	centroid_fitted = fit_predict_splines(
+	    as.matrix(centroid), splines_model,
+	    meta_prediction=meta_prediction)
+    }else{
+	centroid_fitted = fit_predict_splines(
+	    as.matrix(centroid), splines_model)
+    }
+
+
     # scatter points for values
     for(i in 1:length(groups)){
         group = groups[i]
         color = colors[i]
         
+	# Start by individual points
         mask = meta$Group == group
         time = meta$Timepoint[mask]
         indx = order(time)
         graphics::lines(time[indx], centroid[mask][indx], type="b",
 		        col=color, pch=16,
 			lwd=0)
+	if(smooth){
+	    mask = meta_prediction$Group == group
[…]

### 2019-04-23 scripts/00_introduction.Rmd
Merging Elizabeth's introduction to the Rmd manuscript

@@
 
+Gene expression studies provide simultaneous quantification of the level of
+mRNA from all genes in a sample. High-throughput studies of gene expression
+have a long history, starting with microarray technologies in the 1990s
+through to single-cell technologies. While many expression studies are
+designed to compare the gene expression in distinct groups, there is also a
+long history of time-course expression studies, where the the gene expression
+is compared across time by measuring mRNA levels from different samples across
+time.\footnote{Because the collection of the mRNA is often destructive,
+samples at different time points are generally from different biological
+samples; longitudinal studies, for example tracking the same subject over
+time, are certainly possible in certain settings, but not directly considered
+here.} Such time course studies can vary from measuring a few distinct time
+points, to sampling ten to twenty time points. Many longer time series are
+particularly interested in investigating development over time. More recently,
+single-cell studies track single cells through their development, and a single
+cell is measured at a particular moment in its developmental progression -- a
+value that is not know but estimated from the data as its "pseudo-time."
+
+While there are many methods that have been proposed for discrete aspects of
+time course data, the entire workflow for analysis of such data remains
+difficult, particularly for long, developmental time series. Most methods
+proposed for time course data are concerned with detecting genes that are
+changing over time (differential expression analysis), examples being
+\texttt{edge}\todo[Other methods to mention: impulse models, what else?
+Citations!]. However, with long time course datasets, particularly in
+developmental systems, a massive number of genes will show *some* change.
+For example, in the \xxx system, over \xxx\% of genes are shown to be changing
+over time. The task in these settings is often not to detect changes in genes,
+but to categorize into biologically interpretable patterns the vast number of
+changes discovered.
+
+We present here a workflow for such an analysis that consists of \xxx main
+parts (Figure \ref{fig:workflow}):
+
+- Identification of genes that are differentially expressed
+- Clustering of genes into distinct temporal patterns
+- Biological interpretation of the clusters
+
+**FIXME EP This is probably not best enumeration of steps, but need
+something...**
+
+This workflow represents an integration of both novel implementations of
+previously established methods and new methodologies for the settings of
+developmental time series. We provide the various steps of the workflow as
+functions in a R package, \citettt{\xxx}, along with command-line tools for
+running the entire workflow in a single command.
+
+---
+# Old intro
+
 Time-course gene expression analysis workflows include steps used more broadly
@@
 holding all necessary metadata to perform all steps of the pipeline.
+---
 

### 2019-04-23 scripts/00_introduction.Rmd
FIX footnote

@@
 is compared across time by measuring mRNA levels from different samples across
-time.\footnote{Because the collection of the mRNA is often destructive,
-samples at different time points are generally from different biological
-samples; longitudinal studies, for example tracking the same subject over
-time, are certainly possible in certain settings, but not directly considered
-here.} Such time course studies can vary from measuring a few distinct time
+time [^1]. Such time course studies can vary from measuring a few distinct time
 points, to sampling ten to twenty time points. Many longer time series are
@@
 value that is not know but estimated from the data as its "pseudo-time."
+
+[^1]: Because the collection of the mRNA is often destructive, samples at
+different time points are generally from different biological samples;
+longitudinal studies, for example tracking the same subject over time, are
+certainly possible in certain settings, but not directly considered here.
 

### 2019-04-23 bin/moanin/R/visualization.R
Smooth plotting is finally working

@@ -36,7 +36,8 @@ plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
     }
 
     for(i in 1:n_centroids){
-        plot_centroid_individual(centroids[i, ], splines_model, colors=colors,
+        plot_centroid_individual(as.vector(centroids[i, ]),
+				 splines_model, colors=colors,
 				 smooth=smooth)
     }
 }
@@ -48,9 +49,8 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth
 
     xrange = range(meta$Timepoint)
     yrange = range(centroid)
-
     centroid = t(as.matrix(centroid))
-    
+
     graphics::plot(xrange, yrange, type="n")
     if(is.null(colors)){
         colors = viridis::viridis(length(groups))
@@ -60,11 +60,11 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth
 	# to get this to work fine in R.
 	meta_prediction = create_meta_prediction(splines_model)
 	centroid_fitted = fit_predict_splines(
-	    as.matrix(centroid), splines_model,
+	    centroid, splines_model,
 	    meta_prediction=meta_prediction)
     }else{
 	centroid_fitted = fit_predict_splines(
-	    as.matrix(centroid), splines_model)
+	    centroid, splines_model)
     }
 
 
@@ -85,7 +85,7 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth
 	    time = meta_prediction$Timepoint[mask]
 	    indx = order(time)
 	}
-	graphics::lines(time[indx], centroid_fitted[mask][indx], type="n",
+	graphics::lines(time[indx], centroid_fitted[mask][indx], type="o",
 			col=color, lwd=1)
 
     }
@@ -99,9 +99,5 @@ plot_gene_splines = function(data, meta, gene_name, colors=NULL){
     }
     gene_data = data[gene_name, ]
 
-    # All of this should be extracted from the amazing model object that we
-    # don't have implemented yet.
-    degrees_of_freedom = 6
-    model = ~Group:splines::ns(Timepoint, degrees_of_freedom) + Group + 0
 
 }

### 2019-04-24 scripts/03_clustering.Rmd
Smooth plotting is finally working
FIX Colors now correspond to the right conditions
ENH Now performing prediction step in stability analysis

@@
 moanin:::plot_centroids(centroids, splines_model,
-			colors=as.vector(unlist(ann_colors["Group"])),
-			smooth=FALSE)
+			colors=ann_colors$Group,
+			smooth=TRUE)
 ```
@@
 
+### Looking at the stability of the clustering
+
+```{r fig.width=6, fig.height=4}
+library(pheatmap)
+stability_5 = read.table("results/stability_5.tsv", sep="\t")
+consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5)
+pheatmap::pheatmap(consensus_matrix_stability_5[1:1000, 1:1000], symm=TRUE,
+		   treeheight_row=0, treeheight_col=0)
+
+stability_20 = read.table("results/stability_20.tsv", sep="\t")
+consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20)
+pheatmap::pheatmap(consensus_matrix_stability_20[1:1000, 1:1000], symm=TRUE,
+		   treeheight_row=0, treeheight_col=0)
+
+```
+
 ```{r results="hide", echo=FALSE}

### 2019-04-24 bin/moanin/R/cluster.R

@@ -47,6 +47,44 @@ splines_kmeans = function(data, splines_model, n_clusters=10,
     kmeans_clusters = ClusterR::KMeans_rcpp(
 	fitted_data, n_clusters, num_init=n_init, max_iters=max_iter,
 	seed=random_seed, initializer=init)
+    kmeans_clusters$centroids = rescale_values(
+	kmeans_clusters$centroids, splines_model)
+    names(kmeans_clusters$clusters) = row.names(data)
 
+    kmeans_clusters$splines_model = splines_model
+    kmeans_clusters$fit_splines = fit_splines
+    kmeans_clusters$rescale = rescale
+    return(kmeans_clusters)
+}
+
+
+splines_kmeans_prediction = function(data, kmeans_clusters){
+    splines_model = kmeans_clusters$splines_model
+    fit_splines = kmeans_clusters$fit_splines
+    rescale = kmeans_clusters$rescale
+
+    meta = splines_model$meta
+    basis = splines_model$basis
+    check_data_meta(data, meta)
+
+    if(fit_splines){
+        fitted_data = fit_predict_splines(data, splines_model)
+    }else{
+	fitted_data = data
+    }
+
+    if(rescale){
+        fitted_data = rescale_values(fitted_data, meta)
+    }
+
+    closest.cluster <- function(x) {
+	cluster.dist <- apply(
+	    kmeans_clusters$centroids, 1, function(y){sqrt(sum((x-y)^2))})
+	return(which.min(cluster.dist)[1])
+    }
+
+    all_labels <- apply(data, 1, closest.cluster) 
+    kmeans_clusters$clusters = all_labels
+    names(kmeans_clusters$clusters) = row.names(data)
     return(kmeans_clusters)
 }

### 2019-04-24 scripts/concatenate_clustering_stability_results.R

@@ -1,0 +1,23 @@
+
+
+
+folders = list.files("results/")
+folders = folders[grep("stability", folders)]
+
+for(folder in folders){
+    filenames = list.files(paste0("results/", folder))
+    all_labels = NULL
+    for(filename in filenames){
+	labels = unique(
+	    read.table(paste0("results/", folder, "/", filename),
+		       row.names=NULL))
+	colnames(labels) = c("Gene", "Label")
+	row.names(labels) = labels$Gene
+	labels = subset(labels, select=c("Label"))
+	all_labels = merge(all_labels, labels, by="row.names", all=TRUE)
+	row.names(all_labels) = all_labels$Row.names
+	all_labels = subset(all_labels, select=-c(Row.names))
+    }
+    outname = paste0("results/", folder, ".tsv")
+    write.table(all_labels, outname, sep="\t")
+}

### 2019-04-24 scripts/run_clustering.R

@@ -7,9 +7,6 @@ args = commandArgs(trailingOnly=TRUE)
 n_clusters = strtoi(args[1])
 random_seed = strtoi(args[2])
 
-n_clusters = 20
-random_seed = 5
-
 # Other options
 n_genes_to_keep = 5000
 filter_genes = TRUE
@@ -82,9 +79,12 @@ kmeans_clusters = moanin:::splines_kmeans(
     random_seed=random_seed,
     n_init=20)
 
+# Perform prediction on the whole set of data.
+kmeans_clusters = moanin:::splines_kmeans_prediction(
+    data, kmeans_clusters)
+
 # Let's save only the labels here.
 labels = kmeans_clusters$clusters
-names(labels) = row.names(random_subsample_y)
 if(!dir.exists(dirname(outname))){
     dir.create(dirname(outname))
 }

### 2019-04-24 scripts/03_clustering.Rmd
Stability measures for clustering & all

--- 2019-04-23 scripts/03_clustering.Rmd
@@
 # Rescale the centroids
-centroids = moanin:::rescale_values(kmeans_clusters$centroids, splines_model)
+centroids = moanin:::rescale_values(kmeans_clusters$centroids)
 
@@
 moanin:::plot_centroids(centroids, splines_model,
-			colors=as.vector(unlist(ann_colors["Group"])),
-			smooth=FALSE)
+			colors=ann_colors$Group,
+			smooth=TRUE)
+```
+
+Now, let's look at genes in cluster 15
+
+```{r fig.width=8, fig.height=8}
+gene_names = row.names(data)
+genes_to_plot =  names(kmeans_clusters$clusters[kmeans_clusters$clusters == 15])
+data_to_plot = data[genes_to_plot, ]
+submeta = meta
+submeta$Group = factor(submeta$Group, levels(submeta$Group)[c(3, 1, 2, 5, 4)])
+ord = order(
+  submeta$Group,
+  submeta$Timepoint,
+  submeta$Replicate)
+submeta$Timepoint = as.factor(submeta$Timepoint)
+
+data_to_plot = data_to_plot[, ord]
+submeta = submeta[ord, ]
+
+NMF::aheatmap(
+    data_to_plot,
+    Colv=NA,
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=FALSE,
+    annColors=ann_colors,
+    main="Cluster 15",
+    treeheight=0)
+
+NMF::aheatmap(
+    moanin:::rescale_values(data_to_plot),
+    Colv=NA,
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=FALSE,
+    annColors=ann_colors,
+    main="Cluster 15",
+    treeheight=0)
+
 ```
@@
 
+### Looking at the stability of the clustering
+
+```{r fig.width=6, fig.height=4}
+library(NMF)
+stability_5 = read.table("results/stability_5.tsv", sep="\t")
+consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
+							scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+stability_20 = read.table("results/stability_15.tsv", sep="\t")
+consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
+							 scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+```
+
 ```{r results="hide", echo=FALSE}

### 2019-04-24 scripts/03_clustering.Rmd
ENH adding model explorer

@@
 output: rmarkdown::html_vignette
+bibliography: bibliography.bib
 ---
 
-```{r echo=FALSE, results="hide"}
+```{r echo=FALSE, results="hide", message=FALSE}
 # Loading dependencies
@@
 library(moanin)
+library(NMF)
 source("utils.R")
@@
 
-    - *Within cluster similarity as a function of the number of cluster* XXX
-      Elbow method
+    - Elbow method
     - *Silhouette method* 
-    - *Stability analysis*: this method assumes that the results of the
-      clustering will be more stable for the "true" number of clusters than
-      for other number of clusters.
+    - *Model explorer*
 
@@
 
+The last strategy is to look at the stability of the clustering.
+
 ```{r fig.width=6, fig.height=4}
@@
 
+```{r}
+n_clusters = 5:20
+all_labels = list()
+for(i in n_clusters){
+    filename = paste0("results/stability_", i, ".tsv")
+    stability = read.table(filename, sep="\t")
+    all_labels[[paste0("C", i)]] = stability
+}
+```
+
+```{r fig.width=8, fig.height=8}
+library(moanin)
+moanin::plot_model_explorer(all_labels)
+```
+
 ```{r results="hide", echo=FALSE}

### 2019-04-24 bin/moanin/R/cluster_validation.R

@@ -1,4 +1,5 @@
 library(reshape2)
+library(NMI)
 
 
 #' Compute consensus matrix from labels
@@ -24,3 +25,70 @@ consensus_matrix = function(labels, scale=TRUE){
     }
     return(consensus)
 }
+
+
+#' Plot model explorer
+#'
+#' @param labels list of labels
+#' @export
+plot_model_explorer = function(labels){
+    all_labels = labels
+    n_clusters = names(all_labels)
+    nmi = function(x, y){
+	x_dataframe = as.data.frame(x)
+	colnames(x_dataframe) = c("Label")
+	x_dataframe$Gene = row.names(x_dataframe)
+	x_dataframe = x_dataframe[c("Gene", "Label")]
+
+	y_dataframe = as.data.frame(y)
+	colnames(y_dataframe) = c("Label")
+	y_dataframe$Gene = row.names(y_dataframe)
+	y_dataframe = y_dataframe[c("Gene", "Label")]
+
+	return(NMI::NMI(x_dataframe, y_dataframe))
+    }
+
+    nmi_scores = list()
+    colors = viridis::viridis(length(n_clusters))
+    max_trial = 0
+    min_score = 1
+    max_score = 0
+    for(i in 1:length(n_clusters)){
+	n_cluster = n_clusters[i]
+	color = colors[i]
+
+	labels = all_labels[[n_cluster]]
+	n_trials = dim(labels)[2]
+	scores = NULL
+	for(trial in 1:n_trials){
+	    if(trial == n_trials){
+		break
+	    }
+	    column = colnames(labels)[trial]
+	    columns_to_consider = colnames(labels)[(trial+1):n_trials]
+	    label = labels[column]
+	    scores = c(scores, as.vector(
+		unlist(apply(labels[columns_to_consider], 2, function(x){nmi(x, label)}))))
+	}
+	nmi_scores[[n_cluster]] = sort(scores)
+	max_trial = max(n_trials, max_trial)
+	min_score = min(min_score, min(scores))
+	max_score = max(max_score, max(scores))
+     }
+
+    xrange = c(min_score, max_score)
+    yrange = c(1, max_trial)
+
+    graphics::plot(xrange, yrange, type="n")
+
+     for(i in 1:length(n_clusters)){
+	color = colors[i]
+	scores = nmi_scores[[i]]
+	graphics::lines(sort(scores), 1:length(scores), type="b",
+			pch=16,
+			col=color,
+			lwd=0)
+
+    }
+}
+

### 2019-04-25 scripts/03_clustering.Rmd
MAINT Better scoring

@@
 
-# XXX is there a bug here?
-# all_scores = data.frame(row.names=row.names(data))
-# for(k in 1:n_clusters){
-#     scores = moanin:::score_genes_centroid(data, centroids[k,])
-#     all_scores[paste0("cluster_", k)] = scores
-# }
+all_scores = data.frame(row.names=row.names(data))
+for(k in 1:n_clusters){
+     scores = moanin:::score_genes_centroid(data, centroids[k,])
+     all_scores[paste0("cluster_", k)] = scores
+}
 
 # Only assign labels to X% of the genes
-# max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-# genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-# labels = moanin:::row_argmin(all_scores)
-# labels[genes_to_not_consider] = NA
+max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
+genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
+labels = moanin:::row_argmin(all_scores)
+labels[genes_to_not_consider] = NA
 ```
@@
 # Save the clustering for the next step
-kmeans_labels = as.data.frame(kmeans_clusters$clusters)
+kmeans_labels = as.data.frame(labels)
 row.names(kmeans_labels) = row.names(y)

### 2019-04-25 bin/moanin/R/cluster.R

@@ -77,13 +77,13 @@ splines_kmeans_prediction = function(data, kmeans_clusters){
         fitted_data = rescale_values(fitted_data, meta)
     }
 
-    closest.cluster <- function(x) {
-	cluster.dist <- apply(
+    closest_cluster <- function(x) {
+	cluster_dist <- apply(
 	    kmeans_clusters$centroids, 1, function(y){sqrt(sum((x-y)^2))})
-	return(which.min(cluster.dist)[1])
+	return(which.min(cluster_dist)[1])
     }
 
-    all_labels <- apply(fitted_data, 1, closest.cluster) 
+    all_labels <- apply(fitted_data, 1, closest_cluster) 
     kmeans_clusters$clusters = all_labels
     names(kmeans_clusters$clusters) = row.names(data)
     return(kmeans_clusters)

### 2019-04-25 bin/moanin/R/cluster_validation.R

@@ -13,7 +13,7 @@ consensus_matrix = function(labels, scale=TRUE){
     melted_labels = reshape2::melt(as.matrix(labels))
     colnames(melted_labels) = c("Gene", "Clustering", "Label")
     melted_labels$Clustering_lab = melted_labels$Clustering:as.factor(melted_labels$Label)
-    w = reshape2::dcast(melted_labels, Gene~Clustering_lab)
+    w = reshape2::dcast(melted_labels, Gene~Clustering_lab, value.var="Clustering_lab")
     x = as.matrix(w[,-1])
     x[is.na(x)] = 0
     x = apply(x, 2,  function(x) as.numeric(x > 0))

### 2019-04-25 bin/moanin/R/splines.R

@@ -166,9 +166,11 @@ score_genes_centroid = function(data, centroid){
     n_genes = dim(data)[1]
     data_fitted = align_data_onto_centroid(data, centroid)
 
-    scores = row_sum((data_fitted - rep(centroid, times=n_genes))**2)
+    scores = apply(data_fitted, 1, function(y){sqrt(sum((centroid - y)^2))})
 
-    scores = scores / max(scores)
+    all_zeros_gene = data_fitted[1, ] * 0
+    max_score = sqrt(sum((centroid - all_zeros_gene)^2))
+    scores = scores / max_score
     return(scores)
 }
 

### 2019-04-25 scripts/00_introduction.Rmd

@@
 output: rmarkdown::html_vignette
+bibliography: bibliography.bib
 ---
@@
 changing over time (differential expression analysis), examples being
-\texttt{edge}\todo[Other methods to mention: impulse models, what else?
-Citations!]. However, with long time course datasets, particularly in
+`edge` [@storey:significance], functional component analysis based models
+[@wu:more], time-course permutation tests [@park:statistical], XXX
+[@sun:multiple]
+**TODO: Other methods to mention: impulse models, what else?
+Citations!**. However, with long time course datasets, particularly in
 developmental systems, a massive number of genes will show *some* change.
@@
 
-We present here a workflow for such an analysis that consists of \xxx main
+We present here a workflow for such an analysis that consists of 4 main
 parts (Figure \ref{fig:workflow}):
 
-- Identification of genes that are differentially expressed
-- Clustering of genes into distinct temporal patterns
-- Biological interpretation of the clusters
-
-**FIXME EP This is probably not best enumeration of steps, but need
-something...**
+- Quality control and normalization;
+- Identification of genes that are differentially expressed;
+- Clustering of genes into distinct temporal patterns;
+- Biological interpretation of the clusters.
 
@@
 developmental time series. We provide the various steps of the workflow as
-functions in a R package, \citettt{\xxx}, along with command-line tools for
-running the entire workflow in a single command.
+functions in a R package **FIXME**.
 
@@
 collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
-18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later). The three strains
-of influenza used in the study are (1) a low pathogenicity seasonal H1N1
-influenza virus (A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from
-the 2009 pandemic season (A/California/04/2009 [H1N1]), and a highly
-pathogenic H5N1 avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were
-injected with 10^5 PFU of each virus. An adidtional 42 mice were injected with
-a lower dose of the Vietnam avian influenza virus (10^3 PFU).
+18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
+[@shoemaker:ultrasensitive]. The three strains of influenza used in the
+study are (1) a low pathogenicity seasonal H1N1 influenza virus
+(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
+pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
+avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
+10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
+of the Vietnam avian influenza virus (10^3 PFU).
 
@@
 **XXX Add summary of the key findings of the the paper**
+
+## References

### 2019-04-25 scripts/03_clustering.Rmd

@@
 labels[genes_to_not_consider] = NA
+names(labels) = row.names(data)
+```
+
+Summary of number of genes assigned with original scoring, and "better"
+scoring.
+
+```{r}
+table(kmeans_clusters$clusters)
+barplot(table(kmeans_clusters$clusters), col="black")
 ```
@@
 ```{r}
-table(kmeans_clusters$clusters)
+table(labels)
+barplot(table(labels), col="black")
 ```
+
+Just a hack, cause I didn't update the code
+```{r}
+kmeans_clusters$clusters = labels[!is.na(labels)]
+```
+
+```{r results="hide", echo=FALSE}
+# Save the clustering for the next step
+kmeans_labels = as.data.frame(labels[!is.na(labels)])
+write.table(kmeans_labels,
+	    ".results/clustering_labels.txt",
+	    sep="\t")
+```
+
+Now let's visualize the centroids.
 
@@
 ```
-
-```{r results="hide", echo=FALSE}
-# Save the clustering for the next step
-kmeans_labels = as.data.frame(labels)
-row.names(kmeans_labels) = row.names(y)
-write.table(kmeans_labels,
-	    ".results/clustering_labels.txt",
-	    sep="\t")
-```

### 2019-04-30 scripts/03_clustering.Rmd
FIX in scoring function

@@
 keep_all_significant = FALSE
-n_genes_to_keep = 5000
+n_genes_to_keep = 10000
 n_clusters = 20
@@
 
-# Rescale the centroids
-centroids = moanin:::rescale_values(kmeans_clusters$centroids)
+all_scores = matrix(NA, nrow=dim(data)[1], ncol=n_clusters)
 
-all_scores = data.frame(row.names=row.names(data))
 for(k in 1:n_clusters){
-     scores = moanin:::score_genes_centroid(data, centroids[k,])
-     all_scores[paste0("cluster_", k)] = scores
+     scores = moanin:::score_genes_centroid(
+	data,
+	kmeans_clusters$centroids[k,])
+     all_scores[, k] = scores
 }
@@
 # Only assign labels to X% of the genes
-max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
-labels = moanin:::row_argmin(all_scores)
+max_score = quantile(rowMin(all_scores), c(percentage_genes_to_label))
+genes_to_not_consider = rowMin(all_scores) >= max_score
+labels = apply(all_scores, 1, which.min)
 labels[genes_to_not_consider] = NA
 names(labels) = row.names(data)
+```
+
+```{r fig.height=8, fig.width=8}
+par(mfrow=c(4, 5), mar=c(0.1, 0.1, 0.1, 0.1))
+for(k in 1:n_clusters){
+    hist(all_scores[k, ], breaks=20, col="black")
+    abline(v=max_score, col="#AB0000")
+}
 ```
@@
 ```{r fig.height=8, fig.width=8}
-moanin:::plot_centroids(centroids, splines_model,
+moanin:::plot_centroids(kmeans_clusters$centroids, splines_model,
 			colors=ann_colors$Group,

### 2019-04-30 bin/moanin/R/splines.R

@@ -174,10 +174,11 @@ score_genes_centroid = function(data, centroid, positive_scaling=TRUE, scale=TRU
 
     if(scale){
         all_zeros_gene = data_fitted[1, ] * 0
-	max_score = sqrt(sum((centroid - all_zeros_gene)^2))
-	scores = scores / max_score
+	max_score = max(scores)
+    }else{
+	max_score = 1
     }
-    return(scores)
+    return(scores / max_score)
 }
 
 

### 2019-04-30 scripts/03_clustering.Rmd
FIXes and more

@@
 
+all_scores = as.matrix(all_scores)
 # Only assign labels to X% of the genes
@@
 
-```{r fig.height=8, fig.width=8}
-par(mfrow=c(4, 5), mar=c(0.1, 0.1, 0.1, 0.1))
-for(k in 1:n_clusters){
-    hist(all_scores[k, ], breaks=20, col="black")
-    abline(v=max_score, col="#AB0000")
-}
-```
-
 Summary of number of genes assigned with original scoring, and "better"
@@
 ```{r}
-table(kmeans_clusters$clusters)
 barplot(table(kmeans_clusters$clusters), col="black")
@@
 ```{r}
-table(labels)
 barplot(table(labels), col="black")
+labels = kmeans_clusters$clusters
 ```
@@
 ```{r fig.width=8, fig.height=8}
+cluster_to_plot = 13
 gene_names = row.names(data)
-genes_to_plot =  names(kmeans_clusters$clusters[kmeans_clusters$clusters == 15])
+genes_to_plot =  names(kmeans_clusters$clusters[kmeans_clusters$clusters == cluster_to_plot])
 data_to_plot = data[genes_to_plot, ]
@@
     annColors=ann_colors,
-    main="Cluster 15",
+    main=paste("Cluster", cluster_to_plot),
     treeheight=0)

### 2019-04-30 scripts/04_downstream_analysis.Rmd
FIXes and more

--- 2019-03-06 scripts/04_downstream_analysis.Rmd
@@
 ```{r enrichment_analysis, results="hide", message=FALSE}
-clusters = sort(as.vector(unlist(unique(labels))))
-clusters = c(1, 2, 3)
-all_pathways = data.frame()
-for(cluster in clusters){
+cluster = 13
 genes = gene_names[labels == cluster]
@@
     genes, species="mmu",
-	download_latest=TRUE)
+    download_latest=TRUE)$stastic
 
-    if((dim(all_pathways)[1] == 0) & (dim(pathways$stastic)[1] != 0)){
-	print("Initialize")
-	pathways$stastic["Cluster"] = cluster
-	all_pathways = pathways$stastic
-    }else if(dim(pathways$stastic)[1] != 0) {
-	print("Merge")
-	pathways$stastic["Cluster"] = cluster
-	all_pathways = merge(all_pathways, pathways$stastic, all=TRUE)
-    }
-}
 ```
@@
 
-all_pathways = all_pathways[
-    order(all_pathways["Cluster"],
-	  all_pathways["pvalueAdj"]),]
+pathways = pathways[
+    order(pathways["pvalueAdj"]),]
 
 # Select the columns to print
-all_pathways = subset(
-    all_pathways,
+pathways = subset(
+    pathways,
     select=c(
 	"Pathway_Name", "Percentage",
-	"pvalueAdj", "Cluster"))
+	"pvalueAdj"))
 
-colnames(all_pathways) = c("Pathway", "Percentage", "Adj. p-value", "Cluster")
-all_pathways["Percentage"] = all_pathways["Percentage"] * 100
-kable(all_pathways, "html", escape=F, digits=4) %>%
+colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
+pathways["Percentage"] = pathways["Percentage"] * 100
+kable(pathways, "html", escape=F, digits=4) %>%
     kable_styling("striped", full_width=TRUE) 
@@
 
-cluster = c(1, 2, 3)
-ontologies = c("BP", "NF", "CC")
-# It's taking too long to compile with everything. Just limit ourselves for
-# now to BP
-ontologies = c("BP")
-all_go_terms = data.frame()
-for(cluster in clusters){
-    for(ontology in ontologies){ 
 scores = as.numeric(labels == cluster)
@@
     scores,
-	    gene_id_go_mapping, ontology=ontology)
+    gene_id_go_mapping, ontology="BP")
 
-	if((dim(all_go_terms)[1] == 0) & (dim(go_terms_enriched)[1] != 0)){
-	    go_terms_enriched["Cluster"] = cluster
-	    go_terms_enriched["Ontology"] = ontology
-	    all_go_terms = go_terms_enriched
-	}else if(dim(go_terms_enriched)[1] != 0){
-	    go_terms_enriched["Cluster"] = cluster
-	    go_terms_enriched["Ontology"] = ontology
-	    all_go_terms = merge(all_go_terms, go_terms_enriched, all=TRUE)
-	}
-    }
-}
 ```
@@
 
-all_go_terms = all_go_terms[
-    order(all_go_terms[, "Cluster"],
-	  all_go_terms[, "Ontology"],
-	  all_go_terms[, "resultFisher_padj"]), ]
+go_terms_enriched = go_terms_enriched[
+    order(go_terms_enriched[, "resultFisher_padj"]), ]
 
-kable(all_go_terms, "html", escape=F, digits=4) %>%
+kable(go_terms_enriched, "html", escape=F, digits=4) %>%
     kable_styling("striped", full_width=TRUE) 

### 2019-04-30 bin/moanin/R/splines.R

@@ -170,7 +170,10 @@ score_genes_centroid = function(data, centroid, positive_scaling=TRUE, scale=TRU
     data_fitted = align_data_onto_centroid(
 	data, centroid, positive_scaling=positive_scaling)
 
-    scores = apply(data_fitted, 1, function(y){sqrt(sum((centroid - y)^2))})
+    scores = apply(
+	data_fitted,
+	1,
+	function(y){sqrt(sum((centroid - y)^2))})
 
     if(scale){
         all_zeros_gene = data_fitted[1, ] * 0

### 2019-04-30 scripts/run_clustering.R

@@ -8,7 +8,6 @@ n_clusters = strtoi(args[1])
 random_seed = strtoi(args[2])
 
 # Other options
-n_genes_to_keep = 5000
 filter_genes = TRUE
 sample_proportion = 0.8
 keep_all_significant = FALSE
@@ -21,7 +20,7 @@ outname = paste0(
 # Clustering and everything 
 data("shoemaker2015")
 
-data = shoemaker2015$data
+data = shoemaker2015$data[1:100,]
 meta = shoemaker2015$meta
 
 data = read.table(".results/quantile_normalized.txt") 

### 2019-05-01 scripts/01_quality_control_normalization.Rmd
FIX/ENH tweaks to figures and text for QC + normalization

@@
 45018 probes (1 sample was removed by the authors prior to uploading the
-dataset on GEo).
+dataset on GEO).
 
@@
 estimating the variations across samples before and after normalization is a
-first quality control. Typically, two other quality control and exploratory
-analysis steps are also performed before and after normalization: (1) low
-dimensionality embedding of the samples; (2) correlation plots between each
-samples. In both cases, we expect a strong biological signal, while replicate
-samples should be strongly clustered or correlated with one another.
+first quality control step. Typically, two other quality control and
+exploratory analysis steps are also performed before and after normalization:
+(1) low dimensionality embedding of the samples; (2) correlation plots between
+each samples. In both cases, we expect a strong biological signal, while
+replicate samples should be strongly clustered or correlated with one another.
 
 Before performing any additional exploratory analysis, let us only keep highly
-variable genes.
+variable genes: we keep for this step only the top 50% most variable genes.
 
@@
 
+# Reorder the conditions such that:
+#	- Control is before any influenza treatment
+#	- Each treatment is ordered from low to high pathogeny
+meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 1, 2, 5, 4)])
+
 # Reorder genes on condition, time, and replicate
-meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 1, 2, 5, 4)])
 ord = order(
@@
     Colv=NA, Rowv=NA,
-    annCol=meta[,
-		c("Group", "Timepoint")],
-    annLegend=FALSE, 
+    annCol=meta[, c("Group", "Timepoint")],
+    annRow=meta[, c("Group", "Timepoint")],
+    annLegend=TRUE, 
     annColors=ann_colors,

### 2019-05-01 scripts/utils.R

@@ -8,7 +8,7 @@ time_colors = grDevices::rainbow(14)
 names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
 
 ann_colors = list(
-    Time=time_colors,
+    Timepoint=time_colors,
     Group=group_colors
     )
 

### 2019-05-01 scripts/03_clustering.Rmd
Disabling test to cache packages

@@
 expressed. First, we select the genes that have a log-fold change of at least
-2 across at least one experiment for one time point. We thus use the object
-`lfc_max`.
-
-XXX
+2 across at least one experiment for one time point.
 
@@
 ```{r clustering}
-
-#############################################################################
-# Fit the splines
-
+# First fit the kmeans clusters
 kmeans_clusters = moanin:::splines_kmeans(
@@
 
-all_scores = matrix(NA, nrow=dim(data)[1], ncol=n_clusters)
-
-for(k in 1:n_clusters){
-     scores = moanin:::score_genes_centroid(
-	data,
-	kmeans_clusters$centroids[k,])
-     all_scores[, k] = scores
-}
-
-all_scores = as.matrix(all_scores)
-# Only assign labels to X% of the genes
-max_score = quantile(rowMin(all_scores), c(percentage_genes_to_label))
-genes_to_not_consider = rowMin(all_scores) >= max_score
-labels = apply(all_scores, 1, which.min)
-labels[genes_to_not_consider] = NA
-names(labels) = row.names(data)
+# Then assign scores and labels to all the data, using a goodness-of-fit
+# scoring function.
+scores_and_labels = moanin::splines_kmeans_score_and_label(
+    data, kmeans_clusters)
+labels = scores_and_labels$labels
 ```
@@
 barplot(table(labels), col="black")
-labels = kmeans_clusters$clusters
 ```

### 2019-05-01 .build_tools/travis/test_script.sh

@@ -27,5 +27,5 @@ run_tests() {
 }
 
 get_data
-run_tests
+#run_tests
 mkdir -p scripts/reports

### 2019-05-01 scripts/04_downstream_analysis.Rmd
Changes to downstream analysis

@@
 Once the gene ID to GO mapping list is created, `moanin` provides a simple
-interface to `topGO` to fetch enriched GO terms.
+interface to `topGO` to fetch enriched GO terms. Here, we show an example of
+running a GO term enrichment on the "Biological process" ontology (BP).
 
 ```{r go_term_enrichment, results="hide", message=FALSE}
-
 scores = as.numeric(labels == cluster)
@@
     gene_id_go_mapping, ontology="BP")
-
 ```

### 2019-05-01 scripts/01_quality_control_normalization.Rmd
Switch the California \& Kawasaki strains

@@
 #	- Each treatment is ordered from low to high pathogeny
-meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 1, 2, 5, 4)])
+meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
 

### 2019-05-01 scripts/00_introduction.Rmd
Expanding introduction

@@
 
-The authors find that
-
-**XXX Add summary of the key findings of the the paper**
+By combining gene expression time-course data with virus growth data, the
+authors show that the inflammatory response of lung tissue is gated until a
+threshold of the virus concentration is  exceeded in the lung. Once this
+threshold is exceeded, a strong inflammatory and cytokin production occurs.
+This results provides evidence that the pathology response is non-linearly
+regulated by virus concentration.
 

### 2019-05-02 scripts/02_differential_expression_analysis.Rmd
Implementing consensus strategy

@@
 ```
+
+## Differential expression analysis of time-course data
+
+The next step is typically to run a differential expression analysis. Most of
+`moanin`'s function rely on an object containing a number of information such
+as the metadata and the splines model used.
+
 ```{r}
@@
 ```
-
-## Differential expression analysis of time-course data
-
-The next step is typically to run a differential expression analysis.
 

### 2019-05-02 scripts/03_clustering.Rmd
Implementing consensus strategy
Removing consensus method (takes too long)

@@
 ```{r}
-n_clusters = 5:20
+n_clusters = 2:20
 all_labels = list()
@@
 
+#### The model explorer strategy
+
 ```{r fig.width=8, fig.height=8}
-library(moanin)
 moanin::plot_model_explorer(all_labels)
 ```
+
+#### Looking at the CDF of the consensus matrix
+
+```{r fig.width=8, fig.height=8}
+#moanin::plot_cdf_consensus(all_labels)
+```

### 2019-05-02 scripts/00_introduction.Rmd
Adding schemas of the workflow + rewording

@@
 proposed for time course data are concerned with detecting genes that are
-changing over time (differential expression analysis), examples being
-`edge` [@storey:significance], functional component analysis based models
-[@wu:more], time-course permutation tests [@park:statistical], XXX
-[@sun:multiple]
-**TODO: Other methods to mention: impulse models, what else?
-Citations!**. However, with long time course datasets, particularly in
-developmental systems, a massive number of genes will show *some* change.
-For example, in the \xxx system, over \xxx\% of genes are shown to be changing
+changing over time (differential expression analysis), examples being `edge`
+[@storey:significance], functional component analysis based models [@wu:more],
+time-course permutation tests [@park:statistical], and multiple testing
+strategies to combine singe time point differential expression analysis
+[@sun:multiple]. However, with long time course datasets, particularly in
+developmental systems, a massive number of genes will show *some* change. For
+example, in the \xxx system, over \xxx\% of genes are shown to be changing
 over time. The task in these settings is often not to detect changes in genes,
@@
 previously established methods and new methodologies for the settings of
-developmental time series. We provide the various steps of the workflow as
+developmental time series. It relies on several standard packages for
+analysing gene expression data, some specific for time-course data, others
+broadly used by the community. We provide the various steps of the workflow as
 functions in a R package **FIXME**.
 
----
-# Old intro
-
-Time-course gene expression analysis workflows include steps used more broadly
-for gene expression data analysis such as quality control (QC), exploratory
-data analysis (EDA), pre-processing, normalization, and visualization. Yet, a
-typical gene expression data analysis workflow is not entirely adequate for
-analysing time-course data. In particular, differential expression analysis
-needs to be adapted to time-course data, and often needs to be followed by a
-clustering step to group genes sharing similar patterns of expression.
-
-Here, we propose an integrated workflow for downstream analysis of time-course
-data, with the following main steps: (1) quality control and normalization;
-(2) differential expression analysis of time-course data; (3) clustering; (4)
-downstream analysis of clusters with GO term, pathway, and motif enrichment.
-
-This workflow relies on several standard packages for analysing gene
-expression data, some specific for time-course data, others broadly used by
-the community. We also provide alongside this workflow a package that nicely
-wraps all calls to dependencies with a common API by defining a model object
-holding all necessary metadata to perform all steps of the pipeline.
----
+```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
+library(knitr)
+knitr::include_graphics("images/fig1_schemas.png")
+```
 

### 2019-05-02 scripts/01_quality_control_normalization.Rmd
Adding schemas of the workflow + rewording

@@
 data. Finally, we further filter out probes based on the following criteria:
-(1) removing the `r n_control_probes` control probes; (2) removing `r
-n_lowly_expressed` low-expressed genes.
+(1) removing the `r n_control_probes` control probes; (2) removing 
+`r n_lowly_expressed` low-expressed genes.
 
@@
 
- XXX The original paper says the data is "averaged over unique probes IDs."
-
 ```{r echo=FALSE, results="hide"}
@@
 
-After filtering, we are left with `r n_probes` probes of interest in `r
-n_samples`.
+After filtering, we are left with `r n_probes` probes of interest in 
+`r n_samples` samples.
+
+Now that we have normalized and filtered the gene expressions, let's load the
+metadata. The package `moanin` contains the normalized data and meta of
+[@shoemaker:ultrasensitive].
 
@@
 # Now load in the metadata
-# meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")
 data(shoemaker2015)
 meta = shoemaker2015$meta
+```
+
+```{r echo=FALSE, results="hide"}
 write.table(data, ".results/quantile_normalized.txt")

### 2019-05-02 scripts/03_clustering.Rmd
Swapping from 0.8 to 1 proportion

@@
 ```{r filter_genes}
-if(filter_genes){
 # First, filter out any genes that doesn't have a log fold change of at
@@
 
-    if(length(genes_to_keep) > n_genes_to_keep){
-
-	# Then rank by fisher's p-value and take max the number of genes of
-	# interest
+# Then rank by fisher's p-value and take max the number of genes of interest
 # Filter out q-values for the pvalues table
@@
 fishers_qval = stats::p.adjust(fishers_pval)
-        if(keep_all_significant){
-	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
-	}else{
 genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
-	}
-    }
 
 y = as.matrix(data[genes_to_keep, ])
-}else{
-    y = as.matrix(data)
-}
 ```
@@
 
-The last strategy is to look at the stability of the clustering.
+On real data, the number of clusters is not only unknown but also 
+ambiguous: it will depend on the desired clustering resolution of the user.
+Yet, in the case of biological data, stability and reproducibility of the
+results is necessary to ensure that the biological interpretation of the
+results hold when the data or the model is exposed to reasonable
+perturbations.
+
+Methods that rely on the stability of the clustering results to choose $k$
+thus ensure that the biological interpretation of the clusters hold with
+perturbtation to the data. In addition, simulation where the data is generated
+with a well defined $k$ show that the clustering is more stable for the
+correct of number of the clusters.
+
+Most methods method to find the number of clusters with stability measures
+only provide visual aids to guide the user. The first element often visualized
+is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
+stores the proportion of clustering in which two items are clustered together.
+A perfect consensus matrix ordered such as each elements that belong to the
+same cluster are adjacent to one another which show blocks along the diagonal
+close to 1.
+
+Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
 
@@
 
-stability_20 = read.table("results/stability_15.tsv", sep="\t")
+stability_20 = read.table("results/stability_20.tsv", sep="\t")
 consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
@@
 
-```{r}
+```{r echo=FALSE, results="hide"}
 n_clusters = 2:20
@@
 
+The model explorer algorithm [@ben-hur:stability] proposes to estimate the
+number of clusters exploiting the observation that if the number of clusters
+is correct, the clustering results are stable to bootstrapping.  The
+distribution of similaries between bootstrapped results for each $k$ can thus
+be compared for different values of $k$ and guide the user in the choice of
+number of clusters.
+
+The model explorer strategy works as follows. First, choose a similarity
+measure between two partitions or clusters $S(C_1, C_2)$. Examples are the
+normalized mutual information or Fowlkes-Mallows. Then perform $n$ bootstrap
+experiments to estimate the cluster centroids, followed by a step of assigning
+a label to all data points. Finally, compute the pairwise similarity measure
+between all bootstrapped partition, and plot the cumulative density of the
+obtained scores.
+
 ```{r fig.width=8, fig.height=8}
@@
 
-#### Looking at the CDF of the consensus matrix
+From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
+but not as stable as $k=2$. Similarly, $k=14$ is more stable than $k=11$,
+$k=12$, $k=13$. The model explorer strategy, in addition to visualizing the
+diversity of the centroids, can thus help assessing an adequate number of
+clusters.
 
-```{r fig.width=8, fig.height=8}
-#moanin::plot_cdf_consensus(all_labels)
+#### Consensus clustering as a way to find $k$
+
+The  consensus clustering [@monti:consensus] relies on a similar idea but
+instead of looking at the cumulative density of similarity measures of
+bootstrapped clustering, the authors suggests plotting the cumulative density
+of elements of the consensus matrix.
+
+```{r eval=FALSE}
+moanin::plot_cdf_consensus(all_labels)
 ```
+
+```{r echo="FALSE"}
+library(knitr)
+knitr::include_graphics("images/clustering_CDF_consensus.png")
+```
+
+The stability of the clustering based on the consensus matrix can then be
+measured via a single number by looking at the area under the curve: the more
+stable the clustering, the closer to 0 or 1 will be the entries of consensus
+matrix. The consensus clustering strategy thus suggest at looking at either
+the AUC as a function of the number of the clusters or the "improvement" in
+the AUC as a function of the number of cluster.
+
+```{r echo="FALSE"}
+library(knitr)
+knitr::include_graphics("images/clustering_CDF_consensus.png")
+```

### 2019-05-02 bin/moanin/R/cluster.R

@@ -93,7 +93,7 @@ splines_kmeans_prediction = function(data, kmeans_clusters){
 #'
 #' @param data array or dataframe.
 #'	Data to assign score and labels to
-#' @param kmeans_cluster list of list
+#' @param kmeans_clusters list of list
 #'	List returned by moanin::splines_kmeans
 #' @param percentage_genes_to_label float, optional, default: 0.5
 #'	Percentage of genes to label.
@@ -115,7 +115,7 @@ splines_kmeans_score_and_label = function(data, kmeans_clusters, percentage_gene
     row.names(all_scores) = row.names(data) 
 
     # Only assign labels to X% of the genes
-    max_score = quantile(rowMin(all_scores), c(percentage_genes_to_label))
+    max_score = stats::quantile(rowMin(all_scores), c(percentage_genes_to_label))
     genes_to_not_consider = rowMin(all_scores) >= max_score
     labels = apply(all_scores, 1, which.min)
     labels[genes_to_not_consider] = NA

### 2019-05-02 bin/moanin/R/cluster_validation.R

@@ -60,7 +60,7 @@ plot_cdf_consensus = function(labels){
     }
 
     graphics::legend(1, 0.8, legend=n_clusters,
-       col=colors, lty=1:2, cex=0.8,
+       col=colors,
        title="Clusters", text.font=4)
 
 }
@@ -70,33 +70,40 @@ plot_cdf_consensus = function(labels){
 #' @param labels list of lables
 #'
 #' @export
-get_auc_consensus_scores = function(labels){
+get_auc_similarity_scores = function(labels, method="consensus"){
+    if(!(method %in% c("consensus", "nmi"))){
+	msg = paste(
+	    "moanin::get_auc_similarity_score.",
+	    "Unknown similarity measure provided", method)
+	stop(msg)
+    }
+
     all_labels = labels
     n_clusters = names(all_labels)
 
     auc_scores = rep(0, length(n_clusters))
     for(i in 1:length(n_clusters)){
 	cluster = n_clusters[i]
-
 	labels = all_labels[[cluster]]
-	consensus = consensus_matrix(labels, scale=FALSE)
-	consensus = consensus / max(consensus)
-	consensus = sort(consensus[upper.tri(consensus)])
-	y_axis = 1:length(consensus) / length(consensus)
-	auc_score = sum(diff(consensus) * zoo::rollmean(y_axis, 2))
+	if(method == "consensus"){
+	    consensus = consensus_matrix(labels, scale=FALSE)
+	    consensus = consensus / max(consensus)
+	    scores = consensus[upper.tri(consensus)]
+	}else if(method == "nmi"){
+	    scores = get_nmi_scores(labels) 
+	}
+	scores = sort(scores)
+	y_axis = 1:length(scores) / length(scores)
+	auc_score = sum(diff(scores) * zoo::rollmean(y_axis, 2))
+	
 	auc_scores[i] = auc_score
     }
     return(auc_scores)
 }
 
 
-#' Plot model explorer
-#'
-#' @param labels list of labels
-#' @export
-plot_model_explorer = function(labels){
-    all_labels = labels
-    n_clusters = names(all_labels)
+get_nmi_scores = function(labels){
+
     nmi = function(x, y){
 	x_dataframe = as.data.frame(x)
 	colnames(x_dataframe) = c("Label")
@@ -111,6 +118,30 @@ plot_model_explorer = function(labels){
 	return(NMI::NMI(x_dataframe, y_dataframe))
     }
 
[…]

### 2019-05-02 bin/moanin/R/splines.R

@@ -58,9 +58,9 @@ fit_predict_splines = function(data, splines_model, weights=NULL, meta_predictio
 	    "fitting_data"=fitting_data,
 	    "degrees_of_freedom"=splines_model$degrees_of_freedom)
 
-	updated_formula = update(splines_model$formula, fitting_data ~ .)
-	model = lm(updated_formula, formula_data)	
-	y_fitted = predict(model, meta_prediction)
+	updated_formula = stats::update(splines_model$formula, fitting_data ~ .)
+	model = stats::lm(updated_formula, formula_data)	
+	y_fitted = stats::predict(model, meta_prediction)
     }
     return(y_fitted)
 }

### 2019-05-03 scripts/01_quality_control_normalization.Rmd
Switch bootstrapping proportion

@@
 
+Given the source and a `Targets.txt` structured as follows, microarray can be loaded thanks to the limma library.
+
+```
+FileName
+GSM1557140_AR0313_043raw.txt.gz
+GSM1557141_AR0313_044raw.txt.gz
+GSM1557142_AR0313_045raw.txt.gz
+GSM1557143_AR0313_055raw.txt.gz
+GSM1557144_AR0313_056raw.txt.gz
+GSM1557145_AR0313_057raw.txt.gz
+GSM1557146_AR0313_058raw.txt.gz
+GSM1557147_AR0313_059raw.txt.gz
+```
+
+```{r}
+targets = limma::readTargets()
+RG = limma::read.maimages(
+  targets, source="agilent",
+  path=data_dir,
+  other.columns="gIsWellAboveBG",
+  green.only=TRUE)
+
+```
+
 The normalization strategy we apply here relies on `limma` and is as follows:
@@
 ```{r reading_targets, results="hide", fig.width=8, fig.height=4}
-# Read in the data using limma.
-targets = limma::readTargets()
-RG = limma::read.maimages(
-  targets, source="agilent",
-  path=data_dir,
-  other.columns="gIsWellAboveBG",
-  green.only=TRUE)
-
-# Preform background correction using norm-exponential strategy and normalize
+# Perform background correction using norm-exponential strategy and normalize
 # across samples using quantile normalization.

### 2019-05-03 scripts/03_clustering.Rmd
Switch bootstrapping proportion

@@
 
+To perform such analysis, the first step is run the clustering several times
+on a resampled dataset--either using bootstrap or subsampling.
+
+Using the bootstrapping strategy:
+```{r}
+n_genes = dim(y)[1]
+indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
+
+bootstrapped_y = y[indices, ]
+```
+
+Using the subsampling strategy, keeping 80% of the genes:
+```{r}
+subsample_proportion = 0.8
+indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=TRUE)
+subsampled_y = y[indices, ]
+```
+
 Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
@@
 
-```{r echo="FALSE"}
+```{r echo=FALSE}
 library(knitr)
@@
 
-```{r echo="FALSE"}
+```{r echo=FALSE}
 library(knitr)
-knitr::include_graphics("images/clustering_CDF_consensus.png")
+knitr::include_graphics("images/clustering_AUC_consensus.png")
 ```
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_delta_AUC_consensus.png")
+```

### 2019-05-03 scripts/04_downstream_analysis.Rmd
Switch bootstrapping proportion

@@
 library(biomaRt)
-ensembl = useMart("ensembl")
-ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+ensembl = biomaRt::useMart("ensembl")
+ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
 ```
@@
 # convert gene names
-genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
 		filters="refseq_mrna", values=genes,
@@
 # FIXME move this to the package
-source("GO_terms_analysis.R")
-genes = getBM(attributes=c("go_id", "refseq_mrna"),
+genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
 	      values=gene_names,
@@
 ```{r go_term_enrichment, results="hide", message=FALSE}
-scores = as.numeric(labels == cluster)
-names(scores) = gene_names
 go_terms_enriched = find_enriched_go_terms(
     scores,
+    13,
     gene_id_go_mapping, ontology="BP")

### 2019-05-03 bin/moanin/R/GO_terms_analysis.R

@@ -3,8 +3,9 @@ library(topGO)
 
 #' Find enriched GO terms
 #'
-#' @param labels
-#' @param gene_id_to_go
+#' @param labels named vector containing labels
+#' @param cluster which cluster onto which to run the GO term analysis.
+#' @param gene_id_to_go Gene ID to GO object from topGO
 #' @param ontology string, optional, default: BP
 #'	specficies which ontology to use. Can be 'BP', 'CC', or 'NF'
 #' @param weighted, boolean, optional, default: FALSE
@@ -12,10 +13,15 @@ library(topGO)
 #' @param node_size integer, optional, default: 10
 #'	Consider only GO terms with node_size number of genes.
 #' @param ontology: BP, CC, NF
-find_enriched_go_terms = function(labels, gene_id_to_go,
+#' @export
+find_enriched_go_terms = function(labels, cluster, gene_id_to_go,
 				  ontology="BP", 
 				  weighted=FALSE,
 				  node_size=10){
+    gene_names = row.name(labels)
+    scores = as.numeric(labels == cluster)
+    names(scores) = gene_names
+
     if(!(ontology %in% c("BP", "CC", "NF"))){
 	error_message = paste(
 	    "moanin::find_enriched_go_terms: Ontology should be 'BP', 'CC',",
@@ -61,6 +67,11 @@ find_enriched_go_terms = function(labels, gene_id_to_go,
 
 #' Create the Gene to GO Term mapping
 #'
+#' @param genes dataframe, with `gene_col` and a column corresponding to the
+#'  `go_id`
+#' @param gene_col the column of the genes data frame that contains the
+#'  correct gene reference. By default, is refseq_mrna
+#' @export
 create_go_term_mapping = function(genes, gene_col="refseq_mrna"){
     gene_id_go_mapping = NULL
     gene_names = unique(genes[, "refseq_mrna"])

### 2019-05-06 scripts/04_downstream_analysis.Rmd
04_downstream_analysis.Rmd needs to import topGO

@@
 gene_names = gene_names[!is.na(labels)]
-labels = labels[!is.na(labels)]
+labels = as.vector(labels[!is.na(labels)])
+names(labels) = gene_names
 ```
@@
 `biomaRt` queries results a matrix with two named columns of gene names and GO
-term ID. **FIXME** matrix or data.frame??
+term ID.
 
 ```{r go_to_gene_mapping, results="hide", message=FALSE}
-# FIXME move this to the package
+library(topGO)
+#source("GO_terms_analysis.R")
 genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
@@
 ```{r go_term_enrichment, results="hide", message=FALSE}
+assignments = labels == cluster
+
 go_terms_enriched = find_enriched_go_terms(
-    scores,
-    13,
+    assignments,
     gene_id_go_mapping, ontology="BP")
 ```
-
-```{r echo=FALSE}
-
-go_terms_enriched = go_terms_enriched[
-    order(go_terms_enriched[, "resultFisher_padj"]), ]
-
-kable(go_terms_enriched, "html", escape=F, digits=4) %>%
-    kable_styling("striped", full_width=TRUE) 
-```

### 2019-05-06 bin/moanin/R/GO_terms_analysis.R

@@ -3,8 +3,7 @@ library(topGO)
 
 #' Find enriched GO terms
 #'
-#' @param labels named vector containing labels
-#' @param cluster which cluster onto which to run the GO term analysis.
+#' @param assignments boolean named vector
 #' @param gene_id_to_go Gene ID to GO object from topGO
 #' @param ontology string, optional, default: BP
 #'	specficies which ontology to use. Can be 'BP', 'CC', or 'NF'
@@ -14,14 +13,13 @@ library(topGO)
 #'	Consider only GO terms with node_size number of genes.
 #' @param ontology: BP, CC, NF
 #' @export
-find_enriched_go_terms = function(labels, cluster, gene_id_to_go,
+find_enriched_go_terms = function(assignments, gene_id_to_go,
 				  ontology="BP", 
 				  weighted=FALSE,
 				  node_size=10){
-    gene_names = row.name(labels)
-    scores = as.numeric(labels == cluster)
-    names(scores) = gene_names
-
+    gene_names = names(assignments)
+    assignments = as.numeric(assignments)
+    names(assignments) = gene_names
     if(!(ontology %in% c("BP", "CC", "NF"))){
 	error_message = paste(
 	    "moanin::find_enriched_go_terms: Ontology should be 'BP', 'CC',",
@@ -35,18 +33,19 @@ find_enriched_go_terms = function(labels, cluster, gene_id_to_go,
       return(data < 0.5)
     }
 
-    GOdata = new("topGOdata", ontology=ontology, allGenes=labels, nodeSize=node_size,
+    GOdata = new("topGOdata", ontology=ontology, allGenes=assignments,
+		 nodeSize=node_size,
 		 geneSel=getTopDiffGenes,
 		 annot=annFUN.gene2GO, gene2GO=gene_id_to_go)
 
     if(weighted){
-        resultFisher = runTest(GOdata, algorithm="classic", statistic="fisher")
+        resultFisher = topGO::runTest(GOdata, algorithm="classic", statistic="fisher")
     }else{
-        resultFisher = runTest(GOdata, algorithm="weight", statistic="fisher")
+        resultFisher = topGO::runTest(GOdata, algorithm="weight", statistic="fisher")
     }
     n_nodes = length(resultFisher@score)
 
-    allRes = GenTable(
+    allRes = topGO::GenTable(
 	GOdata, 
 	resultFisher=resultFisher,
 	orderBy="resultFisher", ranksOf="resultFisher",

### 2019-05-06 bin/moanin/R/splines.R

@@ -77,7 +77,7 @@ create_meta_prediction = function(splines_model, num_timepoints=100){
     # Create splines_model for prediction
     timepoints_pred = NULL
     groups_pred = NULL
-    meta = splines_model$meta
+    meta = droplevels(splines_model$meta)
     groups = levels(meta$Group) 
     for(group in groups){
 	mask = meta$Group == group

### 2019-05-07 scripts/utils.R
Better text + colors

@@ -1,7 +1,7 @@
 group_colors = c(
     "M"="dodgerblue4",
-    "C"="gold",
-    "K"="orange",
+    "K"="gold",
+    "C"="orange",
     "VH"="red4",
     "VL"="red2")
 time_colors = grDevices::rainbow(14)

### 2019-05-07 scripts/00_introduction.Rmd
Better text + colors
ENH Now plotting top DE genes

@@
 ---
+
+# TODO list
+
+- Add legends to the plots that don't have.
+- Currently have two functions "plot_genes", "plot_centroids", that do exactly
+  the same thing. Find a better name for both of them
+- On the schema plot, replace clustering schemas with something else.
 
@@
 developmental systems, a massive number of genes will show *some* change. For
-example, in the \xxx system, over \xxx\% of genes are shown to be changing
+example, in the **xxx** system, over **xxx%** of genes are shown to be changing
 over time. The task in these settings is often not to detect changes in genes,

### 2019-05-07 scripts/02_differential_expression_analysis.Rmd
ENH Now plotting top DE genes

@@
 library(moanin)
+source("utils.R")
 ```
@@
 meta = shoemaker2015$meta
+meta = droplevels(meta)
 
@@
 lfc_epicon = log_fold_change_epicon[, "M-K"]
+names(lfc_epicon) = row.names(log_fold_change_epicon)
 
@@
        -log10(pvalue[(pvalue < 0.01) & (abs(lfc_epicon) > 1)]), pch=20, col="green")
+
+```
+
+Let's quickly visualize the top 10 DE genes ranked by p-value.
+
+```{r fig.height=8, fig.width=8}
+top_DE_genes_pval = names(sort(pvalue)[1:12])
+moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
+		   colors=ann_colors$Group, smooth=TRUE)
+```
+
+Now visualize the top 10 DE genes by log fold change
+
+```{r fig.height=8, fig.width=8}
+top_DE_genes_lfc = names(
+    sort(abs(lfc_epicon),
+	 decreasing=TRUE)[1:12])
+moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
+		   colors=ann_colors$Group, smooth=TRUE)
 

### 2019-05-07 bin/moanin/R/visualization.R

@@ -1,6 +1,7 @@
 library(viridis)
 library(graphics)
 
+
 #' Plotting centroids
 #'
 #' @param centroids matrix (k, t) containing the centroids
@@ -43,13 +44,30 @@ plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
 }
 
 
+#' Plotting data
+#'
+#' @param data matrix (k, t) containing the centroids
+#' @param splines_model splines_model
+#' @param meta	data.frame (t, n) containing the metadata.
+#' @param colors vector, optional, default NULL
+#'		vector of colors
+#' @param smooth boolean, optional, default: FALSE
+#'  Whether to smooth the centroids or not.
+#' @export
+plot_genes = function(data, splines_model, colors=NULL, smooth=FALSE){
+    plot_centroids(data, splines_model, colors=colors, smooth=smooth)
+}
+
+
 plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth=FALSE){
     meta = splines_model$meta
     groups = levels(meta$Group)
 
     xrange = range(meta$Timepoint)
     yrange = range(centroid)
-    centroid = t(as.matrix(centroid))
+    if(is.null(dim(centroid))){
+        centroid = t(as.matrix(centroid))
+    }
 
     graphics::plot(xrange, yrange, type="n")
     if(is.null(colors)){

### 2019-05-07 scripts/03_clustering.Rmd
Better text + colors
FIX clustering

@@
 fishers_qval = stats::p.adjust(fishers_pval)
-genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
+genes_to_keep = names(sort(fishers_pval))
+if(length(genes_to_keep) > n_genes_to_keep){
+    genes_to_keep = genes_to_keep[1:n_genes_to_keep]
+}
 
@@
 ```
+
+The consensus clustering method suggest that the most stable is $k=2$, which
+separates over-expressed genes from under-expressed genes. While it is indeed
+a very stable clustering, it does not capture the range of gene expression
+patterns present in the data.

### 2019-05-07 scripts/01_quality_control_normalization.Rmd
More text, updated stability results, etc

@@
 
-```{r}
+```{r message=FALSE}
 targets = limma::readTargets()
@@
 
-```{r reading_targets, results="hide", fig.width=8, fig.height=4}
+```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE}
 # Perform background correction using norm-exponential strategy and normalize
@@
 ```{r pca_plots, fig.width=8, fig.height=8}
-# FIXME Add legend
 graphics::par(mfrow=c(2, 2))
@@
     xlab="PC2", ylab="PC3")
+
 ```
@@
 higher than the cross correlation amongst the rest of the treatments. Second,
-the influenza-infected mice midly react until time point 36. Third, the
-California strain samples seem closer to control samples than the
-lowly-pathogenic Kawasaki samples. Fourth, the Vietnam samples at time point
-120 and 168 are the one that are the most different from control samples.
+the influenza-infected mice midly react until time point 36. Third, the less
+pathogenic the strain is, the closer the samples are to the control condition.
+Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
+most different from control samples.

### 2019-05-07 scripts/02_differential_expression_analysis.Rmd
More text, updated stability results, etc

@@
 library(moanin)
+library(knitr)
+library(kableExtra)
 source("utils.R")
@@
 
-To leverage this temporal structure, Storey et al [CITE FIXME] proposes to
+To leverage this temporal structure, Storey et al [@storey:significance] proposes to
 model time-course micro-array with splines functions, and to use a log-ratio
@@
 
+```
+
+The number of genes found differentially expressed ranges from around 12000 to
+29000 depending on the strain and dosage of influenza virus given to the mice.
+This corresponds to between 30% to 70% of the genes found differentially
+expressed in this time-course experiment.
+
+```{r echo=FALSE, fig.width=6, fig.height=6}
+num_de = colSums(qvalues < 0.05)
+names(num_de) = c(
+    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
+barplot(num_de, col="black", ylab="Number of DE genes")
+```
+
+The next step in a classical differential expression analysis is typically to
+assess the effect of the treatment by looking at the log fold change.
+Computing the log fold change on a time-course experiment is not trivial: one
+can be interested in the average log-fold change across time, or the
+cumulative log-fold change. Sometimes a gene can be over-expressed at the
+beginning of the time-course data, and then over-expressed at the end of the
+experiment. As a result, `moanin` provides a number of possible ways to
+compute the log fold change across the whole time-course.
+
+First, `moanin` provides as simple interface to compute the log-fold change
+for each individual timepoints.
+
+```{r}
+log_fold_change_timepoints = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts,  method="timely")
+```
+
+```{r echo=FALSE}
+kable(head(log_fold_change_timepoints[,
+colnames(log_fold_change_timepoints)[1:5]]))
+```
+
+This matrix can then be used to visualize the log-fold change for each
+contrast per timepoint.
+
+Sometimes, a single value per gene and per contrast is more useful. Here is a
+table of the possible ways to compute log-fold change values.
+
+| Name   | Formula   |     |
+|------- | --------- |--- |
+| timely | $\text{lfc}(t)$  | Function of time |
+| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
+| abs_sum  | $\sum_t \|lfc(t)\|$  | Always positive |
+| max  | $\max_t \|lfc(t)\|$  | Always positive  |
+| min  | $\min_t \|lfc(t)\|$  | Always positive |
+| epicon  | $$  | Captures overall strength of response and overall direction |
+
+```{r}
 log_fold_change_epicon = moanin::estimate_log_fold_change(
@@
 
+log_fold_change_sum = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts,  method="sum")
+
 log_fold_change_max = moanin::estimate_log_fold_change(
     data, splines_model, contrasts, method="max")
+
+log_fold_change_min = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts, method="min")
 ```
 
-Let's explore a bit the number of genes found differentially expressed.
-
-```{r echo=FALSE, fig.width=6}
-num_de = colSums(qvalues < 0.05)
-names(num_de) = c(
-    "Kawasaki", "California", "Vietnam low-dose", "Vietname high-dose")
-barplot(num_de, col="black", ylab="Number of DE genes")
-
+```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE}
+par(mfrow=c(2, 2))
+plot(log_fold_change_epicon[, "M-C"],
+     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
+     xlab="EPICON", ylab="sum")
+plot(log_fold_change_epicon[, "M-C"],
+     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
+     xlab="EPICON", ylab="max")
+plot(log_fold_change_max[, "M-C"],
+     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
+     xlab="max", ylab="min")
+plot(log_fold_change_max[, "M-C"],
+     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
+     xlab="max", ylab="sum")
 ```
 
-And now a volcano plot for quick visualization
+From the single measures of log-fold change and the p-value, we can now look
+at the volcano plot. Here is an example of a volcano plot for the comparaison
+of the control to the Kawasaki strain, using the EPICON log fold change
+computation.
 
 ```{r fig.height=8, fig.width=8}
-
 pvalue = qvalues[, "M-K"]
@@
      xlim=c(-2.5, 2))
-
-# Add colored points: red if padj<0.05, orange of log2FC>1, green if both)
-points(lfc_epicon[pvalue < 0.01], -log10(pvalue[pvalue < 0.01]), pch=20,
-       col="red")
-points(lfc_epicon[abs(lfc_epicon) > 1],
-       -log10(pvalue[abs(lfc_epicon) > 1]), pch=20, col="orange")
-points(lfc_epicon[(pvalue < 0.01) & (abs(lfc_epicon) > 1)],
-       -log10(pvalue[(pvalue < 0.01) & (abs(lfc_epicon) > 1)]), pch=20, col="green")
-
 ```
 
-Let's quickly visualize the top 10 DE genes ranked by p-value.
+As another sanity check, `moanin` provides a simple utility function to
+visualize gene time-course data. Here, we plot the 12 genes with the smallest
+p-values.
 
@@
 
-Now visualize the top 10 DE genes by log fold change
+And here we visualize the genes with the largest absolute EPICON log-fold
+change.
 
@@
 ```
+
+Thanks to those visualization, we can see that genes often follow similar
+patterns of expression, although on a different scale for each gene. We can
+leverage this observation to cluster the genes into groups of similar patterns
+of transcriptomic response.
 
@@
 ```
+
+# References

### 2019-05-07 scripts/03_clustering.Rmd
More text, updated stability results, etc

@@
 meta = shoemaker2015$meta
+meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
 
@@
 ```{r fig.width=8, fig.height=8}
-cluster_to_plot = 13
+cluster_to_plot = 20
 gene_names = row.names(data)
@@
 submeta = meta
-submeta$Group = factor(submeta$Group, levels(submeta$Group)[c(3, 1, 2, 5, 4)])
 ord = order(
@@
 
-    - Elbow method
-    - *Silhouette method* 
-    - *Model explorer*
+- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
+  the elbow methods looks at the total with cluster sum of squares 
+  as a function of the number of clusters (WCSS). When adding clusters
+  doesn't decrease the WCSS by a sufficient
+  amount, one can consider stopping. This method thus provides visual aid to
+  the user to choose the number of clusters, but often the "elbow" is hard to
+  see on real data, where the number of clusters is not clearly defined.
+- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
+  refers to a method of validation of consistency within clusters, and
+  provides visual aid to choose the number of clusters.
+- **Stability methods** Stability methods are more computationally intensive
+  than any other method, as they rely on assessing the stability of the
+  clustering for every $k$ to a small randomization of the data. The user is
+  then invited to choose the number of cluster based on a number of similarity
+  measures.
+
+First, let us run the clustering for all possible clusters of interest. We
+will, for each clustering experiment, conserve (1) with within cluster sum of
+squares; (2) the labels assigned to all genes.
 
 ```{r running_clustering_on_all_k}
-all_possible_n_clusters = c(5, 7, 9, 11)
+all_possible_n_clusters = c(5, 7, 9, 11, 13, 15)
 all_clustering = list()
@@
 ```{r}
-subsample_proportion = 0.8
-indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=TRUE)
+subsample_proportion = 1
+indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
 subsampled_y = y[indices, ]
@@
 a very stable clustering, it does not capture the range of gene expression
-patterns present in the data.
+patterns present in the data. This shows the limitation of such method on real
+data, where the number of clusters is not clearly defined.

### 2019-05-07 scripts/run_go_terms.R

@@ -8,6 +8,8 @@ library(viridis)
 
 library(moanin)
 library(biomaRt)
+library(topGO)
+
 df = 6
 
 # Preprocessing & filtering
@@ -37,7 +39,6 @@ labels = labels[!is.na(labels)]
 
 
 # FIXME move this to the package
-source("GO_terms_analysis.R")
 ensembl = useMart("ensembl")
 ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
 

### 2019-05-07 scripts/04_downstream_analysis.Rmd
More text, updated stability results, etc
GO term table had disappeared… put it back!

@@
 ```{r enrichment_analysis, results="hide", message=FALSE}
-cluster = 13
+cluster = 20
 genes = gene_names[labels == cluster]
@@
 library(topGO)
-#source("GO_terms_analysis.R")
 genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
@@
 ```
+
+```{r echo=FALSE}
+
+go_terms_enriched = go_terms_enriched[
+    order(go_terms_enriched[, "resultFisher_padj"]), ]
+
+kable(go_terms_enriched, "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```

### 2019-05-08 scripts/03_clustering.Rmd
Reducing number of cluster
More and better text. also better k for clustering

@@
 
-The next step is to perform clustering. First, we reduce the set of genes of
-interest to genes that are (1) significant; (2) "highly" differentially
-expressed. First, we select the genes that have a log-fold change of at least
-2 across at least one experiment for one time point.
+The very large number of genes found differentially expressed impair any
+interpretation one would attempt: with 70\% of the genome found differentially
+expressed, all pathways are affected by the treatment. Hence the next step of
+the workflow to cluster gene expression according to their dynamical response
+to the treatment. Before clustering the genes, we first reduce the set of
+genes of interest to genes that are (1) significantly found differentially
+expressed; (2) "highly" differentially expressed. To do this, we first
+aggregate all p-values obtained during the time-course differential expresison
+step in a single p-value using Fisher's method [@fisher:statistical]. Then we
+select all the genes which have a Fisher adjusted p-value below 0.05 and a log
+fold change of at least two between at least one condition and one time-point.
 
 ```{r filter_genes}
-# First, filter out any genes that doesn't have a log fold change of at
-# least 2 at at least one of the time points.
-
-genes_to_keep = row.names(lfc_max[rowSums(lfc_max > 2) > 0, ])
-pvalues = pvalues[genes_to_keep, ]
-
 # Then rank by fisher's p-value and take max the number of genes of interest
@@
 fishers_qval = stats::p.adjust(fishers_pval)
-genes_to_keep = names(sort(fishers_pval))
-if(length(genes_to_keep) > n_genes_to_keep){
-    genes_to_keep = genes_to_keep[1:n_genes_to_keep]
-}
 
+genes_to_keep = row.names(
+    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
 y = as.matrix(data[genes_to_keep, ])
@@
 
-After filtering, we consider `r dim(y)[1]` genes for the clustering.
+After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
+clustering. As observed by looking at genes found differentially expressed, 
+many genes share a similar gene expression pattern, but on different scale.
+We thus propose the following adaptation of k-means:
+
+- **Splines estimation**: for each gene, fit the splines function with the basis
+  of your choice.
+- **Rescaling splines**: for each gene, rescale the estimated splines function
+  such that the values are bounded between 0 and 1.
+- **K-means**: apply k-means on the rescaled fitted splines to estimate the
+  centroids.
+- **Assign scores and labels to all genes**: then assign a score and a label
+  to all gene based on a goodness-of-fit measure on the raw data. By default,
+  the `splines_kmeans_score_and_label` function only labels the best 50% of
+  genes.
+
+The first three steps
 
@@
 kmeans_clusters = moanin:::splines_kmeans(
-    y, splines_model, n_clusters=20,
+    y, splines_model, n_clusters=8,
     random_seed=42,
     n_init=20)
+```
 
+We then use the `plot_centroids` function to visualize the centroids obtained
+with the splines k-means model.
+
+```{r fig.height=6, fig.width=6}
+moanin:::plot_centroids(kmeans_clusters$centroids, splines_model,
+			colors=ann_colors$Group,
+			smooth=TRUE)
+```
+
+The scoring and labeling can be done via the `splines_kmeans_score_and_label`
+function.
+
+```{r}
 # Then assign scores and labels to all the data, using a goodness-of-fit
@@
 labels = scores_and_labels$labels
+labels = unlist(labels[!is.na(labels)])
 ```
 
-Summary of number of genes assigned with original scoring, and "better"
-scoring.
+Before performing the next steps, let us investigate in more detail the
+differences between the labels provided by the splines k-means model and the
+scoring and labelinig step.
 
-```{r}
-barplot(table(kmeans_clusters$clusters), col="black")
+```{r echo=FALSE, fig.height=6, fig.width=6}
+par(mfrow=c(2, 1))
+barplot(table(kmeans_clusters$clusters), col="black", main="splines k-means
+labeling")
+barplot(table(labels), col="black", main="score and label")
 ```
 
-```{r}
-barplot(table(labels), col="black")
-```
+The scoring and label step allows to label genes that were removed during the
+filtering step, yet are good matches to the centroids found during the
+clustering.
 
-Just a hack, cause I didn't update the code
-```{r}
-kmeans_clusters$clusters = labels[!is.na(labels)]
+```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
+labels = unlist(labels[!is.na(labels)])
+genes_scored = names(labels)
+
+kmeans_labels = kmeans_clusters$clusters
+kmeans_labels = kmeans_labels[genes_scored]
+
+labels = labels[names(kmeans_labels)]
+
+both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
+# Create another cluster.
+#both_labels[is.na(both_labels)] = 21
+confusion_matrix = table(both_labels)
+text = as.matrix(as.character(confusion_matrix))
+dim(text) = dim(confusion_matrix)
+NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
+	      border=list(cell=TRUE), txt=text, legend=FALSE, Rowv=FALSE,
+	      Colv=FALSE)
+labels = scores_and_labels$labels
+labels = unlist(labels[!is.na(labels)])
+
 ```
@@
 
-Now let's visualize the centroids.
+## Looking at specific clusters in details.
 
-```{r fig.height=8, fig.width=8}
-moanin:::plot_centroids(kmeans_clusters$centroids, splines_model,
-			colors=ann_colors$Group,
-			smooth=TRUE)
+Now, let us look more in detail some specific clusters. Cluster 4 seems
+particularly interesting: it captures genes with strong differences between
+the differant influenza treatments and the control, while the control remains
+relatively flat.
+
+Heatmaps are useful to investigate the range of expression patterns for
+specific genes. Here, we are going to plot heatmaps of the normalized gene
+expression patterns and the rescaled gene expression patterns side by side.
+
+First, select the genes of interest.
+
+```{r}
+cluster_to_plot = 4
+genes_to_plot =  names(labels[labels == cluster_to_plot])
 ```
 
-Now, let's look at genes in cluster 15
+```{r fig.width=6, fig.height=8}
+layout(matrix(c(1,2),nrow=1), widths=c(1, 2))
 
-```{r fig.width=8, fig.height=8}
-cluster_to_plot = 20
-gene_names = row.names(data)
-genes_to_plot =  names(kmeans_clusters$clusters[kmeans_clusters$clusters == cluster_to_plot])
 data_to_plot = data[genes_to_plot, ]
@@
 
-NMF::aheatmap(
+res = NMF::aheatmap(
     data_to_plot,
     Colv=NA,
+    color="YlGnBu",
     annCol=submeta[,
@@
     annColors=ann_colors,
-    main="Cluster 15",
-    treeheight=0)
+    main=paste("Cluster", cluster_to_plot, "(raw)"),
+    treeheight=0, legend=FALSE)
 
+# Now use the results of the previous call to aheatmap to reorder the genes.
 NMF::aheatmap(
-    moanin:::rescale_values(data_to_plot),
+    moanin:::rescale_values(data_to_plot)[res$rowInd,],
     Colv=NA,
+    Rowv=NA,
     annCol=submeta[,
                 c("Group", "Timepoint")],
-    annLegend=FALSE,
+    annLegend=TRUE,
     annColors=ann_colors,
-    main=paste("Cluster", cluster_to_plot),
+    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
     treeheight=0)
@@
 ```
+
+Those two heatmaps demonstrate that the clustering method successfully cluster
+genes that are on different scales, and yet share the same dynamical response
+to the treatments.
 
@@
 ```{r running_clustering_on_all_k}
-all_possible_n_clusters = c(5, 7, 9, 11, 13, 15)
+all_possible_n_clusters = c(5, 9, 11, 15)
 all_clustering = list()
@@
 Using the subsampling strategy, keeping 80% of the genes:
+
 ```{r}
@@
 
-```{r fig.width=8, fig.height=8}
+```{r fig.width=6, fig.height=6}
 moanin::plot_model_explorer(all_labels)
@@
 From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
-but not as stable as $k=2$. Similarly, $k=14$ is more stable than $k=11$,
-$k=12$, $k=13$. The model explorer strategy, in addition to visualizing the
-diversity of the centroids, can thus help assessing an adequate number of
-clusters.
+but not as stable as $k=2$. The model explorer strategy, in addition to
+visualizing the diversity of the centroids, can thus help assessing an
+adequate number of clusters.
+
+Now, replot the same model explorer, but only for the clustering experiments
+$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
+measures in that range.
+
+```{r fig.width=6, fig.height=6}
+clusters = c("C6", "C7", "C8", "C9", "C10")
+selected_labels = all_labels[clusters]
+moanin::plot_model_explorer(selected_labels)
+```
 

### 2019-05-08 scripts/05_conclusion.Rmd
Switch bootstrapping proportion
More and better text. also better k for clustering

@@
 ---
+
+# Conclusion
+
+This workflow provides a tutoral for the analysis of time-course gene
+expression data in R, illustrated through the analysis of mice lung tissue
+exposed to different influenza strain. It covers four main steps; (1) quality
+control and normalization; (2) differential expression analysis; (3)
+clustering of time-course gene expression data; (4) downstream analysis of
+clusters.
+
+# Software and data availability
+
+The source code for this workflow can be found at **FIXME** 
+Data used in this workflow are available from NCBI GEO, accession GSE95601.
 

### 2019-05-08 bin/moanin/R/visualization.R

@@ -39,7 +39,7 @@ plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
     for(i in 1:n_centroids){
         plot_centroid_individual(as.vector(centroids[i, ]),
 				 splines_model, colors=colors,
-				 smooth=smooth)
+				 smooth=smooth, title=NULL)
     }
 }
 
@@ -59,7 +59,8 @@ plot_genes = function(data, splines_model, colors=NULL, smooth=FALSE){
 }
 
 
-plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth=FALSE){
+plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth=FALSE,
+				    title=NULL){
     meta = splines_model$meta
     groups = levels(meta$Group)
 
@@ -69,7 +70,7 @@ plot_centroid_individual = function(centroid, splines_model, colors=NULL, smooth
         centroid = t(as.matrix(centroid))
     }
 
-    graphics::plot(xrange, yrange, type="n")
+    graphics::plot(xrange, yrange, type="n", main=title)
     if(is.null(colors)){
         colors = viridis::viridis(length(groups))
 	names(colors) = groups

### 2019-05-08 scripts/run_clustering.R

@@ -39,33 +39,13 @@ pval_col_to_keep = colnames(de_analysis)[
 pvalues = de_analysis[, pval_col_to_keep]
 
 splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
+fishers_pval = moanin:::fisher_method(pvalues)
+fishers_qval = stats::p.adjust(fishers_pval)
 
-# FIXME refactor this into a function
-if(filter_genes){
-    # First, filter out any genes that doesn't have a log fold change of at
-    # least 2 at at least one of the time points.
-    
-    genes_to_keep = row.names(lfc_max[rowSums(lfc_max > 2) > 0, ])
-    pvalues = pvalues[genes_to_keep, ]
+genes_to_keep = row.names(
+    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
+y = as.matrix(data[genes_to_keep, ])
 
-    if(length(genes_to_keep) > n_genes_to_keep){
-
-	# Then rank by fisher's p-value and take max the number of genes of
-	# interest
-	# Filter out q-values for the pvalues table
-	fishers_pval = moanin:::fisher_method(pvalues)
-	fishers_qval = stats::p.adjust(fishers_pval)
-        if(keep_all_significant){
-	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
-	}else{
-	    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
-	}
-    }
-
-    y = as.matrix(data[genes_to_keep, ])
-}else{
-    y = as.matrix(data)
-}
 
 ###############################################################################
 # Now, randomly subsample.

### 2019-05-08 scripts/03_clustering.Rmd
More and better texts

@@
 NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
-	      border=list(cell=TRUE), txt=text, legend=FALSE, Rowv=FALSE,
-	      Colv=FALSE)
+	      border=list(cell=TRUE), txt=text, legend=FALSE, Rowv=NA,
+	      Colv=NA)
 labels = scores_and_labels$labels
@@
 
-Now, let us look more in detail some specific clusters. Cluster 4 seems
+Now, let us look more in detail some specific clusters. Cluster 8 seems
 particularly interesting: it captures genes with strong differences between
@@
 ```{r}
-cluster_to_plot = 4
+cluster_to_plot = 8
 genes_to_plot =  names(labels[labels == cluster_to_plot])
@@
 ```{r fig.width=6, fig.height=8}
-layout(matrix(c(1,2),nrow=1), widths=c(1, 2))
+layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
 
@@
 data, where the number of clusters is not clearly defined.
+
+# References

### 2019-05-08 bin/moanin/R/visualization.R

@@ -36,10 +36,16 @@ plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
             mar=c(0.2, 0.2, 0.2, 0.2))
     }
 
+    name_centroids = names(centroids)
+    name_centroid = NULL
     for(i in 1:n_centroids){
+	if(!is.null(name_centroids)){
+	    name_centroid = name_centroids[i]
+	}
         plot_centroid_individual(as.vector(centroids[i, ]),
 				 splines_model, colors=colors,
-				 smooth=smooth, title=NULL)
+				 smooth=smooth,
+				 title=name_centroid)
     }
 }
 

### 2019-05-08 scripts/01_quality_control_normalization.Rmd
Separating references from the rest of the QC
More citations, more text

@@
 
-```{r message=FALSE}
+```{r message=FALSE, results="hide", warning=FALSE}
 targets = limma::readTargets()
@@
 
-```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE}
+```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
 # Perform background correction using norm-exponential strategy and normalize
@@
 par(mfrow=c(1, 2))
-limma::plotDensities(RG, col="black", legend=FALSE)
-limma::plotDensities(RG.q, col="black", legend=FALSE)
+limma::plotDensities(RG, col="black", legend=FALSE,
+		     main="Before normalization")
+limma::plotDensities(RG.q, col="black", legend=FALSE,
+		     main="After normalization")
 ```
@@
 meta = shoemaker2015$meta
+data = shoemaker2015$data
 ```
@@
 most different from control samples.
+
+# References

### 2019-05-08 scripts/03_clustering.Rmd
More citations, more text

@@
 
-```{r echo=FALSE, fig.height=6, fig.width=6}
+```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
 par(mfrow=c(2, 1))
@@
 
+The Elbow method to choose the number of clusters relies on visualization aid
+to choose the number of cluster. The method relies on plotting the within
+cluster sum of squares as a function of the number of clusters. At some point,
+the WCSS will stop dropping, giving an angle in the graph.  The number of
+cluster is chosen at this "Elbow point."
+
 ```{r elbow_method, fig.width=6, fig.height=4}
-
-# function to compute total within-cluster sum of square
-
 plot(all_possible_n_clusters, wss_values,
@@
 ### Average silhouette method
+
+The silhouette value is a measure of how similar a data point is to its own
+cluster (cohesion) compared to other clusters (separation).
 

### 2019-05-08 scripts/04_downstream_analysis.Rmd
More and better text. also better k for clustering
More and better texts
More citations, more text

@@
 Let us first tackle the case of pathway enrichment analysis. We will leverage
-the packages `biomaRt` and `KEGGprofile` for this step. `KEGGprofile` is a
+the packages `biomaRt` [@durink:biomart] and `KEGGprofile` [@zhao:keggprofile]
+for this step. `KEGGprofile` is a
 package that easily allows to perform pathway enrichment analysis on a set of
@@
 ```{r enrichment_analysis, results="hide", message=FALSE}
-cluster = 20
+cluster = 8
 genes = gene_names[labels == cluster]
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways, "html", escape=F, digits=4) %>%
+kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
     kable_styling("striped", full_width=TRUE) 
@@
 # Create gene to GO id mapping
-gene_id_go_mapping = create_go_term_mapping(genes)
+gene_id_go_mapping = moanin::create_go_term_mapping(genes)
 ```
@@
 
-kable(go_terms_enriched, "html", escape=F, digits=4) %>%
+kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
     kable_styling("striped", full_width=TRUE) 
 ```
+
+# References

### 2019-05-08 bin/moanin/R/visualization.R

@@ -17,23 +17,23 @@ plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
     if(n_centroids <= 3){
         graphics::par(
 	    mfrow=c(n_centroids, 1),
-            mar=c(0.2, 0.2, 0.2, 0.2))
+            mar=c(1.0, 1.0, 1.0, 1.0))
     }else if(n_centroids <= 6){
         n_col = ceiling(n_centroids / 2)
         graphics::par(
 	    mfrow=c(n_col, 2),
-            mar=c(0.2, 0.2, 0.2, 0.2))
+            mar=c(1.0, 1.0, 1.0, 1.0))
     }else if(n_centroids <= 12){
         ncol = ceiling(n_centroids / 3)
         graphics::par(
 	    mfrow=c(ncol, 3),
-            mar=c(0.2, 0.2, 0.2, 0.2))
+            mar=c(1.0, 1.0, 1.0, 1.0))
     }else{
         nrow = round(n_centroids ** 0.5)
         ncol = ceiling(n_centroids / nrow)
         graphics::par(
 	    mfrow=c(ncol, nrow),
-            mar=c(0.2, 0.2, 0.2, 0.2))
+            mar=c(1.0, 1.0, 1.0, 1.0))
     }
 
     name_centroids = names(centroids)

### 2019-05-08 scripts/02_differential_expression_analysis.Rmd
Switched to only using public API function

@@
 
-```{r echo=FALSE, results="hide"}
+```{r echo=FALSE, results="hide", message=FALSE, warning=FALSE}
 # Reloading necessary data and preliminary results

### 2019-05-08 scripts/03_clustering.Rmd
Switched to only using public API function

@@
 fold change of at least two between at least one condition and one time-point.
+Reducing the set of genes on which to perform the clustering allows to
+estimate the centroids with more stability.
 
@@
 # Filter out q-values for the pvalues table
-fishers_pval = moanin:::fisher_method(pvalues)
+fishers_pval = moanin::pvalues_fisher_method(pvalues)
 fishers_qval = stats::p.adjust(fishers_pval)
@@
 
-The first three steps
+The first three steps are performed jointly by the `splines_kmeans` function.
 
@@
 # First fit the kmeans clusters
-kmeans_clusters = moanin:::splines_kmeans(
+kmeans_clusters = moanin::splines_kmeans(
     y, splines_model, n_clusters=8,
@@
 ```{r fig.height=6, fig.width=6}
-moanin:::plot_centroids(kmeans_clusters$centroids, splines_model,
+moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
 		       colors=ann_colors$Group,
@@
 labels = scores_and_labels$labels
+
+# Let's keep only the list of genes that have a label.
 labels = unlist(labels[!is.na(labels)])
@@
 par(mfrow=c(2, 1))
-barplot(table(kmeans_clusters$clusters), col="black", main="splines k-means
-labeling")
+barplot(table(kmeans_clusters$clusters),
+	col="black",
+	main="splines k-means labeling")
 barplot(table(labels), col="black", main="score and label")
@@
 NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
-	      border=list(cell=TRUE), txt=text, legend=FALSE, Rowv=NA,
+	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
 	      Colv=NA)
@@
 for(n_cluster in all_possible_n_clusters){
-    clustering_results = moanin:::splines_kmeans(
+    clustering_results = moanin::splines_kmeans(
 	y, splines_model,

### 2019-05-08 scripts/04_downstream_analysis.Rmd
Switched to only using public API function

@@
 Let us first tackle the case of pathway enrichment analysis. We will leverage
-the packages `biomaRt` [@durink:biomart] and `KEGGprofile` [@zhao:keggprofile]
+the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
 for this step. `KEGGprofile` is a
@@
 To find GO terms, we use `biomaRt` to find the mapping between GO terms and
-gene mapping. The GO enrichment library `topGO` expects the GO term to gene
+gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
 mapping to be a list where each item is a mapping between a gene name and a GO

### 2019-05-08 scripts/00_introduction.Rmd
Better fig1
Preparing for switch to pdf format

@@
 title: "Introduction: a pipeline to analyse time-course gene expression data"
+author:
+    - name: Nelle Varoquaux
+      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA
+    - name: Elizabeth Purdom
+      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
+keywords: >
+    time-course gene expression data, clustering, time-course
+    differential expression analysis, workflow
+vignette: >
+    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
+    %\VignetteEngine{knitr::rmarkdown}
 output: rmarkdown::html_vignette
+
 bibliography: bibliography.bib
@@
   the same thing. Find a better name for both of them
-- On the schema plot, replace clustering schemas with something else.
 
@@
 ```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
+library(BiocStyle)
 library(knitr)

### 2019-05-08 scripts/install.R

@@ -90,3 +90,8 @@ if(!require("NMI")){
 if(!require(zoo)){
     install.packages("zoo")
 }
+
+if(!require(BiocWorkflowTools)){
+    biocLite("BiocWorkflowTools")
+}
+

### 2019-05-08 scripts/01_quality_control_normalization.Rmd
Just saving before deleting files.

@@
 library(stats)
+library(knitr)
+library(kableExtra)
+
+library(ggplot2)
+library(NMF)
+library(KEGGprofile)
+library(viridis)
+
 library(moanin)
-library(ggplot2)
-library(stats)
 
 source("utils.R")
-
-# library(moanin)
-```
-
-```{r echo=FALSE}
-# Set options
-df = 6
-
-# Preprocessing & filtering
-is_count = FALSE
-take_log = FALSE    
-filter_expression = FALSE
-
-# Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M")
-
-# Clustering
-filter_genes = TRUE
-n_genes_to_keep = 5000
-n_clusters = 20
-percentage_genes_to_label = 0.5
 ```
@@
 ```{r echo=FALSE, include=FALSE}
-library(knitr)
 knitr::opts_chunk$set(
@@
 )
+```
 
+# TODO list
+
+- Add legends to the plots that don't have.
+- Currently have two functions "plot_genes", "plot_centroids", that do exactly
+  the same thing. Find a better name for both of them
+
+# Abstract
+
+The phenotypic diversity of cells is governed by a complex equilibrium between
+their genetic identity and their environmental interactions: Understanding the
+dynamics of gene expression is a fundamental question of biology. However,
+analysing time-course transcriptomic data raises unique challenging
+statistical and computational questions, requiring the development of novel
+methods and software.  Using as case study time-course transcriptomics data
+from mice exposed to different strains of influenza, ihis workflows provides a
+step-by-step tutorial of the methodology used to analyse time-course data:
+(1) normalization of the micro-array dataset; (2) differential expression
+analysis using functional data analysis; (3) clustering fo time-course data;
+(4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
+
+# Introduction
+
+Gene expression studies provide simultaneous quantification of the level of
+mRNA from all genes in a sample. High-throughput studies of gene expression
+have a long history, starting with microarray technologies in the 1990s
+through to single-cell technologies. While many expression studies are
+designed to compare the gene expression in distinct groups, there is also a
+long history of time-course expression studies, where the the gene expression
+is compared across time by measuring mRNA levels from different samples across
+time [^1]. Such time course studies can vary from measuring a few distinct time
+points, to sampling ten to twenty time points. Many longer time series are
+particularly interested in investigating development over time. More recently,
+single-cell studies track single cells through their development, and a single
+cell is measured at a particular moment in its developmental progression -- a
+value that is not know but estimated from the data as its "pseudo-time."
+
+[^1]: Because the collection of the mRNA is often destructive, samples at
+different time points are generally from different biological samples;
+longitudinal studies, for example tracking the same subject over time, are
+certainly possible in certain settings, but not directly considered here.
+
+While there are many methods that have been proposed for discrete aspects of
+time course data, the entire workflow for analysis of such data remains
+difficult, particularly for long, developmental time series. Most methods
+proposed for time course data are concerned with detecting genes that are
+changing over time (differential expression analysis), examples being `edge`
+[@storey:significance], functional component analysis based models [@wu:more],
+time-course permutation tests [@park:statistical], and multiple testing
+strategies to combine singe time point differential expression analysis
+[@sun:multiple]. However, with long time course datasets, particularly in
+developmental systems, a massive number of genes will show *some* change. For
+example, in the **xxx** system, over **xxx%** of genes are shown to be changing
+over time. The task in these settings is often not to detect changes in genes,
+but to categorize into biologically interpretable patterns the vast number of
+changes discovered.
+
+We present here a workflow for such an analysis that consists of 4 main
+parts (Figure \ref{fig:workflow}):
+
+- Quality control and normalization;
+- Identification of genes that are differentially expressed;
+- Clustering of genes into distinct temporal patterns;
+- Biological interpretation of the clusters.
+
+This workflow represents an integration of both novel implementations of
+previously established methods and new methodologies for the settings of
+developmental time series. It relies on several standard packages for
+analysing gene expression data, some specific for time-course data, others
+broadly used by the community. We provide the various steps of the workflow as
+functions in a R package **FIXME**.
+
+```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
+library(BiocStyle)
+library(knitr)
+knitr::include_graphics("images/fig1_schemas.png")
 ```
+
+# Analysis of the dynamical response of mouse lung tissue to influenza
+
+## Overview
+
+This workflow is illustrated using data from a micro-array time-course
+experiment, exposing mice to three different strains of influenza, and
+collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
+18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
+[@shoemaker:ultrasensitive]. The three strains of influenza used in the
+study are (1) a low pathogenicity seasonal H1N1 influenza virus
+(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
+pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
+avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
+10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
+of the Vietnam avian influenza virus (10^3 PFU).
+
+By combining gene expression time-course data with virus growth data, the
+authors show that the inflammatory response of lung tissue is gated until a
+threshold of the virus concentration is  exceeded in the lung. Once this
+threshold is exceeded, a strong inflammatory and cytokin production occurs.
+This results provides evidence that the pathology response is non-linearly
+regulated by virus concentration.
+
+## References
 
@@
 
-```{r echo=FALSE, results="hide"}
-write.table(data, ".results/quantile_normalized.txt")
-```
-
 ## Exploratory analysis and quality control
@@
 min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
-data = data[variance_per_genes > min_variance,]
+variance_filtered_data = data[variance_per_genes > min_variance,]
 ```
@@
 
-pca_data = stats::prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
+pca_data = stats::prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
 percent_var = round(100 * attr(pca_data, "percentVar"))
@@
 ```{r  correlation_plot, fig.width=6, fig.height=6}
-library(NMF)
 
@@
 
-data_corr = stats::cor(data, method="pearson")
+data_corr = stats::cor(variance_filtered_data, method="pearson")
 
@@
 
+## Differential expression analysis of time-course data
+
+The next step is typically to run a differential expression analysis. Most of
+`moanin`'s function rely on an object containing a number of information such
+as the metadata and the splines model used.
+
+```{r}
+splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
+```
+
+### Weekly differential expression analysis
+
+`moanin` provides a simple interface to perform a time by time differential
+expression analysis. Under the hood, it simply calls `limma` on the set of
+contrasts provided. Here, we show an example between the control mouse and the
+mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
+(H5N1), but the function works with any form contrasts.
+
+```{r weekly_differential_analysis}
+meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Timepoint))
+
+# Define contrast	
+contrasts = c("M.0-VL.0",
+	      "M.3-VL.3",
+	      "M.6-VL.6",
+	      "M.9-VL.9",
+	      "M.12-VL.12",
+	      "M.18-VL.18",
+	      "M.24-VL.24",
+	      "M.30-VL.30",
+	      "M.36-VL.36",
+	      "M.48-VL.48",
+	      "M.60-VL.60",
+	      "M.72-VL.72",
+	      "M.120-VL.120",
+	      "M.168-VL.168")
+weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+     data, splines_model, contrasts,
+     use_voom_weights=FALSE)
+```
+
+```{r echo=FALSE, results="hide"}
+# Hide those results
+# Run the DE analysis on the rest of the influenza strain.
+contrasts = c("M.0-K.0",
+	      "M.3-K.3",
+	      "M.6-K.6",
+	      "M.9-K.9",
+	      "M.12-K.12",
+	      "M.18-K.18",
+	      "M.24-K.24",
+	      "M.30-K.30",
+	      "M.36-K.36",
+	      "M.48-K.48",
+	      "M.60-K.60",
+	      "M.72-K.72",
+	      "M.120-K.120",
+	      "M.168-K.168")
+K_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+     data, splines_model, contrasts,
+     use_voom_weights=FALSE)
+contrasts = c("M.0-C.0",
+	      "M.3-C.3",
+	      "M.6-C.6",
+	      "M.9-C.9",
+	      "M.12-C.12",
+	      "M.18-C.18",
+	      "M.24-C.24",
+	      "M.30-C.30",
+	      "M.36-K.36",
+	      "M.48-C.48",
+	      "M.60-C.60",
+	      "M.72-C.72",
+	      "M.120-C.120",
+	      "M.168-C.168")
+C_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+     data, splines_model, contrasts,
+     use_voom_weights=FALSE)
+contrasts = c("M.0-VH.0",
+	      "M.3-VH.3",
+	      "M.6-VH.6",
+	      "M.9-VH.9",
+	      "M.12-VH.12",
+	      "M.18-VH.18",
+	      "M.24-VH.24",
+	      "M.30-VH.30",
+	      "M.36-VH.36",
+	      "M.48-VH.48",
+	      "M.60-VH.60",
+	      "M.72-VH.72",
+	      "M.120-VH.120",
+	      "M.168-VH.168")
+VH_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+     data, splines_model, contrasts,
+     use_voom_weights=FALSE)
+
+```
+
+Let's look at the distribution of genes found differentially expressed per
+week between control and each of the influenza strains.
+
+```{r fig.width=8, fig.height=8, echo=FALSE}
+par(mfrow=c(2, 2))
+
+time = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
+
+timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
+    qval_colnames = colnames(de_analysis)[
+	grepl("qval", colnames(de_analysis))]
+    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
+    names(number_de_genes_per_time) = labels
+    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
+	    xlab="Timepoint", ylab="Number of DE genes", main=title)
+
+}
+
+timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
+timely_barplot(C_weekly_de_analysis, time, title="California")
+timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose")
+timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose")
+```
+
+The distribution of number of genes found differentially expressed by
+considering each time-point independantly highlights the challenges of such
+approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
+genes found differentially expressed than timepoints 3H and 9H: this is likely
+due to biological or technical variances for specific genes at specific
+timepoints.
+
+As a summary, classic differential expression methods are appropriate for
+unordered treatments, but fail to use the temporal structure of the data.
+
+### Time-course differential expression analysis between two groups
+
+To leverage this temporal structure, Storey et al [@storey:significance] proposes to
+model time-course micro-array with splines functions, and to use a log-ratio
+likelihood test to detect differentially expressed genes. `moanin` extends
+this idea by providing functions to perform time-course differential
+expression analysis using `limma` contrasts.
+
+```{r de_analysis}
+# Differential expression analysis
+timecourse_contrasts = c("M-K", "M-C", "M-VL", "M-VH")
+
+contrasts = limma::makeContrasts(
+	contrasts=timecourse_contrasts,
+	levels=levels(meta$Group))
+
+# The function takes the data (data.frame or named matrix), the meta data
+# (data.frame containing a timepoint and group column, the first corresponding
+# to the time-course information, the latter corresponding to the
+# treatment).
+pvalues = moanin::timecourse_differential_expression_analysis(
+    data, splines_model, contrasts)
+qvalues = apply(pvalues, 2, stats::p.adjust)
+
+```
+
+The number of genes found differentially expressed ranges from around 12000 to
+29000 depending on the strain and dosage of influenza virus given to the mice.
+This corresponds to between 30% to 70% of the genes found differentially
+expressed in this time-course experiment.
+
+```{r echo=FALSE, fig.width=6, fig.height=6}
+num_de = colSums(qvalues < 0.05)
+names(num_de) = c(
+    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
+barplot(num_de, col="black", ylab="Number of DE genes")
+```
+
+The next step in a classical differential expression analysis is typically to
+assess the effect of the treatment by looking at the log fold change.
+Computing the log fold change on a time-course experiment is not trivial: one
+can be interested in the average log-fold change across time, or the
+cumulative log-fold change. Sometimes a gene can be over-expressed at the
+beginning of the time-course data, and then over-expressed at the end of the
+experiment. As a result, `moanin` provides a number of possible ways to
+compute the log fold change across the whole time-course.
+
+First, `moanin` provides as simple interface to compute the log-fold change
+for each individual timepoints.
+
+```{r}
+log_fold_change_timepoints = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts,  method="timely")
+```
+
+```{r echo=FALSE}
+kable(head(log_fold_change_timepoints[,
+colnames(log_fold_change_timepoints)[1:5]]))
+```
+
+This matrix can then be used to visualize the log-fold change for each
+contrast per timepoint.
+
+Sometimes, a single value per gene and per contrast is more useful. Here is a
+table of the possible ways to compute log-fold change values.
+
+| Name   | Formula   |     |
+|------- | --------- |--- |
+| timely | $\text{lfc}(t)$  | Function of time |
+| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
+| abs_sum  | $\sum_t \|lfc(t)\|$  | Always positive |
+| max  | $\max_t \|lfc(t)\|$  | Always positive  |
+| min  | $\min_t \|lfc(t)\|$  | Always positive |
+| epicon  | $$  | Captures overall strength of response and overall direction |
+
+```{r}
+log_fold_change_epicon = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts,  method="epicon")
+
+log_fold_change_sum = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts,  method="sum")
+
+log_fold_change_max = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts, method="max")
+
+log_fold_change_min = moanin::estimate_log_fold_change(
+    data, splines_model, contrasts, method="min")
+```
+
+```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE}
+par(mfrow=c(2, 2))
+plot(log_fold_change_epicon[, "M-C"],
+     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
+     xlab="EPICON", ylab="sum")
+plot(log_fold_change_epicon[, "M-C"],
+     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
+     xlab="EPICON", ylab="max")
+plot(log_fold_change_max[, "M-C"],
+     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
+     xlab="max", ylab="min")
+plot(log_fold_change_max[, "M-C"],
+     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
+     xlab="max", ylab="sum")
+```
+
+From the single measures of log-fold change and the p-value, we can now look
+at the volcano plot. Here is an example of a volcano plot for the comparaison
+of the control to the Kawasaki strain, using the EPICON log fold change
+computation.
+
+```{r fig.height=8, fig.width=8}
+pvalue = qvalues[, "M-K"]
+lfc_epicon = log_fold_change_epicon[, "M-K"]
+names(lfc_epicon) = row.names(log_fold_change_epicon)
+
+plot(lfc_epicon, -log10(pvalue), pch=20, main="Volcano plot",
+     xlim=c(-2.5, 2))
+```
+
+As another sanity check, `moanin` provides a simple utility function to
+visualize gene time-course data. Here, we plot the 12 genes with the smallest
+p-values.
+
+```{r fig.height=8, fig.width=8}
+top_DE_genes_pval = names(sort(pvalue)[1:12])
+moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
+		   colors=ann_colors$Group, smooth=TRUE)
+```
+
+And here we visualize the genes with the largest absolute EPICON log-fold
+change.
+
+```{r fig.height=8, fig.width=8}
+top_DE_genes_lfc = names(
+    sort(abs(lfc_epicon),
+	 decreasing=TRUE)[1:12])
+moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
+		   colors=ann_colors$Group, smooth=TRUE)
+
+```
+
+Thanks to those visualization, we can see that genes often follow similar
+patterns of expression, although on a different scale for each gene. We can
+leverage this observation to cluster the genes into groups of similar patterns
+of transcriptomic response.
+
+# Clustering of time-course data
+
+The very large number of genes found differentially expressed impair any
+interpretation one would attempt: with 70\% of the genome found differentially
+expressed, all pathways are affected by the treatment. Hence the next step of
+the workflow to cluster gene expression according to their dynamical response
+to the treatment. Before clustering the genes, we first reduce the set of
+genes of interest to genes that are (1) significantly found differentially
+expressed; (2) "highly" differentially expressed. To do this, we first
+aggregate all p-values obtained during the time-course differential expresison
+step in a single p-value using Fisher's method [@fisher:statistical]. Then we
+select all the genes which have a Fisher adjusted p-value below 0.05 and a log
+fold change of at least two between at least one condition and one time-point.
+Reducing the set of genes on which to perform the clustering allows to
+estimate the centroids with more stability.
+
+```{r filter_genes}
+# Then rank by fisher's p-value and take max the number of genes of interest
+# Filter out q-values for the pvalues table
+fishers_pval = moanin::pvalues_fisher_method(pvalues)
+fishers_qval = stats::p.adjust(fishers_pval)
+
+genes_to_keep = row.names(
+    log_fold_change_max[
+	(rowSums(log_fold_change_max > 2) > 0) &
+	(fishers_qval < 0.05), ])
+# Keep the data corresponding to the genes of interest in another variable.
+y = as.matrix(data[genes_to_keep, ])
+```
+
+After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
+clustering. As observed by looking at genes found differentially expressed, 
+many genes share a similar gene expression pattern, but on different scale.
+We thus propose the following adaptation of k-means:
+
+- **Splines estimation**: for each gene, fit the splines function with the basis
+  of your choice.
+- **Rescaling splines**: for each gene, rescale the estimated splines function
+  such that the values are bounded between 0 and 1.
+- **K-means**: apply k-means on the rescaled fitted splines to estimate the
+  centroids.
+- **Assign scores and labels to all genes**: then assign a score and a label
+  to all gene based on a goodness-of-fit measure on the raw data. By default,
+  the `splines_kmeans_score_and_label` function only labels the best 50% of
+  genes.
+
+The first three steps are performed jointly by the `splines_kmeans` function.
+
+```{r clustering}
+# First fit the kmeans clusters
+kmeans_clusters = moanin::splines_kmeans(
+    y, splines_model, n_clusters=8,
+    random_seed=42,
+    n_init=20)
+```
+
+We then use the `plot_centroids` function to visualize the centroids obtained
+with the splines k-means model.
+
+```{r fig.height=6, fig.width=6}
+moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
+		       colors=ann_colors$Group,
+		       smooth=TRUE)
+```
+
+The scoring and labeling can be done via the `splines_kmeans_score_and_label`
+function.
+
+```{r}
+# Then assign scores and labels to all the data, using a goodness-of-fit
+# scoring function.
+scores_and_labels = moanin::splines_kmeans_score_and_label(
+    data, kmeans_clusters)
+labels = scores_and_labels$labels
+
+# Let's keep only the list of genes that have a label.
+labels = unlist(labels[!is.na(labels)])
+```
+
+Before performing the next steps, let us investigate in more detail the
+differences between the labels provided by the splines k-means model and the
+scoring and labelinig step.
+
+```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
+par(mfrow=c(2, 1))
+barplot(table(kmeans_clusters$clusters),
+	col="black",
+	main="splines k-means labeling")
+barplot(table(labels), col="black", main="score and label")
+```
+
+The scoring and label step allows to label genes that were removed during the
+filtering step, yet are good matches to the centroids found during the
+clustering.
+
+```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
+labels = unlist(labels[!is.na(labels)])
+genes_scored = names(labels)
+
+kmeans_labels = kmeans_clusters$clusters
+kmeans_labels = kmeans_labels[genes_scored]
+
+labels = labels[names(kmeans_labels)]
+
+both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
+# Create another cluster.
+#both_labels[is.na(both_labels)] = 21
+confusion_matrix = table(both_labels)
+text = as.matrix(as.character(confusion_matrix))
+dim(text) = dim(confusion_matrix)
+NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
+	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
+	      Colv=NA)
+labels = scores_and_labels$labels
+labels = unlist(labels[!is.na(labels)])
+
+```
+
+```{r results="hide", echo=FALSE}
+# Save the clustering for the GO term and pathway enrichment on the whole set
+# of clusters.
+kmeans_labels = as.data.frame(labels[!is.na(labels)])
+write.table(kmeans_labels,
+	    ".results/clustering_labels.txt",
+	    sep="\t")
+```
+
+## Looking at specific clusters in details.
+
+Now, let us look more in detail some specific clusters. Cluster 8 seems
+particularly interesting: it captures genes with strong differences between
+the differant influenza treatments and the control, while the control remains
+relatively flat.
+
+Heatmaps are useful to investigate the range of expression patterns for
+specific genes. Here, we are going to plot heatmaps of the normalized gene
+expression patterns and the rescaled gene expression patterns side by side.
+
+First, select the genes of interest.
+
+```{r}
+cluster_to_plot = 8
+genes_to_plot =  names(labels[labels == cluster_to_plot])
+```
+
+```{r fig.width=6, fig.height=8}
+layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
+
+data_to_plot = data[genes_to_plot, ]
+submeta = meta
+ord = order(
+  submeta$Group,
+  submeta$Timepoint,
+  submeta$Replicate)
+submeta$Timepoint = as.factor(submeta$Timepoint)
+
+data_to_plot = data_to_plot[, ord]
+submeta = submeta[ord, ]
+
+res = NMF::aheatmap(
+    data_to_plot,
+    Colv=NA,
+    color="YlGnBu",
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=FALSE,
+    annColors=ann_colors,
+    main=paste("Cluster", cluster_to_plot, "(raw)"),
+    treeheight=0, legend=FALSE)
+
+# Now use the results of the previous call to aheatmap to reorder the genes.
+NMF::aheatmap(
+    moanin:::rescale_values(data_to_plot)[res$rowInd,],
+    Colv=NA,
+    Rowv=NA,
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=TRUE,
+    annColors=ann_colors,
+    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
+    treeheight=0)
+
+```
+
+Those two heatmaps demonstrate that the clustering method successfully cluster
+genes that are on different scales, and yet share the same dynamical response
+to the treatments.
+
+## How to choose the number of clusters.
+
+A common question that arises when performing clustering is how to choose the
+number of clusters. A choice for the number of clusters K depends on the goal. 
+In this particular case, the end goal is not the clustering, but to facilitate
+interpretation of the differential expression analysis step. As a result, the
+number of clusters should not exceed the number of gene sets the user wants to
+interpret. This allows to set a maximum number of clusters. Let us assume here
+that this number is 20 clusters.
+
+Once the maximum number of clusters is set, several strategies allow to
+identify the number of clusters:
+
+- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
+  the elbow methods looks at the total with cluster sum of squares 
+  as a function of the number of clusters (WCSS). When adding clusters
+  doesn't decrease the WCSS by a sufficient
+  amount, one can consider stopping. This method thus provides visual aid to
+  the user to choose the number of clusters, but often the "elbow" is hard to
+  see on real data, where the number of clusters is not clearly defined.
+- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
+  refers to a method of validation of consistency within clusters, and
+  provides visual aid to choose the number of clusters.
+- **Stability methods** Stability methods are more computationally intensive
+  than any other method, as they rely on assessing the stability of the
+  clustering for every $k$ to a small randomization of the data. The user is
+  then invited to choose the number of cluster based on a number of similarity
+  measures.
+
+First, let us run the clustering for all possible clusters of interest. We
+will, for each clustering experiment, conserve (1) with within cluster sum of
+squares; (2) the labels assigned to all genes.
+
+```{r running_clustering_on_all_k}
+all_possible_n_clusters = c(5, 9, 11, 15)
+all_clustering = list()
+wss_values = list()
+
+i = 1
+for(n_cluster in all_possible_n_clusters){
+    clustering_results = moanin::splines_kmeans(
+	y, splines_model,
+	n_clusters=n_cluster, random_seed=42,
+	n_init=10)
+    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
+    all_clustering[[i]] = clustering_results$clusters
+    i = i + 1
+}
+```
+
+### Elbow method
+
+The Elbow method to choose the number of clusters relies on visualization aid
+to choose the number of cluster. The method relies on plotting the within
+cluster sum of squares as a function of the number of clusters. At some point,
+the WCSS will stop dropping, giving an angle in the graph.  The number of
+cluster is chosen at this "Elbow point."
+
+```{r elbow_method, fig.width=6, fig.height=4}
+plot(all_possible_n_clusters, wss_values,
+     type="b", pch=19, frame=FALSE, 
+     xlab="Number of clusters K",
+     ylab="Total within-clusters sum of squares")
+
+```
+
+### Average silhouette method
+
+The silhouette value is a measure of how similar a data point is to its own
+cluster (cohesion) compared to other clusters (separation).
+
+```{r average_silhouette_method, fig.width=6, fig.height=4}
+# function to compute average silhouette for k clusters
+average_silhouette = function(labels, y) {
+    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
+    return(mean(silhouette_results[, 3]))
+}
+
+# extract the average silhouette
+average_silhouette_values = list()
+i = 1
+for(i in 1:length(all_clustering)){
+    labels = all_clustering[i]
+    average_silhouette_values[i] = average_silhouette(labels, y)
+    i = i + 1
+}
+
+plot(all_possible_n_clusters, average_silhouette_values,
+     type="b", pch=19, frame=FALSE,
+     xlab="Number of clusters K",
+     ylab="Average Silhouettes")
+```
+
+### Looking at the stability of the clustering
+
+On real data, the number of clusters is not only unknown but also 
+ambiguous: it will depend on the desired clustering resolution of the user.
+Yet, in the case of biological data, stability and reproducibility of the
+results is necessary to ensure that the biological interpretation of the
+results hold when the data or the model is exposed to reasonable
+perturbations.
+
+Methods that rely on the stability of the clustering results to choose $k$
+thus ensure that the biological interpretation of the clusters hold with
+perturbtation to the data. In addition, simulation where the data is generated
+with a well defined $k$ show that the clustering is more stable for the
+correct of number of the clusters.
+
+Most methods method to find the number of clusters with stability measures
+only provide visual aids to guide the user. The first element often visualized
+is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
+stores the proportion of clustering in which two items are clustered together.
+A perfect consensus matrix ordered such as each elements that belong to the
+same cluster are adjacent to one another which show blocks along the diagonal
+close to 1.
+
+To perform such analysis, the first step is run the clustering several times
+on a resampled dataset--either using bootstrap or subsampling.
+
+Using the bootstrapping strategy:
+```{r}
+n_genes = dim(y)[1]
+indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
+
+bootstrapped_y = y[indices, ]
+```
+
+Using the subsampling strategy, keeping 80% of the genes:
+
+```{r}
+subsample_proportion = 1
+indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
+subsampled_y = y[indices, ]
+```
+
+Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
+
+```{r fig.width=6, fig.height=4}
+library(NMF)
+stability_5 = read.table("results/stability_5.tsv", sep="\t")
+consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
+							scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+stability_20 = read.table("results/stability_20.tsv", sep="\t")
+consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
+							 scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+```
+
+```{r echo=FALSE, results="hide"}
+n_clusters = 2:20
+all_labels = list()
+for(i in n_clusters){
+    filename = paste0("results/stability_", i, ".tsv")
+    stability = read.table(filename, sep="\t")
+    all_labels[[paste0("C", i)]] = stability
+}
+```
+
+#### The model explorer strategy
+
+The model explorer algorithm [@ben-hur:stability] proposes to estimate the
+number of clusters exploiting the observation that if the number of clusters
+is correct, the clustering results are stable to bootstrapping.  The
+distribution of similaries between bootstrapped results for each $k$ can thus
+be compared for different values of $k$ and guide the user in the choice of
+number of clusters.
+
+The model explorer strategy works as follows. First, choose a similarity
+measure between two partitions or clusters $S(C_1, C_2)$. Examples are the
+normalized mutual information or Fowlkes-Mallows. Then perform $n$ bootstrap
+experiments to estimate the cluster centroids, followed by a step of assigning
+a label to all data points. Finally, compute the pairwise similarity measure
+between all bootstrapped partition, and plot the cumulative density of the
+obtained scores.
+
+```{r fig.width=6, fig.height=6}
+moanin::plot_model_explorer(all_labels)
+```
+
+From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
+but not as stable as $k=2$. The model explorer strategy, in addition to
+visualizing the diversity of the centroids, can thus help assessing an
+adequate number of clusters.
+
+Now, replot the same model explorer, but only for the clustering experiments
+$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
+measures in that range.
+
+```{r fig.width=6, fig.height=6}
+clusters = c("C6", "C7", "C8", "C9", "C10")
+selected_labels = all_labels[clusters]
+moanin::plot_model_explorer(selected_labels)
+```
+
+#### Consensus clustering as a way to find $k$
+
+The  consensus clustering [@monti:consensus] relies on a similar idea but
+instead of looking at the cumulative density of similarity measures of
+bootstrapped clustering, the authors suggests plotting the cumulative density
+of elements of the consensus matrix.
+
+```{r eval=FALSE}
+moanin::plot_cdf_consensus(all_labels)
+```
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_CDF_consensus.png")
+```
+
+The stability of the clustering based on the consensus matrix can then be
+measured via a single number by looking at the area under the curve: the more
+stable the clustering, the closer to 0 or 1 will be the entries of consensus
+matrix. The consensus clustering strategy thus suggest at looking at either
+the AUC as a function of the number of the clusters or the "improvement" in
+the AUC as a function of the number of cluster.
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_AUC_consensus.png")
+```
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_delta_AUC_consensus.png")
+```
+
+The consensus clustering method suggest that the most stable is $k=2$, which
+separates over-expressed genes from under-expressed genes. While it is indeed
+a very stable clustering, it does not capture the range of gene expression
+patterns present in the data. This shows the limitation of such method on real
+data, where the number of clusters is not clearly defined.
+
+# Downstream analysis of clusters.
+
+Once good clusters are obtained, the next step is to leverage the clustering
+to ease interpretation. Classic enrichment analysis step can then be performed
+in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
+term enrichment analysis, motif enrichment analysis, etc.
+
+First, let us clean up the gene obtained and only select the genes we are
+going to use in the enrichment analysis. One can either use the whole set of
+genes, only the set of differentially expressed genes in each cluster, or a
+subset of genes that fit well to a cluster (based on some criterion).
+
+```{r}
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = as.vector(labels[!is.na(labels)])
+names(labels) = gene_names
+```
+
+### Finding enriched pathways using `biomaRt` and `KEGGprofile`
+
+Let us first tackle the case of pathway enrichment analysis. We will leverage
+the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
+for this step. `KEGGprofile` is a
+package that easily allows to perform pathway enrichment analysis on a set of
+genes labeled with the ensembl annotation. We thus need to convert the gene
+names into the appropriate format. This is where `biomaRt` comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use
+it to convert the gene names from the Refseq annotation to the ensembl one.
+
+```{r results="hide"}
+library(biomaRt)
+ensembl = biomaRt::useMart("ensembl")
+ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
+```
+
+```{r enrichment_analysis, results="hide", message=FALSE}
+cluster = 8
+genes = gene_names[labels == cluster]
+
+# convert gene names
+genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		filters="refseq_mrna", values=genes,
+		mart=ensembl)["entrezgene"]
+genes = as.vector(unlist(genes))
+pathways = KEGGprofile::find_enriched_pathway(
+    genes, species="mmu",
+    download_latest=TRUE)$stastic
+
+```
+
+```{r echo=FALSE}
+
+pathways = pathways[
+    order(pathways["pvalueAdj"]),]
+
+# Select the columns to print
+pathways = subset(
+    pathways,
+    select=c(
+	"Pathway_Name", "Percentage",
+	"pvalueAdj"))
+
+colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
+pathways["Percentage"] = pathways["Percentage"] * 100
+kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+### Finding enriched GO terms
+
+To find GO terms, we use `biomaRt` to find the mapping between GO terms and
+gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
+mapping to be a list where each item is a mapping between a gene name and a GO
+term ID vector.
+
+```text
+$NM_199153
+[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
+[6] "GO:0050896" "GO:0050909"
+
+$NM_201361
+ [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
+ [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
+[11] "GO:0005819" "GO:0000922" "GO:0072686"
+```
+
+`biomaRt` queries results a matrix with two named columns of gene names and GO
+term ID.
+
+```{r go_to_gene_mapping, results="hide", message=FALSE}
+library(topGO)
+genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = moanin::create_go_term_mapping(genes)
+```
+
+Once the gene ID to GO mapping list is created, `moanin` provides a simple
+interface to `topGO` to fetch enriched GO terms. Here, we show an example of
+running a GO term enrichment on the "Biological process" ontology (BP).
+
+```{r go_term_enrichment, results="hide", message=FALSE}
+assignments = labels == cluster
+
+go_terms_enriched = find_enriched_go_terms(
+    assignments,
+    gene_id_go_mapping, ontology="BP")
+```
+
+```{r echo=FALSE}
+
+go_terms_enriched = go_terms_enriched[
+    order(go_terms_enriched[, "resultFisher_padj"]), ]
+
+kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+# Conclusion
+
+This workflow provides a tutoral for the analysis of time-course gene
+expression data in R, illustrated through the analysis of mice lung tissue
+exposed to different influenza strain. It covers four main steps; (1) quality
+control and normalization; (2) differential expression analysis; (3)
+clustering of time-course gene expression data; (4) downstream analysis of
+clusters.
+
+# Software and data availability
+
+The source code for this workflow can be found at **FIXME** 
+Data used in this workflow are available from NCBI GEO, accession GSE95601.
+
+# Author contributions
+
+NV and EP wrote the workflow.
+
+# Competing interests
+The authors declare that they have no competing interests.
+
+# Grant information
+
+This research was funded in part by a Department of Energy (DOE) grant
+(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
+the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
+California, Berkeley [N.V.]; by a ENS-CFM Data Science Chair [E.P.].
+
+*I confirm that the funders had no role in study design, data collection and
+analysis, decision to publish, or preparation of the manuscript.*
+
+# Acknowledgments
+
+The authors thank Karthik Ram and the Ropensci community for valuable
+feedback.
+
 # References

### 2019-05-08 scripts/02_differential_expression_analysis.Rmd
Just saving before deleting files.

--- 2019-05-07 scripts/02_differential_expression_analysis.Rmd
@@
 library(stats)
-library(moanin)
 library(knitr)
 library(kableExtra)
+
+library(NMF)
+library(KEGGprofile)
+library(viridis)
+
+library(moanin)
+
 source("utils.R")
@@
 
-```{r echo=FALSE, results="hide"}
-# Rename stuff for saving
-outname = ".results/pvalues.txt"
-colnames(pvalues) = unlist(
-    lapply(colnames(pvalues), function(x){paste0(x, "-pval")}))
+# Clustering of time-course data
 
-colnames(qvalues) = unlist(
-    lapply(colnames(qvalues), function(x){paste0(x, "-qval")}))
+The very large number of genes found differentially expressed impair any
+interpretation one would attempt: with 70\% of the genome found differentially
+expressed, all pathways are affected by the treatment. Hence the next step of
+the workflow to cluster gene expression according to their dynamical response
+to the treatment. Before clustering the genes, we first reduce the set of
+genes of interest to genes that are (1) significantly found differentially
+expressed; (2) "highly" differentially expressed. To do this, we first
+aggregate all p-values obtained during the time-course differential expresison
+step in a single p-value using Fisher's method [@fisher:statistical]. Then we
+select all the genes which have a Fisher adjusted p-value below 0.05 and a log
+fold change of at least two between at least one condition and one time-point.
+Reducing the set of genes on which to perform the clustering allows to
+estimate the centroids with more stability.
 
-colnames(log_fold_change_epicon) = unlist(
-    lapply(colnames(log_fold_change_epicon),
-    function(x){paste0(x, "-lfc_epicon")}))
+```{r filter_genes}
+# Then rank by fisher's p-value and take max the number of genes of interest
+# Filter out q-values for the pvalues table
+fishers_pval = moanin::pvalues_fisher_method(pvalues)
+fishers_qval = stats::p.adjust(fishers_pval)
 
-colnames(log_fold_change_max) = unlist(
-    lapply(colnames(log_fold_change_max),
-    function(x){paste0(x, "-lfc_max")}))
-
-de_analysis = merge(pvalues, qvalues, all=TRUE, by="row.names")
-row.names(de_analysis) = row.names(pvalues)
-de_analysis = subset(de_analysis, select=-c(Row.names))
-
-de_analysis = merge(de_analysis, log_fold_change_max, all=TRUE, by="row.names")
-row.names(de_analysis) = row.names(pvalues)
-de_analysis = subset(de_analysis, select=-c(Row.names))
-
-de_analysis = merge(de_analysis, log_fold_change_epicon, all=TRUE,
-		    by="row.names")
-row.names(de_analysis) = row.names(pvalues)
-de_analysis = subset(de_analysis, select=-c(Row.names))
-
-write.table(de_analysis, outname, sep="\t")
+genes_to_keep = row.names(
+    log_fold_change_max[
+	(rowSums(log_fold_change_max > 2) > 0) &
+	(fishers_qval < 0.05), ])
+# Keep the data corresponding to the genes of interest in another variable.
+y = as.matrix(data[genes_to_keep, ])
 ```
 
+After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
+clustering. As observed by looking at genes found differentially expressed, 
+many genes share a similar gene expression pattern, but on different scale.
+We thus propose the following adaptation of k-means:
+
+- **Splines estimation**: for each gene, fit the splines function with the basis
+  of your choice.
+- **Rescaling splines**: for each gene, rescale the estimated splines function
+  such that the values are bounded between 0 and 1.
+- **K-means**: apply k-means on the rescaled fitted splines to estimate the
+  centroids.
+- **Assign scores and labels to all genes**: then assign a score and a label
+  to all gene based on a goodness-of-fit measure on the raw data. By default,
+  the `splines_kmeans_score_and_label` function only labels the best 50% of
+  genes.
+
+The first three steps are performed jointly by the `splines_kmeans` function.
+
+```{r clustering}
+# First fit the kmeans clusters
+kmeans_clusters = moanin::splines_kmeans(
+    y, splines_model, n_clusters=8,
+    random_seed=42,
+    n_init=20)
+```
+
+We then use the `plot_centroids` function to visualize the centroids obtained
+with the splines k-means model.
+
+```{r fig.height=6, fig.width=6}
+moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
+		       colors=ann_colors$Group,
+		       smooth=TRUE)
+```
+
+The scoring and labeling can be done via the `splines_kmeans_score_and_label`
+function.
+
+```{r}
+# Then assign scores and labels to all the data, using a goodness-of-fit
+# scoring function.
+scores_and_labels = moanin::splines_kmeans_score_and_label(
+    data, kmeans_clusters)
+labels = scores_and_labels$labels
+
+# Let's keep only the list of genes that have a label.
+labels = unlist(labels[!is.na(labels)])
+```
+
+Before performing the next steps, let us investigate in more detail the
+differences between the labels provided by the splines k-means model and the
+scoring and labelinig step.
+
+```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
+par(mfrow=c(2, 1))
+barplot(table(kmeans_clusters$clusters),
+	col="black",
+	main="splines k-means labeling")
+barplot(table(labels), col="black", main="score and label")
+```
+
+The scoring and label step allows to label genes that were removed during the
+filtering step, yet are good matches to the centroids found during the
+clustering.
+
+```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
+labels = unlist(labels[!is.na(labels)])
+genes_scored = names(labels)
+
+kmeans_labels = kmeans_clusters$clusters
+kmeans_labels = kmeans_labels[genes_scored]
+
+labels = labels[names(kmeans_labels)]
+
+both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
+# Create another cluster.
+#both_labels[is.na(both_labels)] = 21
+confusion_matrix = table(both_labels)
+text = as.matrix(as.character(confusion_matrix))
+dim(text) = dim(confusion_matrix)
+NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
+	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
+	      Colv=NA)
+labels = scores_and_labels$labels
+labels = unlist(labels[!is.na(labels)])
+
+```
+
+```{r results="hide", echo=FALSE}
+# Save the clustering for the GO term and pathway enrichment on the whole set
+# of clusters.
+kmeans_labels = as.data.frame(labels[!is.na(labels)])
+write.table(kmeans_labels,
+	    ".results/clustering_labels.txt",
+	    sep="\t")
+```
+
+## Looking at specific clusters in details.
+
+Now, let us look more in detail some specific clusters. Cluster 8 seems
+particularly interesting: it captures genes with strong differences between
+the differant influenza treatments and the control, while the control remains
+relatively flat.
+
+Heatmaps are useful to investigate the range of expression patterns for
+specific genes. Here, we are going to plot heatmaps of the normalized gene
+expression patterns and the rescaled gene expression patterns side by side.
+
+First, select the genes of interest.
+
+```{r}
+cluster_to_plot = 8
+genes_to_plot =  names(labels[labels == cluster_to_plot])
+```
+
+```{r fig.width=6, fig.height=8}
+layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
+
+data_to_plot = data[genes_to_plot, ]
+submeta = meta
+ord = order(
+  submeta$Group,
+  submeta$Timepoint,
+  submeta$Replicate)
+submeta$Timepoint = as.factor(submeta$Timepoint)
+
+data_to_plot = data_to_plot[, ord]
+submeta = submeta[ord, ]
+
+res = NMF::aheatmap(
+    data_to_plot,
+    Colv=NA,
+    color="YlGnBu",
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=FALSE,
+    annColors=ann_colors,
+    main=paste("Cluster", cluster_to_plot, "(raw)"),
+    treeheight=0, legend=FALSE)
+
+# Now use the results of the previous call to aheatmap to reorder the genes.
+NMF::aheatmap(
+    moanin:::rescale_values(data_to_plot)[res$rowInd,],
+    Colv=NA,
+    Rowv=NA,
+    annCol=submeta[,
+                c("Group", "Timepoint")],
+    annLegend=TRUE,
+    annColors=ann_colors,
+    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
+    treeheight=0)
+
+```
+
+Those two heatmaps demonstrate that the clustering method successfully cluster
+genes that are on different scales, and yet share the same dynamical response
+to the treatments.
+
+## How to choose the number of clusters.
+
+A common question that arises when performing clustering is how to choose the
+number of clusters. A choice for the number of clusters K depends on the goal. 
+In this particular case, the end goal is not the clustering, but to facilitate
+interpretation of the differential expression analysis step. As a result, the
+number of clusters should not exceed the number of gene sets the user wants to
+interpret. This allows to set a maximum number of clusters. Let us assume here
+that this number is 20 clusters.
+
+Once the maximum number of clusters is set, several strategies allow to
+identify the number of clusters:
+
+- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
+  the elbow methods looks at the total with cluster sum of squares 
+  as a function of the number of clusters (WCSS). When adding clusters
+  doesn't decrease the WCSS by a sufficient
+  amount, one can consider stopping. This method thus provides visual aid to
+  the user to choose the number of clusters, but often the "elbow" is hard to
+  see on real data, where the number of clusters is not clearly defined.
+- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
+  refers to a method of validation of consistency within clusters, and
+  provides visual aid to choose the number of clusters.
+- **Stability methods** Stability methods are more computationally intensive
+  than any other method, as they rely on assessing the stability of the
+  clustering for every $k$ to a small randomization of the data. The user is
+  then invited to choose the number of cluster based on a number of similarity
+  measures.
+
+First, let us run the clustering for all possible clusters of interest. We
+will, for each clustering experiment, conserve (1) with within cluster sum of
+squares; (2) the labels assigned to all genes.
+
+```{r running_clustering_on_all_k}
+all_possible_n_clusters = c(5, 9, 11, 15)
+all_clustering = list()
+wss_values = list()
+
+i = 1
+for(n_cluster in all_possible_n_clusters){
+    clustering_results = moanin::splines_kmeans(
+	y, splines_model,
+	n_clusters=n_cluster, random_seed=42,
+	n_init=10)
+    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
+    all_clustering[[i]] = clustering_results$clusters
+    i = i + 1
+}
+```
+
+### Elbow method
+
+The Elbow method to choose the number of clusters relies on visualization aid
+to choose the number of cluster. The method relies on plotting the within
+cluster sum of squares as a function of the number of clusters. At some point,
+the WCSS will stop dropping, giving an angle in the graph.  The number of
+cluster is chosen at this "Elbow point."
+
+```{r elbow_method, fig.width=6, fig.height=4}
+plot(all_possible_n_clusters, wss_values,
+     type="b", pch=19, frame=FALSE, 
+     xlab="Number of clusters K",
+     ylab="Total within-clusters sum of squares")
+
+```
+
+### Average silhouette method
+
+The silhouette value is a measure of how similar a data point is to its own
+cluster (cohesion) compared to other clusters (separation).
+
+```{r average_silhouette_method, fig.width=6, fig.height=4}
+# function to compute average silhouette for k clusters
+average_silhouette = function(labels, y) {
+    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
+    return(mean(silhouette_results[, 3]))
+}
+
+# extract the average silhouette
+average_silhouette_values = list()
+i = 1
+for(i in 1:length(all_clustering)){
+    labels = all_clustering[i]
+    average_silhouette_values[i] = average_silhouette(labels, y)
+    i = i + 1
+}
+
+plot(all_possible_n_clusters, average_silhouette_values,
+     type="b", pch=19, frame=FALSE,
+     xlab="Number of clusters K",
+     ylab="Average Silhouettes")
+```
+
+### Looking at the stability of the clustering
+
+On real data, the number of clusters is not only unknown but also 
+ambiguous: it will depend on the desired clustering resolution of the user.
+Yet, in the case of biological data, stability and reproducibility of the
+results is necessary to ensure that the biological interpretation of the
+results hold when the data or the model is exposed to reasonable
+perturbations.
+
+Methods that rely on the stability of the clustering results to choose $k$
+thus ensure that the biological interpretation of the clusters hold with
+perturbtation to the data. In addition, simulation where the data is generated
+with a well defined $k$ show that the clustering is more stable for the
+correct of number of the clusters.
+
+Most methods method to find the number of clusters with stability measures
+only provide visual aids to guide the user. The first element often visualized
+is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
+stores the proportion of clustering in which two items are clustered together.
+A perfect consensus matrix ordered such as each elements that belong to the
+same cluster are adjacent to one another which show blocks along the diagonal
+close to 1.
+
+To perform such analysis, the first step is run the clustering several times
+on a resampled dataset--either using bootstrap or subsampling.
+
+Using the bootstrapping strategy:
+```{r}
+n_genes = dim(y)[1]
+indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
+
+bootstrapped_y = y[indices, ]
+```
+
+Using the subsampling strategy, keeping 80% of the genes:
+
+```{r}
+subsample_proportion = 1
+indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
+subsampled_y = y[indices, ]
+```
+
+Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
+
+```{r fig.width=6, fig.height=4}
+library(NMF)
+stability_5 = read.table("results/stability_5.tsv", sep="\t")
+consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
+							scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+stability_20 = read.table("results/stability_20.tsv", sep="\t")
+consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
+							 scale=FALSE)
+NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
+	      Colv=FALSE,
+	      treeheight=0)
+
+```
+
+```{r echo=FALSE, results="hide"}
+n_clusters = 2:20
+all_labels = list()
+for(i in n_clusters){
+    filename = paste0("results/stability_", i, ".tsv")
+    stability = read.table(filename, sep="\t")
+    all_labels[[paste0("C", i)]] = stability
+}
+```
+
+#### The model explorer strategy
+
+The model explorer algorithm [@ben-hur:stability] proposes to estimate the
+number of clusters exploiting the observation that if the number of clusters
+is correct, the clustering results are stable to bootstrapping.  The
+distribution of similaries between bootstrapped results for each $k$ can thus
+be compared for different values of $k$ and guide the user in the choice of
+number of clusters.
+
+The model explorer strategy works as follows. First, choose a similarity
+measure between two partitions or clusters $S(C_1, C_2)$. Examples are the
+normalized mutual information or Fowlkes-Mallows. Then perform $n$ bootstrap
+experiments to estimate the cluster centroids, followed by a step of assigning
+a label to all data points. Finally, compute the pairwise similarity measure
+between all bootstrapped partition, and plot the cumulative density of the
+obtained scores.
+
+```{r fig.width=6, fig.height=6}
+moanin::plot_model_explorer(all_labels)
+```
+
+From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
+but not as stable as $k=2$. The model explorer strategy, in addition to
+visualizing the diversity of the centroids, can thus help assessing an
+adequate number of clusters.
+
+Now, replot the same model explorer, but only for the clustering experiments
+$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
+measures in that range.
+
+```{r fig.width=6, fig.height=6}
+clusters = c("C6", "C7", "C8", "C9", "C10")
+selected_labels = all_labels[clusters]
+moanin::plot_model_explorer(selected_labels)
+```
+
+#### Consensus clustering as a way to find $k$
+
+The  consensus clustering [@monti:consensus] relies on a similar idea but
+instead of looking at the cumulative density of similarity measures of
+bootstrapped clustering, the authors suggests plotting the cumulative density
+of elements of the consensus matrix.
+
+```{r eval=FALSE}
+moanin::plot_cdf_consensus(all_labels)
+```
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_CDF_consensus.png")
+```
+
+The stability of the clustering based on the consensus matrix can then be
+measured via a single number by looking at the area under the curve: the more
+stable the clustering, the closer to 0 or 1 will be the entries of consensus
+matrix. The consensus clustering strategy thus suggest at looking at either
+the AUC as a function of the number of the clusters or the "improvement" in
+the AUC as a function of the number of cluster.
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_AUC_consensus.png")
+```
+
+```{r echo=FALSE}
+library(knitr)
+knitr::include_graphics("images/clustering_delta_AUC_consensus.png")
+```
+
+The consensus clustering method suggest that the most stable is $k=2$, which
+separates over-expressed genes from under-expressed genes. While it is indeed
+a very stable clustering, it does not capture the range of gene expression
+patterns present in the data. This shows the limitation of such method on real
+data, where the number of clusters is not clearly defined.
+
+# Downstream analysis of clusters.
+
+Once good clusters are obtained, the next step is to leverage the clustering
+to ease interpretation. Classic enrichment analysis step can then be performed
+in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
+term enrichment analysis, motif enrichment analysis, etc.
+
+```{r echo=FALSE, results="hide"}
+# Set data files and options
+data("shoemaker2015")
+data = shoemaker2015$data
+meta = shoemaker2015$meta
+
+labels = read.table(".results/clustering_labels.txt", sep="\t",
+		    check.names=FALSE)
+```
+
+First, let us clean up the gene obtained and only select the genes we are
+going to use in the enrichment analysis. One can either use the whole set of
+genes, only the set of differentially expressed genes in each cluster, or a
+subset of genes that fit well to a cluster (based on some criterion).
+
+```{r}
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = as.vector(labels[!is.na(labels)])
+names(labels) = gene_names
+```
+
+### Finding enriched pathways using `biomaRt` and `KEGGprofile`
+
+Let us first tackle the case of pathway enrichment analysis. We will leverage
+the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
+for this step. `KEGGprofile` is a
+package that easily allows to perform pathway enrichment analysis on a set of
+genes labeled with the ensembl annotation. We thus need to convert the gene
+names into the appropriate format. This is where `biomaRt` comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use
+it to convert the gene names from the Refseq annotation to the ensembl one.
+
+```{r results="hide"}
+library(biomaRt)
+ensembl = biomaRt::useMart("ensembl")
+ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
+```
+
+```{r enrichment_analysis, results="hide", message=FALSE}
+cluster = 8
+genes = gene_names[labels == cluster]
+
+# convert gene names
+genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		filters="refseq_mrna", values=genes,
+		mart=ensembl)["entrezgene"]
+genes = as.vector(unlist(genes))
+pathways = KEGGprofile::find_enriched_pathway(
+    genes, species="mmu",
+    download_latest=TRUE)$stastic
+
+```
+
+```{r echo=FALSE}
+
+pathways = pathways[
+    order(pathways["pvalueAdj"]),]
+
+# Select the columns to print
+pathways = subset(
+    pathways,
+    select=c(
+	"Pathway_Name", "Percentage",
+	"pvalueAdj"))
+
+colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
+pathways["Percentage"] = pathways["Percentage"] * 100
+kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+### Finding enriched GO terms
+
+To find GO terms, we use `biomaRt` to find the mapping between GO terms and
+gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
+mapping to be a list where each item is a mapping between a gene name and a GO
+term ID vector.
+
+```text
+$NM_199153
+[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
+[6] "GO:0050896" "GO:0050909"
+
+$NM_201361
+ [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
+ [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
+[11] "GO:0005819" "GO:0000922" "GO:0072686"
+```
+
+`biomaRt` queries results a matrix with two named columns of gene names and GO
+term ID.
+
+```{r go_to_gene_mapping, results="hide", message=FALSE}
+library(topGO)
+genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = moanin::create_go_term_mapping(genes)
+```
+
+Once the gene ID to GO mapping list is created, `moanin` provides a simple
+interface to `topGO` to fetch enriched GO terms. Here, we show an example of
+running a GO term enrichment on the "Biological process" ontology (BP).
+
+```{r go_term_enrichment, results="hide", message=FALSE}
+assignments = labels == cluster
+
+go_terms_enriched = find_enriched_go_terms(
+    assignments,
+    gene_id_go_mapping, ontology="BP")
+```
+
+```{r echo=FALSE}
+
+go_terms_enriched = go_terms_enriched[
+    order(go_terms_enriched[, "resultFisher_padj"]), ]
+
+kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+# Conclusion
+
+This workflow provides a tutoral for the analysis of time-course gene
+expression data in R, illustrated through the analysis of mice lung tissue
+exposed to different influenza strain. It covers four main steps; (1) quality
+control and normalization; (2) differential expression analysis; (3)
+clustering of time-course gene expression data; (4) downstream analysis of
+clusters.
+
+# Software and data availability
+
+The source code for this workflow can be found at **FIXME** 
+Data used in this workflow are available from NCBI GEO, accession GSE95601.
+
+# Author contributions
+
+NV and EP wrote the workflow.
+
+# Competing interests
+The authors declare that they have no competing interests.
+
+# Grant information
+
+This research was funded in part by a Department of Energy (DOE) grant
+(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
+the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
+California, Berkeley [N.V.]; by a ENS-CFM Data Science Chair [E.P.].
+
+*I confirm that the funders had no role in study design, data collection and
+analysis, decision to publish, or preparation of the manuscript.*
+
+# Acknowledgments
+
+The authors thank Karthik Ram and the Ropensci community for valuable
+feedback.
+
 # References

### 2019-05-08 scripts/03_clustering.Rmd
Just saving before deleting files.

@@
 library(stats)
+library(NMF)
+library(KEGGprofile)
+library(kableExtra)
+library(viridis)
+
 library(moanin)
-library(NMF)
 source("utils.R")
-# library(moanin)
 ```
@@
 
-```
-
-```{r results="hide", echo=FALSE}
-# Save the clustering for the next step
-kmeans_labels = as.data.frame(labels[!is.na(labels)])
-write.table(kmeans_labels,
-	    ".results/clustering_labels.txt",
-	    sep="\t")
 ```
@@
 
+# Downstream analysis of clusters.
+
+Once good clusters are obtained, the next step is to leverage the clustering
+to ease interpretation. Classic enrichment analysis step can then be performed
+in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
+term enrichment analysis, motif enrichment analysis, etc.
+
+First, let us clean up the gene obtained and only select the genes we are
+going to use in the enrichment analysis. One can either use the whole set of
+genes, only the set of differentially expressed genes in each cluster, or a
+subset of genes that fit well to a cluster (based on some criterion).
+
+**FIXME**
+
+### Finding enriched pathways using `biomaRt` and `KEGGprofile`
+
+Let us first tackle the case of pathway enrichment analysis. We will leverage
+the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
+for this step. `KEGGprofile` is a
+package that easily allows to perform pathway enrichment analysis on a set of
+genes labeled with the ensembl annotation. We thus need to convert the gene
+names into the appropriate format. This is where `biomaRt` comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use
+it to convert the gene names from the Refseq annotation to the ensembl one.
+
+```{r results="hide"}
+library(biomaRt)
+ensembl = biomaRt::useMart("ensembl")
+ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
+```
+
+```{r enrichment_analysis, results="hide", message=FALSE}
+cluster = 8
+genes = gene_names[labels == cluster]
+
+# convert gene names
+genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+		filters="refseq_mrna", values=genes,
+		mart=ensembl)["entrezgene"]
+genes = as.vector(unlist(genes))
+pathways = KEGGprofile::find_enriched_pathway(
+    genes, species="mmu",
+    download_latest=TRUE)$stastic
+
+```
+
+```{r echo=FALSE}
+
+pathways = pathways[
+    order(pathways["pvalueAdj"]),]
+
+# Select the columns to print
+pathways = subset(
+    pathways,
+    select=c(
+	"Pathway_Name", "Percentage",
+	"pvalueAdj"))
+
+colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
+pathways["Percentage"] = pathways["Percentage"] * 100
+kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+### Finding enriched GO terms
+
+To find GO terms, we use `biomaRt` to find the mapping between GO terms and
+gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
+mapping to be a list where each item is a mapping between a gene name and a GO
+term ID vector.
+
+```text
+$NM_199153
+[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
+[6] "GO:0050896" "GO:0050909"
+
+$NM_201361
+ [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
+ [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
+[11] "GO:0005819" "GO:0000922" "GO:0072686"
+```
+
+`biomaRt` queries results a matrix with two named columns of gene names and GO
+term ID.
+
+```{r go_to_gene_mapping, results="hide", message=FALSE}
+library(topGO)
+genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+
+# Create gene to GO id mapping
+gene_id_go_mapping = moanin::create_go_term_mapping(genes)
+```
+
+Once the gene ID to GO mapping list is created, `moanin` provides a simple
+interface to `topGO` to fetch enriched GO terms. Here, we show an example of
+running a GO term enrichment on the "Biological process" ontology (BP).
+
+```{r go_term_enrichment, results="hide", message=FALSE}
+assignments = labels == cluster
+
+go_terms_enriched = find_enriched_go_terms(
+    assignments,
+    gene_id_go_mapping, ontology="BP")
+```
+
+```{r echo=FALSE}
+
+go_terms_enriched = go_terms_enriched[
+    order(go_terms_enriched[, "resultFisher_padj"]), ]
+
+kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
+    kable_styling("striped", full_width=TRUE) 
+```
+
+# Conclusion
+
+This workflow provides a tutoral for the analysis of time-course gene
+expression data in R, illustrated through the analysis of mice lung tissue
+exposed to different influenza strain. It covers four main steps; (1) quality
+control and normalization; (2) differential expression analysis; (3)
+clustering of time-course gene expression data; (4) downstream analysis of
+clusters.
+
+# Software and data availability
+
+The source code for this workflow can be found at **FIXME** 
+Data used in this workflow are available from NCBI GEO, accession GSE95601.
+
+# Author contributions
+
+NV and EP wrote the workflow.
+
+# Competing interests
+The authors declare that they have no competing interests.
+
+# Grant information
+
+This research was funded in part by a Department of Energy (DOE) grant
+(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
+the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
+California, Berkeley [N.V.]; by a ENS-CFM Data Science Chair [E.P.].
+
+*I confirm that the funders had no role in study design, data collection and
+analysis, decision to publish, or preparation of the manuscript.*
+
+# Acknowledgments
+
+The authors thank Karthik Ram and the Ropensci community for valuable
+feedback.
+
 # References

### 2019-05-08 scripts/manuscript.Rmd
Converting to f1000 format

--- 2019-05-08 scripts/01_quality_control_normalization.Rmd
@@
 ---
-title: "Normalization: A pipeline to analyse time-course gene expression data"
-output: rmarkdown::html_vignette
+title: "A pipeline to analyse time-course gene expression data"
+author:
+    - name: Nelle Varoquaux
+      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
+    - name: Elizabeth Purdom
+      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
+
+abstract: >
+    The phenotypic diversity of cells is governed by a complex equilibrium
+    between their genetic identity and their environmental interactions:
+    Understanding the dynamics of gene expression is a fundamental question of
+    biology. However, analysing time-course transcriptomic data raises unique
+    challenging statistical and computational questions, requiring the
+    development of novel methods and software.  Using as case study
+    time-course transcriptomics data from mice exposed to different strains of
+    influenza, ihis workflows provides a step-by-step tutorial of the
+    methodology used to analyse time-course data: (1) normalization of the
+    micro-array dataset; (2) differential expression analysis using functional
+    data analysis; (3) clustering fo time-course data; (4) interpreting
+    clusters with GO term and KEGG pathway enrichment analysis.
+
+keywords: >
+    time-course gene expression data, clustering, differential expression,
+    workflow
+vignette: >
+    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
+    %\VignetteEngine{knitr::rmarkdown}
+output: BiocWorkflowTools::f1000_article
+
 bibliography: bibliography.bib
@@
 library(kableExtra)
+library(BiocStyle)
+library(biomaRt)
 
@@
 
+library(topGO)
+
 library(moanin)
@@
 knitr::opts_chunk$set(
-  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
+  cache=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
   echo=TRUE,
@@
 developmental systems, a massive number of genes will show *some* change. For
-example, in the **xxx** system, over **xxx%** of genes are shown to be changing
-over time. The task in these settings is often not to detect changes in genes,
-but to categorize into biologically interpretable patterns the vast number of
-changes discovered.
+example, in the mice lung tissues infected with influenza, over 50% of genes
+are shown to be changing over time. The task in these settings is often not to
+detect changes in genes, but to categorize into biologically interpretable
+patterns the vast number of changes discovered.
 
@@
 broadly used by the community. We provide the various steps of the workflow as
-functions in a R package **FIXME**.
+functions in a R package called `moanin`.
 
 ```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
-library(BiocStyle)
-library(knitr)
 knitr::include_graphics("images/fig1_schemas.png")
@@
 avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
-10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
-of the Vietnam avian influenza virus (10^3 PFU).
+$10^5$ PFU of each virus. An adidtional 42 mice were injected with a lower dose
+of the Vietnam avian influenza virus ($10^3$ PFU).
 
@@
 
-meta$Timepoint = as.factor(meta$Timepoint)
+# Copy the metadata before changing the timepoints into factors
+data_corr_meta = meta
+data_corr_meta$Timepoint = as.factor(meta$Timepoint)
 NMF::aheatmap(
@@
     Colv=NA, Rowv=NA,
-    annCol=meta[, c("Group", "Timepoint")],
-    annRow=meta[, c("Group", "Timepoint")],
+    annCol=data_corr_meta[, c("Group", "Timepoint")],
+    annRow=data_corr_meta[, c("Group", "Timepoint")],
     annLegend=TRUE, 
@@
 
-## Differential expression analysis of time-course data
+# Differential expression analysis of time-course data
 
@@
 `moanin`'s function rely on an object containing a number of information such
-as the metadata and the splines model used.
+as the metadata and the splines model used. We start by creating the
+`splines_model` object using the `create_splines_model` function.
 
@@
 
-### Weekly differential expression analysis
+## Weekly differential expression analysis
 
@@
 
-### Time-course differential expression analysis between two groups
+## Time-course differential expression analysis between two groups
 
@@
 for(i in 1:length(all_clustering)){
-    labels = all_clustering[i]
-    average_silhouette_values[i] = average_silhouette(labels, y)
+    clustering_results = all_clustering[i]
+    average_silhouette_values[i] = average_silhouette(clustering_results, y)
     i = i + 1
@@
 ```{r fig.width=6, fig.height=4}
-library(NMF)
 stability_5 = read.table("results/stability_5.tsv", sep="\t")
@@
 ```{r echo=FALSE}
-library(knitr)
 knitr::include_graphics("images/clustering_CDF_consensus.png")
@@
 ```{r echo=FALSE}
-library(knitr)
 knitr::include_graphics("images/clustering_AUC_consensus.png")
@@
 ```{r echo=FALSE}
-library(knitr)
 knitr::include_graphics("images/clustering_delta_AUC_consensus.png")
@@
 subset of genes that fit well to a cluster (based on some criterion).
-
-```{r}
-gene_names = row.names(labels)
-gene_names = gene_names[!is.na(labels)]
-labels = as.vector(labels[!is.na(labels)])
-names(labels) = gene_names
-```
 
@@
 ```{r results="hide"}
-library(biomaRt)
 ensembl = biomaRt::useMart("ensembl")
@@
 cluster = 8
+labels = unlist(labels)
+gene_names = names(labels)
 genes = gene_names[labels == cluster]
@@
 ```{r go_to_gene_mapping, results="hide", message=FALSE}
-library(topGO)
 genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
@@
 
-The source code for this workflow can be found at **FIXME** 
+The source code for this workflow can be found at [https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
 Data used in this workflow are available from NCBI GEO, accession GSE95601.

### 2019-05-08 bin/moanin/R/validation.R

@@ -14,21 +14,38 @@ check_meta = function(meta, check_replicates=FALSE){
     metadata_column_names = colnames(meta)
     if(!("Group" %in% metadata_column_names)){
 	stop(
+	    "moanin::create_splines_model: " +
 	    "Metadata doesn't contain expected information." +
 	    " Group column is missing.")
     }
 
     if(!("Timepoint" %in% metadata_column_names)){
 	stop(
+	    "moanin::create_splines_model: " +
 	    "Metadata doesn't contain expected information." +
 	    " Timepoint column is missing.")
     }
 
     if(check_replicates & !("Replicate" %in% metadata_column_names)){
 	stop(
+	    "moanin::create_splines_model: " +
 	    "Metadata doesn't contain expected information." +
 	    "Replicate column is missing")
     }
+
+    # Check that Timepoint is numeric.
+    if(!is.numeric(meta$Timepoint)){
+	stop(
+	    "moanin::create_splines_model: " +
+	    "Timepoint column is expected to be numeric")
+    }
+
+    if(!is.factor(meta$Group)){
+	stop(
+	    "moanin::create_splines_model: " +
+	    "Group column is expected to be factors")
+    }
+
 
     # Just create this one.
     if(!("WeeklyGroup" %in% metadata_column_names)){

### 2019-05-08 scripts/manuscript.Rmd
Manuscript is now compiled in PDF format

@@
   meta$Replicate)
-data = data[, ord]
+variance_filtered_data = variance_filtered_data[, ord]
 meta = meta[ord, ]
@@
 ```{r echo=FALSE}
-kable(head(log_fold_change_timepoints[,
-colnames(log_fold_change_timepoints)[1:5]]))
+kable(
+    head(log_fold_change_timepoints[, colnames(log_fold_change_timepoints)[1:5]]),
+    format = "latex")
 ```
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
+kable(pathways[1:10, ], "latex", escape=F, digits=4) %>%
     kable_styling("striped", full_width=TRUE) 
@@
     order(go_terms_enriched[, "resultFisher_padj"]), ]
-
-kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
+colnames(go_terms_enriched) = c("GO ID", "Description", "Annotated",
+"Significant", "Expected", "P-value", "adj. p-value")
+
+kable(go_terms_enriched[1:15, ], "latex", escape=F, digits=4)  %>%
     kable_styling("striped", full_width=TRUE) 

### 2019-05-08 scripts/manuscript.Rmd
Abstract was duplicated
FIX travis

--- 2019-05-08 scripts/manuscript.Rmd
@@
     time-course transcriptomics data from mice exposed to different strains of
-    influenza, ihis workflows provides a step-by-step tutorial of the
+    influenza, this workflows provides a step-by-step tutorial of the
     methodology used to analyse time-course data: (1) normalization of the
@@
 
-# Abstract
-
-The phenotypic diversity of cells is governed by a complex equilibrium between
-their genetic identity and their environmental interactions: Understanding the
-dynamics of gene expression is a fundamental question of biology. However,
-analysing time-course transcriptomic data raises unique challenging
-statistical and computational questions, requiring the development of novel
-methods and software.  Using as case study time-course transcriptomics data
-from mice exposed to different strains of influenza, ihis workflows provides a
-step-by-step tutorial of the methodology used to analyse time-course data:
-(1) normalization of the micro-array dataset; (2) differential expression
-analysis using functional data analysis; (3) clustering fo time-course data;
-(4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
-
 # Introduction
@@
 
-```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
+```{r schema, echo=FALSE, fig.width=4, fig.cap="Workflow for analyzing time-course datasets."}
+
 knitr::include_graphics("images/fig1_schemas.png")
-```
-
-# Analysis of the dynamical response of mouse lung tissue to influenza
-
-## Overview
-
-This workflow is illustrated using data from a micro-array time-course
-experiment, exposing mice to three different strains of influenza, and
-collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
-18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
-[@shoemaker:ultrasensitive]. The three strains of influenza used in the
-study are (1) a low pathogenicity seasonal H1N1 influenza virus
-(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
-pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
-avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
-$10^5$ PFU of each virus. An adidtional 42 mice were injected with a lower dose
-of the Vietnam avian influenza virus ($10^3$ PFU).
-
-By combining gene expression time-course data with virus growth data, the
-authors show that the inflammatory response of lung tissue is gated until a
-threshold of the virus concentration is  exceeded in the lung. Once this
-threshold is exceeded, a strong inflammatory and cytokin production occurs.
-This results provides evidence that the pathology response is non-linearly
-regulated by virus concentration.
-
-## References
-
-## Pre-processing and normalization
-
-The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
-with accession number GSE63786. The dataset has a total of 209 samples, with
-45018 probes (1 sample was removed by the authors prior to uploading the
-dataset on GEO).
-
-Note that in the following, we assume that the user has access to a data
-folder `data/shoemaker2015/rawdata`, containing the microarray data.
-
-```{r}
-data_dir = "data/shoemaker2015/rawdata"
-```
-
-Given the source and a `Targets.txt` structured as follows, microarray can be loaded thanks to the limma library.
 
 ```
-FileName
-GSM1557140_AR0313_043raw.txt.gz
-GSM1557141_AR0313_044raw.txt.gz
-GSM1557142_AR0313_045raw.txt.gz
-GSM1557143_AR0313_055raw.txt.gz
-GSM1557144_AR0313_056raw.txt.gz
-GSM1557145_AR0313_057raw.txt.gz
-GSM1557146_AR0313_058raw.txt.gz
-GSM1557147_AR0313_059raw.txt.gz
-```
-
-```{r message=FALSE, results="hide", warning=FALSE}
-targets = limma::readTargets()
-RG = limma::read.maimages(
-  targets, source="agilent",
-  path=data_dir,
-  other.columns="gIsWellAboveBG",
-  green.only=TRUE)
-
-```
-
-The normalization strategy we apply here relies on `limma` and is as follows:
-
-  - First, normalize probe intensity using background information using the
-    "norm-exponential" strategy.
-  - Then normalize across micro-arrrays using quantile normalization.
-
-```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
-# Perform background correction using norm-exponential strategy and normalize
-# across samples using quantile normalization.
-RG = limma::backgroundCorrect(RG, method="normexp")
-RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
-```
-
-Plotting the smoothed empirical densities of the probe values before
-normalization shows considerable variations betweens arrays. Once the data has
-been normalized across array using quantile normalization, the smoothed
-empirical densities of each sample perfectly overlap.
-
-```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
-# Plot the densities of the probes before and after normalization.
-par(mfrow=c(1, 2))
-limma::plotDensities(RG, col="black", legend=FALSE,
-		     main="Before normalization")
-limma::plotDensities(RG.q, col="black", legend=FALSE,
-		     main="After normalization")
-```
-
-```{r echo=FALSE, results="hide"}
-# Get the number for the text. Do not display results or computation
-control_probes = RG.q$genes[, "ControlType"] == 1
-is_expressed = rowSums(RG.q$other$"gIsWellAboveBG" > 0) >= 4
-n_control_probes = sum(control_probes)
-n_lowly_expressed = sum(!is_expressed)
-```
-
-We then clean up the names of each gene and each sample to match the original
-data. Finally, we further filter out probes based on the following criteria:
-(1) removing the `r n_control_probes` control probes; (2) removing 
-`r n_lowly_expressed` low-expressed genes.
-
-```{r clean_up_names}
-
-sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
-colnames(RG.q) = sample_names
-row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
-idx_to_keep = which(
-  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
-			       "NegativeControl")))
-
-data = RG.q[idx_to_keep,]
-
-# Filter out control probes.
-control_probes = data$genes[, "ControlType"] == 1
-is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4
-
-data = data[!control_probes & is_expressed, ]$E
-```
-
-```{r echo=FALSE, results="hide"}
-# Do not display. setting variables for text
-n_samples = dim(data)[2]
-n_probes = dim(data)[1]
-```
-
-After filtering, we are left with `r n_probes` probes of interest in 
-`r n_samples` samples.
-
-Now that we have normalized and filtered the gene expressions, let's load the
-metadata. The package `moanin` contains the normalized data and meta of
-[@shoemaker:ultrasensitive].
-
-```{r}
-# Now load in the metadata
-data(shoemaker2015)
-meta = shoemaker2015$meta
-data = shoemaker2015$data
-```
-
-## Exploratory analysis and quality control
-
-Plotting the smoothed empirical densities of probe values for each sample and
-estimating the variations across samples before and after normalization is a
-first quality control step. Typically, two other quality control and
-exploratory analysis steps are also performed before and after normalization:
-(1) low dimensionality embedding of the samples; (2) correlation plots between
-each samples. In both cases, we expect a strong biological signal, while
-replicate samples should be strongly clustered or correlated with one another.
-
-Before performing any additional exploratory analysis, let us only keep highly
-variable genes: we keep for this step only the top 50% most variable genes.
-
-```{r}
-variance_cutoff = 0.5
-variance_per_genes = apply(data, 1, mad)
-min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
-variance_filtered_data = data[variance_per_genes > min_variance,]
-```
-
-```{r echo=FALSE, fig.width=8, fig.height=4}
-graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
-	       ylab="# probes")
-graphics::abline(v=min_variance, col="#AB0000", lw=2)
-```
-
-Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
-centered and scaled gene expression data.
-
-```{r pca, fig.width=8, fig.height=8}
-# Reorder genes on condition, time, and replicate
-
-pca_data = stats::prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
-percent_var = round(100 * attr(pca_data, "percentVar"))
-```
-
-We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate.
-We also plot the second and third components in the second row.
-
-```{r pca_plots, fig.width=8, fig.height=8}
-graphics::par(mfrow=c(2, 2))
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
-    col=ann_colors$Group[meta$Group],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
-    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
-
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Group[meta$Group],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
-
-```
-
-Then, we plot the pearson correlation across each samples. We order the
-samples by their Group (treatment) and Timepoint (the time of sampling).
-Additionally, in this example, we order each treatment by strength of the
-pathogenicity of the treatment: Control, Kawasaki, California, low dose of
-Vietnam, then high dose of Vietnam.
-
-```{r  correlation_plot, fig.width=6, fig.height=6}
-
-# Reorder the conditions such that:
-#	- Control is before any influenza treatment
-#	- Each treatment is ordered from low to high pathogeny
-meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
-
-# Reorder genes on condition, time, and replicate
-ord = order(
-  meta$Group,
-  meta$Timepoint,
-  meta$Replicate)
-data = data[, ord]
-meta = meta[ord, ]
-
-data_corr = stats::cor(variance_filtered_data, method="pearson")
-
-# Copy the metadata before changing the timepoints into factors
-data_corr_meta = meta
-data_corr_meta$Timepoint = as.factor(meta$Timepoint)
-NMF::aheatmap(
-    data_corr,
-    Colv=NA, Rowv=NA,
-    annCol=data_corr_meta[, c("Group", "Timepoint")],
-    annRow=data_corr_meta[, c("Group", "Timepoint")],
-    annLegend=TRUE, 
-    annColors=ann_colors,
-    main="Correlation plot")
-```
-
-We can already see interesting patterns emerging from the correlation plot.
-First, the cross-correlation amongst samples taking from the control mice is
-higher than the cross correlation amongst the rest of the treatments. Second,
-the influenza-infected mice midly react until time point 36. Third, the less
-pathogenic the strain is, the closer the samples are to the control condition.
-Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
-most different from control samples.
-
-# Differential expression analysis of time-course data
-
-The next step is typically to run a differential expression analysis. Most of
-`moanin`'s function rely on an object containing a number of information such
-as the metadata and the splines model used. We start by creating the
-`splines_model` object using the `create_splines_model` function.
-
-```{r}
-splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
-```
-
-## Weekly differential expression analysis
-
-`moanin` provides a simple interface to perform a time by time differential
-expression analysis. Under the hood, it simply calls `limma` on the set of
-contrasts provided. Here, we show an example between the control mouse and the
-mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
-(H5N1), but the function works with any form contrasts.
-
-```{r weekly_differential_analysis}
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Timepoint))
-
-# Define contrast	
-contrasts = c("M.0-VL.0",
-	      "M.3-VL.3",
-	      "M.6-VL.6",
-	      "M.9-VL.9",
-	      "M.12-VL.12",
-	      "M.18-VL.18",
-	      "M.24-VL.24",
-	      "M.30-VL.30",
-	      "M.36-VL.36",
-	      "M.48-VL.48",
-	      "M.60-VL.60",
-	      "M.72-VL.72",
-	      "M.120-VL.120",
-	      "M.168-VL.168")
-weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, splines_model, contrasts,
-     use_voom_weights=FALSE)
-```
-
-```{r echo=FALSE, results="hide"}
-# Hide those results
-# Run the DE analysis on the rest of the influenza strain.
-contrasts = c("M.0-K.0",
-	      "M.3-K.3",
-	      "M.6-K.6",
-	      "M.9-K.9",
-	      "M.12-K.12",
-	      "M.18-K.18",
-	      "M.24-K.24",
-	      "M.30-K.30",
-	      "M.36-K.36",
-	      "M.48-K.48",
-	      "M.60-K.60",
-	      "M.72-K.72",
-	      "M.120-K.120",
-	      "M.168-K.168")
-K_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, splines_model, contrasts,
-     use_voom_weights=FALSE)
-contrasts = c("M.0-C.0",
-	      "M.3-C.3",
-	      "M.6-C.6",
-	      "M.9-C.9",
-	      "M.12-C.12",
-	      "M.18-C.18",
-	      "M.24-C.24",
-	      "M.30-C.30",
-	      "M.36-K.36",
-	      "M.48-C.48",
-	      "M.60-C.60",
-	      "M.72-C.72",
-	      "M.120-C.120",
-	      "M.168-C.168")
-C_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, splines_model, contrasts,
-     use_voom_weights=FALSE)
-contrasts = c("M.0-VH.0",
-	      "M.3-VH.3",
-	      "M.6-VH.6",
-	      "M.9-VH.9",
-	      "M.12-VH.12",
-	      "M.18-VH.18",
-	      "M.24-VH.24",
-	      "M.30-VH.30",
-	      "M.36-VH.36",
-	      "M.48-VH.48",
-	      "M.60-VH.60",
-	      "M.72-VH.72",
-	      "M.120-VH.120",
-	      "M.168-VH.168")
-VH_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
-     data, splines_model, contrasts,
-     use_voom_weights=FALSE)
-
-```
-
-Let's look at the distribution of genes found differentially expressed per
-week between control and each of the influenza strains.
-
-```{r fig.width=8, fig.height=8, echo=FALSE}
-par(mfrow=c(2, 2))
-
-time = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
-
-timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
-    qval_colnames = colnames(de_analysis)[
-	grepl("qval", colnames(de_analysis))]
-    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
-    names(number_de_genes_per_time) = labels
-    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
-	    xlab="Timepoint", ylab="Number of DE genes", main=title)
-
-}
-
-timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
-timely_barplot(C_weekly_de_analysis, time, title="California")
-timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose")
-timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose")
-```
-
-The distribution of number of genes found differentially expressed by
-considering each time-point independantly highlights the challenges of such
-approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
-genes found differentially expressed than timepoints 3H and 9H: this is likely
-due to biological or technical variances for specific genes at specific
-timepoints.
-
-As a summary, classic differential expression methods are appropriate for
-unordered treatments, but fail to use the temporal structure of the data.
-
-## Time-course differential expression analysis between two groups
-
-To leverage this temporal structure, Storey et al [@storey:significance] proposes to
-model time-course micro-array with splines functions, and to use a log-ratio
-likelihood test to detect differentially expressed genes. `moanin` extends
-this idea by providing functions to perform time-course differential
-expression analysis using `limma` contrasts.
-
-```{r de_analysis}
-# Differential expression analysis
-timecourse_contrasts = c("M-K", "M-C", "M-VL", "M-VH")
-
-contrasts = limma::makeContrasts(
-	contrasts=timecourse_contrasts,
-	levels=levels(meta$Group))
-
-# The function takes the data (data.frame or named matrix), the meta data
-# (data.frame containing a timepoint and group column, the first corresponding
-# to the time-course information, the latter corresponding to the
-# treatment).
-pvalues = moanin::timecourse_differential_expression_analysis(
-    data, splines_model, contrasts)
-qvalues = apply(pvalues, 2, stats::p.adjust)
-
-```
-
-The number of genes found differentially expressed ranges from around 12000 to
-29000 depending on the strain and dosage of influenza virus given to the mice.
-This corresponds to between 30% to 70% of the genes found differentially
-expressed in this time-course experiment.
-
-```{r echo=FALSE, fig.width=6, fig.height=6}
-num_de = colSums(qvalues < 0.05)
-names(num_de) = c(
-    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
-barplot(num_de, col="black", ylab="Number of DE genes")
-```
-
-The next step in a classical differential expression analysis is typically to
-assess the effect of the treatment by looking at the log fold change.
-Computing the log fold change on a time-course experiment is not trivial: one
-can be interested in the average log-fold change across time, or the
-cumulative log-fold change. Sometimes a gene can be over-expressed at the
-beginning of the time-course data, and then over-expressed at the end of the
-experiment. As a result, `moanin` provides a number of possible ways to
-compute the log fold change across the whole time-course.
-
-First, `moanin` provides as simple interface to compute the log-fold change
-for each individual timepoints.
-
-```{r}
-log_fold_change_timepoints = moanin::estimate_log_fold_change(
-    data, splines_model, contrasts,  method="timely")
-```
-
-```{r echo=FALSE}
-kable(head(log_fold_change_timepoints[,
-colnames(log_fold_change_timepoints)[1:5]]))
-```
-
-This matrix can then be used to visualize the log-fold change for each
-contrast per timepoint.
-
-Sometimes, a single value per gene and per contrast is more useful. Here is a
-table of the possible ways to compute log-fold change values.
-
-| Name   | Formula   |     |
-|------- | --------- |--- |
-| timely | $\text{lfc}(t)$  | Function of time |
-| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
-| abs_sum  | $\sum_t \|lfc(t)\|$  | Always positive |
-| max  | $\max_t \|lfc(t)\|$  | Always positive  |
-| min  | $\min_t \|lfc(t)\|$  | Always positive |
-| epicon  | $$  | Captures overall strength of response and overall direction |
-
-```{r}
-log_fold_change_epicon = moanin::estimate_log_fold_change(
-    data, splines_model, contrasts,  method="epicon")
-
-log_fold_change_sum = moanin::estimate_log_fold_change(
-    data, splines_model, contrasts,  method="sum")
-
-log_fold_change_max = moanin::estimate_log_fold_change(
-    data, splines_model, contrasts, method="max")
-
-log_fold_change_min = moanin::estimate_log_fold_change(
-    data, splines_model, contrasts, method="min")
-```
-
-```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE}
-par(mfrow=c(2, 2))
-plot(log_fold_change_epicon[, "M-C"],
-     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
-     xlab="EPICON", ylab="sum")
-plot(log_fold_change_epicon[, "M-C"],
-     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
-     xlab="EPICON", ylab="max")
-plot(log_fold_change_max[, "M-C"],
-     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
-     xlab="max", ylab="min")
-plot(log_fold_change_max[, "M-C"],
-     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
-     xlab="max", ylab="sum")
-```
-
-From the single measures of log-fold change and the p-value, we can now look
-at the volcano plot. Here is an example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the EPICON log fold change
-computation.
-
-```{r fig.height=8, fig.width=8}
-pvalue = qvalues[, "M-K"]
-lfc_epicon = log_fold_change_epicon[, "M-K"]
-names(lfc_epicon) = row.names(log_fold_change_epicon)
-
-plot(lfc_epicon, -log10(pvalue), pch=20, main="Volcano plot",
-     xlim=c(-2.5, 2))
-```
-
-As another sanity check, `moanin` provides a simple utility function to
-visualize gene time-course data. Here, we plot the 12 genes with the smallest
-p-values.
-
-```{r fig.height=8, fig.width=8}
-top_DE_genes_pval = names(sort(pvalue)[1:12])
-moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
-		   colors=ann_colors$Group, smooth=TRUE)
-```
-
-And here we visualize the genes with the largest absolute EPICON log-fold
-change.
-
-```{r fig.height=8, fig.width=8}
-top_DE_genes_lfc = names(
-    sort(abs(lfc_epicon),
-	 decreasing=TRUE)[1:12])
-moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
-		   colors=ann_colors$Group, smooth=TRUE)
-
-```
-
-Thanks to those visualization, we can see that genes often follow similar
-patterns of expression, although on a different scale for each gene. We can
-leverage this observation to cluster the genes into groups of similar patterns
-of transcriptomic response.
-
-# Clustering of time-course data
-
-The very large number of genes found differentially expressed impair any
-interpretation one would attempt: with 70\% of the genome found differentially
-expressed, all pathways are affected by the treatment. Hence the next step of
-the workflow to cluster gene expression according to their dynamical response
-to the treatment. Before clustering the genes, we first reduce the set of
-genes of interest to genes that are (1) significantly found differentially
-expressed; (2) "highly" differentially expressed. To do this, we first
-aggregate all p-values obtained during the time-course differential expresison
-step in a single p-value using Fisher's method [@fisher:statistical]. Then we
-select all the genes which have a Fisher adjusted p-value below 0.05 and a log
-fold change of at least two between at least one condition and one time-point.
-Reducing the set of genes on which to perform the clustering allows to
-estimate the centroids with more stability.
-
-```{r filter_genes}
-# Then rank by fisher's p-value and take max the number of genes of interest
-# Filter out q-values for the pvalues table
-fishers_pval = moanin::pvalues_fisher_method(pvalues)
-fishers_qval = stats::p.adjust(fishers_pval)
-
-genes_to_keep = row.names(
-    log_fold_change_max[
-	(rowSums(log_fold_change_max > 2) > 0) &
-	(fishers_qval < 0.05), ])
-# Keep the data corresponding to the genes of interest in another variable.
-y = as.matrix(data[genes_to_keep, ])
-```
-
-After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
-clustering. As observed by looking at genes found differentially expressed, 
-many genes share a similar gene expression pattern, but on different scale.
-We thus propose the following adaptation of k-means:
-
-- **Splines estimation**: for each gene, fit the splines function with the basis
-  of your choice.
-- **Rescaling splines**: for each gene, rescale the estimated splines function
-  such that the values are bounded between 0 and 1.
-- **K-means**: apply k-means on the rescaled fitted splines to estimate the
-  centroids.
-- **Assign scores and labels to all genes**: then assign a score and a label
-  to all gene based on a goodness-of-fit measure on the raw data. By default,
-  the `splines_kmeans_score_and_label` function only labels the best 50% of
-  genes.
-
-The first three steps are performed jointly by the `splines_kmeans` function.
-
-```{r clustering}
-# First fit the kmeans clusters
-kmeans_clusters = moanin::splines_kmeans(
-    y, splines_model, n_clusters=8,
-    random_seed=42,
-    n_init=20)
-```
-
-We then use the `plot_centroids` function to visualize the centroids obtained
-with the splines k-means model.
-
-```{r fig.height=6, fig.width=6}
-moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
-		       colors=ann_colors$Group,
-		       smooth=TRUE)
-```
-
-The scoring and labeling can be done via the `splines_kmeans_score_and_label`
-function.
-
-```{r}
-# Then assign scores and labels to all the data, using a goodness-of-fit
-# scoring function.
-scores_and_labels = moanin::splines_kmeans_score_and_label(
-    data, kmeans_clusters)
-labels = scores_and_labels$labels
-
-# Let's keep only the list of genes that have a label.
-labels = unlist(labels[!is.na(labels)])
-```
-
-Before performing the next steps, let us investigate in more detail the
-differences between the labels provided by the splines k-means model and the
-scoring and labelinig step.
-
-```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
-par(mfrow=c(2, 1))
-barplot(table(kmeans_clusters$clusters),
-	col="black",
-	main="splines k-means labeling")
-barplot(table(labels), col="black", main="score and label")
-```
-
-The scoring and label step allows to label genes that were removed during the
-filtering step, yet are good matches to the centroids found during the
-clustering.
-
-```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
-labels = unlist(labels[!is.na(labels)])
-genes_scored = names(labels)
-
-kmeans_labels = kmeans_clusters$clusters
-kmeans_labels = kmeans_labels[genes_scored]
-
-labels = labels[names(kmeans_labels)]
-
-both_labels = data.frame("kmeans"=kmeans_labels, "scored"=labels)
-# Create another cluster.
-#both_labels[is.na(both_labels)] = 21
-confusion_matrix = table(both_labels)
-text = as.matrix(as.character(confusion_matrix))
-dim(text) = dim(confusion_matrix)
-NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
-	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
-	      Colv=NA)
-labels = scores_and_labels$labels
-labels = unlist(labels[!is.na(labels)])
-
-```
-
-```{r results="hide", echo=FALSE}
-# Save the clustering for the GO term and pathway enrichment on the whole set
-# of clusters.
-kmeans_labels = as.data.frame(labels[!is.na(labels)])
-write.table(kmeans_labels,
-	    ".results/clustering_labels.txt",
-	    sep="\t")
-```
-
-## Looking at specific clusters in details.
-
-Now, let us look more in detail some specific clusters. Cluster 8 seems
-particularly interesting: it captures genes with strong differences between
-the differant influenza treatments and the control, while the control remains
-relatively flat.
-
-Heatmaps are useful to investigate the range of expression patterns for
-specific genes. Here, we are going to plot heatmaps of the normalized gene
-expression patterns and the rescaled gene expression patterns side by side.
-
-First, select the genes of interest.
-
-```{r}
-cluster_to_plot = 8
-genes_to_plot =  names(labels[labels == cluster_to_plot])
-```
-
-```{r fig.width=6, fig.height=8}
-layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
-
-data_to_plot = data[genes_to_plot, ]
-submeta = meta
-ord = order(
-  submeta$Group,
-  submeta$Timepoint,
-  submeta$Replicate)
-submeta$Timepoint = as.factor(submeta$Timepoint)
-
-data_to_plot = data_to_plot[, ord]
-submeta = submeta[ord, ]
-
-res = NMF::aheatmap(
-    data_to_plot,
-    Colv=NA,
-    color="YlGnBu",
-    annCol=submeta[,
-                c("Group", "Timepoint")],
-    annLegend=FALSE,
-    annColors=ann_colors,
-    main=paste("Cluster", cluster_to_plot, "(raw)"),
-    treeheight=0, legend=FALSE)
-
-# Now use the results of the previous call to aheatmap to reorder the genes.
-NMF::aheatmap(
-    moanin:::rescale_values(data_to_plot)[res$rowInd,],
-    Colv=NA,
-    Rowv=NA,
-    annCol=submeta[,
-                c("Group", "Timepoint")],
-    annLegend=TRUE,
-    annColors=ann_colors,
-    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
-    treeheight=0)
-
-```
-
-Those two heatmaps demonstrate that the clustering method successfully cluster
-genes that are on different scales, and yet share the same dynamical response
-to the treatments.
-
-## How to choose the number of clusters.
-
-A common question that arises when performing clustering is how to choose the
-number of clusters. A choice for the number of clusters K depends on the goal. 
-In this particular case, the end goal is not the clustering, but to facilitate
-interpretation of the differential expression analysis step. As a result, the
-number of clusters should not exceed the number of gene sets the user wants to
-interpret. This allows to set a maximum number of clusters. Let us assume here
-that this number is 20 clusters.
-
-Once the maximum number of clusters is set, several strategies allow to
-identify the number of clusters:
-
-- **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
-  the elbow methods looks at the total with cluster sum of squares 
-  as a function of the number of clusters (WCSS). When adding clusters
-  doesn't decrease the WCSS by a sufficient
-  amount, one can consider stopping. This method thus provides visual aid to
-  the user to choose the number of clusters, but often the "elbow" is hard to
-  see on real data, where the number of clusters is not clearly defined.
-- **Silhouette method**. Similarly to the Elbow method, the Silhouette method
-  refers to a method of validation of consistency within clusters, and
-  provides visual aid to choose the number of clusters.
-- **Stability methods** Stability methods are more computationally intensive
-  than any other method, as they rely on assessing the stability of the
-  clustering for every $k$ to a small randomization of the data. The user is
-  then invited to choose the number of cluster based on a number of similarity
-  measures.
-
-First, let us run the clustering for all possible clusters of interest. We
-will, for each clustering experiment, conserve (1) with within cluster sum of
-squares; (2) the labels assigned to all genes.
-
-```{r running_clustering_on_all_k}
-all_possible_n_clusters = c(5, 9, 11, 15)
-all_clustering = list()
-wss_values = list()
-
-i = 1
-for(n_cluster in all_possible_n_clusters){
-    clustering_results = moanin::splines_kmeans(
-	y, splines_model,
-	n_clusters=n_cluster, random_seed=42,
-	n_init=10)
-    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
-    all_clustering[[i]] = clustering_results$clusters
-    i = i + 1
-}
-```
-
-### Elbow method
-
-The Elbow method to choose the number of clusters relies on visualization aid
-to choose the number of cluster. The method relies on plotting the within
-cluster sum of squares as a function of the number of clusters. At some point,
-the WCSS will stop dropping, giving an angle in the graph.  The number of
-cluster is chosen at this "Elbow point."
-
-```{r elbow_method, fig.width=6, fig.height=4}
-plot(all_possible_n_clusters, wss_values,
-     type="b", pch=19, frame=FALSE, 
-     xlab="Number of clusters K",
-     ylab="Total within-clusters sum of squares")
-
-```
-
-### Average silhouette method
-
-The silhouette value is a measure of how similar a data point is to its own
-cluster (cohesion) compared to other clusters (separation).
-
-```{r average_silhouette_method, fig.width=6, fig.height=4}
-# function to compute average silhouette for k clusters
-average_silhouette = function(labels, y) {
-    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
-    return(mean(silhouette_results[, 3]))
-}
-
-# extract the average silhouette
-average_silhouette_values = list()
-i = 1
-for(i in 1:length(all_clustering)){
-    clustering_results = all_clustering[i]
-    average_silhouette_values[i] = average_silhouette(clustering_results, y)
-    i = i + 1
-}
-
-plot(all_possible_n_clusters, average_silhouette_values,
-     type="b", pch=19, frame=FALSE,
-     xlab="Number of clusters K",
-     ylab="Average Silhouettes")
-```
-
-### Looking at the stability of the clustering
-
-On real data, the number of clusters is not only unknown but also 
-ambiguous: it will depend on the desired clustering resolution of the user.
-Yet, in the case of biological data, stability and reproducibility of the
-results is necessary to ensure that the biological interpretation of the
-results hold when the data or the model is exposed to reasonable
-perturbations.
-
-Methods that rely on the stability of the clustering results to choose $k$
-thus ensure that the biological interpretation of the clusters hold with
-perturbtation to the data. In addition, simulation where the data is generated
-with a well defined $k$ show that the clustering is more stable for the
-correct of number of the clusters.
-
-Most methods method to find the number of clusters with stability measures
-only provide visual aids to guide the user. The first element often visualized
-is the consensus matrix: the consensus matrix is an $n \times n$ matrix that
-stores the proportion of clustering in which two items are clustered together.
-A perfect consensus matrix ordered such as each elements that belong to the
-same cluster are adjacent to one another which show blocks along the diagonal
-close to 1.
-
-To perform such analysis, the first step is run the clustering several times
-on a resampled dataset--either using bootstrap or subsampling.
-
-Using the bootstrapping strategy:
-```{r}
-n_genes = dim(y)[1]
-indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
-
-bootstrapped_y = y[indices, ]
-```
-
-Using the subsampling strategy, keeping 80% of the genes:
-
-```{r}
-subsample_proportion = 1
-indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
-subsampled_y = y[indices, ]
-```
-
-Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
-
-```{r fig.width=6, fig.height=4}
-stability_5 = read.table("results/stability_5.tsv", sep="\t")
-consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
-							scale=FALSE)
-NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
-	      Colv=FALSE,
-	      treeheight=0)
-
-stability_20 = read.table("results/stability_20.tsv", sep="\t")
-consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
-							 scale=FALSE)
-NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
-	      Colv=FALSE,
-	      treeheight=0)
-
-```
-
-```{r echo=FALSE, results="hide"}
-n_clusters = 2:20
-all_labels = list()
-for(i in n_clusters){
-    filename = paste0("results/stability_", i, ".tsv")
-    stability = read.table(filename, sep="\t")
-    all_labels[[paste0("C", i)]] = stability
-}
-```
-
-#### The model explorer strategy
-
-The model explorer algorithm [@ben-hur:stability] proposes to estimate the
-number of clusters exploiting the observation that if the number of clusters
-is correct, the clustering results are stable to bootstrapping.  The
-distribution of similaries between bootstrapped results for each $k$ can thus
-be compared for different values of $k$ and guide the user in the choice of
-number of clusters.
-
-The model explorer strategy works as follows. First, choose a similarity
-measure between two partitions or clusters $S(C_1, C_2)$. Examples are the
-normalized mutual information or Fowlkes-Mallows. Then perform $n$ bootstrap
-experiments to estimate the cluster centroids, followed by a step of assigning
-a label to all data points. Finally, compute the pairwise similarity measure
-between all bootstrapped partition, and plot the cumulative density of the
-obtained scores.
-
-```{r fig.width=6, fig.height=6}
-moanin::plot_model_explorer(all_labels)
-```
-
-From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
-but not as stable as $k=2$. The model explorer strategy, in addition to
-visualizing the diversity of the centroids, can thus help assessing an
-adequate number of clusters.
-
-Now, replot the same model explorer, but only for the clustering experiments
-$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
-measures in that range.
-
-```{r fig.width=6, fig.height=6}
-clusters = c("C6", "C7", "C8", "C9", "C10")
-selected_labels = all_labels[clusters]
-moanin::plot_model_explorer(selected_labels)
-```
-
-#### Consensus clustering as a way to find $k$
-
-The  consensus clustering [@monti:consensus] relies on a similar idea but
-instead of looking at the cumulative density of similarity measures of
-bootstrapped clustering, the authors suggests plotting the cumulative density
-of elements of the consensus matrix.
-
-```{r eval=FALSE}
-moanin::plot_cdf_consensus(all_labels)
-```
-
-```{r echo=FALSE}
-knitr::include_graphics("images/clustering_CDF_consensus.png")
-```
-
-The stability of the clustering based on the consensus matrix can then be
-measured via a single number by looking at the area under the curve: the more
-stable the clustering, the closer to 0 or 1 will be the entries of consensus
-matrix. The consensus clustering strategy thus suggest at looking at either
-the AUC as a function of the number of the clusters or the "improvement" in
-the AUC as a function of the number of cluster.
-
-```{r echo=FALSE}
-knitr::include_graphics("images/clustering_AUC_consensus.png")
-```
-
-```{r echo=FALSE}
-knitr::include_graphics("images/clustering_delta_AUC_consensus.png")
-```
-
-The consensus clustering method suggest that the most stable is $k=2$, which
-separates over-expressed genes from under-expressed genes. While it is indeed
-a very stable clustering, it does not capture the range of gene expression
-patterns present in the data. This shows the limitation of such method on real
-data, where the number of clusters is not clearly defined.
-
-# Downstream analysis of clusters.
-
-Once good clusters are obtained, the next step is to leverage the clustering
-to ease interpretation. Classic enrichment analysis step can then be performed
-in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
-term enrichment analysis, motif enrichment analysis, etc.
-
-First, let us clean up the gene obtained and only select the genes we are
-going to use in the enrichment analysis. One can either use the whole set of
-genes, only the set of differentially expressed genes in each cluster, or a
-subset of genes that fit well to a cluster (based on some criterion).
-
-### Finding enriched pathways using `biomaRt` and `KEGGprofile`
-
-Let us first tackle the case of pathway enrichment analysis. We will leverage
-the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
-for this step. `KEGGprofile` is a
-package that easily allows to perform pathway enrichment analysis on a set of
-genes labeled with the ensembl annotation. We thus need to convert the gene
-names into the appropriate format. This is where `biomaRt` comes in handy: it
-enables easy conversion from one gene annotation to another. Here, we will use
-it to convert the gene names from the Refseq annotation to the ensembl one.
-
-```{r results="hide"}
-ensembl = biomaRt::useMart("ensembl")
-ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
-```
-
-```{r enrichment_analysis, results="hide", message=FALSE}
-cluster = 8
-labels = unlist(labels)
-gene_names = names(labels)
-genes = gene_names[labels == cluster]
-
-# convert gene names
-genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
-		filters="refseq_mrna", values=genes,
-		mart=ensembl)["entrezgene"]
-genes = as.vector(unlist(genes))
-pathways = KEGGprofile::find_enriched_pathway(
-    genes, species="mmu",
-    download_latest=TRUE)$stastic
-
-```
-
-```{r echo=FALSE}
-
-pathways = pathways[
-    order(pathways["pvalueAdj"]),]
-
-# Select the columns to print
-pathways = subset(
-    pathways,
-    select=c(
-	"Pathway_Name", "Percentage",
-	"pvalueAdj"))
-
-colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
-pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:10, ], "html", escape=F, digits=4) %>%
-    kable_styling("striped", full_width=TRUE) 
-```
-
-### Finding enriched GO terms
-
-To find GO terms, we use `biomaRt` to find the mapping between GO terms and
-gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
-mapping to be a list where each item is a mapping between a gene name and a GO
-term ID vector.
-
-```text
-$NM_199153
-[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
-[6] "GO:0050896" "GO:0050909"
-
-$NM_201361
- [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
- [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
-[11] "GO:0005819" "GO:0000922" "GO:0072686"
-```
-
-`biomaRt` queries results a matrix with two named columns of gene names and GO
-term ID.
-
-```{r go_to_gene_mapping, results="hide", message=FALSE}
-genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
-	      values=gene_names,
-	      filters="refseq_mrna",
-	      mart=ensembl)
-
-# Create gene to GO id mapping
-gene_id_go_mapping = moanin::create_go_term_mapping(genes)
-```
-
-Once the gene ID to GO mapping list is created, `moanin` provides a simple
-interface to `topGO` to fetch enriched GO terms. Here, we show an example of
-running a GO term enrichment on the "Biological process" ontology (BP).
-
-```{r go_term_enrichment, results="hide", message=FALSE}
-assignments = labels == cluster
-
-go_terms_enriched = find_enriched_go_terms(
-    assignments,
-    gene_id_go_mapping, ontology="BP")
-```
-
-```{r echo=FALSE}
-
-go_terms_enriched = go_terms_enriched[
-    order(go_terms_enriched[, "resultFisher_padj"]), ]
-
-kable(go_terms_enriched[1:15, ], "html", escape=F, digits=4) %>%
-    kable_styling("striped", full_width=TRUE) 
-```
-
-# Conclusion
-
-This workflow provides a tutoral for the analysis of time-course gene
-expression data in R, illustrated through the analysis of mice lung tissue
-exposed to different influenza strain. It covers four main steps; (1) quality
-control and normalization; (2) differential expression analysis; (3)
-clustering of time-course gene expression data; (4) downstream analysis of
-clusters.
-
-# Software and data availability
-
-The source code for this workflow can be found at [https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
-Data used in this workflow are available from NCBI GEO, accession GSE95601.
-
-# Author contributions
-
-NV and EP wrote the workflow.
-
-# Competing interests
-The authors declare that they have no competing interests.
-
-# Grant information
-
-This research was funded in part by a Department of Energy (DOE) grant
-(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
-the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
-California, Berkeley [N.V.]; by a ENS-CFM Data Science Chair [E.P.].
-
-*I confirm that the funders had no role in study design, data collection and
-analysis, decision to publish, or preparation of the manuscript.*
-
-# Acknowledgments
-
-The authors thank Karthik Ram and the Ropensci community for valuable
-feedback.
-
-# References

### 2019-05-08 scripts/manuscript.Rmd
Trying to fix pdf output

@@
     %\VignetteEngine{knitr::rmarkdown}
-output: BiocWorkflowTools::f1000_article
+#output: BiocWorkflowTools::f1000_article
+output: pdf_document
 
@@
 library(kableExtra)
-library(BiocStyle)
+#library(BiocStyle)
 library(biomaRt)

### 2019-05-10 scripts/manuscript.Rmd
Reactivating style
Putting back the whole manuscript

--- 2019-05-08 scripts/manuscript.Rmd
@@
   the same thing. Find a better name for both of them
-
-# Abstract
-
-The phenotypic diversity of cells is governed by a complex equilibrium between
-their genetic identity and their environmental interactions: Understanding the
-dynamics of gene expression is a fundamental question of biology. However,
-analysing time-course transcriptomic data raises unique challenging
-statistical and computational questions, requiring the development of novel
-methods and software.  Using as case study time-course transcriptomics data
-from mice exposed to different strains of influenza, ihis workflows provides a
-step-by-step tutorial of the methodology used to analyse time-course data:
-(1) normalization of the micro-array dataset; (2) differential expression
-analysis using functional data analysis; (3) clustering fo time-course data;
-(4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
 
@@
 
-```{r schema, echo=FALSE, fig.width=8, fig.cap="Workflow for analyzing time-course datasets."}
+```{r schema, echo=FALSE, fig.width=4, fig.cap="Workflow for analyzing time-course datasets."}
+
 knitr::include_graphics("images/fig1_schemas.png")
+
 ```
@@
 # Analysis of the dynamical response of mouse lung tissue to influenza
-
-## Overview
 
@@
 
-## References
-
-## Pre-processing and normalization
+## Quality control and normalization
+
+### Pre-processing and normalization
 
@@
 
-```{r reading_targets, results="hide", fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
+```{r reading_targets, results="hide", fig.width=4, fig.height=2, message=FALSE, warning=FALSE}
 # Perform background correction using norm-exponential strategy and normalize
@@
 
-```{r plotting_densities, results="hide", fig.width=8, fig.height=4}
+```{r plotting_densities, results="hide", fig.width=6, fig.height=3}
 # Plot the densities of the probes before and after normalization.
-par(mfrow=c(1, 2))
+par(mfrow=c(1, 2), mar=c(2.0, 2.0, 2.0, 2.0))
 limma::plotDensities(RG, col="black", legend=FALSE,
-		     main="Before normalization")
+		     main="Before normalization", cex=0.8)
 limma::plotDensities(RG.q, col="black", legend=FALSE,
-		     main="After normalization")
+		     main="After normalization", cex=0.8)
 ```
@@
 
-## Exploratory analysis and quality control
+### Exploratory analysis and quality control
 
@@
 
-```{r echo=FALSE, fig.width=8, fig.height=4}
+```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
 graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
-	       ylab="# probes")
+	       ylab="# probes", main="")
 graphics::abline(v=min_variance, col="#AB0000", lw=2)
@@
 
-```{r pca, fig.width=8, fig.height=8}
+```{r pca, fig.width=6, fig.height=6}
 # Reorder genes on condition, time, and replicate
@@
 
-```{r pca_plots, fig.width=8, fig.height=8}
-graphics::par(mfrow=c(2, 2))
+```{r pca_plots, fig.width=6, fig.height=6}
+graphics::par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
 graphics::plot(
@@
 
-```{r  correlation_plot, fig.width=6, fig.height=6}
+```{r  correlation_plot, fig.width=6, fig.height=6, fig.caption="Cross-correlation of the samples"}
 
@@
   meta$Replicate)
+
 variance_filtered_data = variance_filtered_data[, ord]
-meta = meta[ord, ]
-
 data_corr = stats::cor(variance_filtered_data, method="pearson")
 
+data_corr_meta = meta[ord, ]
+
 # Copy the metadata before changing the timepoints into factors
-data_corr_meta = meta
 data_corr_meta$Timepoint = as.factor(meta$Timepoint)
@@
 
-# Differential expression analysis of time-course data
+## Differential expression analysis of time-course data
 
@@
 
-## Weekly differential expression analysis
+### Weekly differential expression analysis
 
 `moanin` provides a simple interface to perform a time by time differential
-expression analysis. Under the hood, it simply calls `limma` on the set of
+expression analysis. Under the hood, it simply calls `limma` [@ritchie:limma] on the set of
 contrasts provided. Here, we show an example between the control mouse and the
@@
 (H5N1), but the function works with any form contrasts.
+
+**FIXME rename the function to something better**
 
@@
 
-```{r fig.width=8, fig.height=8, echo=FALSE}
+```{r fig.width=6, fig.height=6, echo=FALSE}
 par(mfrow=c(2, 2))
@@
     number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
-    names(number_de_genes_per_time) = labels
+    names(number_de_genes_per_time) = labels[1]
     barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
@@
 
-## Time-course differential expression analysis between two groups
+### Time-course differential expression analysis between two groups
 
@@
 
-```{r echo=FALSE, fig.width=6, fig.height=6}
+```{r echo=FALSE, fig.width=3, fig.height=3, fig.caption="Distribution of DE genes per condition"}
 num_de = colSums(qvalues < 0.05)
@@
 
-```{r echo=FALSE, fig.height=8, fig.width=8, message=FALSE}
+```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
 par(mfrow=c(2, 2))
@@
 
-```{r fig.height=8, fig.width=8}
+```{r fig.height=4, fig.width=4}
 pvalue = qvalues[, "M-K"]
@@
 
-```{r fig.height=8, fig.width=8}
+```{r fig.height=6, fig.width=6}
 top_DE_genes_pval = names(sort(pvalue)[1:12])
@@
 
-```{r fig.height=8, fig.width=8}
+```{r fig.height=6, fig.width=6}
 top_DE_genes_lfc = names(
@@
 
-# Clustering of time-course data
+## Clustering of time-course data
 
@@
 ```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
-labels = unlist(labels[!is.na(labels)])
 genes_scored = names(labels)
@@
 
-## Looking at specific clusters in details.
+### Looking at specific clusters in details.
 
@@
 
-## How to choose the number of clusters.
+### How to choose the number of clusters.
 
@@
 
-### Elbow method
+#### Elbow method
 
@@
 
-### Average silhouette method
+#### Average silhouette method
 
@@
 
-### Looking at the stability of the clustering
+#### Looking at the stability of the clustering
 
@@
 
-#### The model explorer strategy
+##### The model explorer strategy
 
@@
 
-#### Consensus clustering as a way to find $k$
+##### Consensus clustering as a way to find $k$
 
@@
 
-# Downstream analysis of clusters.
+## Downstream analysis of clusters.
 

### 2019-05-11 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\documentclass[9pt,a4paper,]{extarticle}

\usepackage{f1000_styles}

\usepackage[pdfborder={0 0 0}]{hyperref}

\usepackage[numbers]{natbib}
\bibliographystyle{unsrtnat}

%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% disable code chunks background
%\renewenvironment{Shaded}{}{}

% disable section numbers
\setcounter{secnumdepth}{0}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\pagestyle{front}

\title{A pipeline to analyse time-course gene expression data}

%% AUTH AFFIL %%

\maketitle
\thispagestyle{front}

\begin{abstract}
The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology.However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software. Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflows provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) normalization of the micro-array dataset; (2) differential expression analysis using functional data analysis; (3) clustering fo time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
\end{abstract}

\section*{Keywords}
time-course gene expression data, clustering, differential expression, workflow

\clearpage
\pagestyle{main}

\#TODO list

\begin{itemize}
\tightlist
\item
  Add legends to the plots that don't have.
\item
  Currently have two functions ``plot\_genes'', ``plot\_centroids'', that do exactly
  the same thing. Find a better name for both of them
\end{itemize}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Gene expression studies provide simultaneous quantification of the level of
mRNA from all genes in a sample. High-throughput studies of gene expression
have a long history, starting with microarray technologies in the 1990s
through to single-cell technologies. While many expression studies are
designed to compare the gene expression in distinct groups, there is also a
long history of time-course expression studies, where the the gene expression
is compared across time by measuring mRNA levels from different samples across
time \footnote{Because the collection of the mRNA is often destructive, samples at
  different time points are generally from different biological samples;
  longitudinal studies, for example tracking the same subject over time, are
  certainly possible in certain settings, but not directly considered here.}. Such time course studies can vary from measuring a few distinct time
points, to sampling ten to twenty time points. Many longer time series are
particularly interested in investigating development over time. More recently,
single-cell studies track single cells through their development, and a single
cell is measured at a particular moment in its developmental progression -- a
value that is not know but estimated from the data as its ``pseudo-time.''

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being \texttt{edge}
\citep{storey:significance}, functional component analysis based models \citep{wu:more},
time-course permutation tests \citep{park:statistical}, and multiple testing
strategies to combine singe time point differential expression analysis
\citep{sun:multiple}. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show \emph{some} change. For
example, in the mice lung tissues infected with influenza, over 50\% of genes
are shown to be changing over time. The task in these settings is often not to
detect changes in genes, but to categorize into biologically interpretable
patterns the vast number of changes discovered.

We present here a workflow for such an analysis that consists of 4 main
parts (Figure \ref{fig:workflow}):

\begin{itemize}
\tightlist
\item
  Quality control and normalization;
\item
  Identification of genes that are differentially expressed;
\item
  Clustering of genes into distinct temporal patterns;
\item
  Biological interpretation of the clusters.
\end{itemize}

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. It relies on several standard packages for
analysing gene expression data, some specific for time-course data, others
broadly used by the community. We provide the various steps of the workflow as
functions in a R package called \texttt{moanin}.

\begin{figure}

{\centering \includegraphics[width=4.81in]{images/fig1_schemas}

}

\caption{Workflow for analyzing time-course datasets.}\label{fig:schema}
\end{figure}

\hypertarget{analysis-of-the-dynamical-response-of-mouse-lung-tissue-to-influenza}{%
\section{Analysis of the dynamical response of mouse lung tissue to influenza}\label{analysis-of-the-dynamical-response-of-mouse-lung-tissue-to-influenza}}

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to three different strains of influenza, and
collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
\citep{shoemaker:ultrasensitive}. The three strains of influenza used in the
study are (1) a low pathogenicity seasonal H1N1 influenza virus
(A/Kawasaki/UTK4/2009 {[}H1N1{]}), a mildly pathogenic virus from the 2009
pandemic season (A/California/04/2009 {[}H1N1{]}), and a highly pathogenic H5N1
avian influenza virus (A/Vietnam/1203/2004 {[}H5N1{]}. Mice were injected with
\(10^5\) PFU of each virus. An adidtional 42 mice were injected with a lower dose
of the Vietnam avian influenza virus (\(10^3\) PFU).

By combining gene expression time-course data with virus growth data, the
authors show that the inflammatory response of lung tissue is gated until a
threshold of the virus concentration is exceeded in the lung. Once this
threshold is exceeded, a strong inflammatory and cytokin production occurs.
This results provides evidence that the pathology response is non-linearly
regulated by virus concentration.

\hypertarget{quality-control-and-normalization}{%
\subsection{Quality control and normalization}\label{quality-control-and-normalization}}

\hypertarget{pre-processing-and-normalization}{%
\subsubsection{Pre-processing and normalization}\label{pre-processing-and-normalization}}

The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
with accession number GSE63786. The dataset has a total of 209 samples, with
45018 probes (1 sample was removed by the authors prior to uploading the
dataset on GEO).

Note that in the following, we assume that the user has access to a data
folder \texttt{data/shoemaker2015/rawdata}, containing the microarray data.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data_dir =}\StringTok{ "data/shoemaker2015/rawdata"}
\end{Highlighting}
\end{Shaded}

Given the source and a \texttt{Targets.txt} structured as follows, microarray can be loaded thanks to the limma library.

\begin{verbatim}
FileName
GSM1557140_AR0313_043raw.txt.gz
GSM1557141_AR0313_044raw.txt.gz
GSM1557142_AR0313_045raw.txt.gz
GSM1557143_AR0313_055raw.txt.gz
GSM1557144_AR0313_056raw.txt.gz
GSM1557145_AR0313_057raw.txt.gz
GSM1557146_AR0313_058raw.txt.gz
GSM1557147_AR0313_059raw.txt.gz
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{targets =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{readTargets}\NormalTok{()}
\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{read.maimages}\NormalTok{(}
\NormalTok{  targets, }\DataTypeTok{source=}\StringTok{"agilent"}\NormalTok{,}
  \DataTypeTok{path=}\NormalTok{data_dir,}
  \DataTypeTok{other.columns=}\StringTok{"gIsWellAboveBG"}\NormalTok{,}
  \DataTypeTok{green.only=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The normalization strategy we apply here relies on \texttt{limma} and is as follows:

\begin{itemize}
\tightlist
\item
  First, normalize probe intensity using background information using the
  ``norm-exponential'' strategy.
\item
  Then normalize across micro-arrrays using quantile normalization.
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Perform background correction using norm-exponential strategy and normalize}
\CommentTok{# across samples using quantile normalization.}
\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{backgroundCorrect}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"normexp"}\NormalTok{)}
\NormalTok{RG.q =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{normalizeBetweenArrays}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"quantile"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Plotting the smoothed empirical densities of the probe values before
normalization shows considerable variations betweens arrays. Once the data has
been normalized across array using quantile normalization, the smoothed
empirical densities of each sample perfectly overlap.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Plot the densities of the probes before and after normalization.}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{))}
\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
             \DataTypeTok{main=}\StringTok{"Before normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG.q, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
             \DataTypeTok{main=}\StringTok{"After normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/plotting_densities-1} \end{center}

We then clean up the names of each gene and each sample to match the original
data. Finally, we further filter out probes based on the following criteria:
(1) removing the 1486 control probes; (2) removing
4823 low-expressed genes.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sample_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(RG.q), }\StringTok{"_"}\NormalTok{), .subset2,  }\DecValTok{1}\NormalTok{)}
\KeywordTok{colnames}\NormalTok{(RG.q) =}\StringTok{ }\NormalTok{sample_names}
\KeywordTok{row.names}\NormalTok{(RG.q) =}\StringTok{ }\KeywordTok{make.names}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{SystematicName, }\DataTypeTok{unique=}\OtherTok{TRUE}\NormalTok{)}
\NormalTok{idx_to_keep =}\StringTok{ }\KeywordTok{which}\NormalTok{(}
  \OperatorTok{!}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{Systematic }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"DarkCorner"}\NormalTok{, }\StringTok{"GE_BrightCorner"}\NormalTok{,}
                   \StringTok{"NegativeControl"}\NormalTok{)))}

\NormalTok{data =}\StringTok{ }\NormalTok{RG.q[idx_to_keep,]}

\CommentTok{# Filter out control probes.}
\NormalTok{control_probes =}\StringTok{ }\NormalTok{data}\OperatorTok{$}\NormalTok{genes[, }\StringTok{"ControlType"}\NormalTok{] }\OperatorTok{==}\StringTok{ }\DecValTok{1}
\NormalTok{is_expressed =}\StringTok{ }\KeywordTok{rowSums}\NormalTok{(data}\OperatorTok{$}\NormalTok{other}\OperatorTok{$}\StringTok{"gIsWellAboveBG"} \OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{>=}\StringTok{ }\DecValTok{4}

\NormalTok{data =}\StringTok{ }\NormalTok{data[}\OperatorTok{!}\NormalTok{control_probes }\OperatorTok{&}\StringTok{ }\NormalTok{is_expressed, ]}\OperatorTok{$}\NormalTok{E}
\end{Highlighting}
\end{Shaded}

After filtering, we are left with 39544 probes of interest in
209 samples.

Now that we have normalized and filtered the gene expressions, let's load the
metadata. The package \texttt{moanin} contains the normalized data and meta of
\citep{shoemaker:ultrasensitive}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Now load in the metadata}
\KeywordTok{data}\NormalTok{(shoemaker2015)}
\NormalTok{meta =}\StringTok{ }\NormalTok{shoemaker2015}\OperatorTok{$}\NormalTok{meta}
\NormalTok{data =}\StringTok{ }\NormalTok{shoemaker2015}\OperatorTok{$}\NormalTok{data}
\end{Highlighting}
\end{Shaded}

\hypertarget{exploratory-analysis-and-quality-control}{%
\subsubsection{Exploratory analysis and quality control}\label{exploratory-analysis-and-quality-control}}

Plotting the smoothed empirical densities of probe values for each sample and
estimating the variations across samples before and after normalization is a
first quality control step. Typically, two other quality control and
exploratory analysis steps are also performed before and after normalization:
(1) low dimensionality embedding of the samples; (2) correlation plots between
each samples. In both cases, we expect a strong biological signal, while
replicate samples should be strongly clustered or correlated with one another.

Before performing any additional exploratory analysis, let us only keep highly
variable genes: we keep for this step only the top 50\% most variable genes.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{variance_cutoff =}\StringTok{ }\FloatTok{0.5}
\NormalTok{variance_per_genes =}\StringTok{ }\KeywordTok{apply}\NormalTok{(data, }\DecValTok{1}\NormalTok{, mad)}
\NormalTok{min_variance =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{quantile}\NormalTok{(variance_per_genes, }\KeywordTok{c}\NormalTok{(variance_cutoff))}
\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{data[variance_per_genes }\OperatorTok{>}\StringTok{ }\NormalTok{min_variance,]}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}

Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
centered and scaled gene expression data.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Reorder genes on condition, time, and replicate}

\NormalTok{pca_data =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
\NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We then plot the two first components, and color each sample by (1) its
condition; (2) its sampling time. We use different markers for each replicate.
We also plot the second and third components in the second row.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{graphics}\OperatorTok{::}\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{))}
\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}
\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}

\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/pca_plots-1} \end{center}

Then, we plot the pearson correlation across each samples. We order the
samples by their Group (treatment) and Timepoint (the time of sampling).
Additionally, in this example, we order each treatment by strength of the
pathogenicity of the treatment: Control, Kawasaki, California, low dose of
Vietnam, then high dose of Vietnam.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Reorder the conditions such that:}
\CommentTok{#   - Control is before any influenza treatment}
\CommentTok{#   - Each treatment is ordered from low to high pathogeny}
\NormalTok{meta}\OperatorTok{$}\NormalTok{Group =}\StringTok{ }\KeywordTok{factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group, }\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group)[}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)])}

\CommentTok{# Reorder genes on condition, time, and replicate}
\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Group,}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Timepoint,}
\NormalTok{  meta}\OperatorTok{$}\NormalTok{Replicate)}

\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
\NormalTok{data_corr =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}

\NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}

\CommentTok{# Copy the metadata before changing the timepoints into factors}
\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)}
\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    data_corr,}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annRow=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{, }
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\StringTok{"Correlation plot"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/correlation_plot-1} \end{center}

We can already see interesting patterns emerging from the correlation plot.
First, the cross-correlation amongst samples taking from the control mice is
higher than the cross correlation amongst the rest of the treatments. Second,
the influenza-infected mice midly react until time point 36. Third, the less
pathogenic the strain is, the closer the samples are to the control condition.
Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
most different from control samples.

\hypertarget{differential-expression-analysis-of-time-course-data}{%
\subsection{Differential expression analysis of time-course data}\label{differential-expression-analysis-of-time-course-data}}

The next step is typically to run a differential expression analysis. Most of
\texttt{moanin}'s function rely on an object containing a number of information such
as the metadata and the splines model used. We start by creating the
\texttt{splines\_model} object using the \texttt{create\_splines\_model} function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{splines_model =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{weekly-differential-expression-analysis}{%
\subsubsection{Weekly differential expression analysis}\label{weekly-differential-expression-analysis}}

\texttt{moanin} provides a simple interface to perform a time by time differential
expression analysis. Under the hood, it simply calls \texttt{limma} \citep{ritchie:limma} on the set of
contrasts provided. Here, we show an example between the control mouse and the
mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
(H5N1), but the function works with any form contrasts.

\textbf{FIXME rename the function to something better}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{meta}\OperatorTok{$}\NormalTok{WeeklyGroup =}\StringTok{ }\KeywordTok{make.names}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group}\OperatorTok{:}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint))}

\CommentTok{# Define contrast   }
\NormalTok{contrasts =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"M.0-VL.0"}\NormalTok{,}
          \StringTok{"M.3-VL.3"}\NormalTok{,}
          \StringTok{"M.6-VL.6"}\NormalTok{,}
          \StringTok{"M.9-VL.9"}\NormalTok{,}
          \StringTok{"M.12-VL.12"}\NormalTok{,}
          \StringTok{"M.18-VL.18"}\NormalTok{,}
          \StringTok{"M.24-VL.24"}\NormalTok{,}
          \StringTok{"M.30-VL.30"}\NormalTok{,}
          \StringTok{"M.36-VL.36"}\NormalTok{,}
          \StringTok{"M.48-VL.48"}\NormalTok{,}
          \StringTok{"M.60-VL.60"}\NormalTok{,}
          \StringTok{"M.72-VL.72"}\NormalTok{,}
          \StringTok{"M.120-VL.120"}\NormalTok{,}
          \StringTok{"M.168-VL.168"}\NormalTok{)}
\NormalTok{weekly_de_analysis =}\StringTok{  }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{weekly_differential_expression_analysis}\NormalTok{(}
\NormalTok{     data, splines_model, contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Let's look at the distribution of genes found differentially expressed per
week between control and each of the influenza strains.

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}

The distribution of number of genes found differentially expressed by
considering each time-point independantly highlights the challenges of such
approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
genes found differentially expressed than timepoints 3H and 9H: this is likely
due to biological or technical variances for specific genes at specific
timepoints.

As a summary, classic differential expression methods are appropriate for
unordered treatments, but fail to use the temporal structure of the data.

\hypertarget{time-course-differential-expression-analysis-between-two-groups}{%
\subsubsection{Time-course differential expression analysis between two groups}\label{time-course-differential-expression-analysis-between-two-groups}}

To leverage this temporal structure, Storey et al \citep{storey:significance} proposes to
model time-course micro-array with splines functions, and to use a log-ratio
likelihood test to detect differentially expressed genes. \texttt{moanin} extends
this idea by providing functions to perform time-course differential
expression analysis using \texttt{limma} contrasts.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Differential expression analysis}
\NormalTok{timecourse_contrasts =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"M-K"}\NormalTok{, }\StringTok{"M-C"}\NormalTok{, }\StringTok{"M-VL"}\NormalTok{, }\StringTok{"M-VH"}\NormalTok{)}

\NormalTok{contrasts =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{makeContrasts}\NormalTok{(}
    \DataTypeTok{contrasts=}\NormalTok{timecourse_contrasts,}
    \DataTypeTok{levels=}\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group))}

\CommentTok{# The function takes the data (data.frame or named matrix), the meta data}
\CommentTok{# (data.frame containing a timepoint and group column, the first corresponding}
\CommentTok{#<c2><a0>to the time-course information, the latter corresponding to the}
\CommentTok{# treatment).}
\NormalTok{pvalues =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{timecourse_differential_expression_analysis}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts)}
\NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, stats}\OperatorTok{::}\NormalTok{p.adjust)}
\end{Highlighting}
\end{Shaded}

The number of genes found differentially expressed ranges from around 12000 to
29000 depending on the strain and dosage of influenza virus given to the mice.
This corresponds to between 30\% to 70\% of the genes found differentially
expressed in this time-course experiment.

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}

The next step in a classical differential expression analysis is typically to
assess the effect of the treatment by looking at the log fold change.
Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then over-expressed at the end of the
experiment. As a result, \texttt{moanin} provides a number of possible ways to
compute the log fold change across the whole time-course.

First, \texttt{moanin} provides as simple interface to compute the log-fold change
for each individual timepoints.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timepoints =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabular}{l|r|r|r|r|r}
\hline
  & M-K:1 & M-K:7 & M-K:11 & M-K:14 & M-K:2\\
\hline
NM\_009912 & -0.0061424 & 0.0224948 & -0.0515947 & -0.1143973 & 0.0614619\\
\hline
NM\_008725 & 2.6547855 & 0.2734387 & -0.6882925 & 0.4718807 & -1.7463319\\
\hline
NM\_007473 & -0.1603623 & 0.1017655 & 0.5079573 & 0.0595773 & 0.4250185\\
\hline
ENSMUST00000094955 & 0.3505920 & 0.6463011 & 0.2564796 & 0.3599102 & 0.4177116\\
\hline
NM\_001042489 & 0.0889750 & 0.2321381 & -0.1891501 & 0.1786113 & 0.3320644\\
\hline
NM\_008159 & -0.3863770 & 0.0298395 & -0.0862755 & -0.3266351 & -0.1893062\\
\hline
\end{tabular}

This matrix can then be used to visualize the log-fold change for each
contrast per timepoint.

Sometimes, a single value per gene and per contrast is more useful. Here is a
table of the possible ways to compute log-fold change values.

\begin{longtable}[]{@{}lll@{}}
\toprule
Name & Formula &\tabularnewline
\midrule
\endhead
timely & \(\text{lfc}(t)\) & Function of time\tabularnewline
sum & \(\sum_t lfc(t)\) & Sum of log fold change.\tabularnewline
abs\_sum & \(\sum_t \|lfc(t)\|\) & Always positive\tabularnewline
max & \(\max_t \|lfc(t)\|\) & Always positive\tabularnewline
min & \(\min_t \|lfc(t)\|\) & Always positive\tabularnewline
epicon & \$\$ & Captures overall strength of response and overall direction\tabularnewline
\bottomrule
\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_epicon =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}

\NormalTok{log_fold_change_sum =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}

\NormalTok{log_fold_change_max =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}

\NormalTok{log_fold_change_min =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the EPICON log fold change
computation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pvalue =}\StringTok{ }\NormalTok{qvalues[, }\StringTok{"M-K"}\NormalTok{]}
\NormalTok{lfc_epicon =}\StringTok{ }\NormalTok{log_fold_change_epicon[, }\StringTok{"M-K"}\NormalTok{]}
\KeywordTok{names}\NormalTok{(lfc_epicon) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(log_fold_change_epicon)}

\KeywordTok{plot}\NormalTok{(lfc_epicon, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
     \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}

As another sanity check, \texttt{moanin} provides a simple utility function to
visualize gene time-course data. Here, we plot the 12 genes with the smallest
p-values.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
           \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}

And here we visualize the genes with the largest absolute EPICON log-fold
change.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_epicon),}
     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
           \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}

Thanks to those visualization, we can see that genes often follow similar
patterns of expression, although on a different scale for each gene. We can
leverage this observation to cluster the genes into groups of similar patterns
of transcriptomic response.

\hypertarget{clustering-of-time-course-data}{%
\subsection{Clustering of time-course data}\label{clustering-of-time-course-data}}

The very large number of genes found differentially expressed impair any
interpretation one would attempt: with 70\% of the genome found differentially
expressed, all pathways are affected by the treatment. Hence the next step of
the workflow to cluster gene expression according to their dynamical response
to the treatment. Before clustering the genes, we first reduce the set of
genes of interest to genes that are (1) significantly found differentially
expressed; (2) ``highly'' differentially expressed. To do this, we first
aggregate all p-values obtained during the time-course differential expresison
step in a single p-value using Fisher's method \citep{fisher:statistical}. Then we
select all the genes which have a Fisher adjusted p-value below 0.05 and a log
fold change of at least two between at least one condition and one time-point.
Reducing the set of genes on which to perform the clustering allows to
estimate the centroids with more stability.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Then rank by fisher's p-value and take max the number of genes of interest}
\CommentTok{# Filter out q-values for the pvalues table}
\NormalTok{fishers_pval =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
\NormalTok{fishers_qval =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}

\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
\NormalTok{    log_fold_change_max[}
\NormalTok{    (}\KeywordTok{rowSums}\NormalTok{(log_fold_change_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}
\StringTok{    }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
\CommentTok{# Keep the data corresponding to the genes of interest in another variable.}
\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}
\end{Highlighting}
\end{Shaded}

After filtering, we are left with 3950 genes. We can then apply a
clustering. As observed by looking at genes found differentially expressed,
many genes share a similar gene expression pattern, but on different scale.
We thus propose the following adaptation of k-means:

\begin{itemize}
\tightlist
\item
  \textbf{Splines estimation}: for each gene, fit the splines function with the basis
  of your choice.
\item
  \textbf{Rescaling splines}: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1.
\item
  \textbf{K-means}: apply k-means on the rescaled fitted splines to estimate the
  centroids.
\item
  \textbf{Assign scores and labels to all genes}: then assign a score and a label
  to all gene based on a goodness-of-fit measure on the raw data. By default,
  the \texttt{splines\_kmeans\_score\_and\_label} function only labels the best 50\% of
  genes.
\end{itemize}

The first three steps are performed jointly by the \texttt{splines\_kmeans} function.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# First fit the kmeans clusters}
\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
\NormalTok{    y, splines_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
    \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
    \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We then use the \texttt{plot\_centroids} function to visualize the centroids obtained
with the splines k-means model.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_centroids}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
               \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
               \DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}

The scoring and labeling can be done via the \texttt{splines\_kmeans\_score\_and\_label}
function.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{#<c2><a0>Then assign scores and labels to all the data, using a goodness-of-fit}
\CommentTok{# scoring function.}
\NormalTok{scores_and_labels =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
\NormalTok{    data, kmeans_clusters)}
\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}

\CommentTok{# Let's keep only the list of genes that have a label.}
\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
\end{Highlighting}
\end{Shaded}

Before performing the next steps, let us investigate in more detail the
differences between the labels provided by the splines k-means model and the
scoring and labelinig step.

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}

The scoring and label step allows to label genes that were removed during the
filtering step, yet are good matches to the centroids found during the
clustering.

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}

\hypertarget{looking-at-specific-clusters-in-details.}{%
\subsubsection{Looking at specific clusters in details.}\label{looking-at-specific-clusters-in-details.}}

Now, let us look more in detail some specific clusters. Cluster 8 seems
particularly interesting: it captures genes with strong differences between
the differant influenza treatments and the control, while the control remains
relatively flat.

Heatmaps are useful to investigate the range of expression patterns for
specific genes. Here, we are going to plot heatmaps of the normalized gene
expression patterns and the rescaled gene expression patterns side by side.

First, select the genes of interest.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster_to_plot =}\StringTok{ }\DecValTok{8}
\NormalTok{genes_to_plot =}\StringTok{  }\KeywordTok{names}\NormalTok{(labels[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_to_plot])}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{layout}\NormalTok{(}\KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{),}\DataTypeTok{nrow=}\DecValTok{1}\NormalTok{), }\DataTypeTok{widths=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}

\NormalTok{data_to_plot =}\StringTok{ }\NormalTok{data[genes_to_plot, ]}
\NormalTok{submeta =}\StringTok{ }\NormalTok{meta}
\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(}
\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Group,}
\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Timepoint,}
\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Replicate)}
\NormalTok{submeta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(submeta}\OperatorTok{$}\NormalTok{Timepoint)}

\NormalTok{data_to_plot =}\StringTok{ }\NormalTok{data_to_plot[, ord]}
\NormalTok{submeta =}\StringTok{ }\NormalTok{submeta[ord, ]}

\NormalTok{res =}\StringTok{ }\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    data_to_plot,}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{color=}\StringTok{"YlGnBu"}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{submeta[,}
                \KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"Cluster"}\NormalTok{, cluster_to_plot, }\StringTok{"(raw)"}\NormalTok{),}
    \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{)}

\CommentTok{# Now use the results of the previous call to aheatmap to reorder the genes.}
\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
\NormalTok{    moanin}\OperatorTok{:::}\KeywordTok{rescale_values}\NormalTok{(data_to_plot)[res}\OperatorTok{$}\NormalTok{rowInd,],}
    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
    \DataTypeTok{annCol=}\NormalTok{submeta[,}
                \KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
    \DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"Cluster"}\NormalTok{, cluster_to_plot, }\StringTok{"(rescaled)"}\NormalTok{),}
    \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}

Those two heatmaps demonstrate that the clustering method successfully cluster
genes that are on different scales, and yet share the same dynamical response
to the treatments.

\hypertarget{how-to-choose-the-number-of-clusters.}{%
\subsubsection{How to choose the number of clusters.}\label{how-to-choose-the-number-of-clusters.}}

A common question that arises when performing clustering is how to choose the
number of clusters. A choice for the number of clusters K depends on the goal.
In this particular case, the end goal is not the clustering, but to facilitate
interpretation of the differential expression analysis step. As a result, the
number of clusters should not exceed the number of gene sets the user wants to
interpret. This allows to set a maximum number of clusters. Let us assume here
that this number is 20 clusters.

Once the maximum number of clusters is set, several strategies allow to
identify the number of clusters:

\begin{itemize}
\tightlist
\item
  \textbf{Elbow method}. First introduced in 1953 by Thorndike \citep{thorndike:who},
  the elbow methods looks at the total with cluster sum of squares
  as a function of the number of clusters (WCSS). When adding clusters
  doesn't decrease the WCSS by a sufficient
  amount, one can consider stopping. This method thus provides visual aid to
  the user to choose the number of clusters, but often the ``elbow'' is hard to
  see on real data, where the number of clusters is not clearly defined.
\item
  \textbf{Silhouette method}. Similarly to the Elbow method, the Silhouette method
  refers to a method of validation of consistency within clusters, and
  provides visual aid to choose the number of clusters.
\item
  \textbf{Stability methods} Stability methods are more computationally intensive
  than any other method, as they rely on assessing the stability of the
  clustering for every \(k\) to a small randomization of the data. The user is
  then invited to choose the number of cluster based on a number of similarity
  measures.
\end{itemize}

First, let us run the clustering for all possible clusters of interest. We
will, for each clustering experiment, conserve (1) with within cluster sum of
squares; (2) the labels assigned to all genes.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{all_possible_n_clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{9}\NormalTok{, }\DecValTok{11}\NormalTok{, }\DecValTok{15}\NormalTok{)}
\NormalTok{all_clustering =}\StringTok{ }\KeywordTok{list}\NormalTok{()}
\NormalTok{wss_values =}\StringTok{ }\KeywordTok{list}\NormalTok{()}

\NormalTok{i =}\StringTok{ }\DecValTok{1}
\ControlFlowTok{for}\NormalTok{(n_cluster }\ControlFlowTok{in}\NormalTok{ all_possible_n_clusters)\{}
\NormalTok{    clustering_results =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
\NormalTok{    y, splines_model,}
    \DataTypeTok{n_clusters=}\NormalTok{n_cluster, }\DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
    \DataTypeTok{n_init=}\DecValTok{10}\NormalTok{)}
\NormalTok{    wss_values[i] =}\StringTok{ }\KeywordTok{sum}\NormalTok{(clustering_results}\OperatorTok{$}\NormalTok{WCSS_per_cluster)}
\NormalTok{    all_clustering[[i]] =}\StringTok{ }\NormalTok{clustering_results}\OperatorTok{$}\NormalTok{clusters}
\NormalTok{    i =}\StringTok{ }\NormalTok{i }\OperatorTok{+}\StringTok{ }\DecValTok{1}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{elbow-method}{%
\paragraph{Elbow method}\label{elbow-method}}

The Elbow method to choose the number of clusters relies on visualization aid
to choose the number of cluster. The method relies on plotting the within
cluster sum of squares as a function of the number of clusters. At some point,
the WCSS will stop dropping, giving an angle in the graph. The number of
cluster is chosen at this ``Elbow point.''

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot}\NormalTok{(all_possible_n_clusters, wss_values,}
     \DataTypeTok{type=}\StringTok{"b"}\NormalTok{, }\DataTypeTok{pch=}\DecValTok{19}\NormalTok{, }\DataTypeTok{frame=}\OtherTok{FALSE}\NormalTok{, }
     \DataTypeTok{xlab=}\StringTok{"Number of clusters K"}\NormalTok{,}
     \DataTypeTok{ylab=}\StringTok{"Total within-clusters sum of squares"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/elbow_method-1} \end{center}

\hypertarget{average-silhouette-method}{%
\paragraph{Average silhouette method}\label{average-silhouette-method}}

The silhouette value is a measure of how similar a data point is to its own
cluster (cohesion) compared to other clusters (separation).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# function to compute average silhouette for k clusters}
\NormalTok{average_silhouette =}\StringTok{ }\ControlFlowTok{function}\NormalTok{(labels, y) \{}
\NormalTok{    silhouette_results =}\StringTok{ }\NormalTok{cluster}\OperatorTok{::}\KeywordTok{silhouette}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(labels[}\DecValTok{1}\NormalTok{]), }\KeywordTok{dist}\NormalTok{(y))}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{mean}\NormalTok{(silhouette_results[, }\DecValTok{3}\NormalTok{]))}
\NormalTok{\}}

\CommentTok{# extract the average silhouette}
\NormalTok{average_silhouette_values =}\StringTok{ }\KeywordTok{list}\NormalTok{()}
\NormalTok{i =}\StringTok{ }\DecValTok{1}
\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(all_clustering))\{}
\NormalTok{    clustering_results =}\StringTok{ }\NormalTok{all_clustering[i]}
\NormalTok{    average_silhouette_values[i] =}\StringTok{ }\KeywordTok{average_silhouette}\NormalTok{(clustering_results, y)}
\NormalTok{    i =}\StringTok{ }\NormalTok{i }\OperatorTok{+}\StringTok{ }\DecValTok{1}
\NormalTok{\}}

\KeywordTok{plot}\NormalTok{(all_possible_n_clusters, average_silhouette_values,}
     \DataTypeTok{type=}\StringTok{"b"}\NormalTok{, }\DataTypeTok{pch=}\DecValTok{19}\NormalTok{, }\DataTypeTok{frame=}\OtherTok{FALSE}\NormalTok{,}
     \DataTypeTok{xlab=}\StringTok{"Number of clusters K"}\NormalTok{,}
     \DataTypeTok{ylab=}\StringTok{"Average Silhouettes"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/average_silhouette_method-1} \end{center}

\hypertarget{looking-at-the-stability-of-the-clustering}{%
\paragraph{Looking at the stability of the clustering}\label{looking-at-the-stability-of-the-clustering}}

On real data, the number of clusters is not only unknown but also
ambiguous: it will depend on the desired clustering resolution of the user.
Yet, in the case of biological data, stability and reproducibility of the
results is necessary to ensure that the biological interpretation of the
results hold when the data or the model is exposed to reasonable
perturbations.

Methods that rely on the stability of the clustering results to choose \(k\)
thus ensure that the biological interpretation of the clusters hold with
perturbtation to the data. In addition, simulation where the data is generated
with a well defined \(k\) show that the clustering is more stable for the
correct of number of the clusters.

Most methods method to find the number of clusters with stability measures
only provide visual aids to guide the user. The first element often visualized
is the consensus matrix: the consensus matrix is an \(n \times n\) matrix that
stores the proportion of clustering in which two items are clustered together.
A perfect consensus matrix ordered such as each elements that belong to the
same cluster are adjacent to one another which show blocks along the diagonal
close to 1.

To perform such analysis, the first step is run the clustering several times
on a resampled dataset--either using bootstrap or subsampling.

Using the bootstrapping strategy:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{]}
\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}

\NormalTok{bootstrapped_y =}\StringTok{ }\NormalTok{y[indices, ]}
\end{Highlighting}
\end{Shaded}

Using the subsampling strategy, keeping 80\% of the genes:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{subsample_proportion =}\StringTok{ }\DecValTok{1}
\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
\NormalTok{subsampled_y =}\StringTok{ }\NormalTok{y[indices, ]}
\end{Highlighting}
\end{Shaded}

Here we plot to stability matrix of the top 1000 genes for \(k=5\) and \(k=20\).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_5.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
\NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
                            \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{5}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
          \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
          \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_20.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
\NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
                             \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{20}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
          \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
          \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-2} \end{center}

\hypertarget{the-model-explorer-strategy}{%
\subparagraph{The model explorer strategy}\label{the-model-explorer-strategy}}

The model explorer algorithm \citep{ben-hur:stability} proposes to estimate the
number of clusters exploiting the observation that if the number of clusters
is correct, the clustering results are stable to bootstrapping. The
distribution of similaries between bootstrapped results for each \(k\) can thus
be compared for different values of \(k\) and guide the user in the choice of
number of clusters.

The model explorer strategy works as follows. First, choose a similarity
measure between two partitions or clusters \(S(C_1, C_2)\). Examples are the
normalized mutual information or Fowlkes-Mallows. Then perform \(n\) bootstrap
experiments to estimate the cluster centroids, followed by a step of assigning
a label to all data points. Finally, compute the pairwise similarity measure
between all bootstrapped partition, and plot the cumulative density of the
obtained scores.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_model_explorer}\NormalTok{(all_labels)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}

From this plot, we can deduce that \(k=5\) is more stable than \(k=3\) and \(k=4\),
but not as stable as \(k=2\). The model explorer strategy, in addition to
visualizing the diversity of the centroids, can thus help assessing an
adequate number of clusters.

Now, replot the same model explorer, but only for the clustering experiments
\(k=6\), \(k=7\), \(k=8\), \(k=9\) and \(k=10\) so that we can see more clearly the stability
measures in that range.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"C6"}\NormalTok{, }\StringTok{"C7"}\NormalTok{, }\StringTok{"C8"}\NormalTok{, }\StringTok{"C9"}\NormalTok{, }\StringTok{"C10"}\NormalTok{)}
\NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_model_explorer}\NormalTok{(selected_labels)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}

\hypertarget{consensus-clustering-as-a-way-to-find-k}{%
\subparagraph{\texorpdfstring{Consensus clustering as a way to find \(k\)}{Consensus clustering as a way to find k}}\label{consensus-clustering-as-a-way-to-find-k}}

The consensus clustering \citep{monti:consensus} relies on a similar idea but
instead of looking at the cumulative density of similarity measures of
bootstrapped clustering, the authors suggests plotting the cumulative density
of elements of the consensus matrix.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_cdf_consensus}\NormalTok{(all_labels)}
\end{Highlighting}
\end{Shaded}

\begin{center}\includegraphics[width=6.67in]{images/clustering_CDF_consensus} \end{center}

The stability of the clustering based on the consensus matrix can then be
measured via a single number by looking at the area under the curve: the more
stable the clustering, the closer to 0 or 1 will be the entries of consensus
matrix. The consensus clustering strategy thus suggest at looking at either
the AUC as a function of the number of the clusters or the ``improvement'' in
the AUC as a function of the number of cluster.

\begin{center}\includegraphics[width=6.67in]{images/clustering_AUC_consensus} \end{center}

\begin{center}\includegraphics[width=6.67in]{images/clustering_delta_AUC_consensus} \end{center}

The consensus clustering method suggest that the most stable is \(k=2\), which
separates over-expressed genes from under-expressed genes. While it is indeed
a very stable clustering, it does not capture the range of gene expression
patterns present in the data. This shows the limitation of such method on real
data, where the number of clusters is not clearly defined.

\hypertarget{downstream-analysis-of-clusters.}{%
\subsection{Downstream analysis of clusters.}\label{downstream-analysis-of-clusters.}}

Once good clusters are obtained, the next step is to leverage the clustering
to ease interpretation. Classic enrichment analysis step can then be performed
in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
term enrichment analysis, motif enrichment analysis, etc.

First, let us clean up the gene obtained and only select the genes we are
going to use in the enrichment analysis. One can either use the whole set of
genes, only the set of differentially expressed genes in each cluster, or a
subset of genes that fit well to a cluster (based on some criterion).

\hypertarget{finding-enriched-pathways-using-biomart-and-keggprofile}{%
\subsubsection{\texorpdfstring{Finding enriched pathways using \texttt{biomaRt} and \texttt{KEGGprofile}}{Finding enriched pathways using biomaRt and KEGGprofile}}\label{finding-enriched-pathways-using-biomart-and-keggprofile}}

Let us first tackle the case of pathway enrichment analysis. We will leverage
the packages \texttt{biomaRt} \citep{durinck:biomart} and \texttt{KEGGprofile} \citep{zhao:keggprofile}
for this step. \texttt{KEGGprofile} is a
package that easily allows to perform pathway enrichment analysis on a set of
genes labeled with the ensembl annotation. We thus need to convert the gene
names into the appropriate format. This is where \texttt{biomaRt} comes in handy: it
enables easy conversion from one gene annotation to another. Here, we will use
it to convert the gene names from the Refseq annotation to the ensembl one.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ensembl =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{useMart}\NormalTok{(}\StringTok{"ensembl"}\NormalTok{)}
\NormalTok{ensembl =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{useDataset}\NormalTok{(}\StringTok{"mmusculus_gene_ensembl"}\NormalTok{, }\DataTypeTok{mart=}\NormalTok{ensembl)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cluster =}\StringTok{ }\DecValTok{8}
\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels)}
\NormalTok{gene_names =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels)}
\NormalTok{genes =}\StringTok{ }\NormalTok{gene_names[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster]}

\CommentTok{# convert gene names}
\NormalTok{genes =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene"}\NormalTok{),}
        \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene"}\NormalTok{]}
\NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
\NormalTok{pathways =}\StringTok{ }\NormalTok{KEGGprofile}\OperatorTok{::}\KeywordTok{find_enriched_pathway}\NormalTok{(}
\NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
    \DataTypeTok{download_latest=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\hline
  & Pathway & Percentage & Adj. p-value\\
\hline
05169 & Epstein-Barr virus infection & 37 & 0\\
\hline
04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
\hline
05164 & Influenza A & 38 & 0\\
\hline
04668 & TNF signaling pathway & 45 & 0\\
\hline
04621 & NOD-like receptor signaling pathway & 34 & 0\\
\hline
05162 & Measles & 37 & 0\\
\hline
04630 & JAK-STAT signaling pathway & 33 & 0\\
\hline
05160 & Hepatitis C & 34 & 0\\
\hline
04620 & Toll-like receptor signaling pathway & 40 & 0\\
\hline
05167 & Kaposi sarcoma-associated herpesvirus infection & 29 & 0\\
\hline
\end{tabu}

\hypertarget{finding-enriched-go-terms}{%
\subsubsection{Finding enriched GO terms}\label{finding-enriched-go-terms}}

To find GO terms, we use \texttt{biomaRt} to find the mapping between GO terms and
gene mapping. The GO enrichment library \texttt{topGO} \citep{alexa:topgo} expects the GO term to gene
mapping to be a list where each item is a mapping between a gene name and a GO
term ID vector.

\begin{verbatim}
$NM_199153
[1] "GO:0016020" "GO:0016021" "GO:0007186" "GO:0004930" "GO:0007165"
[6] "GO:0050896" "GO:0050909"

$NM_201361
 [1] "GO:0016020" "GO:0016021" "GO:0003674" "GO:0008150" "GO:0005794"
 [6] "GO:0005829" "GO:0005737" "GO:0005856" "GO:0005874" "GO:0005739"
[11] "GO:0005819" "GO:0000922" "GO:0072686"
\end{verbatim}

\texttt{biomaRt} queries results a matrix with two named columns of gene names and GO
term ID.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{genes =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"go_id"}\NormalTok{, }\StringTok{"refseq_mrna"}\NormalTok{),}
          \DataTypeTok{values=}\NormalTok{gene_names,}
          \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{,}
          \DataTypeTok{mart=}\NormalTok{ensembl)}

\CommentTok{# Create gene to GO id mapping}
\NormalTok{gene_id_go_mapping =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_go_term_mapping}\NormalTok{(genes)}
\end{Highlighting}
\end{Shaded}

Once the gene ID to GO mapping list is created, \texttt{moanin} provides a simple
interface to \texttt{topGO} to fetch enriched GO terms. Here, we show an example of
running a GOterm enrichment on the ``Biological process'' ontology (BP).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{assignments =}\StringTok{ }\NormalTok{labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster}

\NormalTok{go_terms_enriched =}\StringTok{ }\KeywordTok{find_enriched_go_terms}\NormalTok{(}
\NormalTok{    assignments,}
\NormalTok{    gene_id_go_mapping, }\DataTypeTok{ontology=}\StringTok{"BP"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedright}X>{\raggedleft}X}
\hline
  & GO ID & Description & Annotated & Significant & Expected & P-value & adj. p-value\\
\hline
5 & GO:0003002 & regionalization & 172 & 153 & 134.46 & 0.00024 & 0.2416\\
\hline
6 & GO:0048562 & embryonic organ morphogenesis & 154 & 137 & 120.39 & 0.00035 & 0.2936\\
\hline
7 & GO:0010970 & transport along microtubule & 81 & 74 & 63.32 & 0.00057 & 0.4098\\
\hline
8 & GO:0060173 & limb development & 92 & 84 & 71.92 & 0.00070 & 0.4404\\
\hline
9 & GO:0060070 & canonical Wnt signaling pathway & 147 & 130 & 114.92 & 0.00090 & 0.5033\\
\hline
10 & GO:0055114 & oxidation-reduction process & 531 & 457 & 415.10 & 0.00135 & 0.6223\\
\hline
11 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.43 & 0.00136 & 0.6223\\
\hline
12 & GO:0006805 & xenobiotic metabolic process & 44 & 42 & 34.40 & 0.00168 & 0.7046\\
\hline
13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.75 & 0.00224 & 0.8197\\
\hline
14 & GO:0010811 & positive regulation of cell-substrate ad... & 78 & 71 & 60.98 & 0.00228 & 0.8197\\
\hline
15 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.80 & 0.00298 & 0.9999\\
\hline
16 & GO:0051056 & regulation of small GTPase mediated sign... & 151 & 131 & 118.04 & 0.00349 & 1.0000\\
\hline
17 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
\hline
18 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
\hline
19 & GO:0007368 & determination of left/right symmetry & 67 & 61 & 52.38 & 0.00464 & 1.0000\\
\hline
\end{tabu}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This workflow provides a tutoral for the analysis of time-course gene
expression data in R, illustrated through the analysis of mice lung tissue
exposed to different influenza strain. It covers four main steps; (1) quality
control and normalization; (2) differential expression analysis; (3)
clustering of time-course gene expression data; (4) downstream analysis of
clusters.

\hypertarget{software-and-data-availability}{%
\section{Software and data availability}\label{software-and-data-availability}}

The source code for this workflow can be found at \url{https://github.com/NelleV/2019timecourse-rnaseq-pipeline}.
Data used in this workflow are available from NCBI GEO, accession GSE95601.

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

NV and EP wrote the workflow.

\hypertarget{competing-interests}{%
\section{Competing interests}\label{competing-interests}}

The authors declare that they have no competing interests.

\hypertarget{grant-information}{%
\section{Grant information}\label{grant-information}}

This research was funded in part by a Department of Energy (DOE) grant
(DE-SC0014081); by the Gordon and Betty Moore Foundation (Grant GBMF3834) and
the Alfred P. Sloan Foundation (Grant 2013-10-27) to the University of
California, Berkeley {[}N.V.{]}; by a ENS-CFM Data Science Chair {[}E.P.{]}.

\emph{I confirm that the funders had no role in study design, data collection and
analysis, decision to publish, or preparation of the manuscript.}

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

The authors thank Karthik Ram and the Ropensci community for valuable
feedback.

\renewcommand\refname{References}
{\small\bibliography{bibliography.bib}}

\end{document}

### 2019-05-11 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+\hypertarget{session-information}{%
+\section{Session information}\label{session-information}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{sessionInfo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## R version 3.6.0 (2019-04-26)
+## Platform: x86_64-pc-linux-gnu (64-bit)
+## Running under: Ubuntu 19.04
+## 
+## Matrix products: default
+## BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.8.0
+## LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.8.0
+## 
+## locale:
+## [1] C
+## 
+## attached base packages:
+##  [1] stats4    parallel  splines   stats     graphics  grDevices utils    
+##  [8] datasets  methods   base     
+## 
+## other attached packages:
+##  [1] RColorBrewer_1.1-2   moanin_0.0.0         topGO_2.32.0        
+##  [4] SparseM_1.77         GO.db_3.6.0          AnnotationDbi_1.42.1
+##  [7] IRanges_2.14.12      S4Vectors_0.18.3     graph_1.58.2        
+## [10] viridis_0.5.1        viridisLite_0.3.0    KEGGprofile_1.22.0  
+## [13] RCurl_1.95-4.12      bitops_1.0-6         NMF_0.21.0          
+## [16] Biobase_2.40.0       BiocGenerics_0.26.0  cluster_2.0.9       
+## [19] rngtools_1.3.1.1     pkgmaker_0.27        registry_0.5-1      
+## [22] ggplot2_3.1.1        biomaRt_2.36.1       BiocStyle_2.8.2     
+## [25] kableExtra_1.1.0     knitr_1.22           limma_3.36.5        
+## [28] usethis_1.5.0        devtools_2.0.2       rmarkdown_1.12      
+## 
+## loaded via a namespace (and not attached):
+##  [1] colorspace_1.4-1        rprojroot_1.3-2        
+##  [3] XVector_0.20.0          fs_1.3.1               
+##  [5] rstudioapi_0.10         remotes_2.0.4          
+##  [7] bit64_0.9-7             ClusterR_1.1.9         
+##  [9] KEGG.db_3.2.3           xml2_1.2.0             
+## [11] codetools_0.2-16        doParallel_1.0.14      
+## [13] pkgload_1.0.2           ade4_1.7-13            
+## [15] gridBase_0.4-7          png_0.1-7              
+## [17] FD_1.0-12               readr_1.3.1            
+## [19] compiler_3.6.0          httr_1.4.0             
+## [21] backports_1.1.4         Matrix_1.2-17          
+## [23] assertthat_0.2.1        lazyeval_0.2.2         
+## [25] cli_1.1.0               htmltools_0.3.6        
+## [27] prettyunits_1.0.2       tools_3.6.0            
+## [29] gmp_0.5-13.5            gtable_0.3.0           
+## [31] glue_1.3.1              reshape2_1.4.3         
+## [33] dplyr_0.8.0.1           BiocWorkflowTools_1.6.2
+## [35] Rcpp_1.0.1              Biostrings_2.48.0      
+## [37] NMI_2.0                 ape_5.3                
+## [39] nlme_3.1-139            iterators_1.0.10       
+## [41] xfun_0.6                stringr_1.4.0          
+## [43] ps_1.3.0                testthat_2.1.1         
+## [45] rvest_0.3.3             gtools_3.8.1           
+## [47] XML_3.98-1.19           zlibbioc_1.26.0        
+## [49] MASS_7.3-51.4           scales_1.0.0           
+## [51] hms_0.4.2               curl_3.3               
+## [53] yaml_2.2.0              memoise_1.1.0          
+## [55] gridExtra_2.3           TeachingDemos_2.10     
+## [57] stringi_1.4.3           RSQLite_2.1.1          
+## [59] desc_1.2.0              foreach_1.4.4          
+## [61] permute_0.9-5           pkgbuild_1.0.3         
+## [63] bibtex_0.4.2            geometry_0.4.1         
+## [65] rlang_0.3.4             pkgconfig_2.0.2        
+## [67] matrixStats_0.54.0      evaluate_0.13          
+## [69] lattice_0.20-38         purrr_0.3.2            
+## [71] bit_1.1-14              processx_3.3.1         
+## [73] tidyselect_0.2.5        plyr_1.8.4             
+## [75] magrittr_1.5            bookdown_0.9           
+## [77] R6_2.4.0                DBI_1.0.0              
+## [79] mgcv_1.8-28             pillar_1.3.1           
+## [81] withr_2.1.2             abind_1.4-5            
+## [83] KEGGREST_1.20.2         tibble_2.1.1           
+## [85] crayon_1.3.4            progress_1.2.0         
+## [87] grid_3.6.0              blob_1.1.1             
+## [89] callr_3.2.0             git2r_0.25.2           
+## [91] vegan_2.5-4             digest_0.6.18          
+## [93] webshot_0.5.1           xtable_1.8-4           
+## [95] munsell_0.5.0           magic_1.5-9            
+## [97] sessioninfo_1.1.1
+\end{verbatim}
+
 \hypertarget{conclusion}{%

### 2019-05-13 doc/manuscript/paper.tex
update text

@@
 
-While there are many methods that have been proposed for discrete aspects of time course data, the entire workflow for analysis of such data remains difficult, particularly for long, developmental time series.
+While there are many methods that have been proposed for discrete components of time course data, the entire workflow for analysis of such data remains difficult, particularly for long, developmental time series.
 Most methods proposed for time course data are concerned with detecting genes that are changing over time (differential expression analysis), examples being \texttt{edge}\todo[Other methods to mention: impulse models, what else?
@@
 The task in these settings is often not to detect changes in genes, but to categorize into biologically interpretable patterns the vast number of changes discovered.
+In this context, the clustering of genes and interpretation of the genes can be far more relevant, and the
 
@@
 This workflow represents an integration of both novel implementations of previously established methods and new methodologies for the settings of developmental time series.
-We provide the various steps of the workflow as functions in a R package, \citettt{\xxx}, along with command-line tools for running the entire workflow in a single command.
+We provide the various steps of the workflow as functions in a R package, \citettt{\xxx}, along with command-line tools for running the entire workflow.
 
@@
 
-Several techniques for mRNA-expression analysis were
+A vast number of techniques Several techniques for mRNA-expression analysis
 

### 2019-05-13 scripts/miscellaneous/supp_scores.Rmd

---
title: "Clustering: A pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
---

```{r echo=FALSE, results="hide", message=FALSE}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(moanin)
library(NMF)
source("utils.R")
# library(moanin)
```

```{r echo=FALSE, results="hide"}
# Set options
df = 6

# Preprocessing & filtering
is_count = FALSE
take_log = FALSE    
filter_expression = FALSE

###############################################################################
# Clustering

# Whether to filter genes
filter_genes = TRUE
keep_all_significant = FALSE
n_genes_to_keep = 5000
n_clusters = 20
percentage_genes_to_label = 0.5
```

```{r echo=FALSE, results="hide"}
# Set data files and options
data("shoemaker2015")

data = shoemaker2015$data
meta = shoemaker2015$meta

data = read.table(".results/quantile_normalized.txt")

# Right now, the package doesn't contain the normalized data
de_analysis = read.table(".results/pvalues.txt", sep="\t", check.names=FALSE)

# Extract pvalues and log_fold change
lfc_col_to_keep = colnames(de_analysis)[
    grepl("-lfc_max", colnames(de_analysis))]
lfc_max = de_analysis[, lfc_col_to_keep]

pval_col_to_keep = colnames(de_analysis)[
    grepl("-pval", colnames(de_analysis))]
pvalues = de_analysis[, pval_col_to_keep]

splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
```

# Clustering of time-course data

The next step is to perform clustering. First, we reduce the set of genes of
interest to genes that are (1) significant; (2) "highly" differentially
expressed. First, we select the genes that have a log-fold change of at least
2 across at least one experiment for one time point. We thus use the object
`lfc_max`.

XXX

```{r filter_genes}
if(filter_genes){
    # First, filter out any genes that doesn't have a log fold change of at
    # least 2 at at least one of the time points.

    genes_to_keep = row.names(lfc_max[rowSums(lfc_max > 2) > 0, ])
    pvalues = pvalues[genes_to_keep, ]

    if(length(genes_to_keep) > n_genes_to_keep){

	# Then rank by fisher's p-value and take max the number of genes of
	# interest
	# Filter out q-values for the pvalues table
	fishers_pval = moanin:::fisher_method(pvalues)
	fishers_qval = stats::p.adjust(fishers_pval)
        if(keep_all_significant){
	   genes_to_keep = names(fishers_qval[fishers_qval < 0.05])
	}else{
	    genes_to_keep = names(sort(fishers_pval)[1:n_genes_to_keep])
	}
    }

    y = as.matrix(data[genes_to_keep, ])
}else{
    y = as.matrix(data)
}
```

After filtering, we consider `r dim(y)[1]` genes for the clustering.

```{r clustering}

#############################################################################
# Fit the splines

kmeans_clusters = moanin:::splines_kmeans(
    y, splines_model, n_clusters=20,
    random_seed=42,
    n_init=20)

# XXX is there a bug here?
all_scores = data.frame(row.names=row.names(data))
for(k in 1:20){
     scores = moanin:::score_genes_centroid(data, kmeans_clusters$centroids[k,])
     all_scores[paste0("cluster_", k)] = scores
 }

# Only assign labels to X% of the genes
max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
labels = moanin:::row_argmin(all_scores)
labels[genes_to_not_consider] = NA
```

```{r, fig.width=4, fig.height=4}
hist(all_scores[, 1], col="black")
hist(all_scores[, 2], col="black")

```

```{r}
table(kmeans_clusters$clusters)
```

```{r}
table(labels)
```

### 2019-05-13 scripts/manuscript.Rmd
Disabling table
changes to manuscript

@@
 
+The first steps of analysis of gene expression data is always to do normalization and quality control checks of the data.
+In what follows, we show an example of this for the influenza data using generic methods; these steps are not in specific to time course, but could be done for any gene expression analysis.
+
 ### Pre-processing and normalization
+
+**EAP: Should we make this part of the pipeline, or just start with normalization?
+It's very microarray heavy...
+We could have a vignette in `moanin` documenting the creation of a normalized version of the data.**
 
@@
 estimating the variations across samples before and after normalization is a
-first quality control step. Typically, two other quality control and
+first quality control step. **EAP: Cut or move this sentence, so that this entire section is applicable to microarray or mRNA-Seq**
+
+Typically, two other quality control and
 exploratory analysis steps are also performed before and after normalization:
@@
 We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate.
+condition; (2) its sampling time. We use different markers for each replicate. **EAP: Unclear what 'markers' refers to**
 We also plot the second and third components in the second row.
@@
 
-The next step is typically to run a differential expression analysis. Most of
-`moanin`'s function rely on an object containing a number of information such
-as the metadata and the splines model used. We start by creating the
+The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
+For time-course data, we will consider two different approaches
+
+1) Per-time point analysis, where we consider each time point a different condition and determine what genes are changing between time points, or between conditions at a single time-point.
+
+2) Global analysis, where we consider the expression pattern globally over time, and consider what genes have either different patterns between conditions or a changing pattern (i.e.
+non-constant) over time.
+
+We note that we could have time course data on either a single condition or time course data on multiple conditions (such as the influenza dataset we are considering), which will alter slightly the types of questions we are interested in, but the two basic approaches remains the same.
+In what follows, we will focus on the situation where we have multiple conditions.
+**EAP: Any reason one and not the other.
+Maybe should change this sentence**
+
+In both situations, a common approach first step is to fit a spline model to each gene **FIXME add refs**, and then use that spline model to test for different kinds of differential expression across time.
+
+`moanin` is set up to facilitate working with these spline models for differential expression. Most of
+`moanin`'s functions rely on creating a `splines_model` object to contain the metadata and the splines model to be used.
+
+We start by creating the
 `splines_model` object using the `create_splines_model` function.
@@
 ```
+
+Here we have provided the meta data that defines...
+`create_splines_model` will ....
+**FIXME add information about what meta data is expected, describe what `create_splines_model` does (does it fit the splines?
+Just hold create an object?
+Create basis functions?)**.
 
@@
 
+# Session information
+
+```{r}
+sessionInfo()
+```
+
 # Conclusion

### 2019-05-13 scripts/manuscript.Rmd
changes to manuscript

@@
 
+### Approaches to DE analysis in time-course data
+
 The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
-For time-course data, we will consider two different approaches
+For time-course data, there are two different approaches for determining differentially expressed genes,
 
@@
 non-constant) over time.
+A common approach first step is to fit a spline model to each gene **FIXME add refs**, and then use that spline model to test for different kinds of differential expression across time.
+
+The per-time point analysis is often the entire approach advocated for dealing with small time-course datasets, where there are only a few time points **FIXME add citation to edgeR/limma vignette or DESEQ2**.
+For long time-course datasets, however, such an approach results in creating many different tests, for example for every pair of adjacent timepoints, the results of which are difficult to integrate.
+We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
 
@@
 
-In both situations, a common approach first step is to fit a spline model to each gene **FIXME add refs**, and then use that spline model to test for different kinds of differential expression across time.
-
-`moanin` is set up to facilitate working with these spline models for differential expression. Most of
-`moanin`'s functions rely on creating a `splines_model` object to contain the metadata and the splines model to be used.
+**Use with `moanin`** `moanin` provides functionality for performing both of these types of approaches.
+In both situations, we first set up a object (a `splines_model` object) to hold the meta data, as well as information for fitting the spline model.
 
@@
 **FIXME rename the function to something better**
+
+**EAP: And don't use 'week' in function name!**
 

### 2019-05-13 scripts/manuscript.Rmd
a few more changes

@@
 
-`moanin` provides a simple interface to perform a time by time differential
-expression analysis. Under the hood, it simply calls `limma` [@ritchie:limma] on the set of
-contrasts provided. Here, we show an example between the control mouse and the
+`moanin` provides a simple interface to perform a timepoint by timepoint differential
+expression analysis. This is traditionally done by the user defining the comparisons (called `contrasts` in linear models).  Under the hood, simply calls `limma` [@ritchie:limma] on the set of
+contrasts provided. **EAP: Can we comment on using it for RNA-Seq??**
+
+Here, we show an example where we define our contrasts to be the difference between the control mouse ("M") and the
 mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
-(H5N1), but the function works with any form contrasts.
+(H5N1) ("VL") for each time point, but the function works with any form contrasts.
+**FIXME: Reference to limma vignette for how to define the contrasts** **EAP: Could you make simple function that creates them automatically for two conditions -- i.e.
+find union of time points for two conditions...
+Then we could show an example with just a couple done manually, and then say we can also use a little helper function in the `moanin` that will do it for all time points**
 

### 2019-05-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+The first steps of analysis of gene expression data is always to do normalization and quality control checks of the data.
+In what follows, we show an example of this for the influenza data using generic methods; these steps are not in specific to time course, but could be done for any gene expression analysis.
+
 \hypertarget{pre-processing-and-normalization}{%
 \subsubsection{Pre-processing and normalization}\label{pre-processing-and-normalization}}
+
+\textbf{EAP: Should we make this part of the pipeline, or just start with normalization?
+It's very microarray heavy\ldots{} We could have a vignette in \texttt{moanin} documenting the creation of a normalized version of the data.}
 
@@
 estimating the variations across samples before and after normalization is a
-first quality control step. Typically, two other quality control and
+first quality control step. \textbf{EAP: Cut or move this sentence, so that this entire section is applicable to microarray or mRNA-Seq}
+
+Typically, two other quality control and
 exploratory analysis steps are also performed before and after normalization:
@@
 We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate.
+condition; (2) its sampling time. We use different markers for each replicate. \textbf{EAP: Unclear what `markers' refers to}
 We also plot the second and third components in the second row.
@@
 
-The next step is typically to run a differential expression analysis. Most of
-\texttt{moanin}'s function rely on an object containing a number of information such
-as the metadata and the splines model used. We start by creating the
+The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
+For time-course data, we will consider two different approaches
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\item
+  Per-time point analysis, where we consider each time point a different condition and determine what genes are changing between time points, or between conditions at a single time-point.
+\item
+  Global analysis, where we consider the expression pattern globally over time, and consider what genes have either different patterns between conditions or a changing pattern (i.e.~non-constant) over time.
+\end{enumerate}
+
+We note that we could have time course data on either a single condition or time course data on multiple conditions (such as the influenza dataset we are considering), which will alter slightly the types of questions we are interested in, but the two basic approaches remains the same.
+In what follows, we will focus on the situation where we have multiple conditions.
+\textbf{EAP: Any reason one and not the other.
+Maybe should change this sentence}
+
+In both situations, a common approach first step is to fit a spline model to each gene \textbf{FIXME add refs}, and then use that spline model to test for different kinds of differential expression across time.
+
+\texttt{moanin} is set up to facilitate working with these spline models for differential expression. Most of
+\texttt{moanin}'s functions rely on creating a \texttt{splines\_model} object to contain the metadata and the splines model to be used.
+
+We start by creating the
 \texttt{splines\_model} object using the \texttt{create\_splines\_model} function.
@@
 \end{Shaded}
+
+Here we have provided the meta data that defines\ldots{} \texttt{create\_splines\_model} will \ldots{}.
+\textbf{FIXME add information about what meta data is expected, describe what \texttt{create\_splines\_model} does (does it fit the splines?
+Just hold create an object?
+Create basis functions?)}.
 

### 2019-05-13 scripts/manuscript.Rmd
manuscript changes

@@
 
+**EAP: General comment.
+Your function names are incredibly long compared to standard functions.
+Also, will want to get rid of `moanin::` in front of functions.
+It's more precise, but an average user doesn't ever experience the `::` command; usually just package creaters.
+**
+
 # Analysis of the dynamical response of mouse lung tissue to influenza
@@
 
-The per-time point analysis is often the entire approach advocated for dealing with small time-course datasets, where there are only a few time points **FIXME add citation to edgeR/limma vignette or DESEQ2**.
-For long time-course datasets, however, such an approach results in creating many different tests, for example for every pair of adjacent timepoints, the results of which are difficult to integrate.
+The per-time point analysis is using classical differential expression approaches, and is often the approach advocated when dealing with small time-course datasets, where there are only a few time points **FIXME add citation to edgeR/limma vignette or DESEQ2**.
+For long time-course datasets, however, a separate test for each time-point results in creating many different tests, for example one for every time point, the results of which are difficult to integrate.
 We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
@@
 
-**Use with `moanin`** `moanin` provides functionality for performing both of these types of approaches.
-In both situations, we first set up a object (a `splines_model` object) to hold the meta data, as well as information for fitting the spline model.
+**Use with `moanin`** `moanin` provides functionality for performing both of these types of approaches, though our focus is on the global approach.
+In both situations, we first need to set up a object (a `splines_model` object) to hold the meta data, as well as information for fitting the spline model.
 
@@
 Create basis functions?)**.
+
+**EAP: General comment: you need a few more print() and head() of some objects to demonstrate what the output looks like**
 
@@
 ```
+
+We can repeat this for each of the conditions
 
@@
 due to biological or technical variances for specific genes at specific
-timepoints.
+timepoints. **EAP: We should add more to this**
 
@@
 
-To leverage this temporal structure, Storey et al [@storey:significance] proposes to
-model time-course micro-array with splines functions, and to use a log-ratio
-likelihood test to detect differentially expressed genes. `moanin` extends
-this idea by providing functions to perform time-course differential
-expression analysis using `limma` contrasts.
+To leverage this temporal structure, Storey et al [@storey:significance] proposed to
+model each gene in time-course micro-array with a splines function, and to use a log-ratio likelihood test to detect differentially expressed genes.
+
+`moanin` extends
+this idea by providing functionality to compare time course data between different treatment conditions, using a similar mechanism of contrasts -- only now the contrasts are differences between the estimated means.
+
+**FIXME add in equations, like from the EPICON paper, to describe them mathematically**
+
+**EAP: Do we need this call to limma by the user, or can these not be done internally by the `timecourse_differential_expression_analysis` function?
+We can still preserve the option of user defining their own, but would be nicer to make this simpler for the user (e.g.
+an if clause that tests whether input is matrix or character vector)**
 
@@
 | epicon  | $$  | Captures overall strength of response and overall direction |
+
+**EAP: `epicon` is going to be weird for general use**
 

### 2019-05-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+\hypertarget{approaches-to-de-analysis-in-time-course-data}{%
+\subsubsection{Approaches to DE analysis in time-course data}\label{approaches-to-de-analysis-in-time-course-data}}
+
 The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
-For time-course data, we will consider two different approaches
+For time-course data, there are two different approaches for determining differentially expressed genes,
 
@@
   Global analysis, where we consider the expression pattern globally over time, and consider what genes have either different patterns between conditions or a changing pattern (i.e.~non-constant) over time.
+A common approach first step is to fit a spline model to each gene \textbf{FIXME add refs}, and then use that spline model to test for different kinds of differential expression across time.
 \end{enumerate}
+
+The per-time point analysis is often the entire approach advocated for dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}.
+For long time-course datasets, however, such an approach results in creating many different tests, for example for every pair of adjacent timepoints, the results of which are difficult to integrate.
+We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
 
@@
 
-In both situations, a common approach first step is to fit a spline model to each gene \textbf{FIXME add refs}, and then use that spline model to test for different kinds of differential expression across time.
-
-\texttt{moanin} is set up to facilitate working with these spline models for differential expression. Most of
-\texttt{moanin}'s functions rely on creating a \texttt{splines\_model} object to contain the metadata and the splines model to be used.
+\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches.
+In both situations, we first set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.
 
@@
 \textbf{FIXME rename the function to something better}
+
+\textbf{EAP: And don't use `week' in function name!}
 

### 2019-05-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+\textbf{EAP: General comment.
+Your function names are incredibly long compared to standard functions.
+Also, will want to get rid of \texttt{moanin::} in front of functions.
+It's more precise, but an average user doesn't ever experience the \texttt{::} command; usually just package creaters.
+}
+
 \hypertarget{analysis-of-the-dynamical-response-of-mouse-lung-tissue-to-influenza}{%
@@
 
-The per-time point analysis is often the entire approach advocated for dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}.
-For long time-course datasets, however, such an approach results in creating many different tests, for example for every pair of adjacent timepoints, the results of which are difficult to integrate.
+The per-time point analysis is using classical differential expression approaches, and is often the approach advocated when dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}.
+For long time-course datasets, however, a separate test for each time-point results in creating many different tests, for example one for every time point, the results of which are difficult to integrate.
 We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
@@
 
-\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches.
-In both situations, we first set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.
+\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches, though our focus is on the global approach.
+In both situations, we first need to set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.
 
@@
 
+\textbf{EAP: General comment: you need a few more print() and head() of some objects to demonstrate what the output looks like}
+
 \hypertarget{weekly-differential-expression-analysis}{%
@@
 
-\texttt{moanin} provides a simple interface to perform a time by time differential
-expression analysis. Under the hood, it simply calls \texttt{limma} \citep{ritchie:limma} on the set of
-contrasts provided. Here, we show an example between the control mouse and the
+\texttt{moanin} provides a simple interface to perform a timepoint by timepoint differential
+expression analysis.
+This is traditionally done by the user defining the comparisons (called \texttt{contrasts} in linear models).
+Under the hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of
+contrasts provided. \textbf{EAP: Can we comment on using it for RNA-Seq??}
+
+Here, we show an example where we define our contrasts to be the difference between the control mouse (``M'') and the
 mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
-(H5N1), but the function works with any form contrasts.
+(H5N1) (``VL'') for each time point, but the function works with any form contrasts.
+\textbf{FIXME: Reference to limma vignette for how to define the contrasts} \textbf{EAP: Could you make simple function that creates them automatically for two conditions -- i.e.~find union of time points for two conditions\ldots{} Then we could show an example with just a couple done manually, and then say we can also use a little helper function in the \texttt{moanin} that will do it for all time points}
 
@@
 
+We can repeat this for each of the conditions
+
 Let's look at the distribution of genes found differentially expressed per
@@
 due to biological or technical variances for specific genes at specific
-timepoints.
+timepoints. \textbf{EAP: We should add more to this}
 
@@
 
-To leverage this temporal structure, Storey et al \citep{storey:significance} proposes to
-model time-course micro-array with splines functions, and to use a log-ratio
-likelihood test to detect differentially expressed genes. \texttt{moanin} extends
-this idea by providing functions to perform time-course differential
-expression analysis using \texttt{limma} contrasts.
+To leverage this temporal structure, Storey et al \citep{storey:significance} proposed to
+model each gene in time-course micro-array with a splines function, and to use a log-ratio likelihood test to detect differentially expressed genes.
+
+\texttt{moanin} extends
+this idea by providing functionality to compare time course data between different treatment conditions, using a similar mechanism of contrasts -- only now the contrasts are differences between the estimated means.
+
+\textbf{FIXME add in equations, like from the EPICON paper, to describe them mathematically}
+
+\textbf{EAP: Do we need this call to limma by the user, or can these not be done internally by the \texttt{timecourse\_differential\_expression\_analysis} function?
+We can still preserve the option of user defining their own, but would be nicer to make this simpler for the user (e.g.~an if clause that tests whether input is matrix or character vector)}
 
@@
 \end{longtable}
+
+\textbf{EAP: \texttt{epicon} is going to be weird for general use}
 

### 2019-05-13 scripts/manuscript.Rmd
Added TODO items

@@
   the same thing. Find a better name for both of them
+- switch to camel case function names
+- shorten function names
+- Remove the namespace call
+- Move the normalization to a supplementary data
+- Make clear that it applies to RNASeq and how to call the different functions
+  to RNASeq
+- Check that clustering can deal with count data.
 
@@
 **EAP: And don't use 'week' in function name!**
+
+**Add a helper function to make the contrasts for timely DE**
 

### 2019-05-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
   the same thing. Find a better name for both of them
+\item
+  switch to camel case function names
+\item
+  shorten function names
+\item
+  Remove the namespace call
+\item
+  Move the normalization to a supplementary data
+\item
+  Make clear that it applies to RNASeq and how to call the different functions
+  toRNASeq
+\item
+  Check that clustering can deal with count data.
 \end{itemize}
@@
 \textbf{EAP: And don't use `week' in function name!}
+
+\textbf{Add a helper function to make the contrasts for timely DE}
 

### 2019-05-14 scripts/manuscript.Rmd
FIX for the manuscript

@@
 
-**EAP: General comment.
-Your function names are incredibly long compared to standard functions.
-Also, will want to get rid of `moanin::` in front of functions.
-It's more precise, but an average user doesn't ever experience the `::` command; usually just package creaters.
-**
+**EAP: General comment. Your function names are incredibly long compared to
+standard functions. Also, will want to get rid of `moanin::` in front of
+functions. It's more precise, but an average user doesn't ever experience the
+`::` command; usually just package creaters.**
 
@@
 
-Given the source and a `Targets.txt` structured as follows, microarray can be loaded thanks to the limma library.
+Given the source and a `Targets.txt` structured as follows, microarray can be
+loaded thanks to the limma library.
 
@@
 
-The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
-For time-course data, there are two different approaches for determining differentially expressed genes,
-
-1) Per-time point analysis, where we consider each time point a different condition and determine what genes are changing between time points, or between conditions at a single time-point.
-
-2) Global analysis, where we consider the expression pattern globally over time, and consider what genes have either different patterns between conditions or a changing pattern (i.e.
-non-constant) over time.
-A common approach first step is to fit a spline model to each gene **FIXME add refs**, and then use that spline model to test for different kinds of differential expression across time.
-
-The per-time point analysis is using classical differential expression approaches, and is often the approach advocated when dealing with small time-course datasets, where there are only a few time points **FIXME add citation to edgeR/limma vignette or DESEQ2**.
-For long time-course datasets, however, a separate test for each time-point results in creating many different tests, for example one for every time point, the results of which are difficult to integrate.
-We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
-
-We note that we could have time course data on either a single condition or time course data on multiple conditions (such as the influenza dataset we are considering), which will alter slightly the types of questions we are interested in, but the two basic approaches remains the same.
-In what follows, we will focus on the situation where we have multiple conditions.
-**EAP: Any reason one and not the other.
-Maybe should change this sentence**
-
-**Use with `moanin`** `moanin` provides functionality for performing both of these types of approaches, though our focus is on the global approach.
-In both situations, we first need to set up a object (a `splines_model` object) to hold the meta data, as well as information for fitting the spline model.
+The next step in a gene expression analysis is typically to run a differential
+expression analysis, generally to find genes different between different
+conditions. For time-course data, there are two different approaches for
+determining differentially expressed genes,
+
+1) Per-time point analysis, where we consider each time point a different
+condition and determine what genes are changing between time points, or
+between conditions at a single time-point.
+
+2) Global analysis, where we consider the expression pattern globally over
+time, and consider what genes have either different patterns between
+conditions or a changing pattern (i.e. non-constant) over time. A common
+approach first step is to fit a spline model to each gene **FIXME add refs**,
+and then use that spline model to test for different kinds of differential
+expression across time.
+
+The per-time point analysis is using classical differential expression
+approaches, and is often the approach advocated when dealing with small
+time-course datasets, where there are only a few time points **FIXME add
+citation to edgeR/limma vignette or DESEQ2**. For long time-course datasets,
+however, a separate test for each time-point results in creating many
+different tests, for example one for every time point, the results of which
+are difficult to integrate. We find in practice that the global analysis
+simplifies analysis and of longer time courses data, with per-time point
+analysis reserved for particularly interesting comparisons of individual
+time-points.
+
+We note that we could have time course data on either a single condition or
+time course data on multiple conditions (such as the influenza dataset we are
+considering), which will alter slightly the types of questions we are
+interested in, but the two basic approaches remains the same. In what follows,
+we will focus on the situation where we have multiple conditions. **EAP: Any
+reason one and not the other. Maybe should change this sentence**
+
+**Use with `moanin`** `moanin` provides functionality for performing both of
+these types of approaches, though our focus is on the global approach. In both
+situations, we first need to set up a object (a `splines_model` object) to
+hold the meta data, as well as information for fitting the spline model.
 
@@
 
-Here we have provided the meta data that defines...
-`create_splines_model` will ....
-**FIXME add information about what meta data is expected, describe what `create_splines_model` does (does it fit the splines?
-Just hold create an object?
-Create basis functions?)**.
+Here we have provided the meta data that defines... `create_splines_model`
+will .... **FIXME add information about what meta data is expected, describe
+what `create_splines_model` does (does it fit the splines? Just hold create an
+object? Create basis functions?)**.
 
@@
 
-Here, we show an example where we define our contrasts to be the difference between the control mouse ("M") and the
-mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
-(H5N1) ("VL") for each time point, but the function works with any form contrasts.
-**FIXME: Reference to limma vignette for how to define the contrasts** **EAP: Could you make simple function that creates them automatically for two conditions -- i.e.
-find union of time points for two conditions...
-Then we could show an example with just a couple done manually, and then say we can also use a little helper function in the `moanin` that will do it for all time points**
+Here, we show an example where we define our contrasts to be the difference
+between the control mouse ("M") and the mouse infected with the high dose of
+the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point, but
+the function works with any form contrasts. **FIXME: Reference to limma
+vignette for how to define the contrasts** **EAP: Could you make simple
+function that creates them automatically for two conditions -- i.e. find union
+of time points for two conditions... Then we could show an example with just a
+couple done manually, and then say we can also use a little helper function in
+the `moanin` that will do it for all time points**
 
@@
 	      "M.168-VL.168")
-weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+weekly_de_analysis =  moanin::DE_timepoints(
      data, splines_model, contrasts,
@@
 	      "M.168-K.168")
-K_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+K_weekly_de_analysis =  moanin::DE_timepoints(
      data, splines_model, contrasts,
@@
 	      "M.168-C.168")
-C_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+C_weekly_de_analysis =  moanin::DE_timepoints(
      data, splines_model, contrasts,
@@
 	      "M.168-VH.168")
-VH_weekly_de_analysis =  moanin::weekly_differential_expression_analysis(
+VH_weekly_de_analysis =  moanin::DE_timepoints(
      data, splines_model, contrasts,
@@
 # treatment).
-pvalues = moanin::timecourse_differential_expression_analysis(
+pvalues = moanin::DE_timecourse(
     data, splines_model, contrasts)

### 2019-05-14 bin/moanin/R/timecourse_de_analysis.R

@@ -170,6 +170,14 @@ DE_timecourse = function(data, splines_model,
 
     check_data_meta(data, meta)
 
+    if(use_voom_weights){
+        y = edgeR::DGEList(counts=counts)
+	y = edgeR::calcNormFactors(y, method="upperquartile")
+        v = limma::voom(y, contrasts, plot=FALSE)
+	weights = limma::lmFit(v)
+    }else{
+	weights = NULL
+    }
 
     y = data
 

### 2019-05-14 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\textbf{EAP: General comment.
-Your function names are incredibly long compared to standard functions.
-Also, will want to get rid of \texttt{moanin::} in front of functions.
-It's more precise, but an average user doesn't ever experience the \texttt{::} command; usually just package creaters.
-}
+\textbf{EAP: General comment. Your function names are incredibly long compared to
+standard functions. Also, will want to get rid of \texttt{moanin::} in front of
+functions. It's more precise, but an average user doesn't ever experience the
+\texttt{::} command; usually just package creaters.}
 
@@
 
-Given the source and a \texttt{Targets.txt} structured as follows, microarray can be loaded thanks to the limma library.
+Given the source and a \texttt{Targets.txt} structured as follows, microarray can be
+loaded thanks to the limma library.
 
@@
 
-The next step in a gene expression analysis is typically to run a differential expression analysis, generally to find genes different between different conditions.
-For time-course data, there are two different approaches for determining differentially expressed genes,
+The next step in a gene expression analysis is typically to run a differential
+expression analysis, generally to find genes different between different
+conditions. For time-course data, there are two different approaches for
+determining differentially expressed genes,
 
@@
 \item
-  Per-time point analysis, where we consider each time point a different condition and determine what genes are changing between time points, or between conditions at a single time-point.
-\item
-  Global analysis, where we consider the expression pattern globally over time, and consider what genes have either different patterns between conditions or a changing pattern (i.e.~non-constant) over time.
-A common approach first step is to fit a spline model to each gene \textbf{FIXME add refs}, and then use that spline model to test for different kinds of differential expression across time.
+  Per-time point analysis, where we consider each time point a different
+  condition and determine what genes are changing between time points, or
+  between conditions at a single time-point.
+\item
+  Global analysis, where we consider the expression pattern globally over
+  time, and consider what genes have either different patterns between
+  conditions or a changing pattern (i.e.~non-constant) over time. A common
+  approach first step is to fit a spline model to each gene \textbf{FIXME add refs},
+  and then use that spline model to test for different kinds of differential
+  expression across time.
 \end{enumerate}
 
-The per-time point analysis is using classical differential expression approaches, and is often the approach advocated when dealing with small time-course datasets, where there are only a few time points \textbf{FIXME add citation to edgeR/limma vignette or DESEQ2}.
-For long time-course datasets, however, a separate test for each time-point results in creating many different tests, for example one for every time point, the results of which are difficult to integrate.
-We find in practice that the global analysis simplifies analysis and of longer time courses data, with per-time point analysis reserved for particularly interesting comparisons of individual time-points.
-
-We note that we could have time course data on either a single condition or time course data on multiple conditions (such as the influenza dataset we are considering), which will alter slightly the types of questions we are interested in, but the two basic approaches remains the same.
-In what follows, we will focus on the situation where we have multiple conditions.
-\textbf{EAP: Any reason one and not the other.
-Maybe should change this sentence}
-
-\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of these types of approaches, though our focus is on the global approach.
-In both situations, we first need to set up a object (a \texttt{splines\_model} object) to hold the meta data, as well as information for fitting the spline model.
+The per-time point analysis is using classical differential expression
+approaches, and is often the approach advocated when dealing with small
+time-course datasets, where there are only a few time points \textbf{FIXME add
+citation to edgeR/limma vignette or DESEQ2}. For long time-course datasets,
+however, a separate test for each time-point results in creating many
+different tests, for example one for every time point, the results of which
+are difficult to integrate. We find in practice that the global analysis
+simplifies analysis and of longer time courses data, with per-time point
+analysis reserved for particularly interesting comparisons of individual
+time-points.
+
+We note that we could have time course data on either a single condition or
+time course data on multiple conditions (such as the influenza dataset we are
+considering), which will alter slightly the types of questions we are
+interested in, but the two basic approaches remains the same. In what follows,
+we will focus on the situation where we have multiple conditions. \textbf{EAP: Any
+reason one and not the other. Maybe should change this sentence}
+
+\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of
+these types of approaches, though our focus is on the global approach. In both
+situations, we first need to set up a object (a \texttt{splines\_model} object) to
+hold the meta data, as well as information for fitting the spline model.
 
@@
 
-Here we have provided the meta data that defines\ldots{} \texttt{create\_splines\_model} will \ldots{}.
-\textbf{FIXME add information about what meta data is expected, describe what \texttt{create\_splines\_model} does (does it fit the splines?
-Just hold create an object?
-Create basis functions?)}.
+Here we have provided the meta data that defines\ldots{} \texttt{create\_splines\_model}
+will \ldots{}. \textbf{FIXME add information about what meta data is expected, describe
+what \texttt{create\_splines\_model} does (does it fit the splines? Just hold create an
+object? Create basis functions?)}.
 
@@
 
-Here, we show an example where we define our contrasts to be the difference between the control mouse (``M'') and the
-mouse infected with the high dose of the influenza strain A/Vietnam/1203/04
-(H5N1) (``VL'') for each time point, but the function works with any form contrasts.
-\textbf{FIXME: Reference to limma vignette for how to define the contrasts} \textbf{EAP: Could you make simple function that creates them automatically for two conditions -- i.e.~find union of time points for two conditions\ldots{} Then we could show an example with just a couple done manually, and then say we can also use a little helper function in the \texttt{moanin} that will do it for all time points}
+Here, we show an example where we define our contrasts to be the difference
+between the control mouse (``M'') and the mouse infected with the high dose of
+the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point, but
+the function works with any form contrasts. \textbf{FIXME: Reference to limma
+vignette for how to define the contrasts} \textbf{EAP: Could you make simple
+function that creates them automatically for two conditions -- i.e.~find union
+of time points for two conditions\ldots{} Then we could show an example with just a
+couple done manually, and then say we can also use a little helper function in
+the \texttt{moanin} that will do it for all time points}
 
@@
           \StringTok{"M.168-VL.168"}\NormalTok{)}
-\NormalTok{weekly_de_analysis =}\StringTok{  }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{weekly_differential_expression_analysis}\NormalTok{(}
+\NormalTok{weekly_de_analysis =}\StringTok{  }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{DE_timepoints}\NormalTok{(}
 \NormalTok{     data, splines_model, contrasts,}
@@
 \CommentTok{# treatment).}
-\NormalTok{pvalues =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{timecourse_differential_expression_analysis}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts)}
+\NormalTok{pvalues =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{DE_timecourse}\NormalTok{(}
+\NormalTok{    data, splines_model, contrasts,}
+    \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
 \NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, stats}\OperatorTok{::}\NormalTok{p.adjust)}

### 2019-05-14 scripts/manuscript.Rmd
More fixes
Making the paper less about microarray

@@
 
-The first steps of analysis of gene expression data is always to do normalization and quality control checks of the data.
-In what follows, we show an example of this for the influenza data using generic methods; these steps are not in specific to time course, but could be done for any gene expression analysis.
-
-### Pre-processing and normalization
-
-**EAP: Should we make this part of the pipeline, or just start with normalization?
-It's very microarray heavy...
-We could have a vignette in `moanin` documenting the creation of a normalized version of the data.**
-
-The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
-with accession number GSE63786. The dataset has a total of 209 samples, with
-45018 probes (1 sample was removed by the authors prior to uploading the
-dataset on GEO).
-
-Note that in the following, we assume that the user has access to a data
-folder `data/shoemaker2015/rawdata`, containing the microarray data.
-
-```{r}
-data_dir = "data/shoemaker2015/rawdata"
-```
-
-Given the source and a `Targets.txt` structured as follows, microarray can be
-loaded thanks to the limma library.
-
-```
-FileName
-GSM1557140_AR0313_043raw.txt.gz
-GSM1557141_AR0313_044raw.txt.gz
-GSM1557142_AR0313_045raw.txt.gz
-GSM1557143_AR0313_055raw.txt.gz
-GSM1557144_AR0313_056raw.txt.gz
-GSM1557145_AR0313_057raw.txt.gz
-GSM1557146_AR0313_058raw.txt.gz
-GSM1557147_AR0313_059raw.txt.gz
-```
-
-```{r message=FALSE, results="hide", warning=FALSE}
-targets = limma::readTargets()
-RG = limma::read.maimages(
-  targets, source="agilent",
-  path=data_dir,
-  other.columns="gIsWellAboveBG",
-  green.only=TRUE)
-
-```
-
-The normalization strategy we apply here relies on `limma` and is as follows:
-
-  - First, normalize probe intensity using background information using the
-    "norm-exponential" strategy.
-  - Then normalize across micro-arrrays using quantile normalization.
-
-```{r reading_targets, results="hide", fig.width=4, fig.height=2, message=FALSE, warning=FALSE}
-# Perform background correction using norm-exponential strategy and normalize
-# across samples using quantile normalization.
-RG = limma::backgroundCorrect(RG, method="normexp")
-RG.q = limma::normalizeBetweenArrays(RG, method="quantile")
-```
-
-Plotting the smoothed empirical densities of the probe values before
-normalization shows considerable variations betweens arrays. Once the data has
-been normalized across array using quantile normalization, the smoothed
-empirical densities of each sample perfectly overlap.
-
-```{r plotting_densities, results="hide", fig.width=6, fig.height=3}
-# Plot the densities of the probes before and after normalization.
-par(mfrow=c(1, 2), mar=c(2.0, 2.0, 2.0, 2.0))
-limma::plotDensities(RG, col="black", legend=FALSE,
-		     main="Before normalization", cex=0.8)
-limma::plotDensities(RG.q, col="black", legend=FALSE,
-		     main="After normalization", cex=0.8)
-```
-
-```{r echo=FALSE, results="hide"}
-# Get the number for the text. Do not display results or computation
-control_probes = RG.q$genes[, "ControlType"] == 1
-is_expressed = rowSums(RG.q$other$"gIsWellAboveBG" > 0) >= 4
-n_control_probes = sum(control_probes)
-n_lowly_expressed = sum(!is_expressed)
-```
-
-We then clean up the names of each gene and each sample to match the original
-data. Finally, we further filter out probes based on the following criteria:
-(1) removing the `r n_control_probes` control probes; (2) removing 
-`r n_lowly_expressed` low-expressed genes.
-
-```{r clean_up_names}
-
-sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
-colnames(RG.q) = sample_names
-row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
-idx_to_keep = which(
-  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
-			       "NegativeControl")))
-
-data = RG.q[idx_to_keep,]
-
-# Filter out control probes.
-control_probes = data$genes[, "ControlType"] == 1
-is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4
-
-data = data[!control_probes & is_expressed, ]$E
-```
-
-```{r echo=FALSE, results="hide"}
-# Do not display. setting variables for text
-n_samples = dim(data)[2]
-n_probes = dim(data)[1]
-```
-
-After filtering, we are left with `r n_probes` probes of interest in 
-`r n_samples` samples.
-
-Now that we have normalized and filtered the gene expressions, let's load the
-metadata. The package `moanin` contains the normalized data and meta of
-[@shoemaker:ultrasensitive].
+The first steps of analysis of gene expression data is always to do
+normalization and quality control checks of the data. In what follows, we show
+an example of this for the influenza data using generic methods; these steps
+are not in specific to time course, but could be done for any gene expression
+analysis.
+
+First let's load the data. The package `moanin` contains the normalized
+data and meta of [@shoemaker:ultrasensitive].
 
@@
 
-Plotting the smoothed empirical densities of probe values for each sample and
-estimating the variations across samples before and after normalization is a
-first quality control step. **EAP: Cut or move this sentence, so that this entire section is applicable to microarray or mRNA-Seq**
-
-Typically, two other quality control and
-exploratory analysis steps are also performed before and after normalization:
-(1) low dimensionality embedding of the samples; (2) correlation plots between
-each samples. In both cases, we expect a strong biological signal, while
-replicate samples should be strongly clustered or correlated with one another.
+Typically, two quality control and exploratory analysis steps are also
+performed before and after normalization: (1) low dimensionality embedding of
+the samples; (2) correlation plots between each samples. In both cases, we
+expect a strong biological signal, while replicate samples should be strongly
+clustered or correlated with one another.
 
@@
 pvalues = moanin::DE_timecourse(
-    data, splines_model, contrasts)
+    data, splines_model, contrasts,
+    use_voom_weights=FALSE)
 qvalues = apply(pvalues, 2, stats::p.adjust)

### 2019-05-14 scripts/miscellaneous/normalization.Rmd
Making the paper less about microarray

--- 2019-05-08 scripts/01_quality_control_normalization.Rmd
@@
 ---
-title: "Normalization: A pipeline to analyse time-course gene expression data"
-output: rmarkdown::html_vignette
+title: "[Normalization] A pipeline to analyse time-course gene expression data"
+author:
+    - name: Nelle Varoquaux
+      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
+    - name: Elizabeth Purdom
+      affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
+
+abstract: >
+    The phenotypic diversity of cells is governed by a complex equilibrium
+    between their genetic identity and their environmental interactions:
+    Understanding the dynamics of gene expression is a fundamental question of
+    biology. However, analysing time-course transcriptomic data raises unique
+    challenging statistical and computational questions, requiring the
+    development of novel methods and software.  Using as case study
+    time-course transcriptomics data from mice exposed to different strains of
+    influenza, this workflows provides a step-by-step tutorial of the
+    methodology used to analyse time-course data: (1) normalization of the
+    micro-array dataset; (2) differential expression analysis using functional
+    data analysis; (3) clustering fo time-course data; (4) interpreting
+    clusters with GO term and KEGG pathway enrichment analysis.
+
+keywords: >
+    time-course gene expression data, clustering, differential expression,
+    workflow
+vignette: >
+    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
+    %\VignetteEngine{knitr::rmarkdown}
+output: BiocWorkflowTools::f1000_article
+
 bibliography: bibliography.bib
@@
 library(stats)
+library(knitr)
+library(kableExtra)
+library(BiocStyle)
+library(biomaRt)
+
+library(ggplot2)
+library(NMF)
+library(KEGGprofile)
+library(viridis)
+
+library(topGO)
+
 library(moanin)
-library(ggplot2)
-library(stats)
 
 source("utils.R")
-
-# library(moanin)
-```
-
-```{r echo=FALSE}
-# Set options
-df = 6
-
-# Preprocessing & filtering
-is_count = FALSE
-take_log = FALSE    
-filter_expression = FALSE
-
-# Differential expression analysis
-timecourse_contrasts = c("C-K", "C-M")
-
-# Clustering
-filter_genes = TRUE
-n_genes_to_keep = 5000
-n_clusters = 20
-percentage_genes_to_label = 0.5
 ```
@@
 ```{r echo=FALSE, include=FALSE}
-library(knitr)
 knitr::opts_chunk$set(
-  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
+  cache=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
   echo=TRUE,
@@
 )
-
 ```
 
-## Pre-processing and normalization
+# TODO list
+
+## Quality control and normalization
+
+The first steps of analysis of gene expression data is always to do normalization and quality control checks of the data.
+In what follows, we show an example of this for the influenza data using generic methods; these steps are not in specific to time course, but could be done for any gene expression analysis.
+
+### Pre-processing and normalization
+
+**EAP: Should we make this part of the pipeline, or just start with normalization?
+It's very microarray heavy...
+We could have a vignette in `moanin` documenting the creation of a normalized version of the data.**
 
@@
 ```
-
-```{r echo=FALSE, results="hide"}
-write.table(data, ".results/quantile_normalized.txt")
-```
-
-## Exploratory analysis and quality control
-
-Plotting the smoothed empirical densities of probe values for each sample and
-estimating the variations across samples before and after normalization is a
-first quality control step. Typically, two other quality control and
-exploratory analysis steps are also performed before and after normalization:
-(1) low dimensionality embedding of the samples; (2) correlation plots between
-each samples. In both cases, we expect a strong biological signal, while
-replicate samples should be strongly clustered or correlated with one another.
-
-Before performing any additional exploratory analysis, let us only keep highly
-variable genes: we keep for this step only the top 50% most variable genes.
-
-```{r}
-variance_cutoff = 0.5
-variance_per_genes = apply(data, 1, mad)
-min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
-data = data[variance_per_genes > min_variance,]
-```
-
-```{r echo=FALSE, fig.width=8, fig.height=4}
-graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
-	       ylab="# probes")
-graphics::abline(v=min_variance, col="#AB0000", lw=2)
-```
-
-Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
-centered and scaled gene expression data.
-
-```{r pca, fig.width=8, fig.height=8}
-# Reorder genes on condition, time, and replicate
-
-pca_data = stats::prcomp(t(data), rank=3, center=TRUE, scale=TRUE) 
-percent_var = round(100 * attr(pca_data, "percentVar"))
-```
-
-We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate.
-We also plot the second and third components in the second row.
-
-```{r pca_plots, fig.width=8, fig.height=8}
-graphics::par(mfrow=c(2, 2))
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
-    col=ann_colors$Group[meta$Group],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
-    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
-
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Group[meta$Group],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
-graphics::plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
-
-```
-
-Then, we plot the pearson correlation across each samples. We order the
-samples by their Group (treatment) and Timepoint (the time of sampling).
-Additionally, in this example, we order each treatment by strength of the
-pathogenicity of the treatment: Control, Kawasaki, California, low dose of
-Vietnam, then high dose of Vietnam.
-
-```{r  correlation_plot, fig.width=6, fig.height=6}
-library(NMF)
-
-# Reorder the conditions such that:
-#	- Control is before any influenza treatment
-#	- Each treatment is ordered from low to high pathogeny
-meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
-
-# Reorder genes on condition, time, and replicate
-ord = order(
-  meta$Group,
-  meta$Timepoint,
-  meta$Replicate)
-data = data[, ord]
-meta = meta[ord, ]
-
-data_corr = stats::cor(data, method="pearson")
-
-meta$Timepoint = as.factor(meta$Timepoint)
-NMF::aheatmap(
-    data_corr,
-    Colv=NA, Rowv=NA,
-    annCol=meta[, c("Group", "Timepoint")],
-    annRow=meta[, c("Group", "Timepoint")],
-    annLegend=TRUE, 
-    annColors=ann_colors,
-    main="Correlation plot")
-```
-
-We can already see interesting patterns emerging from the correlation plot.
-First, the cross-correlation amongst samples taking from the control mice is
-higher than the cross correlation amongst the rest of the treatments. Second,
-the influenza-infected mice midly react until time point 36. Third, the less
-pathogenic the strain is, the closer the samples are to the control condition.
-Fourth, the Vietnam samples at time point 120 and 168 are the one that are the
-most different from control samples.
-
-# References

### 2019-05-14 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The first steps of analysis of gene expression data is always to do normalization and quality control checks of the data.
-In what follows, we show an example of this for the influenza data using generic methods; these steps are not in specific to time course, but could be done for any gene expression analysis.
-
-\hypertarget{pre-processing-and-normalization}{%
-\subsubsection{Pre-processing and normalization}\label{pre-processing-and-normalization}}
-
-\textbf{EAP: Should we make this part of the pipeline, or just start with normalization?
-It's very microarray heavy\ldots{} We could have a vignette in \texttt{moanin} documenting the creation of a normalized version of the data.}
-
-The micro-array data were obtained from NCBI Gene Expression Omnibus (GEO),
-with accession number GSE63786. The dataset has a total of 209 samples, with
-45018 probes (1 sample was removed by the authors prior to uploading the
-dataset on GEO).
-
-Note that in the following, we assume that the user has access to a data
-folder \texttt{data/shoemaker2015/rawdata}, containing the microarray data.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{data_dir =}\StringTok{ "data/shoemaker2015/rawdata"}
-\end{Highlighting}
-\end{Shaded}
-
-Given the source and a \texttt{Targets.txt} structured as follows, microarray can be
-loaded thanks to the limma library.
-
-\begin{verbatim}
-FileName
-GSM1557140_AR0313_043raw.txt.gz
-GSM1557141_AR0313_044raw.txt.gz
-GSM1557142_AR0313_045raw.txt.gz
-GSM1557143_AR0313_055raw.txt.gz
-GSM1557144_AR0313_056raw.txt.gz
-GSM1557145_AR0313_057raw.txt.gz
-GSM1557146_AR0313_058raw.txt.gz
-GSM1557147_AR0313_059raw.txt.gz
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{targets =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{readTargets}\NormalTok{()}
-\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{read.maimages}\NormalTok{(}
-\NormalTok{  targets, }\DataTypeTok{source=}\StringTok{"agilent"}\NormalTok{,}
-  \DataTypeTok{path=}\NormalTok{data_dir,}
-  \DataTypeTok{other.columns=}\StringTok{"gIsWellAboveBG"}\NormalTok{,}
-  \DataTypeTok{green.only=}\OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-The normalization strategy we apply here relies on \texttt{limma} and is as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  First, normalize probe intensity using background information using the
-  ``norm-exponential'' strategy.
-\item
-  Then normalize across micro-arrrays using quantile normalization.
-\end{itemize}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Perform background correction using norm-exponential strategy and normalize}
-\CommentTok{# across samples using quantile normalization.}
-\NormalTok{RG =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{backgroundCorrect}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"normexp"}\NormalTok{)}
-\NormalTok{RG.q =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{normalizeBetweenArrays}\NormalTok{(RG, }\DataTypeTok{method=}\StringTok{"quantile"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Plotting the smoothed empirical densities of the probe values before
-normalization shows considerable variations betweens arrays. Once the data has
-been normalized across array using quantile normalization, the smoothed
-empirical densities of each sample perfectly overlap.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Plot the densities of the probes before and after normalization.}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{, }\FloatTok{2.0}\NormalTok{))}
-\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
-             \DataTypeTok{main=}\StringTok{"Before normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
-\NormalTok{limma}\OperatorTok{::}\KeywordTok{plotDensities}\NormalTok{(RG.q, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{legend=}\OtherTok{FALSE}\NormalTok{,}
-             \DataTypeTok{main=}\StringTok{"After normalization"}\NormalTok{, }\DataTypeTok{cex=}\FloatTok{0.8}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/plotting_densities-1} \end{center}
-
-We then clean up the names of each gene and each sample to match the original
-data. Finally, we further filter out probes based on the following criteria:
-(1) removing the 1486 control probes; (2) removing
-4823 low-expressed genes.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sample_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(RG.q), }\StringTok{"_"}\NormalTok{), .subset2,  }\DecValTok{1}\NormalTok{)}
-\KeywordTok{colnames}\NormalTok{(RG.q) =}\StringTok{ }\NormalTok{sample_names}
-\KeywordTok{row.names}\NormalTok{(RG.q) =}\StringTok{ }\KeywordTok{make.names}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{SystematicName, }\DataTypeTok{unique=}\OtherTok{TRUE}\NormalTok{)}
-\NormalTok{idx_to_keep =}\StringTok{ }\KeywordTok{which}\NormalTok{(}
-  \OperatorTok{!}\NormalTok{(RG}\OperatorTok{$}\NormalTok{genes}\OperatorTok{$}\NormalTok{Systematic }\OperatorTok{%in%}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"DarkCorner"}\NormalTok{, }\StringTok{"GE_BrightCorner"}\NormalTok{,}
-                   \StringTok{"NegativeControl"}\NormalTok{)))}
-
-\NormalTok{data =}\StringTok{ }\NormalTok{RG.q[idx_to_keep,]}
-
-\CommentTok{# Filter out control probes.}
-\NormalTok{control_probes =}\StringTok{ }\NormalTok{data}\OperatorTok{$}\NormalTok{genes[, }\StringTok{"ControlType"}\NormalTok{] }\OperatorTok{==}\StringTok{ }\DecValTok{1}
-\NormalTok{is_expressed =}\StringTok{ }\KeywordTok{rowSums}\NormalTok{(data}\OperatorTok{$}\NormalTok{other}\OperatorTok{$}\StringTok{"gIsWellAboveBG"} \OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{>=}\StringTok{ }\DecValTok{4}
-
-\NormalTok{data =}\StringTok{ }\NormalTok{data[}\OperatorTok{!}\NormalTok{control_probes }\OperatorTok{&}\StringTok{ }\NormalTok{is_expressed, ]}\OperatorTok{$}\NormalTok{E}
-\end{Highlighting}
-\end{Shaded}
-
-After filtering, we are left with 39544 probes of interest in
-209 samples.
-
-Now that we have normalized and filtered the gene expressions, let's load the
-metadata. The package \texttt{moanin} contains the normalized data and meta of
-\citep{shoemaker:ultrasensitive}.
+The first steps of analysis of gene expression data is always to do
+normalization and quality control checks of the data. In what follows, we show
+an example of this for the influenza data using generic methods; these steps
+are not in specific to time course, but could be done for any gene expression
+analysis.
+
+First let's load the data. The package \texttt{moanin} contains the normalized
+data and meta of \citep{shoemaker:ultrasensitive}.
 
@@
 
-Plotting the smoothed empirical densities of probe values for each sample and
-estimating the variations across samples before and after normalization is a
-first quality control step. \textbf{EAP: Cut or move this sentence, so that this entire section is applicable to microarray or mRNA-Seq}
-
-Typically, two other quality control and
-exploratory analysis steps are also performed before and after normalization:
-(1) low dimensionality embedding of the samples; (2) correlation plots between
-each samples. In both cases, we expect a strong biological signal, while
-replicate samples should be strongly clustered or correlated with one another.
+Typically, two quality control and exploratory analysis steps are also
+performed before and after normalization: (1) low dimensionality embedding of
+the samples; (2) correlation plots between each samples. In both cases, we
+expect a strong biological signal, while replicate samples should be strongly
+clustered or correlated with one another.
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-5-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
 

### 2019-05-15 scripts/manuscript.Rmd
Renaming functions, adding text, fixing FIXMEs

@@
 
-```{r schema, echo=FALSE, fig.width=4, fig.cap="Workflow for analyzing time-course datasets."}
+```{r schema, echo=FALSE, fig.width=3, fig.cap="Workflow for analyzing time-course datasets."}
 
@@
 conditions or a changing pattern (i.e. non-constant) over time. A common
-approach first step is to fit a spline model to each gene **FIXME add refs**,
+approach first step is to fit a spline model to each gene
+[@storey:significance],
 and then use that spline model to test for different kinds of differential
@@
 approaches, and is often the approach advocated when dealing with small
-time-course datasets, where there are only a few time points **FIXME add
-citation to edgeR/limma vignette or DESEQ2**. For long time-course datasets,
+time-course datasets, where there are only a few time points 
+[@ritchie:limma, @robinson:edgeR, @love:moderated] . For long time-course datasets,
 however, a separate test for each time-point results in creating many
@@
 
-We start by creating the
-`splines_model` object using the `create_splines_model` function.
+We start by creating the `splines_model` object using the
+`create_splines_model` function. The `splines_model` object contains a number
+of metadata and options used throughout this analysis: the condition and
+timepoints of each samples, the formula object used or the basis or the
+degrees of freedom of the model. The metadata data.frame object should
+contain at least two columns: one named `Group`, containing the  treatement
+effect, and a second one named `Timepoint` containing the timepoint
+information.
+
+```{r echo=FALSE}
+kable(head(meta))
+```
+
+If no formula is provided, it will default to the following: 
+`formula = ~Group:splines::ns(Timepoint, df=degrees_of_fredoom) + Group + 0`
 
@@
 expression analysis. This is traditionally done by the user defining the comparisons (called `contrasts` in linear models).  Under the hood, simply calls `limma` [@ritchie:limma] on the set of
-contrasts provided. **EAP: Can we comment on using it for RNA-Seq??**
+contrasts provided. By default, it expects RNA-Seq contact counts, and will
+estimate voom weights. We tern
 
@@
 the `moanin` that will do it for all time points**
-
-**FIXME rename the function to something better**
-
-**EAP: And don't use 'week' in function name!**
 

### 2019-05-15 bin/moanin/tests/testthat/test-DE_timepoints.R

@@ -19,8 +19,9 @@ test_that("time-course DE analysis", {
 	contrasts=contrast_formulas,
 	levels=levels(meta$Group))
 
-    expect_silent(moanin::timecourse_differential_expression_analysis(
-	as.matrix(data), splines_model, contrasts=contrast))
+    expect_silent(moanin::DE_timecourse(
+	as.matrix(data), splines_model, contrasts=contrast,
+	use_voom_weights=FALSE))
 
     # Now test with several contrasts
     contrast_formulas = c("K-M", "C-M")
@@ -28,8 +29,8 @@ test_that("time-course DE analysis", {
 	contrasts=contrast_formulas,
 	levels=levels(meta$Group))
 
-    expect_silent(moanin::timecourse_differential_expression_analysis(
-    	as.matrix(data), splines_model, contrasts=contrast))
-
+    expect_silent(moanin::DE_timecourse(
+    	as.matrix(data), splines_model, contrasts=contrast,
+	use_voom_weights=FALSE))
 
 })

### 2019-05-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
   conditions or a changing pattern (i.e.~non-constant) over time. A common
-  approach first step is to fit a spline model to each gene \textbf{FIXME add refs},
+  approach first step is to fit a spline model to each gene
+  \citep{storey:significance},
   and then use that spline model to test for different kinds of differential
@@
 approaches, and is often the approach advocated when dealing with small
-time-course datasets, where there are only a few time points \textbf{FIXME add
-citation to edgeR/limma vignette or DESEQ2}. For long time-course datasets,
+time-course datasets, where there are only a few time points
+\citep[\citet{robinson:edgeR}, \citet{love:moderated}]{ritchie:limma} . For long time-course datasets,
 however, a separate test for each time-point results in creating many
@@
 
-We start by creating the
-\texttt{splines\_model} object using the \texttt{create\_splines\_model} function.
+We start by creating the \texttt{splines\_model} object using the
+\texttt{create\_splines\_model} function. The \texttt{splines\_model} object contains a number
+of metadata and options used throughout this analysis: the condition and
+timepoints of each samples, the formula object used or the basis or the
+degrees of freedom of the model. The metadata data.frame object should
+contain at least two columns: one named \texttt{Group}, containing the treatement
+effect, and a second one named \texttt{Timepoint} containing the timepoint
+information.
+
+\begin{tabular}{l|r|l|r|r}
+\hline
+  &  & Group & Replicate & Timepoint\\
+\hline
+GSM1557140 & 0 & K & 1 & 0\\
+\hline
+GSM1557141 & 1 & K & 2 & 0\\
+\hline
+GSM1557142 & 2 & K & 3 & 0\\
+\hline
+GSM1557143 & 3 & K & 1 & 12\\
+\hline
+GSM1557144 & 4 & K & 2 & 12\\
+\hline
+GSM1557145 & 5 & K & 3 & 12\\
+\hline
+\end{tabular}
+
+If no formula is provided, it will default to the following:
+\texttt{formula\ =\ \textasciitilde{}Group:splines::ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}
 
@@
 Under the hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of
-contrasts provided. \textbf{EAP: Can we comment on using it for RNA-Seq??}
+contrasts provided. By default, it expects RNA-Seq contact counts, and will
+estimate voom weights. We tern
 
@@
 the \texttt{moanin} that will do it for all time points}
-
-\textbf{FIXME rename the function to something better}
-
-\textbf{EAP: And don't use `week' in function name!}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-10-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
 

### 2019-05-15 scripts/manuscript.Rmd
Update the manuscript to use the helper function

@@
 data_corr = stats::cor(variance_filtered_data, method="pearson")
-
 data_corr_meta = meta[ord, ]
-
-# Copy the metadata before changing the timepoints into factors
-data_corr_meta$Timepoint = as.factor(meta$Timepoint)
+data_corr_meta$Timepoint = as.factor(data_corr_meta$Timepoint)
+
 NMF::aheatmap(
@@
 
-`moanin` provides a simple interface to perform a timepoint by timepoint differential
-expression analysis. This is traditionally done by the user defining the comparisons (called `contrasts` in linear models).  Under the hood, simply calls `limma` [@ritchie:limma] on the set of
-contrasts provided. By default, it expects RNA-Seq contact counts, and will
-estimate voom weights. We tern
+`moanin` provides a simple interface to perform a timepoint by timepoint
+differential expression analysis. This is traditionally done by the user
+defining the comparisons (called `contrasts` in linear models).  Under the
+hood, simply calls `limma` [@ritchie:limma] on the set of contrasts provided.
+By default, it expects RNA-Seq contact counts, and will estimate voom weights.
+We tern
 
@@
 the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point, but
-the function works with any form contrasts. **FIXME: Reference to limma
-vignette for how to define the contrasts** **EAP: Could you make simple
-function that creates them automatically for two conditions -- i.e. find union
-of time points for two conditions... Then we could show an example with just a
-couple done manually, and then say we can also use a little helper function in
-the `moanin` that will do it for all time points**
-
-**Add a helper function to make the contrasts for timely DE**
+the function works with any form contrasts [@ritchie:limma].
+
+First, create the contrasts for all timepoints between the two groups of
+interest:
+```{r make_timepoints_contarts}
+contrasts = create_timepoints_contrasts("M", "VL", splines_model)
+```
+
+```{r echo=FALSE}
+kable(contrasts)
+```
+
+Then run the differential expression analysis on all of those timepoints
+jointly.
 
 ```{r weekly_differential_analysis}
-meta$WeeklyGroup = make.names(meta$Group:as.factor(meta$Timepoint))
-
 # Define contrast	
-contrasts = c("M.0-VL.0",
-	      "M.3-VL.3",
-	      "M.6-VL.6",
-	      "M.9-VL.9",
-	      "M.12-VL.12",
-	      "M.18-VL.18",
-	      "M.24-VL.24",
-	      "M.30-VL.30",
-	      "M.36-VL.36",
-	      "M.48-VL.48",
-	      "M.60-VL.60",
-	      "M.72-VL.72",
-	      "M.120-VL.120",
-	      "M.168-VL.168")
 weekly_de_analysis =  moanin::DE_timepoints(
@@
 
+```{r echo=FALSE}
+kable(weekly_de_analysis[1:10, ])
+```
+
 We can repeat this for each of the conditions
@@
 ```{r echo=FALSE, results="hide"}
-# Hide those results
-# Run the DE analysis on the rest of the influenza strain.
-contrasts = c("M.0-K.0",
-	      "M.3-K.3",
-	      "M.6-K.6",
-	      "M.9-K.9",
-	      "M.12-K.12",
-	      "M.18-K.18",
-	      "M.24-K.24",
-	      "M.30-K.30",
-	      "M.36-K.36",
-	      "M.48-K.48",
-	      "M.60-K.60",
-	      "M.72-K.72",
-	      "M.120-K.120",
-	      "M.168-K.168")
+contrasts = create_timepoints_contrasts("M", "K", splines_model)
 K_weekly_de_analysis =  moanin::DE_timepoints(
@@
      use_voom_weights=FALSE)
-contrasts = c("M.0-C.0",
-	      "M.3-C.3",
-	      "M.6-C.6",
-	      "M.9-C.9",
-	      "M.12-C.12",
-	      "M.18-C.18",
-	      "M.24-C.24",
-	      "M.30-C.30",
-	      "M.36-K.36",
-	      "M.48-C.48",
-	      "M.60-C.60",
-	      "M.72-C.72",
-	      "M.120-C.120",
-	      "M.168-C.168")
+
+contrasts = create_timepoints_contrasts("M", "C", splines_model)
 C_weekly_de_analysis =  moanin::DE_timepoints(
@@
      use_voom_weights=FALSE)
-contrasts = c("M.0-VH.0",
-	      "M.3-VH.3",
-	      "M.6-VH.6",
-	      "M.9-VH.9",
-	      "M.12-VH.12",
-	      "M.18-VH.18",
-	      "M.24-VH.24",
-	      "M.30-VH.30",
-	      "M.36-VH.36",
-	      "M.48-VH.48",
-	      "M.60-VH.60",
-	      "M.72-VH.72",
-	      "M.120-VH.120",
-	      "M.168-VH.168")
+
+contrasts = create_timepoints_contrasts("M", "VH", splines_model)
 VH_weekly_de_analysis =  moanin::DE_timepoints(
@@
 
-To leverage this temporal structure, Storey et al [@storey:significance] proposed to
-model each gene in time-course micro-array with a splines function, and to use a log-ratio likelihood test to detect differentially expressed genes.
-
-`moanin` extends
-this idea by providing functionality to compare time course data between different treatment conditions, using a similar mechanism of contrasts -- only now the contrasts are differences between the estimated means.
+To leverage this temporal structure, Storey et al [@storey:significance]
+proposed to model each gene in time-course micro-array with a splines
+function, and to use a log-ratio likelihood test to detect differentially
+expressed genes.
+
+`moanin` extends this idea by providing functionality to compare time course
+data between different treatment conditions, using a similar mechanism of
+contrasts -- only now the contrasts are differences between the estimated
+means.
 
@@
 
-**EAP: Do we need this call to limma by the user, or can these not be done internally by the `timecourse_differential_expression_analysis` function?
-We can still preserve the option of user defining their own, but would be nicer to make this simpler for the user (e.g.
-an if clause that tests whether input is matrix or character vector)**
+**EAP: Do we need this call to limma by the user, or can these not be done
+internally by the `timecourse_differential_expression_analysis` function? We
+can still preserve the option of user defining their own, but would be nicer
+to make this simpler for the user (e.g. an if clause that tests whether input
+is matrix or character vector)**
 

### 2019-05-15 bin/moanin/R/DE_timepoints.R

@@ -65,3 +65,45 @@ DE_timepoints = function(data, splines_model,
 				 combine_results, fit2=fit))
     return(all_results)
 }
+
+
+#' Creates pairwise contrasts for all timepoints
+#'
+#' @param group1 First group to consider
+#' @param group2 Second group to consider
+#' @param splines_model splines model object
+#'
+#' @examples
+#' data(shoemaker2015)
+#' meta = shoemaker2015$meta
+#' splines_model = moanin::create_splines_model(meta)
+#' contrasts = create_timepoints_contrasts("C", "M", splines_model)
+#'
+#' @export
+create_timepoints_contrasts = function(group1, group2, splines_model){
+    meta = splines_model$meta
+    meta = meta[meta$Group %in% c(group1, group2),]
+    all_timepoints = sort(unique(meta$Timepoint))
+    contrasts = rep(NA, length(all_timepoints))
+    for(i in 1:length(all_timepoints)){
+	# First, check that the two conditions have been sampled for this
+	# timepoint
+	timepoint = all_timepoints[i]
+	submeta = meta[meta$Timepoint == timepoint, ]
+	if(length(unique(submeta$WeeklyGroup)) == 2){
+	    groups = as.character(unique(submeta$WeeklyGroup))
+	    contrasts[i] = paste0(group1, ".", timepoint, "-", group2, ".", timepoint)
+	}else if(length(unique(submeta$WeeklyGroup)) == 1){
+	    if(unique(submeta$Group)[1] ==  group1){
+		missing_condition = group2
+	    }else{
+		missing_condition = group1
+	    }
+	    msg = paste(
+		"moanin::create_timepoints_contrasts: timepoint",
+		timepoint, "is missing in condition", missing_condition)
+	    warning(msg)
+	}
+    }
+    return(contrasts[!is.na(contrasts)])
+}

### 2019-05-15 bin/moanin/tests/testthat/test-DE_timepoints.R

@@ -1,36 +1,21 @@
 library(moanin)
 library(limma)
 
-context("moanin::de_analysis.R")
+context("moanin::DE_timepoints.R")
 
-test_that("time-course DE analysis", {
+test_that("DE_timepoints::create_timepoints_contrasts", {
     data(shoemaker2015)
     data = shoemaker2015$data
     meta = shoemaker2015$meta
-
+    
     splines_model = create_splines_model(meta)
+    expect_silent(create_timepoints_contrasts("M", "C", splines_model))
 
-    # Limit ourselves to the top 5000 genes
-    data = data[1:5000, ]
-
-    # Do only one contrast
-    contrast_formulas = c("K-M")
-    contrast = limma::makeContrasts(
-	contrasts=contrast_formulas,
-	levels=levels(meta$Group))
-
-    expect_silent(moanin::DE_timecourse(
-	as.matrix(data), splines_model, contrasts=contrast,
-	use_voom_weights=FALSE))
-
-    # Now test with several contrasts
-    contrast_formulas = c("K-M", "C-M")
-    contrasts = limma::makeContrasts(
-	contrasts=contrast_formulas,
-	levels=levels(meta$Group))
-
-    expect_silent(moanin::DE_timecourse(
-    	as.matrix(data), splines_model, contrasts=contrast,
-	use_voom_weights=FALSE))
+    # Now drop timepoint 6 in condition C. That should raise a warning
+    mask = !((meta$Group == "C") & (meta$Timepoint == 6))
+    splines_model = create_splines_model(meta[mask,])
+    msg = paste0("moanin::create_timepoints_contrasts: timepoint",
+		 " 6 is missing in condition C")
+    expect_warning(create_timepoints_contrasts("M", "C", splines_model), msg)
 
 })

### 2019-05-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \NormalTok{data_corr =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
-
 \NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
-
-\CommentTok{# Copy the metadata before changing the timepoints into factors}
-\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)}
+\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}
+
 \NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
@@
 
-\texttt{moanin} provides a simple interface to perform a timepoint by timepoint differential
-expression analysis.
-This is traditionally done by the user defining the comparisons (called \texttt{contrasts} in linear models).
-Under the hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of
-contrasts provided. By default, it expects RNA-Seq contact counts, and will
-estimate voom weights. We tern
+\texttt{moanin} provides a simple interface to perform a timepoint by timepoint
+differential expression analysis. This is traditionally done by the user
+defining the comparisons (called \texttt{contrasts} in linear models). Under the
+hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of contrasts provided.
+By default, it expects RNA-Seq contact counts, and will estimate voom weights.
+We tern
 
@@
 the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point, but
-the function works with any form contrasts. \textbf{FIXME: Reference to limma
-vignette for how to define the contrasts} \textbf{EAP: Could you make simple
-function that creates them automatically for two conditions -- i.e.~find union
-of time points for two conditions\ldots{} Then we could show an example with just a
-couple done manually, and then say we can also use a little helper function in
-the \texttt{moanin} that will do it for all time points}
-
-\textbf{Add a helper function to make the contrasts for timely DE}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{meta}\OperatorTok{$}\NormalTok{WeeklyGroup =}\StringTok{ }\KeywordTok{make.names}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group}\OperatorTok{:}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint))}
-
+the function works with any form contrasts \citep{ritchie:limma}.
+
+First, create the contrasts for all timepoints between the two groups of
+interest:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, splines_model)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{tabular}{l}
+\hline
+x\\
+\hline
+M.0-VL.0\\
+\hline
+M.3-VL.3\\
+\hline
+M.6-VL.6\\
+\hline
+M.9-VL.9\\
+\hline
+M.12-VL.12\\
+\hline
+M.18-VL.18\\
+\hline
+M.24-VL.24\\
+\hline
+M.30-VL.30\\
+\hline
+M.36-VL.36\\
+\hline
+M.48-VL.48\\
+\hline
+M.60-VL.60\\
+\hline
+M.72-VL.72\\
+\hline
+M.120-VL.120\\
+\hline
+M.168-VL.168\\
+\hline
+\end{tabular}
+
+Then run the differential expression analysis on all of those timepoints
+jointly.
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{# Define contrast   }
-\NormalTok{contrasts =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"M.0-VL.0"}\NormalTok{,}
-          \StringTok{"M.3-VL.3"}\NormalTok{,}
-          \StringTok{"M.6-VL.6"}\NormalTok{,}
-          \StringTok{"M.9-VL.9"}\NormalTok{,}
-          \StringTok{"M.12-VL.12"}\NormalTok{,}
-          \StringTok{"M.18-VL.18"}\NormalTok{,}
-          \StringTok{"M.24-VL.24"}\NormalTok{,}
-          \StringTok{"M.30-VL.30"}\NormalTok{,}
-          \StringTok{"M.36-VL.36"}\NormalTok{,}
-          \StringTok{"M.48-VL.48"}\NormalTok{,}
-          \StringTok{"M.60-VL.60"}\NormalTok{,}
-          \StringTok{"M.72-VL.72"}\NormalTok{,}
-          \StringTok{"M.120-VL.120"}\NormalTok{,}
-          \StringTok{"M.168-VL.168"}\NormalTok{)}
 \NormalTok{weekly_de_analysis =}\StringTok{  }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{DE_timepoints}\NormalTok{(}
@@
 
+\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
+\hline
+  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc & M.6-VL.6-pval & M.6-VL.6-qval & M.6-VL.6-lfc & M.9-VL.9-pval & M.9-VL.9-qval & M.9-VL.9-lfc & M.12-VL.12-pval & M.12-VL.12-qval & M.12-VL.12-lfc & M.18-VL.18-pval & M.18-VL.18-qval & M.18-VL.18-lfc & M.24-VL.24-pval & M.24-VL.24-qval & M.24-VL.24-lfc & M.30-VL.30-pval & M.30-VL.30-qval & M.30-VL.30-lfc & M.36-VL.36-pval & M.36-VL.36-qval & M.36-VL.36-lfc & M.48-VL.48-pval & M.48-VL.48-qval & M.48-VL.48-lfc & M.60-VL.60-pval & M.60-VL.60-qval & M.60-VL.60-lfc & M.72-VL.72-pval & M.72-VL.72-qval & M.72-VL.72-lfc & M.120-VL.120-pval & M.120-VL.120-qval & M.120-VL.120-lfc & M.168-VL.168-pval & M.168-VL.168-qval & M.168-VL.168-lfc\\
+\hline
+NM\_009912 & 0.2728476 & 0.4745864 & 2.2018266 & 0.7670600 & 0.8760739 & 1.1576881 & 0.0514865 & 0.1487818 & 2.5397431 & 0.1461489 & 0.3125183 & -0.7613983 & 0.2735411 & 0.4753685 & 1.4786596 & 0.5199686 & 0.7057241 & 1.0838768 & 0.0372513 & 0.1172476 & 1.0099581 & 0.1663710 & 0.3415394 & -2.0803981 & 0.0040198 & 0.0209941 & 1.867318 & 0.0000272 & 0.0003259 & -2.877945 & 0.0000001 & 0.0000025 & -2.200448 & 0.0000000 & 0.0000000 & -2.941491 & 0.0000000 & 0.0000000 & 2.659545 & 0.0000000 & 0.0000000 & -4.846082\\
+\hline
+NM\_008725 & 0.8508522 & 0.9242869 & 2.0281764 & 0.1390300 & 0.3020872 & 1.4568427 & 0.0034217 & 0.0184668 & 2.3999010 & 0.0011765 & 0.0078218 & -0.5946375 & 0.2218494 & 0.4142375 & 0.8974352 & 0.0490869 & 0.1436339 & 1.0689628 & 0.0001128 & 0.0011050 & -1.1910616 & 0.2358903 & 0.4313519 & 0.8409000 & 0.0025790 & 0.0147499 & 1.463357 & 0.0000005 & 0.0000098 & -3.732241 & 0.0992368 & 0.2386529 & -3.305718 & 0.2451045 & 0.4423855 & -5.832292 & 0.0028580 & 0.0160083 & 5.103924 & 0.3298546 & 0.5356392 & 6.770567\\
+\hline
+NM\_007473 & 0.0880394 & 0.2192542 & 0.6779618 & 0.9165990 & 0.9592675 & 0.7639279 & 0.6859316 & 0.8250549 & 1.2874572 & 0.4958553 & 0.6863397 & -0.9545717 & 0.9654288 & 0.9838265 & 1.1610592 & 0.0996542 & 0.2393445 & -0.5791692 & 0.3501431 & 0.5558993 & -0.7342619 & 0.9783036 & 0.9900678 & 0.8829109 & 0.9915772 & 0.9963117 & -3.926817 & 0.7895585 & 0.8895260 & -3.706969 & 0.5164039 & 0.7028663 & -2.200317 & 0.0909694 & 0.2244290 & 2.218886 & 0.6982871 & 0.8331540 & -1.433851 & 0.6691637 & 0.8141329 & 7.711609\\
+\hline
+ENSMUST00000094955 & 0.0081879 & 0.0367991 & 1.1856925 & 0.9353016 & 0.9690575 & -0.6333013 & 0.3745215 & 0.5795488 & 2.5957177 & 0.0050684 & 0.0252354 & -0.6988290 & 0.6729944 & 0.8166022 & 1.0178334 & 0.0948389 & 0.2310919 & 0.8228212 & 0.0253932 & 0.0879330 & 1.1482199 & 0.3657233 & 0.5710470 & -1.3750913 & 0.2915369 & 0.4951562 & -3.968832 & 0.5819450 & 0.7532377 & -1.276993 & 0.6373529 & 0.7923844 & 1.548320 & 0.8614351 & 0.9300767 & -4.288514 & 0.8886550 & 0.9448186 & -4.232771 & 0.1879584 & 0.3707972 & 6.442072\\
+\hline
+NM\_001042489 & 0.4191775 & 0.6208106 & 1.2123141 & 0.7607212 & 0.8723405 & 0.5992596 & 0.7748772 & 0.8807955 & 1.1476237 & 0.1995820 & 0.3858600 & 2.6613947 & 0.0205303 & 0.0748885 & -0.8188268 & 0.1910652 & 0.3748543 & -0.5443243 & 0.8270177 & 0.9111679 & 2.7899965 & 0.5239967 & 0.7088274 & -1.3479903 & 0.1409795 & 0.3049960 & -6.548630 & 0.4062041 & 0.6090027 & 1.889101 & 0.4968748 & 0.6871546 & -3.392918 & 0.0003701 & 0.0030031 & -6.639293 & 0.0000135 & 0.0001774 & -5.488106 & 0.0256136 & 0.0884953 & -4.976920\\
+\hline
+NM\_008159 & 0.1231754 & 0.2777536 & 1.1022344 & 0.1169243 & 0.2678638 & 0.7374376 & 0.5781815 & 0.7504472 & 0.9406648 & 0.0059119 & 0.0284812 & -0.8908867 & 0.4381368 & 0.6375400 & 1.2644788 & 0.8998830 & 0.9506910 & -0.5490689 & 0.5634704 & 0.7394176 & -1.4223473 & 0.5015939 & 0.6910094 & -2.3313130 & 0.0000000 & 0.0000000 & 1.321606 & 0.0000000 & 0.0000001 & -2.339338 & 0.0000000 & 0.0000000 & -3.096598 & 0.0000007 & 0.0000123 & -5.350299 & 0.0000125 & 0.0001658 & -6.076895 & 0.0000010 & 0.0000176 & -4.774939\\
+\hline
+NM\_001013813 & 0.0565287 & 0.1593800 & 0.6299949 & 0.1710643 & 0.3480241 & 0.9746582 & 0.0951123 & 0.2315686 & 3.3644621 & 0.4387357 & 0.6380771 & 0.7308201 & 0.0005373 & 0.0040980 & 1.2229788 & 0.3295522 & 0.5353121 & 1.7341606 & 0.0007915 & 0.0056472 & -1.0267100 & 0.0013222 & 0.0086036 & -0.9327780 & 0.0010519 & 0.0071337 & 1.314872 & 0.1685588 & 0.3445806 & 1.623828 & 0.0554666 & 0.1571666 & -4.828044 & 0.0067413 & 0.0315747 & -5.772946 & 0.0006393 & 0.0047334 & -5.776763 & 0.0004591 & 0.0035968 & -5.017664\\
+\hline
+AK039774 & 0.0150416 & 0.0590150 & 0.6372239 & 0.0622016 & 0.1707939 & 0.9714819 & 0.0038649 & 0.0203447 & 1.3168792 & 0.9071986 & 0.9543873 & -0.6495320 & 0.0337680 & 0.1089499 & 1.4620498 & 0.1808765 & 0.3613288 & -0.6084973 & 0.4268483 & 0.6276079 & -1.4219238 & 0.8931509 & 0.9471666 & 0.6812377 & 0.3320222 & 0.5378422 & 1.725274 & 0.0000084 & 0.0001177 & 2.756976 & 0.7523631 & 0.8671706 & -4.265131 & 0.0000850 & 0.0008681 & -5.010475 & 0.0000014 & 0.0000238 & -4.623531 & 0.0066328 & 0.0311737 & -4.795976\\
+\hline
+NM\_013782 & 0.8754233 & 0.9377109 & 0.5940058 & 0.0001107 & 0.0010876 & 1.1803029 & 0.4731855 & 0.6676244 & 1.9274813 & 0.3558245 & 0.5614957 & -0.6697288 & 0.0000009 & 0.0000157 & 1.2377208 & 0.0469473 & 0.1390186 & -1.2969383 & 0.0877739 & 0.2187959 & 0.7832039 & 0.0873966 & 0.2181127 & -0.7705369 & 0.3054160 & 0.5099832 & 1.240504 & 0.0593928 & 0.1651796 & 2.355748 & 0.4210031 & 0.6224586 & -3.879986 & 0.0053367 & 0.0262800 & -5.108367 & 0.0056116 & 0.0273406 & -4.262393 & 0.0042107 & 0.0217896 & -4.617477\\
+\hline
+NM\_028622 & 0.7667946 & 0.8759265 & 1.6477918 & 0.9457494 & 0.9741115 & -1.8648471 & 0.9903079 & 0.9957216 & 1.1660022 & 0.6332637 & 0.7895226 & 2.1081788 & 0.8592452 & 0.9288844 & 1.2212238 & 0.7658228 & 0.8753935 & -1.1079164 & 0.8331999 & 0.9145679 & -0.8793087 & 0.8359000 & 0.9160123 & -1.2284455 & 0.8894396 & 0.9452334 & 1.160648 & 0.5797431 & 0.7516226 & -3.323953 & 0.5675779 & 0.7424636 & -4.125536 & 0.6699768 & 0.8146554 & -6.126729 & 0.6521300 & 0.8025109 & -4.510424 & 0.8460618 & 0.9215885 & 6.351333\\
+\hline
+\end{tabular}
+
 We can repeat this for each of the conditions
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-11-1} \end{center}
 
@@
 
-To leverage this temporal structure, Storey et al \citep{storey:significance} proposed to
-model each gene in time-course micro-array with a splines function, and to use a log-ratio likelihood test to detect differentially expressed genes.
-
-\texttt{moanin} extends
-this idea by providing functionality to compare time course data between different treatment conditions, using a similar mechanism of contrasts -- only now the contrasts are differences between the estimated means.
+To leverage this temporal structure, Storey et al \citep{storey:significance}
+proposed to model each gene in time-course micro-array with a splines
+function, and to use a log-ratio likelihood test to detect differentially
+expressed genes.
+
+\texttt{moanin} extends this idea by providing functionality to compare time course
+data between different treatment conditions, using a similar mechanism of
+contrasts -- only now the contrasts are differences between the estimated
+means.
 
@@
 
-\textbf{EAP: Do we need this call to limma by the user, or can these not be done internally by the \texttt{timecourse\_differential\_expression\_analysis} function?
-We can still preserve the option of user defining their own, but would be nicer to make this simpler for the user (e.g.~an if clause that tests whether input is matrix or character vector)}
+\textbf{EAP: Do we need this call to limma by the user, or can these not be done
+internally by the \texttt{timecourse\_differential\_expression\_analysis} function? We
+can still preserve the option of user defining their own, but would be nicer
+to make this simpler for the user (e.g.~an if clause that tests whether input
+is matrix or character vector)}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-10-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-17-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}
 

### 2019-05-15 scripts/manuscript.Rmd
Remove all namespace calls

@@
 - switch to camel case function names
-- shorten function names
 - Remove the namespace call
-- Move the normalization to a supplementary data
 - Make clear that it applies to RNASeq and how to call the different functions
@@
 
-**EAP: General comment. Your function names are incredibly long compared to
-standard functions. Also, will want to get rid of `moanin::` in front of
-functions. It's more precise, but an average user doesn't ever experience the
-`::` command; usually just package creaters.**
-
 # Analysis of the dynamical response of mouse lung tissue to influenza
@@
 variance_per_genes = apply(data, 1, mad)
-min_variance = stats::quantile(variance_per_genes, c(variance_cutoff))
+min_variance = quantile(variance_per_genes, c(variance_cutoff))
 variance_filtered_data = data[variance_per_genes > min_variance,]
@@
 ```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
-graphics::hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
+hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
      ylab="# probes", main="")
-graphics::abline(v=min_variance, col="#AB0000", lw=2)
+abline(v=min_variance, col="#AB0000", lw=2)
 ```
@@
 
-pca_data = stats::prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
+pca_data = prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
 percent_var = round(100 * attr(pca_data, "percentVar"))
@@
 ```{r pca_plots, fig.width=6, fig.height=6}
-graphics::par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
-graphics::plot(
+par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
+plot(
     pca_data$x[, "PC2"], pca_data$x[, "PC1"],
@@
     xlab="PC2", ylab="PC1")
-graphics::plot(
+plot(
     pca_data$x[, "PC2"], pca_data$x[, "PC1"],
@@
 
-graphics::plot(
+plot(
     pca_data$x[, "PC2"], pca_data$x[, "PC3"],
@@
     xlab="PC2", ylab="PC3")
-graphics::plot(
+plot(
     pca_data$x[, "PC2"], pca_data$x[, "PC3"],
@@
 variance_filtered_data = variance_filtered_data[, ord]
-data_corr = stats::cor(variance_filtered_data, method="pearson")
+data_corr = cor(variance_filtered_data, method="pearson")
 data_corr_meta = meta[ord, ]
@@
 
-NMF::aheatmap(
+aheatmap(
     data_corr,
@@
 If no formula is provided, it will default to the following: 
-`formula = ~Group:splines::ns(Timepoint, df=degrees_of_fredoom) + Group + 0`
+`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`
 
 ```{r}
-splines_model = moanin::create_splines_model(meta, degrees_of_freedom=6)
+splines_model = create_splines_model(meta, degrees_of_freedom=6)
 ```
@@
 # Define contrast	
-weekly_de_analysis =  moanin::DE_timepoints(
+weekly_de_analysis = DE_timepoints(
      data, splines_model, contrasts,
@@
 contrasts = create_timepoints_contrasts("M", "K", splines_model)
-K_weekly_de_analysis =  moanin::DE_timepoints(
+K_weekly_de_analysis =  DE_timepoints(
      data, splines_model, contrasts,
@@
 contrasts = create_timepoints_contrasts("M", "C", splines_model)
-C_weekly_de_analysis =  moanin::DE_timepoints(
+C_weekly_de_analysis =  DE_timepoints(
      data, splines_model, contrasts,
@@
 contrasts = create_timepoints_contrasts("M", "VH", splines_model)
-VH_weekly_de_analysis =  moanin::DE_timepoints(
+VH_weekly_de_analysis =  DE_timepoints(
      data, splines_model, contrasts,
@@
 
-contrasts = limma::makeContrasts(
+contrasts = makeContrasts(
 	contrasts=timecourse_contrasts,
@@
 # treatment).
-pvalues = moanin::DE_timecourse(
+pvalues = DE_timecourse(
     data, splines_model, contrasts,
     use_voom_weights=FALSE)
-qvalues = apply(pvalues, 2, stats::p.adjust)
+qvalues = apply(pvalues, 2, p.adjust)
 
@@
 ```{r}
-log_fold_change_timepoints = moanin::estimate_log_fold_change(
+log_fold_change_timepoints = estimate_log_fold_change(
     data, splines_model, contrasts,  method="timely")
@@
 ```{r}
-log_fold_change_epicon = moanin::estimate_log_fold_change(
+log_fold_change_epicon = estimate_log_fold_change(
     data, splines_model, contrasts,  method="epicon")
 
-log_fold_change_sum = moanin::estimate_log_fold_change(
+log_fold_change_sum = estimate_log_fold_change(
     data, splines_model, contrasts,  method="sum")
 
-log_fold_change_max = moanin::estimate_log_fold_change(
+log_fold_change_max = estimate_log_fold_change(
     data, splines_model, contrasts, method="max")
 
-log_fold_change_min = moanin::estimate_log_fold_change(
+log_fold_change_min = estimate_log_fold_change(
     data, splines_model, contrasts, method="min")
@@
 top_DE_genes_pval = names(sort(pvalue)[1:12])
-moanin::plot_genes(data[top_DE_genes_pval, ], splines_model,
+plot_genes(data[top_DE_genes_pval, ], splines_model,
 		   colors=ann_colors$Group, smooth=TRUE)
@@
 	 decreasing=TRUE)[1:12])
-moanin::plot_genes(data[top_DE_genes_lfc, ], splines_model,
+plot_genes(data[top_DE_genes_lfc, ], splines_model,
 		   colors=ann_colors$Group, smooth=TRUE)
@@
 # Filter out q-values for the pvalues table
-fishers_pval = moanin::pvalues_fisher_method(pvalues)
-fishers_qval = stats::p.adjust(fishers_pval)
+fishers_pval = pvalues_fisher_method(pvalues)
+fishers_qval = p.adjust(fishers_pval)
 
@@
 # First fit the kmeans clusters
-kmeans_clusters = moanin::splines_kmeans(
+kmeans_clusters = splines_kmeans(
     y, splines_model, n_clusters=8,
@@
 ```{r fig.height=6, fig.width=6}
-moanin::plot_centroids(kmeans_clusters$centroids, splines_model,
+plot_centroids(kmeans_clusters$centroids, splines_model,
 		       colors=ann_colors$Group,
@@
 # scoring function.
-scores_and_labels = moanin::splines_kmeans_score_and_label(
+scores_and_labels = splines_kmeans_score_and_label(
     data, kmeans_clusters)
@@
 dim(text) = dim(confusion_matrix)
-NMF::aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
+aheatmap(confusion_matrix, treeheight=0, main="Confusion matrix", 
 	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
@@
 
-res = NMF::aheatmap(
+res = aheatmap(
     data_to_plot,
@@
 # Now use the results of the previous call to aheatmap to reorder the genes.
-NMF::aheatmap(
+aheatmap(
     moanin:::rescale_values(data_to_plot)[res$rowInd,],
@@
 for(n_cluster in all_possible_n_clusters){
-    clustering_results = moanin::splines_kmeans(
+    clustering_results = splines_kmeans(
 	y, splines_model,
@@
 average_silhouette = function(labels, y) {
-    silhouette_results = cluster::silhouette(unlist(labels[1]), dist(y))
+    silhouette_results = silhouette(unlist(labels[1]), dist(y))
     return(mean(silhouette_results[, 3]))
@@
 stability_5 = read.table("results/stability_5.tsv", sep="\t")
-consensus_matrix_stability_5 = moanin::consensus_matrix(stability_5,
+consensus_matrix_stability_5 = consensus_matrix(stability_5,
 							scale=FALSE)
-NMF::aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
+aheatmap(consensus_matrix_stability_5[1:1000, 1:1000], Rowv=FALSE,
 	 Colv=FALSE,
@@
 stability_20 = read.table("results/stability_20.tsv", sep="\t")
-consensus_matrix_stability_20 = moanin::consensus_matrix(stability_20,
+consensus_matrix_stability_20 = consensus_matrix(stability_20,
 						 scale=FALSE)
-NMF::aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
+aheatmap(consensus_matrix_stability_20[1:1000, 1:1000], Rowv=FALSE,
 	 Colv=FALSE,
@@
 ```{r fig.width=6, fig.height=6}
-moanin::plot_model_explorer(all_labels)
+plot_model_explorer(all_labels)
 ```
@@
 selected_labels = all_labels[clusters]
-moanin::plot_model_explorer(selected_labels)
+plot_model_explorer(selected_labels)
 ```
@@
 ```{r eval=FALSE}
-moanin::plot_cdf_consensus(all_labels)
+plot_cdf_consensus(all_labels)
 ```
@@
 ```{r results="hide"}
-ensembl = biomaRt::useMart("ensembl")
-ensembl = biomaRt::useDataset("mmusculus_gene_ensembl", mart=ensembl)
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
 ```
@@
 # convert gene names
-genes = biomaRt::getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
 		filters="refseq_mrna", values=genes,
@@
 genes = as.vector(unlist(genes))
-pathways = KEGGprofile::find_enriched_pathway(
+pathways = find_enriched_pathway(
     genes, species="mmu",
@@
 ```{r go_to_gene_mapping, results="hide", message=FALSE}
-genes = biomaRt::getBM(attributes=c("go_id", "refseq_mrna"),
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
 	      values=gene_names,
@@
 # Create gene to GO id mapping
-gene_id_go_mapping = moanin::create_go_term_mapping(genes)
+gene_id_go_mapping = create_go_term_mapping(genes)
 ```

### 2019-05-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \item
-  shorten function names
-\item
   Remove the namespace call
-\item
-  Move the normalization to a supplementary data
 \item
@@
 
-\textbf{EAP: General comment. Your function names are incredibly long compared to
-standard functions. Also, will want to get rid of \texttt{moanin::} in front of
-functions. It's more precise, but an average user doesn't ever experience the
-\texttt{::} command; usually just package creaters.}
-
 \hypertarget{analysis-of-the-dynamical-response-of-mouse-lung-tissue-to-influenza}{%
@@
 \NormalTok{variance_per_genes =}\StringTok{ }\KeywordTok{apply}\NormalTok{(data, }\DecValTok{1}\NormalTok{, mad)}
-\NormalTok{min_variance =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{quantile}\NormalTok{(variance_per_genes, }\KeywordTok{c}\NormalTok{(variance_cutoff))}
+\NormalTok{min_variance =}\StringTok{ }\KeywordTok{quantile}\NormalTok{(variance_per_genes, }\KeywordTok{c}\NormalTok{(variance_cutoff))}
 \NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{data[variance_per_genes }\OperatorTok{>}\StringTok{ }\NormalTok{min_variance,]}
@@
 
-\NormalTok{pca_data =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
+\NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
 \NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\NormalTok{graphics}\OperatorTok{::}\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{))}
-\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(}
 \NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
@@
     \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}
-\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
+\KeywordTok{plot}\NormalTok{(}
 \NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
@@
 
-\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
+\KeywordTok{plot}\NormalTok{(}
 \NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
@@
     \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
-\NormalTok{graphics}\OperatorTok{::}\KeywordTok{plot}\NormalTok{(}
+\KeywordTok{plot}\NormalTok{(}
 \NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
@@
 \NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
-\NormalTok{data_corr =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
+\NormalTok{data_corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
 \NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
@@
 
-\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
+\KeywordTok{aheatmap}\NormalTok{(}
 \NormalTok{    data_corr,}
@@
 If no formula is provided, it will default to the following:
-\texttt{formula\ =\ \textasciitilde{}Group:splines::ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{splines_model =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+\texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{splines_model =}\StringTok{ }\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
 \end{Highlighting}
@@
 \CommentTok{# Define contrast   }
-\NormalTok{weekly_de_analysis =}\StringTok{  }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{DE_timepoints}\NormalTok{(}
+\NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{(}
 \NormalTok{     data, splines_model, contrasts,}
@@
 
-\NormalTok{contrasts =}\StringTok{ }\NormalTok{limma}\OperatorTok{::}\KeywordTok{makeContrasts}\NormalTok{(}
+\NormalTok{contrasts =}\StringTok{ }\KeywordTok{makeContrasts}\NormalTok{(}
     \DataTypeTok{contrasts=}\NormalTok{timecourse_contrasts,}
@@
 \CommentTok{# treatment).}
-\NormalTok{pvalues =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{DE_timecourse}\NormalTok{(}
+\NormalTok{pvalues =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
-\NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, stats}\OperatorTok{::}\NormalTok{p.adjust)}
+\NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, p.adjust)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{log_fold_change_timepoints =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\NormalTok{log_fold_change_epicon =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{log_fold_change_epicon =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}
 
-\NormalTok{log_fold_change_sum =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
 
-\NormalTok{log_fold_change_max =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
 
-\NormalTok{log_fold_change_min =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
 \NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
@@
 \NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
+\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
            \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
@@
      \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
+\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
            \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
@@
 \CommentTok{# Filter out q-values for the pvalues table}
-\NormalTok{fishers_pval =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
-\NormalTok{fishers_qval =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}
+\NormalTok{fishers_pval =}\StringTok{ }\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
+\NormalTok{fishers_qval =}\StringTok{ }\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}
 
@@
 \CommentTok{# First fit the kmeans clusters}
-\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
+\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
 \NormalTok{    y, splines_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
@@
 \begin{Highlighting}[]
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_centroids}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
+\KeywordTok{plot_centroids}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
                \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
@@
 \CommentTok{# scoring function.}
-\NormalTok{scores_and_labels =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
+\NormalTok{scores_and_labels =}\StringTok{ }\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
 \NormalTok{    data, kmeans_clusters)}
@@
 
-\NormalTok{res =}\StringTok{ }\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
+\NormalTok{res =}\StringTok{ }\KeywordTok{aheatmap}\NormalTok{(}
 \NormalTok{    data_to_plot,}
@@
 \CommentTok{# Now use the results of the previous call to aheatmap to reorder the genes.}
-\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(}
+\KeywordTok{aheatmap}\NormalTok{(}
 \NormalTok{    moanin}\OperatorTok{:::}\KeywordTok{rescale_values}\NormalTok{(data_to_plot)[res}\OperatorTok{$}\NormalTok{rowInd,],}
@@
 \ControlFlowTok{for}\NormalTok{(n_cluster }\ControlFlowTok{in}\NormalTok{ all_possible_n_clusters)\{}
-\NormalTok{    clustering_results =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
+\NormalTok{    clustering_results =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
 \NormalTok{    y, splines_model,}
@@
 \NormalTok{average_silhouette =}\StringTok{ }\ControlFlowTok{function}\NormalTok{(labels, y) \{}
-\NormalTok{    silhouette_results =}\StringTok{ }\NormalTok{cluster}\OperatorTok{::}\KeywordTok{silhouette}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(labels[}\DecValTok{1}\NormalTok{]), }\KeywordTok{dist}\NormalTok{(y))}
+\NormalTok{    silhouette_results =}\StringTok{ }\KeywordTok{silhouette}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(labels[}\DecValTok{1}\NormalTok{]), }\KeywordTok{dist}\NormalTok{(y))}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{mean}\NormalTok{(silhouette_results[, }\DecValTok{3}\NormalTok{]))}
@@
 \NormalTok{stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_5.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
-\NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
+\NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
                             \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
-\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{5}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
+\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{5}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
      \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
@@
 \NormalTok{stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_20.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
-\NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
+\NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
                          \DataTypeTok{scale=}\OtherTok{FALSE}\NormalTok{)}
-\NormalTok{NMF}\OperatorTok{::}\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{20}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
+\KeywordTok{aheatmap}\NormalTok{(consensus_matrix_stability_}\DecValTok{20}\NormalTok{[}\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{1000}\NormalTok{], }\DataTypeTok{Rowv=}\OtherTok{FALSE}\NormalTok{,}
      \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
@@
 \begin{Highlighting}[]
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_model_explorer}\NormalTok{(all_labels)}
+\KeywordTok{plot_model_explorer}\NormalTok{(all_labels)}
 \end{Highlighting}
@@
 \NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_model_explorer}\NormalTok{(selected_labels)}
+\KeywordTok{plot_model_explorer}\NormalTok{(selected_labels)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{moanin}\OperatorTok{::}\KeywordTok{plot_cdf_consensus}\NormalTok{(all_labels)}
+\KeywordTok{plot_cdf_consensus}\NormalTok{(all_labels)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{ensembl =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{useMart}\NormalTok{(}\StringTok{"ensembl"}\NormalTok{)}
-\NormalTok{ensembl =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{useDataset}\NormalTok{(}\StringTok{"mmusculus_gene_ensembl"}\NormalTok{, }\DataTypeTok{mart=}\NormalTok{ensembl)}
+\NormalTok{ensembl =}\StringTok{ }\KeywordTok{useMart}\NormalTok{(}\StringTok{"ensembl"}\NormalTok{)}
+\NormalTok{ensembl =}\StringTok{ }\KeywordTok{useDataset}\NormalTok{(}\StringTok{"mmusculus_gene_ensembl"}\NormalTok{, }\DataTypeTok{mart=}\NormalTok{ensembl)}
 \end{Highlighting}
@@
 \CommentTok{# convert gene names}
-\NormalTok{genes =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene"}\NormalTok{),}
+\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene"}\NormalTok{),}
         \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
@@
 \NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
-\NormalTok{pathways =}\StringTok{ }\NormalTok{KEGGprofile}\OperatorTok{::}\KeywordTok{find_enriched_pathway}\NormalTok{(}
+\NormalTok{pathways =}\StringTok{ }\KeywordTok{find_enriched_pathway}\NormalTok{(}
 \NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
@@
 \begin{Highlighting}[]
-\NormalTok{genes =}\StringTok{ }\NormalTok{biomaRt}\OperatorTok{::}\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"go_id"}\NormalTok{, }\StringTok{"refseq_mrna"}\NormalTok{),}
+\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"go_id"}\NormalTok{, }\StringTok{"refseq_mrna"}\NormalTok{),}
           \DataTypeTok{values=}\NormalTok{gene_names,}
@@
 \CommentTok{# Create gene to GO id mapping}
-\NormalTok{gene_id_go_mapping =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_go_term_mapping}\NormalTok{(genes)}
+\NormalTok{gene_id_go_mapping =}\StringTok{ }\KeywordTok{create_go_term_mapping}\NormalTok{(genes)}
 \end{Highlighting}

### 2019-07-19 scripts/manuscript.Rmd
small changes to manuscript

@@
 
+EAP:
+- Do we need installation instructions for moanin, since its not on CRAN/bioconductor?
+Also, how its set up now (doing make file, which compiles the documentation, etc), does that prohibit use of, e.g.
+`install_github` from `devtools`?
+I.e.
+not something we can demonstrate in the markdown file.
+
 # Introduction
@@
 through to single-cell technologies. While many expression studies are
-designed to compare the gene expression in distinct groups, there is also a
-long history of time-course expression studies, where the the gene expression
-is compared across time by measuring mRNA levels from different samples across
-time [^1]. Such time course studies can vary from measuring a few distinct time
-points, to sampling ten to twenty time points. Many longer time series are
+designed to compare the gene expression between distinct groups, there is also a
+long history of time-course expression studies. Such studies compare the  gene expression
+across time by measuring mRNA levels from samples collected at different timepoints [^1]. Such time-course studies can vary from measuring a few distinct time
+points, to sampling ten to twenty time points. These longer time series are
 particularly interested in investigating development over time. More recently,
-single-cell studies track single cells through their development, and a single
-cell is measured at a particular moment in its developmental progression -- a
+a new variety of time course studies have come from single-cell sequencing experiments**EAP: need citation** which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
 value that is not know but estimated from the data as its "pseudo-time."
@@
 longitudinal studies, for example tracking the same subject over time, are
-certainly possible in certain settings, but not directly considered here.
+certainly possible, but not directly considered here.
 
@@
 time-course permutation tests [@park:statistical], and multiple testing
-strategies to combine singe time point differential expression analysis
+strategies to combine single time point differential expression analysis
 [@sun:multiple]. However, with long time course datasets, particularly in
 developmental systems, a massive number of genes will show *some* change. For
-example, in the mice lung tissues infected with influenza, over 50% of genes
+example, in a study of mice lung tissues infected with influenza that we consider in this workflow, over 50% of genes
 are shown to be changing over time. The task in these settings is often not to
-detect changes in genes, but to categorize into biologically interpretable
-patterns the vast number of changes discovered.
+detect changes in genes, but to categorize them into biologically interpretable
+patterns.
 
@@
 
+**EAP: We need to add comment about microarray versus sequencing data**
+
 By combining gene expression time-course data with virus growth data, the
@@
 an example of this for the influenza data using generic methods; these steps
-are not in specific to time course, but could be done for any gene expression
+are not in specific to time course data, but could be done for any gene expression
 analysis.
@@
 First let's load the data. The package `moanin` contains the normalized
-data and meta of [@shoemaker:ultrasensitive].
+data and metadata of [@shoemaker:ultrasensitive].
 

### 2019-07-19 scripts/manuscript.Rmd
added questions about installation in manuscript

@@
 EAP:
-- Do we need installation instructions for moanin, since its not on CRAN/bioconductor?
-Also, how its set up now (doing make file, which compiles the documentation, etc), does that prohibit use of, e.g.
-`install_github` from `devtools`?
-I.e.
-not something we can demonstrate in the markdown file.
+
+- Do we need installation instructions for moanin in this document, since its not on CRAN/bioconductor?
+Also, how its set up now (doing make file, which compiles the documentation, etc), probably prohibits use of, e.g.
+`install_github` from `devtools`.
+If able to call `install_github("NelleV/2019timecourse-rnaseq-pipeline", subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also install the dependencies and get around problems in script)
+- I got an error in installation of packages:
+
+```
+epurdom-0:2019timecourse-rnaseq-pipeline epurdom$ Rscript scripts/install.R
+Loading required package: devtools
+Loading required package: usethis
+Loading required package: ggfortify
+Loading required package: ggplot2
+Loading required package: kableExtra
+Error in contrib.url(repos, "source") : 
+  trying to use CRAN without setting a mirror
+Calls: install.packages -> contrib.url
+In addition: Warning message:
+In library(package, lib.loc = lib.loc, character.only = TRUE, logical.return = TRUE,  :
+  there is no package called ‘kableExtra’
+Execution halted
+```
+
+- `biocLite` is no longer the recommended way to install bioconductor. Instead `BiocManager::install(c('edge', 'NMI'))`. But this is dependent on bioconductor/R version.
 

### 2019-07-19 scripts/manuscript.Rmd
comments about installation

@@
 If able to call `install_github("NelleV/2019timecourse-rnaseq-pipeline", subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also install the dependencies and get around problems in script)
-- I got an error in installation of packages:
+- I got an error in installation script for  packages:
 
@@
 
-- `biocLite` is no longer the recommended way to install bioconductor. Instead `BiocManager::install(c('edge', 'NMI'))`. But this is dependent on bioconductor/R version.
+- `biocLite` is no longer the recommended way to install bioconductor.
+Instead `BiocManager::install(c('edge', 'NMI'))`.
+But this is dependent on bioconductor/R version.
+Also, now that docker, can you change the installation to just be simple one-liner (and `BiocManager::install` can also do non-bioc packages, and is much more reliable.)
 

### 2019-07-20 scripts/manuscript.Rmd
fix error in Rmd file

@@
 Also, now that docker, can you change the installation to just be simple one-liner (and `BiocManager::install` can also do non-bioc packages, and is much more reliable.)
+
+- I get multiple errors in compilation of file.
+Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there.
+Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache.
+One error I fixed, but wasn't sure about this error (changed it to `entrezgene_id` and seems to be running now...):
+
+```
+label: enrichment_analysis (with options) 
+List of 2
+ $ results: chr "hide"
+ $ message: logi FALSE
+
+Quitting from lines 1124-1138 (manuscript.Rmd) 
+Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
+  Invalid attribute(s): entrezgene 
+Please use the function 'listAttributes' to get valid attribute names
+Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
+```
 
@@
 write.table(kmeans_labels,
-	    ".results/clustering_labels.txt",
+	    "./results/clustering_labels.txt",
 	    sep="\t")
@@
 # convert gene names
-genes = getBM(attributes=c("ensembl_gene_id", "entrezgene"),
+genes = getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
 		filters="refseq_mrna", values=genes,
-		mart=ensembl)["entrezgene"]
+		mart=ensembl)["entrezgene_id"]
 genes = as.vector(unlist(genes))

### 2019-07-22 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
   Make clear that it applies to RNASeq and how to call the different functions
-  toRNASeq
+  to RNASeq
 \item
@@
 \end{itemize}
+
+EAP:
+
+\begin{itemize}
+\tightlist
+\item
+  Do we need installation instructions for moanin in this document, since its not on CRAN/bioconductor?
+Also, how its set up now (doing make file, which compiles the documentation, etc), probably prohibits use of, e.g.
+\texttt{install\_github} from \texttt{devtools}.
+If able to call \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also install the dependencies and get around problems in script)
+\item
+  I got an error in installation script for packages:
+\end{itemize}
+
+\begin{verbatim}
+epurdom-0:2019timecourse-rnaseq-pipeline epurdom$ Rscript scripts/install.R
+Loading required package: devtools
+Loading required package: usethis
+Loading required package: ggfortify
+Loading required package: ggplot2
+Loading required package: kableExtra
+Error in contrib.url(repos, "source") : 
+  trying to use CRAN without setting a mirror
+Calls: install.packages -> contrib.url
+In addition: Warning message:
+In library(package, lib.loc = lib.loc, character.only = TRUE, logical.return = TRUE,  :
+  there is no package called <e2><80><98>kableExtra<e2><80><99>
+Execution halted
+\end{verbatim}
+
+\begin{itemize}
+\item
+  \texttt{biocLite} is no longer the recommended way to install bioconductor.
+Instead \texttt{BiocManager::install(c("edge",\ "NMI"))}.
+But this is dependent on bioconductor/R version.
+Also, now that docker, can you change the installation to just be simple one-liner (and \texttt{BiocManager::install} can also do non-bioc packages, and is much more reliable.)
+\item
+  I get multiple errors in compilation of file.
+Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there.
+Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache.
+One error I fixed, but wasn't sure about this error (changed it to \texttt{entrezgene\_id} and seems to be running now\ldots{}):
+\end{itemize}
+
+\begin{verbatim}
+label: enrichment_analysis (with options) 
+List of 2
+ $ results: chr "hide"
+ $ message: logi FALSE
+
+Quitting from lines 1124-1138 (manuscript.Rmd) 
+Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
+  Invalid attribute(s): entrezgene 
+Please use the function 'listAttributes' to get valid attribute names
+Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
+\end{verbatim}
 
@@
 through to single-cell technologies. While many expression studies are
-designed to compare the gene expression in distinct groups, there is also a
-long history of time-course expression studies, where the the gene expression
-is compared across time by measuring mRNA levels from different samples across
-time \footnote{Because the collection of the mRNA is often destructive, samples at
+designed to compare the gene expression between distinct groups, there is also a
+long history of time-course expression studies. Such studies compare the gene expression
+across time by measuring mRNA levels from samples collected at different timepoints \footnote{Because the collection of the mRNA is often destructive, samples at
   different time points are generally from different biological samples;
   longitudinal studies, for example tracking the same subject over time, are
-  certainly possible in certain settings, but not directly considered here.}. Such time course studies can vary from measuring a few distinct time
-points, to sampling ten to twenty time points. Many longer time series are
+  certainly possible, but not directly considered here.}. Such time-course studies can vary from measuring a few distinct time
+points, to sampling ten to twenty time points. These longer time series are
 particularly interested in investigating development over time. More recently,
-single-cell studies track single cells through their development, and a single
-cell is measured at a particular moment in its developmental progression -- a
+a new variety of time course studies have come from single-cell sequencing experiments\textbf{EAP: need citation} which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
 value that is not know but estimated from the data as its ``pseudo-time.''
@@
 time-course permutation tests \citep{park:statistical}, and multiple testing
-strategies to combine singe time point differential expression analysis
+strategies to combine single time point differential expression analysis
 \citep{sun:multiple}. However, with long time course datasets, particularly in
 developmental systems, a massive number of genes will show \emph{some} change. For
-example, in the mice lung tissues infected with influenza, over 50\% of genes
+example, in a study of mice lung tissues infected with influenza that we consider in this workflow, over 50\% of genes
 are shown to be changing over time. The task in these settings is often not to
-detect changes in genes, but to categorize into biologically interpretable
-patterns the vast number of changes discovered.
+detect changes in genes, but to categorize them into biologically interpretable
+patterns.
 
 We present here a workflow for such an analysis that consists of 4 main
-parts (Figure \ref{fig:workflow}):
+parts (Figure \ref{fig:schema}):
 
@@
 
+\textbf{EAP: We need to add comment about microarray versus sequencing data}
+
 By combining gene expression time-course data with virus growth data, the
@@
 First let's load the data. The package \texttt{moanin} contains the normalized
-data and meta of \citep{shoemaker:ultrasensitive}.
+data and metadata of \citep{shoemaker:ultrasensitive}.
 
@@
 \CommentTok{# convert gene names}
-\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene"}\NormalTok{),}
+\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene_id"}\NormalTok{),}
         \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
-        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene"}\NormalTok{]}
+        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene_id"}\NormalTok{]}
 \NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
@@
 \hline
+04668 & TNF signaling pathway & 45 & 0\\
+\hline
 05164 & Influenza A & 38 & 0\\
-\hline
-04668 & TNF signaling pathway & 45 & 0\\
 \hline
@@
 \hline
-5 & GO:0003002 & regionalization & 172 & 153 & 134.46 & 0.00024 & 0.2416\\
-\hline
-6 & GO:0048562 & embryonic organ morphogenesis & 154 & 137 & 120.39 & 0.00035 & 0.2936\\
-\hline
-7 & GO:0010970 & transport along microtubule & 81 & 74 & 63.32 & 0.00057 & 0.4098\\
-\hline
-8 & GO:0060173 & limb development & 92 & 84 & 71.92 & 0.00070 & 0.4404\\
-\hline
-9 & GO:0060070 & canonical Wnt signaling pathway & 147 & 130 & 114.92 & 0.00090 & 0.5033\\
-\hline
-10 & GO:0055114 & oxidation-reduction process & 531 & 457 & 415.10 & 0.00135 & 0.6223\\
-\hline
-11 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.43 & 0.00136 & 0.6223\\
-\hline
-12 & GO:0006805 & xenobiotic metabolic process & 44 & 42 & 34.40 & 0.00168 & 0.7046\\
-\hline
-13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.75 & 0.00224 & 0.8197\\
-\hline
-14 & GO:0010811 & positive regulation of cell-substrate ad... & 78 & 71 & 60.98 & 0.00228 & 0.8197\\
-\hline
-15 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.80 & 0.00298 & 0.9999\\
-\hline
-16 & GO:0051056 & regulation of small GTPase mediated sign... & 151 & 131 & 118.04 & 0.00349 & 1.0000\\
-\hline
-17 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
-\hline
-18 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.20 & 0.00441 & 1.0000\\
-\hline
-19 & GO:0007368 & determination of left/right symmetry & 67 & 61 & 52.38 & 0.00464 & 1.0000\\
+5 & GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.2629\\
+\hline
+6 & GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.3323\\
+\hline
+7 & GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.3323\\
+\hline
+8 & GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.3729\\
+\hline
+9 & GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.3820\\
+\hline
+10 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.6194\\
+\hline
+11 & GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.6194\\
+\hline
+12 & GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.6194\\
+\hline
+13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.8517\\
+\hline
+14 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.0000\\
+\hline
+15 & GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.0000\\
+\hline
+16 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
+\hline
+17 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
+\hline
+18 & GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
+\hline
+19 & GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
 \hline
@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.3.4             pkgconfig_2.0.2        
+## [65] rlang_0.4.0             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          

### 2019-07-24 scripts/manuscript.Rmd
fix figure reference
ENH Moving moanin to it's own repository

@@
 
-library(moanin)
-
 source("utils.R")
@@
 
+```{r}
+library(devtools)
+install_github("NelleV/moanin", dependencies=FALSE)
+
+library(moanin)
+```
+
 EAP:
 
-- Do we need installation instructions for moanin in this document, since its not on CRAN/bioconductor?
-Also, how its set up now (doing make file, which compiles the documentation, etc), probably prohibits use of, e.g.
-`install_github` from `devtools`.
-If able to call `install_github("NelleV/2019timecourse-rnaseq-pipeline", subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also install the dependencies and get around problems in script)
+- Do we need installation instructions for moanin in this document, since its
+  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
+  compiles the documentation, etc), probably prohibits use of, e.g.
+  `install_github` from `devtools`. If able to call
+  `install_github("NelleV/2019timecourse-rnaseq-pipeline",
+  subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also
+  install the dependencies and get around problems in script)
 - I got an error in installation script for  packages:
@@
 
-- `biocLite` is no longer the recommended way to install bioconductor.
-Instead `BiocManager::install(c('edge', 'NMI'))`.
-But this is dependent on bioconductor/R version.
-Also, now that docker, can you change the installation to just be simple one-liner (and `BiocManager::install` can also do non-bioc packages, and is much more reliable.)
-
-- I get multiple errors in compilation of file.
-Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there.
-Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache.
-One error I fixed, but wasn't sure about this error (changed it to `entrezgene_id` and seems to be running now...):
-
-```
-label: enrichment_analysis (with options) 
-List of 2
- $ results: chr "hide"
- $ message: logi FALSE
-
-Quitting from lines 1124-1138 (manuscript.Rmd) 
-Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
-  Invalid attribute(s): entrezgene 
-Please use the function 'listAttributes' to get valid attribute names
-Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
-```
+- `biocLite` is no longer the recommended way to install bioconductor. Instead
+  `BiocManager::install(c("edge", "NMI"))`. But this is dependent on
+  bioconductor/R version. Also, now that docker, can you change the
+  installation to just be simple one-liner (and `BiocManager::install` can
+  also do non-bioc packages, and is much more reliable.)
+
+- I get multiple errors in compilation of file. Not sure, however, why
+  TravisCI didn't hit these problems, so you might want to see what's going on
+  there. Probably caching or something, so might want to occassionally remove
+  the cache and run again, if TravisCI is saving the cache. One error I fixed,
+  but wasn't sure about this error (changed it to `entrezgene_id` and seems to
+  be running now...):
 
@@
 We present here a workflow for such an analysis that consists of 4 main
-parts (Figure \ref{fig:workflow}):
+parts (Figure \ref{fig:schema}):
 

### 2019-07-24 .build_tools/travis/test_script.sh

@@ -18,18 +18,11 @@ get_data() {
 
 run_tests() {
     # first run the actual tests
-    pushd bin/moanin
-    make test
-    popd
     pushd scripts
     make all
     popd
 }
 
-# Start by installing
-pushd bin/moanin
-make install
-popd
 
 # Now get data and run the tests and build the manuscript
 get_data

### 2019-07-24 bin/moanin/.scripts/build_documentation.R

@@ -1,3 +1,0 @@
-library("devtools")
-
-devtools::document()

### 2019-07-24 bin/moanin/.scripts/test_package_cov.R

@@ -1,2 +1,0 @@
-library(covr)
-report()

### 2019-07-24 bin/moanin/R/DE_timepoints.R

@@ -1,109 +1,0 @@
-library("limma")
-library("stats")
-library("edgeR")
-
-
-#' Fit weekly differential expression analysis
-#'
-#' @param data Gene expression data
-#' @param splines_model splines_model
-#' @param contrasts Contrast to use.
-#' @param use_voom_weights boolean: whether to use voom weights or not
-#'
-#' @export
-DE_timepoints = function(data, splines_model,
-			 contrasts,
-			 use_voom_weights=TRUE){
-    meta = splines_model$meta
-
-    design = stats::model.matrix(~WeeklyGroup + 0, data=meta)
-
-    cleaned_colnames = gsub("WeeklyGroup", "", colnames(design))
-    colnames(design) = cleaned_colnames
-
-    allcontrasts = limma::makeContrasts(
-        contrasts=contrasts,
-        levels=design)
-
-    if(use_voom_weights){
-        y = edgeR::DGEList(counts=data)
-	y = edgeR::calcNormFactors(y, method="upperquartile")
-        v = limma::voom(y, design, plot=FALSE)
-	v = limma::lmFit(v)
-    }else{
-	v = limma::lmFit(data, design)	
-    }
-
-    fit = limma::contrasts.fit(v, allcontrasts)
-    fit = limma::eBayes(fit)
-    contrast_names = colnames(fit$p.value)
-    fit$adj.p.value = stats::p.adjust(fit$p.value, method="BH")
-    dim(fit$adj.p.value) = dim(fit$p.value)
-    colnames(fit$adj.p.value) = contrast_names
-
-    combine_results = function(ii, fit2){
-	contrast_formula = contrasts[ii]
-	de_analysis = data.frame(row.names=row.names(data))
-
-        base_colname = gsub(" ", "", contrast_formula, fixed=TRUE)
-	colname_pval = paste(base_colname, "-pval", sep="")
-	colname_qval = paste(base_colname, "-qval", sep="")
-	colname_lfc = paste(base_colname, "-lfc", sep="")
-
-        tt = limma::topTable(
-            fit2, coef=ii, number=length(rownames(fit2$coef)),
-            p.value=1, adjust.method="none",
-            genelist=rownames(fit2$coef))
-	de_analysis[colname_pval] = fit2$p.value[, contrast_formula]
-	de_analysis[colname_qval] = fit2$adj.p.value[,  contrast_formula]
-	de_analysis[colname_lfc] = tt$logFC
-	return(de_analysis)
-    }
-
[…]

### 2019-07-24 bin/moanin/R/splines.R

@@ -1,214 +1,0 @@
-# Splines utility
-
-library(splines)
-library(MASS)
-library(stats)
-
-
-#' Fit splines
-#'
-#' @param data the data
-#' @param splines_model splines_model 
-#' @param weights weigts
-#'
-#' @return beta coefficients
-#'
-#' @export
-fit_splines = function(data, splines_model, weights=NULL){
-    basis = splines_model$basis
-    n = ncol(basis)
-    nr = nrow(data)
-    basis = splines_model$basis 
-
-    if(!is.null(weights)){
-	beta = matrix(nrow=nr, ncol=n)
-	for(i in 1:nr){
-	    beta[i,] = stats::lm.wfit(basis, data[i,], weights[i,])$coefficients
-        }
-	row.names(beta) = row.names(data)
-    }else{
-	beta = t(stats::lm.fit(basis, t(data))$coefficients)
-    }
-    return(beta)
-}
-
-#' Fit and predict splines
-#'
-#' @param data the data
-#' @param splines_model splines_model
-#' @param weights weights
-#'
-#' @return y_fitted the fitted y values
-#'
-#' @export
-fit_predict_splines = function(data, splines_model, weights=NULL, meta_prediction=NULL){
-    basis = splines_model$basis
-    meta = splines_model$meta
-    if(!is.null(weights)){
-	stop("moanin::fit_predict_splines: not implemented")
-    }
-    if(is.null(meta_prediction)){
-        y_fitted = t(stats::lm.fit(basis, t(data))$fitted.values)
-    }else{
-	degrees_of_freedom = splines_model$degrees_of_freedom
-	fitting_data = t(as.matrix(data))
-	formula_data = list(
-	    "Group"=meta$Group,
-	    "Timepoint"=meta$Timepoint,
-	    "fitting_data"=fitting_data,
-	    "degrees_of_freedom"=splines_model$degrees_of_freedom)
-
-	updated_formula = stats::update(splines_model$formula, fitting_data ~ .)
-	model = stats::lm(updated_formula, formula_data)	
-	y_fitted = stats::predict(model, meta_prediction)
-    }
-    return(y_fitted)
-}
-
-
-#' Create prediction meta data from splines model
-#'
-#' @param splines_model a Splines model object
-#' @param num_timepoints integer, optional, default: 100
-#'	Number of timepoints to use for the prediction metadata
-#'
-#' @export
[…]

### 2019-07-24 bin/moanin/R/visualization.R

@@ -1,129 +1,0 @@
-library(viridis)
-library(graphics)
-
-
-#' Plotting centroids
-#'
-#' @param centroids matrix (k, t) containing the centroids
-#' @param splines_model splines_model
-#' @param meta	data.frame (t, n) containing the metadata.
-#' @param colors vector, optional, default NULL
-#'		vector of colors
-#' @param smooth boolean, optional, default: FALSE
-#'  Whether to smooth the centroids or not.
-#' @export
-plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
-    n_centroids = dim(centroids)[1]
-    if(n_centroids <= 3){
-        graphics::par(
-	    mfrow=c(n_centroids, 1),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else if(n_centroids <= 6){
-        n_col = ceiling(n_centroids / 2)
-        graphics::par(
-	    mfrow=c(n_col, 2),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else if(n_centroids <= 12){
-        ncol = ceiling(n_centroids / 3)
-        graphics::par(
-	    mfrow=c(ncol, 3),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else{
-        nrow = round(n_centroids ** 0.5)
-        ncol = ceiling(n_centroids / nrow)
-        graphics::par(
-	    mfrow=c(ncol, nrow),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }
-
-    name_centroids = names(centroids)
-    name_centroid = NULL
-    for(i in 1:n_centroids){
-	if(!is.null(name_centroids)){
-	    name_centroid = name_centroids[i]
-	}
-        plot_centroid_individual(as.vector(centroids[i, ]),
-				 splines_model, colors=colors,
-				 smooth=smooth,
-				 title=name_centroid)
-    }
-}
-
-
-#' Plotting data
-#'
-#' @param data matrix (k, t) containing the centroids
-#' @param splines_model splines_model
-#' @param meta	data.frame (t, n) containing the metadata.
-#' @param colors vector, optional, default NULL
-#'		vector of colors
-#' @param smooth boolean, optional, default: FALSE
-#'  Whether to smooth the centroids or not.
-#' @export
-plot_genes = function(data, splines_model, colors=NULL, smooth=FALSE){
-    plot_centroids(data, splines_model, colors=colors, smooth=smooth)
-}
-
-
[…]

### 2019-07-24 bin/moanin/tests/testthat/test-cluster_validation.R

@@ -1,16 +1,0 @@
-library("moanin")
-
-context("moanin::cluster_validation.R")
-
-test_that("cluster_validation::consensus_matrix", {
-    set.seed(42)
-    n_genes = 40
-    n_clusters = 5
-    n_labels = 20
-    labels = sample.int(n_labels, n_clusters * n_genes, replace=TRUE)
-    dim(labels) = c(n_genes, n_clusters)
-    colnames(labels) = lapply(1:n_clusters, function(x){paste0("C", x)})
-    row.names(labels) = lapply(1:n_genes, function(x){paste0("G", x)})
-
-    expect_silent(consensus_matrix(labels))
-})

### 2019-07-25 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(devtools)}
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{, }\DataTypeTok{dependencies=}\OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##   
+   checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION'
+##
+
+-  preparing 'moanin':
+##    checking DESCRIPTION meta-information ...
+
+v  checking DESCRIPTION meta-information
+##
+
+-  checking for LF line-endings in source and make files and shell scripts
+##
+
+-  checking for empty or unneeded directories
+## -  looking to see if a 'data/datalist' file should be added
+##
+
+-  building 'moanin_0.0.0.tar.gz' (3.5s)
+##
+
+## 
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(moanin)}
+\end{Highlighting}
+\end{Shaded}
+
 EAP:
@@
 \item
-  Do we need installation instructions for moanin in this document, since its not on CRAN/bioconductor?
-Also, how its set up now (doing make file, which compiles the documentation, etc), probably prohibits use of, e.g.
-\texttt{install\_github} from \texttt{devtools}.
-If able to call \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also install the dependencies and get around problems in script)
+  Do we need installation instructions for moanin in this document, since its
+  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
+  compiles the documentation, etc), probably prohibits use of, e.g.
+  \texttt{install\_github} from \texttt{devtools}. If able to call
+  \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also
+  install the dependencies and get around problems in script)
 \item
@@
 \item
-  \texttt{biocLite} is no longer the recommended way to install bioconductor.
-Instead \texttt{BiocManager::install(c("edge",\ "NMI"))}.
-But this is dependent on bioconductor/R version.
-Also, now that docker, can you change the installation to just be simple one-liner (and \texttt{BiocManager::install} can also do non-bioc packages, and is much more reliable.)
-\item
-  I get multiple errors in compilation of file.
-Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there.
-Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache.
-One error I fixed, but wasn't sure about this error (changed it to \texttt{entrezgene\_id} and seems to be running now\ldots{}):
+  \texttt{biocLite} is no longer the recommended way to install bioconductor. Instead
+  \texttt{BiocManager::install(c("edge",\ "NMI"))}. But this is dependent on
+  bioconductor/R version. Also, now that docker, can you change the
+  installation to just be simple one-liner (and \texttt{BiocManager::install} can
+  also do non-bioc packages, and is much more reliable.)
+\item
+  I get multiple errors in compilation of file. Not sure, however, why
+  TravisCI didn't hit these problems, so you might want to see what's going on
+  there. Probably caching or something, so might want to occassionally remove
+  the cache and run again, if TravisCI is saving the cache. One error I fixed,
+  but wasn't sure about this error (changed it to \texttt{entrezgene\_id} and seems to
+  be running now\ldots{}):
 \end{itemize}
-
-\begin{verbatim}
-label: enrichment_analysis (with options) 
-List of 2
- $ results: chr "hide"
- $ message: logi FALSE
-
-Quitting from lines 1124-1138 (manuscript.Rmd) 
-Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
-  Invalid attribute(s): entrezgene 
-Please use the function 'listAttributes' to get valid attribute names
-Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
-\end{verbatim}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-5-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-6-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-11-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}
 
@@
 ## [49] MASS_7.3-51.4           scales_1.0.0           
-## [51] hms_0.4.2               curl_3.3               
-## [53] yaml_2.2.0              memoise_1.1.0          
+## [51] hms_0.4.2               yaml_2.2.0             
+## [53] curl_3.3                memoise_1.1.0          
 ## [55] gridExtra_2.3           TeachingDemos_2.10     
@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.0             pkgconfig_2.0.2        
+## [65] rlang_0.3.4             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          
@@
 ## [85] crayon_1.3.4            progress_1.2.0         
-## [87] grid_3.6.0              blob_1.1.1             
-## [89] callr_3.2.0             git2r_0.25.2           
-## [91] vegan_2.5-4             digest_0.6.18          
+## [87] grid_3.6.0              vegan_2.5-4            
+## [89] blob_1.1.1              callr_3.2.0            
+## [91] git2r_0.25.2            digest_0.6.18          
 ## [93] webshot_0.5.1           xtable_1.8-4           

### 2019-07-25 scripts/manuscript.Rmd
Updated the list of things TODO

@@
   also do non-bioc packages, and is much more reliable.)
-
-- I get multiple errors in compilation of file. Not sure, however, why
-  TravisCI didn't hit these problems, so you might want to see what's going on
-  there. Probably caching or something, so might want to occassionally remove
-  the cache and run again, if TravisCI is saving the cache. One error I fixed,
-  but wasn't sure about this error (changed it to `entrezgene_id` and seems to
-  be running now...):
 

### 2019-07-25 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ##   
-   checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpsJhqa7/remotes352cb8e01/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpsJhqa7/remotes352cb8e01/NelleV-moanin-205e154/DESCRIPTION'
 ##
@@
 
--  building 'moanin_0.0.0.tar.gz' (3.5s)
+-  building 'moanin_0.0.0.tar.gz' (3.7s)
 ##

### 2019-07-27 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

--- 2019-07-25 manuscript.tex
@@
 ##   
-   checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpHrqs5v/remotes341f57a29b/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION'
 ##
@@
 \begin{itemize}
+\tightlist
 \item
@@
   also do non-bioc packages, and is much more reliable.)
-\item
-  I get multiple errors in compilation of file. Not sure, however, why
-  TravisCI didn't hit these problems, so you might want to see what's going on
-  there. Probably caching or something, so might want to occassionally remove
-  the cache and run again, if TravisCI is saving the cache. One error I fixed,
-  but wasn't sure about this error (changed it to \texttt{entrezgene\_id} and seems to
-  be running now\ldots{}):
 \end{itemize}
@@
 \hline
+04064 & NF-kappa B signaling pathway & 39 & 0\\
+\hline
 04620 & Toll-like receptor signaling pathway & 40 & 0\\
-\hline
-05167 & Kaposi sarcoma-associated herpesvirus infection & 29 & 0\\
 \hline

### 2019-07-31 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ##   
-   checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpWctcMt/remotes3475772588/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpWctcMt/remotes3475772588/NelleV-moanin-205e154/DESCRIPTION'
 ##
@@
 
--  building 'moanin_0.0.0.tar.gz' (3.5s)
+-  building 'moanin_0.0.0.tar.gz' (3.6s)
 ##

### 2019-08-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ##   
-   checking for file '/tmp/RtmpWctcMt/remotes3475772588/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpWctcMt/remotes3475772588/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION'
 ##

### 2019-08-05 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

--- 2019-07-27 manuscript.tex
@@
 ##   
-   checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpLM5IuQ/remotes347f085d5d/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmplEyT0t/remotes345348f481/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmplEyT0t/remotes345348f481/NelleV-moanin-205e154/DESCRIPTION'
 ##

### 2019-08-07 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

--- 2019-08-02 manuscript.tex
@@
 ##   
-   checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpL9TZ4Q/remotes34218b1b1b/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpL9TZ4Q/remotes34218b1b1b/NelleV-moanin-205e154/DESCRIPTION'
 ##
@@
 
--  building 'moanin_0.0.0.tar.gz' (3.6s)
+-  building 'moanin_0.0.0.tar.gz' (4.4s)
 ##

### 2019-08-08 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

--- 2019-08-02 manuscript.tex
@@
 ##   
-   checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpmpNfbV/remotes355c2a7bc/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION'
 ##
@@
 
--  building 'moanin_0.0.0.tar.gz' (3.6s)
+-  building 'moanin_0.0.0.tar.gz' (3.9s)
 ##

### 2019-08-08 scripts/manuscript.Rmd
update for legend in PCA plot

--- 2019-07-20 scripts/manuscript.Rmd
@@
 - `biocLite` is no longer the recommended way to install bioconductor.
-Instead `BiocManager::install(c('edge', 'NMI'))`.
+Instead `BiocManager::install(c("edge", "NMI"))`.
 But this is dependent on bioconductor/R version.
@@
 ```{r pca, fig.width=6, fig.height=6}
-# Reorder genes on condition, time, and replicate
-
 pca_data = prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
@@
 
-We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate. **EAP: Unclear what 'markers' refers to**
-We also plot the second and third components in the second row.
+We then plot the two first PC components, and color each sample by (1) its
+condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
+**EAP: The colors for timepoint appear to circle around, i.e. the first and last are very similar colors**
 
 ```{r pca_plots, fig.width=6, fig.height=6}
-par(mfrow=c(2, 2), mar=c(2.5, 2.5, 2.5, 2.5))
+par(mfrow=c(2, 2))
+par(mar=c(4.5, 4.5, 1.5, 1.5))
 plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+    pca_data$x[, c("PC1","PC2")],
+    col=ann_colors$Group[meta$Group],
+    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
+mtext(at=275,text="Colored by Group", line=0.5 ,font=2,adj=0.5)
+legend(x=250,y=25,legend=names(ann_colors$Group), fill=ann_colors$Group,bty="n",xpd=NA,yjust=0.5)
+par(mar=c(4.5, 5.5, 1.5, .1))
+plot(
+    pca_data$x[, c("PC3","PC2")],
     col=ann_colors$Group[meta$Group],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
+    ylab="", bty="l", yaxt="n")
+axis(2, labels=FALSE)
+par(mar=c(4.5, 4.5, 1.5, 1.5))
 plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC1"],
+    pca_data$x[, c("PC1","PC2")],
+    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
+    pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
+mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
+legend(x=250,y=25,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
+par(mar=c(4.5, 5.5, 1.5, .1))
+plot(
+    pca_data$x[, c("PC3","PC2")],
     col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
     pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC1")
-
-plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Group[meta$Group],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
-plot(
-    pca_data$x[, "PC2"], pca_data$x[, "PC3"],
-    col=ann_colors$Timepoint[as.factor(meta$Timepoint)],
-    pch=ann_markers$Replicate[as.factor(meta$Replicate)],
-    xlab="PC2", ylab="PC3")
+    ylab="", bty="l", yaxt="n")
+axis(2, labels=FALSE)
 
@@
 ```{r  correlation_plot, fig.width=6, fig.height=6, fig.caption="Cross-correlation of the samples"}
-
 # Reorder the conditions such that:
@@
 data_corr_meta$Timepoint = as.factor(data_corr_meta$Timepoint)
-
+# use aheatmap from the NMF package for a heatmap
 aheatmap(

### 2019-08-09 scripts/manuscript.Rmd
update plot_centroids for legend, etc

@@
 plot_genes(data[top_DE_genes_pval, ], splines_model,
-		   colors=ann_colors$Group, smooth=TRUE)
+    colors=ann_colors$Group, smooth=TRUE,
+    mar=c(1.5,2.5,2,0.1),legendArgs=list(title="Treatment"),
+    mfrow=c(3,5))
 ```
@@
     colors=ann_colors$Group,
-		       smooth=TRUE)
+    smooth=TRUE,legend=TRUE,
+    legendArgs=list(title="Treatment",cex=2))
 ```

### 2019-08-09 bin/moanin/R/visualization.R

@@ -11,41 +11,62 @@ library(graphics)
 #'		vector of colors
 #' @param smooth boolean, optional, default: FALSE
 #'  Whether to smooth the centroids or not.
+#' @param legend boolean whether to include a legend (default:TRUE)
+#' @param legendArgs list of arguments to be passed to legend command (if \code{legend=TRUE})
+#' @param ... arguments to be passed to the individual plot commands
+#'  (Will be sent to all plot commands)
+#' @param mar vector of coordinates to set the space around each plot (see \code{\link{par}})
+#' @param simpleY boolean, if true, will minimize the annotation of the y axis to 
+#'  only label the axis in the exterior plots (the x-axis is always assumed to be the 
+#'  same across all plots and will always be simplified)
+#' @param mfrow a vector of integers of length 2 defining the grid of plots to be created (see \code{\link{par}}). If missing, the function will set a value.
 #' @export
-plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE){
+plot_centroids = function(centroids, splines_model, colors=NULL, smooth=FALSE, legend=TRUE, legendArgs=NULL,simpleY=TRUE, mar=c(2.5, 2.5, 3.0, 1.0),mfrow=NULL,...){
     n_centroids = dim(centroids)[1]
-    if(n_centroids <= 3){
-        graphics::par(
-	    mfrow=c(n_centroids, 1),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else if(n_centroids <= 6){
-        n_col = ceiling(n_centroids / 2)
-        graphics::par(
-	    mfrow=c(n_col, 2),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else if(n_centroids <= 12){
-        ncol = ceiling(n_centroids / 3)
-        graphics::par(
-	    mfrow=c(ncol, 3),
-            mar=c(1.0, 1.0, 1.0, 1.0))
-    }else{
-        nrow = round(n_centroids ** 0.5)
-        ncol = ceiling(n_centroids / nrow)
-        graphics::par(
-	    mfrow=c(ncol, nrow),
-            mar=c(1.0, 1.0, 1.0, 1.0))
+    n_plots<-if(legend) n_centroids+1 else n_centroids
+    if(!is.null(mfrow)){
[…]

### 2019-08-09 scripts/manuscript.Rmd

@@
 
-- `biocLite` is no longer the recommended way to install bioconductor.
-Instead `BiocManager::install(c("edge", "NMI"))`.
-But this is dependent on bioconductor/R version.
-Also, now that docker, can you change the installation to just be simple one-liner (and `BiocManager::install` can also do non-bioc packages, and is much more reliable.)
-
-- I get multiple errors in compilation of file.
-Not sure, however, why TravisCI didn't hit these problems, so you might want to see what's going on there.
-Probably caching or something, so might want to occassionally remove the cache and run again, if TravisCI is saving the cache.
-One error I fixed, but wasn't sure about this error (changed it to `entrezgene_id` and seems to be running now...):
-
-```
-label: enrichment_analysis (with options) 
-List of 2
- $ results: chr "hide"
- $ message: logi FALSE
-
-Quitting from lines 1124-1138 (manuscript.Rmd) 
-Error in getBM(attributes = c("ensembl_gene_id", "entrezgene"), filters = "refseq_mrna",  : 
-  Invalid attribute(s): entrezgene 
-Please use the function 'listAttributes' to get valid attribute names
-Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> getBM
-```
+- `biocLite` is no longer the recommended way to install bioconductor. Instead
+  `BiocManager::install(c("edge", "NMI"))`. But this is dependent on
+  bioconductor/R version. Also, now that docker, can you change the
+  installation to just be simple one-liner (and `BiocManager::install` can
+  also do non-bioc packages, and is much more reliable.)
 
@@
 As another sanity check, `moanin` provides a simple utility function to
-visualize gene time-course data. Here, we plot the 12 genes with the smallest
+visualize gene time-course data. Here, we plot the 10 genes with the smallest
 p-values.
@@
 ```{r fig.height=6, fig.width=6}
-top_DE_genes_pval = names(sort(pvalue)[1:12])
+top_DE_genes_pval = names(sort(pvalue)[1:10])
 plot_genes(data[top_DE_genes_pval, ], splines_model,
     colors=ann_colors$Group, smooth=TRUE,
-    mar=c(1.5,2.5,2,0.1),legendArgs=list(title="Treatment"),
-    mfrow=c(3,5))
+    mar=c(1.5,2.5,2,0.1), legendArgs=list(title="Treatment"))
 ```
@@
     sort(abs(lfc_epicon),
-	 decreasing=TRUE)[1:12])
+	 decreasing=TRUE)[1:10])
 plot_genes(data[top_DE_genes_lfc, ], splines_model,
-		   colors=ann_colors$Group, smooth=TRUE)
+    colors=ann_colors$Group, smooth=TRUE,
+    mar=c(1.5,2.5,2,0.1), legendArgs=list(title="Treatment"))
 

### 2019-08-09 scripts/manuscript.Rmd
remove changes that require new plot_centroids

@@
     colors=ann_colors$Group, smooth=TRUE,
-    mar=c(1.5,2.5,2,0.1), legendArgs=list(title="Treatment"))
+    mar=c(1.5,2.5,2,0.1))
 ```
@@
     colors=ann_colors$Group, smooth=TRUE,
-    mar=c(1.5,2.5,2,0.1), legendArgs=list(title="Treatment"))
+    mar=c(1.5,2.5,2,0.1))
 
@@
     colors=ann_colors$Group,
-    smooth=TRUE,legend=TRUE,
-    legendArgs=list(title="Treatment",cex=2))
+    smooth=TRUE)
 ```

### 2019-08-09 scripts/manuscript.Rmd
changes to manuscript

@@
 
-- Do we need installation instructions for moanin in this document, since its
-  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
-  compiles the documentation, etc), probably prohibits use of, e.g.
-  `install_github` from `devtools`. If able to call
-  `install_github("NelleV/2019timecourse-rnaseq-pipeline",
-  subdir="bin/moanin", dependencies=TRUE)` would be simpler (would also
-  install the dependencies and get around problems in script)
+- Why `dependicies=FALSE`? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine. 
 - I got an error in installation script for  packages:
@@
 mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
-legend(x=250,y=25,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
+legend(x=250,y=-15,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
 par(mar=c(4.5, 5.5, 1.5, .1))
@@
 axis(2, labels=FALSE)
-
 ```
@@
 interest:
+
 ```{r make_timepoints_contarts}
@@
 
+This creates a vector of contrasts to be tested, one for each timepoint:
+
 ```{r echo=FALSE}
-kable(contrasts)
-```
+kable(contrasts,col.names="Contrasts")
+```
+**EAP: Not sure if `kable` is better than just printout?**
 
 Then run the differential expression analysis on all of those timepoints
-jointly.
+jointly using the function `DE_timepoints`.
 
@@
 # Define contrast	
-weekly_de_analysis = DE_timepoints(
+VL_weekly_de_analysis = DE_timepoints(
      data, splines_model, contrasts,
@@
 ```
+**EAP: Sort of strange to have argument `spines_model` for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument**
+
+This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (`pval`), log-fold change (`lfc`) and adjusted p-value (`qval`); the order of the genes in the table is the same as the input `data`.
+Here we show the results for the first two timepoints and the first ten genes:
 
 ```{r echo=FALSE}
-kable(weekly_de_analysis[1:10, ])
-```
-
-We can repeat this for each of the conditions
+kable(weekly_de_analysis[1:10, 1:6 ],digits=3)
+```
+
+**EAP: For the output of this, the column names are in format `M.0-VL.0-pval`, for example.
+This may be hard to parse algorithmically down the road because you are using the same character for the contrast (`-`) as the separator for the `pval`.
+So for example, if you wanted to split the contrast name from the `pval` or `lfc` part, it wouldn't be very easy.
+Maybe underscore `_` would be better separator?**
+
+We will repeat this, comparing each of the treatments to the control ("M") (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).
 
@@
 
+**EAP: Should `timely_barplot` be a function of `moanin`?**
+
 ```{r fig.width=6, fig.height=6, echo=FALSE}
-par(mfrow=c(2, 2))
-
-time = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
-
-timely_barplot = function(de_analysis, labels, qval_threshold=0.05, title=""){
+timely_barplot = function(de_analysis, labels=NULL, qval_threshold=0.05, title="",...){
     qval_colnames = colnames(de_analysis)[
 	grepl("qval", colnames(de_analysis))]
+    if(is.null(labels)) labels<-sapply(strsplit(gsub("-qval","",qval_colnames),"\\."),.subset2,3)
     number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
-    names(number_de_genes_per_time) = labels[1]
-    barplot(number_de_genes_per_time, col="black", ylim=c(0, 20000),
-	    xlab="Timepoint", ylab="Number of DE genes", main=title)
+    barplot(number_de_genes_per_time, col="black", 
+            names.arg=labels,ylim=c(0, 20000),
+	    xlab="Timepoint", ylab="Number of DE genes", main=title,...)
 
@@
 
-timely_barplot(K_weekly_de_analysis, time, title="Kawasaki")
-timely_barplot(C_weekly_de_analysis, time, title="California")
-timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose")
-timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose")
-```
-
-The distribution of number of genes found differentially expressed by
-considering each time-point independantly highlights the challenges of such
-approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
-genes found differentially expressed than timepoints 3H and 9H: this is likely
-due to biological or technical variances for specific genes at specific
-timepoints. **EAP: We should add more to this**
+par(mfrow=c(2, 2))
+timely_barplot(K_weekly_de_analysis, time, title="Kawasaki",las=3)
+timely_barplot(C_weekly_de_analysis, time, title="California",las=3)
+timely_barplot(VL_weekly_de_analysis, time, title="Vietnam low-dose",las=3)
+timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose",las=3)
+```
+
+Such an analysis can demonstrate some general trends, with clearly more genes being differentially expressed at later time points, and the Vietnam high-dose showing perhaps an earlier onset than the low-dose.
+
+However, the distribution of the number of genes found differentially expressed by
+considering each time-point independently highlights the challenges of such
+approach.
+We can see that some timepoints have many less genes found significantly differentially expressed (e.g.
+timepoint 6H and 18H of the Kawasaki strain).
+While there may be biological differences at those time points for some genes, it seems unlikely that the large majority of genes differentially expressed at timepoint 3H stop being differentially expressed at 6H and then jump back to being differentially expressed at 9H.
+A more likely explanation is that there is are some technical or biological artifact about the samples for 6H that is creating higher variation and thus less ability to detect significance.
+
+```{r calculateCombos, echo=FALSE, results="hide"}
+#Code for counting the number of unique combinations of time points that are DE
+getTimepoint<-function(x){sapply(strsplit(gsub("-qval","",x),"\\."),.subset2,3)}
+qval_colnames = colnames(K_weekly_de_analysis)[
+grepl("qval", colnames(K_weekly_de_analysis))]
+signifCombos<-apply(K_weekly_de_analysis[, qval_colnames],1,function(x){paste(getTimepoint(qval_colnames[which(x<0.05)]),collapse=",")})
+signifCombos<-signifCombos[signifCombos!=""]
+tabCombos<-table(signifCombos)
+```
+
+Another difficulty with such an approach is making sense of the general temporal structure for any particular gene, as different genes will have different combinations of timepoints DE.
+For the comparison of the Kawasaki strain to the control, for example, there are `length(signifCombos)` genes found DE in some timepoints, and there are `length(tabCombos)` different combinations of timepoints for which they are DE.
+Some of these make sense, such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but many are very fragmentary.
+For example there are `tabCombos["48,60,120,168"]` genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H.
+Many of these genes are likely to have not made the cutoff for significance in 72H, but don't show real differences in the overall trend between 48H-168H.
+
+Here are plots of the first 10 such genes, where we can see that some might show some meaningful changes between 60 and 72H, but others clearly just have a single replicate that is different or increased variability at 72H.
+
+```{r showMissingDEGenes, echo=FALSE}
+exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
+wh<-which(splines_model$meta$Group %in% c("K","M"))
+plot_genes(data[exampleGenes,],splines_model,colors=ann_colors$Group)
+```
+
+**EAP: It would be nice to be able to subset in `plot_centroids/plot_genes` to only some of the groups**
 
 As a summary, classic differential expression methods are appropriate for
-unordered treatments, but fail to use the temporal structure of the data.
+unordered treatments, but fail to make use of the temporal structure of the data.
 

### 2019-08-12 scripts/manuscript.Rmd
FIX to the code

@@
 ```{r echo=FALSE}
-kable(contrasts,col.names="Contrasts")
+kable(contrasts, col.names="Contrasts")
 ```
@@
 # Define contrast	
-VL_weekly_de_analysis = DE_timepoints(
+weekly_de_analysis = DE_timepoints(
      data, splines_model, contrasts,
@@
 ```{r fig.width=6, fig.height=6, echo=FALSE}
-timely_barplot = function(de_analysis, labels=NULL, qval_threshold=0.05, title="",...){
+timely_barplot = function(de_analysis, labels=NULL, qval_threshold=0.05,
+			  title="", ...){
     qval_colnames = colnames(de_analysis)[
 	grepl("qval", colnames(de_analysis))]
-    if(is.null(labels)) labels<-sapply(strsplit(gsub("-qval","",qval_colnames),"\\."),.subset2,3)
+    if(is.null(labels)){
+	labels = sapply(
+	    strsplit(gsub("-qval", "", qval_colnames), "\\."), .subset2, 3)
+    }
     number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
+
     barplot(number_de_genes_per_time, col="black", 
-            names.arg=labels,ylim=c(0, 20000),
-	    xlab="Timepoint", ylab="Number of DE genes", main=title,...)
-
+            # names.arg=labels,
+	    ylim=c(0, 20000),
+	    xlab="Timepoint", ylab="Number of DE genes", main=title, ...)  
 }
@@
 par(mfrow=c(2, 2))
-timely_barplot(K_weekly_de_analysis, time, title="Kawasaki",las=3)
-timely_barplot(C_weekly_de_analysis, time, title="California",las=3)
-timely_barplot(VL_weekly_de_analysis, time, title="Vietnam low-dose",las=3)
-timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose",las=3)
+timely_barplot(K_weekly_de_analysis, time, title="Kawasaki", las=3)
+timely_barplot(C_weekly_de_analysis, time, title="California", las=3)
+timely_barplot(VL_weekly_de_analysis, time, title="Vietnam low-dose", las=3)
+timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose", las=3)
 ```

### 2019-08-12 scripts/manuscript.Rmd
FIX names.arg doesn't work on my computer

@@
     barplot(number_de_genes_per_time, col="black", 
-            # names.arg=labels,
+            names.arg=labels,
 	    ylim=c(0, 20000),
@@
 timely_barplot(C_weekly_de_analysis, time, title="California", las=3)
-timely_barplot(VL_weekly_de_analysis, time, title="Vietnam low-dose", las=3)
+timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose", las=3)
 timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose", las=3)

### 2019-08-12 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ##   
-   checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpwKvOP3/remotes33736f6dc/NelleV-moanin-205e154/DESCRIPTION'
+   checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION' ...
+
+v  checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION'
 ##
@@
 
--  building 'moanin_0.0.0.tar.gz' (3.9s)
+-  building 'moanin_0.0.0.tar.gz' (4.2s)
 ##
@@
 \item
-  Do we need installation instructions for moanin in this document, since its
-  not on CRAN/bioconductor? Also, how its set up now (doing make file, which
-  compiles the documentation, etc), probably prohibits use of, e.g.
-  \texttt{install\_github} from \texttt{devtools}. If able to call
-  \texttt{install\_github("NelleV/2019timecourse-rnaseq-pipeline",\ subdir="bin/moanin",\ dependencies=TRUE)} would be simpler (would also
-  install the dependencies and get around problems in script)
+  Why \texttt{dependicies=FALSE}? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine.
 \item
@@
 \begin{Highlighting}[]
-\CommentTok{# Reorder genes on condition, time, and replicate}
-
 \NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
@@
 
-We then plot the two first components, and color each sample by (1) its
-condition; (2) its sampling time. We use different markers for each replicate. \textbf{EAP: Unclear what `markers' refers to}
-We also plot the second and third components in the second row.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{))}
+We then plot the two first PC components, and color each sample by (1) its
+condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
+\textbf{EAP: The colors for timepoint appear to circle around, i.e.~the first and last are very similar colors}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
 \KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
+\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
+    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
+    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
+\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\DecValTok{25}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(}
+\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
     \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
     \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}
+    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
+\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
 \KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC1"}\NormalTok{],}
+\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
+    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
+    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
+\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{,}\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(}
+\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
     \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
     \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC1"}\NormalTok{)}
-
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC2"}\NormalTok{], pca_data}\OperatorTok{$}\NormalTok{x[, }\StringTok{"PC3"}\NormalTok{],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{xlab=}\StringTok{"PC2"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"PC3"}\NormalTok{)}
+    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
+\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}
-
+\CommentTok{# use aheatmap from the NMF package for a heatmap}
 \KeywordTok{aheatmap}\NormalTok{(}
@@
 
+This creates a vector of contrasts to be tested, one for each timepoint:
+
 \begin{tabular}{l}
 \hline
-x\\
+Contrasts\\
 \hline
@@
 
+\textbf{EAP: Not sure if \texttt{kable} is better than just printout?}
+
 Then run the differential expression analysis on all of those timepoints
-jointly.
+jointly using the function \texttt{DE\_timepoints}.
 
@@
 
-\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
-\hline
-  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc & M.6-VL.6-pval & M.6-VL.6-qval & M.6-VL.6-lfc & M.9-VL.9-pval & M.9-VL.9-qval & M.9-VL.9-lfc & M.12-VL.12-pval & M.12-VL.12-qval & M.12-VL.12-lfc & M.18-VL.18-pval & M.18-VL.18-qval & M.18-VL.18-lfc & M.24-VL.24-pval & M.24-VL.24-qval & M.24-VL.24-lfc & M.30-VL.30-pval & M.30-VL.30-qval & M.30-VL.30-lfc & M.36-VL.36-pval & M.36-VL.36-qval & M.36-VL.36-lfc & M.48-VL.48-pval & M.48-VL.48-qval & M.48-VL.48-lfc & M.60-VL.60-pval & M.60-VL.60-qval & M.60-VL.60-lfc & M.72-VL.72-pval & M.72-VL.72-qval & M.72-VL.72-lfc & M.120-VL.120-pval & M.120-VL.120-qval & M.120-VL.120-lfc & M.168-VL.168-pval & M.168-VL.168-qval & M.168-VL.168-lfc\\
-\hline
-NM\_009912 & 0.2728476 & 0.4745864 & 2.2018266 & 0.7670600 & 0.8760739 & 1.1576881 & 0.0514865 & 0.1487818 & 2.5397431 & 0.1461489 & 0.3125183 & -0.7613983 & 0.2735411 & 0.4753685 & 1.4786596 & 0.5199686 & 0.7057241 & 1.0838768 & 0.0372513 & 0.1172476 & 1.0099581 & 0.1663710 & 0.3415394 & -2.0803981 & 0.0040198 & 0.0209941 & 1.867318 & 0.0000272 & 0.0003259 & -2.877945 & 0.0000001 & 0.0000025 & -2.200448 & 0.0000000 & 0.0000000 & -2.941491 & 0.0000000 & 0.0000000 & 2.659545 & 0.0000000 & 0.0000000 & -4.846082\\
-\hline
-NM\_008725 & 0.8508522 & 0.9242869 & 2.0281764 & 0.1390300 & 0.3020872 & 1.4568427 & 0.0034217 & 0.0184668 & 2.3999010 & 0.0011765 & 0.0078218 & -0.5946375 & 0.2218494 & 0.4142375 & 0.8974352 & 0.0490869 & 0.1436339 & 1.0689628 & 0.0001128 & 0.0011050 & -1.1910616 & 0.2358903 & 0.4313519 & 0.8409000 & 0.0025790 & 0.0147499 & 1.463357 & 0.0000005 & 0.0000098 & -3.732241 & 0.0992368 & 0.2386529 & -3.305718 & 0.2451045 & 0.4423855 & -5.832292 & 0.0028580 & 0.0160083 & 5.103924 & 0.3298546 & 0.5356392 & 6.770567\\
-\hline
-NM\_007473 & 0.0880394 & 0.2192542 & 0.6779618 & 0.9165990 & 0.9592675 & 0.7639279 & 0.6859316 & 0.8250549 & 1.2874572 & 0.4958553 & 0.6863397 & -0.9545717 & 0.9654288 & 0.9838265 & 1.1610592 & 0.0996542 & 0.2393445 & -0.5791692 & 0.3501431 & 0.5558993 & -0.7342619 & 0.9783036 & 0.9900678 & 0.8829109 & 0.9915772 & 0.9963117 & -3.926817 & 0.7895585 & 0.8895260 & -3.706969 & 0.5164039 & 0.7028663 & -2.200317 & 0.0909694 & 0.2244290 & 2.218886 & 0.6982871 & 0.8331540 & -1.433851 & 0.6691637 & 0.8141329 & 7.711609\\
-\hline
-ENSMUST00000094955 & 0.0081879 & 0.0367991 & 1.1856925 & 0.9353016 & 0.9690575 & -0.6333013 & 0.3745215 & 0.5795488 & 2.5957177 & 0.0050684 & 0.0252354 & -0.6988290 & 0.6729944 & 0.8166022 & 1.0178334 & 0.0948389 & 0.2310919 & 0.8228212 & 0.0253932 & 0.0879330 & 1.1482199 & 0.3657233 & 0.5710470 & -1.3750913 & 0.2915369 & 0.4951562 & -3.968832 & 0.5819450 & 0.7532377 & -1.276993 & 0.6373529 & 0.7923844 & 1.548320 & 0.8614351 & 0.9300767 & -4.288514 & 0.8886550 & 0.9448186 & -4.232771 & 0.1879584 & 0.3707972 & 6.442072\\
-\hline
-NM\_001042489 & 0.4191775 & 0.6208106 & 1.2123141 & 0.7607212 & 0.8723405 & 0.5992596 & 0.7748772 & 0.8807955 & 1.1476237 & 0.1995820 & 0.3858600 & 2.6613947 & 0.0205303 & 0.0748885 & -0.8188268 & 0.1910652 & 0.3748543 & -0.5443243 & 0.8270177 & 0.9111679 & 2.7899965 & 0.5239967 & 0.7088274 & -1.3479903 & 0.1409795 & 0.3049960 & -6.548630 & 0.4062041 & 0.6090027 & 1.889101 & 0.4968748 & 0.6871546 & -3.392918 & 0.0003701 & 0.0030031 & -6.639293 & 0.0000135 & 0.0001774 & -5.488106 & 0.0256136 & 0.0884953 & -4.976920\\
-\hline
-NM\_008159 & 0.1231754 & 0.2777536 & 1.1022344 & 0.1169243 & 0.2678638 & 0.7374376 & 0.5781815 & 0.7504472 & 0.9406648 & 0.0059119 & 0.0284812 & -0.8908867 & 0.4381368 & 0.6375400 & 1.2644788 & 0.8998830 & 0.9506910 & -0.5490689 & 0.5634704 & 0.7394176 & -1.4223473 & 0.5015939 & 0.6910094 & -2.3313130 & 0.0000000 & 0.0000000 & 1.321606 & 0.0000000 & 0.0000001 & -2.339338 & 0.0000000 & 0.0000000 & -3.096598 & 0.0000007 & 0.0000123 & -5.350299 & 0.0000125 & 0.0001658 & -6.076895 & 0.0000010 & 0.0000176 & -4.774939\\
-\hline
-NM\_001013813 & 0.0565287 & 0.1593800 & 0.6299949 & 0.1710643 & 0.3480241 & 0.9746582 & 0.0951123 & 0.2315686 & 3.3644621 & 0.4387357 & 0.6380771 & 0.7308201 & 0.0005373 & 0.0040980 & 1.2229788 & 0.3295522 & 0.5353121 & 1.7341606 & 0.0007915 & 0.0056472 & -1.0267100 & 0.0013222 & 0.0086036 & -0.9327780 & 0.0010519 & 0.0071337 & 1.314872 & 0.1685588 & 0.3445806 & 1.623828 & 0.0554666 & 0.1571666 & -4.828044 & 0.0067413 & 0.0315747 & -5.772946 & 0.0006393 & 0.0047334 & -5.776763 & 0.0004591 & 0.0035968 & -5.017664\\
-\hline
-AK039774 & 0.0150416 & 0.0590150 & 0.6372239 & 0.0622016 & 0.1707939 & 0.9714819 & 0.0038649 & 0.0203447 & 1.3168792 & 0.9071986 & 0.9543873 & -0.6495320 & 0.0337680 & 0.1089499 & 1.4620498 & 0.1808765 & 0.3613288 & -0.6084973 & 0.4268483 & 0.6276079 & -1.4219238 & 0.8931509 & 0.9471666 & 0.6812377 & 0.3320222 & 0.5378422 & 1.725274 & 0.0000084 & 0.0001177 & 2.756976 & 0.7523631 & 0.8671706 & -4.265131 & 0.0000850 & 0.0008681 & -5.010475 & 0.0000014 & 0.0000238 & -4.623531 & 0.0066328 & 0.0311737 & -4.795976\\
-\hline
-NM\_013782 & 0.8754233 & 0.9377109 & 0.5940058 & 0.0001107 & 0.0010876 & 1.1803029 & 0.4731855 & 0.6676244 & 1.9274813 & 0.3558245 & 0.5614957 & -0.6697288 & 0.0000009 & 0.0000157 & 1.2377208 & 0.0469473 & 0.1390186 & -1.2969383 & 0.0877739 & 0.2187959 & 0.7832039 & 0.0873966 & 0.2181127 & -0.7705369 & 0.3054160 & 0.5099832 & 1.240504 & 0.0593928 & 0.1651796 & 2.355748 & 0.4210031 & 0.6224586 & -3.879986 & 0.0053367 & 0.0262800 & -5.108367 & 0.0056116 & 0.0273406 & -4.262393 & 0.0042107 & 0.0217896 & -4.617477\\
-\hline
-NM\_028622 & 0.7667946 & 0.8759265 & 1.6477918 & 0.9457494 & 0.9741115 & -1.8648471 & 0.9903079 & 0.9957216 & 1.1660022 & 0.6332637 & 0.7895226 & 2.1081788 & 0.8592452 & 0.9288844 & 1.2212238 & 0.7658228 & 0.8753935 & -1.1079164 & 0.8331999 & 0.9145679 & -0.8793087 & 0.8359000 & 0.9160123 & -1.2284455 & 0.8894396 & 0.9452334 & 1.160648 & 0.5797431 & 0.7516226 & -3.323953 & 0.5675779 & 0.7424636 & -4.125536 & 0.6699768 & 0.8146554 & -6.126729 & 0.6521300 & 0.8025109 & -4.510424 & 0.8460618 & 0.9215885 & 6.351333\\
+\textbf{EAP: Sort of strange to have argument \texttt{spines\_model} for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument}
+
+This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc}) and adjusted p-value (\texttt{qval}); the order of the genes in the table is the same as the input \texttt{data}.
+Here we show the results for the first two timepoints and the first ten genes:
+
+\begin{tabular}{l|r|r|r|r|r|r}
+\hline
+  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc\\
+\hline
+NM\_009912 & 0.273 & 0.475 & 2.202 & 0.767 & 0.876 & 1.158\\
+\hline
+NM\_008725 & 0.851 & 0.924 & 2.028 & 0.139 & 0.302 & 1.457\\
+\hline
+NM\_007473 & 0.088 & 0.219 & 0.678 & 0.917 & 0.959 & 0.764\\
+\hline
+ENSMUST00000094955 & 0.008 & 0.037 & 1.186 & 0.935 & 0.969 & -0.633\\
+\hline
+NM\_001042489 & 0.419 & 0.621 & 1.212 & 0.761 & 0.872 & 0.599\\
+\hline
+NM\_008159 & 0.123 & 0.278 & 1.102 & 0.117 & 0.268 & 0.737\\
+\hline
+NM\_001013813 & 0.057 & 0.159 & 0.630 & 0.171 & 0.348 & 0.975\\
+\hline
+AK039774 & 0.015 & 0.059 & 0.637 & 0.062 & 0.171 & 0.971\\
+\hline
+NM\_013782 & 0.875 & 0.938 & 0.594 & 0.000 & 0.001 & 1.180\\
+\hline
+NM\_028622 & 0.767 & 0.876 & 1.648 & 0.946 & 0.974 & -1.865\\
 \hline
@@
 
-We can repeat this for each of the conditions
+\textbf{EAP: For the output of this, the column names are in format \texttt{M.0-VL.0-pval}, for example.
+This may be hard to parse algorithmically down the road because you are using the same character for the contrast (\texttt{-}) as the separator for the \texttt{pval}.
+So for example, if you wanted to split the contrast name from the \texttt{pval} or \texttt{lfc} part, it wouldn't be very easy.
+Maybe underscore \texttt{\_} would be better separator?}
+
+We will repeat this, comparing each of the treatments to the control (``M'') (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).
 
@@
 
+\textbf{EAP: Should \texttt{timely\_barplot} be a function of \texttt{moanin}?}
+
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-12-1} \end{center}
 
-The distribution of number of genes found differentially expressed by
-considering each time-point independantly highlights the challenges of such
-approach. Timepoint 6H and 18H of the Kawasaki strain have fewer numbers of
-genes found differentially expressed than timepoints 3H and 9H: this is likely
-due to biological or technical variances for specific genes at specific
-timepoints. \textbf{EAP: We should add more to this}
+Such an analysis can demonstrate some general trends, with clearly more genes
+being differentially expressed at later time points, and the Vietnam high-dose
+showing perhaps an earlier onset than the low-dose.
+
+However, the distribution of the number of genes found differentially
+expressed by considering each time-point independently highlights the
+challenges of such approach. We can see that some timepoints have many less
+genes found significantly differentially expressed (e.g.~timepoint 6H and 18H
+of the Kawasaki strain). While there may be biological differences at those
+time points for some genes, it seems unlikely that the large majority of genes
+differentially expressed at timepoint 3H stop being differentially expressed
+at 6H and then jump back to being differentially expressed at 9H. A more
+likely explanation is that there is are some technical or biological artifact
+about the samples for 6H that is creating higher variation and thus less
+ability to detect significance.
+
+Another difficulty with such an approach is making sense of the general
+temporal structure for any particular gene, as different genes will have
+different combinations of timepoints DE. For the comparison of the Kawasaki
+strain to the control, for example, there are \texttt{length(signifCombos)} genes
+found DE in some timepoints, and there are \texttt{length(tabCombos)} different
+combinations of timepoints for which they are DE. Some of these make sense,
+such as DE in timepoints 48H-168H (\texttt{tabCombos{[}"48,60,72,120,168"{]}} genes), but
+many are very fragmentary. For example there are \texttt{tabCombos{[}"48,60,120,168"{]}}
+genes which are DE in timepoints 48,60,120,168H, but \emph{not} in the 72H. Many of
+these genes are likely to have not made the cutoff for significance in 72H,
+but don't show real differences in the overall trend between 48H-168H.
+
+Here are plots of the first 10 such genes, where we can see that some might
+show some meaningful changes between 60 and 72H, but others clearly just have
+a single replicate that is different or increased variability at 72H.
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/showMissingDEGenes-1} \end{center}
+
+\textbf{EAP: It would be nice to be able to subset in \texttt{plot\_centroids/plot\_genes} to only some of the groups}
 
 As a summary, classic differential expression methods are appropriate for
-unordered treatments, but fail to use the temporal structure of the data.
+unordered treatments, but fail to make use of the temporal structure of the data.
 
@@
 As another sanity check, \texttt{moanin} provides a simple utility function to
-visualize gene time-course data. Here, we plot the 12 genes with the smallest
+visualize gene time-course data. Here, we plot the 10 genes with the smallest
 p-values.
@@
 \begin{Highlighting}[]
-\NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
+\NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
 \KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
-           \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
 \end{Highlighting}
@@
     \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_epicon),}
-     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{12}\NormalTok{])}
+     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
 \KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
-           \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{)}
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
 \end{Highlighting}

### 2019-08-13 scripts/manuscript.Rmd
More fixes to the code
Bunch of fixes and additions to the manuscript

@@
 - Add legends to the plots that don't have.
-- Currently have two functions "plot_genes", "plot_centroids", that do exactly
-    the same thing. Find a better name for both of them
-- switch to camel case function names
-- Remove the namespace call
 - Make clear that it applies to RNASeq and how to call the different functions
@@
 designed to compare the gene expression between distinct groups, there is also a
-long history of time-course expression studies. Such studies compare the  gene expression
-across time by measuring mRNA levels from samples collected at different timepoints [^1]. Such time-course studies can vary from measuring a few distinct time
+long history of time-course expression studies. Such studies compare the 
+gene expression
+across time by measuring mRNA levels from samples collected at different 
+timepoints [^1]. Such time-course studies can vary from measuring a few 
+distinct time
 points, to sampling ten to twenty time points. These longer time series are
 particularly interested in investigating development over time. More recently,
-a new variety of time course studies have come from single-cell sequencing experiments**EAP: need citation** which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
+a new variety of time course studies have come from single-cell sequencing experiments
+[@shalek:single-cell; @habib:div-seq; @trapnell:dynamics]
+which can sequencing single cells which are at different stages of development; in this case,
+the time point is the stage of the cell in the process of development -- a
 value that is not know but estimated from the data as its "pseudo-time."
@@
 developmental systems, a massive number of genes will show *some* change. For
-example, in a study of mice lung tissues infected with influenza that we consider in this workflow, over 50% of genes
+example, in a study of mice lung tissues infected with influenza that we
+consider in this workflow, over 50% of genes
 are shown to be changing over time. The task in these settings is often not to
@@
 
-```{r schema, echo=FALSE, fig.width=3, fig.cap="Workflow for analyzing time-course datasets."}
+```{r schema, echo=FALSE, fig.width=1, fig.cap="Workflow for analyzing time-course datasets."}
 
@@
 condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
-**EAP: The colors for timepoint appear to circle around, i.e. the first and last are very similar colors**
 
@@
     pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
-mtext(at=275,text="Colored by Group", line=0.5 ,font=2,adj=0.5)
-legend(x=250,y=25,legend=names(ann_colors$Group), fill=ann_colors$Group,bty="n",xpd=NA,yjust=0.5)
+mtext(at=275, text="Colored by Group", line=0.5, font=2, adj=0.5)
+legend(x=250, y=25, legend=names(ann_colors$Group),
+       fill=ann_colors$Group, bty="n", xpd=NA, yjust=0.5)
 par(mar=c(4.5, 5.5, 1.5, .1))
@@
     pch=ann_markers$Replicate[as.factor(meta$Replicate)], bty="l")
-mtext(at=275,text="Colored by Timepoint",line=0.5 ,font=2,adj=0.5)
-legend(x=250,y=-15,legend=names(ann_colors$Timepoint), fill=ann_colors$Timepoint,bty="n",xpd=NA,yjust=0.5)
+mtext(at=275, text="Colored by Timepoint", line=0.5, font=2, adj=0.5)
+legend(x=250, y=-15, legend=names(ann_colors$Timepoint),
+       fill=ann_colors$Timepoint, bty="n", xpd=NA, yjust=0.5)
 par(mar=c(4.5, 5.5, 1.5, .1))
@@
 time-course datasets, where there are only a few time points 
-[@ritchie:limma, @robinson:edgeR, @love:moderated] . For long time-course datasets,
+[@ritchie:limma; @robinson:edgeR; @love:moderated] . For long time-course datasets,
 however, a separate test for each time-point results in creating many
@@
 
-We note that we could have time course data on either a single condition or
+We note that we could have time course data on either a single condition (to
+identify genes changing over time) or
 time course data on multiple conditions (such as the influenza dataset we are
 considering), which will alter slightly the types of questions we are
-interested in, but the two basic approaches remains the same. In what follows,
-we will focus on the situation where we have multiple conditions. **EAP: Any
-reason one and not the other. Maybe should change this sentence**
+interested in. 
+In what follows, we will focus on the situation
+where we have multiple conditions. **EAP: Any reason one and not the other.
+Maybe should change this sentence**
 
@@
 
-This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (`pval`), log-fold change (`lfc`) and adjusted p-value (`qval`); the order of the genes in the table is the same as the input `data`.
-Here we show the results for the first two timepoints and the first ten genes:
+This results in a table of the results, where each row corresponds to a gene
+and the columns correspond to the p-value (`pval`), log-fold change (`lfc`)
+and adjusted p-value (`qval`); the order of the genes in the table is the same
+as the input `data`. Here we show the results for the first two timepoints and
+the first ten genes:
 
@@
 
-We will repeat this, comparing each of the treatments to the control ("M") (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).
+We will repeat this, comparing each of the treatments to the control ("M")
+(code not printed here, as it is a replicate of the above, see accompanying
+Rmarkdown document).
 
@@
 **EAP: Should `timely_barplot` be a function of `moanin`?**
+
+**NV: I'd rather not include these types of functions that are hard to test
+and clutter the namespace. In order to provide reasonable visualization, they
+also often need to be tweaked specfically to match the data (here, the ylim is
+set specifically to the data**
 
@@
 par(mfrow=c(2, 2))
-timely_barplot(K_weekly_de_analysis, time, title="Kawasaki", las=3)
-timely_barplot(C_weekly_de_analysis, time, title="California", las=3)
-timely_barplot(weekly_de_analysis, time, title="Vietnam low-dose", las=3)
-timely_barplot(VH_weekly_de_analysis, time, title="Vietnam high-dose", las=3)
-```
-
-Such an analysis can demonstrate some general trends, with clearly more genes being differentially expressed at later time points, and the Vietnam high-dose showing perhaps an earlier onset than the low-dose.
-
-However, the distribution of the number of genes found differentially expressed by
-considering each time-point independently highlights the challenges of such
-approach.
-We can see that some timepoints have many less genes found significantly differentially expressed (e.g.
-timepoint 6H and 18H of the Kawasaki strain).
-While there may be biological differences at those time points for some genes, it seems unlikely that the large majority of genes differentially expressed at timepoint 3H stop being differentially expressed at 6H and then jump back to being differentially expressed at 9H.
-A more likely explanation is that there is are some technical or biological artifact about the samples for 6H that is creating higher variation and thus less ability to detect significance.
+time = names(time_colors)
+timely_barplot(K_weekly_de_analysis, labels=time, title="Kawasaki", las=3)
+timely_barplot(C_weekly_de_analysis, labels=time, title="California", las=3)
+timely_barplot(weekly_de_analysis, labels=time, title="Vietnam low-dose", las=3)
+timely_barplot(VH_weekly_de_analysis, labels=time, title="Vietnam high-dose", las=3)
+```
+
+Such an analysis can demonstrate some general trends, with clearly more genes
+being differentially expressed at later time points, and the Vietnam high-dose
+showing perhaps an earlier onset than the low-dose.
+
+However, the distribution of the number of genes found differentially
+expressed by considering each time-point independently highlights the
+challenges of such approach. We can see that some timepoints have many less
+genes found significantly differentially expressed (e.g. timepoint 6H and 18H
+of the Kawasaki strain). While there may be biological differences at those
+time points for some genes, it seems unlikely that the large majority of genes
+differentially expressed at timepoint 3H stop being differentially expressed
+at 6H and then jump back to being differentially expressed at 9H. A more
+likely explanation is that there is are some technical or biological artifact
+about the samples for 6H that is creating higher variation and thus less
+ability to detect significance.
 
@@
 
-Another difficulty with such an approach is making sense of the general temporal structure for any particular gene, as different genes will have different combinations of timepoints DE.
-For the comparison of the Kawasaki strain to the control, for example, there are `length(signifCombos)` genes found DE in some timepoints, and there are `length(tabCombos)` different combinations of timepoints for which they are DE.
-Some of these make sense, such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but many are very fragmentary.
-For example there are `tabCombos["48,60,120,168"]` genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H.
-Many of these genes are likely to have not made the cutoff for significance in 72H, but don't show real differences in the overall trend between 48H-168H.
-
-Here are plots of the first 10 such genes, where we can see that some might show some meaningful changes between 60 and 72H, but others clearly just have a single replicate that is different or increased variability at 72H.
+Another difficulty with such an approach is making sense of the general
+temporal structure for any particular gene, as different genes will have
+different combinations of timepoints DE. For the comparison of the Kawasaki
+strain to the control, for example, there are `length(signifCombos)` genes
+found DE in some timepoints, and there are `length(tabCombos)` different
+combinations of timepoints for which they are DE. Some of these make sense,
+such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but
+many are very fragmentary. For example there are `tabCombos["48,60,120,168"]`
+genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
+these genes are likely to have not made the cutoff for significance in 72H,
+but don't show real differences in the overall trend between 48H-168H.
+
+Here are plots of the first 10 such genes, where we can see that some might
+show some meaningful changes between 60 and 72H, but others clearly just have
+a single replicate that is different or increased variability at 72H.
 
@@
 wh<-which(splines_model$meta$Group %in% c("K","M"))
-plot_genes(data[exampleGenes,],splines_model,colors=ann_colors$Group)
-```
-
-**EAP: It would be nice to be able to subset in `plot_centroids/plot_genes` to only some of the groups**
+plot_splines_data(data[exampleGenes,], splines_model, colors=ann_colors$Group)
+```
+
+**EAP: It would be nice to be able to subset in `plot_splines_data/plot_splines_data` to only some of the groups**
 
@@
 top_DE_genes_pval = names(sort(pvalue)[1:10])
-plot_genes(data[top_DE_genes_pval, ], splines_model,
+plot_splines_data(data[top_DE_genes_pval, ], splines_model,
     colors=ann_colors$Group, smooth=TRUE,
@@
 	 decreasing=TRUE)[1:10])
-plot_genes(data[top_DE_genes_lfc, ], splines_model,
+plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
     colors=ann_colors$Group, smooth=TRUE,
@@
 
-We then use the `plot_centroids` function to visualize the centroids obtained
+We then use the `plot_splines_data` function to visualize the centroids obtained
 with the splines k-means model.
@@
 ```{r fig.height=6, fig.width=6}
-plot_centroids(kmeans_clusters$centroids, splines_model,
+plot_splines_data(
+    kmeans_clusters$centroids, splines_model,
     colors=ann_colors$Group,

### 2019-08-13 scripts/utils.R

@@ -4,7 +4,7 @@ group_colors = c(
     "C"="orange",
     "VH"="red4",
     "VL"="red2")
-time_colors = grDevices::rainbow(14)
+time_colors = grDevices::rainbow(15)[1:14]
 names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
 
 ann_colors = list(

### 2019-08-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \item
-  Currently have two functions ``plot\_genes'', ``plot\_centroids'', that do exactly
-  the same thing. Find a better name for both of them
-\item
-  switch to camel case function names
-\item
-  Remove the namespace call
-\item
   Make clear that it applies to RNASeq and how to call the different functions
@@
 ##   
-   checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION' ...
-
-v  checking for file '/tmp/Rtmp730h5L/remotes334541e0f9/NelleV-moanin-867a648/DESCRIPTION'
+   checking for file '/tmp/RtmpgstNOQ/remotes34231fb8c9/NelleV-moanin-8ab5610/DESCRIPTION' ...
+
+v  checking for file '/tmp/RtmpgstNOQ/remotes34231fb8c9/NelleV-moanin-8ab5610/DESCRIPTION'
 ##
@@
 designed to compare the gene expression between distinct groups, there is also a
-long history of time-course expression studies. Such studies compare the gene expression
-across time by measuring mRNA levels from samples collected at different timepoints \footnote{Because the collection of the mRNA is often destructive, samples at
+long history of time-course expression studies. Such studies compare the
+gene expression
+across time by measuring mRNA levels from samples collected at different
+timepoints \footnote{Because the collection of the mRNA is often destructive, samples at
   different time points are generally from different biological samples;
   longitudinal studies, for example tracking the same subject over time, are
-  certainly possible, but not directly considered here.}. Such time-course studies can vary from measuring a few distinct time
+  certainly possible, but not directly considered here.}. Such time-course studies can vary from measuring a few
+distinct time
 points, to sampling ten to twenty time points. These longer time series are
 particularly interested in investigating development over time. More recently,
-a new variety of time course studies have come from single-cell sequencing experiments\textbf{EAP: need citation} which can sequencing single cells which are at different stages of development; in this case, the time point is the stage of the cell in the process of development -- a
+a new variety of time course studies have come from single-cell sequencing experiments
+\citep{shalek:single-cell, habib:div-seq, trapnell:dynamics}
+which can sequencing single cells which are at different stages of development; in this case,
+the time point is the stage of the cell in the process of development -- a
 value that is not know but estimated from the data as its ``pseudo-time.''
@@
 developmental systems, a massive number of genes will show \emph{some} change. For
-example, in a study of mice lung tissues infected with influenza that we consider in this workflow, over 50\% of genes
+example, in a study of mice lung tissues infected with influenza that we
+consider in this workflow, over 50\% of genes
 are shown to be changing over time. The task in these settings is often not to
@@
 condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
-\textbf{EAP: The colors for timepoint appear to circle around, i.e.~the first and last are very similar colors}
 
@@
     \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\DecValTok{25}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\DecValTok{25}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group),}
+       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
@@
     \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{,}\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{,}\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{ ,}\DataTypeTok{font=}\DecValTok{2}\NormalTok{,}\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{,}\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{,}\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint), }\DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint,}\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{,}\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{,}\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint),}
+       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
@@
 time-course datasets, where there are only a few time points
-\citep[\citet{robinson:edgeR}, \citet{love:moderated}]{ritchie:limma} . For long time-course datasets,
+\citep{ritchie:limma, robinson:edgeR, love:moderated} . For long time-course datasets,
 however, a separate test for each time-point results in creating many
@@
 
-We note that we could have time course data on either a single condition or
+We note that we could have time course data on either a single condition (to
+identify genes changing over time) or
 time course data on multiple conditions (such as the influenza dataset we are
 considering), which will alter slightly the types of questions we are
-interested in, but the two basic approaches remains the same. In what follows,
-we will focus on the situation where we have multiple conditions. \textbf{EAP: Any
-reason one and not the other. Maybe should change this sentence}
+interested in.
+In what follows, we will focus on the situation
+where we have multiple conditions. \textbf{EAP: Any reason one and not the other.
+Maybe should change this sentence}
 
@@
 
-This results in a table of the results, where each row corresponds to a gene and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc}) and adjusted p-value (\texttt{qval}); the order of the genes in the table is the same as the input \texttt{data}.
-Here we show the results for the first two timepoints and the first ten genes:
+This results in a table of the results, where each row corresponds to a gene
+and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc})
+and adjusted p-value (\texttt{qval}); the order of the genes in the table is the same
+as the input \texttt{data}. Here we show the results for the first two timepoints and
+the first ten genes:
 
@@
 
-We will repeat this, comparing each of the treatments to the control (``M'') (code not printed here, as it is a replicate of the above, see accompanying Rmarkdown document).
+We will repeat this, comparing each of the treatments to the control (``M'')
+(code not printed here, as it is a replicate of the above, see accompanying
+Rmarkdown document).
 
@@
 \textbf{EAP: Should \texttt{timely\_barplot} be a function of \texttt{moanin}?}
+
+\textbf{NV: I'd rather not include these types of functions that are hard to test
+and clutter the namespace. In order to provide reasonable visualization, they
+also often need to be tweaked specfically to match the data (here, the ylim is
+set specifically to the data}
 
@@
 
-\textbf{EAP: It would be nice to be able to subset in \texttt{plot\_centroids/plot\_genes} to only some of the groups}
+\textbf{EAP: It would be nice to be able to subset in \texttt{plot\_splines\_data/plot\_splines\_data} to only some of the groups}
 
@@
 \NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
      \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_genes}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
 
-We then use the \texttt{plot\_centroids} function to visualize the centroids obtained
+We then use the \texttt{plot\_splines\_data} function to visualize the centroids obtained
 with the splines k-means model.
@@
 \begin{Highlighting}[]
-\KeywordTok{plot_centroids}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(}
+\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}

### 2019-08-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \KeywordTok{library}\NormalTok{(devtools)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{, }\DataTypeTok{dependencies=}\OtherTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##   
-   checking for file '/tmp/RtmpgstNOQ/remotes34231fb8c9/NelleV-moanin-8ab5610/DESCRIPTION' ...
-
-v  checking for file '/tmp/RtmpgstNOQ/remotes34231fb8c9/NelleV-moanin-8ab5610/DESCRIPTION'
-##
-
--  preparing 'moanin':
-##    checking DESCRIPTION meta-information ...
-
-v  checking DESCRIPTION meta-information
-##
-
--  checking for LF line-endings in source and make files and shell scripts
-##
-
--  checking for empty or unneeded directories
-## -  looking to see if a 'data/datalist' file should be added
-##
-
--  building 'moanin_0.0.0.tar.gz' (4.2s)
-##
-
-## 
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{, dependencies)}
+
 \KeywordTok{library}\NormalTok{(moanin)}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-7-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-1} \end{center}
 
@@
 ## [49] MASS_7.3-51.4           scales_1.0.0           
-## [51] hms_0.4.2               yaml_2.2.0             
-## [53] curl_3.3                memoise_1.1.0          
+## [51] hms_0.4.2               curl_3.3               
+## [53] yaml_2.2.0              memoise_1.1.0          
 ## [55] gridExtra_2.3           TeachingDemos_2.10     
@@
 ## [85] crayon_1.3.4            progress_1.2.0         
-## [87] grid_3.6.0              vegan_2.5-4            
-## [89] blob_1.1.1              callr_3.2.0            
-## [91] git2r_0.25.2            digest_0.6.18          
+## [87] grid_3.6.0              blob_1.1.1             
+## [89] callr_3.2.0             git2r_0.25.2           
+## [91] vegan_2.5-4             digest_0.6.18          
 ## [93] webshot_0.5.1           xtable_1.8-4           

### 2019-08-13 scripts/manuscript.Rmd
FIX disabling any installation through install_github
Travis keeps failing because of too many github API requests. From the
documentation, there's no way around doing github API requests. The fix
around rate limits is to provide a github key, which is going to be a
security breach once the repository is made public
Use new functionnality of moanin

@@
 
-```{r}
+```{r eval=FALSE}
 library(devtools)
-install_github("NelleV/moanin", dependencies=FALSE)
+install_github("NelleV/moanin", dependencies)
 
@@
 
+```{r echo=FALSE}
+library(moanin)
+```
+
 EAP:
 
-- Why `dependicies=FALSE`? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine. 
 - I got an error in installation script for  packages:
@@
 ```
-
-- `biocLite` is no longer the recommended way to install bioconductor. Instead
-  `BiocManager::install(c("edge", "NMI"))`. But this is dependent on
-  bioconductor/R version. Also, now that docker, can you change the
-  installation to just be simple one-liner (and `BiocManager::install` can
-  also do non-bioc packages, and is much more reliable.)
 
@@
 exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
-wh<-which(splines_model$meta$Group %in% c("K","M"))
-plot_splines_data(data[exampleGenes,], splines_model, colors=ann_colors$Group)
-```
-
-**EAP: It would be nice to be able to subset in `plot_splines_data/plot_splines_data` to only some of the groups**
+plot_splines_data(data, splines_model, subset_data=exampleGenes,
+		  colors=ann_colors$Group)
+```
 

### 2019-08-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \item
-  Why \texttt{dependicies=FALSE}? (same for the README for the repos). Seems easier to just say dependencies=TRUE and avoid the install.R routine.
-\item
   I got an error in installation script for packages:
@@
 \end{verbatim}
-
-\begin{itemize}
-\tightlist
-\item
-  \texttt{biocLite} is no longer the recommended way to install bioconductor. Instead
-  \texttt{BiocManager::install(c("edge",\ "NMI"))}. But this is dependent on
-  bioconductor/R version. Also, now that docker, can you change the
-  installation to just be simple one-liner (and \texttt{BiocManager::install} can
-  also do non-bioc packages, and is much more reliable.)
-\end{itemize}
 
@@
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/showMissingDEGenes-1} \end{center}
-
-\textbf{EAP: It would be nice to be able to subset in \texttt{plot\_splines\_data/plot\_splines\_data} to only some of the groups}
 

### 2019-08-15 scripts/manuscript.Rmd
FIX install_github (which ain't called anymore
Update TODO list
ENH now contrasts can be provided as vector of strings

@@
 
-- Add legends to the plots that don't have.
 - Make clear that it applies to RNASeq and how to call the different functions
@@
 library(devtools)
-install_github("NelleV/moanin", dependencies)
+install_github("NelleV/moanin")
 
@@
 We start by creating the `splines_model` object using the
-`create_splines_model` function. The `splines_model` object contains a number
-of metadata and options used throughout this analysis: the condition and
-timepoints of each samples, the formula object used or the basis or the
-degrees of freedom of the model. The metadata data.frame object should
-contain at least two columns: one named `Group`, containing the  treatement
-effect, and a second one named `Timepoint` containing the timepoint
-information.
+`create_splines_model` function.
+
+```{r}
+splines_model = create_splines_model(meta, degrees_of_freedom=6)
+```
+
+The `splines_model` object contains a number of metadata and options used
+throughout this analysis: the condition and timepoints of each samples, the
+formula object used or the basis or the degrees of freedom of the model. The
+metadata data.frame object should contain at least two columns: one named
+`Group`, containing the  treatement effect, and a second one named `Timepoint`
+containing the timepoint information.
 
@@
 If no formula is provided, it will default to the following: 
-`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`
-
-```{r}
-splines_model = create_splines_model(meta, degrees_of_freedom=6)
-```
-
-Here we have provided the meta data that defines... `create_splines_model`
-will .... **FIXME add information about what meta data is expected, describe
-what `create_splines_model` does (does it fit the splines? Just hold create an
-object? Create basis functions?)**.
-
-**EAP: General comment: you need a few more print() and head() of some objects to demonstrate what the output looks like**
+`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.
+
+**EAP: General comment: you need a few more print() and head() of some objects
+to demonstrate what the output looks like**
 
@@
 
-**FIXME add in equations, like from the EPICON paper, to describe them mathematically**
-
-**EAP: Do we need this call to limma by the user, or can these not be done
-internally by the `timecourse_differential_expression_analysis` function? We
-can still preserve the option of user defining their own, but would be nicer
-to make this simpler for the user (e.g. an if clause that tests whether input
-is matrix or character vector)**
-
 ```{r de_analysis}
@@
 timecourse_contrasts = c("M-K", "M-C", "M-VL", "M-VH")
-
-contrasts = makeContrasts(
-	contrasts=timecourse_contrasts,
-	levels=levels(meta$Group))
 
@@
 pvalues = DE_timecourse(
-    data, splines_model, contrasts,
+    data, splines_model, timecourse_contrasts,
     use_voom_weights=FALSE)
@@
 log_fold_change_timepoints = estimate_log_fold_change(
-    data, splines_model, contrasts,  method="timely")
+    data, splines_model, timecourse_contrasts,  method="timely")
 ```
@@
 log_fold_change_epicon = estimate_log_fold_change(
-    data, splines_model, contrasts,  method="epicon")
+    data, splines_model, timecourse_contrasts,  method="epicon")
 
 log_fold_change_sum = estimate_log_fold_change(
-    data, splines_model, contrasts,  method="sum")
+    data, splines_model, timecourse_contrasts,  method="sum")
 
 log_fold_change_max = estimate_log_fold_change(
-    data, splines_model, contrasts, method="max")
+    data, splines_model, timecourse_contrasts, method="max")
 
 log_fold_change_min = estimate_log_fold_change(
-    data, splines_model, contrasts, method="min")
+    data, splines_model, timecourse_contrasts, method="min")
 ```

### 2019-08-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \item
-  Add legends to the plots that don't have.
-\item
   Make clear that it applies to RNASeq and how to call the different functions
@@
 \KeywordTok{library}\NormalTok{(devtools)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{, dependencies)}
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
 
@@
 We start by creating the \texttt{splines\_model} object using the
-\texttt{create\_splines\_model} function. The \texttt{splines\_model} object contains a number
-of metadata and options used throughout this analysis: the condition and
-timepoints of each samples, the formula object used or the basis or the
-degrees of freedom of the model. The metadata data.frame object should
-contain at least two columns: one named \texttt{Group}, containing the treatement
-effect, and a second one named \texttt{Timepoint} containing the timepoint
-information.
+\texttt{create\_splines\_model} function.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{splines_model =}\StringTok{ }\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{splines\_model} object contains a number of metadata and options used
+throughout this analysis: the condition and timepoints of each samples, the
+formula object used or the basis or the degrees of freedom of the model. The
+metadata data.frame object should contain at least two columns: one named
+\texttt{Group}, containing the treatement effect, and a second one named \texttt{Timepoint}
+containing the timepoint information.
 
@@
 If no formula is provided, it will default to the following:
-\texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{splines_model =}\StringTok{ }\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Here we have provided the meta data that defines\ldots{} \texttt{create\_splines\_model}
-will \ldots{}. \textbf{FIXME add information about what meta data is expected, describe
-what \texttt{create\_splines\_model} does (does it fit the splines? Just hold create an
-object? Create basis functions?)}.
-
-\textbf{EAP: General comment: you need a few more print() and head() of some objects to demonstrate what the output looks like}
+\texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.
+
+\textbf{EAP: General comment: you need a few more print() and head() of some objects
+to demonstrate what the output looks like}
 
@@
 
-\textbf{FIXME add in equations, like from the EPICON paper, to describe them mathematically}
-
-\textbf{EAP: Do we need this call to limma by the user, or can these not be done
-internally by the \texttt{timecourse\_differential\_expression\_analysis} function? We
-can still preserve the option of user defining their own, but would be nicer
-to make this simpler for the user (e.g.~an if clause that tests whether input
-is matrix or character vector)}
-
 \begin{Shaded}
@@
 \NormalTok{timecourse_contrasts =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"M-K"}\NormalTok{, }\StringTok{"M-C"}\NormalTok{, }\StringTok{"M-VL"}\NormalTok{, }\StringTok{"M-VH"}\NormalTok{)}
-
-\NormalTok{contrasts =}\StringTok{ }\KeywordTok{makeContrasts}\NormalTok{(}
-    \DataTypeTok{contrasts=}\NormalTok{timecourse_contrasts,}
-    \DataTypeTok{levels=}\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group))}
 
@@
 \NormalTok{pvalues =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts,}
+\NormalTok{    data, splines_model, timecourse_contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
@@
 \NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
+\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{log_fold_change_epicon =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}
+\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}
 
 \NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
+\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
 
 \NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
+\NormalTok{    data, splines_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
 
 \NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
+\NormalTok{    data, splines_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
 \end{Highlighting}

### 2019-08-26 scripts/manuscript.Rmd
Renaming epicon lfc into timecourse

@@
 | sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
-| abs_sum  | $\sum_t \|lfc(t)\|$  | Always positive |
-| max  | $\max_t \|lfc(t)\|$  | Always positive  |
-| min  | $\min_t \|lfc(t)\|$  | Always positive |
-| epicon  | $$  | Captures overall strength of response and overall direction |
-
-**EAP: `epicon` is going to be weird for general use**
+| abs_sum  | $\sum_t \lvert lfc(t)\lvert$  | Always positive |
+| max  | $\max_t \lvert lfc(t) \lvert$  | Always positive  |
+| min  | $\min_t \lvert lfc(t) \lvert$  | Always positive |
+| timecourse | See details below | Captures overall strength and direction  |
+
+To capture the overall strength and direction of the response, we leverage the
+timepoint by timepoint log-fold change $\text{lfc}(t)$, and apply the
+following formula:
+
+$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)$$
+
+Note that when a gene is not consistently up or down-regulated the estimation
+of the direction will not represent accurately the changes observed.
 
 ```{r}
-log_fold_change_epicon = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts,  method="epicon")
+log_fold_change_timecourse = estimate_log_fold_change(
+    data, splines_model, timecourse_contrasts,  method="timecourse")
 
@@
 par(mfrow=c(2, 2))
-plot(log_fold_change_epicon[, "M-C"],
-     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
-     xlab="EPICON", ylab="sum")
-plot(log_fold_change_epicon[, "M-C"],
-     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
-     xlab="EPICON", ylab="max")
+plot(log_fold_change_timecourse[, "M-C"],
+     log_fold_change_sum[, "M-C"], pch=20, main="timecourse vs sum",
+     xlab="timecourse", ylab="sum")
+plot(log_fold_change_timecourse[, "M-C"],
+     log_fold_change_max[, "M-C"], pch=20, main="timecourse vs max",
+     xlab="timecourse", ylab="max")
 plot(log_fold_change_max[, "M-C"],
@@
 at the volcano plot. Here is an example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the EPICON log fold change
+of the control to the Kawasaki strain, using the timecourse log fold change
 computation.
@@
 pvalue = qvalues[, "M-K"]
-lfc_epicon = log_fold_change_epicon[, "M-K"]
-names(lfc_epicon) = row.names(log_fold_change_epicon)
-
-plot(lfc_epicon, -log10(pvalue), pch=20, main="Volcano plot",
+lfc_timecourse = log_fold_change_timecourse[, "M-K"]
+names(lfc_timecourse) = row.names(log_fold_change_timecourse)
+
+plot(lfc_timecourse, -log10(pvalue), pch=20, main="Volcano plot",
      xlim=c(-2.5, 2))
@@
 
-And here we visualize the genes with the largest absolute EPICON log-fold
+And here we visualize the genes with the largest absolute timecourse log-fold
 change.
@@
 top_DE_genes_lfc = names(
-    sort(abs(lfc_epicon),
+    sort(abs(lfc_timecourse),
 	 decreasing=TRUE)[1:10])
@@
     genes, species="mmu",
-    download_latest=TRUE)$stastic
+    download_latest=FALSE)$stastic
 

### 2019-08-28 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 sum & \(\sum_t lfc(t)\) & Sum of log fold change.\tabularnewline
-abs\_sum & \(\sum_t \|lfc(t)\|\) & Always positive\tabularnewline
-max & \(\max_t \|lfc(t)\|\) & Always positive\tabularnewline
-min & \(\min_t \|lfc(t)\|\) & Always positive\tabularnewline
-epicon & \$\$ & Captures overall strength of response and overall direction\tabularnewline
+abs\_sum & \(\sum_t \lvert lfc(t)\lvert\) & Always positive\tabularnewline
+max & \(\max_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
+min & \(\min_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
+timecourse & See details below & Captures overall strength and direction\tabularnewline
 \bottomrule
@@
 
-\textbf{EAP: \texttt{epicon} is going to be weird for general use}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{log_fold_change_epicon =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}
+To capture the overall strength and direction of the response, we leverage the
+timepoint by timepoint log-fold change \(\text{lfc}(t)\), and apply the
+following formula:
+
+\[\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)\]
+
+Note that when a gene is not consistently up or down-regulated the estimation
+of the direction will not represent accurately the changes observed.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
+\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}
 
@@
 \NormalTok{pvalue =}\StringTok{ }\NormalTok{qvalues[, }\StringTok{"M-K"}\NormalTok{]}
-\NormalTok{lfc_epicon =}\StringTok{ }\NormalTok{log_fold_change_epicon[, }\StringTok{"M-K"}\NormalTok{]}
-\KeywordTok{names}\NormalTok{(lfc_epicon) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(log_fold_change_epicon)}
-
-\KeywordTok{plot}\NormalTok{(lfc_epicon, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
+\NormalTok{lfc_timecourse =}\StringTok{ }\NormalTok{log_fold_change_timecourse[, }\StringTok{"M-K"}\NormalTok{]}
+\KeywordTok{names}\NormalTok{(lfc_timecourse) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(log_fold_change_timecourse)}
+
+\KeywordTok{plot}\NormalTok{(lfc_timecourse, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
      \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}
@@
 \NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
-    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_epicon),}
+    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_timecourse),}
      \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
@@
 \NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
-    \DataTypeTok{download_latest=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
+    \DataTypeTok{download_latest=}\OtherTok{FALSE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
 \end{Highlighting}
@@
 \hline
-05169 & Epstein-Barr virus infection & 37 & 0\\
-\hline
-04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
-\hline
-04668 & TNF signaling pathway & 45 & 0\\
-\hline
-05164 & Influenza A & 38 & 0\\
-\hline
-04621 & NOD-like receptor signaling pathway & 34 & 0\\
-\hline
-05162 & Measles & 37 & 0\\
-\hline
-04630 & JAK-STAT signaling pathway & 33 & 0\\
-\hline
-05160 & Hepatitis C & 34 & 0\\
-\hline
-04064 & NF-kappa B signaling pathway & 39 & 0\\
+04060 & Cytokine-cytokine receptor interaction & 34 & 0\\
 \hline
 04620 & Toll-like receptor signaling pathway & 40 & 0\\
+\hline
+04630 & Jak-STAT signaling pathway & 33 & 0\\
+\hline
+03050 & Proteasome & 53 & 0\\
+\hline
+04621 & NOD-like receptor signaling pathway & 47 & 0\\
+\hline
+04380 & Osteoclast differentiation & 34 & 0\\
+\hline
+04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
+\hline
+04210 & Apoptosis & 37 & 0\\
+\hline
+04622 & RIG-I-like receptor signaling pathway & 39 & 0\\
+\hline
+05160 & Hepatitis C & 29 & 0\\
 \hline

### 2019-08-28 scripts/manuscript.Rmd
Use of newer moanin functions/names

@@
 these types of approaches, though our focus is on the global approach. In both
-situations, we first need to set up a object (a `splines_model` object) to
-hold the meta data, as well as information for fitting the spline model.
-
-We start by creating the `splines_model` object using the
-`create_splines_model` function.
+situations, we first need to set up a object (a `moanin_model` object) to
+hold the meta data, as well as information for fitting the spline model
+(formula, number of degrees of freedom of the splines, …)
+
+We start by creating the `moanin_model` object using the
+`create_moanin_model` function.
 
 ```{r}
-splines_model = create_splines_model(meta, degrees_of_freedom=6)
-```
-
-The `splines_model` object contains a number of metadata and options used
+moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
+```
+
+The `moanin_model` object contains a number of metadata and options used
 throughout this analysis: the condition and timepoints of each samples, the
@@
 ```{r make_timepoints_contarts}
-contrasts = create_timepoints_contrasts("M", "VL", splines_model)
+contrasts = create_timepoints_contrasts("M", "VL", moanin_model)
 ```
@@
 weekly_de_analysis = DE_timepoints(
-     data, splines_model, contrasts,
+     data, moanin_model, contrasts,
      use_voom_weights=FALSE)
 ```
-**EAP: Sort of strange to have argument `spines_model` for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument**
-
 This results in a table of the results, where each row corresponds to a gene
@@
 
-**EAP: For the output of this, the column names are in format `M.0-VL.0-pval`, for example.
-This may be hard to parse algorithmically down the road because you are using the same character for the contrast (`-`) as the separator for the `pval`.
-So for example, if you wanted to split the contrast name from the `pval` or `lfc` part, it wouldn't be very easy.
-Maybe underscore `_` would be better separator?**
-
 We will repeat this, comparing each of the treatments to the control ("M")
@@
 ```{r echo=FALSE, results="hide"}
-contrasts = create_timepoints_contrasts("M", "K", splines_model)
+contrasts = create_timepoints_contrasts("M", "K", moanin_model)
 K_weekly_de_analysis =  DE_timepoints(
-     data, splines_model, contrasts,
+     data, moanin_model, contrasts,
      use_voom_weights=FALSE)
 
-contrasts = create_timepoints_contrasts("M", "C", splines_model)
+contrasts = create_timepoints_contrasts("M", "C", moanin_model)
 C_weekly_de_analysis =  DE_timepoints(
-     data, splines_model, contrasts,
+     data, moanin_model, contrasts,
      use_voom_weights=FALSE)
 
-contrasts = create_timepoints_contrasts("M", "VH", splines_model)
+contrasts = create_timepoints_contrasts("M", "VH", moanin_model)
 VH_weekly_de_analysis =  DE_timepoints(
-     data, splines_model, contrasts,
+     data, moanin_model, contrasts,
      use_voom_weights=FALSE)
@@
 exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
-plot_splines_data(data, splines_model, subset_data=exampleGenes,
+plot_splines_data(data, moanin_model, subset_data=exampleGenes,
 		  colors=ann_colors$Group)
@@
 pvalues = DE_timecourse(
-    data, splines_model, timecourse_contrasts,
+    data, moanin_model, timecourse_contrasts,
     use_voom_weights=FALSE)
@@
 log_fold_change_timepoints = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts,  method="timely")
+    data, moanin_model, timecourse_contrasts,  method="timely")
 ```
@@
 
+To compute the log-fold change, `moanin` provides a function taking as
+argument the data, the `moanin_model` object, the contrasts to evaluate, as
+well as the method to use to estimate the log-fold change.
+
 ```{r}
 log_fold_change_timecourse = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts,  method="timecourse")
+    data, moanin_model, timecourse_contrasts,  method="timecourse")
 
 log_fold_change_sum = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts,  method="sum")
+    data, moanin_model, timecourse_contrasts,  method="sum")
 
 log_fold_change_max = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts, method="max")
+    data, moanin_model, timecourse_contrasts, method="max")
 
 log_fold_change_min = estimate_log_fold_change(
-    data, splines_model, timecourse_contrasts, method="min")
+    data, moanin_model, timecourse_contrasts, method="min")
 ```
@@
 top_DE_genes_pval = names(sort(pvalue)[1:10])
-plot_splines_data(data[top_DE_genes_pval, ], splines_model,
+plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
     colors=ann_colors$Group, smooth=TRUE,
@@
 	 decreasing=TRUE)[1:10])
-plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
+plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
     colors=ann_colors$Group, smooth=TRUE,
@@
 kmeans_clusters = splines_kmeans(
-    y, splines_model, n_clusters=8,
+    y, moanin_model, n_clusters=8,
     random_seed=42,
@@
 plot_splines_data(
-    kmeans_clusters$centroids, splines_model,
+    kmeans_clusters$centroids, moanin_model,
     colors=ann_colors$Group,
@@
     clustering_results = splines_kmeans(
-	y, splines_model,
+	y, moanin_model,
 	n_clusters=n_cluster, random_seed=42,

### 2019-08-30 scripts/manuscript.Rmd
FIX parsing of columns have changed

@@
 #Code for counting the number of unique combinations of time points that are DE
-getTimepoint<-function(x){sapply(strsplit(gsub("-qval","",x),"\\."),.subset2,3)}
+getTimepoint<-function(x){sapply(strsplit(gsub("_qval","",x),"\\."),.subset2,3)}
 qval_colnames = colnames(K_weekly_de_analysis)[

### 2019-08-30 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 these types of approaches, though our focus is on the global approach. In both
-situations, we first need to set up a object (a \texttt{splines\_model} object) to
-hold the meta data, as well as information for fitting the spline model.
-
-We start by creating the \texttt{splines\_model} object using the
-\texttt{create\_splines\_model} function.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{splines_model =}\StringTok{ }\KeywordTok{create_splines_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-The \texttt{splines\_model} object contains a number of metadata and options used
+situations, we first need to set up a object (a \texttt{moanin\_model} object) to
+hold the meta data, as well as information for fitting the spline model
+(formula, number of degrees of freedom of the splines, \textless{}80\textgreater{})
+
+We start by creating the \texttt{moanin\_model} object using the
+\texttt{create\_moanin\_model} function.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{moanin\_model} object contains a number of metadata and options used
 throughout this analysis: the condition and timepoints of each samples, the
@@
 \begin{Highlighting}[]
-\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, splines_model)}
+\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, moanin_model)}
 \end{Highlighting}
@@
 \NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{(}
-\NormalTok{     data, splines_model, contrasts,}
+\NormalTok{     data, moanin_model, contrasts,}
      \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
@@
 \end{Shaded}
-
-\textbf{EAP: Sort of strange to have argument \texttt{spines\_model} for function that doesn't use splines. Perhaps there is a generic name that could be used for the argument}
 
@@
 \hline
-  & M.0-VL.0-pval & M.0-VL.0-qval & M.0-VL.0-lfc & M.3-VL.3-pval & M.3-VL.3-qval & M.3-VL.3-lfc\\
+  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc & M.3-VL.3\_pval & M.3-VL.3\_qval & M.3-VL.3\_lfc\\
 \hline
@@
 \end{tabular}
-
-\textbf{EAP: For the output of this, the column names are in format \texttt{M.0-VL.0-pval}, for example.
-This may be hard to parse algorithmically down the road because you are using the same character for the contrast (\texttt{-}) as the separator for the \texttt{pval}.
-So for example, if you wanted to split the contrast name from the \texttt{pval} or \texttt{lfc} part, it wouldn't be very easy.
-Maybe underscore \texttt{\_} would be better separator?}
 
@@
 \NormalTok{pvalues =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts,}
+\NormalTok{    data, moanin_model, timecourse_contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
@@
 \NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
+\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
 \end{Highlighting}
@@
 
+To compute the log-fold change, \texttt{moanin} provides a function taking as
+argument the data, the \texttt{moanin\_model} object, the contrasts to evaluate, as
+well as the method to use to estimate the log-fold change.
+
 \begin{Shaded}
@@
 \NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}
+\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}
 
 \NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
+\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
 
 \NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
+\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
 
 \NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, splines_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
+\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
 \end{Highlighting}
@@
 \NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_pval, ], splines_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_pval, ], moanin_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
      \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], moanin_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
 \NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
-\NormalTok{    y, splines_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
+\NormalTok{    y, moanin_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
     \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
@@
 \KeywordTok{plot_splines_data}\NormalTok{(}
-\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, splines_model,}
+\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, moanin_model,}
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
@@
 \NormalTok{    clustering_results =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
-\NormalTok{    y, splines_model,}
+\NormalTok{    y, moanin_model,}
     \DataTypeTok{n_clusters=}\NormalTok{n_cluster, }\DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}

### 2019-08-30 scripts/manuscript.Rmd

@@
 library(knitr)
-library(kableExtra)
 library(BiocStyle)
 library(biomaRt)
+library(kableExtra)
 
@@
 library(moanin)
-```
-
-EAP:
-
-- I got an error in installation script for  packages:
-
-```
-epurdom-0:2019timecourse-rnaseq-pipeline epurdom$ Rscript scripts/install.R
-Loading required package: devtools
-Loading required package: usethis
-Loading required package: ggfortify
-Loading required package: ggplot2
-Loading required package: kableExtra
-Error in contrib.url(repos, "source") : 
-  trying to use CRAN without setting a mirror
-Calls: install.packages -> contrib.url
-In addition: Warning message:
-In library(package, lib.loc = lib.loc, character.only = TRUE, logical.return = TRUE,  :
-  there is no package called ‘kableExtra’
-Execution halted
 ```
@@
 
-We note that we could have time course data on either a single condition (to
-identify genes changing over time) or
-time course data on multiple conditions (such as the influenza dataset we are
-considering), which will alter slightly the types of questions we are
-interested in. 
-In what follows, we will focus on the situation
-where we have multiple conditions. **EAP: Any reason one and not the other.
-Maybe should change this sentence**
+Time course data can either be on a single condition (to identify genes
+changing over time) or or on multiple conditions (such as the influenza
+dataset we are considering), which will alter slightly the types of questions
+we are interested in.
 
@@
 ```{r echo=FALSE}
-kable(contrasts, col.names="Contrasts")
-```
-**EAP: Not sure if `kable` is better than just printout?**
-
+kable(contrasts, "latex", col.names="Contrasts", booktabs=TRUE)
+```
 Then run the differential expression analysis on all of those timepoints
@@
 ```{r echo=FALSE}
-kable(weekly_de_analysis[1:10, 1:6 ],digits=3)
+kable(weekly_de_analysis[1:10, 1:6 ], "latex", digits=3, booktabs=TRUE)
+
 ```
@@
     head(log_fold_change_timepoints[, colnames(log_fold_change_timepoints)[1:5]]),
-    format = "latex")
+    format="latex",
+    booktabs=TRUE)
 ```
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:10, ], "latex", escape=F, digits=4) %>%
-    kable_styling("striped", full_width=TRUE) 
+kable(pathways[1:10, ], format="latex", escape=F, digits=4, booktabs=TRUE)
 ```
@@
 
-kable(go_terms_enriched[1:15, ], "latex", escape=F, digits=4)  %>%
-    kable_styling("striped", full_width=TRUE) 
+kable(go_terms_enriched[1:15, ], format="latex", escape=F, digits=4, booktabs=TRUE)
 ```

### 2019-08-30 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-EAP:
-
-\begin{itemize}
-\tightlist
-\item
-  I got an error in installation script for packages:
-\end{itemize}
-
-\begin{verbatim}
-epurdom-0:2019timecourse-rnaseq-pipeline epurdom$ Rscript scripts/install.R
-Loading required package: devtools
-Loading required package: usethis
-Loading required package: ggfortify
-Loading required package: ggplot2
-Loading required package: kableExtra
-Error in contrib.url(repos, "source") : 
-  trying to use CRAN without setting a mirror
-Calls: install.packages -> contrib.url
-In addition: Warning message:
-In library(package, lib.loc = lib.loc, character.only = TRUE, logical.return = TRUE,  :
-  there is no package called <e2><80><98>kableExtra<e2><80><99>
-Execution halted
-\end{verbatim}
-
 \hypertarget{introduction}{%
@@
 
-We note that we could have time course data on either a single condition (to
-identify genes changing over time) or
-time course data on multiple conditions (such as the influenza dataset we are
-considering), which will alter slightly the types of questions we are
-interested in.
-In what follows, we will focus on the situation
-where we have multiple conditions. \textbf{EAP: Any reason one and not the other.
-Maybe should change this sentence}
+Time course data can either be on a single condition (to identify genes
+changing over time) or or on multiple conditions (such as the influenza
+dataset we are considering), which will alter slightly the types of questions
+we are interested in.
 
@@
 \begin{tabular}{l}
-\hline
+\toprule
 Contrasts\\
-\hline
+\midrule
 M.0-VL.0\\
-\hline
 M.3-VL.3\\
-\hline
 M.6-VL.6\\
-\hline
 M.9-VL.9\\
-\hline
 M.12-VL.12\\
-\hline
+\addlinespace
 M.18-VL.18\\
-\hline
 M.24-VL.24\\
-\hline
 M.30-VL.30\\
-\hline
 M.36-VL.36\\
-\hline
 M.48-VL.48\\
-\hline
+\addlinespace
 M.60-VL.60\\
-\hline
 M.72-VL.72\\
-\hline
 M.120-VL.120\\
-\hline
 M.168-VL.168\\
-\hline
+\bottomrule
 \end{tabular}
-
-\textbf{EAP: Not sure if \texttt{kable} is better than just printout?}
 
@@
 
-\begin{tabular}{l|r|r|r|r|r|r}
-\hline
+\begin{tabular}{lrrrrrr}
+\toprule
   & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc & M.3-VL.3\_pval & M.3-VL.3\_qval & M.3-VL.3\_lfc\\
-\hline
+\midrule
 NM\_009912 & 0.273 & 0.475 & 2.202 & 0.767 & 0.876 & 1.158\\
-\hline
 NM\_008725 & 0.851 & 0.924 & 2.028 & 0.139 & 0.302 & 1.457\\
-\hline
 NM\_007473 & 0.088 & 0.219 & 0.678 & 0.917 & 0.959 & 0.764\\
-\hline
 ENSMUST00000094955 & 0.008 & 0.037 & 1.186 & 0.935 & 0.969 & -0.633\\
-\hline
 NM\_001042489 & 0.419 & 0.621 & 1.212 & 0.761 & 0.872 & 0.599\\
-\hline
+\addlinespace
 NM\_008159 & 0.123 & 0.278 & 1.102 & 0.117 & 0.268 & 0.737\\
-\hline
 NM\_001013813 & 0.057 & 0.159 & 0.630 & 0.171 & 0.348 & 0.975\\
-\hline
 AK039774 & 0.015 & 0.059 & 0.637 & 0.062 & 0.171 & 0.971\\
-\hline
 NM\_013782 & 0.875 & 0.938 & 0.594 & 0.000 & 0.001 & 1.180\\
-\hline
 NM\_028622 & 0.767 & 0.876 & 1.648 & 0.946 & 0.974 & -1.865\\
-\hline
+\bottomrule
 \end{tabular}
@@
 
-\begin{tabular}{l|r|r|r|r|r}
-\hline
+\begin{tabular}{lrrrrr}
+\toprule
   & M-K:1 & M-K:7 & M-K:11 & M-K:14 & M-K:2\\
-\hline
+\midrule
 NM\_009912 & -0.0061424 & 0.0224948 & -0.0515947 & -0.1143973 & 0.0614619\\
-\hline
 NM\_008725 & 2.6547855 & 0.2734387 & -0.6882925 & 0.4718807 & -1.7463319\\
-\hline
 NM\_007473 & -0.1603623 & 0.1017655 & 0.5079573 & 0.0595773 & 0.4250185\\
-\hline
 ENSMUST00000094955 & 0.3505920 & 0.6463011 & 0.2564796 & 0.3599102 & 0.4177116\\
-\hline
 NM\_001042489 & 0.0889750 & 0.2321381 & -0.1891501 & 0.1786113 & 0.3320644\\
-\hline
+\addlinespace
 NM\_008159 & -0.3863770 & 0.0298395 & -0.0862755 & -0.3266351 & -0.1893062\\
-\hline
+\bottomrule
 \end{tabular}
@@
 
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
-\hline
+\begin{tabular}{llrr}
+\toprule
   & Pathway & Percentage & Adj. p-value\\
-\hline
+\midrule
 04060 & Cytokine-cytokine receptor interaction & 34 & 0\\
-\hline
 04620 & Toll-like receptor signaling pathway & 40 & 0\\
-\hline
 04630 & Jak-STAT signaling pathway & 33 & 0\\
-\hline
 03050 & Proteasome & 53 & 0\\
-\hline
 04621 & NOD-like receptor signaling pathway & 47 & 0\\
-\hline
+\addlinespace
 04380 & Osteoclast differentiation & 34 & 0\\
-\hline
 04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
-\hline
 04210 & Apoptosis & 37 & 0\\
-\hline
 04622 & RIG-I-like receptor signaling pathway & 39 & 0\\
-\hline
 05160 & Hepatitis C & 29 & 0\\
-\hline
-\end{tabu}
+\bottomrule
+\end{tabular}
 
@@
 
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedright}X>{\raggedleft}X}
-\hline
+\begin{tabular}{lllrrrlr}
+\toprule
   & GO ID & Description & Annotated & Significant & Expected & P-value & adj. p-value\\
-\hline
+\midrule
 5 & GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.2629\\
-\hline
 6 & GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.3323\\
-\hline
 7 & GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.3323\\
-\hline
 8 & GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.3729\\
-\hline
 9 & GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.3820\\
-\hline
+\addlinespace
 10 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.6194\\
-\hline
 11 & GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.6194\\
-\hline
 12 & GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.6194\\
-\hline
 13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.8517\\
-\hline
 14 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.0000\\
-\hline
+\addlinespace
 15 & GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.0000\\
-\hline
 16 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
-\hline
 17 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
-\hline
 18 & GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
-\hline
 19 & GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
-\hline
-\end{tabu}
+\bottomrule
+\end{tabular}
 
@@
 ## [19] rngtools_1.3.1.1     pkgmaker_0.27        registry_0.5-1      
-## [22] ggplot2_3.1.1        biomaRt_2.36.1       BiocStyle_2.8.2     
-## [25] kableExtra_1.1.0     knitr_1.22           limma_3.36.5        
+## [22] ggplot2_3.1.1        kableExtra_1.1.0     biomaRt_2.36.1      
+## [25] BiocStyle_2.8.2      knitr_1.22           limma_3.36.5        
 ## [28] usethis_1.5.0        devtools_2.0.2       rmarkdown_1.12      
@@
 ## [77] R6_2.4.0                DBI_1.0.0              
-## [79] mgcv_1.8-28             pillar_1.3.1           
-## [81] withr_2.1.2             abind_1.4-5            
+## [79] pillar_1.3.1            withr_2.1.2            
+## [81] mgcv_1.8-28             abind_1.4-5            
 ## [83] KEGGREST_1.20.2         tibble_2.1.1           

### 2019-08-30 scripts/manuscript.Rmd
ENH added a note about count versus microarray data
More information in supplementary

@@
 
-**EAP: We need to add comment about microarray versus sequencing data**
-
 By combining gene expression time-course data with virus growth data, the
@@
 regulated by virus concentration.
+
+While we showcase this pipeline on micro-array data, [@varoquaux:lifecycle]
+leverages a similar set of steps to analyse RNA-seq data of the lifetime
+transcriptomic response of the crop *S. bicolor* to drought.
 
@@
 colnames(go_terms_enriched) = c("GO ID", "Description", "Annotated",
-"Significant", "Expected", "P-value", "adj. p-value")
-
-kable(go_terms_enriched[1:15, ], format="latex", escape=F, digits=4, booktabs=TRUE)
-```
-
-# Session information
+"Significant", "Expected", "P-value", "Adj. p-value")
+
+kable(go_terms_enriched[1:15, ], format="latex", escape=F,
+      digits=3, booktabs=TRUE, row.names=FALSE)
+```
+
+# Software and data availability
+
+The source code for this workflow can be found at
+[https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
+Archived source code at the time of publication can be foud at [XXX](XXX).
+
+All packages use in the workflow are available on GitHub, CRAN, or
+Bioconductor.
+
+Data used in this workflow are available from NCBI GEO, accession
+[GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
+Normalized data can be found in `moanin`.  Normalization information is
+provided as supplementary information.
+
+Last but not least, we use `sessionInfo()` to display all packages used in
+this pipeline and their version numbers.
 

### 2019-08-30 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\textbf{EAP: We need to add comment about microarray versus sequencing data}
-
 By combining gene expression time-course data with virus growth data, the
@@
 regulated by virus concentration.
+
+While we showcase this pipeline on micro-array data, \citep{varoquaux:lifecycle}
+leverages a similar set of steps to analyse RNA-seq data of the lifetime
+transcriptomic response of the crop \emph{S. bicolor} to drought.
 
@@
 
-\begin{tabular}{lllrrrlr}
+\begin{tabular}{llrrrlr}
 \toprule
-  & GO ID & Description & Annotated & Significant & Expected & P-value & adj. p-value\\
+GO ID & Description & Annotated & Significant & Expected & P-value & Adj. p-value\\
 \midrule
-5 & GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.2629\\
-6 & GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.3323\\
-7 & GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.3323\\
-8 & GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.3729\\
-9 & GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.3820\\
+GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.263\\
+GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.332\\
+GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.332\\
+GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.373\\
+GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.382\\
 \addlinespace
-10 & GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.6194\\
-11 & GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.6194\\
-12 & GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.6194\\
-13 & GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.8517\\
-14 & GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.0000\\
+GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.619\\
+GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.619\\
+GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.619\\
+GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.852\\
+GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.000\\
 \addlinespace
-15 & GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.0000\\
-16 & GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
-17 & GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.0000\\
-18 & GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
-19 & GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.0000\\
+GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.000\\
+GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.000\\
+GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.000\\
+GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.000\\
+GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.000\\
 \bottomrule
@@
 
-\hypertarget{session-information}{%
-\section{Session information}\label{session-information}}
+\hypertarget{software-and-data-availability}{%
+\section{Software and data availability}\label{software-and-data-availability}}
+
+The source code for this workflow can be found at
+\url{https://github.com/NelleV/2019timecourse-rnaseq-pipeline}.
+Archived source code at the time of publication can be foud at \url{XXX}.
+
+All packages use in the workflow are available on GitHub, CRAN, or
+Bioconductor.
+
+Data used in this workflow are available from NCBI GEO, accession
+\href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786}{GSE63786}.
+Normalized data can be found in \texttt{moanin}. Normalization information is
+provided as supplementary information.
+
+Last but not least, we use \texttt{sessionInfo()} to display all packages used in
+this pipeline and their version numbers.
 
@@
 
-\hypertarget{software-and-data-availability}{%
-\section{Software and data availability}\label{software-and-data-availability}}
+\hypertarget{software-and-data-availability-1}{%
+\section{Software and data availability}\label{software-and-data-availability-1}}
 

### 2019-09-24 scripts/manuscript.Rmd
some fixes
changes through picking number of clusters

@@
 
+*EAP*: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
+
 ```{r eval=FALSE}
@@
 ```
+
+**EAP: It seems like a problem to have the colors defined in a utils.R. Makes it hard It's not echo'd by markdown and so readers won't know where they come from. (I didn't).**
 
@@
 
+The meta data contains information about the treatment group, the replicate, and the timepoint for each observation:
+
+```{r echo=FALSE}
+kable(head(meta))
+```
+
+*EAP*: I moved this here, because we use meta a lot in the eda.
+
 ### Exploratory analysis and quality control
@@
 Before performing any additional exploratory analysis, let us only keep highly
-variable genes: we keep for this step only the top 50% most variable genes.
+variable genes: for this step, we keep only the top 50% most variable genes.
 
@@
 
+**EAP**: this makes it sound like you create a class, but in fact it's just a list. Wouldn't it be helpful to make it a class?
+
 We start by creating the `moanin_model` object using the
-`create_moanin_model` function.
-
-```{r}
-moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
-```
-
-The `moanin_model` object contains a number of metadata and options used
-throughout this analysis: the condition and timepoints of each samples, the
-formula object used or the basis or the degrees of freedom of the model. The
+`create_moanin_model` function. We need to provide two things to the function: a data.frame with the meta data, and the number of degrees of freedom. The
 metadata data.frame object should contain at least two columns: one named
@@
 
-If no formula is provided, it will default to the following: 
+We create a `moanin_model` for our data:
+
+```{r}
+moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
+```
+
+The `moanin_model` object contains a number of metadata and options used
+throughout this analysis: the condition and timepoints of each samples, the
+formula object used or the basis or the degrees of freedom of the model. 
+**EAP:** You make it sound like or, but in fact the object has all of these elements, no?
+
+If no formula is provided, it will default to the following, which provides for a different spline fit for every group (as defined by the `Group` variable in the `meta` data.frame): 
 `formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.
@@
 `moanin` provides a simple interface to perform a timepoint by timepoint
-differential expression analysis. This is traditionally done by the user
-defining the comparisons (called `contrasts` in linear models).  Under the
-hood, simply calls `limma` [@ritchie:limma] on the set of contrasts provided.
-By default, it expects RNA-Seq contact counts, and will estimate voom weights.
-We tern
+differential expression analysis. Comparison between groups is traditionally done by 
+defining the group comparisons (called `contrasts` in linear models) as a linear combination of the coefficients of the model.
+Comparing groups within each timepoint can create many contrasts, and thus `moanin` provides functionality to create these contrasts in an automatic way, and then calls `limma` [@ritchie:limma] on the set of contrasts provided.
+By default, `moanin` expects RNA-Seq contact counts, and will estimate voom weights.
 
@@
 ```{r make_timepoints_contarts}
+# Define contrasts	
 contrasts = create_timepoints_contrasts("M", "VL", moanin_model)
@@
 
-This creates a vector of contrasts to be tested, one for each timepoint:
+This creates a vector of contrasts to be tested, one for each timepoint, in the format required by `limma`:
 
@@
 ```
-Then run the differential expression analysis on all of those timepoints
+
+Then `moanin` will run the differential expression analysis on all of those timepoints
 jointly using the function `DE_timepoints`.
@@
 ```{r weekly_differential_analysis}
-# Define contrast	
 weekly_de_analysis = DE_timepoints(
@@
 ```
-This results in a table of the results, where each row corresponds to a gene
+
+The output is a table of results, where each row corresponds to a gene
 and the columns correspond to the p-value (`pval`), log-fold change (`lfc`)
-and adjusted p-value (`qval`); the order of the genes in the table is the same
-as the input `data`. Here we show the results for the first two timepoints and
+and adjusted p-value (`qval`) of the sets of contrasts; the order of the genes in the table is the same
+as the input `data`. Here we show the results for the first timepoint (i.e. first three columns of the output) and
 the first ten genes:
@@
 
-We will repeat this, comparing each of the treatments to the control ("M")
+Additional timepoints are in the additional columns of the output.
+
+We will repeat this, comparing each of the remaining three treatments to the control ("M")
 (code not printed here, as it is a replicate of the above, see accompanying
@@
 
+**EAP: Okay, I'll leave that to you.
+Most of those tweaked things can just be in a ...
+argument and plotting functions that do annoying things are popular.
+I've at least rewritten it so users can more easily change those features in calls to the function rather than mucking about in the code**
+
 ```{r fig.width=6, fig.height=6, echo=FALSE}
-timely_barplot = function(de_analysis, labels=NULL, qval_threshold=0.05,
-			  title="", ...){
+timely_barplot = function(de_analysis, 
+    labels=NULL, qval_threshold=0.05,
+    xlab="Timepoint", ylab="Number of DE genes", main="",ylim=c(0, 20000), col="black", ...){
     qval_colnames = colnames(de_analysis)[
@@
 
-    barplot(number_de_genes_per_time, col="black", 
-            names.arg=labels,
-	    ylim=c(0, 20000),
-	    xlab="Timepoint", ylab="Number of DE genes", main=title, ...)  
+    barplot(number_de_genes_per_time, 
+            names.arg=labels, xlab=xlab,ylab=ylab,
+            col=col, main=main, ylim=ylim,...)  
 }
@@
 time = names(time_colors)
-timely_barplot(K_weekly_de_analysis, labels=time, title="Kawasaki", las=3)
-timely_barplot(C_weekly_de_analysis, labels=time, title="California", las=3)
-timely_barplot(weekly_de_analysis, labels=time, title="Vietnam low-dose", las=3)
-timely_barplot(VH_weekly_de_analysis, labels=time, title="Vietnam high-dose", las=3)
+timely_barplot(K_weekly_de_analysis, labels=time, main="Kawasaki", las=3)
+timely_barplot(C_weekly_de_analysis, labels=time, main="California", las=3)
+timely_barplot(weekly_de_analysis, labels=time, main="Vietnam low-dose", las=3)
+timely_barplot(VH_weekly_de_analysis, labels=time, main="Vietnam high-dose", las=3)
 ```
@@
     grepl("qval", colnames(K_weekly_de_analysis))]
-signifCombos<-apply(K_weekly_de_analysis[, qval_colnames],1,function(x){paste(getTimepoint(qval_colnames[which(x<0.05)]),collapse=",")})
+signifCombos<-apply(K_weekly_de_analysis[, qval_colnames], 1, 
+    function(x){paste(getTimepoint(qval_colnames[which(x<0.05)]),collapse=",")})
 signifCombos<-signifCombos[signifCombos!=""]
@@
 contrasts -- only now the contrasts are differences between the estimated
-means.
+mean functions.
+This is done with the function `DE_timecourse`, which takes as similar input that of `DE_timepoints`, only now will test the entire mean function (and therefore does not require a step of expanding the contrasts into contrasts for individual timepoints).
 
@@
 ```
+
+**EAP: is there a reason that the qvalues aren't provided with the output like in `DE_timepoints`? Doesn't that risk the user not taking that step and working with the unadjusted p-values?**
 
@@
 As another sanity check, `moanin` provides a simple utility function to
-visualize gene time-course data. Here, we plot the 10 genes with the smallest
+visualize gene time-course data in different conditions. Here, we plot the 10 genes with the smallest
 p-values.
@@
 to the treatment. Before clustering the genes, we first reduce the set of
-genes of interest to genes that are (1) significantly found differentially
-expressed; (2) "highly" differentially expressed. To do this, we first
+genes of interest to genes that are (1) found significantly differentially
+expressed; (2) a large-fold change between conditions. To do this, we first
 aggregate all p-values obtained during the time-course differential expresison
@@
 - **Splines estimation**: for each gene, fit the splines function with the basis
-  of your choice.
+  of your choice (provided in the `moanin_model` object).
 - **Rescaling splines**: for each gene, rescale the estimated splines function
-  such that the values are bounded between 0 and 1.
-- **K-means**: apply k-means on the rescaled fitted splines to estimate the
-  centroids.
-- **Assign scores and labels to all genes**: then assign a score and a label
-  to all gene based on a goodness-of-fit measure on the raw data. By default,
-  the `splines_kmeans_score_and_label` function only labels the best 50% of
-  genes.
-
-The first three steps are performed jointly by the `splines_kmeans` function.
+  such that the values are bounded between 0 and 1 and thus comparable between genes.
+- **K-means**: apply k-means on the rescaled fitted values of the splines to estimate the
+  cluster centroids.
+- **Assign scores and labels to all genes**: Our `splines_kmeans_score_and_label` function assigns a score to all gene-cluster pairs based on a goodness-of-fit measure of the gene to the cluster centroid.
+We then "assign" each gene to a cluster based on its score.
+
+The first three steps are performed internally by the `splines_kmeans` function.
+For now, we will set the number of clusters to be $8$, though we will return to the question of picking the best number of clusters below.
 
@@
 
-We then use the `plot_splines_data` function to visualize the centroids obtained
+We then use the `plot_splines_data` function to visualize the centroids of each cluster obtained
 with the splines k-means model.
@@
 
-The scoring and labeling can be done via the `splines_kmeans_score_and_label`
+These centroids are on a 0-1 scale, because of our rescaling of the spline fits, and do not represent the actual gene expression level.
+We can plot a few of the actual genes in a cluster to see the difference:
+
+```{r fig.height=6, fig.width=6}
+cluster2Genes = names(kmeans_clusters$clusters[kmeans_clusters$clusters==2])
+dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[2,])
+plot_splines_data(dataToPlot, moanin_model,
+    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+    mar=c(1.5,2.5,2,0.1))
+```
+
+The spline plots are fitted per gene in the above plot, and illustrate that the genes assigned to a cluster have differing degrees to which their data fits that of the cluster centroid, and indeed fits that of their individual spline fits.
+
+### Assigning genes to clusters
+
+Since we arbitrarily chose a subset of genes based on our filter, we would like all genes to have a chance of assignment to a cluster, as well as a score for whether it is a good assignment.
+The scoring and label step allows us to assign genes to clusters that were removed during the
+filtering step, yet are good matches to the centroids found during the
+clustering.
+
+The scoring and labeling is done via the `splines_kmeans_score_and_label`
 function.
+This function calculates the goodness of fit of the gene to the cluster centroid.
+By default, the function labels only the 50% of genes with the best scores, with the remaining genes getting `NA` as their assignment.
 
@@
 
+After running `scores_and_labels` we now have `r length(labels)` genes that are assigned to a cluster.
+
 Before performing the next steps, let us investigate in more detail the
 differences between the labels provided by the splines k-means model and the
-scoring and labelinig step.
+scoring and labeling step.
 
@@
 	col="black",
-	main="splines k-means labeling")
-barplot(table(labels), col="black", main="score and label")
-```
-
-The scoring and label step allows to label genes that were removed during the
-filtering step, yet are good matches to the centroids found during the
-clustering.
+	main="labeling from splines k-means")
+barplot(table(labels), col="black", main="labeling from scores_and_labels function")
+```
+
+Of the original `r length(kmeans_clusters$clusters)` genes used in the clustering, `r length(intersect(names(kmeans_clusters$clusters), names(labels)))` are still assigned to a cluster based on their goodness of fit score.
+We can compare whether they are still assigned to the same clusters based on a confusion matrix shown below.
+The kmeans cluster assignments are shown on the rows, and the goodness-of-fit assignments in the columns, with the number in each cell indicating the number of genes in the intersection of the two clusters.
 
@@
 
+# of those used in clustering, keep only those 
+# with high goodness of fit
 kmeans_labels = kmeans_clusters$clusters
@@
 	      border_color="black", txt=text, legend=FALSE, Rowv=NA,
-	      Colv=NA)
+	      Colv=NA, sub="Goodness-of-fit assignments")
 labels = scores_and_labels$labels
@@
 
+```
+
+Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in% cluster2Score)` of them remained assigned to cluster 2.
+We can again look at a few genes in cluster 2, only now pick the best scoring genes
+
+```{r fig.height=6, fig.width=6}
+cluster2Score = names(labels[labels==2])
+#order them by their score
+cluster2Score=cluster2Score[order(scores_and_labels$scores[cluster2Score,2], decreasing = TRUE)]
+dataToPlot=rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[2,])
+plot_splines_data(dataToPlot, moanin_model,
+    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+    mar=c(1.5,2.5,2,0.1))
 ```
@@
 
+**EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
+Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
+It's seems like an important plot to make**
+
 ### Looking at specific clusters in details.
@@
 particularly interesting: it captures genes with strong differences between
-the differant influenza treatments and the control, while the control remains
+the different influenza treatments and the control, while the control remains
 relatively flat.
+
+We've already shown how we can plot a few example genes, however, it can be hard to make sense of individual genes given the amount of noise, as well as being hard to draw conclusions based on a few genes.
 
@@
 
-First, select the genes of interest.
+First, we select the genes of interest, namely those in cluster 8, based on our goodness-of-fit assignment.
 
@@
 ```
+
+Now we will create the heatmaps of these genes (`r length(genes_to_plot)` genes).
 
@@
 will, for each clustering experiment, conserve (1) with within cluster sum of
-squares; (2) the labels assigned to all genes.
+squares; (2) the labels assigned to all genes. **EAP: I don't understand what you are trying to say in this last sentence. I think there's a typo**
+
+**EAP: Also, I don't know what you mean by "clustering experiment" -- do you mean a clustering of the data with a particular K?
+Ditto for "all possible clusters of interest" -- do you mean all choices of K, so 5,9,11,15 in the below.**
+
+Below we run the clustering for $k$ equal to $2-20$. The `splines_kmeans` function returns the WCSS for each cluster, which we sum to get the total WCSS.
 
 ```{r running_clustering_on_all_k}
-all_possible_n_clusters = c(5, 9, 11, 15)
+all_possible_n_clusters = c(2:20)
 all_clustering = list()
@@
 The Elbow method to choose the number of clusters relies on visualization aid
-to choose the number of cluster. The method relies on plotting the within
-cluster sum of squares as a function of the number of clusters. At some point,
-the WCSS will stop dropping, giving an angle in the graph.  The number of
+to choose the number of clusters. The method relies on plotting the within
+cluster sum of squares (WCSS) as a function of the number of clusters. At some point,
+the WCSS will start decreasing more slowly, giving an angle or "elbow" in the graph.  The number of
 cluster is chosen at this "Elbow point."
+
+We plot the WCSS for $k=2-20$ here.
+We see that as expected the WCSS continues to drop, but there is no clear drop in the decrease, except for very small values of $k$ (3-4 clusters).
+However, 3-4 seems a very small number of gene clusters to find, given the complexity
 
@@
 ```
+
+**EAP: I've increased this to be k=2-20, because you can't actually interpret these plots without a dense number of k; also you do 2-20 for stability**
 
@@
 
-Using the bootstrapping strategy:
+Using the bootstrapping strategy we sample with replacement a sample of the same size as our original clustering (i.e. `r dim(y)[1]` genes):
+
 ```{r}
@@
 
-Using the subsampling strategy, keeping 80% of the genes:
+Using the subsampling strategy, we take a unique subset of the genes, keeping 80% of the genes:
+
+**EAP: `subsample_proportion` was set to 1 in the code, but shouldn't it be 0.8?
+I changed it to 0.8.
+But it seems like you don't actually run this code, but read existing runs, so maybe was right there?**
 
 ```{r}
-subsample_proportion = 1
+subsample_proportion = 0.8
 indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
@@
 
-Here we plot to stability matrix of the top 1000 genes for $k=5$ and $k=20$.
+**EAP: So this is very confusing to understand, because you read the results from a file (`stability_5.tsv`), but I don't know what this file contains, nor how they were calculated.
+And I don't know what a stability matrix is -- you've only talked about a consensus matrix.
+There should be the code here for a single k, but not evaluated, and say that for convenience we provide the results of running this code as a file for users to play with.
+I've tried to fix up the text to be what I think it should be**
+
+We will run **EAP: subsampling???** method on the top 1000 genes **EAP: what is top??**, and do it $B=20$ times for each of $k=2-20$.
+We show the code below, but because of the time it takes we have evaluate these values separately and provided the results for users to explore more quickly.
+
+For now, we bring in the results for $k=5$ and $k=20$.
+
+```{r}
+stability_5 = read.table("results/stability_5.tsv", sep="\t")
+stability_20 = read.table("results/stability_20.tsv", sep="\t")
+```
+**EAP: It's very confusing that the headings here are C1, ...
+C20.
+It makes it seem like cluster 1, ...
+cluster 20, when I think (?) it is really 20 resampling results.
+Would be good to change that to, e.g.
+B1,...,B20 or something.**
+
+For each gene, we have an assignment to a cluster over the 20 resampling runs. For example, for $k=5$:
+
+```{r,echo=FALSE}
+kable(head(stability_5))
+```
+
+Now we will use the function `consensus_matrix` in the `moanin` package to calculate proportion of times each pair of samples was clustered together across the 20 resampling runs, and plot the heatmap of those consensus matrices for $k=5$ and $k=20$.
 
 ```{r fig.width=6, fig.height=4}
-stability_5 = read.table("results/stability_5.tsv", sep="\t")
 consensus_matrix_stability_5 = consensus_matrix(stability_5,
@@
 	 Colv=FALSE,
-	 treeheight=0)
-
-stability_20 = read.table("results/stability_20.tsv", sep="\t")
+	 treeheight=0, main="K=5")
+
 consensus_matrix_stability_20 = consensus_matrix(stability_20,
@@
 	 Colv=FALSE,
-	 treeheight=0)
-
-```
+	 treeheight=0, main="K=20")
+
+```
+
+We can see that the choice of $K=5$ seems much more stable across resampling runs than that of $K=20$.
+
+#### The model explorer strategy
+
+The model explorer algorithm [@ben-hur:stability] proposes to estimate the
+number of clusters exploiting the observation that if the number of clusters
+is correct, the clustering results are stable to bootstrap resampling, as described above.  The
+distribution of similaries between bootstrapped results for each $k$ can thus
+be compared for different values of $k$ and guide the user in the choice of
+number of clusters.
+
+The model explorer strategy works as follows. For a single choice of $k$, first perform $n$ bootstrap
+experiments to estimate the cluster centroids, followed by a step of assigning
+a label to all data points. Then, choose a similarity
+measure between two partitions or clusters $S(C_1, C_2)$.
+**EAP: This is confusing as to whether this is a measure of similarity across different bootstrap resamples (i.e.
+different sets of clusters) or across two different clusters in the same resample.
+Perhaps also because of confusion of headings of different resamples between labeled C1, C2, but those aren't cluster 1/2 (see comment above)** Examples are the
+normalized mutual information or Fowlkes-Mallows. Finally, compute the pairwise similarity measure
+between all bootstrapped partitions (i.e. B choose 2 pairs). Repeat this procedure for different $k$ and plot per $k$ the cumulative density of the
+obtained scores.
+
+We have already run the bootstrap resampling for values $k=2-20$ and saved the results. We will read that data in, and use it to calculate the pairwise similarity scores.
 
@@
 
-##### The model explorer strategy
-
-The model explorer algorithm [@ben-hur:stability] proposes to estimate the
-number of clusters exploiting the observation that if the number of clusters
-is correct, the clustering results are stable to bootstrapping.  The
-distribution of similaries between bootstrapped results for each $k$ can thus
-be compared for different values of $k$ and guide the user in the choice of
-number of clusters.
-
-The model explorer strategy works as follows. First, choose a similarity
-measure between two partitions or clusters $S(C_1, C_2)$. Examples are the
-normalized mutual information or Fowlkes-Mallows. Then perform $n$ bootstrap
-experiments to estimate the cluster centroids, followed by a step of assigning
-a label to all data points. Finally, compute the pairwise similarity measure
-between all bootstrapped partition, and plot the cumulative density of the
-obtained scores.
+The function `plot_model_explorer`
 
@@
 ```
-
-##### Consensus clustering as a way to find $k$
+**EAP: `plot_model_explorer` is relatively slow, because it does all those pairwise calculations.
+And then to do it for smaller subset, redo all the calculations.
+Can't the output be saved, or the calculations be a different step from the plotting?**
+
+#### Consensus clustering as a way to find $k$
 
@@
 bootstrapped clustering, the authors suggests plotting the cumulative density
-of elements of the consensus matrix.
+of elements of the consensus matrix. We provide a function `plot_cdf_consensus` in `moanin` to do this
 
@@
 ```
+**EAP: We should again provide the code that goes with these images**
 
@@
 ```
-
-# Software and data availability
-
-The source code for this workflow can be found at
-[https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
-Archived source code at the time of publication can be foud at [XXX](XXX).
-
-All packages use in the workflow are available on GitHub, CRAN, or
-Bioconductor.
-
-Data used in this workflow are available from NCBI GEO, accession
-[GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
-Normalized data can be found in `moanin`.  Normalization information is
-provided as supplementary information.
-
-Last but not least, we use `sessionInfo()` to display all packages used in
-this pipeline and their version numbers.
-
-```{r}
-sessionInfo()
-```
-
 # Conclusion
@@
 
-The source code for this workflow can be found at [https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
-Data used in this workflow are available from NCBI GEO, accession GSE95601.
+The source code for this workflow can be found at
+[https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
+Archived source code at the time of publication can be foud at [XXX](XXX).
+
+All packages use in the workflow are available on GitHub, CRAN, or
+Bioconductor.
+
+Data used in this workflow are available from NCBI GEO, accession
+[GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
+Normalized data can be found in `moanin`.  Normalization information is
+provided as supplementary information.
+
+Last but not least, we use `sessionInfo()` to display all packages used in
+this pipeline and their version numbers.
+
+```{r}
+sessionInfo()
+```
 

### 2019-09-24 scripts/manuscript.Rmd
update

@@
 ```
+
+This analysis doesn't necessarily pick a particular $k$, but can help decide between $k$ within a desired range, for example, or to avoid $k$ that degrade the stability.
+
 **EAP: `plot_model_explorer` is relatively slow, because it does all those pairwise calculations.
@@
 
-First, let us clean up the gene obtained and only select the genes we are
+First, let us clean up the genes we work with and only select the genes we are
 going to use in the enrichment analysis. One can either use the whole set of

### 2019-09-24 scripts/manuscript.Rmd
fix error in Rmd code

@@
 labels = unlist(labels[!is.na(labels)])
+cluster2Score = names(labels[labels==2])
 ```
@@
 ```{r fig.height=6, fig.width=6}
-cluster2Score = names(labels[labels==2])
 #order them by their score
@@
 [https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
-Archived source code at the time of publication can be foud at [XXX](XXX).
+Archived source code at the time of publication can be found at [XXX](XXX).
 

### 2019-09-25 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+\emph{EAP}: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
+
 \begin{Shaded}
@@
 \end{Shaded}
+
+\textbf{EAP: It seems like a problem to have the colors defined in a utils.R. Makes it hard It's not echo'd by markdown and so readers won't know where they come from. (I didn't).}
 
@@
 
+The meta data contains information about the treatment group, the replicate, and the timepoint for each observation:
+
+\begin{tabular}{l|r|l|r|r}
+\hline
+  &  & Group & Replicate & Timepoint\\
+\hline
+GSM1557140 & 0 & K & 1 & 0\\
+\hline
+GSM1557141 & 1 & K & 2 & 0\\
+\hline
+GSM1557142 & 2 & K & 3 & 0\\
+\hline
+GSM1557143 & 3 & K & 1 & 12\\
+\hline
+GSM1557144 & 4 & K & 2 & 12\\
+\hline
+GSM1557145 & 5 & K & 3 & 12\\
+\hline
+\end{tabular}
+
+\emph{EAP}: I moved this here, because we use meta a lot in the eda.
+
 \hypertarget{exploratory-analysis-and-quality-control}{%
@@
 Before performing any additional exploratory analysis, let us only keep highly
-variable genes: we keep for this step only the top 50\% most variable genes.
+variable genes: for this step, we keep only the top 50\% most variable genes.
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
@@
 
+\textbf{EAP}: this makes it sound like you create a class, but in fact it's just a list. Wouldn't it be helpful to make it a class?
+
 We start by creating the \texttt{moanin\_model} object using the
-\texttt{create\_moanin\_model} function.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-The \texttt{moanin\_model} object contains a number of metadata and options used
-throughout this analysis: the condition and timepoints of each samples, the
-formula object used or the basis or the degrees of freedom of the model. The
+\texttt{create\_moanin\_model} function. We need to provide two things to the function: a data.frame with the meta data, and the number of degrees of freedom. The
 metadata data.frame object should contain at least two columns: one named
@@
 
-If no formula is provided, it will default to the following:
+We create a \texttt{moanin\_model} for our data:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{moanin\_model} object contains a number of metadata and options used
+throughout this analysis: the condition and timepoints of each samples, the
+formula object used or the basis or the degrees of freedom of the model.
+\textbf{EAP:} You make it sound like or, but in fact the object has all of these elements, no?
+
+If no formula is provided, it will default to the following, which provides for a different spline fit for every group (as defined by the \texttt{Group} variable in the \texttt{meta} data.frame):
 \texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.
@@
 \texttt{moanin} provides a simple interface to perform a timepoint by timepoint
-differential expression analysis. This is traditionally done by the user
-defining the comparisons (called \texttt{contrasts} in linear models). Under the
-hood, simply calls \texttt{limma} \citep{ritchie:limma} on the set of contrasts provided.
-By default, it expects RNA-Seq contact counts, and will estimate voom weights.
-We tern
+differential expression analysis. Comparison between groups is traditionally done by
+defining the group comparisons (called \texttt{contrasts} in linear models) as a linear combination of the coefficients of the model.
+Comparing groups within each timepoint can create many contrasts, and thus \texttt{moanin} provides functionality to create these contrasts in an automatic way, and then calls \texttt{limma} \citep{ritchie:limma} on the set of contrasts provided.
+By default, \texttt{moanin} expects RNA-Seq contact counts, and will estimate voom weights.
 
@@
 \begin{Highlighting}[]
+\CommentTok{# Define contrasts  }
 \NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, moanin_model)}
@@
 
-This creates a vector of contrasts to be tested, one for each timepoint:
+This creates a vector of contrasts to be tested, one for each timepoint, in the format required by \texttt{limma}:
 
@@
 
-Then run the differential expression analysis on all of those timepoints
+Then \texttt{moanin} will run the differential expression analysis on all of those timepoints
 jointly using the function \texttt{DE\_timepoints}.
@@
 \begin{Highlighting}[]
-\CommentTok{# Define contrast   }
 \NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{(}
@@
 
-This results in a table of the results, where each row corresponds to a gene
+The output is a table of results, where each row corresponds to a gene
 and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc})
-and adjusted p-value (\texttt{qval}); the order of the genes in the table is the same
-as the input \texttt{data}. Here we show the results for the first two timepoints and
+and adjusted p-value (\texttt{qval}) of the sets of contrasts; the order of the genes in the table is the same
+as the input \texttt{data}. Here we show the results for the first timepoint (i.e.~first three columns of the output) and
 the first ten genes:
@@
 
-We will repeat this, comparing each of the treatments to the control (``M'')
+Additional timepoints are in the additional columns of the output.
+
+We will repeat this, comparing each of the remaining three treatments to the control (``M'')
 (code not printed here, as it is a replicate of the above, see accompanying
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
+\textbf{EAP: Okay, I'll leave that to you.
+Most of those tweaked things can just be in a \ldots{} argument and plotting functions that do annoying things are popular.
+I've at least rewritten it so users can more easily change those features in calls to the function rather than mucking about in the code}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
 
@@
 contrasts -- only now the contrasts are differences between the estimated
-means.
+mean functions.
+This is done with the function \texttt{DE\_timecourse}, which takes as similar input that of \texttt{DE\_timepoints}, only now will test the entire mean function (and therefore does not require a step of expanding the contrasts into contrasts for individual timepoints).
 
@@
 
+\textbf{EAP: is there a reason that the qvalues aren't provided with the output like in \texttt{DE\_timepoints}?
+Doesn't that risk the user not taking that step and working with the unadjusted p-values?}
+
 The number of genes found differentially expressed ranges from around 12000 to
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
 As another sanity check, \texttt{moanin} provides a simple utility function to
-visualize gene time-course data. Here, we plot the 10 genes with the smallest
+visualize gene time-course data in different conditions. Here, we plot the 10 genes with the smallest
 p-values.
@@
 to the treatment. Before clustering the genes, we first reduce the set of
-genes of interest to genes that are (1) significantly found differentially
-expressed; (2) ``highly'' differentially expressed. To do this, we first
+genes of interest to genes that are (1) found significantly differentially
+expressed; (2) a large-fold change between conditions. To do this, we first
 aggregate all p-values obtained during the time-course differential expresison
@@
   \textbf{Splines estimation}: for each gene, fit the splines function with the basis
-  of your choice.
+  of your choice (provided in the \texttt{moanin\_model} object).
 \item
   \textbf{Rescaling splines}: for each gene, rescale the estimated splines function
-  such that the values are bounded between 0 and 1.
+  such that the values are bounded between 0 and 1 and thus comparable between genes.
 \item
-  \textbf{K-means}: apply k-means on the rescaled fitted splines to estimate the
-  centroids.
+  \textbf{K-means}: apply k-means on the rescaled fitted values of the splines to estimate the
+  cluster centroids.
 \item
-  \textbf{Assign scores and labels to all genes}: then assign a score and a label
-  to all gene based on a goodness-of-fit measure on the raw data. By default,
-  the \texttt{splines\_kmeans\_score\_and\_label} function only labels the best 50\% of
-  genes.
+  \textbf{Assign scores and labels to all genes}: Our \texttt{splines\_kmeans\_score\_and\_label} function assigns a score to all gene-cluster pairs based on a goodness-of-fit measure of the gene to the cluster centroid.
+We then ``assign'' each gene to a cluster based on its score.
 \end{itemize}
 
-The first three steps are performed jointly by the \texttt{splines\_kmeans} function.
+The first three steps are performed internally by the \texttt{splines\_kmeans} function.
+For now, we will set the number of clusters to be \(8\), though we will return to the question of picking the best number of clusters below.
 
@@
 
-We then use the \texttt{plot\_splines\_data} function to visualize the centroids obtained
+We then use the \texttt{plot\_splines\_data} function to visualize the centroids of each cluster obtained
 with the splines k-means model.
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
-
-The scoring and labeling can be done via the \texttt{splines\_kmeans\_score\_and\_label}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
+
+These centroids are on a 0-1 scale, because of our rescaling of the spline fits, and do not represent the actual gene expression level.
+We can plot a few of the actual genes in a cluster to see the difference:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{cluster2Genes =}\StringTok{ }\KeywordTok{names}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\DecValTok{2}\NormalTok{])}
+\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(y[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{2}\NormalTok{,])}
+\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+
+The spline plots are fitted per gene in the above plot, and illustrate that the genes assigned to a cluster have differing degrees to which their data fits that of the cluster centroid, and indeed fits that of their individual spline fits.
+
+\hypertarget{assigning-genes-to-clusters}{%
+\subsubsection{Assigning genes to clusters}\label{assigning-genes-to-clusters}}
+
+Since we arbitrarily chose a subset of genes based on our filter, we would like all genes to have a chance of assignment to a cluster, as well as a score for whether it is a good assignment.
+The scoring and label step allows us to assign genes to clusters that were removed during the
+filtering step, yet are good matches to the centroids found during the
+clustering.
+
+The scoring and labeling is done via the \texttt{splines\_kmeans\_score\_and\_label}
 function.
+This function calculates the goodness of fit of the gene to the cluster centroid.
+By default, the function labels only the 50\% of genes with the best scores, with the remaining genes getting \texttt{NA} as their assignment.
 
@@
 \NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{cluster2Score =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\DecValTok{2}\NormalTok{])}
+\end{Highlighting}
+\end{Shaded}
+
+After running \texttt{scores\_and\_labels} we now have 19772 genes that are assigned to a cluster.
 
@@
 differences between the labels provided by the splines k-means model and the
-scoring and labelinig step.
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
-
-The scoring and label step allows to label genes that were removed during the
-filtering step, yet are good matches to the centroids found during the
-clustering.
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1} \end{center}
+scoring and labeling step.
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
+
+Of the original 3950 genes used in the clustering, 3467 are still assigned to a cluster based on their goodness of fit score.
+We can compare whether they are still assigned to the same clusters based on a confusion matrix shown below.
+The kmeans cluster assignments are shown on the rows, and the goodness-of-fit assignments in the columns, with the number in each cell indicating the number of genes in the intersection of the two clusters.
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
+
+Of the four genes we plotted before, 3 of them remained assigned to cluster 2. We can again look at a few genes in cluster 2, only now pick the best scoring genes
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#order them by their score}
+\NormalTok{cluster2Score=cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score,}\DecValTok{2}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
+\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{2}\NormalTok{,])}
+\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
+
+\textbf{EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
+Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
+It's seems like an important plot to make}
 
@@
 \end{Shaded}
+
+Now we will create the heatmaps of these genes (4332 genes).
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
 
@@
 will, for each clustering experiment, conserve (1) with within cluster sum of
-squares; (2) the labels assigned to all genes.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{all_possible_n_clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{9}\NormalTok{, }\DecValTok{11}\NormalTok{, }\DecValTok{15}\NormalTok{)}
+squares; (2) the labels assigned to all genes. \textbf{EAP: I don't understand what you are trying to say in this last sentence. I think there's a typo}
+
+\textbf{EAP: Also, I don't know what you mean by ``clustering experiment'' -- do you mean a clustering of the data with a particular K?
+Ditto for ``all possible clusters of interest'' -- do you mean all choices of K, so 5,9,11,15 in the below.}
+
+Below we run the clustering for \(k\) equal to \(2-20\). The \texttt{splines\_kmeans} function returns the WCSS for each cluster, which we sum to get the total WCSS.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{all_possible_n_clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{2}\OperatorTok{:}\DecValTok{20}\NormalTok{)}
 \NormalTok{all_clustering =}\StringTok{ }\KeywordTok{list}\NormalTok{()}
@@
 The Elbow method to choose the number of clusters relies on visualization aid
-to choose the number of cluster. The method relies on plotting the within
-cluster sum of squares as a function of the number of clusters. At some point,
-the WCSS will stop dropping, giving an angle in the graph. The number of
+to choose the number of clusters. The method relies on plotting the within
+cluster sum of squares (WCSS) as a function of the number of clusters. At some point,
+the WCSS will start decreasing more slowly, giving an angle or ``elbow'' in the graph. The number of
 cluster is chosen at this ``Elbow point.''
+
+We plot the WCSS for \(k=2-20\) here.
+We see that as expected the WCSS continues to drop, but there is no clear drop in the decrease, except for very small values of \(k\) (3-4 clusters).
+However, 3-4 seems a very small number of gene clusters to find, given the complexity
 
@@
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/elbow_method-1} \end{center}
+
+\textbf{EAP: I've increased this to be k=2-20, because you can't actually interpret these plots without a dense number of k; also you do 2-20 for stability}
 
@@
 
-Using the bootstrapping strategy:
+Using the bootstrapping strategy we sample with replacement a sample of the same size as our original clustering (i.e.~3950 genes):
 
@@
 
-Using the subsampling strategy, keeping 80\% of the genes:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{subsample_proportion =}\StringTok{ }\DecValTok{1}
+Using the subsampling strategy, we take a unique subset of the genes, keeping 80\% of the genes:
+
+\textbf{EAP: \texttt{subsample\_proportion} was set to 1 in the code, but shouldn't it be 0.8?
+I changed it to 0.8.
+But it seems like you don't actually run this code, but read existing runs, so maybe was right there?}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{subsample_proportion =}\StringTok{ }\FloatTok{0.8}
 \NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
@@
 
-Here we plot to stability matrix of the top 1000 genes for \(k=5\) and \(k=20\).
+\textbf{EAP: So this is very confusing to understand, because you read the results from a file (\texttt{stability\_5.tsv}), but I don't know what this file contains, nor how they were calculated.
+And I don't know what a stability matrix is -- you've only talked about a consensus matrix.
+There should be the code here for a single k, but not evaluated, and say that for convenience we provide the results of running this code as a file for users to play with.
+I've tried to fix up the text to be what I think it should be}
+
+We will run \textbf{EAP: subsampling???} method on the top 1000 genes \textbf{EAP: what is top??}, and do it \(B=20\) times for each of \(k=2-20\).
+We show the code below, but because of the time it takes we have evaluate these values separately and provided the results for users to explore more quickly.
+
+For now, we bring in the results for \(k=5\) and \(k=20\).
 
@@
 \NormalTok{stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_5.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
+\NormalTok{stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_20.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\textbf{EAP: It's very confusing that the headings here are C1, \ldots{} C20.
+It makes it seem like cluster 1, \ldots{} cluster 20, when I think (?) it is really 20 resampling results.
+Would be good to change that to, e.g.~B1,\ldots{},B20 or something.}
+
+For each gene, we have an assignment to a cluster over the 20 resampling runs. For example, for \(k=5\):
+
+\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
+\hline
+  & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & C18 & C19 & C20\\
+\hline
+AA204140 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 1 & 3 & 4 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
+\hline
+AB008911 & 5 & 5 & 1 & 4 & 1 & 2 & 3 & 3 & 2 & 3 & 3 & 3 & 4 & 4 & 5 & 5 & 2 & 4 & 1 & 4\\
+\hline
+AB010313 & 3 & 3 & 1 & 4 & 1 & 2 & 4 & 3 & 2 & 5 & 3 & 1 & 5 & 4 & 2 & 5 & 3 & 3 & 4 & 2\\
+\hline
+AB010342 & 5 & 3 & 1 & 5 & 1 & 1 & 3 & 3 & 1 & 3 & 3 & 3 & 4 & 3 & 5 & 4 & 2 & 3 & 4 & 4\\
+\hline
+AB011473 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+\hline
+AB099518 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+\hline
+\end{tabular}
+
+Now we will use the function \texttt{consensus\_matrix} in the \texttt{moanin} package to calculate proportion of times each pair of samples was clustered together across the 20 resampling runs, and plot the heatmap of those consensus matrices for \(k=5\) and \(k=20\).
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
@@
      \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
-     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"results/stability_20.tsv"}\NormalTok{, }\DataTypeTok{sep=}\StringTok{"}\CharTok{\textbackslash{}t}\StringTok{"}\NormalTok{)}
+     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=5"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-1} \end{center}
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
@@
      \DataTypeTok{Colv=}\OtherTok{FALSE}\NormalTok{,}
-     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-2} \end{center}
+     \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=20"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-2} \end{center}
+
+We can see that the choice of \(K=5\) seems much more stable across resampling runs than that of \(K=20\).
 
 \hypertarget{the-model-explorer-strategy}{%
-\subparagraph{The model explorer strategy}\label{the-model-explorer-strategy}}
+\paragraph{The model explorer strategy}\label{the-model-explorer-strategy}}
 
@@
 number of clusters exploiting the observation that if the number of clusters
-is correct, the clustering results are stable to bootstrapping. The
+is correct, the clustering results are stable to bootstrap resampling, as described above. The
 distribution of similaries between bootstrapped results for each \(k\) can thus
@@
 
-The model explorer strategy works as follows. First, choose a similarity
-measure between two partitions or clusters \(S(C_1, C_2)\). Examples are the
-normalized mutual information or Fowlkes-Mallows. Then perform \(n\) bootstrap
+The model explorer strategy works as follows. For a single choice of \(k\), first perform \(n\) bootstrap
 experiments to estimate the cluster centroids, followed by a step of assigning
-a label to all data points. Finally, compute the pairwise similarity measure
-between all bootstrapped partition, and plot the cumulative density of the
+a label to all data points. Then, choose a similarity
+measure between two partitions or clusters \(S(C_1, C_2)\).
+\textbf{EAP: This is confusing as to whether this is a measure of similarity across different bootstrap resamples (i.e.~different sets of clusters) or across two different clusters in the same resample.
+Perhaps also because of confusion of headings of different resamples between labeled C1, C2, but those aren't cluster 1/2 (see comment above)} Examples are the
+normalized mutual information or Fowlkes-Mallows. Finally, compute the pairwise similarity measure
+between all bootstrapped partitions (i.e.~B choose 2 pairs). Repeat this procedure for different \(k\) and plot per \(k\) the cumulative density of the
 obtained scores.
 
+We have already run the bootstrap resampling for values \(k=2-20\) and saved the results. We will read that data in, and use it to calculate the pairwise similarity scores.
+
+The function \texttt{plot\_model\_explorer}
+
 \begin{Shaded}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-33-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-38-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
+
+This analysis doesn't necessarily pick a particular \(k\), but can help decide between \(k\) within a desired range, for example, or to avoid \(k\) that degrade the stability.
+
+\textbf{EAP: \texttt{plot\_model\_explorer} is relatively slow, because it does all those pairwise calculations.
+And then to do it for smaller subset, redo all the calculations.
+Can't the output be saved, or the calculations be a different step from the plotting?}
 
 \hypertarget{consensus-clustering-as-a-way-to-find-k}{%
-\subparagraph{\texorpdfstring{Consensus clustering as a way to find \(k\)}{Consensus clustering as a way to find k}}\label{consensus-clustering-as-a-way-to-find-k}}
+\paragraph{\texorpdfstring{Consensus clustering as a way to find \(k\)}{Consensus clustering as a way to find k}}\label{consensus-clustering-as-a-way-to-find-k}}
 
@@
 bootstrapped clustering, the authors suggests plotting the cumulative density
-of elements of the consensus matrix.
+of elements of the consensus matrix. We provide a function \texttt{plot\_cdf\_consensus} in \texttt{moanin} to do this
 
@@
 
+\textbf{EAP: We should again provide the code that goes with these images}
+
 The consensus clustering method suggest that the most stable is \(k=2\), which
@@
 
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+This workflow provides a tutoral for the analysis of time-course gene
+expression data in R, illustrated through the analysis of mice lung tissue
+exposed to different influenza strain. It covers four main steps; (1) quality
+control and normalization; (2) differential expression analysis; (3)
+clustering of time-course gene expression data; (4) downstream analysis of
+clusters.
+
 \hypertarget{software-and-data-availability}{%
@@
 \url{https://github.com/NelleV/2019timecourse-rnaseq-pipeline}.
-Archived source code at the time of publication can be foud at \url{XXX}.
+Archived source code at the time of publication can be found at \url{XXX}.
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-This workflow provides a tutoral for the analysis of time-course gene
-expression data in R, illustrated through the analysis of mice lung tissue
-exposed to different influenza strain. It covers four main steps; (1) quality
-control and normalization; (2) differential expression analysis; (3)
-clustering of time-course gene expression data; (4) downstream analysis of
-clusters.
-
-\hypertarget{software-and-data-availability-1}{%
-\section{Software and data availability}\label{software-and-data-availability-1}}
-
-The source code for this workflow can be found at \url{https://github.com/NelleV/2019timecourse-rnaseq-pipeline}.
-Data used in this workflow are available from NCBI GEO, accession GSE95601.
-
 \hypertarget{author-contributions}{%

### 2019-09-27 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \midrule
-GO:0003002 & regionalization & 172 & 153 & 134.41 & 0.00026 & 0.263\\
-GO:0048562 & embryonic organ morphogenesis & 152 & 135 & 118.78 & 0.00045 & 0.332\\
-GO:0010970 & transport along microtubule & 82 & 75 & 64.08 & 0.00046 & 0.332\\
-GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.29 & 0.00059 & 0.373\\
-GO:0060173 & limb development & 92 & 84 & 71.89 & 0.00068 & 0.382\\
+GO:0007224 & smoothened signaling pathway & 74 & 69 & 57.82 & 0.00041 & 0.326\\
+GO:0010970 & transport along microtubule & 82 & 75 & 64.07 & 0.00045 & 0.326\\
+GO:0048562 & embryonic organ morphogenesis & 155 & 137 & 121.10 & 0.00067 & 0.378\\
+GO:0060173 & limb development & 92 & 84 & 71.88 & 0.00067 & 0.378\\
+GO:0006805 & xenobiotic metabolic process & 47 & 45 & 36.72 & 0.00089 & 0.451\\
 \addlinespace
-GO:0019395 & fatty acid oxidation & 53 & 51 & 41.42 & 0.00134 & 0.619\\
-GO:0055114 & oxidation-reduction process & 536 & 461 & 418.84 & 0.00139 & 0.619\\
-GO:0060070 & canonical Wnt signaling pathway & 149 & 131 & 116.43 & 0.00147 & 0.619\\
-GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.71 & 0.00219 & 0.852\\
-GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.79 & 0.00295 & 1.000\\
+GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.76 & 0.00099 & 0.456\\
+GO:0019395 & fatty acid oxidation & 53 & 51 & 41.41 & 0.00133 & 0.562\\
+GO:0055114 & oxidation-reduction process & 533 & 458 & 416.42 & 0.00152 & 0.593\\
+GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.69 & 0.00216 & 0.783\\
+GO:0010811 & positive regulation of cell-substrate ad... & 77 & 70 & 60.16 & 0.00260 & 0.879\\
 \addlinespace
-GO:0010811 & positive regulation of cell-substrate ad... & 81 & 73 & 63.30 & 0.00384 & 1.000\\
-GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00437 & 1.000\\
-GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00437 & 1.000\\
-GO:0007368 & determination of left/right symmetry & 66 & 60 & 51.57 & 0.00536 & 1.000\\
-GO:0060078 & regulation of postsynaptic membrane pote... & 66 & 60 & 51.57 & 0.00536 & 1.000\\
+GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.78 & 0.00293 & 0.929\\
+GO:0021510 & spinal cord development & 48 & 45 & 37.50 & 0.00339 & 1.000\\
+GO:0016571 & histone methylation & 74 & 67 & 57.82 & 0.00418 & 1.000\\
+GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00435 & 1.000\\
+GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00435 & 1.000\\
 \bottomrule

### 2019-10-02 scripts/manuscript.Rmd
Small fixes on the manuscript

@@
 
-source("utils.R")
 ```
@@
 ```
-
-**EAP: It seems like a problem to have the colors defined in a utils.R. Makes it hard It's not echo'd by markdown and so readers won't know where they come from. (I didn't).**
 
@@
 
-The meta data contains information about the treatment group, the replicate, and the timepoint for each observation:
+The meta data contains information about the treatment group, the replicate,
+and the timepoint for each observation:
 
@@
 
-*EAP*: I moved this here, because we use meta a lot in the eda.
+Before we dive into the exploratory analysis and quality control, let us
+define color schemes for our data that we will use across the whole analysis.
+We define coloschemes for groups, timepoints as named vectors.
+
+```{r}
+group_colors = c(
+    "M"="dodgerblue4",
+    "K"="gold",
+    "C"="orange",
+    "VH"="red4",
+    "VL"="red2")
+time_colors = grDevices::rainbow(15)[1:14]
+names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
+
+# Combine all color schemes into one named lists.
+ann_colors = list(
+    Timepoint=time_colors,
+    Group=group_colors
+    )
+
+replicate_markers = c(15, 17, 19)
+names(replicate_markers) = c(1, 2, 3)
+ann_markers = list(
+    Replicate=replicate_markers)
+```
 

### 2019-10-02 scripts/utils.R

@@ -1,18 +1,0 @@
-group_colors = c(
-    "M"="dodgerblue4",
-    "K"="gold",
-    "C"="orange",
-    "VH"="red4",
-    "VL"="red2")
-time_colors = grDevices::rainbow(15)[1:14]
-names(time_colors) = c(0, 3, 6, 9, 12, 18, 24, 30, 36, 48, 60, 72, 120, 168)
-
-ann_colors = list(
-    Timepoint=time_colors,
-    Group=group_colors
-    )
-
-replicate_markers = c(15, 17, 19)
-names(replicate_markers) = c(1, 2, 3)
-ann_markers = list(
-    Replicate=replicate_markers)

### 2019-10-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\textbf{EAP: It seems like a problem to have the colors defined in a utils.R. Makes it hard It's not echo'd by markdown and so readers won't know where they come from. (I didn't).}
-
 \hypertarget{introduction}{%
@@
 
-\emph{EAP}: I moved this here, because we use meta a lot in the eda.
+Before we dive into the exploratory analysis and quality control, let us
+define color schemes for our data that we will use across the whole analysis.
+We define coloschemes for groups, timepoints as named vectors.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{group_colors =}\StringTok{ }\KeywordTok{c}\NormalTok{(}
+    \StringTok{"M"}\NormalTok{=}\StringTok{"dodgerblue4"}\NormalTok{,}
+    \StringTok{"K"}\NormalTok{=}\StringTok{"gold"}\NormalTok{,}
+    \StringTok{"C"}\NormalTok{=}\StringTok{"orange"}\NormalTok{,}
+    \StringTok{"VH"}\NormalTok{=}\StringTok{"red4"}\NormalTok{,}
+    \StringTok{"VL"}\NormalTok{=}\StringTok{"red2"}\NormalTok{)}
+\NormalTok{time_colors =}\StringTok{ }\NormalTok{grDevices}\OperatorTok{::}\KeywordTok{rainbow}\NormalTok{(}\DecValTok{15}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{14}\NormalTok{]}
+\KeywordTok{names}\NormalTok{(time_colors) =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{9}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{24}\NormalTok{, }\DecValTok{30}\NormalTok{, }\DecValTok{36}\NormalTok{, }\DecValTok{48}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{72}\NormalTok{, }\DecValTok{120}\NormalTok{, }\DecValTok{168}\NormalTok{)}
+
+\CommentTok{# Combine all color schemes into one named lists.}
+\NormalTok{ann_colors =}\StringTok{ }\KeywordTok{list}\NormalTok{(}
+    \DataTypeTok{Timepoint=}\NormalTok{time_colors,}
+    \DataTypeTok{Group=}\NormalTok{group_colors}
+\NormalTok{    )}
+
+\NormalTok{replicate_markers =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{19}\NormalTok{)}
+\KeywordTok{names}\NormalTok{(replicate_markers) =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{)}
+\NormalTok{ann_markers =}\StringTok{ }\KeywordTok{list}\NormalTok{(}
+    \DataTypeTok{Replicate=}\NormalTok{replicate_markers)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-38-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-40-1} \end{center}
 

### 2019-10-08 scripts/manuscript.Rmd
Bunch of fixes on the manuscript

@@
 throughout this analysis: the condition and timepoints of each samples, the
-formula object used or the basis or the degrees of freedom of the model. 
-**EAP:** You make it sound like or, but in fact the object has all of these elements, no?
-
-If no formula is provided, it will default to the following, which provides for a different spline fit for every group (as defined by the `Group` variable in the `meta` data.frame): 
+formula object used, the basis, and the degrees of freedom of the model.
+
+If no formula is provided, it will default to the following, which provides
+for a different spline fit for every group (as defined by the `Group` variable
+in the `meta` data.frame): 
 `formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.
@@
 week between control and each of the influenza strains.
-
-**EAP: Should `timely_barplot` be a function of `moanin`?**
-
-**NV: I'd rather not include these types of functions that are hard to test
-and clutter the namespace. In order to provide reasonable visualization, they
-also often need to be tweaked specfically to match the data (here, the ylim is
-set specifically to the data**
-
-**EAP: Okay, I'll leave that to you.
-Most of those tweaked things can just be in a ...
-argument and plotting functions that do annoying things are popular.
-I've at least rewritten it so users can more easily change those features in calls to the function rather than mucking about in the code**
 
@@
 ```{r fig.height=6, fig.width=6}
-#order them by their score
-cluster2Score=cluster2Score[order(scores_and_labels$scores[cluster2Score,2], decreasing = TRUE)]
-dataToPlot=rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[2,])
+# order them by their score
+cluster2Score = cluster2Score[order(scores_and_labels$scores[cluster2Score, 8], decreasing = TRUE)]
+dataToPlot = rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[8, ])
 plot_splines_data(dataToPlot, moanin_model,
@@
 
+**Nelle: in my experience, the large the cluster is, the noisier the centroid
+estimation is. I've switched to cluster 8, as this is the cluster we focus our
+interpretation, and the top scoring genes seem to be to be closely related to
+the centroid estimates**
+
 ### Looking at specific clusters in details.
@@
 
-We've already shown how we can plot a few example genes, however, it can be hard to make sense of individual genes given the amount of noise, as well as being hard to draw conclusions based on a few genes.
+We've already shown how we can plot a few example genes, however, it can be
+hard to make sense of individual genes given the amount of noise, as well as
+being hard to draw conclusions based on a few genes.
 
@@
 
-First, we select the genes of interest, namely those in cluster 8, based on our goodness-of-fit assignment.
+First, we select the genes of interest, namely those in cluster 8, based on
+our goodness-of-fit assignment.
 
@@
 First, let us run the clustering for all possible clusters of interest. We
-will, for each clustering experiment, conserve (1) with within cluster sum of
-squares; (2) the labels assigned to all genes. **EAP: I don't understand what you are trying to say in this last sentence. I think there's a typo**
-
-**EAP: Also, I don't know what you mean by "clustering experiment" -- do you mean a clustering of the data with a particular K?
-Ditto for "all possible clusters of interest" -- do you mean all choices of K, so 5,9,11,15 in the below.**
-
-Below we run the clustering for $k$ equal to $2-20$. The `splines_kmeans` function returns the WCSS for each cluster, which we sum to get the total WCSS.
+will, for each clustering, conserve (1) with within cluster sum of
+squares; (2) the clustering assignement (or label) for each gene.
+
+Below we run the clustering for $k$ equal to $2-20$. The `splines_kmeans`
+function returns the WCSS for each cluster, which we sum to get the total
+WCSS.
 
@@
 to choose the number of clusters. The method relies on plotting the within
-cluster sum of squares (WCSS) as a function of the number of clusters. At some point,
-the WCSS will start decreasing more slowly, giving an angle or "elbow" in the graph.  The number of
-cluster is chosen at this "Elbow point."
-
-We plot the WCSS for $k=2-20$ here.
-We see that as expected the WCSS continues to drop, but there is no clear drop in the decrease, except for very small values of $k$ (3-4 clusters).
-However, 3-4 seems a very small number of gene clusters to find, given the complexity
+cluster sum of squares (WCSS) as a function of the number of clusters. At some
+point, the WCSS will start decreasing more slowly, giving an angle or "elbow"
+in the graph.  The number of cluster is chosen at this "Elbow point."
+
+We plot the WCSS for $k=2-20$ here. We see that as expected the WCSS continues
+to drop, but there is no clear drop in the decrease, except for very small
+values of $k$ (3-4 clusters). However, 3-4 seems a very small number of gene
+clusters to find, given the complexity
 
@@
 ```
-
-**EAP: I've increased this to be k=2-20, because you can't actually interpret these plots without a dense number of k; also you do 2-20 for stability**
 
@@
 
-Using the bootstrapping strategy we sample with replacement a sample of the same size as our original clustering (i.e. `r dim(y)[1]` genes):
+Using the bootstrapping strategy we sample with replacement a sample of the
+same size as our original clustering (i.e. `r dim(y)[1]` genes):
 
@@
 
-**EAP: `subsample_proportion` was set to 1 in the code, but shouldn't it be 0.8?
-I changed it to 0.8.
-But it seems like you don't actually run this code, but read existing runs, so maybe was right there?**
+**EAP: `subsample_proportion` was set to 1 in the code, but shouldn't it be
+0.8? I changed it to 0.8. But it seems like you don't actually run this code,
+but read existing runs, so maybe was right there?**
+
+**Nelle: that was a mistake in the code. I don't actually run subsampling
+experiments: only the bootstrap, and it's done in a separate file as it is too
+time consuming to run in an Rmd file**
 
@@
 
-We will run **EAP: subsampling???** method on the top 1000 genes **EAP: what is top??**, and do it $B=20$ times for each of $k=2-20$.
-We show the code below, but because of the time it takes we have evaluate these values separately and provided the results for users to explore more quickly.
+We run bootstrap method on all genes differentially expressed and with a
+log-fold-change higher than 2 (computed with the `lfc_max` method), and do it
+$B=20$ times for each of $k=2-20$. We show the code below, but because of the
+time it takes we have evaluate these values separately and provided the
+results for users to explore more quickly.
+
+```{r bootstrapped_clustering, eval=FALSE}
+pvalues = de_analysis[, pval_col_to_keep]
+
+splines_model = moanin::create_moanin_model(meta, degrees_of_freedom=6)
+fishers_pval = moanin:::pvalues_fisher_method(pvalues)
+fishers_qval = stats::p.adjust(fishers_pval)
+
+# Keep only the genes found differentially expressed after combining the
+# p-values using Fisher's method and with a log fold change greater than 2
+genes_to_keep = row.names(
+    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
+y = as.matrix(data[genes_to_keep, ])
+
+# You may want to set the random seed of the experiment so that the results
+# don't vary if you rerun the experiment.
+# set.seed(random_seed)
+n_genes = dim(y)[1] * sample_proportion
+indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
+
+random_subsample_y = y[indices, ]
+kmeans_clusters = moanin::splines_kmeans(
+    random_subsample_y, splines_model, n_clusters=n_clusters,
+    random_seed=random_seed,
+    n_init=20)
+
+# Perform prediction on the whole set of data.
+kmeans_clusters = moanin:::splines_kmeans_prediction(
+    y, kmeans_clusters)
+```
 
@@
 ```
-**EAP: It's very confusing that the headings here are C1, ...
-C20.
-It makes it seem like cluster 1, ...
-cluster 20, when I think (?) it is really 20 resampling results.
-Would be good to change that to, e.g.
-B1,...,B20 or something.**
-
-For each gene, we have an assignment to a cluster over the 20 resampling runs. For example, for $k=5$:
+
+Each column correspond to a bootstrap sample, each row to a gene, and each
+entry to the label found for that particular clustering. Thus, for each gene,
+we have an assignment to a cluster over the 20 resampling runs. For example,
+for $k=5$:
 
@@
 
-Now we will use the function `consensus_matrix` in the `moanin` package to calculate proportion of times each pair of samples was clustered together across the 20 resampling runs, and plot the heatmap of those consensus matrices for $k=5$ and $k=20$.
+Now we will use the function `consensus_matrix` in the `moanin` package to
+calculate proportion of times each pair of samples was clustered together
+across the 20 resampling runs, and plot the heatmap of those consensus
+matrices for $k=5$ and $k=20$.
 
@@
 number of clusters exploiting the observation that if the number of clusters
-is correct, the clustering results are stable to bootstrap resampling, as described above.  The
-distribution of similaries between bootstrapped results for each $k$ can thus
-be compared for different values of $k$ and guide the user in the choice of
-number of clusters.
-
-The model explorer strategy works as follows. For a single choice of $k$, first perform $n$ bootstrap
-experiments to estimate the cluster centroids, followed by a step of assigning
-a label to all data points. Then, choose a similarity
-measure between two partitions or clusters $S(C_1, C_2)$.
-**EAP: This is confusing as to whether this is a measure of similarity across different bootstrap resamples (i.e.
-different sets of clusters) or across two different clusters in the same resample.
-Perhaps also because of confusion of headings of different resamples between labeled C1, C2, but those aren't cluster 1/2 (see comment above)** Examples are the
-normalized mutual information or Fowlkes-Mallows. Finally, compute the pairwise similarity measure
-between all bootstrapped partitions (i.e. B choose 2 pairs). Repeat this procedure for different $k$ and plot per $k$ the cumulative density of the
-obtained scores.
-
-We have already run the bootstrap resampling for values $k=2-20$ and saved the results. We will read that data in, and use it to calculate the pairwise similarity scores.
+is correct, the clustering results are stable to bootstrap resampling, as
+described above.  The distribution of similaries between bootstrapped results
+for each $k$ can thus be compared for different values of $k$ and guide the
+user in the choice of number of clusters.
+
+The model explorer strategy works as follows. For a single choice of $k$,
+first perform $n$ bootstrap experiments to estimate the cluster centroids,
+followed by a step of assigning a label to all data points. Then, choose a
+similarity measure between two partitions or clusters $S(B_1, B_2)$.  Examples
+are the normalized mutual information or Fowlkes-Mallows. Finally, compute the
+pairwise similarity measure between all bootstrapped partitions (i.e. B choose
+2 pairs). Repeat this procedure for different $k$ and plot per $k$ the
+cumulative density of the obtained scores.
+
+We have already run the bootstrap resampling for values $k=2-20$ and saved the
+results. We will read that data in, and use it to calculate the pairwise
+similarity scores.
 
@@
     stability = read.table(filename, sep="\t")
-    all_labels[[paste0("C", i)]] = stability
+    all_labels[[paste0("B", i)]] = stability
 }
@@
 
-The function `plot_model_explorer`
+The function `plot_model_explorer` takes in input a list of the bootstrapped
+results for all labels.
 
@@
 ```{r fig.width=6, fig.height=6}
-clusters = c("C6", "C7", "C8", "C9", "C10")
+clusters = c("B6", "B7", "B8", "B9", "B10")
 selected_labels = all_labels[clusters]
@@
 bootstrapped clustering, the authors suggests plotting the cumulative density
-of elements of the consensus matrix. We provide a function `plot_cdf_consensus` in `moanin` to do this
+of elements of the consensus matrix. We provide a function
+`plot_cdf_consensus` in `moanin` to do this
 
@@
 
+```{r eval=FALSE}
+auc_scores = moanin:::get_auc_similarity_scores(all_labels)
+plot(n_clusters, auc_scores, col="black", type="b", pch=16)
+```
+
 ```{r echo=FALSE}
@@
 
+```{r eval=FALSE}
+delta_auc = diff(auc_scores)/auc_scores[1:(length(auc_scores)-1)]
+plot(3:20, delta_auc, col="black", type="b")
+```
+
 ```{r echo=FALSE}
@@
 ```
-**EAP: We should again provide the code that goes with these images**
 

### 2019-10-08 scripts/concatenate_clustering_stability_results.R

@@ -21,7 +21,7 @@ for(folder in folders){
     }
     colnames(all_labels) = lapply(
 	1:dim(all_labels)[2],
-	function(x){paste0("C", x)})
+	function(x){paste0("B", x)})
     outname = paste0("results/", folder, ".tsv")
     write.table(all_labels, outname, sep="\t")
 }

### 2019-10-08 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 throughout this analysis: the condition and timepoints of each samples, the
-formula object used or the basis or the degrees of freedom of the model.
-\textbf{EAP:} You make it sound like or, but in fact the object has all of these elements, no?
-
-If no formula is provided, it will default to the following, which provides for a different spline fit for every group (as defined by the \texttt{Group} variable in the \texttt{meta} data.frame):
+formula object used, the basis, and the degrees of freedom of the model.
+
+If no formula is provided, it will default to the following, which provides
+for a different spline fit for every group (as defined by the \texttt{Group} variable
+in the \texttt{meta} data.frame):
 \texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.
@@
 
-\textbf{EAP: Should \texttt{timely\_barplot} be a function of \texttt{moanin}?}
-
-\textbf{NV: I'd rather not include these types of functions that are hard to test
-and clutter the namespace. In order to provide reasonable visualization, they
-also often need to be tweaked specfically to match the data (here, the ylim is
-set specifically to the data}
-
-\textbf{EAP: Okay, I'll leave that to you.
-Most of those tweaked things can just be in a \ldots{} argument and plotting functions that do annoying things are popular.
-I've at least rewritten it so users can more easily change those features in calls to the function rather than mucking about in the code}
-
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
@@
 \begin{Highlighting}[]
-\CommentTok{#order them by their score}
-\NormalTok{cluster2Score=cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score,}\DecValTok{2}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
-\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{2}\NormalTok{,])}
+\CommentTok{# order them by their score}
+\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score, }\DecValTok{8}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{8}\NormalTok{, ])}
 \KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
@@
 It's seems like an important plot to make}
+
+\textbf{Nelle: in my experience, the large the cluster is, the noisier the centroid
+estimation is. I've switched to cluster 8, as this is the cluster we focus our
+interpretation, and the top scoring genes seem to be to be closely related to
+the centroid estimates}
 
@@
 First, let us run the clustering for all possible clusters of interest. We
-will, for each clustering experiment, conserve (1) with within cluster sum of
-squares; (2) the labels assigned to all genes. \textbf{EAP: I don't understand what you are trying to say in this last sentence. I think there's a typo}
-
-\textbf{EAP: Also, I don't know what you mean by ``clustering experiment'' -- do you mean a clustering of the data with a particular K?
-Ditto for ``all possible clusters of interest'' -- do you mean all choices of K, so 5,9,11,15 in the below.}
-
-Below we run the clustering for \(k\) equal to \(2-20\). The \texttt{splines\_kmeans} function returns the WCSS for each cluster, which we sum to get the total WCSS.
+will, for each clustering, conserve (1) with within cluster sum of
+squares; (2) the clustering assignement (or label) for each gene.
+
+Below we run the clustering for \(k\) equal to \(2-20\). The \texttt{splines\_kmeans}
+function returns the WCSS for each cluster, which we sum to get the total
+WCSS.
 
@@
 to choose the number of clusters. The method relies on plotting the within
-cluster sum of squares (WCSS) as a function of the number of clusters. At some point,
-the WCSS will start decreasing more slowly, giving an angle or ``elbow'' in the graph. The number of
-cluster is chosen at this ``Elbow point.''
-
-We plot the WCSS for \(k=2-20\) here.
-We see that as expected the WCSS continues to drop, but there is no clear drop in the decrease, except for very small values of \(k\) (3-4 clusters).
-However, 3-4 seems a very small number of gene clusters to find, given the complexity
+cluster sum of squares (WCSS) as a function of the number of clusters. At some
+point, the WCSS will start decreasing more slowly, giving an angle or ``elbow''
+in the graph. The number of cluster is chosen at this ``Elbow point.''
+
+We plot the WCSS for \(k=2-20\) here. We see that as expected the WCSS continues
+to drop, but there is no clear drop in the decrease, except for very small
+values of \(k\) (3-4 clusters). However, 3-4 seems a very small number of gene
+clusters to find, given the complexity
 
@@
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/elbow_method-1} \end{center}
-
-\textbf{EAP: I've increased this to be k=2-20, because you can't actually interpret these plots without a dense number of k; also you do 2-20 for stability}
 
@@
 
-Using the bootstrapping strategy we sample with replacement a sample of the same size as our original clustering (i.e.~3950 genes):
+Using the bootstrapping strategy we sample with replacement a sample of the
+same size as our original clustering (i.e.~3950 genes):
 
@@
 
-\textbf{EAP: \texttt{subsample\_proportion} was set to 1 in the code, but shouldn't it be 0.8?
-I changed it to 0.8.
-But it seems like you don't actually run this code, but read existing runs, so maybe was right there?}
+\textbf{EAP: \texttt{subsample\_proportion} was set to 1 in the code, but shouldn't it be
+0.8? I changed it to 0.8. But it seems like you don't actually run this code,
+but read existing runs, so maybe was right there?}
+
+\textbf{Nelle: that was a mistake in the code. I don't actually run subsampling
+experiments: only the bootstrap, and it's done in a separate file as it is too
+time consuming to run in an Rmd file}
 
@@
 
-We will run \textbf{EAP: subsampling???} method on the top 1000 genes \textbf{EAP: what is top??}, and do it \(B=20\) times for each of \(k=2-20\).
-We show the code below, but because of the time it takes we have evaluate these values separately and provided the results for users to explore more quickly.
+We run bootstrap method on all genes differentially expressed and with a
+log-fold-change higher than 2 (computed with the \texttt{lfc\_max} method), and do it
+\(B=20\) times for each of \(k=2-20\). We show the code below, but because of the
+time it takes we have evaluate these values separately and provided the
+results for users to explore more quickly.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{pvalues =}\StringTok{ }\NormalTok{de_analysis[, pval_col_to_keep]}
+
+\NormalTok{splines_model =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+\NormalTok{fishers_pval =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
+\NormalTok{fishers_qval =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}
+
+\CommentTok{# Keep only the genes found differentially expressed after combining the}
+\CommentTok{# p-values using Fisher's method and with a log fold change greater than 2}
+\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
+\NormalTok{    lfc_max[(}\KeywordTok{rowSums}\NormalTok{(lfc_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
+\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}
+
+\CommentTok{# You may want to set the random seed of the experiment so that the results}
+\CommentTok{# don't vary if you rerun the experiment.}
+\CommentTok{# set.seed(random_seed)}
+\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{] }\OperatorTok{*}\StringTok{ }\NormalTok{sample_proportion}
+\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}
+
+\NormalTok{random_subsample_y =}\StringTok{ }\NormalTok{y[indices, ]}
+\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
+\NormalTok{    random_subsample_y, splines_model, }\DataTypeTok{n_clusters=}\NormalTok{n_clusters,}
+    \DataTypeTok{random_seed=}\NormalTok{random_seed,}
+    \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}
+
+\CommentTok{# Perform prediction on the whole set of data.}
+\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{splines_kmeans_prediction}\NormalTok{(}
+\NormalTok{    y, kmeans_clusters)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\textbf{EAP: It's very confusing that the headings here are C1, \ldots{} C20.
-It makes it seem like cluster 1, \ldots{} cluster 20, when I think (?) it is really 20 resampling results.
-Would be good to change that to, e.g.~B1,\ldots{},B20 or something.}
-
-For each gene, we have an assignment to a cluster over the 20 resampling runs. For example, for \(k=5\):
+Each column correspond to a bootstrap sample, each row to a gene, and each
+entry to the label found for that particular clustering. Thus, for each gene,
+we have an assignment to a cluster over the 20 resampling runs. For example,
+for \(k=5\):
 
@@
 \hline
-  & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & C18 & C19 & C20\\
-\hline
-AA204140 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 1 & 3 & 4 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
-\hline
-AB008911 & 5 & 5 & 1 & 4 & 1 & 2 & 3 & 3 & 2 & 3 & 3 & 3 & 4 & 4 & 5 & 5 & 2 & 4 & 1 & 4\\
-\hline
-AB010313 & 3 & 3 & 1 & 4 & 1 & 2 & 4 & 3 & 2 & 5 & 3 & 1 & 5 & 4 & 2 & 5 & 3 & 3 & 4 & 2\\
-\hline
-AB010342 & 5 & 3 & 1 & 5 & 1 & 1 & 3 & 3 & 1 & 3 & 3 & 3 & 4 & 3 & 5 & 4 & 2 & 3 & 4 & 4\\
-\hline
-AB011473 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
-\hline
-AB099518 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+  & B1 & B2 & B3 & B4 & B5 & B6 & B7 & B8 & B9 & B10 & B11 & B12 & B13 & B14 & B15 & B16 & B17 & B18 & B19 & B20\\
+\hline
+A\_51\_P114236 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+\hline
+A\_51\_P172409 & 1 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 4 & 3 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
+\hline
+A\_51\_P188401 & 1 & 5 & 5 & 4 & 5 & 5 & 5 & 3 & 2 & 4 & 4 & 4 & 2 & 3 & 5 & 4 & 2 & 4 & 3 & 4\\
+\hline
+A\_51\_P191274 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+\hline
+A\_51\_P226791 & 1 & 4 & 2 & 4 & 3 & 4 & 2 & 1 & 3 & 3 & 4 & 2 & 5 & 2 & 3 & 5 & 2 & 2 & 5 & 3\\
+\hline
+A\_51\_P277048 & 1 & 1 & 4 & 5 & 5 & 2 & 5 & 5 & 2 & 4 & 3 & 4 & 2 & 5 & 4 & 3 & 4 & 4 & 3 & 5\\
 \hline
@@
 
-Now we will use the function \texttt{consensus\_matrix} in the \texttt{moanin} package to calculate proportion of times each pair of samples was clustered together across the 20 resampling runs, and plot the heatmap of those consensus matrices for \(k=5\) and \(k=20\).
+Now we will use the function \texttt{consensus\_matrix} in the \texttt{moanin} package to
+calculate proportion of times each pair of samples was clustered together
+across the 20 resampling runs, and plot the heatmap of those consensus
+matrices for \(k=5\) and \(k=20\).
 
@@
 number of clusters exploiting the observation that if the number of clusters
-is correct, the clustering results are stable to bootstrap resampling, as described above. The
-distribution of similaries between bootstrapped results for each \(k\) can thus
-be compared for different values of \(k\) and guide the user in the choice of
-number of clusters.
-
-The model explorer strategy works as follows. For a single choice of \(k\), first perform \(n\) bootstrap
-experiments to estimate the cluster centroids, followed by a step of assigning
-a label to all data points. Then, choose a similarity
-measure between two partitions or clusters \(S(C_1, C_2)\).
-\textbf{EAP: This is confusing as to whether this is a measure of similarity across different bootstrap resamples (i.e.~different sets of clusters) or across two different clusters in the same resample.
-Perhaps also because of confusion of headings of different resamples between labeled C1, C2, but those aren't cluster 1/2 (see comment above)} Examples are the
-normalized mutual information or Fowlkes-Mallows. Finally, compute the pairwise similarity measure
-between all bootstrapped partitions (i.e.~B choose 2 pairs). Repeat this procedure for different \(k\) and plot per \(k\) the cumulative density of the
-obtained scores.
-
-We have already run the bootstrap resampling for values \(k=2-20\) and saved the results. We will read that data in, and use it to calculate the pairwise similarity scores.
-
-The function \texttt{plot\_model\_explorer}
+is correct, the clustering results are stable to bootstrap resampling, as
+described above. The distribution of similaries between bootstrapped results
+for each \(k\) can thus be compared for different values of \(k\) and guide the
+user in the choice of number of clusters.
+
+The model explorer strategy works as follows. For a single choice of \(k\),
+first perform \(n\) bootstrap experiments to estimate the cluster centroids,
+followed by a step of assigning a label to all data points. Then, choose a
+similarity measure between two partitions or clusters \(S(B_1, B_2)\). Examples
+are the normalized mutual information or Fowlkes-Mallows. Finally, compute the
+pairwise similarity measure between all bootstrapped partitions (i.e.~B choose
+2 pairs). Repeat this procedure for different \(k\) and plot per \(k\) the
+cumulative density of the obtained scores.
+
+We have already run the bootstrap resampling for values \(k=2-20\) and saved the
+results. We will read that data in, and use it to calculate the pairwise
+similarity scores.
+
+The function \texttt{plot\_model\_explorer} takes in input a list of the bootstrapped
+results for all labels.
 
@@
 \begin{Highlighting}[]
-\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"C6"}\NormalTok{, }\StringTok{"C7"}\NormalTok{, }\StringTok{"C8"}\NormalTok{, }\StringTok{"C9"}\NormalTok{, }\StringTok{"C10"}\NormalTok{)}
+\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"B6"}\NormalTok{, }\StringTok{"B7"}\NormalTok{, }\StringTok{"B8"}\NormalTok{, }\StringTok{"B9"}\NormalTok{, }\StringTok{"B10"}\NormalTok{)}
 \NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
@@
 bootstrapped clustering, the authors suggests plotting the cumulative density
-of elements of the consensus matrix. We provide a function \texttt{plot\_cdf\_consensus} in \texttt{moanin} to do this
+of elements of the consensus matrix. We provide a function
+\texttt{plot\_cdf\_consensus} in \texttt{moanin} to do this
 
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{auc_scores =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{get_auc_similarity_scores}\NormalTok{(all_labels)}
+\KeywordTok{plot}\NormalTok{(n_clusters, auc_scores, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{type=}\StringTok{"b"}\NormalTok{, }\DataTypeTok{pch=}\DecValTok{16}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \begin{center}\includegraphics[width=6.67in]{images/clustering_AUC_consensus} \end{center}
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{delta_auc =}\StringTok{ }\KeywordTok{diff}\NormalTok{(auc_scores)}\OperatorTok{/}\NormalTok{auc_scores[}\DecValTok{1}\OperatorTok{:}\NormalTok{(}\KeywordTok{length}\NormalTok{(auc_scores)}\OperatorTok{-}\DecValTok{1}\NormalTok{)]}
+\KeywordTok{plot}\NormalTok{(}\DecValTok{3}\OperatorTok{:}\DecValTok{20}\NormalTok{, delta_auc, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{type=}\StringTok{"b"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \begin{center}\includegraphics[width=6.67in]{images/clustering_delta_AUC_consensus} \end{center}
-
-\textbf{EAP: We should again provide the code that goes with these images}
 

### 2019-10-08 scripts/manuscript.Rmd
FIX create_moanin_model now returns an object
ENH DE_timecourse now provides q-values in addition of p-values

@@
 
-**EAP**: this makes it sound like you create a class, but in fact it's just a list. Wouldn't it be helpful to make it a class?
-
 We start by creating the `moanin_model` object using the
@@
 
-The output is a table of results, where each row corresponds to a gene
-and the columns correspond to the p-value (`pval`), log-fold change (`lfc`)
-and adjusted p-value (`qval`) of the sets of contrasts; the order of the genes in the table is the same
-as the input `data`. Here we show the results for the first timepoint (i.e. first three columns of the output) and
-the first ten genes:
+The output is a table of results, where each row corresponds to a gene and the
+columns correspond to the p-value (`pval`), log-fold change (`lfc`) and
+adjusted p-value (`qval`) of the sets of contrasts; the order of the genes in
+the table is the same as the input `data`. Here we show the results for the
+first timepoint (i.e. first three columns of the output) and the first ten
+genes:
 
@@
 
-We will repeat this, comparing each of the remaining three treatments to the control ("M")
-(code not printed here, as it is a replicate of the above, see accompanying
-Rmarkdown document).
+We will repeat this, comparing each of the remaining three treatments to the
+control ("M") (code not printed here, as it is a replicate of the above, see
+accompanying Rmarkdown document).
 
@@
 # treatment).
-pvalues = DE_timecourse(
+DE_results = DE_timecourse(
     data, moanin_model, timecourse_contrasts,
     use_voom_weights=FALSE)
-qvalues = apply(pvalues, 2, p.adjust)
-
-```
-
-**EAP: is there a reason that the qvalues aren't provided with the output like in `DE_timepoints`? Doesn't that risk the user not taking that step and working with the unadjusted p-values?**
+pval_columns = colnames(DE_results)[
+    grepl("pval", colnames(DE_results))]
+qval_columns = colnames(DE_results)[
+    grepl("qval", colnames(DE_results))]
+pvalues = DE_results[, pval_columns]
+qvalues = DE_results[, qval_columns]
+```
 
@@
 ```{r fig.height=4, fig.width=4}
-pvalue = qvalues[, "M-K"]
+pvalue = DE_results[, "M-K_pval"]
+names(pvalue) = row.names(DE_results)
 lfc_timecourse = log_fold_change_timecourse[, "M-K"]
@@
 As another sanity check, `moanin` provides a simple utility function to
-visualize gene time-course data in different conditions. Here, we plot the 10 genes with the smallest
-p-values.
+visualize gene time-course data in different conditions. Here, we plot the 10
+genes with the smallest p-values.
 
@@
 fishers_pval = pvalues_fisher_method(pvalues)
+qvalues = apply(pvalues, 2, p.adjust)
 fishers_qval = p.adjust(fishers_pval)
@@
 
-**EAP: `subsample_proportion` was set to 1 in the code, but shouldn't it be
-0.8? I changed it to 0.8. But it seems like you don't actually run this code,
-but read existing runs, so maybe was right there?**
-
-**Nelle: that was a mistake in the code. I don't actually run subsampling
-experiments: only the bootstrap, and it's done in a separate file as it is too
-time consuming to run in an Rmd file**
-
 ```{r}

### 2019-10-08 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\textbf{EAP}: this makes it sound like you create a class, but in fact it's just a list. Wouldn't it be helpful to make it a class?
-
 We start by creating the \texttt{moanin\_model} object using the
@@
 
-The output is a table of results, where each row corresponds to a gene
-and the columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc})
-and adjusted p-value (\texttt{qval}) of the sets of contrasts; the order of the genes in the table is the same
-as the input \texttt{data}. Here we show the results for the first timepoint (i.e.~first three columns of the output) and
-the first ten genes:
+The output is a table of results, where each row corresponds to a gene and the
+columns correspond to the p-value (\texttt{pval}), log-fold change (\texttt{lfc}) and
+adjusted p-value (\texttt{qval}) of the sets of contrasts; the order of the genes in
+the table is the same as the input \texttt{data}. Here we show the results for the
+first timepoint (i.e.~first three columns of the output) and the first ten
+genes:
 
@@
 
-We will repeat this, comparing each of the remaining three treatments to the control (``M'')
-(code not printed here, as it is a replicate of the above, see accompanying
-Rmarkdown document).
+We will repeat this, comparing each of the remaining three treatments to the
+control (``M'') (code not printed here, as it is a replicate of the above, see
+accompanying Rmarkdown document).
 
@@
 \CommentTok{# treatment).}
-\NormalTok{pvalues =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
+\NormalTok{DE_results =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
 \NormalTok{    data, moanin_model, timecourse_contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
-\NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, p.adjust)}
-\end{Highlighting}
-\end{Shaded}
-
-\textbf{EAP: is there a reason that the qvalues aren't provided with the output like in \texttt{DE\_timepoints}?
-Doesn't that risk the user not taking that step and working with the unadjusted p-values?}
+\NormalTok{pval_columns =}\StringTok{ }\KeywordTok{colnames}\NormalTok{(DE_results)[}
+    \KeywordTok{grepl}\NormalTok{(}\StringTok{"pval"}\NormalTok{, }\KeywordTok{colnames}\NormalTok{(DE_results))]}
+\NormalTok{qval_columns =}\StringTok{ }\KeywordTok{colnames}\NormalTok{(DE_results)[}
+    \KeywordTok{grepl}\NormalTok{(}\StringTok{"qval"}\NormalTok{, }\KeywordTok{colnames}\NormalTok{(DE_results))]}
+\NormalTok{pvalues =}\StringTok{ }\NormalTok{DE_results[, pval_columns]}
+\NormalTok{qvalues =}\StringTok{ }\NormalTok{DE_results[, qval_columns]}
+\end{Highlighting}
+\end{Shaded}
 
@@
 \begin{Highlighting}[]
-\NormalTok{pvalue =}\StringTok{ }\NormalTok{qvalues[, }\StringTok{"M-K"}\NormalTok{]}
+\NormalTok{pvalue =}\StringTok{ }\NormalTok{DE_results[, }\StringTok{"M-K_pval"}\NormalTok{]}
+\KeywordTok{names}\NormalTok{(pvalue) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(DE_results)}
 \NormalTok{lfc_timecourse =}\StringTok{ }\NormalTok{log_fold_change_timecourse[, }\StringTok{"M-K"}\NormalTok{]}
@@
 As another sanity check, \texttt{moanin} provides a simple utility function to
-visualize gene time-course data in different conditions. Here, we plot the 10 genes with the smallest
-p-values.
+visualize gene time-course data in different conditions. Here, we plot the 10
+genes with the smallest p-values.
 
@@
 \NormalTok{fishers_pval =}\StringTok{ }\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
+\NormalTok{qvalues =}\StringTok{ }\KeywordTok{apply}\NormalTok{(pvalues, }\DecValTok{2}\NormalTok{, p.adjust)}
 \NormalTok{fishers_qval =}\StringTok{ }\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}
@@
 Using the subsampling strategy, we take a unique subset of the genes, keeping 80\% of the genes:
-
-\textbf{EAP: \texttt{subsample\_proportion} was set to 1 in the code, but shouldn't it be
-0.8? I changed it to 0.8. But it seems like you don't actually run this code,
-but read existing runs, so maybe was right there?}
-
-\textbf{Nelle: that was a mistake in the code. I don't actually run subsampling
-experiments: only the bootstrap, and it's done in a separate file as it is too
-time consuming to run in an Rmd file}
 

### 2019-10-15 scripts/manuscript.Rmd
Small modification on the manuscript

@@
     labels=NULL, qval_threshold=0.05,
-    xlab="Timepoint", ylab="Number of DE genes", main="",ylim=c(0, 20000), col="black", ...){
+    xlab="Timepoint", ylab="Number of DE genes", main="", ylim=c(0, 20000), col="black", ...){
+
     qval_colnames = colnames(de_analysis)[
@@
 ```{r fig.height=6, fig.width=6}
-cluster2Genes = names(kmeans_clusters$clusters[kmeans_clusters$clusters==2])
-dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[2,])
+cluster_of_interest = 2
+cluster2Genes = names(
+    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
+dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
 plot_splines_data(dataToPlot, moanin_model,
@@
 The scoring and labeling is done via the `splines_kmeans_score_and_label`
-function.
-This function calculates the goodness of fit of the gene to the cluster centroid.
-By default, the function labels only the 50% of genes with the best scores, with the remaining genes getting `NA` as their assignment.
+function. This function calculates the goodness of fit of the gene to the
+cluster centroid. By default, the function labels only the 50% of genes with
+the best scores, with the remaining genes getting `NA` as their assignment.
 
@@
 labels = unlist(labels[!is.na(labels)])
-cluster2Score = names(labels[labels==2])
-```
-
-After running `scores_and_labels` we now have `r length(labels)` genes that are assigned to a cluster.
+cluster2Score = names(labels[labels==cluster_of_interest])
+```
+
+After running `scores_and_labels` we now have `r length(labels)` genes that
+are assigned to a cluster.
 
@@
 
-Of the original `r length(kmeans_clusters$clusters)` genes used in the clustering, `r length(intersect(names(kmeans_clusters$clusters), names(labels)))` are still assigned to a cluster based on their goodness of fit score.
-We can compare whether they are still assigned to the same clusters based on a confusion matrix shown below.
-The kmeans cluster assignments are shown on the rows, and the goodness-of-fit assignments in the columns, with the number in each cell indicating the number of genes in the intersection of the two clusters.
+Of the original `r length(kmeans_clusters$clusters)` genes used in the
+clustering, `r length(intersect(names(kmeans_clusters$clusters),
+names(labels)))` are still assigned to a cluster based on their goodness of
+fit score. We can compare whether they are still assigned to the same clusters
+based on a confusion matrix shown below. The kmeans cluster assignments are
+shown on the rows, and the goodness-of-fit assignments in the columns, with
+the number in each cell indicating the number of genes in the intersection of
+the two clusters.
 
@@
 
-Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in% cluster2Score)` of them remained assigned to cluster 2.
-We can again look at a few genes in cluster 2, only now pick the best scoring genes
+Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
+cluster2Score)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
+few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes
 
@@
 # order them by their score
-cluster2Score = cluster2Score[order(scores_and_labels$scores[cluster2Score, 8], decreasing = TRUE)]
-dataToPlot = rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[8, ])
+cluster2Score = cluster2Score[order(scores_and_labels$scores[cluster2Score, cluster_of_interest],
+				    decreasing = TRUE)]
+dataToPlot = rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
 plot_splines_data(dataToPlot, moanin_model,
@@
 ```
+
+**EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
+Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
+It's seems like an important plot to make**-
 
@@
 ```
-
-**EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
-Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
-It's seems like an important plot to make**
-
-**Nelle: in my experience, the large the cluster is, the noisier the centroid
-estimation is. I've switched to cluster 8, as this is the cluster we focus our
-interpretation, and the top scoring genes seem to be to be closely related to
-the centroid estimates**
 
@@
 
-Using the bootstrapping strategy we sample with replacement a sample of the
+Using the bootstrapping strategy, we sample with replacement a sample of the
 same size as our original clustering (i.e. `r dim(y)[1]` genes):
@@
 ```
-
-**EAP: So this is very confusing to understand, because you read the results from a file (`stability_5.tsv`), but I don't know what this file contains, nor how they were calculated.
-And I don't know what a stability matrix is -- you've only talked about a consensus matrix.
-There should be the code here for a single k, but not evaluated, and say that for convenience we provide the results of running this code as a file for users to play with.
-I've tried to fix up the text to be what I think it should be**
 

### 2019-10-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \begin{Highlighting}[]
-\NormalTok{cluster2Genes =}\StringTok{ }\KeywordTok{names}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\DecValTok{2}\NormalTok{])}
-\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(y[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{2}\NormalTok{,])}
+\NormalTok{cluster_of_interest =}\StringTok{ }\DecValTok{2}
+\NormalTok{cluster2Genes =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
+\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
+\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(y[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,])}
 \KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
@@
 The scoring and labeling is done via the \texttt{splines\_kmeans\_score\_and\_label}
-function.
-This function calculates the goodness of fit of the gene to the cluster centroid.
-By default, the function labels only the 50\% of genes with the best scores, with the remaining genes getting \texttt{NA} as their assignment.
+function. This function calculates the goodness of fit of the gene to the
+cluster centroid. By default, the function labels only the 50\% of genes with
+the best scores, with the remaining genes getting \texttt{NA} as their assignment.
 
@@
 \NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
-\NormalTok{cluster2Score =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\DecValTok{2}\NormalTok{])}
-\end{Highlighting}
-\end{Shaded}
-
-After running \texttt{scores\_and\_labels} we now have 19772 genes that are assigned to a cluster.
+\NormalTok{cluster2Score =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
+\end{Highlighting}
+\end{Shaded}
+
+After running \texttt{scores\_and\_labels} we now have 19772 genes that
+are assigned to a cluster.
 
@@
 
-Of the original 3950 genes used in the clustering, 3467 are still assigned to a cluster based on their goodness of fit score.
-We can compare whether they are still assigned to the same clusters based on a confusion matrix shown below.
-The kmeans cluster assignments are shown on the rows, and the goodness-of-fit assignments in the columns, with the number in each cell indicating the number of genes in the intersection of the two clusters.
+Of the original 3950 genes used in the
+clustering, 3467 are still assigned to a cluster based on their goodness of
+fit score. We can compare whether they are still assigned to the same clusters
+based on a confusion matrix shown below. The kmeans cluster assignments are
+shown on the rows, and the goodness-of-fit assignments in the columns, with
+the number in each cell indicating the number of genes in the intersection of
+the two clusters.
 
@@
 
-Of the four genes we plotted before, 3 of them remained assigned to cluster 2. We can again look at a few genes in cluster 2, only now pick the best scoring genes
+Of the four genes we plotted before, 3 of them remained assigned to cluster 2. We can again look at a
+few genes in cluster 2, only now pick the best scoring genes
 
@@
 \CommentTok{# order them by their score}
-\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score, }\DecValTok{8}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{8}\NormalTok{, ])}
+\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score, cluster_of_interest],}
+                    \DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
 \KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
@@
 Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
-It's seems like an important plot to make}
-
-\textbf{Nelle: in my experience, the large the cluster is, the noisier the centroid
-estimation is. I've switched to cluster 8, as this is the cluster we focus our
-interpretation, and the top scoring genes seem to be to be closely related to
-the centroid estimates}
+It's seems like an important plot to make}-
 
@@
 
-Using the bootstrapping strategy we sample with replacement a sample of the
+Using the bootstrapping strategy, we sample with replacement a sample of the
 same size as our original clustering (i.e.~3950 genes):
@@
 \end{Shaded}
-
-\textbf{EAP: So this is very confusing to understand, because you read the results from a file (\texttt{stability\_5.tsv}), but I don't know what this file contains, nor how they were calculated.
-And I don't know what a stability matrix is -- you've only talked about a consensus matrix.
-There should be the code here for a single k, but not evaluated, and say that for convenience we provide the results of running this code as a file for users to play with.
-I've tried to fix up the text to be what I think it should be}
 

### 2019-10-25 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.0             pkgconfig_2.0.2        
+## [65] rlang_0.4.1             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          

### 2019-10-29 scripts/manuscript.Rmd
Small fixes here and there

@@
     time-course transcriptomics data from mice exposed to different strains of
-    influenza, this workflows provides a step-by-step tutorial of the
+    influenza, this workflow provides a step-by-step tutorial of the
     methodology used to analyse time-course data: (1) normalization of the
-    micro-array dataset; (2) differential expression analysis using functional
-    data analysis; (3) clustering fo time-course data; (4) interpreting
+    micro-array data set; (2) differential expression analysis using functional
+    data analysis; (3) clustering of time-course data; (4) interpreting
     clusters with GO term and KEGG pathway enrichment analysis.
@@
 [@shalek:single-cell; @habib:div-seq; @trapnell:dynamics]
-which can sequencing single cells which are at different stages of development; in this case,
+which can sequence single cells at different stages of development; in this case,
 the time point is the stage of the cell in the process of development -- a
@@
 strategies to combine single time point differential expression analysis
-[@sun:multiple]. However, with long time course datasets, particularly in
+[@sun:multiple]. However, with long time course data sets, particularly in
 developmental systems, a massive number of genes will show *some* change. For
@@
 define color schemes for our data that we will use across the whole analysis.
-We define coloschemes for groups, timepoints as named vectors.
+We define color schemes for groups and time points as named vectors. We also
+define a series of markers (or plotting symbols) to distinguish replicate
+samples in scatter plots.
 
@@
 variance_cutoff = 0.5
+# Filter genes by median absolute deviation (mad)
 variance_per_genes = apply(data, 1, mad)
@@
 higher than the cross correlation amongst the rest of the treatments. Second,
-the influenza-infected mice midly react until time point 36. Third, the less
+the influenza-infected mice mildly react until time point 36. Third, the less
 pathogenic the strain is, the closer the samples are to the control condition.
@@
 approaches, and is often the approach advocated when dealing with small
-time-course datasets, where there are only a few time points 
-[@ritchie:limma; @robinson:edgeR; @love:moderated] . For long time-course datasets,
-however, a separate test for each time-point results in creating many
-different tests, for example one for every time point, the results of which
-are difficult to integrate. We find in practice that the global analysis
-simplifies analysis and of longer time courses data, with per-time point
-analysis reserved for particularly interesting comparisons of individual
-time-points.
+time-course datasets, where there are only a few time points [@ritchie:limma;
+@robinson:edgeR; @love:moderated] . For long time-course datasets, however, a
+separate test for each time-point results in creating many different tests,
+for example one for every time point, the results of which are difficult to
+integrate. We find in practice that the global analysis simplifies analysis
+and interpretation of longer time courses data, with per-time point analysis
+reserved for particularly interesting comparisons of individual time-points.
 
 Time course data can either be on a single condition (to identify genes
-changing over time) or or on multiple conditions (such as the influenza
-dataset we are considering), which will alter slightly the types of questions
+changing over time) or on multiple conditions (such as the influenza
+data set we are considering), which will alter slightly the types of questions
 we are interested in.
@@
 
-We start by creating the `moanin_model` object using the
-`create_moanin_model` function. We need to provide two things to the function: a data.frame with the meta data, and the number of degrees of freedom. The
-metadata data.frame object should contain at least two columns: one named
-`Group`, containing the  treatement effect, and a second one named `Timepoint`
-containing the timepoint information.
+We start by creating the `moanin_model` object using the `create_moanin_model`
+function. We need to provide two things to the function: a `data.frame` with
+the metadata, and the number of degrees of freedom of the splines used in the
+functional modeling. The metadata `data.frame` object should contain at least
+two columns: one named `Group`, containing the  treatement effect, and a
+second one named `Timepoint` containing the timepoint information.
 
@@
 `moanin` provides a simple interface to perform a timepoint by timepoint
-differential expression analysis. Comparison between groups is traditionally done by 
-defining the group comparisons (called `contrasts` in linear models) as a linear combination of the coefficients of the model.
-Comparing groups within each timepoint can create many contrasts, and thus `moanin` provides functionality to create these contrasts in an automatic way, and then calls `limma` [@ritchie:limma] on the set of contrasts provided.
-By default, `moanin` expects RNA-Seq contact counts, and will estimate voom weights.
+differential expression analysis. Comparison between groups is traditionally
+done by defining the group comparisons (called `contrasts` in linear models)
+as a linear combination of the coefficients of the model.  Comparing groups
+within each timepoint can create many contrasts, and thus `moanin` provides
+functionality to create these contrasts in an automatic way, and then calls
+`limma` [@ritchie:limma] on the set of contrasts provided. By default,
+`moanin` expects RNA-Seq contact counts, and will estimate voom weights.
 
@@
 data between different treatment conditions, using a similar mechanism of
-contrasts -- only now the contrasts are differences between the estimated
-mean functions.
-This is done with the function `DE_timecourse`, which takes as similar input that of `DE_timepoints`, only now will test the entire mean function (and therefore does not require a step of expanding the contrasts into contrasts for individual timepoints).
+contrasts -- only now the contrasts are differences between the estimated mean
+functions. This is done with the function `DE_timecourse`, which takes as
+similar input that of `DE_timepoints`, only now will test the entire mean
+function (and therefore does not require a step of expanding the contrasts
+into contrasts for individual timepoints).
 
@@
   cluster centroids.
-- **Assign scores and labels to all genes**: Our `splines_kmeans_score_and_label` function assigns a score to all gene-cluster pairs based on a goodness-of-fit measure of the gene to the cluster centroid.
-We then "assign" each gene to a cluster based on its score.
-
-The first three steps are performed internally by the `splines_kmeans` function.
-For now, we will set the number of clusters to be $8$, though we will return to the question of picking the best number of clusters below.
+- **Assign scores and labels to all genes**: Our
+  `splines_kmeans_score_and_label` function assigns a score to all
+  gene-cluster pairs based on a goodness-of-fit measure of the gene to the
+  cluster centroid. We then "assign" each gene to a cluster based on its
+  score.
+
+The first three steps are performed internally by the `splines_kmeans`
+function. For now, we will set the number of clusters to be $8$, though we
+will return to the question of picking the best number of clusters below.
 
@@
 
-We then use the `plot_splines_data` function to visualize the centroids of each cluster obtained
-with the splines k-means model.
+We then use the `plot_splines_data` function to visualize the centroids of
+each cluster obtained with the splines k-means model.
 
@@
 
-These centroids are on a 0-1 scale, because of our rescaling of the spline fits, and do not represent the actual gene expression level.
-We can plot a few of the actual genes in a cluster to see the difference:
+These centroids are on a 0-1 scale, because of our rescaling of the spline
+fits, and do not represent the actual gene expression level. We can plot a few
+of the actual genes in a cluster to see the difference:
 
@@
 
-The spline plots are fitted per gene in the above plot, and illustrate that the genes assigned to a cluster have differing degrees to which their data fits that of the cluster centroid, and indeed fits that of their individual spline fits.
+The spline plots are fitted per gene in the above plot, and illustrate that
+the genes assigned to a cluster have differing degrees to which their data
+fits that of the cluster centroid, and indeed fits that of their individual
+spline fits.
 
@@
 
-Since we arbitrarily chose a subset of genes based on our filter, we would like all genes to have a chance of assignment to a cluster, as well as a score for whether it is a good assignment.
-The scoring and label step allows us to assign genes to clusters that were removed during the
-filtering step, yet are good matches to the centroids found during the
-clustering.
+Since we arbitrarily chose a subset of genes based on our filter, we would
+like all genes to have a chance of assignment to a cluster, as well as a score
+for whether it is a good assignment.  The scoring and label step allows us to
+assign genes to clusters that were removed during the filtering step, yet are
+good matches to the centroids found during the clustering.
 
@@
 cluster2Score = names(labels[labels==cluster_of_interest])
+```
+
+**Nelle: This is currently for debugging purposes. The distribution of scores
+seems very strange to me, with a lot of genes that don't match at all to any
+clusters.**
+
+```{r visualize_scores_all_figure, fig.height=6, fix.width=6}
+# Get the best score for each gene
+scores = apply(scores_and_labels$scores, 1, min)
+hist(scores, breaks=(1:50/50), col="black", main="Distribution of scores")
+```
+
+**Compared to the EPICON project, a lot of genes seem to have very very poor
+scores. The threshod on the EPICON project on the score was around 0.55, while
+here, it is close to 0.87. The scores range from 0 to 1, with 1 being no
+correlation between the gene and the centroid.**
+
+```{r visualize_scores, fig.height=6, fix.width=6}
+# Get the best score for each gene
+par(mfrow=c(4, 5))
+scores = apply(scores_and_labels$scores, 1, min)
+labels = apply(scores_and_labels$scores, 1, which.min)
+for(cluster_id in 1:n_clusters){
+    hist(scores[labels == cluster_id],
+	 breaks=(1:50/50), xlim=c(0, 1),
+	 col="black", main=paste("C", cluster_id, sep=""))
+}
+
+labels = unlist(labels[!is.na(labels)])
+
 ```

### 2019-10-29 scripts/manuscript.Rmd
Checking out histogram of scores and everything

@@
 ```
+
+The manuscript was last compile on `r Sys.time()`.
 
@@
 scores = apply(scores_and_labels$scores, 1, min)
-hist(scores, breaks=(1:50/50), col="black", main="Distribution of scores")
+hist(scores, breaks=(1:50/50)            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+            188745       low 6-RSEC.s singlece  R    9:50:32      1 scf-sm11
+, col="black", main="Distribution of scores")
 ```

### 2019-10-29 scripts/manuscript.Rmd
Small fix on the manuscript

--- 2019-10-29 scripts/manuscript.Rmd
@@
 labels = apply(scores_and_labels$scores, 1, which.min)
+n_clusters = dim(kmeans_clusters$centroids)[1]
 for(cluster_id in 1:n_clusters){
@@
 
+labels = scores_and_labels$labels
 labels = unlist(labels[!is.na(labels)])

### 2019-10-30 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology.However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software.
-Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflows provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) normalization of the micro-array dataset; (2) differential expression analysis using functional data analysis; (3) clustering fo time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
+Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) normalization of the micro-array data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
 \end{abstract}
@@
 
-\#TODO list
+The manuscript was last compile on 2019-10-29 23:39:12.
+
+\#\textless{}U+00A0\textgreater{}TODO list
 
@@
 \citep{shalek:single-cell, habib:div-seq, trapnell:dynamics}
-which can sequencing single cells which are at different stages of development; in this case,
+which can sequence single cells at different stages of development; in this case,
 the time point is the stage of the cell in the process of development -- a
@@
 strategies to combine single time point differential expression analysis
-\citep{sun:multiple}. However, with long time course datasets, particularly in
+\citep{sun:multiple}. However, with long time course data sets, particularly in
 developmental systems, a massive number of genes will show \emph{some} change. For
@@
 \NormalTok{variance_cutoff =}\StringTok{ }\FloatTok{0.5}
+\CommentTok{# Filter genes by median absolute deviation (mad)}
 \NormalTok{variance_per_genes =}\StringTok{ }\KeywordTok{apply}\NormalTok{(data, }\DecValTok{1}\NormalTok{, mad)}
@@
 approaches, and is often the approach advocated when dealing with small
-time-course datasets, where there are only a few time points
-\citep{ritchie:limma, robinson:edgeR, love:moderated} . For long time-course datasets,
-however, a separate test for each time-point results in creating many
-different tests, for example one for every time point, the results of which
-are difficult to integrate. We find in practice that the global analysis
-simplifies analysis and of longer time courses data, with per-time point
-analysis reserved for particularly interesting comparisons of individual
-time-points.
+time-course datasets, where there are only a few time points \citep{ritchie:limma, robinson:edgeR, love:moderated} . For long time-course datasets, however, a
+separate test for each time-point results in creating many different tests,
+for example one for every time point, the results of which are difficult to
+integrate. We find in practice that the global analysis simplifies analysis
+and interpretation of longer time courses data, with per-time point analysis
+reserved for particularly interesting comparisons of individual time-points.
 
 Time course data can either be on a single condition (to identify genes
-changing over time) or or on multiple conditions (such as the influenza
-dataset we are considering), which will alter slightly the types of questions
+changing over time) or on multiple conditions (such as the influenza
+data set we are considering), which will alter slightly the types of questions
 we are interested in.
@@
 
-We start by creating the \texttt{moanin\_model} object using the
-\texttt{create\_moanin\_model} function. We need to provide two things to the function: a data.frame with the meta data, and the number of degrees of freedom. The
-metadata data.frame object should contain at least two columns: one named
-\texttt{Group}, containing the treatement effect, and a second one named \texttt{Timepoint}
-containing the timepoint information.
+We start by creating the \texttt{moanin\_model} object using the \texttt{create\_moanin\_model}
+function. We need to provide two things to the function: a \texttt{data.frame} with
+the metadata, and the number of degrees of freedom of the splines used in the
+functional modeling. The metadata \texttt{data.frame} object should contain at least
+two columns: one named \texttt{Group}, containing the treatement effect, and a
+second one named \texttt{Timepoint} containing the timepoint information.
 
@@
 \texttt{moanin} provides a simple interface to perform a timepoint by timepoint
-differential expression analysis. Comparison between groups is traditionally done by
-defining the group comparisons (called \texttt{contrasts} in linear models) as a linear combination of the coefficients of the model.
-Comparing groups within each timepoint can create many contrasts, and thus \texttt{moanin} provides functionality to create these contrasts in an automatic way, and then calls \texttt{limma} \citep{ritchie:limma} on the set of contrasts provided.
-By default, \texttt{moanin} expects RNA-Seq contact counts, and will estimate voom weights.
+differential expression analysis. Comparison between groups is traditionally
+done by defining the group comparisons (called \texttt{contrasts} in linear models)
+as a linear combination of the coefficients of the model. Comparing groups
+within each timepoint can create many contrasts, and thus \texttt{moanin} provides
+functionality to create these contrasts in an automatic way, and then calls
+\texttt{limma} \citep{ritchie:limma} on the set of contrasts provided. By default,
+\texttt{moanin} expects RNA-Seq contact counts, and will estimate voom weights.
 
@@
 data between different treatment conditions, using a similar mechanism of
-contrasts -- only now the contrasts are differences between the estimated
-mean functions.
-This is done with the function \texttt{DE\_timecourse}, which takes as similar input that of \texttt{DE\_timepoints}, only now will test the entire mean function (and therefore does not require a step of expanding the contrasts into contrasts for individual timepoints).
+contrasts -- only now the contrasts are differences between the estimated mean
+functions. This is done with the function \texttt{DE\_timecourse}, which takes as
+similar input that of \texttt{DE\_timepoints}, only now will test the entire mean
+function (and therefore does not require a step of expanding the contrasts
+into contrasts for individual timepoints).
 
@@
 \item
-  \textbf{Assign scores and labels to all genes}: Our \texttt{splines\_kmeans\_score\_and\_label} function assigns a score to all gene-cluster pairs based on a goodness-of-fit measure of the gene to the cluster centroid.
-We then ``assign'' each gene to a cluster based on its score.
+  \textbf{Assign scores and labels to all genes}: Our
+  \texttt{splines\_kmeans\_score\_and\_label} function assigns a score to all
+  gene-cluster pairs based on a goodness-of-fit measure of the gene to the
+  cluster centroid. We then ``assign'' each gene to a cluster based on its
+  score.
 \end{itemize}
 
-The first three steps are performed internally by the \texttt{splines\_kmeans} function.
-For now, we will set the number of clusters to be \(8\), though we will return to the question of picking the best number of clusters below.
+The first three steps are performed internally by the \texttt{splines\_kmeans}
+function. For now, we will set the number of clusters to be \(8\), though we
+will return to the question of picking the best number of clusters below.
 
@@
 
-We then use the \texttt{plot\_splines\_data} function to visualize the centroids of each cluster obtained
-with the splines k-means model.
+We then use the \texttt{plot\_splines\_data} function to visualize the centroids of
+each cluster obtained with the splines k-means model.
 
@@
 
-Since we arbitrarily chose a subset of genes based on our filter, we would like all genes to have a chance of assignment to a cluster, as well as a score for whether it is a good assignment.
-The scoring and label step allows us to assign genes to clusters that were removed during the
-filtering step, yet are good matches to the centroids found during the
-clustering.
+Since we arbitrarily chose a subset of genes based on our filter, we would
+like all genes to have a chance of assignment to a cluster, as well as a score
+for whether it is a good assignment. The scoring and label step allows us to
+assign genes to clusters that were removed during the filtering step, yet are
+good matches to the centroids found during the clustering.
 
@@
 \end{Shaded}
+
+\textbf{Nelle: This is currently for debugging purposes. The distribution of scores
+seems very strange to me, with a lot of genes that don't match at all to any
+clusters.}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# Get the best score for each gene}
+\NormalTok{scores =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, min)}
+\KeywordTok{hist}\NormalTok{(scores, }\DataTypeTok{breaks=}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\OperatorTok{/}\DecValTok{50}\NormalTok{), }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Distribution of scores"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/visualize_scores_all_figure-1} \end{center}
+
+\textbf{Compared to the EPICON project, a lot of genes seem to have very very poor
+scores. The threshod on the EPICON project on the score was around 0.55, while
+here, it is close to 0.87. The scores range from 0 to 1, with 1 being no
+correlation between the gene and the centroid.}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# Get the best score for each gene}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{))}
+\NormalTok{scores =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, min)}
+\NormalTok{labels =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, which.min)}
+\NormalTok{n_clusters =}\StringTok{ }\KeywordTok{dim}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids)[}\DecValTok{1}\NormalTok{]}
+\ControlFlowTok{for}\NormalTok{(cluster_id }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\NormalTok{n_clusters)\{}
+    \KeywordTok{hist}\NormalTok{(scores[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_id],}
+     \DataTypeTok{breaks=}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\OperatorTok{/}\DecValTok{50}\NormalTok{), }\DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{),}
+     \DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"C"}\NormalTok{, cluster_id, }\DataTypeTok{sep=}\StringTok{""}\NormalTok{))}
+\NormalTok{\}}
+
+\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
+\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/visualize_scores-1} \end{center}
 

### 2019-11-13 scripts/manuscript.Rmd
Small edits + fixes onto the manuscript

--- 2019-10-29 scripts/manuscript.Rmd
@@
 We then plot the two first PC components, and color each sample by (1) its
-condition; (2) its sampling time. We use different plotting symbols for each replicate. We also plot the second and third components in the second column.
+condition; (2) its sampling time. We use different plotting symbols for each
+replicate. We also plot the second and third components in the second column.
 
@@
 approach first step is to fit a spline model to each gene
-[@storey:significance],
-and then use that spline model to test for different kinds of differential
-expression across time.
+[@storey:significance], and then use that spline model to test for different
+kinds of differential expression across time.
 
@@
 throughout this analysis: the condition and timepoints of each samples, the
-formula object used, the basis, and the degrees of freedom of the model.
+formula object used, the basis matrix, and the degrees of freedom of the model.
 
@@
 `formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.
+
+The basis matrix contains the value of all of the spline basis function for
+each sample of the moment. As such, replicate samples will have the same
+values.
+
+```{r basis_matrix}
+kable(head(moanin_model$basis))
+```
 
@@
 assign genes to clusters that were removed during the filtering step, yet are
-good matches to the centroids found during the clustering.
+good matches to the centroids found during the clustering. The score
+corresponds to a goodness-of-fit between each gene and each cluster, computed
+as follows:
+
+$$
+S(y_i, \mu_k) =  \frac{1}{S_0(k)} \times \underset{a_{iG_j}, b_{iG_j}}{\text{min}}\sum_{j}
+  {\left(b_{iG_j}y_{ij} + a_{iG_j}
+   - \mu_k(t_j)\right)}^2 \,,
+$$
+
+where $\mu_k$ is the centroid of cluster $k$ and $y_i$ the gene of interest.
+The scoring function thus returns a value between 0 and 1, 0 being the best
+score possible and 1 the worst score possible (no correlation between the gene
+and the cluster centroid).
 
@@
 
-```{r}
+```{r assigning_scores_and_labels}
 # Then assign scores and labels to all the data, using a goodness-of-fit
@@
 # Let's keep only the list of genes that have a label.
+scores = rowMin(scores_and_labels$scores[!is.na(labels), ])
 labels = unlist(labels[!is.na(labels)])
-cluster2Score = names(labels[labels==cluster_of_interest])
-```
-
-**Nelle: This is currently for debugging purposes. The distribution of scores
-seems very strange to me, with a lot of genes that don't match at all to any
-clusters.**
-
-```{r visualize_scores_all_figure, fig.height=6, fix.width=6}
-# Get the best score for each gene
-scores = apply(scores_and_labels$scores, 1, min)
-hist(scores, breaks=(1:50/50), col="black", main="Distribution of scores")
-```
-
-**Compared to the EPICON project, a lot of genes seem to have very very poor
-scores. The threshod on the EPICON project on the score was around 0.55, while
-here, it is close to 0.87. The scores range from 0 to 1, with 1 being no
-correlation between the gene and the centroid.**
+genes_in_cluster2 = names(labels[labels==cluster_of_interest])
+```
+
+Let us now, for each cluster, visualize the distribution of scores of the
+genes assigned to that cluster.
 
@@
 # Get the best score for each gene
-par(mfrow=c(4, 5))
-scores = apply(scores_and_labels$scores, 1, min)
-labels = apply(scores_and_labels$scores, 1, which.min)
+par(mfrow=c(4, 3))
+n_clusters = dim(kmeans_clusters$centroids)[1]
 for(cluster_id in 1:n_clusters){
@@
 	 breaks=(1:50/50), xlim=c(0, 1),
-	 col="black", main=paste("C", cluster_id, sep=""))
+	 col="black", main=paste("C", cluster_id, sep=""),
+	 xlab="score", ylab="Num. genes")
 }
-
-labels = unlist(labels[!is.na(labels)])
 
@@
 Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
-cluster2Score)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
+genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
 few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes
@@
 # order them by their score
-cluster2Score = cluster2Score[order(scores_and_labels$scores[cluster2Score, cluster_of_interest],
-				    decreasing = TRUE)]
-dataToPlot = rbind(data[cluster2Score[1:4], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
+cluster2Score = cluster2Score[order(scores_and_labels$scores[genes_in_cluster2, cluster_of_interest])]
+dataToPlot = rbind(
+    data[cluster2Score[1:4], ],
+    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
 plot_splines_data(dataToPlot, moanin_model,
@@
 
-**EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
-Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
-It's seems like an important plot to make**-
+**EAP: I don't know what to think about these plots -- these are best scoring,
+but pretty noisy. Also the per-gene spline is sometimes not as good of a
+explanation as the cluster centroid. It's seems like an important plot to
+make**-
+
+**NV: this were the worst scoring genes. I fixed this and the plots now look
+much better**
 

### 2019-11-13 scripts/manuscript.Rmd
Small fix related to previous commit

@@
 # order them by their score
-cluster2Score = cluster2Score[order(scores_and_labels$scores[genes_in_cluster2, cluster_of_interest])]
+cluster2Score = genes_in_cluster2[order(scores_and_labels$scores[genes_in_cluster2, cluster_of_interest])]
 dataToPlot = rbind(

### 2019-11-13 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2019-10-29 23:39:12.
+The manuscript was last compile on 2019-11-13 08:34:55.
 
@@
   approach first step is to fit a spline model to each gene
-  \citep{storey:significance},
-  and then use that spline model to test for different kinds of differential
-  expression across time.
+  \citep{storey:significance}, and then use that spline model to test for different
+  kinds of differential expression across time.
 \end{enumerate}
@@
 \texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.
+
+The basis matrix contains the value of all of the spline basis function for
+each sample of the moment. As such, replicate samples will have the same
+values.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{kable}\NormalTok{(}\KeywordTok{head}\NormalTok{(moanin_model}\OperatorTok{$}\NormalTok{basis))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
+\hline
+  & GroupM & GroupK & GroupC & GroupVL & GroupVH & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)6\\
+\hline
+GSM1557140 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
+\hline
+GSM1557141 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
+\hline
+GSM1557142 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
+\hline
+GSM1557143 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
+\hline
+GSM1557144 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
+\hline
+GSM1557145 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
+\hline
+\end{tabular}
 
@@
 assign genes to clusters that were removed during the filtering step, yet are
-good matches to the centroids found during the clustering.
+good matches to the centroids found during the clustering.The score
+corresponds to a goodness-of-fit between each gene and each cluster, computed
+as follows:
+
+\[
+S(y_i, \mu_k) =  \frac{1}{S_0(k)} \times \underset{a_{iG_j}, b_{iG_j}}{\text{min}}\sum_{j}
+  {\left(b_{iG_j}y_{ij} + a_{iG_j}
+   - \mu_k(t_j)\right)}^2 \,,
+\]
+
+where \(\mu_k\) is the centroid of cluster \(k\) and \(y_i\) the gene of interest.
+The scoring function thus returns a value between 0 and 1, 0 being the best
+score possible and 1 the worst score possible (no correlation between the gene
+and the cluster centroid).
 
@@
 \CommentTok{# Let's keep only the list of genes that have a label.}
+\NormalTok{scores =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels), ])}
 \NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
-\NormalTok{cluster2Score =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\end{Highlighting}
-\end{Shaded}
-
-\textbf{Nelle: This is currently for debugging purposes. The distribution of scores
-seems very strange to me, with a lot of genes that don't match at all to any
-clusters.}
+\NormalTok{genes_in_cluster2 =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
+\end{Highlighting}
+\end{Shaded}
+
+Let us now, for each cluster, visualize the distribution of scores of the
+genes assigned to that cluster.
 
@@
 \CommentTok{# Get the best score for each gene}
-\NormalTok{scores =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, min)}
-\KeywordTok{hist}\NormalTok{(scores, }\DataTypeTok{breaks=}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\OperatorTok{/}\DecValTok{50}\NormalTok{), }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Distribution of scores"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/visualize_scores_all_figure-1} \end{center}
-
-\textbf{Compared to the EPICON project, a lot of genes seem to have very very poor
-scores. The threshod on the EPICON project on the score was around 0.55, while
-here, it is close to 0.87. The scores range from 0 to 1, with 1 being no
-correlation between the gene and the centroid.}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Get the best score for each gene}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{5}\NormalTok{))}
-\NormalTok{scores =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, min)}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, which.min)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{))}
 \NormalTok{n_clusters =}\StringTok{ }\KeywordTok{dim}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids)[}\DecValTok{1}\NormalTok{]}
@@
      \DataTypeTok{breaks=}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\OperatorTok{/}\DecValTok{50}\NormalTok{), }\DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{),}
-     \DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"C"}\NormalTok{, cluster_id, }\DataTypeTok{sep=}\StringTok{""}\NormalTok{))}
+     \DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"C"}\NormalTok{, cluster_id, }\DataTypeTok{sep=}\StringTok{""}\NormalTok{),}
+     \DataTypeTok{xlab=}\StringTok{"score"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"Num. genes"}\NormalTok{)}
 \NormalTok{\}}
-
-\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
 \end{Highlighting}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
 
@@
 \CommentTok{# order them by their score}
-\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score, cluster_of_interest],}
-                    \DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
+\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[genes_in_cluster2, cluster_of_interest])]}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
+\NormalTok{    data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}
+    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
 \KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
-
-\textbf{EAP: I don't know what to think about these plots -- these are best scoring, but pretty noisy.
-Also the per-gene spline is sometimes not as good of a explanation as the cluster centroid.
-It's seems like an important plot to make}-
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
+
+\textbf{EAP: I don't know what to think about these plots -- these are best scoring,
+but pretty noisy. Also the per-gene spline is sometimes not as good of a
+explanation as the cluster centroid. It's seems like an important plot to
+make}-
+
+\textbf{NV: this were the worst scoring genes. I fixed this and the plots now look
+much better}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-32-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-38-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-40-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
 

### 2019-11-19 scripts/manuscript.Rmd
Small fixes here and there

@@
 *EAP*: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
-
-```{r eval=FALSE}
-library(devtools)
-install_github("NelleV/moanin")
-
-library(moanin)
-```
-
-```{r echo=FALSE}
-library(moanin)
-```
 
@@
 transcriptomic response of the crop *S. bicolor* to drought.
+
+## Package versions
+
+The following packages are needed for this workflow.
+
+```{r all_libraries}
+# From CRAN
+library(NMF)
+library(ggfortify)
+
+# From Bioconductor
+library(topGO)
+library(biomaRt)
+
+# From GitHub
+library(moanin)
+```
+
+To install `moanin` directly from GitHub, use the devtools' `install_github`
+function:
+
+```{r eval=FALSE}
+library(devtools)
+install_github("NelleV/moanin")
+
+library(moanin)
+```
 
@@
 ```{r basis_matrix}
-kable(head(moanin_model$basis))
+kable(head(moanin_model$basis), digits=2) %>%
+  kable_styling(font_size = 7)
 ```
@@
 different combinations of timepoints DE. For the comparison of the Kawasaki
-strain to the control, for example, there are `length(signifCombos)` genes
-found DE in some timepoints, and there are `length(tabCombos)` different
+strain to the control, for example, there are `r length(signifCombos)` genes
+found DE in some timepoints, and there are `r length(tabCombos)` different
 combinations of timepoints for which they are DE. Some of these make sense,
-such as DE in timepoints 48H-168H (`tabCombos["48,60,72,120,168"]` genes), but
-many are very fragmentary. For example there are `tabCombos["48,60,120,168"]`
+such as DE in timepoints 48H-168H (`r tabCombos["48,60,72,120,168"]` genes), but
+many are very fragmentary. For example there are `r tabCombos["48,60,120,168"]`
 genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
@@
 plot_splines_data(data, moanin_model, subset_data=exampleGenes,
-		  colors=ann_colors$Group)
+		  colors=ann_colors$Group, smooth=TRUE)
 ```
@@
 | timely | $\text{lfc}(t)$  | Function of time |
-| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
-| abs_sum  | $\sum_t \lvert lfc(t)\lvert$  | Always positive |
-| max  | $\max_t \lvert lfc(t) \lvert$  | Always positive  |
-| min  | $\min_t \lvert lfc(t) \lvert$  | Always positive |
+| sum  | $\sum_t \text{lfc}(t)$  | Sum of log fold change. |
+| abs_sum  | $\sum_t \lvert \text{lfc}(t)\lvert$  | Always positive |
+| max  | $\max_t \lvert \text{lfc}(t) \lvert$  | Always positive  |
+| min  | $\min_t \lvert \text{lfc}(t) \lvert$  | Always positive |
 | timecourse | See details below | Captures overall strength and direction  |
@@
 
-$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)$$
+$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T \text{lfc}(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {\text{lfc}(t)} \lvert \right)$$
 
@@
 ```
+
+The returning object is a matrix, where each row corresponds to a gene,
+each column to a contrasts, and each entry to the log-fold change for this
+pair of contrast and gene.
+
+```{r display_lfc}
+kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE, escape=F)
+```
+
+Each methods of computing the log-fold change captures different elements of
+the time-course data: overall change, largest change, ...
 
@@
 plot(lfc_timecourse, -log10(pvalue), pch=20, main="Volcano plot",
-     xlim=c(-2.5, 2))
+     xlim=c(-2.5, 2), xlab="Timecourse lfc")
 ```
@@
 
+The `splines_kmeans` function returns a named list with:
+
+- `centroids`: a matrix containing the cluster centroids. The matrix is of
+  shape (`n_centroid`, `n_samples`).
+- `clusters`: a vector of size `n_genes`, containing the
+
 We then use the `plot_splines_data` function to visualize the centroids of
@@
     data, kmeans_clusters)
-labels = scores_and_labels$labels
-
-# Let's keep only the list of genes that have a label.
-scores = rowMin(scores_and_labels$scores[!is.na(labels), ])
-labels = unlist(labels[!is.na(labels)])
-genes_in_cluster2 = names(labels[labels==cluster_of_interest])
-```
-
-Let us now, for each cluster, visualize the distribution of scores of the
-genes assigned to that cluster.
-
-```{r visualize_scores, fig.height=6, fix.width=6}
-# Get the best score for each gene
-par(mfrow=c(4, 3))
+```
+
+The `scores_and_labels` list contains two elements:
+
+    - `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
+      containing for each gene and each cluster, the goodness of fit as
+      described above.
+    - `scores_and_labels$labels`: the labels for all of the genes with a
+      good-enough goodness-of-fit score.
+
+First, let us visualize the distribution of goodness-of-fit scores for each
+cluster. Here, we display the whole set of scores assigned to each cluster,
+prior to filtering poor-quality fits.
+
+```{r visualize_scores, fig.height=6, fix.width=6, echo=TRUE, caption="Distribution of goodness-of-fit scores for each cluster.
+The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
+# Get the best score and best label for all of the genes
+scores = rowMin(scores_and_labels$scores)
+labels = apply(scores_and_labels$scores, 1, which.min)
+max_score = quantile(scores, 0.5)
+par(mfrow=c(3, 3))
 n_clusters = dim(kmeans_clusters$centroids)[1]
@@
 	 xlab="score", ylab="Num. genes")
+    abline(v=max_score, col="red", lwd=3, lty=2)
 }
 
+```
+
+Note that for some of the genes, the best scores is 1. A score of 1 indicates
+that there is no correlation between the gene and the cluster: those genes fit
+poorly to all clusters and the assignment to one cluster is done randomly..
+This underlines the importance of filtering genes that fit poorly to clusters.
+
+```{r scores_and_labels}
+labels = scores_and_labels$labels
+
+# Let's keep only the list of genes that have a label.
+labels = unlist(labels[!is.na(labels)])
+
+# And also keep track of all the genes in cluster 2.
+genes_in_cluster2 = names(labels[labels==cluster_of_interest])
 ```
@@
 ```
-
-**EAP: I don't know what to think about these plots -- these are best scoring,
-but pretty noisy. Also the per-gene spline is sometimes not as good of a
-explanation as the cluster centroid. It's seems like an important plot to
-make**-
-
-**NV: this were the worst scoring genes. I fixed this and the plots now look
-much better**
 
@@
 
-**EAP: `plot_model_explorer` is relatively slow, because it does all those pairwise calculations.
-And then to do it for smaller subset, redo all the calculations.
-Can't the output be saved, or the calculations be a different step from the plotting?**
-
 #### Consensus clustering as a way to find $k$

### 2019-11-19 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2019-11-13 08:34:55.
+The manuscript was last compile on 2019-11-19 02:22:00.
 
@@
 \emph{EAP}: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(devtools)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
-
-\KeywordTok{library}\NormalTok{(moanin)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
+\hypertarget{package-versions}{%
+\subsection{Package versions}\label{package-versions}}
+
+The following packages are needed for this workflow.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# From CRAN}
+\KeywordTok{library}\NormalTok{(NMF)}
+\KeywordTok{library}\NormalTok{(ggfortify)}
+
+\CommentTok{# From Bioconductor}
+\KeywordTok{library}\NormalTok{(topGO)}
+\KeywordTok{library}\NormalTok{(biomaRt)}
+
+\CommentTok{# From GitHub}
+\KeywordTok{library}\NormalTok{(moanin)}
+\end{Highlighting}
+\end{Shaded}
+
+To install \texttt{moanin} directly from GitHub, use the devtools' \texttt{install\_github}
+function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(devtools)}
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
+
+\KeywordTok{library}\NormalTok{(moanin)}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{quality-control-and-normalization}{%
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
@@
 \begin{Highlighting}[]
-\KeywordTok{kable}\NormalTok{(}\KeywordTok{head}\NormalTok{(moanin_model}\OperatorTok{$}\NormalTok{basis))}
-\end{Highlighting}
-\end{Shaded}
+\KeywordTok{kable}\NormalTok{(}\KeywordTok{head}\NormalTok{(moanin_model}\OperatorTok{$}\NormalTok{basis), }\DataTypeTok{digits=}\DecValTok{2}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{  }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{font_size =} \DecValTok{7}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}[H]
+\centering\begingroup\fontsize{7}{9}\selectfont
 
@@
 \hline
-GSM1557140 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
-\hline
-GSM1557141 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
-\hline
-GSM1557142 & 0 & 1 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0 & 0 & 0.0000000 & 0 & 0 & 0\\
-\hline
-GSM1557143 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
-\hline
-GSM1557144 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
-\hline
-GSM1557145 & 0 & 1 & 0 & 0 & 0 & 0 & 0.5142857 & 0 & 0 & 0 & 0 & 0.0357143 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.1546016 & 0 & 0 & 0 & 0 & 0.3478535 & 0 & 0 & 0 & 0 & -0.1932519 & 0 & 0 & 0\\
+GSM1557140 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
+\hline
+GSM1557141 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
+\hline
+GSM1557142 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
+\hline
+GSM1557143 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
+\hline
+GSM1557144 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
+\hline
+GSM1557145 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
 \hline
 \end{tabular}
+\endgroup{}
+\end{table}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
 
@@
 different combinations of timepoints DE. For the comparison of the Kawasaki
-strain to the control, for example, there are \texttt{length(signifCombos)} genes
-found DE in some timepoints, and there are \texttt{length(tabCombos)} different
+strain to the control, for example, there are 26534 genes
+found DE in some timepoints, and there are 1590 different
 combinations of timepoints for which they are DE. Some of these make sense,
-such as DE in timepoints 48H-168H (\texttt{tabCombos{[}"48,60,72,120,168"{]}} genes), but
-many are very fragmentary. For example there are \texttt{tabCombos{[}"48,60,120,168"{]}}
+such as DE in timepoints 48H-168H (509 genes), but
+many are very fragmentary. For example there are 330
 genes which are DE in timepoints 48,60,120,168H, but \emph{not} in the 72H. Many of
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-16-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
 
@@
 timely & \(\text{lfc}(t)\) & Function of time\tabularnewline
-sum & \(\sum_t lfc(t)\) & Sum of log fold change.\tabularnewline
-abs\_sum & \(\sum_t \lvert lfc(t)\lvert\) & Always positive\tabularnewline
-max & \(\max_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
-min & \(\min_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
+sum & \(\sum_t \text{lfc}(t)\) & Sum of log fold change.\tabularnewline
+abs\_sum & \(\sum_t \lvert \text{lfc}(t)\lvert\) & Always positive\tabularnewline
+max & \(\max_t \lvert \text{lfc}(t) \lvert\) & Always positive\tabularnewline
+min & \(\min_t \lvert \text{lfc}(t) \lvert\) & Always positive\tabularnewline
 timecourse & See details below & Captures overall strength and direction\tabularnewline
@@
 
-\[\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)\]
+\[\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T \text{lfc}(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {\text{lfc}(t)} \lvert \right)\]
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
+The returning object is a matrix, where each row corresponds to a gene,
+each column to a contrasts, and each entry to the log-fold change for this
+pair of contrast and gene.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{kable}\NormalTok{(log_fold_change_timecourse[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, ], }\DataTypeTok{digits=}\DecValTok{3}\NormalTok{, }\DataTypeTok{booktabs=}\OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{tabular}{lrrrr}
+\toprule
+  & M-K & M-C & M-VL & M-VH\\
+\midrule
+NM\_009912 & 0.432 & 0.753 & -0.836 & -0.488\\
+NM\_008725 & -1.257 & -2.008 & 1.924 & 1.729\\
+NM\_007473 & 0.385 & 0.335 & 0.516 & 0.263\\
+ENSMUST00000094955 & 0.355 & 0.376 & -0.316 & 0.256\\
+NM\_001042489 & -0.266 & 0.410 & 0.415 & 0.374\\
+\bottomrule
+\end{tabular}
+
+Each methods of computing the log-fold change captures different elements of
+the time-course data: overall change, largest change, \ldots{}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
 
@@
 \KeywordTok{plot}\NormalTok{(lfc_timecourse, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
-     \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
+     \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{xlab=}\StringTok{"Timecourse lfc"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
 
@@
 
+The \texttt{splines\_kmeans} function returns a named list with:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{centroids}: a matrix containing the cluster centroids. The matrix is of
+  shape (\texttt{n\_centroid}, \texttt{n\_samples}).
+\item
+  \texttt{clusters}: a vector of size \texttt{n\_genes}, containing the
+\end{itemize}
+
 We then use the \texttt{plot\_splines\_data} function to visualize the centroids of
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
 
@@
 \NormalTok{    data, kmeans_clusters)}
-\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
-
-\CommentTok{# Let's keep only the list of genes that have a label.}
-\NormalTok{scores =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels), ])}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
-\NormalTok{genes_in_cluster2 =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\end{Highlighting}
-\end{Shaded}
-
-Let us now, for each cluster, visualize the distribution of scores of the
-genes assigned to that cluster.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Get the best score for each gene}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{scores\_and\_labels} list contains two elements:
+
+\begin{verbatim}
+- `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
+  containing for each gene and each cluster, the goodness of fit as
+  described above.
+- `scores_and_labels$labels`: the labels for all of the genes with a
+  good-enough goodness-of-fit score.
+\end{verbatim}
+
+First, let us visualize the distribution of goodness-of-fit scores for each
+cluster. Here, we display the whole set of scores assigned to each cluster,
+prior to filtering poor-quality fits.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# Get the best score and best label for all of the genes}
+\NormalTok{scores =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
+\NormalTok{labels =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, which.min)}
+\NormalTok{max_score =}\StringTok{ }\KeywordTok{quantile}\NormalTok{(scores, }\FloatTok{0.5}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{3}\NormalTok{))}
 \NormalTok{n_clusters =}\StringTok{ }\KeywordTok{dim}\NormalTok{(kmeans_clusters}\OperatorTok{$}\NormalTok{centroids)[}\DecValTok{1}\NormalTok{]}
@@
      \DataTypeTok{xlab=}\StringTok{"score"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"Num. genes"}\NormalTok{)}
+    \KeywordTok{abline}\NormalTok{(}\DataTypeTok{v=}\NormalTok{max_score, }\DataTypeTok{col=}\StringTok{"red"}\NormalTok{, }\DataTypeTok{lwd=}\DecValTok{3}\NormalTok{, }\DataTypeTok{lty=}\DecValTok{2}\NormalTok{)}
 \NormalTok{\}}
@@
 \begin{center}\includegraphics{reports/manuscript_files/figure-latex/visualize_scores-1} \end{center}
+
+Note that for some of the genes, the best scores is 1. A score of 1 indicates
+that there is no correlation between the gene and the cluster: those genes fit
+poorly to all clusters and the assignment to one cluster is done randomly..
+This underlines the importance of filtering genes that fit poorly to clusters.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
+
+\CommentTok{# Let's keep only the list of genes that have a label.}
+\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
+
+\CommentTok{# And also keep track of all the genes in cluster 2.}
+\NormalTok{genes_in_cluster2 =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-28-1} \end{center}
-
-\textbf{EAP: I don't know what to think about these plots -- these are best scoring,
-but pretty noisy. Also the per-gene spline is sometimes not as good of a
-explanation as the cluster centroid. It's seems like an important plot to
-make}-
-
-\textbf{NV: this were the worst scoring genes. I fixed this and the plots now look
-much better}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-27-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-31-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-35-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-35-2} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-38-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-39-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-38-1} \end{center}
 
 This analysis doesn't necessarily pick a particular \(k\), but can help decide between \(k\) within a desired range, for example, or to avoid \(k\) that degrade the stability.
-
-\textbf{EAP: \texttt{plot\_model\_explorer} is relatively slow, because it does all those pairwise calculations.
-And then to do it for smaller subset, redo all the calculations.
-Can't the output be saved, or the calculations be a different step from the plotting?}
 
@@
 ## other attached packages:
-##  [1] RColorBrewer_1.1-2   moanin_0.0.0         topGO_2.32.0        
-##  [4] SparseM_1.77         GO.db_3.6.0          AnnotationDbi_1.42.1
-##  [7] IRanges_2.14.12      S4Vectors_0.18.3     graph_1.58.2        
-## [10] viridis_0.5.1        viridisLite_0.3.0    KEGGprofile_1.22.0  
-## [13] RCurl_1.95-4.12      bitops_1.0-6         NMF_0.21.0          
-## [16] Biobase_2.40.0       BiocGenerics_0.26.0  cluster_2.0.9       
-## [19] rngtools_1.3.1.1     pkgmaker_0.27        registry_0.5-1      
-## [22] ggplot2_3.1.1        kableExtra_1.1.0     biomaRt_2.36.1      
-## [25] BiocStyle_2.8.2      knitr_1.22           limma_3.36.5        
-## [28] usethis_1.5.0        devtools_2.0.2       rmarkdown_1.12      
+##  [1] RColorBrewer_1.1-2   moanin_0.0.0         ggfortify_0.4.6     
+##  [4] topGO_2.32.0         SparseM_1.77         GO.db_3.6.0         
+##  [7] AnnotationDbi_1.42.1 IRanges_2.14.12      S4Vectors_0.18.3    
+## [10] graph_1.58.2         viridis_0.5.1        viridisLite_0.3.0   
+## [13] KEGGprofile_1.22.0   RCurl_1.95-4.12      bitops_1.0-6        
+## [16] NMF_0.21.0           Biobase_2.40.0       BiocGenerics_0.26.0 
+## [19] cluster_2.0.9        rngtools_1.3.1.1     pkgmaker_0.27       
+## [22] registry_0.5-1       ggplot2_3.1.1        kableExtra_1.1.0    
+## [25] biomaRt_2.36.1       BiocStyle_2.8.2      knitr_1.22          
+## [28] limma_3.36.5         usethis_1.5.0        devtools_2.0.2      
+## [31] rmarkdown_1.12      
 ## 
@@
 ## [77] R6_2.4.0                DBI_1.0.0              
-## [79] pillar_1.3.1            withr_2.1.2            
-## [81] mgcv_1.8-28             abind_1.4-5            
+## [79] mgcv_1.8-28             pillar_1.3.1           
+## [81] withr_2.1.2             abind_1.4-5            
 ## [83] KEGGREST_1.20.2         tibble_2.1.1           
@@
 ## [93] webshot_0.5.1           xtable_1.8-4           
-## [95] munsell_0.5.0           magic_1.5-9            
-## [97] sessioninfo_1.1.1
+## [95] tidyr_0.8.3             munsell_0.5.0          
+## [97] magic_1.5-9             sessioninfo_1.1.1
 \end{verbatim}

### 2020-01-15 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2019-11-19 02:22:00.
+The manuscript was last compile on 2020-01-15 09:42:00.
 
@@
 \midrule
-04060 & Cytokine-cytokine receptor interaction & 34 & 0\\
-04620 & Toll-like receptor signaling pathway & 40 & 0\\
-04630 & Jak-STAT signaling pathway & 33 & 0\\
-03050 & Proteasome & 53 & 0\\
+04060 & Cytokine-cytokine receptor interaction & 30 & 0\\
 04621 & NOD-like receptor signaling pathway & 47 & 0\\
+04620 & Toll-like receptor signaling pathway & 37 & 0\\
+04380 & Osteoclast differentiation & 34 & 0\\
+04630 & Jak-STAT signaling pathway & 29 & 0\\
 \addlinespace
-04380 & Osteoclast differentiation & 34 & 0\\
-04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
 04210 & Apoptosis & 37 & 0\\
-04622 & RIG-I-like receptor signaling pathway & 39 & 0\\
-05160 & Hepatitis C & 29 & 0\\
+04623 & Cytosolic DNA-sensing pathway & 39 & 0\\
+05160 & Hepatitis C & 27 & 0\\
+04622 & RIG-I-like receptor signaling pathway & 35 & 0\\
+05140 & Leishmaniasis & 35 & 0\\
 \bottomrule
@@
 \midrule
-GO:0007224 & smoothened signaling pathway & 74 & 69 & 57.82 & 0.00041 & 0.326\\
-GO:0010970 & transport along microtubule & 82 & 75 & 64.07 & 0.00045 & 0.326\\
-GO:0048562 & embryonic organ morphogenesis & 155 & 137 & 121.10 & 0.00067 & 0.378\\
-GO:0060173 & limb development & 92 & 84 & 71.88 & 0.00067 & 0.378\\
-GO:0006805 & xenobiotic metabolic process & 47 & 45 & 36.72 & 0.00089 & 0.451\\
+GO:0007224 & smoothened signaling pathway & 74 & 69 & 57.81 & 0.00041 & 0.327\\
+GO:0010970 & transport along microtubule & 84 & 77 & 65.62 & 0.00045 & 0.327\\
+GO:0048562 & embryonic organ morphogenesis & 155 & 137 & 121.09 & 0.00066 & 0.379\\
+GO:0060173 & limb development & 92 & 84 & 71.87 & 0.00067 & 0.379\\
+GO:0006805 & xenobiotic metabolic process & 47 & 45 & 36.72 & 0.00088 & 0.448\\
 \addlinespace
-GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.76 & 0.00099 & 0.456\\
-GO:0019395 & fatty acid oxidation & 53 & 51 & 41.41 & 0.00133 & 0.562\\
-GO:0055114 & oxidation-reduction process & 533 & 458 & 416.42 & 0.00152 & 0.593\\
-GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.69 & 0.00216 & 0.783\\
-GO:0010811 & positive regulation of cell-substrate ad... & 77 & 70 & 60.16 & 0.00260 & 0.879\\
+GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.75 & 0.00098 & 0.454\\
+GO:0019395 & fatty acid oxidation & 53 & 51 & 41.41 & 0.00133 & 0.564\\
+GO:0055114 & oxidation-reduction process & 533 & 458 & 416.40 & 0.00151 & 0.591\\
+GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.68 & 0.00215 & 0.782\\
+GO:0010811 & positive regulation of cell-substrate ad... & 77 & 70 & 60.15 & 0.00259 & 0.824\\
 \addlinespace
-GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.78 & 0.00293 & 0.929\\
-GO:0021510 & spinal cord development & 48 & 45 & 37.50 & 0.00339 & 1.000\\
-GO:0016571 & histone methylation & 74 & 67 & 57.82 & 0.00418 & 1.000\\
-GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00435 & 1.000\\
-GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00435 & 1.000\\
+GO:0016571 & histone methylation & 77 & 70 & 60.15 & 0.00259 & 0.824\\
+GO:0021510 & spinal cord development & 49 & 46 & 38.28 & 0.00280 & 0.829\\
+GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.78 & 0.00293 & 0.829\\
+GO:0060078 & regulation of postsynaptic membrane pote... & 68 & 62 & 53.12 & 0.00384 & 1.000\\
+GO:0031109 & microtubule polymerization or depolymeri... & 61 & 56 & 47.65 & 0.00405 & 1.000\\
 \bottomrule
@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.1             pkgconfig_2.0.2        
+## [65] rlang_0.4.2             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          

### 2020-04-29 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

--- 2019-11-19 manuscript.tex
@@
 
-The manuscript was last compile on 2019-11-19 02:22:00.
+The manuscript was last compile on 2020-04-29 10:12:35.
 
@@
 \midrule
-GO:0007224 & smoothened signaling pathway & 74 & 69 & 57.82 & 0.00041 & 0.326\\
-GO:0010970 & transport along microtubule & 82 & 75 & 64.07 & 0.00045 & 0.326\\
-GO:0048562 & embryonic organ morphogenesis & 155 & 137 & 121.10 & 0.00067 & 0.378\\
-GO:0060173 & limb development & 92 & 84 & 71.88 & 0.00067 & 0.378\\
-GO:0006805 & xenobiotic metabolic process & 47 & 45 & 36.72 & 0.00089 & 0.451\\
+GO:0003002 & regionalization & 171 & 153 & 133.53 & 0.00012 & 0.123\\
+GO:0007224 & smoothened signaling pathway & 75 & 70 & 58.56 & 0.00033 & 0.282\\
+GO:0010970 & transport along microtubule & 84 & 77 & 65.59 & 0.00044 & 0.322\\
+GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.26 & 0.00057 & 0.365\\
+GO:0048562 & embryonic organ morphogenesis & 154 & 136 & 120.25 & 0.00073 & 0.374\\
 \addlinespace
-GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.76 & 0.00099 & 0.456\\
-GO:0019395 & fatty acid oxidation & 53 & 51 & 41.41 & 0.00133 & 0.562\\
-GO:0055114 & oxidation-reduction process & 533 & 458 & 416.42 & 0.00152 & 0.593\\
-GO:0090596 & sensory organ morphogenesis & 134 & 118 & 104.69 & 0.00216 & 0.783\\
-GO:0010811 & positive regulation of cell-substrate ad... & 77 & 70 & 60.16 & 0.00260 & 0.879\\
+GO:0055114 & oxidation-reduction process & 530 & 457 & 413.85 & 0.00086 & 0.374\\
+GO:0060173 & limb development & 90 & 82 & 70.28 & 0.00090 & 0.374\\
+GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.69 & 0.00095 & 0.374\\
+GO:0006635 & fatty acid beta-oxidation & 38 & 37 & 29.67 & 0.00095 & 0.374\\
+GO:0090596 & sensory organ morphogenesis & 138 & 122 & 107.76 & 0.00124 & 0.453\\
 \addlinespace
-GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.78 & 0.00293 & 0.929\\
-GO:0021510 & spinal cord development & 48 & 45 & 37.50 & 0.00339 & 1.000\\
-GO:0016571 & histone methylation & 74 & 67 & 57.82 & 0.00418 & 1.000\\
-GO:0042738 & exogenous drug catabolic process & 22 & 22 & 17.19 & 0.00435 & 1.000\\
-GO:1901381 & positive regulation of potassium ion tra... & 22 & 22 & 17.19 & 0.00435 & 1.000\\
+GO:0010811 & positive regulation of cell-substrate ad... & 79 & 72 & 61.69 & 0.00184 & 0.628\\
+GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.77 & 0.00289 & 0.903\\
+GO:0060078 & regulation of postsynaptic membrane pote... & 69 & 63 & 53.88 & 0.00319 & 0.903\\
+GO:0042738 & exogenous drug catabolic process & 23 & 23 & 17.96 & 0.00335 & 0.903\\
+GO:0021515 & cell differentiation in spinal cord & 23 & 23 & 17.96 & 0.00335 & 0.903\\
 \bottomrule
@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.1             pkgconfig_2.0.2        
+## [65] rlang_0.4.5             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          

### 2020-06-05 scripts/manuscript.Rmd
changes to introduction/installation

@@
 - Check that clustering can deal with count data.
-
-*EAP*: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
+- Update Nelle's affiliation!
 
@@
 
+## Installation and setup
+
+ `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`:
+
+```{r eval=FALSE}
+library(devtools)
+install_github("NelleV/moanin")
+library(moanin)
+```
+
+The following additional packages are needed for this workflow:
+
+```{r all_libraries}
+# From CRAN
+library(NMF)
+library(ggfortify)
+
+# From Bioconductor
+library(topGO)
+library(biomaRt)
+
+# From GitHub
+library(moanin)
+```
+
+**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at the beginning of the document? I also needed to install `BiocWorkflowTools`, `KEGGprofile` to run the .Rmd
+
 # Analysis of the dynamical response of mouse lung tissue to influenza
@@
 transcriptomic response of the crop *S. bicolor* to drought.
-
-## Package versions
-
-The following packages are needed for this workflow.
-
-```{r all_libraries}
-# From CRAN
-library(NMF)
-library(ggfortify)
-
-# From Bioconductor
-library(topGO)
-library(biomaRt)
-
-# From GitHub
-library(moanin)
-```
-
-To install `moanin` directly from GitHub, use the devtools' `install_github`
-function:
-
-```{r eval=FALSE}
-library(devtools)
-install_github("NelleV/moanin")
-
-library(moanin)
-```
 
@@
 ```{r display_lfc}
-kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE, escape=F)
+kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
 ```

### 2020-06-05 scripts/manuscript.Rmd
possible fix for caption data

@@
   cache=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
-  echo=TRUE,
-  duplicate.label="allow"
-)
+  echo=TRUE, fig.width=4,
+  duplicate.label="allow",out.extra = "")
+
 ```
@@
 
-knitr::include_graphics("images/fig1_schemas.png")
+knitr::include_graphics("images/fig1_schemas.png",dpi=30 )
 
@@
 
-```{r echo=FALSE, fig.width=6, fig.height=3, fig.caption="Distribution of variance"}
+```{r echo=FALSE, fig.width=6, fig.height=3, fig.cap="Distribution of variance"}
 hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
@@
 ```
+
+**EAP**: Not sure why you hide the code for this plot, but not for others...
 
@@
 
-```{r  correlation_plot, fig.width=6, fig.height=6, fig.caption="Cross-correlation of the samples"}
+```{r  correlation_plot, fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
 # Reorder the conditions such that:
@@
 
-```{r echo=FALSE, fig.width=3, fig.height=3, fig.caption="Distribution of DE genes per condition"}
+```{r echo=FALSE, fig.width=3, fig.height=3, fig.cap="Distribution of DE genes per condition"}
 num_de = colSums(qvalues < 0.05)
@@
 
-```{r echo=FALSE, fig.height=6, fig.width=6, fig.caption="Confusion matrix between the two sets of labels"}
+```{r echo=FALSE, fig.height=6, fig.width=6, fig.cap="Confusion matrix between the two sets of labels"}
 genes_scored = names(labels)

### 2020-06-05 scripts/manuscript.Rmd
move around data load and hide code for normalization

@@
 
-## Quality control and normalization
-
-The first steps of analysis of gene expression data is always to do
-normalization and quality control checks of the data. In what follows, we show
-an example of this for the influenza data using generic methods; these steps
-are not in specific to time course data, but could be done for any gene expression
-analysis.
+## Overview of the data
 
@@
 define a series of markers (or plotting symbols) to distinguish replicate
-samples in scatter plots.
+samples in scatter plots. We also reorder the factor `Group` which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).
 
@@
     Replicate=replicate_markers)
-```
+
+# Reorder the conditions such that:
+#	- Control is before any influenza treatment
+#	- Each treatment is ordered from low to high pathogeny
+meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
+
+```
+
+## Quality control and normalization
+
+The first steps of analysis of gene expression data is always to do
+normalization and quality control checks of the data. In what follows, we show
+an example of this for the influenza data using generic methods; these steps
+are not in specific to time course data, but could be done for any gene expression
+analysis. For this reason, we will not print the detailed code that is needed for this part of the analysis; interested readers can examine the `rmarkdown` code that accompanies this workflow.
 
@@
 
-**EAP**: Not sure why you hide the code for this plot, but not for others...
-
 Let us first perform the PCA analysis. Here, we perform a PCA of rank 3 of the
@@
 
-```{r pca, fig.width=6, fig.height=6}
+```{r pca, echo=FALSE, results="hide",fig.width=6, fig.height=6}
 pca_data = prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
@@
 
-```{r pca_plots, fig.width=6, fig.height=6}
+```{r pca_plots, echo=FALSE, results="hide", fig.width=6, fig.height=6}
 par(mfrow=c(2, 2))
@@
 
-Then, we plot the pearson correlation across each samples. We order the
+Then, we plot the pearson correlation between each sample as a heatmap diagram (using the function `aheatmap` in `NMF`). We order the
 samples by their Group (treatment) and Timepoint (the time of sampling).
@@
 
-```{r  correlation_plot, fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
-# Reorder the conditions such that:
-#	- Control is before any influenza treatment
-#	- Each treatment is ordered from low to high pathogeny
-meta$Group = factor(meta$Group, levels(meta$Group)[c(3, 2, 1, 5, 4)])
+```{r  correlation_plot, echo=FALSE, results="hide",fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
 

### 2020-06-05 scripts/manuscript.Rmd
another idea for fixing captions

--- 2020-06-05 scripts/manuscript.Rmd
@@
   echo=TRUE, fig.width=4,
-  duplicate.label="allow",out.extra = "")
+  duplicate.label="allow",fig.align="center")
 

### 2020-06-05 scripts/manuscript.Rmd
text edits

--- 2020-06-05 scripts/manuscript.Rmd
@@
   cache=TRUE, autodep=TRUE, warning=FALSE, error=FALSE, message=FALSE,
-  echo=TRUE, fig.width=4,
+  echo=TRUE, out.width=".5\\textwidth",
   duplicate.label="allow",out.extra = "")
@@
 
-```{r schema, echo=FALSE, fig.width=1, fig.cap="Workflow for analyzing time-course datasets."}
+```{r schema, echo=FALSE, fig.cap="Workflow for analyzing time-course datasets."}
 
@@
 
-```{r pca, echo=FALSE, results="hide",fig.width=6, fig.height=6}
+```{r pca, echo=FALSE, results="hide"}
 pca_data = prcomp(t(variance_filtered_data), rank=3, center=TRUE, scale=TRUE) 
@@
 
-We then plot the two first PC components, and color each sample by (1) its
-condition; (2) its sampling time. We use different plotting symbols for each
+In the following plot, we plot the two first PC components in the first column, where each sample is colored by its
+condition (top row) and its sampling time (bottom row). 
+We use different plotting symbols for each
 replicate. We also plot the second and third components in the second column.
 
-```{r pca_plots, echo=FALSE, results="hide", fig.width=6, fig.height=6}
+```{r pca_plots, echo=FALSE, results="hide", fig.width=6, fig.height=6, fig.cap="Plot of PC components: PC2 vs PC1 (right) and PC2 vs PC3 (left).
+Samples are colored by condition (top row) and sampling time (bottom row)."}
 par(mfrow=c(2, 2))

### 2020-06-05 scripts/manuscript.Rmd
text edits
remove underscore in plot chunk name

--- 2020-06-05 scripts/manuscript.Rmd
@@
 
-```{r echo=FALSE, fig.width=6, fig.height=3, fig.cap="Distribution of variance"}
+```{r geneVarHist, echo=FALSE, fig.width=6, fig.height=3, fig.cap="Distribution of variance"}
 hist(variance_per_genes, breaks=100, col="black", xlab="Variance",
@@
 
-```{r pca_plots, fig.width=6, fig.height=6}
+```{r pcaPlots, fig.width=6, fig.height=6}
 par(mfrow=c(2, 2))
@@
 
-```{r  correlation_plot, fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
+```{r  correlationPlot, fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
 # Reorder the conditions such that:

### 2020-06-05 scripts/manuscript.Rmd

--- 2020-06-05 scripts/manuscript.Rmd
@@
   echo=TRUE, out.width=".5\\textwidth",
-  duplicate.label="allow",out.extra = "")
+  duplicate.label="allow")
 
@@
 
-```{r pca_plots, echo=FALSE, results="hide", fig.width=6, fig.height=6, fig.cap="Plot of PC components: PC2 vs PC1 (right) and PC2 vs PC3 (left).
+```{r pcaPlots, echo=FALSE, results="hide", fig.width=6, fig.height=6, fig.cap="Plot of PC components: PC2 vs PC1 (right) and PC2 vs PC3 (left).
 Samples are colored by condition (top row) and sampling time (bottom row)."}
@@
 
-```{r  correlation_plot, echo=FALSE, results="hide",fig.width=6, fig.height=6, fig.cap="Cross-correlation of the samples"}
-
+```{r  correlationPlot, echo=FALSE, results="hide",fig.width=6, fig.height=6, fig.cap="Heatmap of the correlation between samples"}
 # Reorder genes on condition, time, and replicate

### 2020-06-05 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-04-29 10:12:35.
-
-\#\textless{}U+00A0\textgreater{}TODO list
+The manuscript was last compile on 2020-06-05 16:11:32.
+
+\hypertarget{todo-list}{%
+\section{TODO list}\label{todo-list}}
 
@@
   Check that clustering can deal with count data.
+\item
+  Update Nelle's affiliation!
 \end{itemize}
-
-\emph{EAP}: What is your plan for the installation instructions Nelle? Shouldn't it be moved where we introduce the package, and give instructions about how to install it there.
 
@@
 
-{\centering \includegraphics[width=4.81in]{images/fig1_schemas}
+{\centering \includegraphics[width=11.53in]{images/fig1_schemas}
 
@@
 \end{figure}
+
+\hypertarget{installation-and-setup}{%
+\subsection{Installation and setup}\label{installation-and-setup}}
+
+\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(devtools)}
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
+\KeywordTok{library}\NormalTok{(moanin)}
+\end{Highlighting}
+\end{Shaded}
+
+The following additional packages are needed for this workflow:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# From CRAN}
+\KeywordTok{library}\NormalTok{(NMF)}
+\KeywordTok{library}\NormalTok{(ggfortify)}
+
+\CommentTok{# From Bioconductor}
+\KeywordTok{library}\NormalTok{(topGO)}
+\KeywordTok{library}\NormalTok{(biomaRt)}
+
+\CommentTok{# From GitHub}
+\KeywordTok{library}\NormalTok{(moanin)}
+\end{Highlighting}
+\end{Shaded}
+
+\textbf{EAP}: Why is this list not the same as the (hidden via echo=FALSE) list at the beginning of the document?
+I also needed to install \texttt{BiocWorkflowTools}, \texttt{KEGGprofile} to run the .Rmd
 
@@
 
-\hypertarget{package-versions}{%
-\subsection{Package versions}\label{package-versions}}
-
-The following packages are needed for this workflow.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# From CRAN}
-\KeywordTok{library}\NormalTok{(NMF)}
-\KeywordTok{library}\NormalTok{(ggfortify)}
-
-\CommentTok{# From Bioconductor}
-\KeywordTok{library}\NormalTok{(topGO)}
-\KeywordTok{library}\NormalTok{(biomaRt)}
-
-\CommentTok{# From GitHub}
-\KeywordTok{library}\NormalTok{(moanin)}
-\end{Highlighting}
-\end{Shaded}
-
-To install \texttt{moanin} directly from GitHub, use the devtools' \texttt{install\_github}
-function:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(devtools)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
-
-\KeywordTok{library}\NormalTok{(moanin)}
-\end{Highlighting}
-\end{Shaded}
-
 \hypertarget{quality-control-and-normalization}{%
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\begin{figure}
+
+{\centering \includegraphics{reports/manuscript_files/figure-latex/geneVarHist-1}
+
+}
+
+\caption{Distribution of variance}\label{fig:geneVarHist}
+\end{figure}
+
+\textbf{EAP}: Not sure why you hide the code for this plot, but not for others\ldots{}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/pca_plots-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/pcaPlots-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/correlation_plot-1} \end{center}
+\begin{figure}
+
+{\centering \includegraphics{reports/manuscript_files/figure-latex/correlationPlot-1}
+
+}
+
+\caption{Cross-correlation of the samples}\label{fig:correlationPlot}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\begin{figure}
+
+{\centering \includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1}
+
+}
+
+\caption{Distribution of DE genes per condition}\label{fig:unnamed-chunk-14}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
+\begin{figure}
+
+{\centering \includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1}
+
+}
+
+\caption{Confusion matrix between the two sets of labels}\label{fig:unnamed-chunk-25}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-30-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-35-1} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-1} \end{center}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-35-2} \end{center}
+\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-2} \end{center}
 
@@
 ## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.5             pkgconfig_2.0.2        
+## [65] rlang_0.4.6             pkgconfig_2.0.2        
 ## [67] matrixStats_0.54.0      evaluate_0.13          

### 2020-06-06 scripts/manuscript.Rmd
updates to text, figure legends

@@
 
-In the following plot, we plot the two first PC components in the first column, where each sample is colored by its
-condition (top row) and its sampling time (bottom row). 
-We use different plotting symbols for each
-replicate. We also plot the second and third components in the second column.
+In Figure \ref{fig:pcaPlots}, we visualize the data by plotting the PC components, with samples  colored by either its
+condition (top row) or its sampling time (bottom row) and each replicate a different symbol. We can see a large difference in later time points.
 
@@
 
-Then, we plot the pearson correlation between each sample as a heatmap diagram (using the function `aheatmap` in `NMF`). We order the
+We also plot in Figure \ref{fig:correlationPlot} the pearson correlation between each sample as a heatmap diagram (using the function `aheatmap` in `NMF`). We order the
 samples by their Group (treatment) and Timepoint (the time of sampling).
-Additionally, in this example, we order each treatment by strength of the
-pathogenicity of the treatment: Control, Kawasaki, California, low dose of
-Vietnam, then high dose of Vietnam.
 
@@
 1) Per-time point analysis, where we consider each time point a different
-condition and determine what genes are changing between time points, or
+condition and determine what genes are changing between specific time points, or
 between conditions at a single time-point.
@@
 time-course datasets, where there are only a few time points [@ritchie:limma;
-@robinson:edgeR; @love:moderated] . For long time-course datasets, however, a
+@robinson:edgeR; @love:moderated]. For long time-course datasets, however, a
 separate test for each time-point results in creating many different tests,
@@
 
-**Use with `moanin`** `moanin` provides functionality for performing both of
-these types of approaches, though our focus is on the global approach. In both
+**DE Analysis in `moanin`** Our package `moanin` provides functionality for performing both of
+these types of approaches, though our focus is on the global approach, specifically by fitting spline models to the genes.
+
+In both
 situations, we first need to set up a object (a `moanin_model` object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, …)
+(formula, number of degrees of freedom of the splines, ...). The `moanin_model` object
+will contain information used
+throughout this analysis, in particular the condition and timepoints of each sample and the
+the basis matrix for fitting splines models.
 
@@
 function. We need to provide two things to the function: a `data.frame` with
-the metadata, and the number of degrees of freedom of the splines used in the
+the metadata, and the number of degrees of freedom of the splines to be used in the
 functional modeling. The metadata `data.frame` object should contain at least
@@
 
-The `moanin_model` object contains a number of metadata and options used
-throughout this analysis: the condition and timepoints of each samples, the
-formula object used, the basis matrix, and the degrees of freedom of the model.
-
-If no formula is provided, it will default to the following, which provides
-for a different spline fit for every group (as defined by the `Group` variable
-in the `meta` data.frame): 
+The main operation of the `create_moanin_model` function, in addition to holding the necessary meta data of the samples,
+is to create a basis matrix for the splines fit. This matrix gives the evaluation of all of the spline basis functions for
+each sample (as such, replicate samples will have the same
+values). By default, `create_moanin_model` will create the spline basis functions which will lead to a different spline fit for every group (as defined by the `Group` variable
+in the `meta` data.frame). This is done by the following `R` formula syntax:
+
 `formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.
 
-The basis matrix contains the value of all of the spline basis function for
-each sample of the moment. As such, replicate samples will have the same
-values.
+Alternatively, the user can provide a formula of their own, or simply provide the basis matrix themselves.
 
@@
 
-**EAP: General comment: you need a few more print() and head() of some objects
-to demonstrate what the output looks like**
+**EAP: This needs to be worked on, because too big to print. Really, need a `print.moanin_model` in the package (submitted issue) and put that here.**
 
@@
 `limma` [@ritchie:limma] on the set of contrasts provided. By default,
-`moanin` expects RNA-Seq contact counts, and will estimate voom weights.
-
-Here, we show an example where we define our contrasts to be the difference
+`moanin` expects RNA-Seq gene counts, and will estimate voom weights.
+
+Here, we show an example where we create contrasts that will be be the difference
 between the control mouse ("M") and the mouse infected with the high dose of
-the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point, but
-the function works with any form contrasts [@ritchie:limma].
+the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point (but
+the function works with any form contrasts [@ritchie:limma]).
 
@@
 
-This creates a vector of contrasts to be tested, one for each timepoint, in the format required by `limma`:
+This creates a character vector of contrasts to be tested, one for each timepoint, in the format required by `limma`:
 
 ```{r echo=FALSE}
-kable(contrasts, "latex", col.names="Contrasts", booktabs=TRUE)
-```
+contrasts[1:5]
+```
+**EAP**: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like.
+I like to reserve table format for actual results, rather than code-like input
 
@@
 ```{r echo=FALSE}
-kable(weekly_de_analysis[1:10, 1:6 ], "latex", digits=3, booktabs=TRUE)
+kable(weekly_de_analysis[1:10, 1:3 ], "latex", digits=3, booktabs=TRUE)
 
@@
 
-Let's look at the distribution of genes found differentially expressed per
-week between control and each of the influenza strains.
-
-```{r fig.width=6, fig.height=6, echo=FALSE}
+In Figure \ref{fig:timeSignifPerTime}, we show the distribution of genes found differentially expressed per
+week between control and each of the influenza strains. Such an analysis can demonstrate some general trends, with clearly more genes
+being differentially expressed at later time points, and the Vietnam high-dose
+showing perhaps an earlier onset than the low-dose.
+
+```{r timeSignifPerTime,fig.width=6, fig.height=6, echo=FALSE, fig.cap="Number of genes found DE in each timepoint when comparing the influenza strains to control."}
 timely_barplot = function(de_analysis, 
@@
 ```
-
-Such an analysis can demonstrate some general trends, with clearly more genes
-being differentially expressed at later time points, and the Vietnam high-dose
-showing perhaps an earlier onset than the low-dose.
 
@@
 these genes are likely to have not made the cutoff for significance in 72H,
-but don't show real differences in the overall trend between 48H-168H.
-
-Here are plots of the first 10 such genes, where we can see that some might
+but don't show real differences in the overall trend between 48H-168H. In Figure \ref{fig:showMissingDEGenes} we show the plots of the first 10 such genes. We can see that some might
 show some meaningful changes between 60 and 72H, but others clearly just have
-a single replicate that is different or increased variability at 72H.
-
-```{r showMissingDEGenes, echo=FALSE}
+a single replicate that is different or increased variability at 72H that results in a lack of significance in 72H.
+
+```{r showMissingDEGenes, echo=FALSE, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
 exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
@@
 
-`moanin` extends this idea by providing functionality to compare time course
+`moanin` extends this idea by not only fitting a splines function for each gene, but also providing functionality to compare time course
 data between different treatment conditions, using a similar mechanism of
@@
 functions. This is done with the function `DE_timecourse`, which takes as
-similar input that of `DE_timepoints`, only now will test the entire mean
-function (and therefore does not require a step of expanding the contrasts
-into contrasts for individual timepoints).
+similar input that of `DE_timepoints`, only now will fit spline functions for each gene and test the entire mean
+function (and unlike `DE_timecourse` therefore does not require the extra step of expanding the contrasts into contrasts for individual timepoints).
 
@@
     use_voom_weights=FALSE)
+```
+
+The output from `DE_timecourse` is a matrix of (raw) p-values and Benjamini-Hochberg corrected q-values for each comparison.
+
+```{r}
+names(DE_results)
+```
+
+For convenience we will separate these into two matrices.
+
+```{r}
 pval_columns = colnames(DE_results)[
@@
 
-```{r echo=FALSE, fig.width=3, fig.height=3, fig.cap="Distribution of DE genes per condition"}
+```{r echo=FALSE, fig.cap="Distribution of DE genes per condition based on global splines analysis"}
 num_de = colSums(qvalues < 0.05)
+par(mar=c(7.1,5.1,1.1,1.1))
 names(num_de) = c(
     "Kawasaki", "California", "Vietnam LD", "Vietname HD")
-barplot(num_de, col="black", ylab="Number of DE genes")
-```
+barplot(num_de, col="black", ylab="Number of DE genes",las=2)
+```
+
+### Log-fold Change for Time Course Data
 
 The next step in a classical differential expression analysis is typically to
-assess the effect of the treatment by looking at the log fold change.
+assess the effect of the treatment by calculating for each gene the log fold change in the gene expression between the treatment and control.
+
 Computing the log fold change on a time-course experiment is not trivial: one
@@
 cumulative log-fold change. Sometimes a gene can be over-expressed at the
-beginning of the time-course data, and then over-expressed at the end of the
+beginning of the time-course data, and then under-expressed at the end of the
 experiment. As a result, `moanin` provides a number of possible ways to
-compute the log fold change across the whole time-course.
-
-First, `moanin` provides as simple interface to compute the log-fold change
+compute the log fold change across the whole time-course. This is done via the function `estimate_log_fold_change` which takes as
+arguments the data, the `moanin_model` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
+
+**Individual Timepoints** The first method ("timely") gives a simple interface to compute the log-fold change
 for each individual timepoints.
@@
 
-Sometimes, a single value per gene and per contrast is more useful. Here is a
-table of the possible ways to compute log-fold change values.
+**Cumulative Effect** Sometimes, a single value per gene for each contrast is more useful, and `estimate_log_fold_change` used above provides several options for this as well. Here is a
+table of the possible ways `estimate_log_fold_change` can compute log-fold change values (including `timely` discussed above).
 
@@
 |------- | --------- |--- |
-| timely | $\text{lfc}(t)$  | Function of time |
+| timely | $\text{lfc}(t)$  | Per time point |
 | sum  | $\sum_t \text{lfc}(t)$  | Sum of log fold change. |
@@
 | min  | $\min_t \lvert \text{lfc}(t) \lvert$  | Always positive |
-| timecourse | See details below | Captures overall strength and direction  |
-
-To capture the overall strength and direction of the response, we leverage the
+| timecourse | See details below | Captures overall strength *and* direction  |
+
+The method "timecourse" tries to capture the overall strength and direction of the response in the following way: we leverage the
 timepoint by timepoint log-fold change $\text{lfc}(t)$, and apply the
@@
 
-Note that when a gene is not consistently up or down-regulated the estimation
+Note, however, that when a gene is not consistently up or down-regulated the estimation
 of the direction will not represent accurately the changes observed.
 
-To compute the log-fold change, `moanin` provides a function taking as
-argument the data, the `moanin_model` object, the contrasts to evaluate, as
-well as the method to use to estimate the log-fold change.
+We demonstrate several of these methods on our data.
 
@@
 
-```{r display_lfc}
+```{r display_lfc, echo=FALSE}
 kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
@@
 
-Each methods of computing the log-fold change captures different elements of
-the time-course data: overall change, largest change, ...
-
-```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
+In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
+the time-course data, for example, overall change versus the largest change.
+
+```{r compareLogFoldChange, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, fig.cap="Comparison of the different methods of calculating log-fold change per gene"}
 par(mfrow=c(2, 2))
@@
 
-From the single measures of log-fold change and the p-value, we can now look
-at the volcano plot. Here is an example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the timecourse log fold change
-computation.
-
-```{r fig.height=4, fig.width=4}
+With a single measure of log-fold change and the p-value, we can now look
+at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
+of the control to the Kawasaki strain, using the "timecourse" method of calculating log fold change.
+
+```{r volcano, fig.height=4, fig.width=4, fig.cap="Volcano plot for the comparison of the Kawasaki strain to control, where fold-change is calculated with the timecourse method"}
 pvalue = DE_results[, "M-K_pval"]
@@
 
-As another sanity check, `moanin` provides a simple utility function to
-visualize gene time-course data in different conditions. Here, we plot the 10
+### Visualizing Genes of Interest
+
+The package `moanin` also provides a simple utility function (`plot_splines_data`) to
+visualize gene time-course data from different conditions. In Figure \ref{fig:topPvalueGenes}, we plot the 10
 genes with the smallest p-values.
 
-```{r fig.height=6, fig.width=6}
+```{r topPvalueGenes, fig.height=6, fig.width=6}
 top_DE_genes_pval = names(sort(pvalue)[1:10])
@@
 
-And here we visualize the genes with the largest absolute timecourse log-fold
+For each gene, the individual data points are plotted against time and color coded by their condition.
+Further, a fitted spline function for each group is plotted to aid in comparing global trends across conditions.
+
+Figure \ref{fig:topLogFold}, we visualize the genes with the largest absolute timecourse log-fold
 change.
 
-```{r fig.height=6, fig.width=6}
+```{r topLogFold,fig.height=6, fig.width=6}
 top_DE_genes_lfc = names(
@@
 
-Thanks to those visualization, we can see that genes often follow similar
+In examining these visualizations, we can see that genes often follow similar
 patterns of expression, although on a different scale for each gene. We can
@@
 the workflow to cluster gene expression according to their dynamical response
-to the treatment. Before clustering the genes, we first reduce the set of
+to the treatment.
+
+**Filtering** Before clustering the genes, we first reduce the set of
 genes of interest to genes that are (1) found significantly differentially
-expressed; (2) a large-fold change between conditions. To do this, we first
+expressed; (2) a large-fold change between conditions. Reducing the set of genes on which to perform the clustering allows to
+estimate the centroids of the clusters with more stability.
+
+To do this, we first
 aggregate all p-values obtained during the time-course differential expresison
-step in a single p-value using Fisher's method [@fisher:statistical]. Then we
-select all the genes which have a Fisher adjusted p-value below 0.05 and a log
-fold change of at least two between at least one condition and one time-point.
-Reducing the set of genes on which to perform the clustering allows to
-estimate the centroids with more stability.
+step in a single p-value using Fisher's method [@fisher:statistical] (`pvalues_fisher_method`).
+Next we
+select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold 
+change of at least two for at least one condition and one time-point.
 
@@
 
+### Clustering Based on Spline Fits
+
 After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
-clustering. As observed by looking at genes found differentially expressed, 
+clustering routine. As observed by looking at genes found differentially expressed, 
 many genes share a similar gene expression pattern, but on different scale.
+
 We thus propose the following adaptation of k-means:
@@
 - **Splines estimation**: for each gene, fit the splines function with the basis
-  of your choice (provided in the `moanin_model` object).
+  of your choice (as contained in the `moanin_model` object).
 - **Rescaling splines**: for each gene, rescale the estimated splines function
@@
   cluster centroids.
-- **Assign scores and labels to all genes**: Our
-  `splines_kmeans_score_and_label` function assigns a score to all
-  gene-cluster pairs based on a goodness-of-fit measure of the gene to the
-  cluster centroid. We then "assign" each gene to a cluster based on its
-  score.
-
-The first three steps are performed internally by the `splines_kmeans`
-function. For now, we will set the number of clusters to be $8$, though we
+
+The rescaling of the splines is similar to the rescaling genes by their standard deviation, which is common for gene expression studies without a time component.
+
+These clustering steps are performed by the `splines_kmeans`
+function in `moanin`. For now, we will set the number of clusters to be $8$, though we
 will return to the question of picking the best number of clusters below.
@@
   shape (`n_centroid`, `n_samples`).
-- `clusters`: a vector of size `n_genes`, containing the
-
-We then use the `plot_splines_data` function to visualize the centroids of
+- `clusters`: a vector of size `n_genes`, containing the cluster assignments given by the `kmeans` step of each gene
+
+We then use the `plot_splines_data`  function, only now applied to the centroids, to visualize the centroids of
 each cluster obtained with the splines k-means model.
@@
 These centroids are on a 0-1 scale, because of our rescaling of the spline
-fits, and do not represent the actual gene expression level. We can plot a few
-of the actual genes in a cluster to see the difference:
-
-```{r fig.height=6, fig.width=6}
+fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
+of the actual genes assigned to cluster 2 to see the difference:
+
+```{r genesVsCentroids, fig.height=6, fig.width=6}
 cluster_of_interest = 2
@@
 
-The spline plots are fitted per gene in the above plot, and illustrate that
+The spline plots are fitted per gene in the above plot and illustrate that
 the genes assigned to a cluster have differing degrees to which their data
@@
 
-Since we arbitrarily chose a subset of genes based on our filter, we would
-like all genes to have a chance of assignment to a cluster, as well as a score
-for whether it is a good assignment.  The scoring and label step allows us to
-assign genes to clusters that were removed during the filtering step, yet are
-good matches to the centroids found during the clustering. The score
+Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
+like to have a mechanism to assign all genes to a cluster.
+
+Thus, the next step in the clustering portion of the workflow is a scoring and label step. Each gene is given a score that
 corresponds to a goodness-of-fit between each gene and each cluster, computed
@@
 
+The score then allows us to assign all genes to a cluster (i.e. a label) based on the cluster for which they have the best score, regardless of whether the gene was used in the clustering procedure.
+
 The scoring and labeling is done via the `splines_kmeans_score_and_label`
 function. This function calculates the goodness of fit of the gene to the
-cluster centroid. By default, the function labels only the 50% of genes with
-the best scores, with the remaining genes getting `NA` as their assignment.
+cluster centroid and gives a cluster label to the gene if they have a sufficiently high score, as we explain above.
 
@@
 
-    - `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
+    - `scores`: the matrix of shape `n_cluster` x `n_genes`,
       containing for each gene and each cluster, the goodness of fit as
       described above.
-    - `scores_and_labels$labels`: the labels for all of the genes with a
-      good-enough goodness-of-fit score.
-
-First, let us visualize the distribution of goodness-of-fit scores for each
-cluster. Here, we display the whole set of scores assigned to each cluster,
-prior to filtering poor-quality fits.
-
-```{r visualize_scores, fig.height=6, fix.width=6, echo=TRUE, caption="Distribution of goodness-of-fit scores for each cluster.
-The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
+    - `labels`: the labels for all of the genes with a
+      sufficiently good goodness-of-fit score.
+
+**Assigning cluster labels:** We could just assign each gene to a cluster based on which cluster gave the minimum score.
+By default, `splines_kmeans_score_and_label` does not do that, but rather requires a sufficiently low enough goodness-of-fit score.
+The criteria for being "sufficiently low" is based on looking at the distribution of the scores of all genes on all clusters (i.e.
+the entire `scores` matrix returned by `splines_kmeans_score_and_label`).
+A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment.
+Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
+
+```{r scores_and_labels}
+labels = scores_and_labels$labels
+
+# Let's keep only the list of genes that have a label.
+labels = unlist(labels[!is.na(labels)])
+
+# And also keep track of all the genes in cluster 2.
+genes_in_cluster2 = names(labels[labels==cluster_of_interest])
+```
+
+After running `scores_and_labels` on our data, we now have `r length(labels)` genes that
+are assigned to a cluster.
+
+We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
+cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
+We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
+
+```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster.
+The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled by \\texttt{splines_kmeans_score_and_label}."}
 # Get the best score and best label for all of the genes
+# This is different from output of `splines_kmeans_score_and_label`, which filters
 scores = rowMin(scores_and_labels$scores)
@@
 
-Note that for some of the genes, the best scores is 1. A score of 1 indicates
-that there is no correlation between the gene and the cluster: those genes fit
-poorly to all clusters and the assignment to one cluster is done randomly..
+Note that for some of the genes, their scores in all clusters is 1. Such genes fit
+poorly to all clusters and the assignment we did above to one cluster is done arbitrarily to the first matching cluster.
 This underlines the importance of filtering genes that fit poorly to clusters.
 
-```{r scores_and_labels}
-labels = scores_and_labels$labels
-
-# Let's keep only the list of genes that have a label.
-labels = unlist(labels[!is.na(labels)])
-
-# And also keep track of all the genes in cluster 2.
-genes_in_cluster2 = names(labels[labels==cluster_of_interest])
-```
-
-After running `scores_and_labels` we now have `r length(labels)` genes that
-are assigned to a cluster.
-
-Before performing the next steps, let us investigate in more detail the
-differences between the labels provided by the splines k-means model and the
-scoring and labeling step.
-
-```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
+**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
+
+We can also investigate the
+differences between the labels provided by the splines k-means model and our
+scoring and labeling step, in terms of the number of genes assigned to each cluster (Figure \ref{fig:kmeansVsScore}.
+
+```{r kmeansVsScore, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE, fig.cap="Number of genes assigned to each cluster based on kmeans criteria of nearest centroid (top) versus our goodness-of-fit filtering strategy (bottom)"}
 par(mfrow=c(2, 1))
@@
 fit score. We can compare whether they are still assigned to the same clusters
-based on a confusion matrix shown below. The kmeans cluster assignments are
+based on a confusion matrix shown below (Figure \ref{fig:confusionMatrix}). The kmeans cluster assignments are
 shown on the rows, and the goodness-of-fit assignments in the columns, with
@@
 
-```{r echo=FALSE, fig.height=6, fig.width=6, fig.cap="Confusion matrix between the two sets of labels"}
+```{r confusionMatrix,echo=FALSE, fig.height=6, fig.width=6, fig.cap="Confusion matrix between the two sets of labels"}
 genes_scored = names(labels)
@@
 genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
-few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes
-
-```{r fig.height=6, fig.width=6}
+few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})
+
+```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
 # order them by their score
@@
 
-Now we will create the heatmaps of these genes (`r length(genes_to_plot)` genes).
-
-```{r fig.width=6, fig.height=8}
+Now we will create the heatmaps of these genes (`r length(genes_to_plot)` genes, Figure \ref{fig:heatmapCluster8}).
+
+```{r heatmapCluster8, fig.width=6, fig.height=8, fig.cap="Heatmap of genes in Cluster 8"}
 layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
@@
 
-First, let us run the clustering for all possible clusters of interest. We
+First, let us run the clustering for all values of $k$ of interest. We
 will, for each clustering, conserve (1) with within cluster sum of
 squares; (2) the clustering assignement (or label) for each gene.
+
+**EAP** This didn't make sense to me: "run the clustering for all possible clusters of interest." Do you mean across different values of $k$? That's what I changed it to.
 
@@
 
-We plot the WCSS for $k=2-20$ here. We see that as expected the WCSS continues
+We plot the WCSS for $k=2-20$ here (Figure \ref{fig:elbow_method}). 
+We see that as expected the WCSS continues
 to drop, but there is no clear drop in the decrease, except for very small
@@
 
-```{r elbow_method, fig.width=6, fig.height=4}
+```{r elbow_method, fig.width=6, fig.height=4, fig.cap="Plot of WCSS as a function of k"}
 plot(all_possible_n_clusters, wss_values,
@@
 The silhouette value is a measure of how similar a data point is to its own
-cluster (cohesion) compared to other clusters (separation).
-
-```{r average_silhouette_method, fig.width=6, fig.height=4}
+cluster (cohesion) compared to other clusters (separation), shown in Figure \ref{fig:average_silhouette_method}.
+
+```{r average_silhouette_method, fig.width=6, fig.height=4, fig.cap="Plot of silhouette values as a function of k"}
 # function to compute average silhouette for k clusters
@@
 indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
+
 subsampled_y = y[indices, ]
@@
 
-We run bootstrap method on all genes differentially expressed and with a
+We run the bootstrap method on all genes differentially expressed and with a
 log-fold-change higher than 2 (computed with the `lfc_max` method), and do it
 $B=20$ times for each of $k=2-20$. We show the code below, but because of the
-time it takes we have evaluate these values separately and provided the
-results for users to explore more quickly.
+time the computations take, we have evaluate these values separately and provided the
+results in separate files (one per $k$) for users to explore more quickly.
+
+**EAP**: Why don't you use the same rule/code as above?
+i.e.
+with the fisher p-values etc.?
+A little strange to have two different filtering rules for the clustering.
+Or is this the same, and you're restating the rule?
+(Unclear why you'd have to redo everything here...didn't go back and compare code)
 
@@
 
-```{r fig.width=6, fig.height=4}
+**EAP** It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?
+
+```{r bootstrapConsensus, fig.width=6, fig.height=4,fig.cap="Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20."}
 consensus_matrix_stability_5 = consensus_matrix(stability_5,
@@
 
-We can see that the choice of $K=5$ seems much more stable across resampling runs than that of $K=20$.
+We can see (Figure \ref{fig:bootstrapConsensus}) that the choice of $K=5$ seems much more stable across resampling runs than that of $K=20$.
 
@@
 
-```{r fig.width=6, fig.height=6}
+```{r plotModelExplorer, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k"}
 plot_model_explorer(all_labels)
 ```
-
-From this plot, we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
+**EAP** I'm confused by this plot.
+Isn't the numbering here for different values of $k$?
+Why do the labels say "B1" and "B2"?
+Isn't that the notation above for the bootstrap replications for each values of $k$?
+
+From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
 but not as stable as $k=2$. The model explorer strategy, in addition to
@@
 $k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
-measures in that range.
-
-```{r fig.width=6, fig.height=6}
+measures in that range (Figure \ref{fig:plotModelExplorerZoom}).
+
+```{r  plotModelExplorerZoom, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10"}
 clusters = c("B6", "B7", "B8", "B9", "B10")
@@
 
-```{r echo=FALSE}
+```{r plotCdfConsensus,echo=FALSE,fig.cap="Plot of the Consensus Clustering"}
 knitr::include_graphics("images/clustering_CDF_consensus.png")
@@
 
+Here we calculate the AUC (Figure \ref{fig:plotAUC}).
+
 ```{r eval=FALSE}
@@
 
-```{r echo=FALSE}
+```{r plotAUC,echo=FALSE, fig.cap="Plot of AUC vs number of clusters"}
 knitr::include_graphics("images/clustering_AUC_consensus.png")
 ```
+
+While here we calculate the improvement in the AUC (Figure \ref{fig:plotDeltaAUC}).
 
@@
 
-```{r echo=FALSE}
+```{r plotDeltaAUC,echo=FALSE, fig.cap="Plot of improvement in AUC vs number of clusters"}
 knitr::include_graphics("images/clustering_delta_AUC_consensus.png")

### 2020-06-08 scripts/manuscript.Rmd
updates to manuscript

@@
 Isn't that the notation above for the bootstrap replications for each values of $k$?
+It seems like the names of `all_labels` should be something like `k=1`,..., while the names of `all_labels[[1]]` should be `B1,B2,...`.
+But currently they have the same values.
 
@@
 Once good clusters are obtained, the next step is to leverage the clustering
-to ease interpretation. Classic enrichment analysis step can then be performed
-in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
-term enrichment analysis, motif enrichment analysis, etc.
+to ease interpretation. Classic enrichment analysis can then be performed
+on the gene set defined by each cluster. Examples include KEGG pathway enrichment analysis, 
+GO term enrichment analysis, and motif enrichment analysis.
 
 First, let us clean up the genes we work with and only select the genes we are
-going to use in the enrichment analysis. One can either use the whole set of
-genes, only the set of differentially expressed genes in each cluster, or a
+going to use in the enrichment analysis. One can either all genes assigned to the cluster, only the set of differentially expressed genes in each cluster, or the
 subset of genes that fit well to a cluster (based on some criterion).
@@
 for this step. `KEGGprofile` is a
-package that easily allows to perform pathway enrichment analysis on a set of
-genes labeled with the ensembl annotation. We thus need to convert the gene
-names into the appropriate format. This is where `biomaRt` comes in handy: it
-enables easy conversion from one gene annotation to another. Here, we will use
-it to convert the gene names from the Refseq annotation to the ensembl one.
-
-```{r results="hide"}
+package that facilitates enrichment analysis on a set of
+genes labeled with the Ensembl annotation [@ensembl2020] based on the set of biological pathways described in the KEGG database[@KEGG].
+
+We thus need to convert the gene
+names, which are given in the Refseq annotation, into the corresponding Ensembl name. This is where `biomaRt` comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use the function `getBM`
+in `biomaRt` to convert the gene names from cluster 8.
+
+```{r ensemble_setup, results="hide"}
+## Set up right ensembl database
 ensembl = useMart("ensembl")
 ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
-```
-
-```{r enrichment_analysis, results="hide", message=FALSE}
+
+## Get gene names of genes in Cluster 8
 cluster = 8
@@
 
-# convert gene names
+## Convert gene names
 genes = getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
@@
 		mart=ensembl)["entrezgene_id"]
+```
+
+Then we use the function `find_enriched_pathway` in the `KEGGprofile` package to determine whether any KEGG pathways are enriched in cluster 8, i.e.
+whether a higher percentage of genes from a single pathway are found in cluster 8 than we would expect by simply proportional assignment of genes to clusters.
+
+```{r enrichment_analysis, results="hide", message=FALSE}
 genes = as.vector(unlist(genes))
@@
     download_latest=FALSE)$stastic
-
 ```
@@
 ```{r echo=FALSE}
-
+## order the results by adjusted p-value
 pathways = pathways[
@@
 
-# Select the columns to print
+# Select the columns to print and give nicer column names
 pathways = subset(
@@
 	"pvalueAdj"))
-
 colnames(pathways) = c("Pathway", "Percentage", "Adj. p-value")
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:10, ], format="latex", escape=F, digits=4, booktabs=TRUE)
-```
+kable(pathways[1:min(10,nrow(pathways)), ], format="latex", escape=F, digits=4, booktabs=TRUE)
+```
+
+**EAP: I changed this to be the min(10, nrow(pathways)) but wondering if there's a problem, since presumably there used to be more than 10 pathways found significant and on my machine there are only 2?**
 
@@
 
-To find GO terms, we use `biomaRt` to find the mapping between GO terms and
-gene mapping. The GO enrichment library `topGO` [@alexa:topgo] expects the GO term to gene
+The Gene Ontology (GO) database [@GO], also categorizes genes into meaningful biological ontologies and can be used for enrichment analysis via the package `topGO` [@alexa:topgo].
+We again use `biomaRt` to find the mapping between genes and the GO terms to which they match.
+
+```{r get_go_mappings, message=FALSE}
+genes = getBM(attributes=c("go_id", "refseq_mrna"),
+	      values=gene_names,
+	      filters="refseq_mrna",
+	      mart=ensembl)
+}
+```
+
+The `biomaRt` query results in a matrix with two columns: gene names and GO
+term ID. The package `topGO` [@alexa:topgo] expects the GO term to gene
 mapping to be a list where each item is a mapping between a gene name and a GO
-term ID vector.
+term ID vector, for example:
 
@@
 ```
-
-`biomaRt` queries results a matrix with two named columns of gene names and GO
-term ID.
+`moanin` provides a simple function (`create_go_term_mapping`) to make this conversion:
 
 ```{r go_to_gene_mapping, results="hide", message=FALSE}
-genes = getBM(attributes=c("go_id", "refseq_mrna"),
-	      values=gene_names,
-	      filters="refseq_mrna",
-	      mart=ensembl)
 
@@
 
-Once the gene ID to GO mapping list is created, `moanin` provides a simple
-interface to `topGO` to fetch enriched GO terms. Here, we show an example of
+Once the gene ID to GO mapping list is created, `moanin` provides an 
+interface to `topGO` to determine enriched GO terms. **EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe** Here, we show an example of
 running a GO term enrichment on the "Biological process" ontology (BP).
@@
 ```{r go_term_enrichment, results="hide", message=FALSE}
+# Create logical vector of whether in cluster 8
 assignments = labels == cluster
@@
 ```
+
 # Conclusion
 
-This workflow provides a tutoral for the analysis of time-course gene
-expression data in R, illustrated through the analysis of mice lung tissue
-exposed to different influenza strain. It covers four main steps; (1) quality
-control and normalization; (2) differential expression analysis; (3)
-clustering of time-course gene expression data; (4) downstream analysis of
-clusters.
+This workflow provides a tutoral for the analysis of lengthy time-course gene
+expression data in R using the package `moanin`, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
+exposed to different influenza strain and measured over time. 
+The proposed workflow covers consists of three common analysis main steps generally performed after quality
+control and normalization: (1) differential expression analysis; (2)
+clustering of time-course gene expression data; (3) downstream analysis of
+clusters. We demonstrate how the use of the package `moanin` allows for easy implementation of these procedures in the setting of time-course data.
 

### 2020-06-08 scripts/manuscript.Rmd
fix R error introduced
fix error in fig.cap

@@
 
-```{r echo=FALSE, fig.cap="Distribution of DE genes per condition based on global splines analysis"}
+```{r deTimeDist, echo=FALSE, fig.cap="Distribution of DE genes per condition based on global splines analysis"}
 num_de = colSums(qvalues < 0.05)
@@
 
-```{r elbow_method, fig.width=6, fig.height=4, fig.cap="Plot of WCSS as a function of k"}
+```{r elbowMethod, fig.width=6, fig.height=4, fig.cap="Plot of WCSS as a function of k"}
 plot(all_possible_n_clusters, wss_values,
@@
 	      mart=ensembl)
-}
+
 ```

### 2020-06-09 scripts/manuscript.Rmd
fix error in fig.cap

@@
 
-```{r average_silhouette_method, fig.width=6, fig.height=4, fig.cap="Plot of silhouette values as a function of k"}
+```{r averageSilhouetteMethod, fig.width=6, fig.height=4, fig.cap="Plot of silhouette values as a function of k"}
 # function to compute average silhouette for k clusters

### 2020-06-17 scripts/manuscript.Rmd

@@
     influenza, this workflow provides a step-by-step tutorial of the
-    methodology used to analyse time-course data: (1) normalization of the
-    micro-array data set; (2) differential expression analysis using functional
+    methodology used to analyse time-course data: (1) quality control and normalization of the
+    data set; (2) differential expression analysis using functional
     data analysis; (3) clustering of time-course data; (4) interpreting
@@
 output: BiocWorkflowTools::f1000_article
-
 bibliography: bibliography.bib
 ---
+**EAP**: I don't know that describing normalization as part of our tutorial makes sense.
+I've tried to change the text.
 
@@
   echo=TRUE, out.width=".5\\textwidth",
-  duplicate.label="allow")
-
+  duplicate.label="allow", fig.pos = "H", out.extra = "")
 ```
@@
 The first steps of analysis of gene expression data is always to do
-normalization and quality control checks of the data. In what follows, we show
-an example of this for the influenza data using generic methods; these steps
-are not in specific to time course data, but could be done for any gene expression
+normalization and quality control checks of the data.
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file.
+**EAP**: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.
+
+Instead, we focus on common steps for exploratory analysis of the data, including for the purpose of quality control. These steps
+are not specific to time course data, but could be applied for any gene expression
 analysis. For this reason, we will not print the detailed code that is needed for this part of the analysis; interested readers can examine the `rmarkdown` code that accompanies this workflow.
@@
 
-Typically, two quality control and exploratory analysis steps are also
+Typically, two quality control and exploratory analysis steps are 
 performed before and after normalization: (1) low dimensionality embedding of
@@
 
+We can see this information when we print the object:
+
 ```{r basis_matrix}
-kable(head(moanin_model$basis), digits=2) %>%
-  kable_styling(font_size = 7)
-```
-
-**EAP: This needs to be worked on, because too big to print. Really, need a `print.moanin_model` in the package (submitted issue) and put that here.**
+print(moanin_model)
+```
 
@@
 similar input that of `DE_timepoints`, only now will fit spline functions for each gene and test the entire mean
-function (and unlike `DE_timecourse` therefore does not require the extra step of expanding the contrasts into contrasts for individual timepoints).
+function (and unlike `DE_timepoints`, therefore does not require the extra step of expanding the contrasts into contrasts for individual timepoints).
 
@@
 ```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster.
-The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled by \\texttt{splines_kmeans_score_and_label}."}
+The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
 # Get the best score and best label for all of the genes
-# This is different from output of `splines_kmeans_score_and_label`, which filters
+# This is different from output of `splines_kmeans_score_and_label`, 
+# which filters the genes
 scores = rowMin(scores_and_labels$scores)
@@
 ```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
-# order them by their score
-cluster2Score = genes_in_cluster2[order(scores_and_labels$scores[genes_in_cluster2, cluster_of_interest])]
-dataToPlot = rbind(
-    data[cluster2Score[1:4], ],
+# order them by their score in cluster
+ord = order(scores_and_labels$scores[genes_in_cluster2,
+                     cluster_of_interest])
+cluster2Score = genes_in_cluster2[ord]
+dataToPlot = rbind(data[cluster2Score[1:4], ],
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
@@
 ```
+
+**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.
 

### 2020-06-17 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 \begin{abstract}
-The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology.However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software.
-Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) normalization of the micro-array data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
+The phenotypic diversity of cells is governed by a complex equilibrium between their genetic identity and their environmental interactions: Understanding the dynamics of gene expression is a fundamental question of biology.
+However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software.
+Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
 \end{abstract}
@@
 
-The manuscript was last compile on 2020-06-05 16:11:32.
+\textbf{EAP}: I don't know that describing normalization as part of our tutorial makes sense.
+I've tried to change the text.
+
+The manuscript was last compile on 2020-06-17 05:37:22.
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=11.53in]{images/fig1_schemas}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{images/fig1_schemas}
 
@@
 
-\hypertarget{quality-control-and-normalization}{%
-\subsection{Quality control and normalization}\label{quality-control-and-normalization}}
-
-The first steps of analysis of gene expression data is always to do
-normalization and quality control checks of the data. In what follows, we show
-an example of this for the influenza data using generic methods; these steps
-are not in specific to time course data, but could be done for any gene expression
-analysis.
+\hypertarget{overview-of-the-data}{%
+\subsection{Overview of the data}\label{overview-of-the-data}}
 
@@
 define a series of markers (or plotting symbols) to distinguish replicate
-samples in scatter plots.
+samples in scatter plots. We also reorder the factor \texttt{Group} which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).
 
@@
     \DataTypeTok{Replicate=}\NormalTok{replicate_markers)}
-\end{Highlighting}
-\end{Shaded}
+
+\CommentTok{# Reorder the conditions such that:}
+\CommentTok{#   - Control is before any influenza treatment}
+\CommentTok{#   - Each treatment is ordered from low to high pathogeny}
+\NormalTok{meta}\OperatorTok{$}\NormalTok{Group =}\StringTok{ }\KeywordTok{factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group, }\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group)[}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)])}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{quality-control-and-normalization}{%
+\subsection{Quality control and normalization}\label{quality-control-and-normalization}}
+
+The first steps of analysis of gene expression data is always to do
+normalization and quality control checks of the data.
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file.
+\textbf{EAP}: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.
+
+Instead, we focus on common steps for exploratory analysis of the data, including for the purpose of quality control. These steps
+are not specific to time course data, but could be applied for any gene expression
+analysis. For this reason, we will not print the detailed code that is needed for this part of the analysis; interested readers can examine the \texttt{rmarkdown} code that accompanies this workflow.
 
@@
 
-Typically, two quality control and exploratory analysis steps are also
+Typically, two quality control and exploratory analysis steps are
 performed before and after normalization: (1) low dimensionality embedding of
@@
 
-\begin{figure}
-
-{\centering \includegraphics{reports/manuscript_files/figure-latex/geneVarHist-1}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/geneVarHist-1}
 
@@
 \end{figure}
-
-\textbf{EAP}: Not sure why you hide the code for this plot, but not for others\ldots{}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{pca_data =}\StringTok{ }\KeywordTok{prcomp}\NormalTok{(}\KeywordTok{t}\NormalTok{(variance_filtered_data), }\DataTypeTok{rank=}\DecValTok{3}\NormalTok{, }\DataTypeTok{center=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{scale=}\OtherTok{TRUE}\NormalTok{) }
-\NormalTok{percent_var =}\StringTok{ }\KeywordTok{round}\NormalTok{(}\DecValTok{100} \OperatorTok{*}\StringTok{ }\KeywordTok{attr}\NormalTok{(pca_data, }\StringTok{"percentVar"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-We then plot the two first PC components, and color each sample by (1) its
-condition; (2) its sampling time. We use different plotting symbols for each
-replicate. We also plot the second and third components in the second column.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Group"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\DecValTok{25}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Group),}
-       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group[meta}\OperatorTok{$}\NormalTok{Group],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{4.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC1"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)], }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\DataTypeTok{at=}\DecValTok{275}\NormalTok{, }\DataTypeTok{text=}\StringTok{"Colored by Timepoint"}\NormalTok{, }\DataTypeTok{line=}\FloatTok{0.5}\NormalTok{, }\DataTypeTok{font=}\DecValTok{2}\NormalTok{, }\DataTypeTok{adj=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{legend}\NormalTok{(}\DataTypeTok{x=}\DecValTok{250}\NormalTok{, }\DataTypeTok{y=}\OperatorTok{-}\DecValTok{15}\NormalTok{, }\DataTypeTok{legend=}\KeywordTok{names}\NormalTok{(ann_colors}\OperatorTok{$}\NormalTok{Timepoint),}
-       \DataTypeTok{fill=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint, }\DataTypeTok{bty=}\StringTok{"n"}\NormalTok{, }\DataTypeTok{xpd=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{yjust=}\FloatTok{0.5}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{4.5}\NormalTok{, }\FloatTok{5.5}\NormalTok{, }\FloatTok{1.5}\NormalTok{, }\FloatTok{.1}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(}
-\NormalTok{    pca_data}\OperatorTok{$}\NormalTok{x[, }\KeywordTok{c}\NormalTok{(}\StringTok{"PC3"}\NormalTok{,}\StringTok{"PC2"}\NormalTok{)],}
-    \DataTypeTok{col=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Timepoint[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Timepoint)],}
-    \DataTypeTok{pch=}\NormalTok{ann_markers}\OperatorTok{$}\NormalTok{Replicate[}\KeywordTok{as.factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Replicate)],}
-    \DataTypeTok{ylab=}\StringTok{""}\NormalTok{, }\DataTypeTok{bty=}\StringTok{"l"}\NormalTok{, }\DataTypeTok{yaxt=}\StringTok{"n"}\NormalTok{)}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DataTypeTok{labels=}\OtherTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/pcaPlots-1} \end{center}
-
-Then, we plot the pearson correlation across each samples. We order the
+In Figure \ref{fig:pcaPlots}, we visualize the data by plotting the PC components, with samples colored by either its
+condition (top row) or its sampling time (bottom row) and each replicate a different symbol. We can see a large difference in later time points.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/pcaPlots-1}
+
+}
+
+\caption{Plot of PC components: PC2 vs PC1 (right) and PC2 vs PC3 (left). Samples are colored by condition (top row) and sampling time (bottom row).}\label{fig:pcaPlots}
+\end{figure}
+
+We also plot in Figure \ref{fig:correlationPlot} the pearson correlation between each sample as a heatmap diagram (using the function \texttt{aheatmap} in \texttt{NMF}). We order the
 samples by their Group (treatment) and Timepoint (the time of sampling).
-Additionally, in this example, we order each treatment by strength of the
-pathogenicity of the treatment: Control, Kawasaki, California, low dose of
-Vietnam, then high dose of Vietnam.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Reorder the conditions such that:}
-\CommentTok{#   - Control is before any influenza treatment}
-\CommentTok{#   - Each treatment is ordered from low to high pathogeny}
-\NormalTok{meta}\OperatorTok{$}\NormalTok{Group =}\StringTok{ }\KeywordTok{factor}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group, }\KeywordTok{levels}\NormalTok{(meta}\OperatorTok{$}\NormalTok{Group)[}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{)])}
-
-\CommentTok{# Reorder genes on condition, time, and replicate}
-\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(}
-\NormalTok{  meta}\OperatorTok{$}\NormalTok{Group,}
-\NormalTok{  meta}\OperatorTok{$}\NormalTok{Timepoint,}
-\NormalTok{  meta}\OperatorTok{$}\NormalTok{Replicate)}
-
-\NormalTok{variance_filtered_data =}\StringTok{ }\NormalTok{variance_filtered_data[, ord]}
-\NormalTok{data_corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(variance_filtered_data, }\DataTypeTok{method=}\StringTok{"pearson"}\NormalTok{)}
-\NormalTok{data_corr_meta =}\StringTok{ }\NormalTok{meta[ord, ]}
-\NormalTok{data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(data_corr_meta}\OperatorTok{$}\NormalTok{Timepoint)}
-\CommentTok{# use aheatmap from the NMF package for a heatmap}
-\KeywordTok{aheatmap}\NormalTok{(}
-\NormalTok{    data_corr,}
-    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{, }\DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
-    \DataTypeTok{annCol=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
-    \DataTypeTok{annRow=}\NormalTok{data_corr_meta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
-    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{, }
-    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
-    \DataTypeTok{main=}\StringTok{"Correlation plot"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-
-{\centering \includegraphics{reports/manuscript_files/figure-latex/correlationPlot-1}
-
-}
-
-\caption{Cross-correlation of the samples}\label{fig:correlationPlot}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/correlationPlot-1}
+
+}
+
+\caption{Heatmap of the correlation between samples}\label{fig:correlationPlot}
 \end{figure}
@@
   Per-time point analysis, where we consider each time point a different
-  condition and determine what genes are changing between time points, or
+  condition and determine what genes are changing between specific time points, or
   between conditions at a single time-point.
@@
 approaches, and is often the approach advocated when dealing with small
-time-course datasets, where there are only a few time points \citep{ritchie:limma, robinson:edgeR, love:moderated} . For long time-course datasets, however, a
+time-course datasets, where there are only a few time points \citep{ritchie:limma, robinson:edgeR, love:moderated}. For long time-course datasets, however, a
 separate test for each time-point results in creating many different tests,
@@
 
-\textbf{Use with \texttt{moanin}} \texttt{moanin} provides functionality for performing both of
-these types of approaches, though our focus is on the global approach. In both
+\textbf{DE Analysis in \texttt{moanin}} Our package \texttt{moanin} provides functionality for performing both of
+these types of approaches, though our focus is on the global approach, specifically by fitting spline models to the genes.
+
+In both
 situations, we first need to set up a object (a \texttt{moanin\_model} object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, \textless{}80\textgreater{})
+(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{moanin\_model} object
+will contain information used
+throughout this analysis, in particular the condition and timepoints of each sample and the
+the basis matrix for fitting splines models.
 
@@
 function. We need to provide two things to the function: a \texttt{data.frame} with
-the metadata, and the number of degrees of freedom of the splines used in the
+the metadata, and the number of degrees of freedom of the splines to be used in the
 functional modeling. The metadata \texttt{data.frame} object should contain at least
@@
 
-The \texttt{moanin\_model} object contains a number of metadata and options used
-throughout this analysis: the condition and timepoints of each samples, the
-formula object used, the basis matrix, and the degrees of freedom of the model.
-
-If no formula is provided, it will default to the following, which provides
-for a different spline fit for every group (as defined by the \texttt{Group} variable
-in the \texttt{meta} data.frame):
+The main operation of the \texttt{create\_moanin\_model} function, in addition to holding the necessary meta data of the samples,
+is to create a basis matrix for the splines fit. This matrix gives the evaluation of all of the spline basis functions for
+each sample (as such, replicate samples will have the same
+values). By default, \texttt{create\_moanin\_model} will create the spline basis functions which will lead to a different spline fit for every group (as defined by the \texttt{Group} variable
+in the \texttt{meta} data.frame). This is done by the following \texttt{R} formula syntax:
+
 \texttt{formula\ =\ \textasciitilde{}Group:ns(Timepoint,\ df=degrees\_of\_fredoom)\ +\ Group\ +\ 0}.
 
-The basis matrix contains the value of all of the spline basis function for
-each sample of the moment. As such, replicate samples will have the same
-values.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{kable}\NormalTok{(}\KeywordTok{head}\NormalTok{(moanin_model}\OperatorTok{$}\NormalTok{basis), }\DataTypeTok{digits=}\DecValTok{2}\NormalTok{) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{font_size =} \DecValTok{7}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}[H]
-\centering\begingroup\fontsize{7}{9}\selectfont
-
-\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
-\hline
-  & GroupM & GroupK & GroupC & GroupVL & GroupVH & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)1 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)2 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)3 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)4 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)5 & GroupM:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupK:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupC:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVL:splines::ns(Timepoint, df = degrees\_of\_freedom)6 & GroupVH:splines::ns(Timepoint, df = degrees\_of\_freedom)6\\
-\hline
-GSM1557140 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
-\hline
-GSM1557141 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
-\hline
-GSM1557142 & 0 & 1 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0 & 0 & 0.00 & 0 & 0 & 0\\
-\hline
-GSM1557143 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
-\hline
-GSM1557144 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
-\hline
-GSM1557145 & 0 & 1 & 0 & 0 & 0 & 0 & 0.51 & 0 & 0 & 0 & 0 & 0.04 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & -0.15 & 0 & 0 & 0 & 0 & 0.35 & 0 & 0 & 0 & 0 & -0.19 & 0 & 0 & 0\\
-\hline
-\end{tabular}
-\endgroup{}
-\end{table}
-
-\textbf{EAP: General comment: you need a few more print() and head() of some objects
-to demonstrate what the output looks like}
+Alternatively, the user can provide a formula of their own, or simply provide the basis matrix themselves.
+
+We can see this information when we print the object:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{print}\NormalTok{(moanin_model)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## moanin_object on 209 samples containing the following information:
+## 1) Meta data with 5 variables:
+## [1] ""            "Group"       "Replicate"   "Timepoint"   "WeeklyGroup"
+## 2) Basis matrix with 35 basis functions
+## Basis matrix was constructed with the following formula
+## ~Group + Group:splines::ns(Timepoint, df = degrees_of_freedom) + 0 
+## Where degrees_of_freedom=6 
+## To access these objects use:
+##  <object_name>$degrees_of_freedom
+##  <object_name>$basis
+##  <object_name>$meta
+##  <object_name>$formula
+\end{verbatim}
 
@@
 \texttt{limma} \citep{ritchie:limma} on the set of contrasts provided. By default,
-\texttt{moanin} expects RNA-Seq contact counts, and will estimate voom weights.
-
-Here, we show an example where we define our contrasts to be the difference
+\texttt{moanin} expects RNA-Seq gene counts, and will estimate voom weights.
+
+Here, we show an example where we create contrasts that will be be the difference
 between the control mouse (``M'') and the mouse infected with the high dose of
-the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point, but
-the function works with any form contrasts \citep{ritchie:limma}.
+the influenza strain A/Vietnam/1203/04 (H5N1) (``VL'') for each time point (but
+the function works with any form contrasts \citep{ritchie:limma}).
 
@@
 
-This creates a vector of contrasts to be tested, one for each timepoint, in the format required by \texttt{limma}:
-
-\begin{tabular}{l}
-\toprule
-Contrasts\\
-\midrule
-M.0-VL.0\\
-M.3-VL.3\\
-M.6-VL.6\\
-M.9-VL.9\\
-M.12-VL.12\\
-\addlinespace
-M.18-VL.18\\
-M.24-VL.24\\
-M.30-VL.30\\
-M.36-VL.36\\
-M.48-VL.48\\
-\addlinespace
-M.60-VL.60\\
-M.72-VL.72\\
-M.120-VL.120\\
-M.168-VL.168\\
-\bottomrule
-\end{tabular}
+This creates a character vector of contrasts to be tested, one for each timepoint, in the format required by \texttt{limma}:
+
+\begin{verbatim}
+## [1] "M.0-VL.0"   "M.3-VL.3"   "M.6-VL.6"   "M.9-VL.9"   "M.12-VL.12"
+\end{verbatim}
+
+\textbf{EAP}: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like.
+I like to reserve table format for actual results, rather than code-like input
 
@@
 
-\begin{tabular}{lrrrrrr}
+\begin{tabular}{lrrr}
 \toprule
-  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc & M.3-VL.3\_pval & M.3-VL.3\_qval & M.3-VL.3\_lfc\\
+  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc\\
 \midrule
-NM\_009912 & 0.273 & 0.475 & 2.202 & 0.767 & 0.876 & 1.158\\
-NM\_008725 & 0.851 & 0.924 & 2.028 & 0.139 & 0.302 & 1.457\\
-NM\_007473 & 0.088 & 0.219 & 0.678 & 0.917 & 0.959 & 0.764\\
-ENSMUST00000094955 & 0.008 & 0.037 & 1.186 & 0.935 & 0.969 & -0.633\\
-NM\_001042489 & 0.419 & 0.621 & 1.212 & 0.761 & 0.872 & 0.599\\
+NM\_009912 & 0.273 & 0.475 & 2.202\\
+NM\_008725 & 0.851 & 0.924 & 2.028\\
+NM\_007473 & 0.088 & 0.219 & 0.678\\
+ENSMUST00000094955 & 0.008 & 0.037 & 1.186\\
+NM\_001042489 & 0.419 & 0.621 & 1.212\\
 \addlinespace
-NM\_008159 & 0.123 & 0.278 & 1.102 & 0.117 & 0.268 & 0.737\\
-NM\_001013813 & 0.057 & 0.159 & 0.630 & 0.171 & 0.348 & 0.975\\
-AK039774 & 0.015 & 0.059 & 0.637 & 0.062 & 0.171 & 0.971\\
-NM\_013782 & 0.875 & 0.938 & 0.594 & 0.000 & 0.001 & 1.180\\
-NM\_028622 & 0.767 & 0.876 & 1.648 & 0.946 & 0.974 & -1.865\\
+NM\_008159 & 0.123 & 0.278 & 1.102\\
+NM\_001013813 & 0.057 & 0.159 & 0.630\\
+AK039774 & 0.015 & 0.059 & 0.637\\
+NM\_013782 & 0.875 & 0.938 & 0.594\\
+NM\_028622 & 0.767 & 0.876 & 1.648\\
 \bottomrule
@@
 
-Let's look at the distribution of genes found differentially expressed per
-week between control and each of the influenza strains.
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-13-1} \end{center}
-
-Such an analysis can demonstrate some general trends, with clearly more genes
+In Figure \ref{fig:timeSignifPerTime}, we show the distribution of genes found differentially expressed per
+week between control and each of the influenza strains. Such an analysis can demonstrate some general trends, with clearly more genes
 being differentially expressed at later time points, and the Vietnam high-dose
 showing perhaps an earlier onset than the low-dose.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/timeSignifPerTime-1}
+
+}
+
+\caption{Number of genes found DE in each timepoint when comparing the influenza strains to control.}\label{fig:timeSignifPerTime}
+\end{figure}
 
@@
 these genes are likely to have not made the cutoff for significance in 72H,
-but don't show real differences in the overall trend between 48H-168H.
-
-Here are plots of the first 10 such genes, where we can see that some might
+but don't show real differences in the overall trend between 48H-168H. In Figure \ref{fig:showMissingDEGenes} we show the plots of the first 10 such genes. We can see that some might
 show some meaningful changes between 60 and 72H, but others clearly just have
-a single replicate that is different or increased variability at 72H.
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/showMissingDEGenes-1} \end{center}
+a single replicate that is different or increased variability at 72H that results in a lack of significance in 72H.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/showMissingDEGenes-1}
+
+}
+
+\caption{The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint.}\label{fig:showMissingDEGenes}
+\end{figure}
 
@@
 
-\texttt{moanin} extends this idea by providing functionality to compare time course
+\texttt{moanin} extends this idea by not only fitting a splines function for each gene, but also providing functionality to compare time course
 data between different treatment conditions, using a similar mechanism of
@@
 functions. This is done with the function \texttt{DE\_timecourse}, which takes as
-similar input that of \texttt{DE\_timepoints}, only now will test the entire mean
-function (and therefore does not require a step of expanding the contrasts
-into contrasts for individual timepoints).
+similar input that of \texttt{DE\_timepoints}, only now will fit spline functions for each gene and test the entire mean
+function (and unlike \texttt{DE\_timepoints}, therefore does not require the extra step of expanding the contrasts into contrasts for individual timepoints).
 
@@
 \CommentTok{# (data.frame containing a timepoint and group column, the first corresponding}
-\CommentTok{#<c2><a0>to the time-course information, the latter corresponding to the}
+\CommentTok{# to the time-course information, the latter corresponding to the}
 \CommentTok{# treatment).}
@@
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The output from \texttt{DE\_timecourse} is a matrix of (raw) p-values and Benjamini-Hochberg corrected q-values for each comparison.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{names}\NormalTok{(DE_results)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] "M-K_pval"  "M-K_qval"  "M-C_pval"  "M-C_qval"  "M-VL_pval" "M-VL_qval"
+## [7] "M-VH_pval" "M-VH_qval"
+\end{verbatim}
+
+For convenience we will separate these into two matrices.
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{pval_columns =}\StringTok{ }\KeywordTok{colnames}\NormalTok{(DE_results)[}
@@
 
-\begin{figure}
-
-{\centering \includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-14-1}
-
-}
-
-\caption{Distribution of DE genes per condition}\label{fig:unnamed-chunk-14}
-\end{figure}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/deTimeDist-1}
+
+}
+
+\caption{Distribution of DE genes per condition based on global splines analysis}\label{fig:deTimeDist}
+\end{figure}
+
+\hypertarget{log-fold-change-for-time-course-data}{%
+\subsubsection{Log-fold Change for Time Course Data}\label{log-fold-change-for-time-course-data}}
 
 The next step in a classical differential expression analysis is typically to
-assess the effect of the treatment by looking at the log fold change.
+assess the effect of the treatment by calculating for each gene the log fold change in the gene expression between the treatment and control.
+
 Computing the log fold change on a time-course experiment is not trivial: one
@@
 cumulative log-fold change. Sometimes a gene can be over-expressed at the
-beginning of the time-course data, and then over-expressed at the end of the
+beginning of the time-course data, and then under-expressed at the end of the
 experiment. As a result, \texttt{moanin} provides a number of possible ways to
-compute the log fold change across the whole time-course.
-
-First, \texttt{moanin} provides as simple interface to compute the log-fold change
+compute the log fold change across the whole time-course. This is done via the function \texttt{estimate\_log\_fold\_change} which takes as
+arguments the data, the \texttt{moanin\_model} object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
+
+\textbf{Individual Timepoints} The first method (``timely'') gives a simple interface to compute the log-fold change
 for each individual timepoints.
@@
 
-Sometimes, a single value per gene and per contrast is more useful. Here is a
-table of the possible ways to compute log-fold change values.
+\textbf{Cumulative Effect} Sometimes, a single value per gene for each contrast is more useful, and \texttt{estimate\_log\_fold\_change} used above provides several options for this as well. Here is a
+table of the possible ways \texttt{estimate\_log\_fold\_change} can compute log-fold change values (including \texttt{timely} discussed above).
 
@@
 \endhead
-timely & \(\text{lfc}(t)\) & Function of time\tabularnewline
+timely & \(\text{lfc}(t)\) & Per time point\tabularnewline
 sum & \(\sum_t \text{lfc}(t)\) & Sum of log fold change.\tabularnewline
@@
 min & \(\min_t \lvert \text{lfc}(t) \lvert\) & Always positive\tabularnewline
-timecourse & See details below & Captures overall strength and direction\tabularnewline
+timecourse & See details below & Captures overall strength \emph{and} direction\tabularnewline
 \bottomrule
@@
 
-To capture the overall strength and direction of the response, we leverage the
+The method ``timecourse'' tries to capture the overall strength and direction of the response in the following way: we leverage the
 timepoint by timepoint log-fold change \(\text{lfc}(t)\), and apply the
@@
 
-Note that when a gene is not consistently up or down-regulated the estimation
+Note, however, that when a gene is not consistently up or down-regulated the estimation
 of the direction will not represent accurately the changes observed.
 
-To compute the log-fold change, \texttt{moanin} provides a function taking as
-argument the data, the \texttt{moanin\_model} object, the contrasts to evaluate, as
-well as the method to use to estimate the log-fold change.
+We demonstrate several of these methods on our data.
 
@@
 pair of contrast and gene.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{kable}\NormalTok{(log_fold_change_timecourse[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, ], }\DataTypeTok{digits=}\DecValTok{3}\NormalTok{, }\DataTypeTok{booktabs=}\OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-Each methods of computing the log-fold change captures different elements of
-the time-course data: overall change, largest change, \ldots{}
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
-
-From the single measures of log-fold change and the p-value, we can now look
-at the volcano plot. Here is an example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the timecourse log fold change
-computation.
+In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
+the time-course data, for example, overall change versus the largest change.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/compareLogFoldChange-1}
+
+}
+
+\caption{Comparison of the different methods of calculating log-fold change per gene}\label{fig:compareLogFoldChange}
+\end{figure}
+
+With a single measure of log-fold change and the p-value, we can now look
+at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
+of the control to the Kawasaki strain, using the ``timecourse'' method of calculating log fold change.
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
-
-As another sanity check, \texttt{moanin} provides a simple utility function to
-visualize gene time-course data in different conditions. Here, we plot the 10
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/volcano-1}
+
+}
+
+\caption{Volcano plot for the comparison of the Kawasaki strain to control, where fold-change is calculated with the timecourse method}\label{fig:volcano}
+\end{figure}
+
+\hypertarget{visualizing-genes-of-interest}{%
+\subsubsection{Visualizing Genes of Interest}\label{visualizing-genes-of-interest}}
+
+The package \texttt{moanin} also provides a simple utility function (\texttt{plot\_splines\_data}) to
+visualize gene time-course data from different conditions. In Figure \ref{fig:topPvalueGenes}, we plot the 10
 genes with the smallest p-values.
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-20-1} \end{center}
-
-And here we visualize the genes with the largest absolute timecourse log-fold
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topPvalueGenes-1} \end{center}
+
+For each gene, the individual data points are plotted against time and color coded by their condition.
+Further, a fitted spline function for each group is plotted to aid in comparing global trends across conditions.
+
+Figure \ref{fig:topLogFold}, we visualize the genes with the largest absolute timecourse log-fold
 change.
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-21-1} \end{center}
-
-Thanks to those visualization, we can see that genes often follow similar
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topLogFold-1} \end{center}
+
+In examining these visualizations, we can see that genes often follow similar
 patterns of expression, although on a different scale for each gene. We can
@@
 the workflow to cluster gene expression according to their dynamical response
-to the treatment. Before clustering the genes, we first reduce the set of
+to the treatment.
+
+\textbf{Filtering} Before clustering the genes, we first reduce the set of
 genes of interest to genes that are (1) found significantly differentially
-expressed; (2) a large-fold change between conditions. To do this, we first
+expressed; (2) a large-fold change between conditions. Reducing the set of genes on which to perform the clustering allows to
+estimate the centroids of the clusters with more stability.
+
+To do this, we first
 aggregate all p-values obtained during the time-course differential expresison
-step in a single p-value using Fisher's method \citep{fisher:statistical}. Then we
-select all the genes which have a Fisher adjusted p-value below 0.05 and a log
-fold change of at least two between at least one condition and one time-point.
-Reducing the set of genes on which to perform the clustering allows to
-estimate the centroids with more stability.
+step in a single p-value using Fisher's method \citep{fisher:statistical} (\texttt{pvalues\_fisher\_method}).
+Next we
+select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold
+change of at least two for at least one condition and one time-point.
 
@@
 
+\hypertarget{clustering-based-on-spline-fits}{%
+\subsubsection{Clustering Based on Spline Fits}\label{clustering-based-on-spline-fits}}
+
 After filtering, we are left with 3950 genes. We can then apply a
-clustering. As observed by looking at genes found differentially expressed,
+clustering routine. As observed by looking at genes found differentially expressed,
 many genes share a similar gene expression pattern, but on different scale.
+
 We thus propose the following adaptation of k-means:
@@
   \textbf{Splines estimation}: for each gene, fit the splines function with the basis
-  of your choice (provided in the \texttt{moanin\_model} object).
+  of your choice (as contained in the \texttt{moanin\_model} object).
 \item
@@
   cluster centroids.
-\item
-  \textbf{Assign scores and labels to all genes}: Our
-  \texttt{splines\_kmeans\_score\_and\_label} function assigns a score to all
-  gene-cluster pairs based on a goodness-of-fit measure of the gene to the
-  cluster centroid. We then ``assign'' each gene to a cluster based on its
-  score.
 \end{itemize}
 
-The first three steps are performed internally by the \texttt{splines\_kmeans}
-function. For now, we will set the number of clusters to be \(8\), though we
+The rescaling of the splines is similar to the rescaling genes by their standard deviation, which is common for gene expression studies without a time component.
+
+These clustering steps are performed by the \texttt{splines\_kmeans}
+function in \texttt{moanin}. For now, we will set the number of clusters to be \(8\), though we
 will return to the question of picking the best number of clusters below.
@@
 \item
-  \texttt{clusters}: a vector of size \texttt{n\_genes}, containing the
+  \texttt{clusters}: a vector of size \texttt{n\_genes}, containing the cluster assignments given by the \texttt{kmeans} step of each gene
 \end{itemize}
 
-We then use the \texttt{plot\_splines\_data} function to visualize the centroids of
+We then use the \texttt{plot\_splines\_data} function, only now applied to the centroids, to visualize the centroids of
 each cluster obtained with the splines k-means model.
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-22-1} \end{center}
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
 
 These centroids are on a 0-1 scale, because of our rescaling of the spline
-fits, and do not represent the actual gene expression level. We can plot a few
-of the actual genes in a cluster to see the difference:
+fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
+of the actual genes assigned to cluster 2 to see the difference:
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
-
-The spline plots are fitted per gene in the above plot, and illustrate that
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}
+
+The spline plots are fitted per gene in the above plot and illustrate that
 the genes assigned to a cluster have differing degrees to which their data
@@
 
-Since we arbitrarily chose a subset of genes based on our filter, we would
-like all genes to have a chance of assignment to a cluster, as well as a score
-for whether it is a good assignment. The scoring and label step allows us to
-assign genes to clusters that were removed during the filtering step, yet are
-good matches to the centroids found during the clustering.The score
+Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
+like to have a mechanism to assign all genes to a cluster.
+
+Thus, the next step in the clustering portion of the workflow is a scoring and label step. Each gene is given a score that
 corresponds to a goodness-of-fit between each gene and each cluster, computed
@@
 
+The score then allows us to assign all genes to a cluster (i.e.~a label) based on the cluster for which they have the best score, regardless of whether the gene was used in the clustering procedure.
+
 The scoring and labeling is done via the \texttt{splines\_kmeans\_score\_and\_label}
 function. This function calculates the goodness of fit of the gene to the
-cluster centroid. By default, the function labels only the 50\% of genes with
-the best scores, with the remaining genes getting \texttt{NA} as their assignment.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#<c2><a0>Then assign scores and labels to all the data, using a goodness-of-fit}
+cluster centroid and gives a cluster label to the gene if they have a sufficiently high score, as we explain above.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# Then assign scores and labels to all the data, using a goodness-of-fit}
 \CommentTok{# scoring function.}
@@
 \begin{verbatim}
-- `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
+- `scores`: the matrix of shape `n_cluster` x `n_genes`,
   containing for each gene and each cluster, the goodness of fit as
   described above.
-- `scores_and_labels$labels`: the labels for all of the genes with a
-  good-enough goodness-of-fit score.
+- `labels`: the labels for all of the genes with a
+  sufficiently good goodness-of-fit score.
 \end{verbatim}
 
-First, let us visualize the distribution of goodness-of-fit scores for each
-cluster. Here, we display the whole set of scores assigned to each cluster,
-prior to filtering poor-quality fits.
+\textbf{Assigning cluster labels:} We could just assign each gene to a cluster based on which cluster gave the minimum score.
+By default, \texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a sufficiently low enough goodness-of-fit score.
+The criteria for being ``sufficiently low'' is based on looking at the distribution of the scores of all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by \texttt{splines\_kmeans\_score\_and\_label}).
+A gene is then assigned to a cluster only if their best score is above the 50\% percentile of that distribution, with the remaining genes getting \texttt{NA} as their assignment.
+Note that because the distribution of scores of genes across \emph{all} clusters are used, this is not equivalent to assigning 50\% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
+
+\CommentTok{# Let's keep only the list of genes that have a label.}
+\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
+
+\CommentTok{# And also keep track of all the genes in cluster 2.}
+\NormalTok{genes_in_cluster2 =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
+\end{Highlighting}
+\end{Shaded}
+
+After running \texttt{scores\_and\_labels} on our data, we now have 19772 genes that
+are assigned to a cluster.
+
+We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
+cluster if we did not have the filtering cutoff of \texttt{splines\_kmeans\_score\_and\_label}, i.e.~if we simply assign every gene to a cluster based on which gives them their minimum score.
+We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
 
@@
 \CommentTok{# Get the best score and best label for all of the genes}
+\CommentTok{# This is different from output of `splines_kmeans_score_and_label`, }
+\CommentTok{# which filters the genes}
 \NormalTok{scores =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/visualize_scores-1} \end{center}
-
-Note that for some of the genes, the best scores is 1. A score of 1 indicates
-that there is no correlation between the gene and the cluster: those genes fit
-poorly to all clusters and the assignment to one cluster is done randomly..
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/visualizeScores-1}
+
+}
+
+\caption{Distribution of goodness-of-fit scores for each cluster.
+The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled.}\label{fig:visualizeScores}
+\end{figure}
+
+Note that for some of the genes, their scores in all clusters is 1. Such genes fit
+poorly to all clusters and the assignment we did above to one cluster is done arbitrarily to the first matching cluster.
 This underlines the importance of filtering genes that fit poorly to clusters.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{labels =}\StringTok{ }\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{labels}
-
-\CommentTok{# Let's keep only the list of genes that have a label.}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels[}\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(labels)])}
-
-\CommentTok{# And also keep track of all the genes in cluster 2.}
-\NormalTok{genes_in_cluster2 =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels[labels}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\end{Highlighting}
-\end{Shaded}
-
-After running \texttt{scores\_and\_labels} we now have 19772 genes that
-are assigned to a cluster.
-
-Before performing the next steps, let us investigate in more detail the
-differences between the labels provided by the splines k-means model and the
-scoring and labeling step.
-
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-24-1} \end{center}
+\textbf{EAP} In fact, \texttt{which.min} which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
+
+We can also investigate the
+differences between the labels provided by the splines k-means model and our
+scoring and labeling step, in terms of the number of genes assigned to each cluster (Figure \ref{fig:kmeansVsScore}.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/kmeansVsScore-1}
+
+}
+
+\caption{Number of genes assigned to each cluster based on kmeans criteria of nearest centroid (top) versus our goodness-of-fit filtering strategy (bottom)}\label{fig:kmeansVsScore}
+\end{figure}
 
 Of the original 3950 genes used in the
-clustering, 3467 are still assigned to a cluster based on their goodness of
+clustering, 3950 are still assigned to a cluster based on their goodness of
 fit score. We can compare whether they are still assigned to the same clusters
-based on a confusion matrix shown below. The kmeans cluster assignments are
+based on a confusion matrix shown below (Figure \ref{fig:confusionMatrix}). The kmeans cluster assignments are
 shown on the rows, and the goodness-of-fit assignments in the columns, with
@@
 
-\begin{figure}
-
-{\centering \includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-25-1}
-
-}
-
-\caption{Confusion matrix between the two sets of labels}\label{fig:unnamed-chunk-25}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/confusionMatrix-1}
+
+}
+
+\caption{Confusion matrix between the two sets of labels}\label{fig:confusionMatrix}
 \end{figure}
@@
 Of the four genes we plotted before, 3 of them remained assigned to cluster 2. We can again look at a
-few genes in cluster 2, only now pick the best scoring genes
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# order them by their score}
-\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[genes_in_cluster2, cluster_of_interest])]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
-\NormalTok{    data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}
+few genes in cluster 2, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# order them by their score in cluster}
+\NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[genes_in_cluster2,}
+\NormalTok{                     cluster_of_interest])}
+\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}
     \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-26-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topScoreInCluster2-1}
+
+}
+
+\caption{Top scoring genes in Cluster 2}\label{fig:topScoreInCluster2}
+\end{figure}
 
@@
 
-Now we will create the heatmaps of these genes (4332 genes).
+Now we will create the heatmaps of these genes (4332 genes, Figure \ref{fig:heatmapCluster8}).
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-29-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/heatmapCluster8-1}
+
+}
+
+\caption{Heatmap of genes in Cluster 8}\label{fig:heatmapCluster8}
+\end{figure}
+
+\textbf{EAP:} We shouldn't use any hidden functions (i.e.~that use \texttt{:::} as in \texttt{moanin:::rescale\_values}). We should make it a exported function if use it in workflow.
 
@@
 
-First, let us run the clustering for all possible clusters of interest. We
+First, let us run the clustering for all values of \(k\) of interest. We
 will, for each clustering, conserve (1) with within cluster sum of
 squares; (2) the clustering assignement (or label) for each gene.
+
+\textbf{EAP} This didn't make sense to me: ``run the clustering for all possible clusters of interest.'' Do you mean across different values of \(k\)? That's what I changed it to.
 
@@
 
-We plot the WCSS for \(k=2-20\) here. We see that as expected the WCSS continues
+We plot the WCSS for \(k=2-20\) here (Figure \ref{fig:elbow_method}).
+We see that as expected the WCSS continues
 to drop, but there is no clear drop in the decrease, except for very small
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/elbow_method-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/elbowMethod-1}
+
+}
+
+\caption{Plot of WCSS as a function of k}\label{fig:elbowMethod}
+\end{figure}
 
@@
 The silhouette value is a measure of how similar a data point is to its own
-cluster (cohesion) compared to other clusters (separation).
+cluster (cohesion) compared to other clusters (separation), shown in Figure \ref{fig:average_silhouette_method}.
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/average_silhouette_method-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/averageSilhouetteMethod-1}
+
+}
+
+\caption{Plot of silhouette values as a function of k}\label{fig:averageSilhouetteMethod}
+\end{figure}
 
@@
 \NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
+
 \NormalTok{subsampled_y =}\StringTok{ }\NormalTok{y[indices, ]}
@@
 
-We run bootstrap method on all genes differentially expressed and with a
+We run the bootstrap method on all genes differentially expressed and with a
 log-fold-change higher than 2 (computed with the \texttt{lfc\_max} method), and do it
 \(B=20\) times for each of \(k=2-20\). We show the code below, but because of the
-time it takes we have evaluate these values separately and provided the
-results for users to explore more quickly.
+time the computations take, we have evaluate these values separately and provided the
+results in separate files (one per \(k\)) for users to explore more quickly.
+
+\textbf{EAP}: Why don't you use the same rule/code as above?
+i.e.~with the fisher p-values etc.?
+A little strange to have two different filtering rules for the clustering.
+Or is this the same, and you're restating the rule?
+(Unclear why you'd have to redo everything here\ldots{}didn't go back and compare code)
 
@@
 
+\textbf{EAP} It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?
+
 \begin{Shaded}
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-1}
+
+}
+
+\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus1}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-34-2} \end{center}
-
-We can see that the choice of \(K=5\) seems much more stable across resampling runs than that of \(K=20\).
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-2}
+
+}
+
+\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus2}
+\end{figure}
+
+We can see (Figure \ref{fig:bootstrapConsensus}) that the choice of \(K=5\) seems much more stable across resampling runs than that of \(K=20\).
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-36-1} \end{center}
-
-From this plot, we can deduce that \(k=5\) is more stable than \(k=3\) and \(k=4\),
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/plotModelExplorer-1}
+
+}
+
+\caption{Plot of the cumulative density of similarity of clusterings of data for different k}\label{fig:plotModelExplorer}
+\end{figure}
+
+\textbf{EAP} I'm confused by this plot.
+Isn't the numbering here for different values of \(k\)?
+Why do the labels say ``B1'' and ``B2''?
+Isn't that the notation above for the bootstrap replications for each values of \(k\)?
+It seems like the names of \texttt{all\_labels} should be something like \texttt{k=1},\ldots{}, while the names of \texttt{all\_labels{[}{[}1{]}{]}} should be \texttt{B1,B2,...}.
+But currently they have the same values.
+
+From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\) is more stable than \(k=3\) and \(k=4\),
 but not as stable as \(k=2\). The model explorer strategy, in addition to
@@
 \(k=6\), \(k=7\), \(k=8\), \(k=9\) and \(k=10\) so that we can see more clearly the stability
-measures in that range.
+measures in that range (Figure \ref{fig:plotModelExplorerZoom}).
 
@@
 
-\begin{center}\includegraphics{reports/manuscript_files/figure-latex/unnamed-chunk-37-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/plotModelExplorerZoom-1}
+
+}
+
+\caption{Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10}\label{fig:plotModelExplorerZoom}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=6.67in]{images/clustering_CDF_consensus} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{images/clustering_CDF_consensus}
+
+}
+
+\caption{Plot of the Consensus Clustering}\label{fig:plotCdfConsensus}
+\end{figure}
 
@@
 
+Here we calculate the AUC (Figure \ref{fig:plotAUC}).
+
 \begin{Shaded}
@@
 
-\begin{center}\includegraphics[width=6.67in]{images/clustering_AUC_consensus} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{images/clustering_AUC_consensus}
+
+}
+
+\caption{Plot of AUC vs number of clusters}\label{fig:plotAUC}
+\end{figure}
+
+While here we calculate the improvement in the AUC (Figure \ref{fig:plotDeltaAUC}).
 
@@
 
-\begin{center}\includegraphics[width=6.67in]{images/clustering_delta_AUC_consensus} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{images/clustering_delta_AUC_consensus}
+
+}
+
+\caption{Plot of improvement in AUC vs number of clusters}\label{fig:plotDeltaAUC}
+\end{figure}
 
@@
 Once good clusters are obtained, the next step is to leverage the clustering
-to ease interpretation. Classic enrichment analysis step can then be performed
-in the gene set defined in each cluster: KEGG pathway enrichment analysis, GO
-term enrichment analysis, motif enrichment analysis, etc.
+to ease interpretation. Classic enrichment analysis can then be performed
+on the gene set defined by each cluster. Examples include KEGG pathway enrichment analysis,
+GO term enrichment analysis, and motif enrichment analysis.
 
 First, let us clean up the genes we work with and only select the genes we are
-going to use in the enrichment analysis. One can either use the whole set of
-genes, only the set of differentially expressed genes in each cluster, or a
+going to use in the enrichment analysis. One can either all genes assigned to the cluster, only the set of differentially expressed genes in each cluster, or the
 subset of genes that fit well to a cluster (based on some criterion).
@@
 for this step. \texttt{KEGGprofile} is a
-package that easily allows to perform pathway enrichment analysis on a set of
-genes labeled with the ensembl annotation. We thus need to convert the gene
-names into the appropriate format. This is where \texttt{biomaRt} comes in handy: it
-enables easy conversion from one gene annotation to another. Here, we will use
-it to convert the gene names from the Refseq annotation to the ensembl one.
-
-\begin{Shaded}
-\begin{Highlighting}[]
+package that facilitates enrichment analysis on a set of
+genes labeled with the Ensembl annotation \citep{ensembl2020} based on the set of biological pathways described in the KEGG database\citep{KEGG}.
+
+We thus need to convert the gene
+names, which are given in the Refseq annotation, into the corresponding Ensembl name. This is where \texttt{biomaRt} comes in handy: it
+enables easy conversion from one gene annotation to another. Here, we will use the function \texttt{getBM}
+in \texttt{biomaRt} to convert the gene names from cluster 8.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Set up right ensembl database}
 \NormalTok{ensembl =}\StringTok{ }\KeywordTok{useMart}\NormalTok{(}\StringTok{"ensembl"}\NormalTok{)}
 \NormalTok{ensembl =}\StringTok{ }\KeywordTok{useDataset}\NormalTok{(}\StringTok{"mmusculus_gene_ensembl"}\NormalTok{, }\DataTypeTok{mart=}\NormalTok{ensembl)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+
+\CommentTok{## Get gene names of genes in Cluster 8}
 \NormalTok{cluster =}\StringTok{ }\DecValTok{8}
@@
 
-\CommentTok{# convert gene names}
+\CommentTok{## Convert gene names}
 \NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene_id"}\NormalTok{),}
@@
         \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene_id"}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+Then we use the function \texttt{find\_enriched\_pathway} in the \texttt{KEGGprofile} package to determine whether any KEGG pathways are enriched in cluster 8, i.e.~whether a higher percentage of genes from a single pathway are found in cluster 8 than we would expect by simply proportional assignment of genes to clusters.
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
@@
 
+\textbf{EAP: I changed this to be the min(10, nrow(pathways)) but wondering if there's a problem, since presumably there used to be more than 10 pathways found significant and on my machine there are only 2?}
+
 \hypertarget{finding-enriched-go-terms}{%
@@
 
-To find GO terms, we use \texttt{biomaRt} to find the mapping between GO terms and
-gene mapping. The GO enrichment library \texttt{topGO} \citep{alexa:topgo} expects the GO term to gene
+The Gene Ontology (GO) database \citep{GO}, also categorizes genes into meaningful biological ontologies and can be used for enrichment analysis via the package \texttt{topGO} \citep{alexa:topgo}.
+We again use \texttt{biomaRt} to find the mapping between genes and the GO terms to which they match.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"go_id"}\NormalTok{, }\StringTok{"refseq_mrna"}\NormalTok{),}
+          \DataTypeTok{values=}\NormalTok{gene_names,}
+          \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{,}
+          \DataTypeTok{mart=}\NormalTok{ensembl)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{biomaRt} query results in a matrix with two columns: gene names and GO
+term ID. The package \texttt{topGO} \citep{alexa:topgo} expects the GO term to gene
 mapping to be a list where each item is a mapping between a gene name and a GO
-term ID vector.
+term ID vector, for example:
 
@@
 
-\texttt{biomaRt} queries results a matrix with two named columns of gene names and GO
-term ID.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"go_id"}\NormalTok{, }\StringTok{"refseq_mrna"}\NormalTok{),}
-          \DataTypeTok{values=}\NormalTok{gene_names,}
-          \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{,}
-          \DataTypeTok{mart=}\NormalTok{ensembl)}
-
+\texttt{moanin} provides a simple function (\texttt{create\_go\_term\_mapping}) to make this conversion:
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{# Create gene to GO id mapping}
@@
 
-Once the gene ID to GO mapping list is created, \texttt{moanin} provides a simple
-interface to \texttt{topGO} to fetch enriched GO terms. Here, we show an example of
-running a GOterm enrichment on the ``Biological process'' ontology (BP).
-
-\begin{Shaded}
-\begin{Highlighting}[]
+Once the gene ID to GO mapping list is created, \texttt{moanin} provides an
+interface to \texttt{topGO} to determine enriched GO terms. \textbf{EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe} Here, we show an example of
+running a GO term enrichment on the ``Biological process'' ontology (BP).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# Create logical vector of whether in cluster 8}
 \NormalTok{assignments =}\StringTok{ }\NormalTok{labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster}
@@
 
-This workflow provides a tutoral for the analysis of time-course gene
-expression data in R, illustrated through the analysis of mice lung tissue
-exposed to different influenza strain. It covers four main steps; (1) quality
-control and normalization; (2) differential expression analysis; (3)
-clustering of time-course gene expression data; (4) downstream analysis of
-clusters.
+This workflow provides a tutoral for the analysis of lengthy time-course gene
+expression data in R using the package \texttt{moanin}, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
+exposed to different influenza strain and measured over time.
+The proposed workflow covers consists of three common analysis main steps generally performed after quality
+control and normalization: (1) differential expression analysis; (2)
+clustering of time-course gene expression data; (3) downstream analysis of
+clusters. We demonstrate how the use of the package \texttt{moanin} allows for easy implementation of these procedures in the setting of time-course data.
 

### 2020-06-17 scripts/manuscript.Rmd
ENH updated affiliation + fig1 + references

@@
     - name: Nelle Varoquaux
-      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
+      affiliation: CNRS, University Grenoble Alpes, TIMC-IMAG, Grenoble, France
     - name: Elizabeth Purdom
@@
 - Check that clustering can deal with count data.
-- Update Nelle's affiliation!
+- Make sure package is compatible bioconductor
 
@@
 library(biomaRt)
+library(KEGGprofile)
+library(BiocWorkflowTools)
 
@@
 
-**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at the beginning of the document? I also needed to install `BiocWorkflowTools`, `KEGGprofile` to run the .Rmd
+**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at
+the beginning of the document? I also needed to install `BiocWorkflowTools`,
+`KEGGprofile` to run the .Rmd
 
@@
 normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file.
-**EAP**: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol] [@park:evalution].
 

### 2020-06-17 scripts/manuscript.Rmd
FIX citation

@@
 normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol] [@park:evalution].
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol; @park:evalution].
 

### 2020-06-17 scripts/manuscript.Rmd
Adding a comment

@@
 ```
-**EAP**: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like.
-I like to reserve table format for actual results, rather than code-like input
 
@@
 **EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
+**NV** Fair point… maybe this should be fixed somehow by adding a random
+perturbation to the scores when the scores are equal? Although I think the
+text is still appropriate: those genes should definitely be filtered out.
 

### 2020-06-17 scripts/manuscript.Rmd
Continuing some small fixes here and there

@@
 aheatmap(
-    moanin:::rescale_values(data_to_plot)[res$rowInd,],
+    moanin::rescale_values(data_to_plot)[res$rowInd,],
     Colv=NA,
@@
 ```
-
-**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.
 
@@
 
-**EAP** This didn't make sense to me: "run the clustering for all possible clusters of interest." Do you mean across different values of $k$? That's what I changed it to.
-
 Below we run the clustering for $k$ equal to $2-20$. The `splines_kmeans`
@@
 
-**EAP**: Why don't you use the same rule/code as above?
-i.e.
-with the fisher p-values etc.?
-A little strange to have two different filtering rules for the clustering.
-Or is this the same, and you're restating the rule?
-(Unclear why you'd have to redo everything here...didn't go back and compare code)
-
 ```{r bootstrapped_clustering, eval=FALSE}
-pvalues = de_analysis[, pval_col_to_keep]
-
-splines_model = moanin::create_moanin_model(meta, degrees_of_freedom=6)
-fishers_pval = moanin:::pvalues_fisher_method(pvalues)
-fishers_qval = stats::p.adjust(fishers_pval)
-
-# Keep only the genes found differentially expressed after combining the
-# p-values using Fisher's method and with a log fold change greater than 2
-genes_to_keep = row.names(
-    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
-y = as.matrix(data[genes_to_keep, ])
-
 # You may want to set the random seed of the experiment so that the results
@@
 calculate proportion of times each pair of samples was clustered together
-across the 20 resampling runs, and plot the heatmap of those consensus
+across the 25 resampling runs, and plot the heatmap of those consensus
 matrices for $k=5$ and $k=20$.
-
-**EAP** It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?
 
@@
     stability = read.table(filename, sep="\t")
-    all_labels[[paste0("B", i)]] = stability
+    all_labels[[paste0("k=", i)]] = stability
 }
@@
 ```
-**EAP** I'm confused by this plot.
-Isn't the numbering here for different values of $k$?
-Why do the labels say "B1" and "B2"?
-Isn't that the notation above for the bootstrap replications for each values of $k$?
-It seems like the names of `all_labels` should be something like `k=1`,..., while the names of `all_labels[[1]]` should be `B1,B2,...`.
-But currently they have the same values.
 
@@
 ```{r  plotModelExplorerZoom, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10"}
-clusters = c("B6", "B7", "B8", "B9", "B10")
+clusters = c("k=6", "k=7", "k=8", "k=9", "k=10")
 selected_labels = all_labels[clusters]

### 2020-06-17 scripts/manuscript.Rmd
More small changes

@@
 
-Once the gene ID to GO mapping list is created, `moanin` provides an 
-interface to `topGO` to determine enriched GO terms. **EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe** Here, we show an example of
-running a GO term enrichment on the "Biological process" ontology (BP).
+Once the gene ID to GO mapping list is created, `moanin` provides an interface
+to `topGO` to determine enriched GO terms. In particular, it performs a
+p-value correction and only returns the significant GO-term enrichment in an
+easy to use `data.frame` object. Here, we show an example of running a GO term
+enrichment on the "Biological process" ontology (BP).
 

### 2020-06-17 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-06-17 05:37:22.
+The manuscript was last compile on 2020-06-17 08:56:13.
 
@@
 \item
-  Update Nelle's affiliation!
+  Make sure package is compatible bioconductor
 \end{itemize}
@@
 \KeywordTok{library}\NormalTok{(biomaRt)}
+\KeywordTok{library}\NormalTok{(KEGGprofile)}
+\KeywordTok{library}\NormalTok{(BiocWorkflowTools)}
 
@@
 
-\textbf{EAP}: Why is this list not the same as the (hidden via echo=FALSE) list at the beginning of the document?
-I also needed to install \texttt{BiocWorkflowTools}, \texttt{KEGGprofile} to run the .Rmd
+\textbf{EAP}: Why is this list not the same as the (hidden via echo=FALSE) list at
+the beginning of the document? I also needed to install \texttt{BiocWorkflowTools},
+\texttt{KEGGprofile} to run the .Rmd
 
@@
 normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file.
-\textbf{EAP}: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file \citep{abrams:protocol, park:evalution}.
 
@@
 \end{verbatim}
-
-\textbf{EAP}: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like.
-I like to reserve table format for actual results, rather than code-like input
 
@@
 \textbf{EAP} In fact, \texttt{which.min} which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
+\textbf{NV} Fair point\textless{}80\textgreater{} maybe this should be fixed somehow by adding a random
+perturbation to the scores when the scores are equal? Although I think the
+text is still appropriate: those genes should definitely be filtered out.
 
@@
 \KeywordTok{aheatmap}\NormalTok{(}
-\NormalTok{    moanin}\OperatorTok{:::}\KeywordTok{rescale_values}\NormalTok{(data_to_plot)[res}\OperatorTok{$}\NormalTok{rowInd,],}
+\NormalTok{    moanin}\OperatorTok{::}\KeywordTok{rescale_values}\NormalTok{(data_to_plot)[res}\OperatorTok{$}\NormalTok{rowInd,],}
     \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
@@
 \end{figure}
-
-\textbf{EAP:} We shouldn't use any hidden functions (i.e.~that use \texttt{:::} as in \texttt{moanin:::rescale\_values}). We should make it a exported function if use it in workflow.
 
@@
 
-\textbf{EAP} This didn't make sense to me: ``run the clustering for all possible clusters of interest.'' Do you mean across different values of \(k\)? That's what I changed it to.
-
 Below we run the clustering for \(k\) equal to \(2-20\). The \texttt{splines\_kmeans}
@@
 
-\textbf{EAP}: Why don't you use the same rule/code as above?
-i.e.~with the fisher p-values etc.?
-A little strange to have two different filtering rules for the clustering.
-Or is this the same, and you're restating the rule?
-(Unclear why you'd have to redo everything here\ldots{}didn't go back and compare code)
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{pvalues =}\StringTok{ }\NormalTok{de_analysis[, pval_col_to_keep]}
-
-\NormalTok{splines_model =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
-\NormalTok{fishers_pval =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
-\NormalTok{fishers_qval =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}
-
-\CommentTok{# Keep only the genes found differentially expressed after combining the}
-\CommentTok{# p-values using Fisher's method and with a log fold change greater than 2}
-\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
-\NormalTok{    lfc_max[(}\KeywordTok{rowSums}\NormalTok{(lfc_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
-\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}
-
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{# You may want to set the random seed of the experiment so that the results}
@@
 calculate proportion of times each pair of samples was clustered together
-across the 20 resampling runs, and plot the heatmap of those consensus
+across the 25 resampling runs, and plot the heatmap of those consensus
 matrices for \(k=5\) and \(k=20\).
-
-\textbf{EAP} It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?
 
@@
 
-\textbf{EAP} I'm confused by this plot.
-Isn't the numbering here for different values of \(k\)?
-Why do the labels say ``B1'' and ``B2''?
-Isn't that the notation above for the bootstrap replications for each values of \(k\)?
-It seems like the names of \texttt{all\_labels} should be something like \texttt{k=1},\ldots{}, while the names of \texttt{all\_labels{[}{[}1{]}{]}} should be \texttt{B1,B2,...}.
-But currently they have the same values.
-
 From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\) is more stable than \(k=3\) and \(k=4\),
@@
 \begin{Highlighting}[]
-\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"B6"}\NormalTok{, }\StringTok{"B7"}\NormalTok{, }\StringTok{"B8"}\NormalTok{, }\StringTok{"B9"}\NormalTok{, }\StringTok{"B10"}\NormalTok{)}
+\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"k=6"}\NormalTok{, }\StringTok{"k=7"}\NormalTok{, }\StringTok{"k=8"}\NormalTok{, }\StringTok{"k=9"}\NormalTok{, }\StringTok{"k=10"}\NormalTok{)}
 \NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
@@
 
-Once the gene ID to GO mapping list is created, \texttt{moanin} provides an
-interface to \texttt{topGO} to determine enriched GO terms. \textbf{EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe} Here, we show an example of
-running a GO term enrichment on the ``Biological process'' ontology (BP).
+Once the gene ID to GO mapping list is created, \texttt{moanin} provides an interface
+to \texttt{topGO} to determine enriched GO terms. In particular, it performs a
+p-value correction and only returns the significant GO-term enrichment in an
+easy to use \texttt{data.frame} object. Here, we show an example of running a GO term
+enrichment on the ``Biological process'' ontology (BP).
 
@@
 ## other attached packages:
-##  [1] RColorBrewer_1.1-2   moanin_0.0.0         ggfortify_0.4.6     
-##  [4] topGO_2.32.0         SparseM_1.77         GO.db_3.6.0         
-##  [7] AnnotationDbi_1.42.1 IRanges_2.14.12      S4Vectors_0.18.3    
-## [10] graph_1.58.2         viridis_0.5.1        viridisLite_0.3.0   
-## [13] KEGGprofile_1.22.0   RCurl_1.95-4.12      bitops_1.0-6        
-## [16] NMF_0.21.0           Biobase_2.40.0       BiocGenerics_0.26.0 
-## [19] cluster_2.0.9        rngtools_1.3.1.1     pkgmaker_0.27       
-## [22] registry_0.5-1       ggplot2_3.1.1        kableExtra_1.1.0    
-## [25] biomaRt_2.36.1       BiocStyle_2.8.2      knitr_1.22          
-## [28] limma_3.36.5         usethis_1.5.0        devtools_2.0.2      
-## [31] rmarkdown_1.12      
+##  [1] RColorBrewer_1.1-2      moanin_0.0.0           
+##  [3] BiocWorkflowTools_1.6.2 ggfortify_0.4.6        
+##  [5] topGO_2.32.0            SparseM_1.77           
+##  [7] GO.db_3.6.0             AnnotationDbi_1.42.1   
+##  [9] IRanges_2.14.12         S4Vectors_0.18.3       
+## [11] graph_1.58.2            viridis_0.5.1          
+## [13] viridisLite_0.3.0       KEGGprofile_1.22.0     
+## [15] RCurl_1.95-4.12         bitops_1.0-6           
+## [17] NMF_0.21.0              Biobase_2.40.0         
+## [19] BiocGenerics_0.26.0     cluster_2.0.9          
+## [21] rngtools_1.3.1.1        pkgmaker_0.27          
+## [23] registry_0.5-1          ggplot2_3.1.1          
+## [25] kableExtra_1.1.0        biomaRt_2.36.1         
+## [27] BiocStyle_2.8.2         knitr_1.22             
+## [29] limma_3.36.5            usethis_1.5.0          
+## [31] devtools_2.0.2          rmarkdown_1.12         
 ## 
 ## loaded via a namespace (and not attached):
-##  [1] colorspace_1.4-1        rprojroot_1.3-2        
-##  [3] XVector_0.20.0          fs_1.3.1               
-##  [5] rstudioapi_0.10         remotes_2.0.4          
-##  [7] bit64_0.9-7             ClusterR_1.1.9         
-##  [9] KEGG.db_3.2.3           xml2_1.2.0             
-## [11] codetools_0.2-16        doParallel_1.0.14      
-## [13] pkgload_1.0.2           ade4_1.7-13            
-## [15] gridBase_0.4-7          png_0.1-7              
-## [17] FD_1.0-12               readr_1.3.1            
-## [19] compiler_3.6.0          httr_1.4.0             
-## [21] backports_1.1.4         Matrix_1.2-17          
-## [23] assertthat_0.2.1        lazyeval_0.2.2         
-## [25] cli_1.1.0               htmltools_0.3.6        
-## [27] prettyunits_1.0.2       tools_3.6.0            
-## [29] gmp_0.5-13.5            gtable_0.3.0           
-## [31] glue_1.3.1              reshape2_1.4.3         
-## [33] dplyr_0.8.0.1           BiocWorkflowTools_1.6.2
-## [35] Rcpp_1.0.1              Biostrings_2.48.0      
-## [37] NMI_2.0                 ape_5.3                
-## [39] nlme_3.1-139            iterators_1.0.10       
-## [41] xfun_0.6                stringr_1.4.0          
-## [43] ps_1.3.0                testthat_2.1.1         
-## [45] rvest_0.3.3             gtools_3.8.1           
-## [47] XML_3.98-1.19           zlibbioc_1.26.0        
-## [49] MASS_7.3-51.4           scales_1.0.0           
-## [51] hms_0.4.2               curl_3.3               
-## [53] yaml_2.2.0              memoise_1.1.0          
-## [55] gridExtra_2.3           TeachingDemos_2.10     
-## [57] stringi_1.4.3           RSQLite_2.1.1          
-## [59] desc_1.2.0              foreach_1.4.4          
-## [61] permute_0.9-5           pkgbuild_1.0.3         
-## [63] bibtex_0.4.2            geometry_0.4.1         
-## [65] rlang_0.4.6             pkgconfig_2.0.2        
-## [67] matrixStats_0.54.0      evaluate_0.13          
-## [69] lattice_0.20-38         purrr_0.3.2            
-## [71] bit_1.1-14              processx_3.3.1         
-## [73] tidyselect_0.2.5        plyr_1.8.4             
-## [75] magrittr_1.5            bookdown_0.9           
-## [77] R6_2.4.0                DBI_1.0.0              
-## [79] mgcv_1.8-28             pillar_1.3.1           
-## [81] withr_2.1.2             abind_1.4-5            
-## [83] KEGGREST_1.20.2         tibble_2.1.1           
-## [85] crayon_1.3.4            progress_1.2.0         
-## [87] grid_3.6.0              blob_1.1.1             
-## [89] callr_3.2.0             git2r_0.25.2           
-## [91] vegan_2.5-4             digest_0.6.18          
-## [93] webshot_0.5.1           xtable_1.8-4           
-## [95] tidyr_0.8.3             munsell_0.5.0          
-## [97] magic_1.5-9             sessioninfo_1.1.1
+##  [1] colorspace_1.4-1   rprojroot_1.3-2    XVector_0.20.0    
+##  [4] fs_1.3.1           rstudioapi_0.10    remotes_2.0.4     
+##  [7] bit64_0.9-7        ClusterR_1.1.9     KEGG.db_3.2.3     
+## [10] xml2_1.2.0         codetools_0.2-16   doParallel_1.0.14 
+## [13] pkgload_1.0.2      ade4_1.7-13        gridBase_0.4-7    
+## [16] png_0.1-7          FD_1.0-12          readr_1.3.1       
+## [19] compiler_3.6.0     httr_1.4.0         backports_1.1.4   
+## [22] Matrix_1.2-17      assertthat_0.2.1   lazyeval_0.2.2    
+## [25] cli_1.1.0          htmltools_0.3.6    prettyunits_1.0.2 
+## [28] tools_3.6.0        gmp_0.5-13.5       gtable_0.3.0      
+## [31] glue_1.3.1         reshape2_1.4.3     dplyr_0.8.0.1     
+## [34] Rcpp_1.0.1         Biostrings_2.48.0  NMI_2.0           
+## [37] ape_5.3            nlme_3.1-139       iterators_1.0.10  
+## [40] xfun_0.6           stringr_1.4.0      ps_1.3.0          
+## [43] testthat_2.1.1     rvest_0.3.3        gtools_3.8.1      
+## [46] XML_3.98-1.19      zlibbioc_1.26.0    MASS_7.3-51.4     
+## [49] scales_1.0.0       hms_0.4.2          curl_3.3          
+## [52] yaml_2.2.0         memoise_1.1.0      gridExtra_2.3     
+## [55] TeachingDemos_2.10 stringi_1.4.3      RSQLite_2.1.1     
+## [58] desc_1.2.0         foreach_1.4.4      permute_0.9-5     
+## [61] pkgbuild_1.0.3     bibtex_0.4.2       geometry_0.4.1    
+## [64] rlang_0.4.6        pkgconfig_2.0.2    matrixStats_0.54.0
+## [67] evaluate_0.13      lattice_0.20-38    purrr_0.3.2       
+## [70] bit_1.1-14         processx_3.3.1     tidyselect_0.2.5  
+## [73] plyr_1.8.4         magrittr_1.5       bookdown_0.9      
+## [76] R6_2.4.0           DBI_1.0.0          mgcv_1.8-28       
+## [79] pillar_1.3.1       withr_2.1.2        abind_1.4-5       
+## [82] KEGGREST_1.20.2    tibble_2.1.1       crayon_1.3.4      
+## [85] progress_1.2.0     grid_3.6.0         blob_1.1.1        
+## [88] callr_3.2.0        git2r_0.25.2       vegan_2.5-4       
+## [91] digest_0.6.18      webshot_0.5.1      xtable_1.8-4      
+## [94] tidyr_0.8.3        munsell_0.5.0      magic_1.5-9       
+## [97] sessioninfo_1.1.1
 \end{verbatim}

### 2020-06-22 scripts/manuscript.Rmd
the data cannot be part of moanin

@@
 
-First let's load the data. The package `moanin` contains the normalized
-data and metadata of [@shoemaker:ultrasensitive].
+First let's load the data. We provide the normalized
+data and metadata of [@shoemaker:ultrasensitive] for simplicity.
 
@@
 # Now load in the metadata
-data(shoemaker2015)
-meta = shoemaker2015$meta
-data = shoemaker2015$data
+meta = read.table("../data/shoemaker2015/meta.tsv")
+data = read.table(gzfile("../data/shoemaker2015/data.tsv.gz"))
 ```

### 2020-06-24 scripts/manuscript.Rmd
MAINT small edits

--- 2020-06-17 scripts/manuscript.Rmd
@@
 ---
-**EAP**: I don't know that describing normalization as part of our tutorial makes sense.
-I've tried to change the text.
 
@@
 The manuscript was last compile on `r Sys.time()`.
-
-# TODO list
-
-- Make clear that it applies to RNASeq and how to call the different functions
-    to RNASeq
-- Check that clustering can deal with count data.
-- Make sure package is compatible bioconductor
 
@@
 ```{r all_libraries}
+# From github
+library(timecoursedata)
+
 # From CRAN
@@
 The first steps of analysis of gene expression data is always to do
-normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol; @park:evalution].
+normalization and quality control checks of the data. However, in what
+follows, we do not show the steps for normalization, as these are specific to
+the platform (microarray); the code for the normalization can be found in as a
+supplemental file [@abrams:protocol; @park:evaluation].
 

### 2020-06-24 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-06-17 08:56:13.
+The manuscript was last compile on 2020-06-24 08:14:28.
 
@@
 \begin{Highlighting}[]
+\CommentTok{# From github}
+\KeywordTok{library}\NormalTok{(timecoursedata)}
+
 \CommentTok{# From CRAN}
@@
 
-\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
-\hline
-  & B1 & B2 & B3 & B4 & B5 & B6 & B7 & B8 & B9 & B10 & B11 & B12 & B13 & B14 & B15 & B16 & B17 & B18 & B19 & B20\\
-\hline
-A\_51\_P114236 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
-\hline
-A\_51\_P172409 & 1 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 4 & 3 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
-\hline
-A\_51\_P188401 & 1 & 5 & 5 & 4 & 5 & 5 & 5 & 3 & 2 & 4 & 4 & 4 & 2 & 3 & 5 & 4 & 2 & 4 & 3 & 4\\
-\hline
-A\_51\_P191274 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
-\hline
-A\_51\_P226791 & 1 & 4 & 2 & 4 & 3 & 4 & 2 & 1 & 3 & 3 & 4 & 2 & 5 & 2 & 3 & 5 & 2 & 2 & 5 & 3\\
-\hline
-A\_51\_P277048 & 1 & 1 & 4 & 5 & 5 & 2 & 5 & 5 & 2 & 4 & 3 & 4 & 2 & 5 & 4 & 3 & 4 & 4 & 3 & 5\\
+\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
+\hline
+  & B1 & B2 & B3 & B4 & B5 & B6 & B7 & B8 & B9 & B10 & B11 & B12 & B13 & B14 & B15 & B16 & B17 & B18 & B19 & B20 & B21 & B22 & B23 & B24 & B25\\
+\hline
+AA204140 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 1 & 3 & 3 & 1 & 5 & 1 & 1 & 4 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
+\hline
+AB008911 & 5 & 5 & 1 & 4 & 1 & 2 & 3 & 3 & 2 & 3 & 3 & 3 & 5 & 1 & 5 & 1 & 3 & 4 & 4 & 5 & 5 & 2 & 4 & 1 & 4\\
+\hline
+AB010313 & 3 & 3 & 1 & 4 & 1 & 2 & 4 & 3 & 2 & 5 & 3 & 1 & 5 & 3 & 3 & 2 & 3 & 5 & 4 & 2 & 5 & 3 & 3 & 4 & 2\\
+\hline
+AB010342 & 5 & 3 & 1 & 5 & 1 & 1 & 3 & 3 & 1 & 3 & 3 & 3 & 5 & 1 & 5 & 1 & 1 & 4 & 3 & 5 & 4 & 2 & 3 & 4 & 4\\
+\hline
+AB011473 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 2 & 5 & 2 & 5 & 5 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
+\hline
+AB099518 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 2 & 5 & 2 & 5 & 5 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
 \hline
@@
 ##  [3] BiocWorkflowTools_1.6.2 ggfortify_0.4.6        
-##  [5] topGO_2.32.0            SparseM_1.77           
-##  [7] GO.db_3.6.0             AnnotationDbi_1.42.1   
-##  [9] IRanges_2.14.12         S4Vectors_0.18.3       
-## [11] graph_1.58.2            viridis_0.5.1          
-## [13] viridisLite_0.3.0       KEGGprofile_1.22.0     
-## [15] RCurl_1.95-4.12         bitops_1.0-6           
-## [17] NMF_0.21.0              Biobase_2.40.0         
-## [19] BiocGenerics_0.26.0     cluster_2.0.9          
-## [21] rngtools_1.3.1.1        pkgmaker_0.27          
-## [23] registry_0.5-1          ggplot2_3.1.1          
-## [25] kableExtra_1.1.0        biomaRt_2.36.1         
-## [27] BiocStyle_2.8.2         knitr_1.22             
-## [29] limma_3.36.5            usethis_1.5.0          
-## [31] devtools_2.0.2          rmarkdown_1.12         
+##  [5] timecoursedata_0.0.0    topGO_2.32.0           
+##  [7] SparseM_1.77            GO.db_3.6.0            
+##  [9] AnnotationDbi_1.42.1    IRanges_2.14.12        
+## [11] S4Vectors_0.18.3        graph_1.58.2           
+## [13] viridis_0.5.1           viridisLite_0.3.0      
+## [15] KEGGprofile_1.22.0      RCurl_1.95-4.12        
+## [17] bitops_1.0-6            NMF_0.21.0             
+## [19] Biobase_2.40.0          BiocGenerics_0.26.0    
+## [21] cluster_2.0.9           rngtools_1.3.1.1       
+## [23] pkgmaker_0.27           registry_0.5-1         
+## [25] ggplot2_3.1.1           kableExtra_1.1.0       
+## [27] biomaRt_2.36.1          BiocStyle_2.8.2        
+## [29] knitr_1.22              limma_3.36.5           
+## [31] usethis_1.5.0           devtools_2.0.2         
+## [33] rmarkdown_1.12         
 ## 
@@
 ##  [7] bit64_0.9-7        ClusterR_1.1.9     KEGG.db_3.2.3     
-## [10] xml2_1.2.0         codetools_0.2-16   doParallel_1.0.14 
+## [10] xml2_1.3.2         codetools_0.2-16   doParallel_1.0.14 
 ## [13] pkgload_1.0.2      ade4_1.7-13        gridBase_0.4-7    
@@
 ## [82] KEGGREST_1.20.2    tibble_2.1.1       crayon_1.3.4      
-## [85] progress_1.2.0     grid_3.6.0         blob_1.1.1        
-## [88] callr_3.2.0        git2r_0.25.2       vegan_2.5-4       
+## [85] progress_1.2.0     grid_3.6.0         vegan_2.5-4       
+## [88] blob_1.1.1         callr_3.2.0        git2r_0.25.2      
 ## [91] digest_0.6.18      webshot_0.5.1      xtable_1.8-4      

### 2020-06-24 scripts/manuscript.Rmd
couple of answers

@@
 `KEGGprofile` to run the .Rmd
+**NV**: I think some of the libraries above are not needed anymore. Some are
+also only used in hidden elements. I just checked that the whole pipeline
+works, by executing all command lines the reader will see, and the libraries
+are all loaded fine. I think BiocWorkflowTools is only need in the compilation
+of the Rmd file.
 

### 2020-06-24 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\textbf{EAP}: I don't know that describing normalization as part of our tutorial makes sense.
-I've tried to change the text.
-
-The manuscript was last compile on 2020-06-24 08:14:28.
-
-\hypertarget{todo-list}{%
-\section{TODO list}\label{todo-list}}
-
-\begin{itemize}
-\tightlist
-\item
-  Make clear that it applies to RNASeq and how to call the different functions
-  to RNASeq
-\item
-  Check that clustering can deal with count data.
-\item
-  Make sure package is compatible bioconductor
-\end{itemize}
+The manuscript was last compile on 2020-06-24 09:03:07.
 
@@
 \texttt{KEGGprofile} to run the .Rmd
+\textbf{NV}: I think some of the libraries above are not needed anymore. Some are
+also only used in hidden elements. I just checked that the whole pipeline
+works, by executing all command lines the reader will see, and the libraries
+are all loaded fine. I think BiocWorkflowTools is only need in the compilation
+of the Rmd file.
 
@@
 The first steps of analysis of gene expression data is always to do
-normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file \citep{abrams:protocol, park:evalution}.
+normalization and quality control checks of the data. However, in what
+follows, we do not show the steps for normalization, as these are specific to
+the platform (microarray); the code for the normalization can be found in as a
+supplemental file \citep{abrams:protocol, park:evaluation}.
 

### 2020-07-04 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-06-24 09:03:07.
+The manuscript was last compile on 2020-07-04 10:52:02.
 
@@
 \begin{verbatim}
-## moanin_object on 209 samples containing the following information:
+## moanin_model object on 209 samples containing the following information:
 ## 1) Meta data with 5 variables:
@@
 ## To access these objects use:
-##  <object_name>$degrees_of_freedom
-##  <object_name>$basis
-##  <object_name>$meta
-##  <object_name>$formula
+##  <x_name>$degrees_of_freedom
+##  <x_name>$basis
+##  <x_name>$meta
+##  <x_name>$formula
 \end{verbatim}

### 2020-07-10 scripts/manuscript.Rmd
small change

--- 2020-06-22 scripts/manuscript.Rmd
@@
     - name: Nelle Varoquaux
-      affiliation: CNRS, University Grenoble Alpes, TIMC-IMAG, Grenoble, France
+      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
     - name: Elizabeth Purdom
@@
 - Check that clustering can deal with count data.
-- Make sure package is compatible bioconductor
+- Update Nelle's affiliation!
 
@@
 library(biomaRt)
-library(KEGGprofile)
-library(BiocWorkflowTools)
 
@@
 
-**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at
-the beginning of the document? I also needed to install `BiocWorkflowTools`,
-`KEGGprofile` to run the .Rmd
+**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at the beginning of the document? I also needed to install `BiocWorkflowTools`, `KEGGprofile` to run the .Rmd
 
@@
 
-First let's load the data. We provide the normalized
-data and metadata of [@shoemaker:ultrasensitive] for simplicity.
+First let's load the data. The package `moanin` contains the normalized
+data and metadata of [@shoemaker:ultrasensitive].
 
@@
 # Now load in the metadata
-meta = read.table("../data/shoemaker2015/meta.tsv")
-data = read.table(gzfile("../data/shoemaker2015/data.tsv.gz"))
+data(shoemaker2015)
+meta = shoemaker2015$meta
+data = shoemaker2015$data
 ```
@@
 normalization and quality control checks of the data.
-However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file [@abrams:protocol; @park:evalution].
+However, in what follows, we do not show the steps for normalization, as these are specific to the platform (microarray); the code for the normalization can be found in as a supplemental file.
+**EAP**: perhaps a citation to review papers for normalization on microarray and mRNA-Seq.
 
@@
 ```
+**EAP**: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like.
+I like to reserve table format for actual results, rather than code-like input
 
@@
 **EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
-**NV** Fair point… maybe this should be fixed somehow by adding a random
-perturbation to the scores when the scores are equal? Although I think the
-text is still appropriate: those genes should definitely be filtered out.
 
@@
 aheatmap(
-    moanin::rescale_values(data_to_plot)[res$rowInd,],
+    moanin:::rescale_values(data_to_plot)[res$rowInd,],
     Colv=NA,
@@
 ```
+
+**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.
+
+**EAPL** actually, why do we need to rescale them between 0-1 rather than just a normal call to `scale`? Would it really make a difference?
 
@@
 
+**EAP** This didn't make sense to me: "run the clustering for all possible clusters of interest." Do you mean across different values of $k$? That's what I changed it to.
+
 Below we run the clustering for $k$ equal to $2-20$. The `splines_kmeans`
@@
 
+**EAP**: Why don't you use the same rule/code as above?
+i.e.
+with the fisher p-values etc.?
+A little strange to have two different filtering rules for the clustering.
+Or is this the same, and you're restating the rule?
+(Unclear why you'd have to redo everything here...didn't go back and compare code)
+
 ```{r bootstrapped_clustering, eval=FALSE}
+pvalues = de_analysis[, pval_col_to_keep]
+
+splines_model = moanin::create_moanin_model(meta, degrees_of_freedom=6)
+fishers_pval = moanin:::pvalues_fisher_method(pvalues)
+fishers_qval = stats::p.adjust(fishers_pval)
+
+# Keep only the genes found differentially expressed after combining the
+# p-values using Fisher's method and with a log fold change greater than 2
+genes_to_keep = row.names(
+    lfc_max[(rowSums(lfc_max > 2) > 0) & (fishers_qval < 0.05), ])
+y = as.matrix(data[genes_to_keep, ])
+
 # You may want to set the random seed of the experiment so that the results
@@
 calculate proportion of times each pair of samples was clustered together
-across the 25 resampling runs, and plot the heatmap of those consensus
+across the 20 resampling runs, and plot the heatmap of those consensus
 matrices for $k=5$ and $k=20$.
+
+**EAP** It's somewhat unfortunately that B=20 and the maximum k=20. Makes it easy to get them confused. Could we switch one of these to be 15 or 25 so as to keep them straight for the reader?
 
@@
     stability = read.table(filename, sep="\t")
-    all_labels[[paste0("k=", i)]] = stability
+    all_labels[[paste0("B", i)]] = stability
 }
@@
 ```
+**EAP** I'm confused by this plot.
+Isn't the numbering here for different values of $k$?
+Why do the labels say "B1" and "B2"?
+Isn't that the notation above for the bootstrap replications for each values of $k$?
+It seems like the names of `all_labels` should be something like `k=1`,..., while the names of `all_labels[[1]]` should be `B1,B2,...`.
+But currently they have the same values.
 
@@
 ```{r  plotModelExplorerZoom, fig.width=6, fig.height=6,fig.cap="Plot of the cumulative density of similarity of clusterings of data for different k, restricted to k=6-10"}
-clusters = c("k=6", "k=7", "k=8", "k=9", "k=10")
+clusters = c("B6", "B7", "B8", "B9", "B10")
 selected_labels = all_labels[clusters]
@@
 
-Once the gene ID to GO mapping list is created, `moanin` provides an interface
-to `topGO` to determine enriched GO terms. In particular, it performs a
-p-value correction and only returns the significant GO-term enrichment in an
-easy to use `data.frame` object. Here, we show an example of running a GO term
-enrichment on the "Biological process" ontology (BP).
+Once the gene ID to GO mapping list is created, `moanin` provides an 
+interface to `topGO` to determine enriched GO terms. **EAP: What does the package do that is an enhancement to topGO output? Would be good to briefly describe** Here, we show an example of
+running a GO term enrichment on the "Biological process" ontology (BP).
 

### 2020-07-10 scripts/manuscript.Rmd
update

--- 2020-06-24 scripts/manuscript.Rmd
@@
 
- `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`:
+ `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`, along with the corresponding package that contains time course datasets we will use:
 
@@
 install_github("NelleV/moanin")
+install_github("NelleV/timecoursedata")
+```
+
+The following additional packages are needed for this workflow:
+
+```{r all_libraries}
+# From Github
 library(moanin)
-```
-
-The following additional packages are needed for this workflow:
-
-```{r all_libraries}
-# From github
 library(timecoursedata)
@@
 library(KEGGprofile)
-library(BiocWorkflowTools)
-
-# From GitHub
-library(moanin)
+library(BiocWorkflowTools) #Needed for compiling the .Rmd script
+```
+
+If Bioconductor is installed, the CRAN and Bioconductor packages above can be installed via
+
+```{r eval=FALSE}
+BiocManager::install(c("NMF","ggfortify","topGO","biomaRt","KEGGprofile","BiocWorkflowTools"))
 ```
@@
 
-```{r}
+```{r loadData}
 # Now load in the metadata

### 2020-07-13 scripts/run_clustering.R
update

@@ -57,7 +57,7 @@ kmeans_clusters = moanin:::splines_kmeans(
     n_init=20)
 
 # Perform prediction on the whole set of data.
-kmeans_clusters = moanin:::splines_kmeans_prediction(
+kmeans_clusters = moanin::: splines_kmeans_predict(
     y, kmeans_clusters)
 
 # Let's save only the labels here.

### 2020-07-14 scripts/manuscript.Rmd
update

@@
 
+**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.
+
+**EAP** actually, why do we need to rescale them between 0-1 rather than just a normal call to `scale`? Would it really make a difference?
+
 Those two heatmaps demonstrate that the clustering method successfully cluster
@@
 # Perform prediction on the whole set of data.
-kmeans_clusters = moanin:::splines_kmeans_prediction(
+kmeans_clusters = moanin::: splines_kmeans_predict(
     y, kmeans_clusters)

### 2020-07-15 scripts/manuscript.Rmd
start updating manuscript

@@
 ```
-
-**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at
-the beginning of the document? I also needed to install `BiocWorkflowTools`,
-`KEGGprofile` to run the .Rmd
-**NV**: I think some of the libraries above are not needed anymore. Some are
-also only used in hidden elements. I just checked that the whole pipeline
-works, by executing all command lines the reader will see, and the libraries
-are all loaded fine. I think BiocWorkflowTools is only need in the compilation
-of the Rmd file.
 
@@
 In both
-situations, we first need to set up a object (a `moanin_model` object) to
+situations, we first need to set up a object (a `Moanin` object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, ...). The `moanin_model` object
+(formula, number of degrees of freedom of the splines, ...). The `Moanin` object
 will contain information used
@@
 
-We start by creating the `moanin_model` object using the `create_moanin_model`
+We start by creating the `Moanin` object using the `create_moanin_model`
 function. We need to provide two things to the function: a `data.frame` with
@@
 
-We create a `moanin_model` for our data:
+We create a `Moanin` object for our data:
 
 ```{r}
-moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
+moanin_model = create_moanin_model(data=data, meta=meta,
+    degrees_of_freedom=6)
 ```
@@
 ```{r basis_matrix}
-print(moanin_model)
+show(moanin_model)
 ```
@@
 # Define contrasts	
-contrasts = create_timepoints_contrasts("M", "VL", moanin_model)
+contrasts = create_timepoints_contrasts(moanin_model,"M", "VL")
 ```
@@
 ```{r weekly_differential_analysis}
-weekly_de_analysis = DE_timepoints(
-     data, moanin_model, contrasts,
+weekly_de_analysis = DE_timepoints( moanin_model, contrasts,
      use_voom_weights=FALSE)
@@
 ```{r echo=FALSE, results="hide"}
-contrasts = create_timepoints_contrasts("M", "K", moanin_model)
+contrasts = create_timepoints_contrasts(moanin_model,"M", "K")
 K_weekly_de_analysis =  DE_timepoints(
-     data, moanin_model, contrasts,
+     moanin_model, contrasts,
      use_voom_weights=FALSE)
 
-contrasts = create_timepoints_contrasts("M", "C", moanin_model)
+contrasts = create_timepoints_contrasts(moanin_model,"M", "C")
 C_weekly_de_analysis =  DE_timepoints(
-     data, moanin_model, contrasts,
+     moanin_model, contrasts,
      use_voom_weights=FALSE)
 
-contrasts = create_timepoints_contrasts("M", "VH", moanin_model)
+contrasts = create_timepoints_contrasts( moanin_model,"M", "VH")
 VH_weekly_de_analysis =  DE_timepoints(
-     data, moanin_model, contrasts,
+     moanin_model, contrasts,
      use_voom_weights=FALSE)
@@
 ```{r timeSignifPerTime,fig.width=6, fig.height=6, echo=FALSE, fig.cap="Number of genes found DE in each timepoint when comparing the influenza strains to control."}
-timely_barplot = function(de_analysis, 
-    labels=NULL, qval_threshold=0.05,
-    xlab="Timepoint", ylab="Number of DE genes", main="", ylim=c(0, 20000), col="black", ...){
-
-    qval_colnames = colnames(de_analysis)[
-	grepl("qval", colnames(de_analysis))]
-    if(is.null(labels)){
-	labels = sapply(
-	    strsplit(gsub("-qval", "", qval_colnames), "\\."), .subset2, 3)
-    }
-    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)
-
-    barplot(number_de_genes_per_time, 
-            names.arg=labels, xlab=xlab,ylab=ylab,
-            col=col, main=main, ylim=ylim,...)  
-}
-
 par(mfrow=c(2, 2))
 time = names(time_colors)
-timely_barplot(K_weekly_de_analysis, labels=time, main="Kawasaki", las=3)
-timely_barplot(C_weekly_de_analysis, labels=time, main="California", las=3)
-timely_barplot(weekly_de_analysis, labels=time, main="Vietnam low-dose", las=3)
-timely_barplot(VH_weekly_de_analysis, labels=time, main="Vietnam high-dose", las=3)
+perWeek_barplot(K_weekly_de_analysis, labels=time, main="Kawasaki", las=3)
+perWeek_barplot(C_weekly_de_analysis, labels=time, main="California", las=3)
+perWeek_barplot(weekly_de_analysis, labels=time, main="Vietnam low-dose", las=3)
+perWeek_barplot(VH_weekly_de_analysis, labels=time, main="Vietnam high-dose", las=3)
 ```
@@
 exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
-plot_splines_data(data, moanin_model, subset_data=exampleGenes,
+plot_splines_data(moanin_model, subset_data=exampleGenes,
 		  colors=ann_colors$Group, smooth=TRUE)
@@
 # treatment).
-DE_results = DE_timecourse(
-    data, moanin_model, timecourse_contrasts,
+DE_results = DE_timecourse( moanin_model, timecourse_contrasts,
     use_voom_weights=FALSE)
@@
 compute the log fold change across the whole time-course. This is done via the function `estimate_log_fold_change` which takes as
-arguments the data, the `moanin_model` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
+arguments the data, the `Moanin` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
 
@@
 log_fold_change_timepoints = estimate_log_fold_change(
-    data, moanin_model, timecourse_contrasts,  method="timely")
+    moanin_model, timecourse_contrasts,  method="timely")
 ```
@@
 log_fold_change_timecourse = estimate_log_fold_change(
-    data, moanin_model, timecourse_contrasts,  method="timecourse")
+    moanin_model, timecourse_contrasts,  method="timecourse")
 
 log_fold_change_sum = estimate_log_fold_change(
-    data, moanin_model, timecourse_contrasts,  method="sum")
+    moanin_model, timecourse_contrasts,  method="sum")
 
 log_fold_change_max = estimate_log_fold_change(
-    data, moanin_model, timecourse_contrasts, method="max")
+    moanin_model, timecourse_contrasts, method="max")
 
 log_fold_change_min = estimate_log_fold_change(
-    data, moanin_model, timecourse_contrasts, method="min")
+    moanin_model, timecourse_contrasts, method="min")
 ```
@@
 top_DE_genes_pval = names(sort(pvalue)[1:10])
-plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
+plot_splines_data(moanin_model,subset_data=top_DE_genes_pval, 
     colors=ann_colors$Group, smooth=TRUE,
@@
 	 decreasing=TRUE)[1:10])
-plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
+plot_splines_data(moanin_model, subset_data=top_DE_genes_lfc, 
     colors=ann_colors$Group, smooth=TRUE,
@@
 - **Splines estimation**: for each gene, fit the splines function with the basis
-  of your choice (as contained in the `moanin_model` object).
+  of your choice (as contained in the `Moanin` object).
 - **Rescaling splines**: for each gene, rescale the estimated splines function
@@
 kmeans_clusters = splines_kmeans(
-    y, moanin_model, n_clusters=8,
+    moanin_model[genes_to_keep, ], n_clusters=8,
     random_seed=42,
@@
 ```{r fig.height=6, fig.width=6}
-plot_splines_data(
-    kmeans_clusters$centroids, moanin_model,
+plot_splines_data(moanin_model,
+    data=kmeans_clusters$centroids, 
     colors=ann_colors$Group,
@@
 dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
-plot_splines_data(dataToPlot, moanin_model,
+plot_splines_data(moanin_model, data=dataToPlot, 
     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
@@
 scores_and_labels = splines_kmeans_score_and_label(
-    data, kmeans_clusters)
+    moanin_model, kmeans_clusters)
 ```

### 2020-07-15 scripts/manuscript.Rmd
updates

@@
 # Keep the data corresponding to the genes of interest in another variable.
-y = as.matrix(data[genes_to_keep, ])
+# by subsetting the `moanin_model`, which contains the data.
+y = moanin_model[genes_to_keep,]
 ```
@@
 # First fit the kmeans clusters
-kmeans_clusters = splines_kmeans(
-    moanin_model[genes_to_keep, ], n_clusters=8,
+kmeans_clusters = splines_kmeans(y, n_clusters=8,
     random_seed=42,
@@
 
-The `scores_and_labels` list contains two elements:
+The `scores_and_labels` list contains three elements:
 
@@
       sufficiently good goodness-of-fit score.
+    - `score_cutoff`: the cutoff used on `scores` to determine whether to 
+      assign a label
 
@@
 the entire `scores` matrix returned by `splines_kmeans_score_and_label`).
-A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment.
+A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment (this choice can be changed by `proportion_genes_to_label`).
 Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
@@
 We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
+We can do this by rerunning `splines_kmeans_score_and_label` and setting the filter `percentage_genes_to_label=1` (and we can speed up the calculation by giving our previously calculated score via the argument `previous_scores`)
 
@@
 # Get the best score and best label for all of the genes
-# This is different from output of `splines_kmeans_score_and_label`, 
-# which filters the genes
-scores = rowMin(scores_and_labels$scores)
-labels = apply(scores_and_labels$scores, 1, which.min)
-max_score = quantile(scores, 0.5)
+# This time without filtering labels
+# We can give the previous calculated scores to `previous_scores` to save time
+unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
+best_score = rowMin(scores_and_labels$scores)
+best_label = unfiltered_scores$labels
 par(mfrow=c(3, 3))
@@
 for(cluster_id in 1:n_clusters){
-    hist(scores[labels == cluster_id],
+    hist(best_score[best_label==cluster_id],
 	 breaks=(1:50/50), xlim=c(0, 1),
@@
 	 xlab="score", ylab="Num. genes")
-    abline(v=max_score, col="red", lwd=3, lty=2)
+    abline(v=scores_and_labels$score_cutoff, col="red", lwd=3, lty=2)
 }
@@
 This underlines the importance of filtering genes that fit poorly to clusters.
-
-**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
-**NV** Fair point… maybe this should be fixed somehow by adding a random
-perturbation to the scores when the scores are equal? Although I think the
-text is still appropriate: those genes should definitely be filtered out.
 
@@
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
-plot_splines_data(dataToPlot, moanin_model,
+plot_splines_data(moanin_model, data=dataToPlot, 
     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
@@
 
-data_to_plot = data[genes_to_plot, ]
-submeta = meta
+# order the observations by Group, then time, then replicate
+orderedMeta<-colData(moanin_model)
 ord = order(
@@
 
-data_to_plot = data_to_plot[, ord]
 submeta = submeta[ord, ]
@@
 res = aheatmap(
-    data_to_plot,
+    assay(moanin_model[genes_to_plot,ord]),
     Colv=NA,
@@
 aheatmap(
-    moanin::rescale_values(data_to_plot)[res$rowInd,],
+    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
     Colv=NA,
@@
 ```
-
-**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.
-
-**EAP** actually, why do we need to rescale them between 0-1 rather than just a normal call to `scale`? Would it really make a difference?
 
@@
 for(n_cluster in all_possible_n_clusters){
-    clustering_results = splines_kmeans(
-	y, moanin_model,
+    clustering_results = splines_kmeans(y,
 	    n_clusters=n_cluster, random_seed=42,
@@
 random_subsample_y = y[indices, ]
-kmeans_clusters = moanin::splines_kmeans(
+kmeans_clusters = splines_kmeans(y[indices,]
     random_subsample_y, splines_model, n_clusters=n_clusters,
@@
 # Perform prediction on the whole set of data.
-kmeans_clusters = moanin::: splines_kmeans_predict(
-    y, kmeans_clusters)
+kmeans_clusters = splines_kmeans_predict(y, kmeans_clusters, method="distance")
 ```

### 2020-07-15 scripts/manuscript.Rmd
continuing to transition manuscript

@@
 
+The `Moanin` class extends the `SummarizedExperiment` class of Bioconductor, so that the data as well as the meta data and our spline information are all held in one object.
+This means that the object can be subsetted just like the data matrix, and the corresponding meta data and basis function evaluations will be similarly subsetted:
+
+```{r demonstrateSubsetting}
+dim(moanin_model)
+moanin_model[1:10,1:3]
+```
+
 ### Weekly differential expression analysis
@@
 # by subsetting the `moanin_model`, which contains the data.
-y = moanin_model[genes_to_keep,]
+de_moanin_model = moanin_model[genes_to_keep,]
 ```
@@
 # First fit the kmeans clusters
-kmeans_clusters = splines_kmeans(y, n_clusters=8,
+kmeans_clusters = splines_kmeans(de_moanin_model, n_clusters=8,
     random_seed=42,
@@
 ```{r fig.height=6, fig.width=6}
-plot_splines_data(moanin_model,
+plot_splines_data(de_moanin_model,
     data=kmeans_clusters$centroids, 
@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
-plot_splines_data(moanin_model, data=dataToPlot, 
+dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
+    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
+plot_splines_data(de_moanin_model,  data=dataToPlot, 
     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
@@
 spline fits.
+
+We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids
+
+```{r}
+de_moanin_rescale = de_moanin_model
+assay(de_moanin_rescale) = rescale_values(data=assay(de_moanin_model))
+plot_splines_data(de_moanin_rescale,  data=kmeans_clusters$centroids[cluster_of_interest,], 
+    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+    data_smooth_only=TRUE, subset_data=cluster2Genes[3:6],
+    mar=c(1.5,2.5,2,0.1))
+```
+**EAP** I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this.
+The `rescale_values` function is too sensitive to outliers.
+Would be better to rescale each centroid to scale of gene.
+If we had a helper function to do that, could easily add that to plot_splines_data (made issue on moanin), but probably right now should remove this.
 
@@
 best_label = unfiltered_scores$labels
+
 par(mfrow=c(3, 3))
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind(data[cluster2Score[1:4], ],
+dataToPlot = rbind([cluster2Score[1:4], ],
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
@@
 res = aheatmap(
-    assay(moanin_model[genes_to_plot,ord]),
+    assay(de_moanin_model[genes_to_plot,ord]),
     Colv=NA,
@@
 aheatmap(
-    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
+    rescale_values(de_moanin_model[genes_to_plot,ord])[res$rowInd,],
     Colv=NA,

### 2020-07-15 scripts/manuscript.Rmd
finish initial switch to new Moanin class

@@
 ```{r assigning_scores_and_labels}
-# Then assign scores and labels to all the data, using a goodness-of-fit
+# Then assign scores and labels to *all* the data, using a goodness-of-fit
 # scoring function.
@@
 	      Colv=NA, sub="Goodness-of-fit assignments")
+
+# recreate, so in same order, etc as before
 labels = scores_and_labels$labels
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind([cluster2Score[1:4], ],
+dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
@@
 res = aheatmap(
-    assay(de_moanin_model[genes_to_plot,ord]),
+    assay(moanin_model[genes_to_plot,ord]),
     Colv=NA,
@@
 aheatmap(
-    rescale_values(de_moanin_model[genes_to_plot,ord])[res$rowInd,],
+    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
     Colv=NA,
@@
 for(n_cluster in all_possible_n_clusters){
-    clustering_results = splines_kmeans(y,
+    clustering_results = splines_kmeans(de_moanin_model,
 	    n_clusters=n_cluster, random_seed=42,
@@
 n_genes = dim(y)[1]
-indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
-
-bootstrapped_y = y[indices, ]
+indices = sample(1:dim(de_moanin_model)[1], n_genes, replace=TRUE)
+
+bootstrapped_y = de_moanin_model[indices, ]
 ```
@@
 subsample_proportion = 0.8
-indices = sample(1:dim(y)[1], n_genes * subsample_proportion, replace=FALSE)
-
-subsampled_y = y[indices, ]
+indices = sample(1:dim(de_moanin_model)[1], n_genes * subsample_proportion, replace=FALSE)
+
+subsampled_y = de_moanin_model[indices, ]
 ```
@@
 # set.seed(random_seed)
-n_genes = dim(y)[1] * sample_proportion
-indices = sample(1:dim(y)[1], n_genes, replace=TRUE)
-
-random_subsample_y = y[indices, ]
-kmeans_clusters = splines_kmeans(y[indices,]
-    random_subsample_y, splines_model, n_clusters=n_clusters,
+n_genes = dim(de_moanin_model)[1] * sample_proportion
+indices = sample(1:dim(de_moanin_model)[1], n_genes, replace=TRUE)
+
+kmeans_clusters = splines_kmeans(de_moanin_model[indices,]
+    n_clusters=n_clusters,
     random_seed=random_seed,

### 2020-07-15 scripts/manuscript.Rmd
finish initial switch to new Moanin class

@@
 ```{r}
-n_genes = dim(y)[1]
+n_genes = dim(de_moanin_model)[1]
 indices = sample(1:dim(de_moanin_model)[1], n_genes, replace=TRUE)
@@
 # set.seed(random_seed)
-n_genes = dim(de_moanin_model)[1] * sample_proportion
+n_genes = dim(de_moanin_model)[1] * subsample_proportion
 indices = sample(1:dim(de_moanin_model)[1], n_genes, replace=TRUE)
 
-kmeans_clusters = splines_kmeans(de_moanin_model[indices,]
+kmeans_clusters = splines_kmeans(de_moanin_model[indices,],
     n_clusters=n_clusters,
@@
 # Perform prediction on the whole set of data.
-kmeans_clusters = splines_kmeans_predict(y, kmeans_clusters, method="distance")
+kmeans_clusters = splines_kmeans_predict(de_moanin_model, kmeans_clusters, method="distance")
 ```

### 2020-07-16 scripts/manuscript.Rmd
changes
update manuscript
fix error in manuscript

@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
+dataToPlot=rbind(as.matrix(assay(de_moanin_model[cluster2Genes[3:6],])), 
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
@@
 assay(de_moanin_rescale) = rescale_values(data=assay(de_moanin_model))
-plot_splines_data(de_moanin_rescale,  data=kmeans_clusters$centroids[cluster_of_interest,], 
+plot_splines_data(de_moanin_rescale,  
+    centroid=kmeans_clusters$centroids[cluster_of_interest,], 
     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
-    data_smooth_only=TRUE, subset_data=cluster2Genes[3:6],
+    subset_data=cluster2Genes[3:6],
     mar=c(1.5,2.5,2,0.1))
@@
     n_clusters=n_clusters,
-    random_seed=random_seed,
+    random_seed=42,
     n_init=20)
@@
 # Perform prediction on the whole set of data.
-kmeans_clusters = splines_kmeans_predict(de_moanin_model, kmeans_clusters, method="distance")
+kmeans_clusters = splines_kmeans_predict(de_moanin_model, kmeans_clusters, 
+    method="distance")
 ```
@@
 # Competing interests
+
 The authors declare that they have no competing interests.

### 2020-07-17 scripts/manuscript.Rmd
fix error

@@
 
-After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
+After filtering, we are left with `r dim(de_moanin_model)[1]` genes. We can then apply a
 clustering routine. As observed by looking at genes found differentially expressed, 
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
+dataToPlot = rbind(as.matrix(assay(moanin_model[cluster2Score[1:4], ])),
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])

### 2020-07-17 scripts/manuscript.Rmd
switch from DataFrame to matrix

--- 2020-07-16 scripts/manuscript.Rmd
@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot=rbind(as.matrix(assay(de_moanin_model[cluster2Genes[3:6],])), 
+dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])

### 2020-07-17 scripts/manuscript.Rmd
fix error in missing submeta

@@
 ord = order(
-  submeta$Group,
-  submeta$Timepoint,
-  submeta$Replicate)
-submeta$Timepoint = as.factor(submeta$Timepoint)
-
-submeta = submeta[ord, ]
+  orderedMeta$Group,
+  orderedMeta$Timepoint,
+  orderedMeta$Replicate)
+orderedMeta$Timepoint = as.factor(orderedMeta$Timepoint)
+
+orderedMeta = orderedMeta[ord, ]
 
@@
     color="YlGnBu",
-    annCol=submeta[,
+    annCol=orderedMeta[,
                 c("Group", "Timepoint")],
@@
     Rowv=NA,
-    annCol=submeta[,
+    annCol=orderedMeta[,
                 c("Group", "Timepoint")],

### 2020-07-17 scripts/manuscript.Rmd
fix errors in heatmap

@@
 # order the observations by Group, then time, then replicate
-orderedMeta<-colData(moanin_model)
+orderedMeta<-data.frame(colData(moanin_model))
 ord = order(
@@
     color="YlGnBu",
-    annCol=orderedMeta[,
-                c("Group", "Timepoint")],
+    annCol=orderedMeta[,c("Group", "Timepoint")],
     annLegend=FALSE,
@@
     Rowv=NA,
-    annCol=orderedMeta[,
-                c("Group", "Timepoint")],
+    annCol=orderedMeta[, c("Group", "Timepoint")],
     annLegend=TRUE,

### 2020-07-17 scripts/manuscript.Rmd
fixes to update manuscript

@@
     clustering_results = all_clustering[i]
-    average_silhouette_values[i] = average_silhouette(clustering_results, y)
+    average_silhouette_values[i] = average_silhouette(clustering_results, assay(de_moanin_model))
     i = i + 1

### 2020-07-20 scripts/manuscript.Rmd
manuscript runs on my computer

@@
 average_silhouette = function(labels, y) {
-    silhouette_results = silhouette(unlist(labels[1]), dist(y))
+    silhouette_results = silhouette(unlist(labels[1]),dist(y))
     return(mean(silhouette_results[, 3]))
@@
     clustering_results = all_clustering[i]
-    average_silhouette_values[i] = average_silhouette(clustering_results, assay(de_moanin_model))
+    average_silhouette_values[i] = average_silhouette(clustering_results,
+        assay(de_moanin_model))
     i = i + 1
@@
 Using the bootstrapping strategy, we sample with replacement a sample of the
-same size as our original clustering (i.e. `r dim(y)[1]` genes):
+same size as our original clustering (i.e. `r dim(de_moanin_model)[1]` genes):
 

### 2020-07-21 scripts/manuscript.Rmd

@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
-    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
-plot_splines_data(de_moanin_model,  data=dataToPlot, 
-    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
-    mar=c(1.5,2.5,2,0.1))
+# FIXME This line fails
+# dataToPlot = rbind(
+#     assay(de_moanin_model[cluster2Genes[3:6],]), 
+#    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
+# plot_splines_data(de_moanin_model, data=dataToPlot, 
+#     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
+#    mar=c(1.5,2.5,2,0.1))
 ```
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
-    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
-plot_splines_data(moanin_model, data=dataToPlot, 
-    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
-    mar=c(1.5,2.5,2,0.1))
+# FIXME this lines failes.
+# dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
+#     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
+# plot_splines_data(moanin_model, data=dataToPlot, 
+#     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+#     mar=c(1.5,2.5,2,0.1))
 ```
@@
 res = aheatmap(
-    assay(moanin_model[genes_to_plot,ord]),
+    as.matrix(assay(moanin_model[genes_to_plot,ord])),
     Colv=NA,
@@
 # Now use the results of the previous call to aheatmap to reorder the genes.
-aheatmap(
-    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
-    Colv=NA,
-    Rowv=NA,
-    annCol=orderedMeta[, c("Group", "Timepoint")],
-    annLegend=TRUE,
-    annColors=ann_colors,
-    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
-    treeheight=0)
+# FIXME this line doesn't work
+# aheatmap(
+#    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
+#    Colv=NA,
+#    Rowv=NA,
+#    annCol=orderedMeta[, c("Group", "Timepoint")],
+#    annLegend=TRUE,
+#    annColors=ann_colors,
+#    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
+#    treeheight=0)
 

### 2020-07-21 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
+%% added by MLS, this is not in the F1000 style by default %%
+
+\hypersetup{unicode=true,
+            pdftitle={A pipeline to analyse time-course gene expression data},
+            pdfkeywords={time-course gene expression data, clustering, differential expression, workflow},
+            pdfborder={0 0 0},
+            breaklinks=true}
+
+%% End added by MLS %%
+
 \setlength{\parindent}{0pt}
@@
 
-The manuscript was last compile on 2020-07-04 10:52:02.
+The manuscript was last compile on 2020-07-21 12:22:19.
 
@@
 
-\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}:
+\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}, along with the corresponding package that contains time course datasets we will use:
 
@@
 \KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
+\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/timecoursedata"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+The following additional packages are needed for this workflow:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# From Github}
 \KeywordTok{library}\NormalTok{(moanin)}
-\end{Highlighting}
-\end{Shaded}
-
-The following additional packages are needed for this workflow:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# From github}
 \KeywordTok{library}\NormalTok{(timecoursedata)}
@@
 \KeywordTok{library}\NormalTok{(KEGGprofile)}
-\KeywordTok{library}\NormalTok{(BiocWorkflowTools)}
-
-\CommentTok{# From GitHub}
-\KeywordTok{library}\NormalTok{(moanin)}
-\end{Highlighting}
-\end{Shaded}
-
-\textbf{EAP}: Why is this list not the same as the (hidden via echo=FALSE) list at
-the beginning of the document? I also needed to install \texttt{BiocWorkflowTools},
-\texttt{KEGGprofile} to run the .Rmd
-\textbf{NV}: I think some of the libraries above are not needed anymore. Some are
-also only used in hidden elements. I just checked that the whole pipeline
-works, by executing all command lines the reader will see, and the libraries
-are all loaded fine. I think BiocWorkflowTools is only need in the compilation
-of the Rmd file.
+\KeywordTok{library}\NormalTok{(BiocWorkflowTools) }\CommentTok{#Needed for compiling the .Rmd script}
+\end{Highlighting}
+\end{Shaded}
+
+If Bioconductor is installed, the CRAN and Bioconductor packages above can be installed via
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"NMF"}\NormalTok{,}\StringTok{"ggfortify"}\NormalTok{,}\StringTok{"topGO"}\NormalTok{,}\StringTok{"biomaRt"}\NormalTok{,}\StringTok{"KEGGprofile"}\NormalTok{,}\StringTok{"BiocWorkflowTools"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 In both
-situations, we first need to set up a object (a \texttt{moanin\_model} object) to
+situations, we first need to set up a object (a \texttt{Moanin} object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{moanin\_model} object
+(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin} object
 will contain information used
@@
 
-We start by creating the \texttt{moanin\_model} object using the \texttt{create\_moanin\_model}
+We start by creating the \texttt{Moanin} object using the \texttt{create\_moanin\_model}
 function. We need to provide two things to the function: a \texttt{data.frame} with
@@
 
-We create a \texttt{moanin\_model} for our data:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
+We create a \texttt{Moanin} object for our data:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{moanin_model =}\StringTok{ }\KeywordTok{create_moanin_model}\NormalTok{(}\DataTypeTok{data=}\NormalTok{data, }\DataTypeTok{meta=}\NormalTok{meta,}
+    \DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\KeywordTok{print}\NormalTok{(moanin_model)}
+\KeywordTok{show}\NormalTok{(moanin_model)}
 \end{Highlighting}
@@
 \begin{verbatim}
-## moanin_model object on 209 samples containing the following information:
-## 1) Meta data with 5 variables:
-## [1] ""            "Group"       "Replicate"   "Timepoint"   "WeeklyGroup"
-## 2) Basis matrix with 35 basis functions
-## Basis matrix was constructed with the following formula
-## ~Group + Group:splines::ns(Timepoint, df = degrees_of_freedom) + 0 
-## Where degrees_of_freedom=6 
-## To access these objects use:
-##  <x_name>$degrees_of_freedom
-##  <x_name>$basis
-##  <x_name>$meta
-##  <x_name>$formula
+## Moanin object on 209 samples containing the following information:
+## Group variable given by 'Group' with the following levels:
+##  M  K  C VL VH 
+## 42 42 42 42 41 
+## Time variable given by 'Timepoint'
+## Basis matrix with 35 basis_matrix functions
+## Basis matrix was constructed with the following spline_formula
+## ~Group + Group:splines::ns(Timepoint, df = 6) + 0 
+## 
+## Information about the data (a SummarizedExperiment object):
+## class: SummarizedExperiment 
+## dim: 39544 209 
+## metadata(0):
+## assays(1): ''
+## rownames(39544): NM_009912 NM_008725 ... NM_010201.1 AK078781
+## rowData names(0):
+## colnames(209): GSM1557140 GSM1557141 ... GSM1557347 GSM1557348
+## colData names(5): '' Group Replicate Timepoint WeeklyGroup
+\end{verbatim}
+
+The \texttt{Moanin} class extends the \texttt{SummarizedExperiment} class of Bioconductor, so that the data as well as the meta data and our spline information are all held in one object.
+This means that the object can be subsetted just like the data matrix, and the corresponding meta data and basis function evaluations will be similarly subsetted:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{dim}\NormalTok{(moanin_model)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 39544   209
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{moanin_model[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{,}\DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Moanin object on 3 samples containing the following information:
+## Group variable given by 'Group' with the following levels:
+##  M  K  C VL VH 
+##  0  3  0  0  0 
+## Time variable given by 'Timepoint'
+## Basis matrix with 35 basis_matrix functions
+## Basis matrix was constructed with the following spline_formula
+## ~Group + Group:splines::ns(Timepoint, df = 6) + 0 
+## 
+## Information about the data (a SummarizedExperiment object):
+## class: SummarizedExperiment 
+## dim: 10 3 
+## metadata(0):
+## assays(1): ''
+## rownames(10): NM_009912 NM_008725 ... NM_013782 NM_028622
+## rowData names(0):
+## colnames(3): GSM1557140 GSM1557141 GSM1557142
+## colData names(5): '' Group Replicate Timepoint WeeklyGroup
 \end{verbatim}
@@
 \CommentTok{# Define contrasts  }
-\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{, moanin_model)}
+\NormalTok{contrasts =}\StringTok{ }\KeywordTok{create_timepoints_contrasts}\NormalTok{(moanin_model,}\StringTok{"M"}\NormalTok{, }\StringTok{"VL"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{(}
-\NormalTok{     data, moanin_model, contrasts,}
+\NormalTok{weekly_de_analysis =}\StringTok{ }\KeywordTok{DE_timepoints}\NormalTok{( moanin_model, contrasts,}
      \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
@@
 \CommentTok{# treatment).}
-\NormalTok{DE_results =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts,}
+\NormalTok{DE_results =}\StringTok{ }\KeywordTok{DE_timecourse}\NormalTok{( moanin_model, timecourse_contrasts,}
     \DataTypeTok{use_voom_weights=}\OtherTok{FALSE}\NormalTok{)}
@@
 compute the log fold change across the whole time-course. This is done via the function \texttt{estimate\_log\_fold\_change} which takes as
-arguments the data, the \texttt{moanin\_model} object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
+arguments the data, the \texttt{Moanin} object, the contrasts to evaluate, and the method to use to estimate the log-fold change.
 
@@
 \NormalTok{log_fold_change_timepoints =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
+\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timely"}\NormalTok{)}
 \end{Highlighting}
@@
 \toprule
-  & M-K:1 & M-K:7 & M-K:11 & M-K:14 & M-K:2\\
+  & M-K:0 & M-K:3 & M-K:6 & M-K:9 & M-K:12\\
 \midrule
-NM\_009912 & -0.0061424 & 0.0224948 & -0.0515947 & -0.1143973 & 0.0614619\\
-NM\_008725 & 2.6547855 & 0.2734387 & -0.6882925 & 0.4718807 & -1.7463319\\
-NM\_007473 & -0.1603623 & 0.1017655 & 0.5079573 & 0.0595773 & 0.4250185\\
-ENSMUST00000094955 & 0.3505920 & 0.6463011 & 0.2564796 & 0.3599102 & 0.4177116\\
-NM\_001042489 & 0.0889750 & 0.2321381 & -0.1891501 & 0.1786113 & 0.3320644\\
+NM\_009912 & -0.5662398 & -0.2454676 & -0.4096932 & -0.3320085 & -0.0502552\\
+NM\_008725 & 0.9947059 & -3.4609546 & 0.6561484 & -1.2106329 & -0.1998796\\
+NM\_007473 & -0.5176866 & -0.3146529 & 0.2327960 & -0.3445581 & 0.0476160\\
+ENSMUST00000094955 & -0.2201803 & 0.2566677 & -0.0358376 & -0.0469726 & -0.0707469\\
+NM\_001042489 & -0.2475888 & -0.2043850 & -0.3428368 & -0.3930846 & -0.4821660\\
 \addlinespace
-NM\_008159 & -0.3863770 & 0.0298395 & -0.0862755 & -0.3266351 & -0.1893062\\
+NM\_008159 & -0.7466435 & -0.0678902 & -0.3072220 & -0.7067393 & -0.4966867\\
 \bottomrule
@@
 \NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}
+\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}
 
 \NormalTok{log_fold_change_sum =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
+\NormalTok{    moanin_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"sum"}\NormalTok{)}
 
 \NormalTok{log_fold_change_max =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
+\NormalTok{    moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"max"}\NormalTok{)}
 
 \NormalTok{log_fold_change_min =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
-\NormalTok{    data, moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
+\NormalTok{    moanin_model, timecourse_contrasts, }\DataTypeTok{method=}\StringTok{"min"}\NormalTok{)}
 \end{Highlighting}
@@
 \midrule
-NM\_009912 & 0.432 & 0.753 & -0.836 & -0.488\\
-NM\_008725 & -1.257 & -2.008 & 1.924 & 1.729\\
-NM\_007473 & 0.385 & 0.335 & 0.516 & 0.263\\
-ENSMUST00000094955 & 0.355 & 0.376 & -0.316 & 0.256\\
-NM\_001042489 & -0.266 & 0.410 & 0.415 & 0.374\\
+NM\_009912 & -0.409 & -0.753 & -1.025 & -1.420\\
+NM\_008725 & 1.575 & 2.008 & 2.543 & 2.618\\
+NM\_007473 & -0.416 & -0.335 & -0.213 & 0.364\\
+ENSMUST00000094955 & -0.113 & -0.376 & -0.156 & -0.309\\
+NM\_001042489 & -0.234 & -0.410 & 0.308 & 0.430\\
 \bottomrule
@@
 \NormalTok{top_DE_genes_pval =}\StringTok{ }\KeywordTok{names}\NormalTok{(}\KeywordTok{sort}\NormalTok{(pvalue)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_pval, ], moanin_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(moanin_model,}\DataTypeTok{subset_data=}\NormalTok{top_DE_genes_pval, }
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
      \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
-\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], moanin_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(moanin_model, }\DataTypeTok{subset_data=}\NormalTok{top_DE_genes_lfc, }
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
@@
 \CommentTok{# Keep the data corresponding to the genes of interest in another variable.}
-\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}
+\CommentTok{# by subsetting the `moanin_model`, which contains the data.}
+\NormalTok{de_moanin_model =}\StringTok{ }\NormalTok{moanin_model[genes_to_keep,]}
 \end{Highlighting}
@@
 
-After filtering, we are left with 3950 genes. We can then apply a
+After filtering, we are left with 5521 genes. We can then apply a
 clustering routine. As observed by looking at genes found differentially expressed,
@@
   \textbf{Splines estimation}: for each gene, fit the splines function with the basis
-  of your choice (as contained in the \texttt{moanin\_model} object).
+  of your choice (as contained in the \texttt{Moanin} object).
 \item
@@
 \CommentTok{# First fit the kmeans clusters}
-\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
-\NormalTok{    y, moanin_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
+\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(de_moanin_model, }\DataTypeTok{n_clusters=}\DecValTok{8}\NormalTok{,}
     \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
@@
 \begin{Highlighting}[]
-\KeywordTok{plot_splines_data}\NormalTok{(}
-\NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, moanin_model,}
+\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_model,}
+    \DataTypeTok{data=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids, }
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group,}
@@
 \NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(y[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,])}
-\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
-    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
-    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}
+\CommentTok{# }\AlertTok{FIXME}\CommentTok{ This line fails}
+\CommentTok{# dataToPlot = rbind(}
+\CommentTok{#     assay(de_moanin_model[cluster2Genes[3:6],]), }
+\CommentTok{#    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])}
+\CommentTok{# plot_splines_data(de_moanin_model, data=dataToPlot, }
+\CommentTok{#     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,}
+\CommentTok{#    mar=c(1.5,2.5,2,0.1))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 spline fits.
+
+We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{de_moanin_rescale =}\StringTok{ }\NormalTok{de_moanin_model}
+\KeywordTok{assay}\NormalTok{(de_moanin_rescale) =}\StringTok{ }\KeywordTok{rescale_values}\NormalTok{(}\DataTypeTok{data=}\KeywordTok{assay}\NormalTok{(de_moanin_model))}
+\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_rescale,  }
+    \DataTypeTok{centroid=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,], }
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
+    \DataTypeTok{subset_data=}\NormalTok{cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{],}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
+
+\textbf{EAP} I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this.
+The \texttt{rescale\_values} function is too sensitive to outliers.
+Would be better to rescale each centroid to scale of gene.
+If we had a helper function to do that, could easily add that to plot\_splines\_data (made issue on moanin), but probably right now should remove this.
 
@@
 \begin{Highlighting}[]
-\CommentTok{# Then assign scores and labels to all the data, using a goodness-of-fit}
+\CommentTok{# Then assign scores and labels to *all* the data, using a goodness-of-fit}
 \CommentTok{# scoring function.}
 \NormalTok{scores_and_labels =}\StringTok{ }\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
-\NormalTok{    data, kmeans_clusters)}
-\end{Highlighting}
-\end{Shaded}
-
-The \texttt{scores\_and\_labels} list contains two elements:
+\NormalTok{    moanin_model, kmeans_clusters)}
+\end{Highlighting}
+\end{Shaded}
+
+The \texttt{scores\_and\_labels} list contains three elements:
 
@@
   sufficiently good goodness-of-fit score.
+- `score_cutoff`: the cutoff used on `scores` to determine whether to 
+  assign a label
 \end{verbatim}
@@
 The criteria for being ``sufficiently low'' is based on looking at the distribution of the scores of all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by \texttt{splines\_kmeans\_score\_and\_label}).
-A gene is then assigned to a cluster only if their best score is above the 50\% percentile of that distribution, with the remaining genes getting \texttt{NA} as their assignment.
+A gene is then assigned to a cluster only if their best score is above the 50\% percentile of that distribution, with the remaining genes getting \texttt{NA} as their assignment (this choice can be changed by \texttt{proportion\_genes\_to\_label}).
 Note that because the distribution of scores of genes across \emph{all} clusters are used, this is not equivalent to assigning 50\% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
@@
 We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
+We can do this by rerunning \texttt{splines\_kmeans\_score\_and\_label} and setting the filter \texttt{percentage\_genes\_to\_label=1} (and we can speed up the calculation by giving our previously calculated score via the argument \texttt{previous\_scores})
 
@@
 \CommentTok{# Get the best score and best label for all of the genes}
-\CommentTok{# This is different from output of `splines_kmeans_score_and_label`, }
-\CommentTok{# which filters the genes}
-\NormalTok{scores =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{apply}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores, }\DecValTok{1}\NormalTok{, which.min)}
-\NormalTok{max_score =}\StringTok{ }\KeywordTok{quantile}\NormalTok{(scores, }\FloatTok{0.5}\NormalTok{)}
+\CommentTok{# This time without filtering labels}
+\CommentTok{# We can give the previous calculated scores to `previous_scores` to save time}
+\NormalTok{unfiltered_scores<-}\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(moanin_model,kmeans_clusters,}\DataTypeTok{proportion_genes_to_label=}\DecValTok{1}\NormalTok{, }\DataTypeTok{previous_scores=}\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
+\NormalTok{best_score =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
+\NormalTok{best_label =}\StringTok{ }\NormalTok{unfiltered_scores}\OperatorTok{$}\NormalTok{labels}
+
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{3}\NormalTok{))}
@@
 \ControlFlowTok{for}\NormalTok{(cluster_id }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\NormalTok{n_clusters)\{}
-    \KeywordTok{hist}\NormalTok{(scores[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_id],}
+    \KeywordTok{hist}\NormalTok{(best_score[best_label}\OperatorTok{==}\NormalTok{cluster_id],}
      \DataTypeTok{breaks=}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{50}\OperatorTok{/}\DecValTok{50}\NormalTok{), }\DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{),}
@@
      \DataTypeTok{xlab=}\StringTok{"score"}\NormalTok{, }\DataTypeTok{ylab=}\StringTok{"Num. genes"}\NormalTok{)}
-    \KeywordTok{abline}\NormalTok{(}\DataTypeTok{v=}\NormalTok{max_score, }\DataTypeTok{col=}\StringTok{"red"}\NormalTok{, }\DataTypeTok{lwd=}\DecValTok{3}\NormalTok{, }\DataTypeTok{lty=}\DecValTok{2}\NormalTok{)}
+    \KeywordTok{abline}\NormalTok{(}\DataTypeTok{v=}\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{score_cutoff, }\DataTypeTok{col=}\StringTok{"red"}\NormalTok{, }\DataTypeTok{lwd=}\DecValTok{3}\NormalTok{, }\DataTypeTok{lty=}\DecValTok{2}\NormalTok{)}
 \NormalTok{\}}
@@
 
-\textbf{EAP} In fact, \texttt{which.min} which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
-\textbf{NV} Fair point\textless{}80\textgreater{} maybe this should be fixed somehow by adding a random
-perturbation to the scores when the scores are equal? Although I think the
-text is still appropriate: those genes should definitely be filtered out.
-
 We can also investigate the
@@
 
-Of the original 3950 genes used in the
-clustering, 3950 are still assigned to a cluster based on their goodness of
+Of the original 5521 genes used in the
+clustering, 4946 are still assigned to a cluster based on their goodness of
 fit score. We can compare whether they are still assigned to the same clusters
@@
 
-Of the four genes we plotted before, 3 of them remained assigned to cluster 2. We can again look at a
+Of the four genes we plotted before, 2 of them remained assigned to cluster 2. We can again look at a
 few genes in cluster 2, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})
@@
 \NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}
-    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
-\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
-    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
-    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topScoreInCluster2-1}
-
-}
-
-\caption{Top scoring genes in Cluster 2}\label{fig:topScoreInCluster2}
-\end{figure}
+\CommentTok{# }\AlertTok{FIXME}\CommentTok{ this lines failes.}
+\CommentTok{# dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),}
+\CommentTok{#     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])}
+\CommentTok{# plot_splines_data(moanin_model, data=dataToPlot, }
+\CommentTok{#     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,}
+\CommentTok{#     mar=c(1.5,2.5,2,0.1))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-Now we will create the heatmaps of these genes (4332 genes, Figure \ref{fig:heatmapCluster8}).
+Now we will create the heatmaps of these genes (1283 genes, Figure \ref{fig:heatmapCluster8}).
 
@@
 
-\NormalTok{data_to_plot =}\StringTok{ }\NormalTok{data[genes_to_plot, ]}
-\NormalTok{submeta =}\StringTok{ }\NormalTok{meta}
+\CommentTok{# order the observations by Group, then time, then replicate}
+\NormalTok{orderedMeta<-}\KeywordTok{data.frame}\NormalTok{(}\KeywordTok{colData}\NormalTok{(moanin_model))}
 \NormalTok{ord =}\StringTok{ }\KeywordTok{order}\NormalTok{(}
-\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Group,}
-\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Timepoint,}
-\NormalTok{  submeta}\OperatorTok{$}\NormalTok{Replicate)}
-\NormalTok{submeta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(submeta}\OperatorTok{$}\NormalTok{Timepoint)}
-
-\NormalTok{data_to_plot =}\StringTok{ }\NormalTok{data_to_plot[, ord]}
-\NormalTok{submeta =}\StringTok{ }\NormalTok{submeta[ord, ]}
+\NormalTok{  orderedMeta}\OperatorTok{$}\NormalTok{Group,}
+\NormalTok{  orderedMeta}\OperatorTok{$}\NormalTok{Timepoint,}
+\NormalTok{  orderedMeta}\OperatorTok{$}\NormalTok{Replicate)}
+\NormalTok{orderedMeta}\OperatorTok{$}\NormalTok{Timepoint =}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(orderedMeta}\OperatorTok{$}\NormalTok{Timepoint)}
+
+\NormalTok{orderedMeta =}\StringTok{ }\NormalTok{orderedMeta[ord, ]}
 
 \NormalTok{res =}\StringTok{ }\KeywordTok{aheatmap}\NormalTok{(}
-\NormalTok{    data_to_plot,}
+    \KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{assay}\NormalTok{(moanin_model[genes_to_plot,ord])),}
     \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
     \DataTypeTok{color=}\StringTok{"YlGnBu"}\NormalTok{,}
-    \DataTypeTok{annCol=}\NormalTok{submeta[,}
-                \KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
+    \DataTypeTok{annCol=}\NormalTok{orderedMeta[,}\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
     \DataTypeTok{annLegend=}\OtherTok{FALSE}\NormalTok{,}
@@
 \CommentTok{# Now use the results of the previous call to aheatmap to reorder the genes.}
-\KeywordTok{aheatmap}\NormalTok{(}
-\NormalTok{    moanin}\OperatorTok{::}\KeywordTok{rescale_values}\NormalTok{(data_to_plot)[res}\OperatorTok{$}\NormalTok{rowInd,],}
-    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
-    \DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
-    \DataTypeTok{annCol=}\NormalTok{submeta[,}
-                \KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
-    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{,}
-    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
-    \DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"Cluster"}\NormalTok{, cluster_to_plot, }\StringTok{"(rescaled)"}\NormalTok{),}
-    \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
+\CommentTok{# }\AlertTok{FIXME}\CommentTok{ this line doesn't work}
+\CommentTok{# aheatmap(}
+\CommentTok{#    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],}
+\CommentTok{#    Colv=NA,}
+\CommentTok{#    Rowv=NA,}
+\CommentTok{#    annCol=orderedMeta[, c("Group", "Timepoint")],}
+\CommentTok{#    annLegend=TRUE,}
+\CommentTok{#    annColors=ann_colors,}
+\CommentTok{#    main=paste("Cluster", cluster_to_plot, "(rescaled)"),}
+\CommentTok{#    treeheight=0)}
 \end{Highlighting}
@@
 \ControlFlowTok{for}\NormalTok{(n_cluster }\ControlFlowTok{in}\NormalTok{ all_possible_n_clusters)\{}
-\NormalTok{    clustering_results =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(}
-\NormalTok{    y, moanin_model,}
+\NormalTok{    clustering_results =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(de_moanin_model,}
         \DataTypeTok{n_clusters=}\NormalTok{n_cluster, }\DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
@@
 \NormalTok{average_silhouette =}\StringTok{ }\ControlFlowTok{function}\NormalTok{(labels, y) \{}
-\NormalTok{    silhouette_results =}\StringTok{ }\KeywordTok{silhouette}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(labels[}\DecValTok{1}\NormalTok{]), }\KeywordTok{dist}\NormalTok{(y))}
+\NormalTok{    silhouette_results =}\StringTok{ }\KeywordTok{silhouette}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(labels[}\DecValTok{1}\NormalTok{]),}\KeywordTok{dist}\NormalTok{(y))}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{mean}\NormalTok{(silhouette_results[, }\DecValTok{3}\NormalTok{]))}
@@
 \NormalTok{    clustering_results =}\StringTok{ }\NormalTok{all_clustering[i]}
-\NormalTok{    average_silhouette_values[i] =}\StringTok{ }\KeywordTok{average_silhouette}\NormalTok{(clustering_results, y)}
+\NormalTok{    average_silhouette_values[i] =}\StringTok{ }\KeywordTok{average_silhouette}\NormalTok{(clustering_results,}
+        \KeywordTok{assay}\NormalTok{(de_moanin_model))}
 \NormalTok{    i =}\StringTok{ }\NormalTok{i }\OperatorTok{+}\StringTok{ }\DecValTok{1}
@@
 Using the bootstrapping strategy, we sample with replacement a sample of the
-same size as our original clustering (i.e.~3950 genes):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{]}
-\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}
-
-\NormalTok{bootstrapped_y =}\StringTok{ }\NormalTok{y[indices, ]}
+same size as our original clustering (i.e.~5521 genes):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{]}
+\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}
+
+\NormalTok{bootstrapped_y =}\StringTok{ }\NormalTok{de_moanin_model[indices, ]}
 \end{Highlighting}
@@
 \NormalTok{subsample_proportion =}\StringTok{ }\FloatTok{0.8}
-\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
-
-\NormalTok{subsampled_y =}\StringTok{ }\NormalTok{y[indices, ]}
+\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
+
+\NormalTok{subsampled_y =}\StringTok{ }\NormalTok{de_moanin_model[indices, ]}
 \end{Highlighting}
@@
 \CommentTok{# set.seed(random_seed)}
-\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{] }\OperatorTok{*}\StringTok{ }\NormalTok{sample_proportion}
-\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}
-
-\NormalTok{random_subsample_y =}\StringTok{ }\NormalTok{y[indices, ]}
-\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
-\NormalTok{    random_subsample_y, splines_model, }\DataTypeTok{n_clusters=}\NormalTok{n_clusters,}
-    \DataTypeTok{random_seed=}\NormalTok{random_seed,}
+\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{] }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion}
+\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}
+
+\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans}\NormalTok{(de_moanin_model[indices,],}
+    \DataTypeTok{n_clusters=}\NormalTok{n_clusters,}
+    \DataTypeTok{random_seed=}\DecValTok{42}\NormalTok{,}
     \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}
@@
 \CommentTok{# Perform prediction on the whole set of data.}
-\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{splines_kmeans_prediction}\NormalTok{(}
-\NormalTok{    y, kmeans_clusters)}
+\NormalTok{kmeans_clusters =}\StringTok{ }\KeywordTok{splines_kmeans_predict}\NormalTok{(de_moanin_model, kmeans_clusters, }
+    \DataTypeTok{method=}\StringTok{"distance"}\NormalTok{)}
 \end{Highlighting}
@@
 
-\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus1}
+\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-1}
 \end{figure}
@@
 
-\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus2}
+\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-2}
 \end{figure}
@@
 \midrule
-04060 & Cytokine-cytokine receptor interaction & 34 & 0\\
-04620 & Toll-like receptor signaling pathway & 40 & 0\\
-04630 & Jak-STAT signaling pathway & 33 & 0\\
-03050 & Proteasome & 53 & 0\\
-04621 & NOD-like receptor signaling pathway & 47 & 0\\
-\addlinespace
-04380 & Osteoclast differentiation & 34 & 0\\
-04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
-04210 & Apoptosis & 37 & 0\\
-04622 & RIG-I-like receptor signaling pathway & 39 & 0\\
-05160 & Hepatitis C & 29 & 0\\
+04740 & Olfactory transduction & 8 & 0\\
+04080 & Neuroactive ligand-receptor interaction & 10 & 0\\
 \bottomrule
@@
 \midrule
-GO:0003002 & regionalization & 171 & 153 & 133.53 & 0.00012 & 0.123\\
-GO:0007224 & smoothened signaling pathway & 75 & 70 & 58.56 & 0.00033 & 0.282\\
-GO:0010970 & transport along microtubule & 84 & 77 & 65.59 & 0.00044 & 0.322\\
-GO:0006805 & xenobiotic metabolic process & 49 & 47 & 38.26 & 0.00057 & 0.365\\
-GO:0048562 & embryonic organ morphogenesis & 154 & 136 & 120.25 & 0.00073 & 0.374\\
+GO:0044260 & cellular macromolecule metabolic process & 3762 & 3607 & 3540.21 & 0.00031 & 0.276\\
+GO:0043603 & cellular amide metabolic process & 467 & 458 & 439.47 & 0.00036 & 0.276\\
+GO:0043161 & proteasome-mediated ubiquitin-dependent ... & 201 & 199 & 189.15 & 0.00042 & 0.276\\
+GO:0016071 & mRNA metabolic process & 285 & 280 & 268.20 & 0.00052 & 0.276\\
+GO:0070925 & organelle assembly & 427 & 417 & 401.83 & 0.00052 & 0.276\\
 \addlinespace
-GO:0055114 & oxidation-reduction process & 530 & 457 & 413.85 & 0.00086 & 0.374\\
-GO:0060173 & limb development & 90 & 82 & 70.28 & 0.00090 & 0.374\\
-GO:0060070 & canonical Wnt signaling pathway & 152 & 134 & 118.69 & 0.00095 & 0.374\\
-GO:0006635 & fatty acid beta-oxidation & 38 & 37 & 29.67 & 0.00095 & 0.374\\
-GO:0090596 & sensory organ morphogenesis & 138 & 122 & 107.76 & 0.00124 & 0.453\\
+GO:0019058 & viral life cycle & 123 & 123 & 115.75 & 0.00054 & 0.276\\
+GO:0018130 & heterocycle biosynthetic process & 1809 & 1733 & 1702.35 & 0.00093 & 0.348\\
+GO:0019438 & aromatic compound biosynthetic process & 1810 & 1734 & 1703.29 & 0.00098 & 0.348\\
+GO:0042110 & T cell activation & 273 & 268 & 256.91 & 0.00102 & 0.348\\
+GO:0006954 & inflammatory response & 410 & 398 & 385.83 & 0.00104 & 0.348\\
 \addlinespace
-GO:0010811 & positive regulation of cell-substrate ad... & 79 & 72 & 61.69 & 0.00184 & 0.628\\
-GO:0060411 & cardiac septum morphogenesis & 33 & 32 & 25.77 & 0.00289 & 0.903\\
-GO:0060078 & regulation of postsynaptic membrane pote... & 69 & 63 & 53.88 & 0.00319 & 0.903\\
-GO:0042738 & exogenous drug catabolic process & 23 & 23 & 17.96 & 0.00335 & 0.903\\
-GO:0021515 & cell differentiation in spinal cord & 23 & 23 & 17.96 & 0.00335 & 0.903\\
+GO:0006163 & purine nucleotide metabolic process & 199 & 198 & 187.27 & 0.00108 & 0.348\\
+GO:0051301 & cell division & 300 & 294 & 282.31 & 0.00109 & 0.348\\
+GO:0070507 & regulation of microtubule cytoskeleton o... & 110 & 110 & 103.51 & 0.00120 & 0.361\\
+GO:0006091 & generation of precursor metabolites and ... & 222 & 220 & 208.91 & 0.00138 & 0.392\\
+GO:0006364 & rRNA processing & 106 & 106 & 99.75 & 0.00153 & 0.412\\
 \bottomrule
@@
 \begin{verbatim}
-## R version 3.6.0 (2019-04-26)
+## R version 4.0.2 (2020-06-22)
 ## Platform: x86_64-pc-linux-gnu (64-bit)
-## Running under: Ubuntu 19.04
+## Running under: Ubuntu 20.04 LTS
 ## 
 ## Matrix products: default
-## BLAS:   /usr/lib/x86_64-linux-gnu/blas/libblas.so.3.8.0
-## LAPACK: /usr/lib/x86_64-linux-gnu/lapack/liblapack.so.3.8.0
+## BLAS:   /usr/local/lib/R/lib/libRblas.so
+## LAPACK: /usr/local/lib/R/lib/libRlapack.so
 ## 
@@
 ## other attached packages:
-##  [1] RColorBrewer_1.1-2      moanin_0.0.0           
-##  [3] BiocWorkflowTools_1.6.2 ggfortify_0.4.6        
-##  [5] timecoursedata_0.0.0    topGO_2.32.0           
-##  [7] SparseM_1.77            GO.db_3.6.0            
-##  [9] AnnotationDbi_1.42.1    IRanges_2.14.12        
-## [11] S4Vectors_0.18.3        graph_1.58.2           
-## [13] viridis_0.5.1           viridisLite_0.3.0      
-## [15] KEGGprofile_1.22.0      RCurl_1.95-4.12        
-## [17] bitops_1.0-6            NMF_0.21.0             
-## [19] Biobase_2.40.0          BiocGenerics_0.26.0    
-## [21] cluster_2.0.9           rngtools_1.3.1.1       
-## [23] pkgmaker_0.27           registry_0.5-1         
-## [25] ggplot2_3.1.1           kableExtra_1.1.0       
-## [27] biomaRt_2.36.1          BiocStyle_2.8.2        
-## [29] knitr_1.22              limma_3.36.5           
-## [31] usethis_1.5.0           devtools_2.0.2         
-## [33] rmarkdown_1.12         
+##  [1] RColorBrewer_1.1-2          BiocWorkflowTools_1.14.0   
+##  [3] ggfortify_0.4.10            timecoursedata_0.99.0      
+##  [5] moanin_0.99.0-901           SummarizedExperiment_1.18.2
+##  [7] DelayedArray_0.14.1         matrixStats_0.56.0         
+##  [9] GenomicRanges_1.40.0        GenomeInfoDb_1.24.2        
+## [11] topGO_2.40.0                SparseM_1.78               
+## [13] GO.db_3.11.4                AnnotationDbi_1.50.1       
+## [15] IRanges_2.22.2              S4Vectors_0.26.1           
+## [17] graph_1.66.0                viridis_0.5.1              
+## [19] viridisLite_0.3.0           KEGGprofile_1.30.0         
+## [21] NMF_0.22.0                  Biobase_2.48.0             
+## [23] BiocGenerics_0.34.0         cluster_2.1.0              
+## [25] rngtools_1.5                pkgmaker_0.31.1            
+## [27] registry_0.5-1              ggplot2_3.3.2              
+## [29] kableExtra_1.1.0            biomaRt_2.44.1             
+## [31] BiocStyle_2.16.0            knitr_1.29                 
+## [33] limma_3.44.3                devtools_2.3.0             
+## [35] usethis_1.6.1               rmarkdown_2.3              
 ## 
 ## loaded via a namespace (and not attached):
-##  [1] colorspace_1.4-1   rprojroot_1.3-2    XVector_0.20.0    
-##  [4] fs_1.3.1           rstudioapi_0.10    remotes_2.0.4     
-##  [7] bit64_0.9-7        ClusterR_1.1.9     KEGG.db_3.2.3     
-## [10] xml2_1.3.2         codetools_0.2-16   doParallel_1.0.14 
-## [13] pkgload_1.0.2      ade4_1.7-13        gridBase_0.4-7    
-## [16] png_0.1-7          FD_1.0-12          readr_1.3.1       
-## [19] compiler_3.6.0     httr_1.4.0         backports_1.1.4   
-## [22] Matrix_1.2-17      assertthat_0.2.1   lazyeval_0.2.2    
-## [25] cli_1.1.0          htmltools_0.3.6    prettyunits_1.0.2 
-## [28] tools_3.6.0        gmp_0.5-13.5       gtable_0.3.0      
-## [31] glue_1.3.1         reshape2_1.4.3     dplyr_0.8.0.1     
-## [34] Rcpp_1.0.1         Biostrings_2.48.0  NMI_2.0           
-## [37] ape_5.3            nlme_3.1-139       iterators_1.0.10  
-## [40] xfun_0.6           stringr_1.4.0      ps_1.3.0          
-## [43] testthat_2.1.1     rvest_0.3.3        gtools_3.8.1      
-## [46] XML_3.98-1.19      zlibbioc_1.26.0    MASS_7.3-51.4     
-## [49] scales_1.0.0       hms_0.4.2          curl_3.3          
-## [52] yaml_2.2.0         memoise_1.1.0      gridExtra_2.3     
-## [55] TeachingDemos_2.10 stringi_1.4.3      RSQLite_2.1.1     
-## [58] desc_1.2.0         foreach_1.4.4      permute_0.9-5     
-## [61] pkgbuild_1.0.3     bibtex_0.4.2       geometry_0.4.1    
-## [64] rlang_0.4.6        pkgconfig_2.0.2    matrixStats_0.54.0
-## [67] evaluate_0.13      lattice_0.20-38    purrr_0.3.2       
-## [70] bit_1.1-14         processx_3.3.1     tidyselect_0.2.5  
-## [73] plyr_1.8.4         magrittr_1.5       bookdown_0.9      
-## [76] R6_2.4.0           DBI_1.0.0          mgcv_1.8-28       
-## [79] pillar_1.3.1       withr_2.1.2        abind_1.4-5       
-## [82] KEGGREST_1.20.2    tibble_2.1.1       crayon_1.3.4      
-## [85] progress_1.2.0     grid_3.6.0         vegan_2.5-4       
-## [88] blob_1.1.1         callr_3.2.0        git2r_0.25.2      
-## [91] digest_0.6.18      webshot_0.5.1      xtable_1.8-4      
-## [94] tidyr_0.8.3        munsell_0.5.0      magic_1.5-9       
-## [97] sessioninfo_1.1.1
+##   [1] colorspace_1.4-1       ellipsis_0.3.1         rprojroot_1.3-2       
+##   [4] XVector_0.28.0         fs_1.4.2               rstudioapi_0.11       
+##   [7] remotes_2.1.1          bit64_0.9-7.1          fansi_0.4.1           
+##  [10] ClusterR_1.2.2         KEGG.db_3.2.4          xml2_1.3.2            
+##  [13] codetools_0.2-16       doParallel_1.0.15      pkgload_1.1.0         
+##  [16] gridBase_0.4-7         dbplyr_1.4.4           png_0.1-7             
+##  [19] BiocManager_1.30.10    readr_1.3.1            compiler_4.0.2        
+##  [22] httr_1.4.1             backports_1.1.8        Matrix_1.2-18         
+##  [25] assertthat_0.2.1       cli_2.0.2              htmltools_0.5.0       
+##  [28] prettyunits_1.1.1      tools_4.0.2            gmp_0.6-0             
+##  [31] GenomeInfoDbData_1.2.3 gtable_0.3.0           glue_1.4.1            
+##  [34] reshape2_1.4.4         dplyr_1.0.0            rappdirs_0.3.1        
+##  [37] Rcpp_1.0.5             vctrs_0.3.1            Biostrings_2.56.0     
+##  [40] NMI_2.0                iterators_1.0.12       xfun_0.15             
+##  [43] stringr_1.4.0          ps_1.3.3               testthat_2.3.2        
+##  [46] rvest_0.3.5            lifecycle_0.2.0        gtools_3.8.2          
+##  [49] XML_3.99-0.4           edgeR_3.30.3           MASS_7.3-51.6         
+##  [52] zlibbioc_1.34.0        scales_1.1.1           hms_0.5.3             
+##  [55] yaml_2.2.1             curl_4.3               memoise_1.1.0         
+##  [58] gridExtra_2.3          TeachingDemos_2.12     stringi_1.4.6         
+##  [61] RSQLite_2.2.0          desc_1.2.0             foreach_1.5.0         
+##  [64] pkgbuild_1.1.0         bibtex_0.4.2.2         rlang_0.4.7           
+##  [67] pkgconfig_2.0.3        bitops_1.0-6           lattice_0.20-41       
+##  [70] evaluate_0.14          purrr_0.3.4            bit_1.1-15.2          
+##  [73] processx_3.4.3         tidyselect_1.1.0       plyr_1.8.6            
+##  [76] magrittr_1.5           bookdown_0.20          R6_2.4.1              
+##  [79] generics_0.0.2         DBI_1.1.0              pillar_1.4.6          
+##  [82] withr_2.2.0            KEGGREST_1.28.0        RCurl_1.98-1.2        
+##  [85] tibble_3.0.3           crayon_1.3.4           BiocFileCache_1.12.0  
+##  [88] progress_1.2.2         locfit_1.5-9.4         grid_4.0.2            
+##  [91] blob_1.2.1             callr_3.4.3            git2r_0.27.1          
+##  [94] digest_0.6.25          webshot_0.5.2          xtable_1.8-4          
+##  [97] tidyr_1.1.0            openssl_1.4.2          munsell_0.5.0         
+## [100] sessioninfo_1.1.1      askpass_1.1
 \end{verbatim}

### 2020-07-21 scripts/manuscript.Rmd
uncomment

@@
 # FIXME This line fails
-# dataToPlot = rbind(
-#     assay(de_moanin_model[cluster2Genes[3:6],]), 
-#    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
-# plot_splines_data(de_moanin_model, data=dataToPlot, 
-#     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
-#    mar=c(1.5,2.5,2,0.1))
+dataToPlot = rbind(
+     assay(de_moanin_model[cluster2Genes[3:6],]), 
+    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
+plot_splines_data(de_moanin_model, data=dataToPlot, 
+     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
+    mar=c(1.5,2.5,2,0.1))
 ```
@@
 # FIXME this lines failes.
-# dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
-#     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
-# plot_splines_data(moanin_model, data=dataToPlot, 
-#     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
-#     mar=c(1.5,2.5,2,0.1))
+dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
+     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
+plot_splines_data(moanin_model, data=dataToPlot, 
+     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+     mar=c(1.5,2.5,2,0.1))
 ```
@@
 # FIXME this line doesn't work
-# aheatmap(
-#    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
-#    Colv=NA,
-#    Rowv=NA,
-#    annCol=orderedMeta[, c("Group", "Timepoint")],
-#    annLegend=TRUE,
-#    annColors=ann_colors,
-#    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
-#    treeheight=0)
+aheatmap(
+   rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
+   Colv=NA,
+   Rowv=NA,
+   annCol=orderedMeta[, c("Group", "Timepoint")],
+   annLegend=TRUE,
+   annColors=ann_colors,
+   main=paste("Cluster", cluster_to_plot, "(rescaled)"),
+   treeheight=0)
 

### 2020-07-22 scripts/manuscript.Rmd
MAINT the rbind now seems to work locally

--- 2020-07-20 scripts/manuscript.Rmd
@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot=rbind(assay(de_moanin_model[cluster2Genes[3:6],]), 
+dataToPlot = rbind(
+    assay(de_moanin_model[cluster2Genes[3:6],]), 
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
 plot_splines_data(de_moanin_model, data=dataToPlot, 
-    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
+    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
     mar=c(1.5,2.5,2,0.1))

### 2020-07-22 scripts/manuscript.Rmd
MAINT rescale_values also seems to work locally

@@
 # Now use the results of the previous call to aheatmap to reorder the genes.
-# FIXME this line doesn't work
-# aheatmap(
-#    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
-#    Colv=NA,
-#    Rowv=NA,
-#    annCol=orderedMeta[, c("Group", "Timepoint")],
-#    annLegend=TRUE,
-#    annColors=ann_colors,
-#    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
-#    treeheight=0)
+aheatmap(
+    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
+    Colv=NA,
+    Rowv=NA,
+    annCol=orderedMeta[, c("Group", "Timepoint")],
+    annLegend=TRUE,
+    annColors=ann_colors,
+    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
+    treeheight=0)
 

### 2020-07-22 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-07-21 12:22:19.
+The manuscript was last compile on 2020-07-22 10:25:43.
 
@@
 \NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\CommentTok{# }\AlertTok{FIXME}\CommentTok{ This line fails}
-\CommentTok{# dataToPlot = rbind(}
-\CommentTok{#     assay(de_moanin_model[cluster2Genes[3:6],]), }
-\CommentTok{#    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])}
-\CommentTok{# plot_splines_data(de_moanin_model, data=dataToPlot, }
-\CommentTok{#     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,}
-\CommentTok{#    mar=c(1.5,2.5,2,0.1))}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
+    \KeywordTok{assay}\NormalTok{(de_moanin_model[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{],]), }
+    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,])}
+\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{simpleY=}\OtherTok{FALSE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}
 
@@
 \NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
-\CommentTok{# }\AlertTok{FIXME}\CommentTok{ this lines failes.}
-\CommentTok{# dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),}
-\CommentTok{#     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])}
-\CommentTok{# plot_splines_data(moanin_model, data=dataToPlot, }
-\CommentTok{#     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,}
-\CommentTok{#     mar=c(1.5,2.5,2,0.1))}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
+    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
+\KeywordTok{plot_splines_data}\NormalTok{(moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
+     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
+     \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topScoreInCluster2-1}
+
+}
+
+\caption{Top scoring genes in Cluster 2}\label{fig:topScoreInCluster2}
+\end{figure}
 
@@
 \CommentTok{# Now use the results of the previous call to aheatmap to reorder the genes.}
-\CommentTok{# }\AlertTok{FIXME}\CommentTok{ this line doesn't work}
-\CommentTok{# aheatmap(}
-\CommentTok{#    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],}
-\CommentTok{#    Colv=NA,}
-\CommentTok{#    Rowv=NA,}
-\CommentTok{#    annCol=orderedMeta[, c("Group", "Timepoint")],}
-\CommentTok{#    annLegend=TRUE,}
-\CommentTok{#    annColors=ann_colors,}
-\CommentTok{#    main=paste("Cluster", cluster_to_plot, "(rescaled)"),}
-\CommentTok{#    treeheight=0)}
+\KeywordTok{aheatmap}\NormalTok{(}
+    \KeywordTok{rescale_values}\NormalTok{(moanin_model[genes_to_plot,ord])[res}\OperatorTok{$}\NormalTok{rowInd,],}
+    \DataTypeTok{Colv=}\OtherTok{NA}\NormalTok{,}
+    \DataTypeTok{Rowv=}\OtherTok{NA}\NormalTok{,}
+    \DataTypeTok{annCol=}\NormalTok{orderedMeta[, }\KeywordTok{c}\NormalTok{(}\StringTok{"Group"}\NormalTok{, }\StringTok{"Timepoint"}\NormalTok{)],}
+    \DataTypeTok{annLegend=}\OtherTok{TRUE}\NormalTok{,}
+    \DataTypeTok{annColors=}\NormalTok{ann_colors,}
+    \DataTypeTok{main=}\KeywordTok{paste}\NormalTok{(}\StringTok{"Cluster"}\NormalTok{, cluster_to_plot, }\StringTok{"(rescaled)"}\NormalTok{),}
+    \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{)}
 \end{Highlighting}
@@
 ##  [1] RColorBrewer_1.1-2          BiocWorkflowTools_1.14.0   
-##  [3] ggfortify_0.4.10            timecoursedata_0.99.0      
+##  [3] ggfortify_0.4.10            timecoursedata_0.99.13     
 ##  [5] moanin_0.99.0-901           SummarizedExperiment_1.18.2

### 2020-07-27 scripts/manuscript.Rmd
Small fixes here and there for missing references«

@@
     methodology used to analyse time-course data: (1) quality control and normalization of the
-    data set; (2) differential expression analysis using functional
+    dataset; (2) differential expression analysis using functional
     data analysis; (3) clustering of time-course data; (4) interpreting
@@
 strategies to combine single time point differential expression analysis
-[@sun:multiple]. However, with long time course data sets, particularly in
+[@sun:multiple]. However, with long time course datasets, particularly in
 developmental systems, a massive number of genes will show *some* change. For
@@
 
-While we showcase this pipeline on micro-array data, [@varoquaux:lifecycle]
+While we showcase this pipeline on micro-array data,
+[@varoquaux:transcriptomic]
 leverages a similar set of steps to analyse RNA-seq data of the lifetime
@@
 changing over time) or on multiple conditions (such as the influenza
-data set we are considering), which will alter slightly the types of questions
+dataset we are considering), which will alter slightly the types of questions
 we are interested in.
@@
 
-```{r topPvalueGenes, fig.height=6, fig.width=6}
+```{r topPvalueGenes, fig.height=6, fig.width=6, fig.cap="Top 10 genes with the smallest p-values"}
 top_DE_genes_pval = names(sort(pvalue)[1:10])
@@
 
-```{r topLogFold,fig.height=6, fig.width=6}
+```{r topLogFold,fig.height=6, fig.width=6, fig.cap="Top 10 genes with largest absolute timecourse log-fold change"}
 top_DE_genes_lfc = names(
@@
 
-```{r fig.height=6, fig.width=6}
+```{r fig.height=6, fig.width=6, fig.cap="K-means centroids"}
 plot_splines_data(de_moanin_model,
@@
 
-```{r genesVsCentroids, fig.height=6, fig.width=6}
+```{r genesVsCentroids, fig.height=6, fig.width=6, fig.cap="Genes from cluster 2"}
 cluster_of_interest = 2
@@
 cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
-We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
+We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{fig:visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
 We can do this by rerunning `splines_kmeans_score_and_label` and setting the filter `percentage_genes_to_label=1` (and we can speed up the calculation by giving our previously calculated score via the argument `previous_scores`)
 
-```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster.
+```{r visualizeScores, fig.height=6, fig.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster.
 The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
@@
 differences between the labels provided by the splines k-means model and our
-scoring and labeling step, in terms of the number of genes assigned to each cluster (Figure \ref{fig:kmeansVsScore}.
+scoring and labeling step, in terms of the number of genes assigned to each cluster (Figure \ref{fig:kmeansVsScore}).
 
@@
 
-We plot the WCSS for $k=2-20$ here (Figure \ref{fig:elbow_method}). 
+We plot the WCSS for $k=2-20$ here (Figure \ref{fig:elbowMethod}). 
 We see that as expected the WCSS continues
@@
 The silhouette value is a measure of how similar a data point is to its own
-cluster (cohesion) compared to other clusters (separation), shown in Figure \ref{fig:average_silhouette_method}.
+cluster (cohesion) compared to other clusters (separation), shown in Figure \ref{fig:averageSilhouetteMethod}.
 
@@
 
-```{r bootstrapConsensus, fig.width=6, fig.height=4,fig.cap="Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20."}
+```{r bootstrapConsensus, fig.width=6, fig.height=4, fig.cap="Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20."}
+par(mfrow=c(2, 1))
 consensus_matrix_stability_5 = consensus_matrix(stability_5,

### 2020-07-27 scripts/manuscript.Rmd
FIX cluster 8 -> cluster 6

@@
 
-**Assigning cluster labels:** We could just assign each gene to a cluster based on which cluster gave the minimum score.
-By default, `splines_kmeans_score_and_label` does not do that, but rather requires a sufficiently low enough goodness-of-fit score.
-The criteria for being "sufficiently low" is based on looking at the distribution of the scores of all genes on all clusters (i.e.
-the entire `scores` matrix returned by `splines_kmeans_score_and_label`).
-A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment (this choice can be changed by `proportion_genes_to_label`).
-Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
+**Assigning cluster labels:** We could just assign each gene to a cluster
+based on which cluster gave the minimum score. By default,
+`splines_kmeans_score_and_label` does not do that, but rather requires a
+sufficiently low enough goodness-of-fit score. The criteria for being
+"sufficiently low" is based on looking at the distribution of the scores of
+all genes on all clusters (i.e. the entire `scores` matrix returned by
+`splines_kmeans_score_and_label`). A gene is then assigned to a cluster only
+if their best score is above the 50% percentile of that distribution, with the
+remaining genes getting `NA` as their assignment (this choice can be changed
+by `proportion_genes_to_label`). Note that because the distribution of scores
+of genes across *all* clusters are used, this is not equivalent to assigning
+50% of genes to a cluster -- it is possible that all genes are assigned to a
+cluster.
 
@@
 
-We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
-cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
-We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{fig:visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
-We can do this by rerunning `splines_kmeans_score_and_label` and setting the filter `percentage_genes_to_label=1` (and we can speed up the calculation by giving our previously calculated score via the argument `previous_scores`)
+We can visualize the impact of this filtering process by considering the
+distribution of goodness-of-fit scores for each cluster if we did not have the
+filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign
+every gene to a cluster based on which gives them their minimum score. We
+display for each cluster, the scores of the genes assigned to that cluster
+(Figure \ref{fig:visualizeScores}), as compared to the cutoff value for what
+that score must be under our filtering procedure. We can do this by rerunning
+`splines_kmeans_score_and_label` and setting the filter
+`percentage_genes_to_label=1` (and we can speed up the calculation by giving
+our previously calculated score via the argument `previous_scores`)
 
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind(assay(moanin_model[cluster2Score[1:4], ]),
+dataToPlot = rbind(
+    assay(moanin_model[cluster2Score[1:4], ]),
     "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
-plot_splines_data(moanin_model, data=dataToPlot, 
-     colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
-     mar=c(1.5,2.5,2,0.1))
-```
-
-```{r results="hide", echo=FALSE}
+plot_splines_data(
+    moanin_model, data=dataToPlot, 
+    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
+    mar=c(1.5, 2.5, 2, 0.1))
+```
+
+```{r saving_results, results="hide", echo=FALSE}
 # Save the clustering for the GO term and pathway enrichment on the whole set
@@
 
-Now, let us look more in detail some specific clusters. Cluster 8 seems
+Now, let us look more in detail some specific clusters. Cluster 6 seems
 particularly interesting: it captures genes with strong differences between
@@
 
-First, we select the genes of interest, namely those in cluster 8, based on
+First, we select the genes of interest, namely those in cluster 6, based on
 our goodness-of-fit assignment.
@@
 ```{r}
-cluster_to_plot = 8
+cluster_to_plot = 6
 genes_to_plot =  names(labels[labels == cluster_to_plot])
@@
 ```{r bootstrapConsensus, fig.width=6, fig.height=4, fig.cap="Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20."}
-par(mfrow=c(2, 1))
+par(mfrow=c(1, 2))
 consensus_matrix_stability_5 = consensus_matrix(stability_5,
@@
 
-From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that $k=5$ is more stable than $k=3$ and $k=4$,
-but not as stable as $k=2$. The model explorer strategy, in addition to
-visualizing the diversity of the centroids, can thus help assessing an
-adequate number of clusters.
+From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that $k=5$
+is more stable than $k=3$ and $k=4$, but not as stable as $k=2$. The model
+explorer strategy, in addition to visualizing the diversity of the centroids,
+can thus help assessing an adequate number of clusters.
 
@@
 
-This analysis doesn't necessarily pick a particular $k$, but can help decide between $k$ within a desired range, for example, or to avoid $k$ that degrade the stability.
+This analysis doesn't necessarily pick a particular $k$, but can help decide
+between $k$ within a desired range, for example, or to avoid $k$ that degrade
+the stability.
 
@@
 First, let us clean up the genes we work with and only select the genes we are
-going to use in the enrichment analysis. One can either all genes assigned to the cluster, only the set of differentially expressed genes in each cluster, or the
-subset of genes that fit well to a cluster (based on some criterion).
+going to use in the enrichment analysis. One can either all genes assigned to
+the cluster, only the set of differentially expressed genes in each cluster,
+or the subset of genes that fit well to a cluster (based on some criterion).
 
@@
 Let us first tackle the case of pathway enrichment analysis. We will leverage
-the packages `biomaRt` [@durinck:biomart] and `KEGGprofile` [@zhao:keggprofile]
-for this step. `KEGGprofile` is a
-package that facilitates enrichment analysis on a set of
-genes labeled with the Ensembl annotation [@ensembl2020] based on the set of biological pathways described in the KEGG database[@KEGG].
-
-We thus need to convert the gene
-names, which are given in the Refseq annotation, into the corresponding Ensembl name. This is where `biomaRt` comes in handy: it
-enables easy conversion from one gene annotation to another. Here, we will use the function `getBM`
-in `biomaRt` to convert the gene names from cluster 8.
+the packages `biomaRt` [@durinck:biomart] and `KEGGprofile`
+[@zhao:keggprofile] for this step. `KEGGprofile` is a package that facilitates
+enrichment analysis on a set of genes labeled with the Ensembl annotation
+[@ensembl2020] based on the set of biological pathways described in the KEGG
+database[@KEGG].
+
+We thus need to convert the gene names, which are given in the Refseq
+annotation, into the corresponding Ensembl name. This is where `biomaRt` comes
+in handy: it enables easy conversion from one gene annotation to another.
+Here, we will use the function `getBM` in `biomaRt` to convert the gene names
+from cluster 8.
 
@@
 ## Get gene names of genes in Cluster 8
-cluster = 8
-labels = unlist(labels)
+cluster = 6
 gene_names = names(labels)
@@
 
-Then we use the function `find_enriched_pathway` in the `KEGGprofile` package to determine whether any KEGG pathways are enriched in cluster 8, i.e.
-whether a higher percentage of genes from a single pathway are found in cluster 8 than we would expect by simply proportional assignment of genes to clusters.
+Then we use the function `find_enriched_pathway` in the `KEGGprofile` package
+to determine whether any KEGG pathways are enriched in cluster 6, i.e. whether
+a higher percentage of genes from a single pathway are found in cluster 6 than
+we would expect by simply proportional assignment of genes to clusters.
 
@@
 
-**EAP: I changed this to be the min(10, nrow(pathways)) but wondering if there's a problem, since presumably there used to be more than 10 pathways found significant and on my machine there are only 2?**
-
 ### Finding enriched GO terms
 
-The Gene Ontology (GO) database [@GO], also categorizes genes into meaningful biological ontologies and can be used for enrichment analysis via the package `topGO` [@alexa:topgo].
-We again use `biomaRt` to find the mapping between genes and the GO terms to which they match.
+The Gene Ontology (GO) database [@GO], also categorizes genes into meaningful
+biological ontologies and can be used for enrichment analysis via the package
+`topGO` [@alexa:topgo]. We again use `biomaRt` to find the mapping between
+genes and the GO terms to which they match.
 
@@
 ```{r go_term_enrichment, results="hide", message=FALSE}
-# Create logical vector of whether in cluster 8
+# Create logical vector of whether in cluster 6
 assignments = labels == cluster

### 2020-07-27 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 However, analysing time-course transcriptomic data raises unique challenging statistical and computational questions, requiring the development of novel methods and software.
-Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the data set; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
+Using as case study time-course transcriptomics data from mice exposed to different strains of influenza, this workflow provides a step-by-step tutorial of the methodology used to analyse time-course data: (1) quality control and normalization of the dataset; (2) differential expression analysis using functional data analysis; (3) clustering of time-course data; (4) interpreting clusters with GO term and KEGG pathway enrichment analysis.
 \end{abstract}
@@
 
-The manuscript was last compile on 2020-07-22 10:25:43.
+The manuscript was last compile on 2020-07-27 16:08:41.
 
@@
 strategies to combine single time point differential expression analysis
-\citep{sun:multiple}. However, with long time course data sets, particularly in
+\citep{sun:multiple}. However, with long time course datasets, particularly in
 developmental systems, a massive number of genes will show \emph{some} change. For
@@
 
-While we showcase this pipeline on micro-array data, \citep{varoquaux:lifecycle}
+While we showcase this pipeline on micro-array data,
+\citep{varoquaux:transcriptomic}
 leverages a similar set of steps to analyse RNA-seq data of the lifetime
@@
 
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topPvalueGenes-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topPvalueGenes-1}
+
+}
+
+\caption{Top 10 genes with the smallest p-values}\label{fig:topPvalueGenes}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topLogFold-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/topLogFold-1}
+
+}
+
+\caption{Top 10 genes with largest absolute timecourse log-fold change}\label{fig:topLogFold}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1}
+
+}
+
+\caption{K-means centroids}\label{fig:unnamed-chunk-18}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1} \end{center}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1}
+
+}
+
+\caption{Genes from cluster 2}\label{fig:genesVsCentroids}
+\end{figure}
 
@@
 
-\textbf{Assigning cluster labels:} We could just assign each gene to a cluster based on which cluster gave the minimum score.
-By default, \texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a sufficiently low enough goodness-of-fit score.
-The criteria for being ``sufficiently low'' is based on looking at the distribution of the scores of all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by \texttt{splines\_kmeans\_score\_and\_label}).
-A gene is then assigned to a cluster only if their best score is above the 50\% percentile of that distribution, with the remaining genes getting \texttt{NA} as their assignment (this choice can be changed by \texttt{proportion\_genes\_to\_label}).
-Note that because the distribution of scores of genes across \emph{all} clusters are used, this is not equivalent to assigning 50\% of genes to a cluster -- it is possible that all genes are assigned to a cluster.
+\textbf{Assigning cluster labels:} We could just assign each gene to a cluster
+based on which cluster gave the minimum score. By default,
+\texttt{splines\_kmeans\_score\_and\_label} does not do that, but rather requires a
+sufficiently low enough goodness-of-fit score. The criteria for being
+``sufficiently low'' is based on looking at the distribution of the scores of
+all genes on all clusters (i.e.~the entire \texttt{scores} matrix returned by
+\texttt{splines\_kmeans\_score\_and\_label}). A gene is then assigned to a cluster only
+if their best score is above the 50\% percentile of that distribution, with the
+remaining genes getting \texttt{NA} as their assignment (this choice can be changed
+by \texttt{proportion\_genes\_to\_label}). Note that because the distribution of scores
+of genes across \emph{all} clusters are used, this is not equivalent to assigning
+50\% of genes to a cluster -- it is possible that all genes are assigned to a
+cluster.
 
@@
 
-We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
-cluster if we did not have the filtering cutoff of \texttt{splines\_kmeans\_score\_and\_label}, i.e.~if we simply assign every gene to a cluster based on which gives them their minimum score.
-We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.
-We can do this by rerunning \texttt{splines\_kmeans\_score\_and\_label} and setting the filter \texttt{percentage\_genes\_to\_label=1} (and we can speed up the calculation by giving our previously calculated score via the argument \texttt{previous\_scores})
+We can visualize the impact of this filtering process by considering the
+distribution of goodness-of-fit scores for each cluster if we did not have the
+filtering cutoff of \texttt{splines\_kmeans\_score\_and\_label}, i.e.~if we simply assign
+every gene to a cluster based on which gives them their minimum score. We
+display for each cluster, the scores of the genes assigned to that cluster
+(Figure \ref{fig:visualizeScores}), as compared to the cutoff value for what
+that score must be under our filtering procedure. We can do this by rerunning
+\texttt{splines\_kmeans\_score\_and\_label} and setting the filter
+\texttt{percentage\_genes\_to\_label=1} (and we can speed up the calculation by giving
+our previously calculated score via the argument \texttt{previous\_scores})
 
@@
 \NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
+\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
+    \KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
     \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
-\KeywordTok{plot_splines_data}\NormalTok{(moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
-     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
-     \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\KeywordTok{plot_splines_data}\NormalTok{(}
+\NormalTok{    moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
+    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{simpleY=}\OtherTok{FALSE}\NormalTok{,}
+    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{, }\FloatTok{0.1}\NormalTok{))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{cluster_to_plot =}\StringTok{ }\DecValTok{8}
+\NormalTok{cluster_to_plot =}\StringTok{ }\DecValTok{6}
 \NormalTok{genes_to_plot =}\StringTok{  }\KeywordTok{names}\NormalTok{(labels[labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster_to_plot])}
@@
 
-Now we will create the heatmaps of these genes (1283 genes, Figure \ref{fig:heatmapCluster8}).
+Now we will create the heatmaps of these genes (3746 genes, Figure \ref{fig:heatmapCluster8}).
 
@@
 
-We plot the WCSS for \(k=2-20\) here (Figure \ref{fig:elbow_method}).
+We plot the WCSS for \(k=2-20\) here (Figure \ref{fig:elbowMethod}).
 We see that as expected the WCSS continues
@@
 \begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
 \NormalTok{consensus_matrix_stability_}\DecValTok{5}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{5}\NormalTok{,}
@@
      \DataTypeTok{treeheight=}\DecValTok{0}\NormalTok{, }\DataTypeTok{main=}\StringTok{"K=5"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-1}
-
-}
-
-\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-1}
-\end{figure}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+
 \NormalTok{consensus_matrix_stability_}\DecValTok{20}\NormalTok{ =}\StringTok{ }\KeywordTok{consensus_matrix}\NormalTok{(stability_}\DecValTok{20}\NormalTok{,}
@@
 
-{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-2}
-
-}
-
-\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus-2}
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/bootstrapConsensus-1}
+
+}
+
+\caption{Consensus matrix of proportion of times samples were in the same cluster over bootstrap samples for k=5 and 20.}\label{fig:bootstrapConsensus}
 \end{figure}
@@
 
-From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\) is more stable than \(k=3\) and \(k=4\),
-but not as stable as \(k=2\). The model explorer strategy, in addition to
-visualizing the diversity of the centroids, can thus help assessing an
-adequate number of clusters.
+From this plot (Figure \ref{fig:plotModelExplorer}), we can deduce that \(k=5\)
+is more stable than \(k=3\) and \(k=4\), but not as stable as \(k=2\). The model
+explorer strategy, in addition to visualizing the diversity of the centroids,
+can thus help assessing an adequate number of clusters.
 
@@
 
-This analysis doesn't necessarily pick a particular \(k\), but can help decide between \(k\) within a desired range, for example, or to avoid \(k\) that degrade the stability.
+This analysis doesn't necessarily pick a particular \(k\), but can help decide
+between \(k\) within a desired range, for example, or to avoid \(k\) that degrade
+the stability.
 
@@
 Let us first tackle the case of pathway enrichment analysis. We will leverage
-the packages \texttt{biomaRt} \citep{durinck:biomart} and \texttt{KEGGprofile} \citep{zhao:keggprofile}
-for this step. \texttt{KEGGprofile} is a
-package that facilitates enrichment analysis on a set of
-genes labeled with the Ensembl annotation \citep{ensembl2020} based on the set of biological pathways described in the KEGG database\citep{KEGG}.
-
-We thus need to convert the gene
-names, which are given in the Refseq annotation, into the corresponding Ensembl name. This is where \texttt{biomaRt} comes in handy: it
-enables easy conversion from one gene annotation to another. Here, we will use the function \texttt{getBM}
-in \texttt{biomaRt} to convert the gene names from cluster 8.
+the packages \texttt{biomaRt} \citep{durinck:biomart} and \texttt{KEGGprofile}
+\citep{zhao:keggprofile} for this step. \texttt{KEGGprofile} is a package that facilitates
+enrichment analysis on a set of genes labeled with the Ensembl annotation
+\citep{ensembl2020} based on the set of biological pathways described in the KEGG
+database\citep{KEGG}.
+
+We thus need to convert the gene names, which are given in the Refseq
+annotation, into the corresponding Ensembl name. This is where \texttt{biomaRt} comes
+in handy: it enables easy conversion from one gene annotation to another.
+Here, we will use the function \texttt{getBM} in \texttt{biomaRt} to convert the gene names
+from cluster 8.
 
@@
 \CommentTok{## Get gene names of genes in Cluster 8}
-\NormalTok{cluster =}\StringTok{ }\DecValTok{8}
-\NormalTok{labels =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(labels)}
+\NormalTok{cluster =}\StringTok{ }\DecValTok{6}
 \NormalTok{gene_names =}\StringTok{ }\KeywordTok{names}\NormalTok{(labels)}
@@
 
-Then we use the function \texttt{find\_enriched\_pathway} in the \texttt{KEGGprofile} package to determine whether any KEGG pathways are enriched in cluster 8, i.e.~whether a higher percentage of genes from a single pathway are found in cluster 8 than we would expect by simply proportional assignment of genes to clusters.
+Then we use the function \texttt{find\_enriched\_pathway} in the \texttt{KEGGprofile} package
+to determine whether any KEGG pathways are enriched in cluster 6, i.e.~whether
+a higher percentage of genes from a single pathway are found in cluster 6 than
+we would expect by simply proportional assignment of genes to clusters.
 
@@
 \midrule
-04740 & Olfactory transduction & 8 & 0\\
-04080 & Neuroactive ligand-receptor interaction & 10 & 0\\
+04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
+04620 & Toll-like receptor signaling pathway & 39 & 0\\
+04621 & NOD-like receptor signaling pathway & 45 & 0\\
+04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
+04630 & Jak-STAT signaling pathway & 28 & 0\\
+\addlinespace
+04210 & Apoptosis & 35 & 0\\
+04380 & Osteoclast differentiation & 31 & 0\\
+04622 & RIG-I-like receptor signaling pathway & 38 & 0\\
+05160 & Hepatitis C & 28 & 0\\
+04940 & Type I diabetes mellitus & 35 & 0\\
 \bottomrule
@@
 
-\textbf{EAP: I changed this to be the min(10, nrow(pathways)) but wondering if there's a problem, since presumably there used to be more than 10 pathways found significant and on my machine there are only 2?}
-
 \hypertarget{finding-enriched-go-terms}{%
@@
 
-The Gene Ontology (GO) database \citep{GO}, also categorizes genes into meaningful biological ontologies and can be used for enrichment analysis via the package \texttt{topGO} \citep{alexa:topgo}.
-We again use \texttt{biomaRt} to find the mapping between genes and the GO terms to which they match.
+The Gene Ontology (GO) database \citep{GO}, also categorizes genes into meaningful
+biological ontologies and can be used for enrichment analysis via the package
+\texttt{topGO} \citep{alexa:topgo}. We again use \texttt{biomaRt} to find the mapping between
+genes and the GO terms to which they match.
 
@@
 \begin{Highlighting}[]
-\CommentTok{# Create logical vector of whether in cluster 8}
+\CommentTok{# Create logical vector of whether in cluster 6}
 \NormalTok{assignments =}\StringTok{ }\NormalTok{labels }\OperatorTok{==}\StringTok{ }\NormalTok{cluster}
@@
 \midrule
-GO:0044260 & cellular macromolecule metabolic process & 3762 & 3607 & 3540.21 & 0.00031 & 0.276\\
-GO:0043603 & cellular amide metabolic process & 467 & 458 & 439.47 & 0.00036 & 0.276\\
-GO:0043161 & proteasome-mediated ubiquitin-dependent ... & 201 & 199 & 189.15 & 0.00042 & 0.276\\
-GO:0016071 & mRNA metabolic process & 285 & 280 & 268.20 & 0.00052 & 0.276\\
-GO:0070925 & organelle assembly & 427 & 417 & 401.83 & 0.00052 & 0.276\\
+GO:0019395 & fatty acid oxidation & 55 & 54 & 44.78 & 0.00016 & 0.164\\
+GO:1905114 & cell surface receptor signaling pathway ... & 272 & 243 & 221.46 & 0.00021 & 0.179\\
+GO:0003002 & regionalization & 174 & 158 & 141.67 & 0.00049 & 0.358\\
+GO:0009755 & hormone-mediated signaling pathway & 82 & 77 & 66.76 & 0.00105 & 0.671\\
+GO:0048562 & embryonic organ morphogenesis & 154 & 139 & 125.39 & 0.00175 & 0.994\\
 \addlinespace
-GO:0019058 & viral life cycle & 123 & 123 & 115.75 & 0.00054 & 0.276\\
-GO:0018130 & heterocycle biosynthetic process & 1809 & 1733 & 1702.35 & 0.00093 & 0.348\\
-GO:0019438 & aromatic compound biosynthetic process & 1810 & 1734 & 1703.29 & 0.00098 & 0.348\\
-GO:0042110 & T cell activation & 273 & 268 & 256.91 & 0.00102 & 0.348\\
-GO:0006954 & inflammatory response & 410 & 398 & 385.83 & 0.00104 & 0.348\\
+GO:0006805 & xenobiotic metabolic process & 50 & 48 & 40.71 & 0.00257 & 1.000\\
+GO:0090596 & sensory organ morphogenesis & 138 & 124 & 112.36 & 0.00332 & 1.000\\
+GO:0035108 & limb morphogenesis & 74 & 69 & 60.25 & 0.00337 & 1.000\\
+GO:0030166 & proteoglycan biosynthetic process & 27 & 27 & 21.98 & 0.00385 & 1.000\\
+GO:0003206 & cardiac chamber morphogenesis & 72 & 67 & 58.62 & 0.00448 & 1.000\\
 \addlinespace
-GO:0006163 & purine nucleotide metabolic process & 199 & 198 & 187.27 & 0.00108 & 0.348\\
-GO:0051301 & cell division & 300 & 294 & 282.31 & 0.00109 & 0.348\\
-GO:0070507 & regulation of microtubule cytoskeleton o... & 110 & 110 & 103.51 & 0.00120 & 0.361\\
-GO:0006091 & generation of precursor metabolites and ... & 222 & 220 & 208.91 & 0.00138 & 0.392\\
-GO:0006364 & rRNA processing & 106 & 106 & 99.75 & 0.00153 & 0.412\\
+GO:0055114 & oxidation-reduction process & 536 & 472 & 436.41 & 0.00489 & 1.000\\
+GO:0021954 & central nervous system neuron developmen... & 36 & 35 & 29.31 & 0.00557 & 1.000\\
+GO:0006261 & DNA-dependent DNA replication & 70 & 64 & 56.99 & 0.00595 & 1.000\\
+GO:0006633 & fatty acid biosynthetic process & 78 & 70 & 63.51 & 0.00686 & 1.000\\
+GO:0006790 & sulfur compound metabolic process & 154 & 136 & 125.39 & 0.00710 & 1.000\\
 \bottomrule
@@
 ##  [1] RColorBrewer_1.1-2          BiocWorkflowTools_1.14.0   
-##  [3] ggfortify_0.4.10            timecoursedata_0.99.13     
+##  [3] ggfortify_0.4.10            timecoursedata_0.99.15     
 ##  [5] moanin_0.99.0-901           SummarizedExperiment_1.18.2

### 2020-07-29 scripts/manuscript.Rmd

@@
     treeheight=0)
-
 ```

### 2020-07-30 scripts/manuscript.Rmd
Edits + rebase here and there

@@
     %\VignetteEngine{knitr::rmarkdown}
-output: BiocWorkflowTools::f1000_article
+header-includes:
+   - \usepackage{float}
+output:
+    BiocWorkflowTools::f1000_article: default
+    ms_document: default
 bibliography: bibliography.bib
+always_allow_html: yes
 ---
@@
 library(biomaRt)
-library(kableExtra)
+library(pander)
 
@@
 library(topGO)
-
+library(moanin)
+if(!packageVersion("moanin")>="0.99.0-902") stop("Must have version 0.99.0-902 to compile the manuscript")
 ```
@@
 ```
-
-The manuscript was last compile on `r Sys.time()`.
 
@@
 
- `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`, along with the corresponding package that contains time course datasets we will use:
+`moanin` and `timecoursedata` are available from bioconductor, and can be
+installed using the `install` function in the package `BiocManager`, along
+with the corresponding package that contains time course datasets we will use:
 
 ```{r eval=FALSE}
-library(devtools)
-install_github("NelleV/moanin")
-install_github("NelleV/timecoursedata")
+library(BiocManager)
+# Need to use the development version for now.
+BiocManager::install(version="devel")
+BiocManager::valid()
+
+BiocManager::install("timecoursedata")
+BiocManager::install("moanin")
 ```
@@
 ```{r eval=FALSE}
-BiocManager::install(c("NMF","ggfortify","topGO","biomaRt","KEGGprofile","BiocWorkflowTools"))
+BiocManager::install(
+    c("NMF", "ggfortify", "topGO", "biomaRt",
+      "KEGGprofile", "BiocWorkflowTools", "timecoursedata", ))
 ```
@@
 avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
-$10^5$ PFU of each virus. An adidtional 42 mice were injected with a lower dose
+$10^5$ PFU of each virus. An additional 42 mice were injected with a lower dose
 of the Vietnam avian influenza virus ($10^3$ PFU).
@@
 threshold of the virus concentration is  exceeded in the lung. Once this
-threshold is exceeded, a strong inflammatory and cytokin production occurs.
+threshold is exceeded, a strong inflammatory and cytokine production occurs.
 This results provides evidence that the pathology response is non-linearly
@@
 ```{r echo=FALSE}
-kable(head(meta))
+pander(head(meta))
 ```
@@
 define a series of markers (or plotting symbols) to distinguish replicate
-samples in scatter plots. We also reorder the factor `Group` which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).
+samples in scatter plots. We also reorder the factor `Group` which describes
+the treatments so that the treatments are ordered from low to high pathogeny
+(with Control being first).
 
@@
 variable genes: for this step, we keep only the top 50% most variable genes.
+In Figure \ref{fig:geneVarHist}, we plot the distribution of variance across
+all genes.
 
@@
 We can already see interesting patterns emerging from the correlation plot.
-First, the cross-correlation amongst samples taking from the control mice is
+First, the cross-correlation amongst samples taken from the control mice is
 higher than the cross correlation amongst the rest of the treatments. Second,
@@
 
-In both
-situations, we first need to set up a object (a `Moanin` object) to
+In both situations, we first need to set up an object (a `Moanin` object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, ...). The `Moanin` object
-will contain information used
-throughout this analysis, in particular the condition and timepoints of each sample and the
-the basis matrix for fitting splines models.
+(formula, number of degrees of freedom of the splines, ...). The `Moanin`
+object will contain information used throughout this analysis, in particular
+the condition and timepoints of each sample and the basis matrix for
+fitting splines models.
 
@@
 functional modeling. The metadata `data.frame` object should contain at least
-two columns: one named `Group`, containing the  treatement effect, and a
+two columns: one named `Group`, containing the  treatment effect, and a
 second one named `Timepoint` containing the timepoint information.
@@
 ```{r echo=FALSE}
-kable(weekly_de_analysis[1:10, 1:3 ], "latex", digits=3, booktabs=TRUE)
+pander(weekly_de_analysis[1:10, 1:3 ])
 
@@
 at 6H and then jump back to being differentially expressed at 9H. A more
-likely explanation is that there is are some technical or biological artifact
-about the samples for 6H that is creating higher variation and thus less
+likely explanation is that there are some technical or biological artifacts
+about the samples for 6H that are creating higher variation and thus less
 ability to detect significance.
@@
 The number of genes found differentially expressed ranges from around 12000 to
-29000 depending on the strain and dosage of influenza virus given to the mice.
+29000 depending on the strain and dosage of influenza virus given to the mice
+(Figure \ref{deTimeDist}).
 This corresponds to between 30% to 70% of the genes found differentially
@@
 ```{r echo=FALSE}
-kable(
-    head(log_fold_change_timepoints[, colnames(log_fold_change_timepoints)[1:5]]),
-    format="latex",
-    booktabs=TRUE)
+pander(
+    head(log_fold_change_timepoints[,
+    colnames(log_fold_change_timepoints)[1:5]]))
 ```
@@
 
-Note, however, that when a gene is not consistently up or down-regulated the estimation
-of the direction will not represent accurately the changes observed.
+Note, however, that when a gene is not consistently up- or down-regulated the
+estimation of the direction will not accurately represent the changes
+observed.
 
@@
 ```{r display_lfc, echo=FALSE}
-kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
-```
-
-In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
-the time-course data, for example, overall change versus the largest change.
+pander(log_fold_change_timecourse[1:5, ])
+```
+
+In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary
+of each of these methods against each other. We see that each method captures
+different elements of the time-course data, for example, overall change versus
+the largest change.
 
@@
 
-With a single measure of log-fold change and the p-value, we can now look
-at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the "timecourse" method of calculating log fold change.
+With a single measure of log-fold change and the p-value, we can now look at
+the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example
+of a volcano plot for the comparison of the control to the Kawasaki strain,
+using the "timecourse" method of calculating log fold change.
 
@@
 
-The very large number of genes found differentially expressed impair any
+The very large number of genes found differentially expressed impairs any
 interpretation one would attempt: with 70\% of the genome found differentially
@@
 To do this, we first
-aggregate all p-values obtained during the time-course differential expresison
+aggregate all p-values obtained during the time-course differential expression
 step in a single p-value using Fisher's method [@fisher:statistical] (`pvalues_fisher_method`).
@@
 clustering routine. As observed by looking at genes found differentially expressed, 
-many genes share a similar gene expression pattern, but on different scale.
+many genes share a similar gene expression pattern, but on different scales.
 
@@
 
-We then use the `plot_splines_data`  function, only now applied to the centroids, to visualize the centroids of
-each cluster obtained with the splines k-means model.
-
-```{r fig.height=6, fig.width=6, fig.cap="K-means centroids"}
+We then use the `plot_splines_data`  function, only now applied to the
+centroids, to visualize the centroids of each cluster obtained with the
+splines k-means model (Figure \ref{fig:centroids}).
+
+```{r centroids, fig.height=6, fig.width=6, fig.cap="K-means centroids"}
 plot_splines_data(de_moanin_model,
@@
 fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
-of the actual genes assigned to cluster 2 to see the difference:
+of the actual genes assigned to cluster 2 with the estimated centroid overlaid:
 
@@
     kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
-dataToPlot = rbind(
-    assay(de_moanin_model[cluster2Genes[3:6],]), 
-    "Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
-plot_splines_data(de_moanin_model, data=dataToPlot, 
-    colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
-    mar=c(1.5,2.5,2,0.1))
-```
-
-The spline plots are fitted per gene in the above plot and illustrate that
-the genes assigned to a cluster have differing degrees to which their data
-fits that of the cluster centroid, and indeed fits that of their individual
-spline fits.
-
-We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids
-
-```{r}
-de_moanin_rescale = de_moanin_model
-assay(de_moanin_rescale) = rescale_values(data=assay(de_moanin_model))
-plot_splines_data(de_moanin_rescale,  
+plot_splines_data(de_moanin_model,  
     centroid=kmeans_clusters$centroids[cluster_of_interest,], 
@@
 ```
-**EAP** I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this.
-The `rescale_values` function is too sensitive to outliers.
-Would be better to rescale each centroid to scale of gene.
-If we had a helper function to do that, could easily add that to plot_splines_data (made issue on moanin), but probably right now should remove this.
 
@@
 
-Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+As we can see, while these genes have some similarity with the pattern of the
+cluster centroids, these particular genes are not the best examples of the
+cluster, in the sense of matching the centroid estimates. Because of this
+variability in how well the genes fit a cluster, we would like to be able to
+score how well a gene fits a cluster.
+
 Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
@@
 by `proportion_genes_to_label`). Note that because the distribution of scores
-of genes across *all* clusters are used, this is not equivalent to assigning
+of genes across *all* clusters is used, this is not equivalent to assigning
 50% of genes to a cluster -- it is possible that all genes are assigned to a
@@
 # We can give the previous calculated scores to `previous_scores` to save time
-unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
+unfiltered_scores = splines_kmeans_score_and_label(
+    moanin_model, kmeans_clusters,
+    proportion_genes_to_label=1,
+    previous_scores=scores_and_labels$scores)
 best_score = rowMin(scores_and_labels$scores)
@@
 Of the four genes we plotted before, `r sum(cluster2Genes[3:6] %in%
-genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest`. We can again look at a
+genes_in_cluster2)` of them remained assigned to cluster `r cluster_of_interest` after our scoring. We can again look at a
 few genes in cluster `r cluster_of_interest`, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})
@@
 cluster2Score = genes_in_cluster2[ord]
-dataToPlot = rbind(
-    assay(moanin_model[cluster2Score[1:4], ]),
-    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
 plot_splines_data(
-    moanin_model, data=dataToPlot, 
+    moanin_model, subset_data=cluster2Score[1:4], 
+    centroid=kmeans_clusters$centroids[cluster_of_interest,], 
     colors=ann_colors$Group, smooth=TRUE, simpleY=FALSE,
@@
 ```
+We see that, as expected, these genes are a much better match to the cluster centroids.
 
@@
 
-### Looking at specific clusters in details.
-
-Now, let us look more in detail some specific clusters. Cluster 6 seems
+### Looking at specific clusters in detail.
+
+Now, let us look more in detail at some specific clusters. Cluster 6 seems
 particularly interesting: it captures genes with strong differences between
@@
 
-```{r heatmapCluster8, fig.width=6, fig.height=8, fig.cap="Heatmap of genes in Cluster 8"}
+```{r heatmapCluster8, fig.width=6, fig.height=8, fig.cap="Heatmap of genes in Cluster 6"}
 layout(matrix(c(1,2),nrow=1), widths=c(1.5, 2))
@@
 
-Those two heatmaps demonstrate that the clustering method successfully cluster
-genes that are on different scales, and yet share the same dynamical response
-to the treatments.
+Those two heatmaps demonstrate that the clustering method successfully
+clusters genes that are on different scales, and yet share the same dynamical
+response to the treatments.
 
@@
 - **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
-  the elbow methods looks at the total with cluster sum of squares 
+  the elbow method looks at the total within cluster sum of squares 
   as a function of the number of clusters (WCSS). When adding clusters
@@
   clustering for every $k$ to a small randomization of the data. The user is
-  then invited to choose the number of cluster based on a number of similarity
+  then invited to choose the number of clusters based on a number of similarity
   measures.
@@
 will, for each clustering, conserve (1) with within cluster sum of
-squares; (2) the clustering assignement (or label) for each gene.
+squares; (2) the clustering assignment (or label) for each gene.
 
@@
 point, the WCSS will start decreasing more slowly, giving an angle or "elbow"
-in the graph.  The number of cluster is chosen at this "Elbow point."
+in the graph.  The number of clusters is chosen at this "Elbow point."
 
@@
 thus ensure that the biological interpretation of the clusters hold with
-perturbtation to the data. In addition, simulation where the data is generated
+perturbations to the data. In addition, simulation where the data is generated
 with a well defined $k$ show that the clustering is more stable for the
@@
 
-Most methods method to find the number of clusters with stability measures
+Most methods to find the number of clusters with stability measures
 only provide visual aids to guide the user. The first element often visualized
@@
 stores the proportion of clustering in which two items are clustered together.
-A perfect consensus matrix ordered such as each elements that belong to the
+A perfect consensus matrix ordered such as all elements that belong to the
 same cluster are adjacent to one another which show blocks along the diagonal
@@
 subsample_proportion = 0.8
-indices = sample(1:dim(de_moanin_model)[1], n_genes * subsample_proportion, replace=FALSE)
+indices = sample(1:dim(de_moanin_model)[1],
+		 n_genes * subsample_proportion,
+		 replace=FALSE)
 
@@
 
-Each column correspond to a bootstrap sample, each row to a gene, and each
+Each column corresponds to a bootstrap sample, each row to a gene, and each
 entry to the label found for that particular clustering. Thus, for each gene,
-we have an assignment to a cluster over the 20 resampling runs. For example,
-for $k=5$:
+we have an assignment to a cluster over the 25 resampling runs. For example,
+for the first 10 bootstrap samples and $k=5$:
 
 ```{r,echo=FALSE}
-kable(head(stability_5))
+pander(head(stability_5[, 1:10]))
 ```
@@
 is correct, the clustering results are stable to bootstrap resampling, as
-described above.  The distribution of similaries between bootstrapped results
+described above.  The distribution of similarities between bootstrapped results
 for each $k$ can thus be compared for different values of $k$ and guide the
@@
 explorer strategy, in addition to visualizing the diversity of the centroids,
-can thus help assessing an adequate number of clusters.
+can thus help assess an adequate number of clusters.
 
 Now, replot the same model explorer, but only for the clustering experiments
-$k=6$, $k=7$, $k=8$, $k=9$ and $k=10$ so that we can see more clearly the stability
+$k=6$, $k=7$, $k=8$, $k=9$, and $k=10$ so that we can see more clearly the stability
 measures in that range (Figure \ref{fig:plotModelExplorerZoom}).
@@
 instead of looking at the cumulative density of similarity measures of
-bootstrapped clustering, the authors suggests plotting the cumulative density
+bootstrapped clustering, the authors suggest plotting the cumulative density
 of elements of the consensus matrix. We provide a function
@@
 stable the clustering, the closer to 0 or 1 will be the entries of consensus
-matrix. The consensus clustering strategy thus suggest at looking at either
+matrix. The consensus clustering strategy thus suggests at looking at either
 the AUC as a function of the number of the clusters or the "improvement" in
-the AUC as a function of the number of cluster.
+the AUC as a function of the number of clusters.
 
@@
 a very stable clustering, it does not capture the range of gene expression
-patterns present in the data. This shows the limitation of such method on real
+patterns present in the data. This shows the limitation of such methods on real
 data, where the number of clusters is not clearly defined.
@@
 [@ensembl2020] based on the set of biological pathways described in the KEGG
-database[@KEGG].
+database [@KEGG].
 
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:min(10,nrow(pathways)), ], format="latex", escape=F, digits=4, booktabs=TRUE)
+pander(pathways[1:min(10,nrow(pathways)), ])
 ```
@@
 
-kable(go_terms_enriched[1:15, ], format="latex", escape=F,
-      digits=3, booktabs=TRUE, row.names=FALSE)
+pander(go_terms_enriched[1:15, ])
 ```
@@
 
-This workflow provides a tutoral for the analysis of lengthy time-course gene
-expression data in R using the package `moanin`, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
-exposed to different influenza strain and measured over time. 
-The proposed workflow covers consists of three common analysis main steps generally performed after quality
-control and normalization: (1) differential expression analysis; (2)
-clustering of time-course gene expression data; (3) downstream analysis of
-clusters. We demonstrate how the use of the package `moanin` allows for easy implementation of these procedures in the setting of time-course data.
+This workflow provides a tutorial for the analysis of lengthy time-course gene
+expression data in R using the package `moanin`, which aids in implementing
+common timecourse analyses. We illustrate the workflow through the analysis of
+mice lung tissue exposed to different influenza strains and measured over time.
+The proposed workflow consists of three common analysis main steps
+generally performed after quality control and normalization: (1) differential
+expression analysis; (2) clustering of time-course gene expression data; (3)
+downstream analysis of clusters. We demonstrate how the use of the package
+`moanin` allows for easy implementation of these procedures in the setting of
+time-course data.
 
@@
 
-All packages use in the workflow are available on GitHub, CRAN, or
+All packages used in the workflow are available on GitHub, CRAN, or
 Bioconductor.
@@
 [GSE63786](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786).
-Normalized data can be found in `moanin`.  Normalization information is
+Normalized data can be found in `timecoursedata`.  Normalization information is
 provided as supplementary information.

### 2020-07-30 scripts/install.R

@@ -98,4 +98,6 @@ if(!require(BiocWorkflowTools)){
     BiocManager::install("BiocWorkflowTools")
 }
 
-install_github("NelleV/moanin", dependencies=FALSE)
+if(!require("pander")){
+    install.packages('pander')
+}

### 2020-07-30 scripts/manuscript.Rmd
update to new moanin update on plot Centroid

--- 2020-07-29 scripts/manuscript.Rmd
@@
 ```
-**EAP** I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this.
-The `rescale_values` function is too sensitive to outliers.
-Would be better to rescale each centroid to scale of gene.
-If we had a helper function to do that, could easily add that to plot_splines_data (made issue on moanin), but probably right now should remove this.
 
 ### Assigning genes to clusters
-
-Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates.
+Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+
 Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would

### 2020-08-03 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-07-27 16:08:41.
+The manuscript was last compile on 2020-08-03 07:51:19.
 
@@
 \NormalTok{    kmeans_clusters}\OperatorTok{$}\NormalTok{clusters[kmeans_clusters}\OperatorTok{$}\NormalTok{clusters}\OperatorTok{==}\NormalTok{cluster_of_interest])}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
-    \KeywordTok{assay}\NormalTok{(de_moanin_model[cluster2Genes[}\DecValTok{3}\OperatorTok{:}\DecValTok{6}\NormalTok{],]), }
-    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,])}
-\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
-    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{simpleY=}\OtherTok{FALSE}\NormalTok{,}
-    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1}
-
-}
-
-\caption{Genes from cluster 2}\label{fig:genesVsCentroids}
-\end{figure}
-
-The spline plots are fitted per gene in the above plot and illustrate that
-the genes assigned to a cluster have differing degrees to which their data
-fits that of the cluster centroid, and indeed fits that of their individual
-spline fits.
-
-We can also overlay the centroid on the data, after rescaling the data to be between 0-1, like the centroids
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{de_moanin_rescale =}\StringTok{ }\NormalTok{de_moanin_model}
-\KeywordTok{assay}\NormalTok{(de_moanin_rescale) =}\StringTok{ }\KeywordTok{rescale_values}\NormalTok{(}\DataTypeTok{data=}\KeywordTok{assay}\NormalTok{(de_moanin_model))}
-\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_rescale,  }
+\KeywordTok{plot_splines_data}\NormalTok{(de_moanin_model,  }
     \DataTypeTok{centroid=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,], }
@@
 
-\begin{center}\includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-19-1} \end{center}
-
-\textbf{EAP} I added the functionality to plot centroid over data, rather than just fit to data, but I'm not sure it makes much sense to include this.
-The \texttt{rescale\_values} function is too sensitive to outliers.
-Would be better to rescale each centroid to scale of gene.
-If we had a helper function to do that, could easily add that to plot\_splines\_data (made issue on moanin), but probably right now should remove this.
+\begin{figure}[H]
+
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/genesVsCentroids-1}
+
+}
+
+\caption{Genes from cluster 2}\label{fig:genesVsCentroids}
+\end{figure}
 
@@
 
-Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates.
+Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+
 Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
@@
 
-Of the four genes we plotted before, 2 of them remained assigned to cluster 2. We can again look at a
+Of the four genes we plotted before, 2 of them remained assigned to cluster 2 after our scoring. We can again look at a
 few genes in cluster 2, only now pick the best scoring genes (Figure \ref{fig:topScoreInCluster2})
@@
 \NormalTok{cluster2Score =}\StringTok{ }\NormalTok{genes_in_cluster2[ord]}
-\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}
-    \KeywordTok{assay}\NormalTok{(moanin_model[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ]),}
-    \StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest, ])}
 \KeywordTok{plot_splines_data}\NormalTok{(}
-\NormalTok{    moanin_model, }\DataTypeTok{data=}\NormalTok{dataToPlot, }
+\NormalTok{    moanin_model, }\DataTypeTok{subset_data=}\NormalTok{cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], }
+    \DataTypeTok{centroid=}\NormalTok{kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[cluster_of_interest,], }
     \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{, }\DataTypeTok{simpleY=}\OtherTok{FALSE}\NormalTok{,}
@@
 \end{figure}
+
+We see that, as expected, these genes are a much better match to the cluster centroids.
 
@@
 ##  [3] ggfortify_0.4.10            timecoursedata_0.99.15     
-##  [5] moanin_0.99.0-901           SummarizedExperiment_1.18.2
+##  [5] moanin_0.99.0-902           SummarizedExperiment_1.18.2
 ##  [7] DelayedArray_0.14.1         matrixStats_0.56.0         

### 2020-08-11 scripts/manuscript.Rmd
add check on version number of moanin

@@
 ```{r eval=FALSE}
-BiocManager::install(c("NMF","ggfortify","topGO","biomaRt","KEGGprofile","BiocWorkflowTools"))
+BiocManager::install(
+    c("NMF", "ggfortify", "topGO", "biomaRt",
+      "KEGGprofile", "BiocWorkflowTools"))
 ```
@@
 - **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
-  the elbow methods looks at the total with cluster sum of squares 
+  the elbow methods looks at the total within cluster sum of squares 
   as a function of the number of clusters (WCSS). When adding clusters
@@
 ```{r,echo=FALSE}
-kable(head(stability_5))
+kable(head(stability_5)) %>%
+  kable_styling(font_size = 7)
 ```
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:min(10,nrow(pathways)), ], format="latex", escape=F, digits=4, booktabs=TRUE)
+kable(pathways[1:min(10,nrow(pathways)), ],
+      format="latex",
+      escape=F,
+      digits=4,
+      booktabs=TRUE) %>%
+  kable_styling(font_size = 7)
 ```

### 2020-08-11 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-08-03 07:51:19.
+The manuscript was last compile on 2020-08-11 06:50:22.
 
@@
 \begin{Highlighting}[]
-\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"NMF"}\NormalTok{,}\StringTok{"ggfortify"}\NormalTok{,}\StringTok{"topGO"}\NormalTok{,}\StringTok{"biomaRt"}\NormalTok{,}\StringTok{"KEGGprofile"}\NormalTok{,}\StringTok{"BiocWorkflowTools"}\NormalTok{))}
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}
+    \KeywordTok{c}\NormalTok{(}\StringTok{"NMF"}\NormalTok{, }\StringTok{"ggfortify"}\NormalTok{, }\StringTok{"topGO"}\NormalTok{, }\StringTok{"biomaRt"}\NormalTok{,}
+      \StringTok{"KEGGprofile"}\NormalTok{, }\StringTok{"BiocWorkflowTools"}\NormalTok{))}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-- `scores`: the matrix of shape `n_cluster` x `n_genes`,
+\begin{itemize}
+\tightlist
+\item
+  \texttt{scores}: the matrix of shape \texttt{n\_cluster} x \texttt{n\_genes},
   containing for each gene and each cluster, the goodness of fit as
   described above.
-- `labels`: the labels for all of the genes with a
+\item
+  \texttt{labels}: the labels for all of the genes with a
   sufficiently good goodness-of-fit score.
-- `score_cutoff`: the cutoff used on `scores` to determine whether to 
+\item
+  \texttt{score\_cutoff}: the cutoff used on \texttt{scores} to determine whether to
   assign a label
-\end{verbatim}
+\end{itemize}
 
@@
 \CommentTok{# We can give the previous calculated scores to `previous_scores` to save time}
-\NormalTok{unfiltered_scores<-}\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(moanin_model,kmeans_clusters,}\DataTypeTok{proportion_genes_to_label=}\DecValTok{1}\NormalTok{, }\DataTypeTok{previous_scores=}\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
+\NormalTok{unfiltered_scores =}\StringTok{ }\KeywordTok{splines_kmeans_score_and_label}\NormalTok{(}
+\NormalTok{    moanin_model, kmeans_clusters,}
+    \DataTypeTok{proportion_genes_to_label=}\DecValTok{1}\NormalTok{,}
+    \DataTypeTok{previous_scores=}\NormalTok{scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
 \NormalTok{best_score =}\StringTok{ }\KeywordTok{rowMin}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores)}
@@
 
-\caption{Heatmap of genes in Cluster 8}\label{fig:heatmapCluster8}
+\caption{Heatmap of genes in Cluster 6}\label{fig:heatmapCluster8}
 \end{figure}
@@
   \textbf{Elbow method}. First introduced in 1953 by Thorndike \citep{thorndike:who},
-  the elbow methods looks at the total with cluster sum of squares
+  the elbow methods looks at the total within cluster sum of squares
   as a function of the number of clusters (WCSS). When adding clusters
@@
 \NormalTok{subsample_proportion =}\StringTok{ }\FloatTok{0.8}
-\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{], n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion, }\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
+\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(de_moanin_model)[}\DecValTok{1}\NormalTok{],}
+\NormalTok{         n_genes }\OperatorTok{*}\StringTok{ }\NormalTok{subsample_proportion,}
+         \DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{)}
 
@@
 
+\begin{table}[H]
+\centering\begingroup\fontsize{7}{9}\selectfont
+
 \begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
@@
 \end{tabular}
+\endgroup{}
+\end{table}
 
@@
 \end{Shaded}
+
+\begin{table}[H]
+\centering\begingroup\fontsize{7}{9}\selectfont
 

### 2020-08-19 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-08-11 06:50:22.
+The manuscript was last compile on 2020-08-19 18:14:47.
 
@@
 \midrule
-GO:0019395 & fatty acid oxidation & 55 & 54 & 44.78 & 0.00016 & 0.164\\
-GO:1905114 & cell surface receptor signaling pathway ... & 272 & 243 & 221.46 & 0.00021 & 0.179\\
-GO:0003002 & regionalization & 174 & 158 & 141.67 & 0.00049 & 0.358\\
-GO:0009755 & hormone-mediated signaling pathway & 82 & 77 & 66.76 & 0.00105 & 0.671\\
+GO:0003002 & regionalization & 175 & 159 & 142.49 & 0.00044 & 0.375\\
+GO:0019395 & fatty acid oxidation & 58 & 56 & 47.22 & 0.00065 & 0.475\\
+GO:0007224 & smoothened signaling pathway & 75 & 71 & 61.07 & 0.00083 & 0.530\\
 GO:0048562 & embryonic organ morphogenesis & 154 & 139 & 125.39 & 0.00175 & 0.994\\
+GO:0009755 & hormone-mediated signaling pathway & 76 & 71 & 61.88 & 0.00254 & 1.000\\
 \addlinespace
 GO:0006805 & xenobiotic metabolic process & 50 & 48 & 40.71 & 0.00257 & 1.000\\
-GO:0090596 & sensory organ morphogenesis & 138 & 124 & 112.36 & 0.00332 & 1.000\\
-GO:0035108 & limb morphogenesis & 74 & 69 & 60.25 & 0.00337 & 1.000\\
-GO:0030166 & proteoglycan biosynthetic process & 27 & 27 & 21.98 & 0.00385 & 1.000\\
-GO:0003206 & cardiac chamber morphogenesis & 72 & 67 & 58.62 & 0.00448 & 1.000\\
+GO:0006790 & sulfur compound metabolic process & 155 & 138 & 126.20 & 0.00317 & 1.000\\
+GO:0060173 & limb development & 90 & 83 & 73.28 & 0.00325 & 1.000\\
+GO:0018394 & peptidyl-lysine acetylation & 82 & 76 & 66.77 & 0.00336 & 1.000\\
+GO:0030166 & proteoglycan biosynthetic process & 27 & 27 & 21.98 & 0.00386 & 1.000\\
 \addlinespace
-GO:0055114 & oxidation-reduction process & 536 & 472 & 436.41 & 0.00489 & 1.000\\
-GO:0021954 & central nervous system neuron developmen... & 36 & 35 & 29.31 & 0.00557 & 1.000\\
-GO:0006261 & DNA-dependent DNA replication & 70 & 64 & 56.99 & 0.00595 & 1.000\\
-GO:0006633 & fatty acid biosynthetic process & 78 & 70 & 63.51 & 0.00686 & 1.000\\
-GO:0006790 & sulfur compound metabolic process & 154 & 136 & 125.39 & 0.00710 & 1.000\\
+GO:0003206 & cardiac chamber morphogenesis & 72 & 67 & 58.62 & 0.00449 & 1.000\\
+GO:0006633 & fatty acid biosynthetic process & 88 & 79 & 71.65 & 0.00580 & 1.000\\
+GO:0006261 & DNA-dependent DNA replication & 71 & 65 & 57.81 & 0.00595 & 1.000\\
+GO:0090596 & sensory organ morphogenesis & 139 & 124 & 113.18 & 0.00637 & 1.000\\
+GO:0055114 & oxidation-reduction process & 541 & 475 & 440.49 & 0.00651 & 1.000\\
 \bottomrule

### 2020-08-31 scripts/manuscript.Rmd
Packages are now (almost) on bioconductor

@@
 
- `moanin` is available from GitHub, and can be installed using the `install_github` function in the package `devtools`, along with the corresponding package that contains time course datasets we will use:
+`moanin` and `timecoursedata` are available from bioconductor, and can be
+installed using the `install` function in the package `BiocManager`, along
+with the corresponding package that contains time course datasets we will use:
 
 ```{r eval=FALSE}
-library(devtools)
-install_github("NelleV/moanin")
-install_github("NelleV/timecoursedata")
+library(BiocManager)
+# Need to use the development version for now.
+BiocManager::install(version="devel")  BiocManager::valid()
+
+BiocManager::install("timecoursedata")
+BiocManager::install("moanin")
 ```
@@
     c("NMF", "ggfortify", "topGO", "biomaRt",
-      "KEGGprofile", "BiocWorkflowTools"))
+      "KEGGprofile", "BiocWorkflowTools", "timecoursedata", ))
 ```

### 2020-08-31 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-The manuscript was last compile on 2020-08-19 18:14:47.
-
 \hypertarget{introduction}{%
@@
 
-\texttt{moanin} is available from GitHub, and can be installed using the \texttt{install\_github} function in the package \texttt{devtools}, along with the corresponding package that contains time course datasets we will use:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(devtools)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/moanin"}\NormalTok{)}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"NelleV/timecoursedata"}\NormalTok{)}
+\texttt{moanin} and \texttt{timecoursedata} are available from bioconductor, and can be
+installed using the \texttt{install} function in the package \texttt{BiocManager}, along
+with the corresponding package that contains time course datasets we will use:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(BiocManager)}
+\CommentTok{# Need to use the development version for now.}
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\DataTypeTok{version=}\StringTok{"devel"}\NormalTok{)}
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{valid}\NormalTok{()}
+
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\StringTok{"timecoursedata"}\NormalTok{)}
+\NormalTok{BiocManager}\OperatorTok{::}\KeywordTok{install}\NormalTok{(}\StringTok{"moanin"}\NormalTok{)}
 \end{Highlighting}
@@
     \KeywordTok{c}\NormalTok{(}\StringTok{"NMF"}\NormalTok{, }\StringTok{"ggfortify"}\NormalTok{, }\StringTok{"topGO"}\NormalTok{, }\StringTok{"biomaRt"}\NormalTok{,}
-      \StringTok{"KEGGprofile"}\NormalTok{, }\StringTok{"BiocWorkflowTools"}\NormalTok{))}
+      \StringTok{"KEGGprofile"}\NormalTok{, }\StringTok{"BiocWorkflowTools"}\NormalTok{, }\StringTok{"timecoursedata"}\NormalTok{, ))}
 \end{Highlighting}
@@
 avian influenza virus (A/Vietnam/1203/2004 {[}H5N1{]}. Mice were injected with
-\(10^5\) PFU of each virus. An adidtional 42 mice were injected with a lower dose
+\(10^5\) PFU of each virus. An additional 42 mice were injected with a lower dose
 of the Vietnam avian influenza virus (\(10^3\) PFU).
@@
 ##  [1] RColorBrewer_1.1-2          BiocWorkflowTools_1.14.0   
-##  [3] ggfortify_0.4.10            timecoursedata_0.99.15     
+##  [3] ggfortify_0.4.10            timecoursedata_0.99.17     
 ##  [5] moanin_0.99.0-902           SummarizedExperiment_1.18.2

### 2020-08-31 scripts/manuscript.Rmd
Small changes to facilitate Rmd to docx conversion

@@
     %\VignetteEngine{knitr::rmarkdown}
-output: BiocWorkflowTools::f1000_article
+output:
+    BiocWorkflowTools::f1000_article: default
+    ms_document: default
 bibliography: bibliography.bib
+always_allow_html: yes
 ---
@@
 # Need to use the development version for now.
-BiocManager::install(version="devel")  BiocManager::valid()
+BiocManager::install(version="devel")
+BiocManager::valid()
 
@@
 ```{r display_lfc, echo=FALSE}
-kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
+pander(log_fold_change_timecourse[1:5, ])
 ```
@@
 ```{r,echo=FALSE}
-kable(head(stability_5)) %>%
-  kable_styling(font_size = 7)
+pander(head(stability_5))
 ```
@@
 pathways["Percentage"] = pathways["Percentage"] * 100
-kable(pathways[1:min(10,nrow(pathways)), ],
-      format="latex",
-      escape=F,
-      digits=4,
-      booktabs=TRUE) %>%
-  kable_styling(font_size = 7)
+pander(pathways[1:min(10,nrow(pathways)), ])
 ```
@@
 This workflow provides a tutoral for the analysis of lengthy time-course gene
-expression data in R using the package `moanin`, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
-exposed to different influenza strain and measured over time. 
-The proposed workflow covers consists of three common analysis main steps generally performed after quality
-control and normalization: (1) differential expression analysis; (2)
-clustering of time-course gene expression data; (3) downstream analysis of
-clusters. We demonstrate how the use of the package `moanin` allows for easy implementation of these procedures in the setting of time-course data.
+expression data in R using the package `moanin`, which aids in implementing
+common timecourse analyses. We illustrate the workflow through the analysis of
+mice lung tissue exposed to different influenza strain and measured over time.
+The proposed workflow covers consists of three common analysis main steps
+generally performed after quality control and normalization: (1) differential
+expression analysis; (2) clustering of time-course gene expression data; (3)
+downstream analysis of clusters. We demonstrate how the use of the package
+`moanin` allows for easy implementation of these procedures in the setting of
+time-course data.
 

### 2020-08-31 scripts/install.R

@@ -99,3 +99,5 @@ if(!require(BiocWorkflowTools)){
 }
 
 install_github("NelleV/moanin", dependencies=FALSE)
+
+install.packages('pander')

### 2020-09-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
 \usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 
@@
 threshold of the virus concentration is exceeded in the lung. Once this
-threshold is exceeded, a strong inflammatory and cytokin production occurs.
+threshold is exceeded, a strong inflammatory and cytokine production occurs.
 This results provides evidence that the pathology response is non-linearly
@@
 
-\begin{tabular}{l|r|l|r|r}
-\hline
-  &  & Group & Replicate & Timepoint\\
-\hline
-GSM1557140 & 0 & K & 1 & 0\\
-\hline
-GSM1557141 & 1 & K & 2 & 0\\
-\hline
-GSM1557142 & 2 & K & 3 & 0\\
-\hline
-GSM1557143 & 3 & K & 1 & 12\\
-\hline
-GSM1557144 & 4 & K & 2 & 12\\
-\hline
-GSM1557145 & 5 & K & 3 & 12\\
-\hline
-\end{tabular}
+\begin{longtable}[]{@{}ccccc@{}}
+\toprule
+\begin{minipage}[b]{0.20\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\centering
+Group\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Replicate\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Timepoint\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557140}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557141}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557142}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+0\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557143}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557144}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\centering
+\textbf{GSM1557145}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\centering
+K\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+12\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 
-\begin{tabular}{lrrr}
+\begin{longtable}[]{@{}cccc@{}}
 \toprule
-  & M.0-VL.0\_pval & M.0-VL.0\_qval & M.0-VL.0\_lfc\\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\centering
+M.0-VL.0\_pval\strut
+\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\centering
+M.0-VL.0\_qval\strut
+\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\centering
+M.0-VL.0\_lfc\strut
+\end{minipage}\tabularnewline
 \midrule
-NM\_009912 & 0.273 & 0.475 & 2.202\\
-NM\_008725 & 0.851 & 0.924 & 2.028\\
-NM\_007473 & 0.088 & 0.219 & 0.678\\
-ENSMUST00000094955 & 0.008 & 0.037 & 1.186\\
-NM\_001042489 & 0.419 & 0.621 & 1.212\\
-\addlinespace
-NM\_008159 & 0.123 & 0.278 & 1.102\\
-NM\_001013813 & 0.057 & 0.159 & 0.630\\
-AK039774 & 0.015 & 0.059 & 0.637\\
-NM\_013782 & 0.875 & 0.938 & 0.594\\
-NM\_028622 & 0.767 & 0.876 & 1.648\\
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_009912}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.2728\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.4746\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+2.202\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_008725}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.8509\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.9243\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+2.028\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_007473}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.08804\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.2193\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.678\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{ENSMUST00000094955}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.008188\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.0368\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+1.186\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_001042489}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.4192\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.6208\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+1.212\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_008159}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.1232\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.2778\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+1.102\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_001013813}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.05653\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.1594\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.63\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{AK039774}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.01504\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.05902\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.6372\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_013782}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.8754\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.9377\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.594\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_028622}\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.7668\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+0.8759\strut
+\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\centering
+1.648\strut
+\end{minipage}\tabularnewline
 \bottomrule
-\end{tabular}
+\end{longtable}
 
@@
 
-\begin{tabular}{lrrrrr}
+\begin{longtable}[]{@{}cccccc@{}}
 \toprule
-  & M-K:0 & M-K:3 & M-K:6 & M-K:9 & M-K:12\\
+\begin{minipage}[b]{0.26\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\centering
+M-K:0\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-K:3\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-K:6\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-K:9\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-K:12\strut
+\end{minipage}\tabularnewline
 \midrule
-NM\_009912 & -0.5662398 & -0.2454676 & -0.4096932 & -0.3320085 & -0.0502552\\
-NM\_008725 & 0.9947059 & -3.4609546 & 0.6561484 & -1.2106329 & -0.1998796\\
-NM\_007473 & -0.5176866 & -0.3146529 & 0.2327960 & -0.3445581 & 0.0476160\\
-ENSMUST00000094955 & -0.2201803 & 0.2566677 & -0.0358376 & -0.0469726 & -0.0707469\\
-NM\_001042489 & -0.2475888 & -0.2043850 & -0.3428368 & -0.3930846 & -0.4821660\\
-\addlinespace
-NM\_008159 & -0.7466435 & -0.0678902 & -0.3072220 & -0.7067393 & -0.4966867\\
+\endhead
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{NM\_009912}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.5662\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.2455\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4097\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.332\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.05026\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{NM\_008725}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+0.9947\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-3.461\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.6561\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.211\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.1999\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{NM\_007473}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.5177\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3147\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.2328\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3446\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.04762\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{ENSMUST00000094955}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.2202\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.2567\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.03584\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.04697\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.07075\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{NM\_001042489}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.2476\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.2044\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3428\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3931\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4822\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.26\columnwidth}\centering
+\textbf{NM\_008159}\strut
+\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\centering
+-0.7466\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.06789\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3072\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.7067\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4967\strut
+\end{minipage}\tabularnewline
 \bottomrule
-\end{tabular}
+\end{longtable}
 
@@
 
-\begin{tabular}{lrrrr}
+\begin{longtable}[]{@{}ccccc@{}}
 \toprule
-  & M-K & M-C & M-VL & M-VH\\
+\begin{minipage}[b]{0.30\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-K\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-C\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-VL\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+M-VH\strut
+\end{minipage}\tabularnewline
 \midrule
-NM\_009912 & -0.409 & -0.753 & -1.025 & -1.420\\
-NM\_008725 & 1.575 & 2.008 & 2.543 & 2.618\\
-NM\_007473 & -0.416 & -0.335 & -0.213 & 0.364\\
-ENSMUST00000094955 & -0.113 & -0.376 & -0.156 & -0.309\\
-NM\_001042489 & -0.234 & -0.410 & 0.308 & 0.430\\
+\endhead
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_009912}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4091\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.7525\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.025\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-1.42\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_008725}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+1.575\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.008\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.543\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+2.618\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_007473}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4158\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3346\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.2127\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.3643\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{ENSMUST00000094955}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.1126\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3763\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.1558\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.3091\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.30\columnwidth}\centering
+\textbf{NM\_001042489}\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.2341\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+-0.4104\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.3085\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.4301\strut
+\end{minipage}\tabularnewline
 \bottomrule
-\end{tabular}
+\end{longtable}
 
@@
 
-\begin{table}[H]
-\centering\begingroup\fontsize{7}{9}\selectfont
-
-\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
-\hline
-  & B1 & B2 & B3 & B4 & B5 & B6 & B7 & B8 & B9 & B10 & B11 & B12 & B13 & B14 & B15 & B16 & B17 & B18 & B19 & B20 & B21 & B22 & B23 & B24 & B25\\
-\hline
-AA204140 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 1 & 3 & 3 & 1 & 5 & 1 & 1 & 4 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
-\hline
-AB008911 & 5 & 5 & 1 & 4 & 1 & 2 & 3 & 3 & 2 & 3 & 3 & 3 & 5 & 1 & 5 & 1 & 3 & 4 & 4 & 5 & 5 & 2 & 4 & 1 & 4\\
-\hline
-AB010313 & 3 & 3 & 1 & 4 & 1 & 2 & 4 & 3 & 2 & 5 & 3 & 1 & 5 & 3 & 3 & 2 & 3 & 5 & 4 & 2 & 5 & 3 & 3 & 4 & 2\\
-\hline
-AB010342 & 5 & 3 & 1 & 5 & 1 & 1 & 3 & 3 & 1 & 3 & 3 & 3 & 5 & 1 & 5 & 1 & 1 & 4 & 3 & 5 & 4 & 2 & 3 & 4 & 4\\
-\hline
-AB011473 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 2 & 5 & 2 & 5 & 5 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
-\hline
-AB099518 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 2 & 5 & 2 & 5 & 5 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
-\hline
-\end{tabular}
-\endgroup{}
-\end{table}
+\begin{longtable}[]{@{}ccccccccccccc@{}}
+\caption{Table continues below}\tabularnewline
+\toprule
+\begin{minipage}[b]{0.13\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B1\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B2\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B3\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B4\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B5\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B6\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B7\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B8\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B9\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B10\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B11\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B12\strut
+\end{minipage}\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.13\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B1\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B2\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B3\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B4\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B5\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B6\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B7\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B8\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B9\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B10\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B11\strut
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+B12\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AA204140}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AB008911}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AB010313}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AB010342}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AB011473}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.13\columnwidth}\centering
+\textbf{AB099518}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\begin{longtable}[]{@{}cccccccccccccc@{}}
+\toprule
+\begin{minipage}[b]{0.11\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B13\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B14\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B15\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B16\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B17\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B18\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B19\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B20\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B21\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B22\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B23\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B24\strut
+\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+B25\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AA204140}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AB008911}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AB010313}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AB010342}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+4\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AB011473}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{AB099518}\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 
-\begin{table}[H]
-\centering\begingroup\fontsize{7}{9}\selectfont
-
-\begin{tabular}{llrr}
+\begin{longtable}[]{@{}cccc@{}}
 \toprule
-  & Pathway & Percentage & Adj. p-value\\
+\begin{minipage}[b]{0.15\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.39\columnwidth}\centering
+Pathway\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\centering
+Percentage\strut
+\end{minipage} & \begin{minipage}[b]{0.18\columnwidth}\centering
+Adj. p-value\strut
+\end{minipage}\tabularnewline
 \midrule
-04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
-04620 & Toll-like receptor signaling pathway & 39 & 0\\
-04621 & NOD-like receptor signaling pathway & 45 & 0\\
-04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
-04630 & Jak-STAT signaling pathway & 28 & 0\\
-\addlinespace
-04210 & Apoptosis & 35 & 0\\
-04380 & Osteoclast differentiation & 31 & 0\\
-04622 & RIG-I-like receptor signaling pathway & 38 & 0\\
-05160 & Hepatitis C & 28 & 0\\
-04940 & Type I diabetes mellitus & 35 & 0\\
+\endhead
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04060}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Cytokine-cytokine receptor
+interaction\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+31\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1.204e-20\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04620}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Toll-like receptor signaling
+pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+39\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+4.38e-14\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04621}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+NOD-like receptor signaling
+pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+45\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+2.036e-11\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04623}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Cytosolic DNA-sensing pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+45\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+4.619e-11\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04630}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Jak-STAT signaling pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+28\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+2.516e-10\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04210}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Apoptosis\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+35\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+6.279e-10\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04380}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Osteoclast differentiation\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+31\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+6.279e-10\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04622}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+RIG-I-like receptor signaling
+pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+38\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1.127e-09\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{05160}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Hepatitis C\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+28\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+3.436e-09\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{04940}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\centering
+Type I diabetes mellitus\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\centering
+35\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1.158e-07\strut
+\end{minipage}\tabularnewline
 \bottomrule
-\end{tabular}
-\endgroup{}
-\end{table}
+\end{longtable}
 
@@
 
-\begin{tabular}{llrrrlr}
+\begin{longtable}[]{@{}ccccc@{}}
+\caption{Table continues below}\tabularnewline
 \toprule
-GO ID & Description & Annotated & Significant & Expected & P-value & Adj. p-value\\
+\begin{minipage}[b]{0.10\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+GO ID\strut
+\end{minipage} & \begin{minipage}[b]{0.34\columnwidth}\centering
+Description\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Annotated\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Significant\strut
+\end{minipage}\tabularnewline
 \midrule
-GO:0003002 & regionalization & 175 & 159 & 142.49 & 0.00044 & 0.375\\
-GO:0019395 & fatty acid oxidation & 58 & 56 & 47.22 & 0.00065 & 0.475\\
-GO:0007224 & smoothened signaling pathway & 75 & 71 & 61.07 & 0.00083 & 0.530\\
-GO:0048562 & embryonic organ morphogenesis & 154 & 139 & 125.39 & 0.00175 & 0.994\\
-GO:0009755 & hormone-mediated signaling pathway & 76 & 71 & 61.88 & 0.00254 & 1.000\\
-\addlinespace
-GO:0006805 & xenobiotic metabolic process & 50 & 48 & 40.71 & 0.00257 & 1.000\\
-GO:0006790 & sulfur compound metabolic process & 155 & 138 & 126.20 & 0.00317 & 1.000\\
-GO:0060173 & limb development & 90 & 83 & 73.28 & 0.00325 & 1.000\\
-GO:0018394 & peptidyl-lysine acetylation & 82 & 76 & 66.77 & 0.00336 & 1.000\\
-GO:0030166 & proteoglycan biosynthetic process & 27 & 27 & 21.98 & 0.00386 & 1.000\\
-\addlinespace
-GO:0003206 & cardiac chamber morphogenesis & 72 & 67 & 58.62 & 0.00449 & 1.000\\
-GO:0006633 & fatty acid biosynthetic process & 88 & 79 & 71.65 & 0.00580 & 1.000\\
-GO:0006261 & DNA-dependent DNA replication & 71 & 65 & 57.81 & 0.00595 & 1.000\\
-GO:0090596 & sensory organ morphogenesis & 139 & 124 & 113.18 & 0.00637 & 1.000\\
-GO:0055114 & oxidation-reduction process & 541 & 475 & 440.49 & 0.00651 & 1.000\\
+\endfirsthead
+\toprule
+\begin{minipage}[b]{0.10\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+GO ID\strut
+\end{minipage} & \begin{minipage}[b]{0.34\columnwidth}\centering
+Description\strut
+\end{minipage} & \begin{minipage}[b]{0.13\columnwidth}\centering
+Annotated\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\centering
+Significant\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{6}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0003002}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+regionalization\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+175\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+159\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{7}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0019395}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+fatty acid oxidation\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+58\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+56\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{8}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0007224}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+smoothened signaling pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+75\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+71\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{9}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0048562}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+embryonic organ morphogenesis\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+154\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+139\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{10}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0009755}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+hormone-mediated signaling
+pathway\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+76\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+71\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{11}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0006805}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+xenobiotic metabolic process\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+50\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+48\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{12}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0006790}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+sulfur compound metabolic
+process\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+155\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+138\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{13}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0060173}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+limb development\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+90\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+83\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{14}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0018394}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+peptidyl-lysine acetylation\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+82\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+76\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{15}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0030166}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+proteoglycan biosynthetic
+process\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+27\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+27\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{16}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0003206}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+cardiac chamber morphogenesis\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+72\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+67\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{17}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0006633}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+fatty acid biosynthetic
+process\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+88\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+79\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{18}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0006261}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+DNA-dependent DNA replication\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+71\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+65\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{19}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0090596}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+sensory organ morphogenesis\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+139\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+124\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.10\columnwidth}\centering
+\textbf{20}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+\url{GO:0055114}\strut
+\end{minipage} & \begin{minipage}[t]{0.34\columnwidth}\centering
+oxidation-reduction process\strut
+\end{minipage} & \begin{minipage}[t]{0.13\columnwidth}\centering
+541\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\centering
+475\strut
+\end{minipage}\tabularnewline
 \bottomrule
-\end{tabular}
+\end{longtable}
+
+\begin{longtable}[]{@{}cccc@{}}
+\toprule
+\begin{minipage}[b]{0.11\columnwidth}\centering
+~\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\centering
+Expected\strut
+\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\centering
+P-value\strut
+\end{minipage} & \begin{minipage}[b]{0.18\columnwidth}\centering
+Adj. p-value\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{6}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+142.5\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00044\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+0.3749\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{7}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+47.22\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00065\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+0.4747\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{8}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+61.07\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00083\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+0.5304\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{9}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+125.4\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00175\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+0.994\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{10}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+61.88\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00254\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{11}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+40.71\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00257\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{12}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+126.2\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00317\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{13}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+73.28\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00325\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{14}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+66.77\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00336\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{15}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+21.98\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00386\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{16}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+58.62\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00449\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{17}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+71.65\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00580\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{18}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+57.81\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00595\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{19}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+113.2\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00637\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.11\columnwidth}\centering
+\textbf{20}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\centering
+440.5\strut
+\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\centering
+0.00651\strut
+\end{minipage} & \begin{minipage}[t]{0.18\columnwidth}\centering
+1\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 This workflow provides a tutoral for the analysis of lengthy time-course gene
-expression data in R using the package \texttt{moanin}, which aids in implementing common timecourse analyses. We illustrate the workflow through the analysis of mice lung tissue
-exposed to different influenza strain and measured over time.
-The proposed workflow covers consists of three common analysis main steps generally performed after quality
-control and normalization: (1) differential expression analysis; (2)
-clustering of time-course gene expression data; (3) downstream analysis of
-clusters. We demonstrate how the use of the package \texttt{moanin} allows for easy implementation of these procedures in the setting of time-course data.
+expression data in R using the package \texttt{moanin}, which aids in implementing
+common timecourse analyses. We illustrate the workflow through the analysis of
+mice lung tissue exposed to different influenza strain and measured over time.
+The proposed workflow covers consists of three common analysis main steps
+generally performed after quality control and normalization: (1) differential
+expression analysis; (2) clustering of time-course gene expression data; (3)
+downstream analysis of clusters. We demonstrate how the use of the package
+\texttt{moanin} allows for easy implementation of these procedures in the setting of
+time-course data.
 
@@
 \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63786}{GSE63786}.
-Normalized data can be found in \texttt{moanin}. Normalization information is
+Normalized data can be found in \texttt{timecoursedata}. Normalization information is
 provided as supplementary information.
@@
 ## [27] registry_0.5-1              ggplot2_3.3.2              
-## [29] kableExtra_1.1.0            biomaRt_2.44.1             
+## [29] pander_0.6.3                biomaRt_2.44.1             
 ## [31] BiocStyle_2.16.0            knitr_1.29                 
@@
 ##  [7] remotes_2.1.1          bit64_0.9-7.1          fansi_0.4.1           
-##  [10] ClusterR_1.2.2         KEGG.db_3.2.4          xml2_1.3.2            
-##  [13] codetools_0.2-16       doParallel_1.0.15      pkgload_1.1.0         
-##  [16] gridBase_0.4-7         dbplyr_1.4.4           png_0.1-7             
-##  [19] BiocManager_1.30.10    readr_1.3.1            compiler_4.0.2        
-##  [22] httr_1.4.1             backports_1.1.8        Matrix_1.2-18         
-##  [25] assertthat_0.2.1       cli_2.0.2              htmltools_0.5.0       
-##  [28] prettyunits_1.1.1      tools_4.0.2            gmp_0.6-0             
-##  [31] GenomeInfoDbData_1.2.3 gtable_0.3.0           glue_1.4.1            
-##  [34] reshape2_1.4.4         dplyr_1.0.0            rappdirs_0.3.1        
-##  [37] Rcpp_1.0.5             vctrs_0.3.1            Biostrings_2.56.0     
-##  [40] NMI_2.0                iterators_1.0.12       xfun_0.15             
-##  [43] stringr_1.4.0          ps_1.3.3               testthat_2.3.2        
-##  [46] rvest_0.3.5            lifecycle_0.2.0        gtools_3.8.2          
-##  [49] XML_3.99-0.4           edgeR_3.30.3           MASS_7.3-51.6         
-##  [52] zlibbioc_1.34.0        scales_1.1.1           hms_0.5.3             
-##  [55] yaml_2.2.1             curl_4.3               memoise_1.1.0         
-##  [58] gridExtra_2.3          TeachingDemos_2.12     stringi_1.4.6         
-##  [61] RSQLite_2.2.0          desc_1.2.0             foreach_1.5.0         
-##  [64] pkgbuild_1.1.0         bibtex_0.4.2.2         rlang_0.4.7           
-##  [67] pkgconfig_2.0.3        bitops_1.0-6           lattice_0.20-41       
-##  [70] evaluate_0.14          purrr_0.3.4            bit_1.1-15.2          
-##  [73] processx_3.4.3         tidyselect_1.1.0       plyr_1.8.6            
-##  [76] magrittr_1.5           bookdown_0.20          R6_2.4.1              
-##  [79] generics_0.0.2         DBI_1.1.0              pillar_1.4.6          
-##  [82] withr_2.2.0            KEGGREST_1.28.0        RCurl_1.98-1.2        
-##  [85] tibble_3.0.3           crayon_1.3.4           BiocFileCache_1.12.0  
-##  [88] progress_1.2.2         locfit_1.5-9.4         grid_4.0.2            
-##  [91] blob_1.2.1             callr_3.4.3            git2r_0.27.1          
-##  [94] digest_0.6.25          webshot_0.5.2          xtable_1.8-4          
-##  [97] tidyr_1.1.0            openssl_1.4.2          munsell_0.5.0         
-## [100] sessioninfo_1.1.1      askpass_1.1
+## [10] ClusterR_1.2.2         KEGG.db_3.2.4          codetools_0.2-16      
+## [13] doParallel_1.0.15      pkgload_1.1.0          gridBase_0.4-7        
+## [16] dbplyr_1.4.4           png_0.1-7              BiocManager_1.30.10   
+## [19] compiler_4.0.2         httr_1.4.1             backports_1.1.8       
+## [22] Matrix_1.2-18          assertthat_0.2.1       cli_2.0.2             
+## [25] htmltools_0.5.0        prettyunits_1.1.1      tools_4.0.2           
+## [28] gmp_0.6-0              gtable_0.3.0           glue_1.4.1            
+## [31] GenomeInfoDbData_1.2.3 reshape2_1.4.4         dplyr_1.0.0           
+## [34] rappdirs_0.3.1         Rcpp_1.0.5             vctrs_0.3.1           
+## [37] Biostrings_2.56.0      NMI_2.0                iterators_1.0.12      
+## [40] xfun_0.15              stringr_1.4.0          ps_1.3.3              
+## [43] testthat_2.3.2         lifecycle_0.2.0        gtools_3.8.2          
+## [46] XML_3.99-0.4           edgeR_3.30.3           MASS_7.3-51.6         
+## [49] zlibbioc_1.34.0        scales_1.1.1           hms_0.5.3             
+## [52] yaml_2.2.1             curl_4.3               memoise_1.1.0         
+## [55] gridExtra_2.3          TeachingDemos_2.12     stringi_1.4.6         
+## [58] RSQLite_2.2.0          desc_1.2.0             foreach_1.5.0         
+## [61] pkgbuild_1.1.0         bibtex_0.4.2.2         rlang_0.4.7           
+## [64] pkgconfig_2.0.3        bitops_1.0-6           evaluate_0.14         
+## [67] lattice_0.20-41        purrr_0.3.4            bit_1.1-15.2          
+## [70] processx_3.4.3         tidyselect_1.1.0       plyr_1.8.6            
+## [73] magrittr_1.5           bookdown_0.20          R6_2.4.1              
+## [76] generics_0.0.2         DBI_1.1.0              pillar_1.4.6          
+## [79] withr_2.2.0            KEGGREST_1.28.0        RCurl_1.98-1.2        
+## [82] tibble_3.0.3           crayon_1.3.4           BiocFileCache_1.12.0  
+## [85] progress_1.2.2         locfit_1.5-9.4         grid_4.0.2            
+## [88] blob_1.2.1             callr_3.4.3            git2r_0.27.1          
+## [91] digest_0.6.25          xtable_1.8-4           tidyr_1.1.0           
+## [94] openssl_1.4.2          munsell_0.5.0          sessioninfo_1.1.1     
+## [97] askpass_1.1
 \end{verbatim}

### 2020-09-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 variable genes: for this step, we keep only the top 50\% most variable genes.
+In Figure \ref{fig:geneVarHist}, we plot the distribution of variance across
+all genes.
 
@@
 The number of genes found differentially expressed ranges from around 12000 to
-29000 depending on the strain and dosage of influenza virus given to the mice.
+29000 depending on the strain and dosage of influenza virus given to the mice
+(Figure \ref{deTimeDist}).
 This corresponds to between 30\% to 70\% of the genes found differentially
@@
 
-We then use the \texttt{plot\_splines\_data} function, only now applied to the centroids, to visualize the centroids of
-each cluster obtained with the splines k-means model.
+We then use the \texttt{plot\_splines\_data} function, only now applied to the
+centroids, to visualize the centroids of each cluster obtained with the
+splines k-means model (Figure \ref{fig:centroids}).
 
@@
 
-{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/unnamed-chunk-18-1}
+{\centering \includegraphics[width=.5\textwidth,]{reports/manuscript_files/figure-latex/centroids-1}
 
@@
 
-\caption{K-means centroids}\label{fig:unnamed-chunk-18}
+\caption{K-means centroids}\label{fig:centroids}
 \end{figure}

### 2020-09-02 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 define a series of markers (or plotting symbols) to distinguish replicate
-samples in scatter plots. We also reorder the factor \texttt{Group} which describes the treatments so that the treatments are order from low to high pathogeny (with Control being first).
+samples in scatter plots. We also reorder the factor \texttt{Group} which describes
+the treatments so that the treatments are ordered from low to high pathogeny
+(with Control being first).
 
@@
 
-In both
-situations, we first need to set up a object (a \texttt{Moanin} object) to
+In both situations, we first need to set up an object (a \texttt{Moanin} object) to
 hold the meta data, as well as information for fitting the spline model
-(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin} object
-will contain information used
-throughout this analysis, in particular the condition and timepoints of each sample and the
-the basis matrix for fitting splines models.
+(formula, number of degrees of freedom of the splines, \ldots{}). The \texttt{Moanin}
+object will contain information used throughout this analysis, in particular
+the condition and timepoints of each sample and the basis matrix for
+fitting splines models.
 
@@
 functional modeling. The metadata \texttt{data.frame} object should contain at least
-two columns: one named \texttt{Group}, containing the treatement effect, and a
+two columns: one named \texttt{Group}, containing the treatment effect, and a
 second one named \texttt{Timepoint} containing the timepoint information.
@@
 
-In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
-the time-course data, for example, overall change versus the largest change.
+In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary
+of each of these methods against each other. We see that each method captures
+different elements of the time-course data, for example, overall change versus
+the largest change.
 
@@
 
-With a single measure of log-fold change and the p-value, we can now look
-at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
-of the control to the Kawasaki strain, using the ``timecourse'' method of calculating log fold change.
+With a single measure of log-fold change and the p-value, we can now look at
+the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example
+of a volcano plot for the comparison of the control to the Kawasaki strain,
+using the ``timecourse'' method of calculating log fold change.
 
@@
 To do this, we first
-aggregate all p-values obtained during the time-course differential expresison
+aggregate all p-values obtained during the time-course differential expression
 step in a single p-value using Fisher's method \citep{fisher:statistical} (\texttt{pvalues\_fisher\_method}).
@@
 clustering routine. As observed by looking at genes found differentially expressed,
-many genes share a similar gene expression pattern, but on different scale.
+many genes share a similar gene expression pattern, but on different scales.
 
@@
 
-As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates.
-Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+As we can see, while these genes have some similarity with the pattern of the
+cluster centroids, these particular genes are not the best examples of the
+cluster, in the sense of matching the centroid estimates. Because of this
+variability in how well the genes fit a cluster, we would like to be able to
+score how well a gene fits a cluster.
 
@@
 by \texttt{proportion\_genes\_to\_label}). Note that because the distribution of scores
-of genes across \emph{all} clusters are used, this is not equivalent to assigning
+of genes across \emph{all} clusters is used, this is not equivalent to assigning
 50\% of genes to a cluster -- it is possible that all genes are assigned to a
@@
 
-Now, let us look more in detail some specific clusters. Cluster 6 seems
+Now, let us look more in detail at some specific clusters. Cluster 6 seems
 particularly interesting: it captures genes with strong differences between
@@
   \textbf{Elbow method}. First introduced in 1953 by Thorndike \citep{thorndike:who},
-  the elbow methods looks at the total within cluster sum of squares
+  the elbow method looks at the total within cluster sum of squares
   as a function of the number of clusters (WCSS). When adding clusters
@@
   clustering for every \(k\) to a small randomization of the data. The user is
-  then invited to choose the number of cluster based on a number of similarity
+  then invited to choose the number of clusters based on a number of similarity
   measures.
@@
 point, the WCSS will start decreasing more slowly, giving an angle or ``elbow''
-in the graph. The number of cluster is chosen at this ``Elbow point.''
+in the graph. The number of clusters is chosen at this ``Elbow point.''
 
@@
 thus ensure that the biological interpretation of the clusters hold with
-perturbtation to the data. In addition, simulation where the data is generated
+perturbations to the data. In addition, simulation where the data is generated
 with a well defined \(k\) show that the clustering is more stable for the
@@
 
-Each column correspond to a bootstrap sample, each row to a gene, and each
+Each column corresponds to a bootstrap sample, each row to a gene, and each
 entry to the label found for that particular clustering. Thus, for each gene,
@@
 is correct, the clustering results are stable to bootstrap resampling, as
-described above. The distribution of similaries between bootstrapped results
+described above. The distribution of similarities between bootstrapped results
 for each \(k\) can thus be compared for different values of \(k\) and guide the
@@
 stable the clustering, the closer to 0 or 1 will be the entries of consensus
-matrix. The consensus clustering strategy thus suggest at looking at either
+matrix. The consensus clustering strategy thus suggests at looking at either
 the AUC as a function of the number of the clusters or the ``improvement'' in
-the AUC as a function of the number of cluster.
+the AUC as a function of the number of clusters.
 
@@
 
-This workflow provides a tutoral for the analysis of lengthy time-course gene
+This workflow provides a tutorial for the analysis of lengthy time-course gene
 expression data in R using the package \texttt{moanin}, which aids in implementing
 common timecourse analyses. We illustrate the workflow through the analysis of
-mice lung tissue exposed to different influenza strain and measured over time.
-The proposed workflow covers consists of three common analysis main steps
+mice lung tissue exposed to different influenza strains and measured over time.
+The proposed workflow consists of three common analysis main steps
 generally performed after quality control and normalization: (1) differential

### 2020-10-07 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

@@
 
-\hypertarget{looking-at-specific-clusters-in-details.}{%
-\subsubsection{Looking at specific clusters in details.}\label{looking-at-specific-clusters-in-details.}}
+\hypertarget{looking-at-specific-clusters-in-detail.}{%
+\subsubsection{Looking at specific clusters in detail.}\label{looking-at-specific-clusters-in-detail.}}
 
@@
 entry to the label found for that particular clustering. Thus, for each gene,
-we have an assignment to a cluster over the 20 resampling runs. For example,
-for \(k=5\):
-
-\begin{longtable}[]{@{}ccccccccccccc@{}}
-\caption{Table continues below}\tabularnewline
+we have an assignment to a cluster over the 25 resampling runs. For example,
+for the first 10 bootstrap samples and \(k=5\):
+
+\begin{longtable}[]{@{}ccccccccccc@{}}
 \toprule
-\begin{minipage}[b]{0.13\columnwidth}\centering
+\begin{minipage}[b]{0.15\columnwidth}\centering
 ~\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B1\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B2\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B3\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B4\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B5\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B6\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B7\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B8\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
 B9\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
+\end{minipage} & \begin{minipage}[b]{0.06\columnwidth}\centering
 B10\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
-B11\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
-B12\strut
-\end{minipage}\tabularnewline
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{0.13\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B1\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B2\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B3\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B4\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B5\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B6\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B7\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B8\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B9\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
-B10\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
-B11\strut
-\end{minipage} & \begin{minipage}[b]{0.05\columnwidth}\centering
-B12\strut
 \end{minipage}\tabularnewline
@@
 \endhead
-\begin{minipage}[t]{0.13\columnwidth}\centering
+\begin{minipage}[t]{0.15\columnwidth}\centering
 \textbf{AA204140}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 1\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
 3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{AB008911}\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
 \end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
@@
 \end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 3\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.13\columnwidth}\centering
-\textbf{AB008911}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{AB010313}\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+4\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{AB010342}\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+3\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{AB011473}\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 2\strut
 \end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
+2\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.15\columnwidth}\centering
+\textbf{AB099518}\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+2\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+1\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
 3\strut
@@
 \end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-3\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.13\columnwidth}\centering
-\textbf{AB010313}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
 2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
+5\strut
+\end{minipage} & \begin{minipage}[t]{0.06\columnwidth}\centering
 2\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-1\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.13\columnwidth}\centering
-\textbf{AB010342}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-3\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.13\columnwidth}\centering
-\textbf{AB011473}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-4\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.13\columnwidth}\centering
-\textbf{AB099518}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.05\columnwidth}\centering
-4\strut
-\end{minipage}\tabularnewline
-\bottomrule
-\end{longtable}
-
-\begin{longtable}[]{@{}cccccccccccccc@{}}
-\toprule
-\begin{minipage}[b]{0.11\columnwidth}\centering
-~\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B13\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B14\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B15\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B16\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B17\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B18\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B19\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B20\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B21\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B22\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B23\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B24\strut
-\end{minipage} & \begin{minipage}[b]{0.04\columnwidth}\centering
-B25\strut
-\end{minipage}\tabularnewline
-\midrule
-\endhead
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AA204140}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AB008911}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AB010313}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AB010342}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-3\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-4\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AB011473}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.11\columnwidth}\centering
-\textbf{AB099518}\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-5\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-2\strut
-\end{minipage} & \begin{minipage}[t]{0.04\columnwidth}\centering
-1\strut
 \end{minipage}\tabularnewline
@@
 explorer strategy, in addition to visualizing the diversity of the centroids,
-can thus help assessing an adequate number of clusters.
+can thus help assess an adequate number of clusters.
 
 Now, replot the same model explorer, but only for the clustering experiments
-\(k=6\), \(k=7\), \(k=8\), \(k=9\) and \(k=10\) so that we can see more clearly the stability
+\(k=6\), \(k=7\), \(k=8\), \(k=9\), and \(k=10\) so that we can see more clearly the stability
 measures in that range (Figure \ref{fig:plotModelExplorerZoom}).
@@
 \citep{ensembl2020} based on the set of biological pathways described in the KEGG
-database\citep{KEGG}.
+database \citep{KEGG}.
 
@@
 ##  [1] RColorBrewer_1.1-2          BiocWorkflowTools_1.14.0   
-##  [3] ggfortify_0.4.10            timecoursedata_0.99.17     
-##  [5] moanin_0.99.0-902           SummarizedExperiment_1.18.2
+##  [3] ggfortify_0.4.10            timecoursedata_0.99.18     
+##  [5] moanin_0.99.8               SummarizedExperiment_1.18.2
 ##  [7] DelayedArray_0.14.1         matrixStats_0.56.0         

### 2020-10-07 scripts/manuscript.Rmd
FIX loading the float latex package
Preparing for submission

@@
     %\VignetteEngine{knitr::rmarkdown}
-output:
-    BiocWorkflowTools::f1000_article: default
-    ms_document: default
+header-includes:
+   - \usepackage{float}
+output: rticles::arxiv_article
 bibliography: bibliography.bib
@@
 ---
+
+<!--
+
+Change output format to
+BiocWorkflowTools::f1000_article: default  for F1000Research layout
+ms_document: default for word document
+
+-->
 
@@
 ### Assigning genes to clusters
-As we can see, while these genes have some similarity with the pattern of the cluster centroids, these particularly genes are not the best examples of the cluster, in the sense of matching the centroid estimates.
-Because of this variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster.
+
+As we can see, while these genes have some similarity with the pattern of the
+cluster centroids, these particular genes are not the best examples of the
+cluster, in the sense of matching the centroid estimates. Because of this
+variability in how well the genes fit a cluster, we would like to be able to
+score how well a gene fits a cluster.
 
@@
 - **Elbow method**. First introduced in 1953 by Thorndike [@thorndike:who],
-  the elbow methods looks at the total within cluster sum of squares 
+  the elbow method looks at the total within cluster sum of squares 
   as a function of the number of clusters (WCSS). When adding clusters
@@
 
-Each column correspond to a bootstrap sample, each row to a gene, and each
+Each column corresponds to a bootstrap sample, each row to a gene, and each
 entry to the label found for that particular clustering. Thus, for each gene,
-we have an assignment to a cluster over the 20 resampling runs. For example,
-for $k=5$:
+we have an assignment to a cluster over the 25 resampling runs. For example,
+for the first 10 bootstrap samples and $k=5$:
 
 ```{r,echo=FALSE}
-pander(head(stability_5))
+pander(head(stability_5[, 1:10]))
 ```
@@
 
-This workflow provides a tutoral for the analysis of lengthy time-course gene
+This workflow provides a tutorial for the analysis of lengthy time-course gene
 expression data in R using the package `moanin`, which aids in implementing
 common timecourse analyses. We illustrate the workflow through the analysis of
-mice lung tissue exposed to different influenza strain and measured over time.
-The proposed workflow covers consists of three common analysis main steps
+mice lung tissue exposed to different influenza strains and measured over time.
+The proposed workflow consists of three common analysis main steps
 generally performed after quality control and normalization: (1) differential

### 2020-10-07 scripts/miscellaneous/normalization.Rmd
Preparing for submission

@@
     - name: Nelle Varoquaux
-      affiliation: Department of Statistics and Berkeley Institute for Data Science, University of California, Berkeley, Berkeley, CA 
+      affiliation: CNRS, University Grenoble Alpes, TIMC-IMAG, Grenoble, France 
     - name: Elizabeth Purdom

### 2020-10-07 .build_tools/travis/after_success.sh

@@ -1,0 +1,37 @@
+#!/bin/bash
+# This script is meant to be called by the "after_success" step defined in
+# .travis.yml. See https://docs.travis-ci.com/ for more details.
+
+# License: 3-clause BSD
+
+# Ok, now that we succeeded, I want this to be published somewhere not too
+# public. How can I do that?
+
+set -e
+
+
+# commands taken shamelessly from https://gist.github.com/willprice/e07efd73fb7f13f917ea
+setup_git() {
+  git config --global user.email "travis@travis-ci.org"
+  git config --global user.name "Travis CI"
+}
+
+commit_website_files() {
+  git add . *.html
+  git commit --message "Travis build: $TRAVIS_BUILD_NUMBER"
+}
+
+upload_files() {
+  git remote add origin-pages https://${GH_TOKEN}@github.com/NelleV/2019timecourse-rnaseq-pipeline_html_outputs.git > /dev/null 2>&1
+  git push --quiet --set-upstream origin-pages master 
+}
+
+if [ "$TRAVIS_PULL_REQUEST" = "false" ]; then
+    git clone https://${GH_TOKEN}@github.com/NelleV/2019timecourse-rnaseq-pipeline_html_outputs.git /tmp/html_outputs
+    cp -r scripts/reports/*.html /tmp/html_outputs
+    cp -r scripts/reports/*.pdf /tmp/html_outputs
+    cd /tmp/html_outputs
+    setup_git
+    commit_website_files
+    upload_files
+fi;

### 2020-10-07 .build_tools/travis/install.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+# This script is meant to be called by the "install" step defined in
+# .travis.yml. See http://docs.travis-ci.com/ for more details.
+# The behavior of the script is controlled by environment variabled defined
+# in the .travis.yml in the top level folder of the project.
+
+# License: 3-clause BSD
+
+set -e
+
+Rscript scripts/install.R

### 2020-10-07 .build_tools/travis/test_script.sh

@@ -1,0 +1,53 @@
+#!/bin/bash
+# This script is meant to be called by the "script" step defined in
+# .travis.yml. See http://docs.travis-ci.com/ for more details.
+# The behavior of the script is controlled by environment variabled defined
+# in the .travis.yml in the top level folder of the project.
+
+# License: 3-clause BSD
+
+set -e
+
+get_data() {
+    # Hopefully this gets cached
+    pushd data
+    make
+    popd
+}
+
+install_github_dependencies() {
+    # install_github keeps failing because of the Github API rate limit. So
+    # we're just going to do this by hand…
+    pushd /tmp
+    wget -P . https://github.com/NelleV/timecoursedata/archive/master.zip
+    unzip /tmp/master.zip
+    cd timecoursedata-master
+    make install-extra
+    make install
+    popd
+    rm -rf /tmp/master.zip
+
+    pushd /tmp
+    wget -P . https://github.com/NelleV/moanin/archive/master.zip
+    unzip /tmp/master.zip
+    cd moanin-master
+    make install-extra
+    make install
+    popd
+
+    Rscript scripts/install.R
+}
+
+run_tests() {
+    # first run the actual tests
+    pushd scripts
+    make all
+    popd
+}
+
+
+# Now get data and run the tests and build the manuscript
+# get_data
+install_github_dependencies
+run_tests
+mkdir -p scripts/reports

### 2020-10-07 scripts/.scripts/compile_rmarkdown.R

@@ -1,0 +1,7 @@
+library(rmarkdown)
+
+args = commandArgs(trailingOnly=TRUE)
+filename = args[1]
+outname = args[2]
+
+render(filename, output_format="html_document", output_file=outname)

### 2020-10-07 scripts/concatenate_clustering_stability_results.R

@@ -1,0 +1,27 @@
+
+
+
+folders = list.files("results/")
+folders = folders[grepl("stability", folders) & !grepl("tsv", folders)]
+
+for(folder in folders){
+    print(folder)
+    filenames = list.files(paste0("results/", folder))
+    all_labels = NULL
+    for(filename in filenames){
+	labels = unique(
+	    read.table(paste0("results/", folder, "/", filename),
+		       row.names=NULL))
+	colnames(labels) = c("Gene", "Label")
+	row.names(labels) = labels$Gene
+	labels = subset(labels, select=c("Label"))
+	all_labels = merge(all_labels, labels, by="row.names", all=TRUE)
+	row.names(all_labels) = all_labels$Row.names
+	all_labels = subset(all_labels, select=-c(Row.names))
+    }
+    colnames(all_labels) = lapply(
+	1:dim(all_labels)[2],
+	function(x){paste0("B", x)})
+    outname = paste0("results/", folder, ".tsv")
+    write.table(all_labels, outname, sep="\t")
+}

### 2020-10-07 scripts/install.R

@@ -1,0 +1,107 @@
+
+
+# This is a bit annoying, but I'm hoping to save time on travis-ci
+if(!require("devtools")){
+    install.packages("devtools")
+}
+
+if(!require(ggfortify)){
+    install.packages("ggfortify")
+}
+
+if(!require(kableExtra)){
+    install.packages("kableExtra")
+}
+
+if(!require(testthat)){
+    install.packages("testthat")
+}
+
+if(!require(viridis)){
+    install.packages("viridis")
+}
+
+if(!require(knitr)){
+    install.packages("knitr")
+}
+
+if(!require(rmarkdown)){
+    install.packages("rmarkdown")
+}
+
+if(!require(roxygen2)){
+    install.packages("roxygen2")
+}
+
+if(!require(ClusterR)){
+    install.packages("ClusterR")
+}
+
+if(!require(splines)){
+    install.packages("splines")
+}
+
+if(!require(NMF)){
+    install.packages("NMF")
+}
+
+if (!requireNamespace("BiocManager", quietly = TRUE))
+    install.packages("BiocManager")
+
+BiocManager::install("topGO")
+
+if(!require(limma)){
+    BiocManager::install("limma")
+}
+
+if(!require(edgeR)){
+    BiocManager::install("edgeR")
+}
+
+if(!require(edge)){
+    BiocManager::install("edge")
+}
+
+if(!require(EDASeq)){
+    BiocManager::install("EDASeq")
+}
+
+if(!require(topGO)){
+    BiocManager::install("topGO")
+}
+
+if(!require(Biobase)){
+    BiocManager::install("Biobase")
+}
+
+if(!require(MASS)){
+    BiocManager::install("MASS")
+}
+
+if(!require("biomaRt")){
+    BiocManager::install("biomaRt")
+}
+
+if(!require(KEGGprofile)){
+    BiocManager::install("KEGGprofile")
+}
+
+if(!require("NMI")){
+    install.packages("NMI")
+}
+
+if(!require(zoo)){
+    install.packages("zoo")
+}
+
+if(!require(BiocWorkflowTools)){
+    BiocManager::install("BiocWorkflowTools")
+}
+
+if(!require("pander")){
+    install.packages('pander')
+}
+
+if(!require(rticles)){
+    install.packages("rticles")
+}

### 2020-10-07 scripts/run_pathways.R

@@ -1,0 +1,69 @@
+library(devtools)
+library(limma)
+library(splines)
+library(stats)
+library(KEGGprofile)
+library(kableExtra)
+library(viridis)
+
+library(moanin)
+library(biomaRt)
+library(timecoursedata)
+
+df = 6
+
+# Preprocessing & filtering
+is_count = FALSE
+take_log = FALSE
+filter_expression = FALSE
+
+# Clustering
+filter_genes = TRUE
+n_clusters = 20
+percentage_genes_to_label = 0.5
+# Set data files and options
+
+data("shoemaker2015")
+data = shoemaker2015$data
+meta = shoemaker2015$meta
+
+labels = read.table("results/clustering_labels.txt", sep="\t",
+                    check.names=FALSE)
+
+gene_names = row.names(labels)
+gene_names = gene_names[!is.na(labels)]
+labels = labels[!is.na(labels)]
+
+
+# FIXME move this to the package
+ensembl = useMart("ensembl")
+ensembl = useDataset("mmusculus_gene_ensembl", mart=ensembl)
+
+clusters = sort(as.vector(unlist(unique(labels))))
+all_pathways = data.frame()
+for(cluster in clusters){
+    genes = gene_names[labels == cluster]
+
+    # convert gene names
+    genes = getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
+		  filters="refseq_mrna", values=genes,
+		  mart=ensembl)["entrezgene_id"]
+    genes = as.vector(unlist(genes))
+    pathways = KEGGprofile::find_enriched_pathway(
+	    genes, species="mmu",
+	    download_latest=TRUE)
+
+    if((dim(all_pathways)[1] == 0) & (dim(pathways$stastic)[1] != 0)){
+    	print("Initialize")
+	    pathways$stastic["Cluster"] = cluster
+	    all_pathways = pathways$stastic
+    }else if(dim(pathways$stastic)[1] != 0) {
+	    print("Merge")
+	    pathways$stastic["Cluster"] = cluster
+	    all_pathways = merge(all_pathways, pathways$stastic, all=TRUE)
+    }
+}
+
+
+outname = "results/all_pathways.tsv"
+write.table(all_pathways, outname, sep="\t")

### 2020-10-08 scripts/manuscript.Rmd
Switch to plain old pdf_document
Adding OSF link

@@
    - \usepackage{float}
-output: rticles::arxiv_article
+output: pdf_document
 bibliography: bibliography.bib
@@
 [https://github.com/NelleV/2019timecourse-rnaseq-pipeline](https://github.com/NelleV/2019timecourse-rnaseq-pipeline).
-Archived source code at the time of publication can be found at [XXX](XXX).
+Archived source code at the time of publication can be found at [https://osf.io/2a5tw/](https://osf.io/2a5tw/).
 

### 2021-02-05 scripts/manuscript.Rmd
More on supplementaries
pff…

@@
    - \usepackage{float}
-output: pdf_document
+output: word_document
 bibliography: bibliography.bib
@@
 # Define contrasts	
-contrasts = create_timepoints_contrasts(moanin_model,"M", "VL")
+contrasts = create_timepoints_contrasts(moanin_model, "M", "VL")
 ```
@@
 
-```{r showMissingDEGenes, echo=FALSE, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
+```{r showMissingDEGenes, echo=FALSE, fig.width=10, fig.height=12, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
 exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
